### 2020-06-08 docs/blog/blogposts/2020-06-04-welcome-to-collapse.Rmd
Add website.

---
title: "Welcome to collapse: Advanced and Fast Data Transformation in R"
author: "Sebastian Krantz"
date: '2020-06-08'
slug: welcome-to-collapse
categories: ["R"]
tags: ["collapse", "advanced", "fast", "transformation", "manipulation", "time-series", "multilevel", "panel", "weighted"]
---

<img src='collapse logo small.png' width="150px" align="right" />

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, fig.width = 8, fig.height = 5, out.width = '100%')

oldopts <- options(width = 101L)

library(data.table)
```

# Introduction

*collapse* was released on CRAN end of March. The current version 1.2.1 performs well on all operating systems, passing > 5600 unit tests covering all core functionality, thus a good point to start introducing the package to a larger audience. *collapse* has 2 main aims:

1. To facilitate complex data transformation and exploration tasks in R. 
*(In particular grouped and weighted statistical computations, advanced aggregation of multi-type data, advanced transformations of time-series and panel-data, and the manipulation of lists)*

2. To help make R code fast, flexible, parsimonious and programmer friendly. *(Provide order of magnitude performance improvements via extensive use of C++ and highly optimized R code, broad object orientation, and infrastructure for grouped programming)*

It can be installed in R using:

```{r, eval=FALSE}
install.packages('collapse')

# See Documentation
help('collapse-documentation')
```

With this post I want to formally introduce *collapse* and share some of the motivation and history of it. Then I will provide a basic demonstration of important features, and end with a small benchmark comparing *collapse* to *dplyr* and *data.table*. The key features and functions of the package are summarized in the figure below.

![*collapse* Core Functions](collapse header.png)

I start with the motivation (you can skip this if you like).

# Some Motivation and History

I work as an applied economist in the broader fields of macroeconomics, trade and development. As such I frequently encounter survey data such as household or enterprise surveys, longitudinal / panel data such as cross-country, trade, geospatial panels or panel-surveys, and multivariate macroeconomic data such as groups of monthly or quarterly economic time-series.

On this data I usually require various grouped computations, such as aggregating multi-type data (including groupwise-weighted statistics for survey data or trade-share weighted stuff), grouped scaling or centering panel data in preparation for some econometric techniques programmed by hand (sometimes with weights on a panel-survey), and various time-series stuff such as computing sequences of lags / leads, (iterated) differences, quasi-differences and growth rates / log-differences on large panel data (i.e. trade or geospatial panels) and large groups of macroeconomic time-series. I also found myself often faced with recursive extraction and unlisting / row-binding problems, particularly for multivariate time-series stuff like impulse responses from a VAR which I like to turn into a data.frame and visualize with *ggplot2*.

<!-- On this data I usually compute various grouped computations, such as aggregating multi-type survey data which requires groupwise-weighted statistics (i.e. weighted mean, or weighted mode for categorical data) or trade-share weighted stuff etc., grouped scaling or centering panel data (sometimes centering on multiple groups and sometimes with weights on a panel-survey) in preparation for some advanced econometric techniques I program by hand, and various time-series stuff such as computing sequences of lags / leags, (iterated) differences, quasi-differnces (for serial correlation issues) and growth rates / log-differences on large panel data (i.e. geospatial panels) and large groups of macroeconomic time-series (i.e. in preparation for dynamic factor analysis and related techniques). -->

<!-- With these demands I soon realized that the *tidyverse*, while in many ways very useful and instructive, was not really serving me well. Why? Because it is focussed on data.frames thus requiring conversion and often loss of attributes, it does not really support weighted or grouped and weighted computations, it is very limited in terms of transformations and time-computations, and the non-standard evaluation and piped syntax does not make for easy or efficient programming. With *data.table* I found more possibilities to do grouped and weighted computations, rolling statistics and some quite fast aggregations, but I soon also got dissatisfied with the requirement of converting everything to data.table first and the limited support for (fast) grouped transformation, time-computations and programming. I also discovered that the methods and classes offered by the *plm* package are very useful to manipulate, explore and program with panel data, but they are also not very broad in scope and quite slow as data grow large. Finally, I noticed that many statistical methods, particularly multivariate time-series stuff like Vector-Autoregressions etc., provide results in terms of (sometimes nested) lists of statistics matrices which I often wanted to turn into a data.frame for *ggplot*-ing or further analysis, and spent quite a lot of time doing that.  -->

With these demands I soon realized that neither the *tidyverse* nor *data.table* were servicing me very well. The reasons being the requirement of converting to data.frame/data.table/tibble, limited support for weighted computations, efficient time-series and panel-data transformations, and, (in my opinion) limited utility for programming because of non-standard evaluation and R overhead. I also found some object-oriented packages quite useful, especially *plm* for panel-data manipulation, but soon bumped into serious constraints regarding performance and scope of implementation. The same holds true for useful model-extraction packages like *broom*, *insight*, *parameters* etc. which don't support multivariate time-series stuff and lack some flexibility in the format and type of information extracted.

<!-- Finally, I noticed that many statistical methods, particularly multivariate time-series stuff like Vector-Autoregressions etc., provide important results in terms of (sometimes nested) lists of statistics matrices which I often wanted to turn into a data.frame for *ggplot*-ing or further analysis, and spent quite a lot of time doing that.  -->

<!-- , while in many ways very useful and instructive, was not really serving me well. Why? Because it is focussed on data.frames thus requiring conversion and often loss of attributes, it does not really support weighted or grouped and weighted computations, it is very limited in terms of transformations and time-computations, and the non-standard evaluation and piped syntax does not make for easy or efficient programming. With *data.table* I found more possibilities to do grouped and weighted computations, rolling statistics and some quite fast aggregations, but I soon also got dissatisfied with the requirement of converting everything to data.table first and the limited support for (fast) grouped transformation, time-computations and programming. I also discovered that the methods and classes offered by the *plm* package are very useful to manipulate, explore and program with panel data, but they are also not very broad in scope and quite slow as data grow large. Finally, I noticed that many statistical methods, particularly multivariate time-series stuff like Vector-Autoregressions etc., provide results in terms of (sometimes nested) lists of statistics matrices which I often wanted to turn into a data.frame for *ggplot*-ing or further analysis, and spent quite a lot of time doing that.  -->

Thus I came to the conclusion that if I wanted to take things further in R, I had to create my own data manipulation package with the following properties:

* A broad object-oriented approach with generic functions supporting vectors, matrices, data.frames and lists. The approach should preserve object classes and avoid unnecessary conversions and loss of attributes (such as attributes of time-series, data.frame's or columns in a data.frame).

<!-- and various useful classes such as time-series, panel-series / panel data.frames and grouped data.frames -->

* Support very fast grouped and weighted computations (aggregations, transformations, time-series and panel-data stuff and general replacing and sweeping out of statistics) on all those objects.

* Provide facilities for efficient (grouped) programming in R using standard evaluation (so that I could write panel-estimators etc. with a few lines of expressive code).

* Provide various useful functions that would facilitate frequent complex tasks, such as (weighted) aggregation of multi-type data, processing lists (i.e. recursive search and extraction from list-like objects and recursive unlisting to data.frame), and some essential tools such as panel-data statistics and autocorrelation functions, support for variable labels etc..

<!-- even packages like *data.table* have some R overhead slowing it down quite a bit on smaller datasets, and that  -->

These objectives synthesized into an ambitious project over time. I started about 2 years ago with the functions `collap` and `qsu` for multi-type data aggregation and panel-summary statistics. I soon discovered that my R code was somewhat complex and did not deliver the performance I had hoped for. For a while I thought about using *data.table* backend, but then I encountered *Rcpp* and studied *data.table* source a bit. I discovered that developing grouped statistical functions with *Rcpp* and combining them with *data.table*'s C-functions for fast radix-order based grouping could produce serially compiled code that was faster than *data.table* itself run with two threads on my laptop. Thus I decided to develop sets of grouped and weighted functions and transformation operators, which allowed a much faster and more flexible package than using a backend. I also discovered that I could optimize critical base R operations such as selecting columns or subsetting data.frames using efficient primitives and some C/C++ code. Thus performance became an objective and I aimed to

<!-- and service them with grouping objects based on *data.table*'s grouping C-code. This -->

* Develop the fastest serially compiled package for data manipulation in R, use C/C++ code whenever sensible and thoroughly optimize the R code.

<!-- (Possibly parallelism will be added later but it is not top of the agenda). -->

Then I thought that this could also become a quite useful add-on package for users of *dplyr* / *tidyverse*, *data.table* and *plm*. Thus I decided to integrate *collapse* into these programming environments:

<!-- To service them I chose to integrate *collapse* with these packages and allow for seamless harmony and coexistence, adding an additional objective. -->

* Offer seamless integration with *dplyr*, *data.table* and *plm* through support of relevant classes (*grouped_df*, *data.table*, *pseries*, *pdata.frame*), some extra methods and added non-standard evaluation features.

I also thought that the *collapse* API for grouped programming in R, with fast C++ based grouped functions serviced with grouping objects efficiently created and accessed in R, could be useful for other developers, both in R and C++. Thus I opted to make this all available to the user and document it. I will shortly write a post covering fast grouped programming with *collapse*.

* Make *collapse* maximally programmer / developer friendly, through making available fast grouping mechanisms, core methods and some utilities facilitating efficient grouped programming in R.

Regarding documentation, I do not really like the default R format of function-Rd pages supplemented by vignettes. I got inspired by *Mathematica* which has a beautiful structured documentation starting from a central overview page. Thus my final development objective:

* Create a structured, horizonatally and vertically integrated package documentation which can be called from the R console.

*collapse* also has some quite extensive vignettes supplementing the built-in documentation and discussing the integration with *dplyr* and *plm*. 
<!-- That's it up to this point in time. *collapse* 1.2.1 is ready to be used.  -->

# Demonstration

I will start by briefly demonstrating the *Fast Statistical Functions*, which are a central feature of *collapse*. Currently there are 13 of them (`fmean`, `fmedian`, `fmode`, `fsum`, `fprod`, `fsd`, `fvar`, `fmin`, `fmax`, `ffirst`, `flast`, `fNobs` and `fNdistinct`), they are all S3 generic and support fast grouped computations on vectors, matrices, data.frame's, lists and grouped tibble's (class *grouped_df*). Calling these functions on different objects yields simple column-wise statistical computations:

```{r, message=FALSE}
library(collapse)
v <- iris$Sepal.Length 
d <- num_vars(iris)    # Saving numeric variables
g <- iris$Species

# Simple statistics
fmean(v)              # vector
fmean(qM(d))          # matrix (qM is a faster as.matrix)
fmean(d)              # data.frame

# Preserving data structure
fmean(qM(d), drop = FALSE)     # still a matrix
fmean(d, drop = FALSE)         # still a data.frame
```
The functions `fmean`, `fsum`, `fprod`, `fmode`, `fvar` and `fsd` additionally support weights^[`fvar` and `fsd` compute frequency weights, the most common form of weighted sample variance. I am working on weights for `fmedian`, but still dealing with performance issues.].

```{r}
# Weighted statistics, similarly for vectors and matrices ...
w <- abs(rnorm(fnrow(iris)))
fmean(d, w = w)     
```

The second argument of these functions is called `g` and supports vectors or lists of grouping variables for grouped computations. For functions supporting weights, `w` is the third argument. I note that all further examples generalize to computations on vectors and matrices. 
<!-- it does not matter anymore on which type of object we are working.   -->

```{r}
# Grouped statistics
fmean(d, g)

# Groupwise-weighted statistics
fmean(d, g, w) 
```

Grouping becomes more efficient when factors or grouping objects are passed to `g`. Factors can efficiently be created using the function `qF`, and grouping objects are efficiently created with the function `GRP`^[Grouping objects are better for programming and for multiple grouping variables. This is demonstrated in the blog post on programming with *collapse*.] (more in a future post about programming). In addition, all functions support transformations through the `TRA` argument.

```{r}
# Simple Transformations
head(fvar(v, TRA = "replace"))  # replacing values with the overall variance
head(fsd(v, TRA = "/"))         # dividing by the overall standard-deviation (scaling)

# Grouped transformations
head(fvar(v, g, TRA = "replace"))  # replacing values with the group variance
head(fsd(v, g, TRA = "/"))         # grouped scaling
head(fmin(v, g, TRA = "-"))        # setting the minimum value in each species to 0
head(fsum(v, g, TRA = "/"))        # dividing by the sum (proportions)
head(fmedian(v, g, TRA = "-"))     # de-median
head(ffirst(v, g, TRA = "%%"))     # taking modulus of first group-value, etc ...

# Grouped and weighted transformations
head(fsd(v, g, w, "/"), 3)         # Weighted scaling
head(fmode(d, g, w, "replace"), 3) # replace with weighted statistical mode
```

Currently there are 10 different replacing or sweeping operations supported by `TRA`, see `?TRA`. `TRA` can also be called directly as a function which performs simple and grouped replacing and sweeping operations:

```{r}
head(TRA(v, fmedian(v), "-"))                        # Same as fmedian(v, TRA = "-")
head(TRA(d, BY(d, g, quantile, 0.05), "replace", g)) # Replace values with 5% percentile by species
```

I have just used the function `BY`, which is also generic for split-apply-combine computing with user-supplied functions. Another useful function is `dapply` (data-apply), which supports quite efficient column- and row-wise operations on matrices and data.frames using user-supplied functions.

<!-- I note that simple row-wise operations on data.frames like row-sums are best performed through efficient matrix conversion i.e. `rowSums(qM(d))` is better than `dapply(d, sum, MARGIN = 1)`.    -->

Some common panel-data transformations like between- and within-transformations (averaging and centering using the mean) are implemented slightly more memory efficient in the functions `fbetween` and `fwithin`. The function `fscale` also exists for fast (grouped, weighted) scaling and centering (standardizing) and mean-preserving scaling. These functions provide further options for data harmonization, such as centering on the overall data mean or scaling to the within-group standard deviation^[The within-group standard deviation is the standard deviation computed on the group-centered data.] (as shown below), as well as scaling / centering to arbitrary supplied means and standard deviations.

```{r, collapse=FALSE}
oldpar <- par(mfrow = c(1,3))
plot(get_vars(d, 1:2), col = g, main = "Raw Data")                      
plot(fwithin(get_vars(d, 1:2), g, mean = "overall.mean"), col = g, 
     main = "Centered on Overall Mean")
plot(fscale(get_vars(d, 1:2), g, mean = "overall.mean", sd = "within.sd"), col = g,    
     main = "Harmonized Mean and Variance")
par(oldpar)

```

The function `get_vars` is 2x faster than `[.data.frame`, attribute-preserving, and also supports column selection using functions or regular expressions. It's replacement version `get_vars<-` is 6x faster than `[<-.data.frame`. Apart from `fbetween` and `fwithin`, the functions `fHDbetween` and `fHDwithin` can average or center data on multiple groups, and they can also project out continuous variables alongside (i.e. they provide fitted values or residuals from regression problems which may or may not involve one or more factors).

For the manipulation of time-series and panel-series, *collapse* offers the functions `flag`, `fdiff` and `fgrowth`.

```{r}
head3 <- function(x) head(x, 3L)
head3(flag(EuStockMarkets, -1:1))      # A sequence of lags and leads
head3(fdiff(EuStockMarkets, 0:1, 1:2)) # First and second difference of each variable
```

I note that all attributes of the time-series matrix `EuStockMarkets` were preserved, the use of `head` just suppresses the print method.
<!-- At this point I will  -->
<!-- ```{r, eval=FALSE} -->
<!-- library(vars) -->
<!-- library(ggplot2) -->
<!-- library(data.table) # for melt function -->

<!-- frequency(EuStockMarkets) -->
<!-- VARselect(EuStockMarkets, type = "both", season = 260) -->
<!-- varmod <- VAR(EuStockMarkets, p = 7, type = "both", season = 260) -->
<!-- serial.test(varmod) -->
<!-- irf <- irf(varmod) -->
<!-- str(irf) -->
<!-- irfdata <- unlist2d(list_elem(irf), idcols = c("bound", "series"), row.names = "time", -->
<!--                     id.factor = TRUE, DT = TRUE) -->
<!-- head(irfdata) -->

<!-- melt(irfdata, 1:3) %>% ggplot(aes(x = time, y = value, colour = series, shape = bound)) + -->
<!--   geom_line() + facet_wrap("variable") -->

<!-- ``` -->

To take things a bit further, let's consider some multilevel / panel data:

```{r}
# World Bank World Development Data - supplied with collapse
head3(wlddev)
```
All variables in this data have labels stored in a 'label' attribute (the default if you import with *haven*). Variable labels can be accessed and set using `vlabels` and `vlabels<-`, and viewed together with names and classes using `namlab`. In general variable labels and other attributes will be preserved in when working with *collapse*. *collapse* provides some of the fastest and most advanced summary statistics:
```{r}
# Distinct value count
fNdistinct(wlddev)
# Use descr(wlddev) for a detailed description of each variable

# Checking for within-country variation
varying(wlddev, ~ iso3c)

# Panel-data statistics: Summarize GDP and GINI overall, between and within countries
qsu(wlddev, pid = PCGDP + GINI ~ iso3c, 
    vlabels = TRUE, higher = TRUE)

# All of that by region, only summarizing GDP, not computing higher moments
aperm(qsu(wlddev, by = ~ region,
                 pid = PCGDP ~ iso3c))[,, 1:2]

# Within-country correlations with p-value and observation count
pwcor(fwithin(get_vars(wlddev, 9:12), wlddev$iso3c), 
      N = TRUE, P = TRUE)

# Panel-data ACF: Efficient grouped standardizing and computing covariance with panel-lags
# (same as stats version, I might add psAcf to also give the forecast version)
psacf(wlddev, ~ iso3c, ~ year, cols = 9:12)
```

For fast grouped statistics we can keep programming in standard evaluation as before, or we can use piped expressions.

```{r, message=FALSE}
head3(fmean(get_vars(wlddev, 9:12), 
            get_vars(wlddev, c("region", "income"))))

`%>%` <- magrittr::`%>%` 
wlddev %>% fgroup_by(region, income) %>% 
  fselect(PCGDP:ODA) %>% fmean %>% head3
```

I note that the default is `na.rm = TRUE` for all *collapse* functions^[Missing values are efficiently skipped at C++ level with hardly any computational cost. This also pertains to missing values occurring in the weight vector. If `na.rm = FALSE`, execution will stop when a missing value is encountered, and `NA` is returned. This also speeds up computations compared to base R, particularly if some columns or some groups have missing values and others not. The fast functions also avoid `NaN`'s being created from computations involving `NA` values, and functions like `fsum` are well behaved (i.e. `fsum(NA)` gives `NA`, not `0` like `sum(NA, na.rm = TRUE)`, similarly for `fmin` and `fmax`).]  I also note that you can also use `dplyr::group_by` and `dplyr::select`, but `fgroup_by` and `fselect` are significantly faster (see benchmark). We can do a weighted aggregation using the variable `ODA` as weights using:

```{r}
# Weighted group mean: Weighted by ODA
wlddev %>% fgroup_by(region, income) %>% 
  fselect(PCGDP:ODA) %>% fmean(ODA) %>% head3
```

Note that in this case by default (`keep.w = TRUE`) the sum of the weights is also computed and saved. *collapse* also has its own data aggregation command called `collap`. It is in many ways more flexible, for example you can apply multiple functions to each column and efficiently reshape the result.

```{r}
collap(wlddev, by = ~ region + income, 
       FUN = list(fmean, fsd), cols = 9:12) %>% head3
collap(wlddev, by = ~ region + income, 
       FUN = list(fmean, fsd), cols = 9:12, return = "long") %>% head3

```

`collap` also supports flexible multi-type aggregation, here applying the mean to numeric and statistical mode to categorical data. The default (`keep.col.order = TRUE`) ensures that the data remains in the same order, and, when working with *Fast Statistical Functions*, all column attributes are preserved.
```{r}
# Applying the mean to numeric and the mode to categorical data (first 2 arguments are 'by' and 'FUN')
collap(wlddev, ~ iso3c + decade, fmean, catFUN = fmode) %>% head3

# Same as a piped call.. without column reordering 
wlddev %>% fgroup_by(iso3c, decade) %>% {
  add_vars(fmode(cat_vars(.)), fmean(get_vars(., 9:12))) # cat_vars selects non-numeric (categorical) columns
} %>% head3

# Adding weights: weighted mean and weighted mode (catFUN is 3rd argument) 
collap(wlddev, ~ iso3c + decade, fmean, fmode, w = ~ ODA) %>% head3

# Fully custom aggregation (also possible with weights)
collap(wlddev, ~ iso3c + decade, 
            custom = list(fmean = 9:12, 
                          fmax = 9:10, 
                          flast = cat_vars(wlddev, "indices"),
                          fmode = "GINI")) %>% head3

```
When aggregating with multiple functions, you can parallelize over them (internally done with `parallel::mclapply`).

Time computations on panel-data are also performed fast and simple.

```{r}
# Panel Lag and lead of PCGDP and LIFEEX
L(wlddev, -1:1, PCGDP + LIFEEX ~ iso3c, ~year) %>% head3

# Equivalent piped call
wlddev %>% fgroup_by(iso3c) %>% fselect(iso3c, year, PCGDP, LIFEEX) %>% 
  flag(-1:1, year) %>% head3

# Growth rates in percentage terms: 1 and 10-year
G(wlddev, c(1, 10), 1, ~ iso3c, ~year, cols = 9:12) %>% head3 # or use Dlog, or G(..., logdiff = TRUE) for percentages
```

Equivalently we can can compute leaded and suitably iterated (log-) differences and growth rates, as well as quasi-(log-)differences of the form $x_t - \rho x_{t-1}$. The operators `L` and `G` are shorthand's for the functions `flag` and `fgrowth` allowing formula input. Similar operators exist for `fwithin`, `fscale`, etc.

# Benchmark

For benchmarking we use some product-level trade data from the UN Comtrade database, processed by \href{https://tradestatistics.io/}{tadestatistics.io}.

```{r, eval=FALSE}
library(tradestatistics)
# US HS4-level trade from 2000 to 2018
us_trade <- ots_create_tidy_data(years = 2000:2018, 
                                 reporters = "usa", 
                                 table = "yrpc")
```
Downloading US product-level trade (HS4) from 2000 to 2018 gives about 2.6 million observations:
```{r, echo=FALSE}
load("C:/Users/Sebastian Krantz/Documents/R/collapse/docs/blog/us_trade.RData")
settransform(us_trade, id = NULL, export_value_usd2 = NULL, import_value_usd2 = NULL)
```
```{r}
fdim(us_trade)
head3(us_trade)

# 19 years, 221 trading partners, 1222 products, unbalanced panel with product-time gaps...
qDF(rbind(class = vclasses(us_trade), 
          Ndistinct = fNdistinct(us_trade)))

# Summarizing data between and within partner-product pairs
qsu(us_trade, pid = export_value_usd + import_value_usd ~ partner_iso + product_code)
```
It would also be interesting to summarize the trade flows for each partner, but that would be too large to print to the console. We can however get the `qsu` output as a list of matrices: 
```{r}
# Doing all of that by partner - variance of flows between and within traded products for each partner
l <- qsu(us_trade, by = export_value_usd + import_value_usd ~ partner_iso, 
                   pid = ~ partner_iso + product_code, array = FALSE)
str(l)
```
Now with the function `unlist2d`, we can efficiently turn this into a tidy data.table:

```{r}
unlist2d(l, idcols = c("Variable", "Trans"), 
         row.names = "Partner", DT = TRUE)
```
If `l` were some statistical object we could first pull out relevant elements using `get_elem`, possibly process those elements using `rapply2d` and then apply `unlist2d` to get the data.frame (or data.table with `DT = TRUE`). These are the main *collapse* list-processing functions.

Now on to the benchmark. It is run on a Windows 8.1 laptop with a 2x 2.2 GHZ Intel i5 processor, 8GB DDR3 RAM and a Samsung 850 EVO SSD hard drive.

```{r, message=FALSE, warning=FALSE, error=FALSE}
library(microbenchmark)
library(dplyr)
library(data.table)
setDTthreads(2)     # Default setting for this machine

# Grouping (data.table:::forderv here does not compute the unique groups yet)
microbenchmark(collapse = fgroup_by(us_trade, partner_iso, group_code, year), 
               data.table = data.table:::forderv(us_trade, c("partner_iso", "group_code", "year"), retGrp = TRUE),
               dplyr = group_by(us_trade, partner_iso, group_code, year), times = 10)

# Sum
microbenchmark(collapse = collap(us_trade, export_value_usd + import_value_usd ~ partner_iso + group_code + year, fsum),
               data.table = us_trade[, list(export_value_usd = sum(export_value_usd, na.rm = TRUE),
                                            import_value_usd = sum(import_value_usd, na.rm = TRUE)),
                                      by = c("partner_iso", "group_code", "year")],
               dplyr = group_by(us_trade, partner_iso, group_code, year) %>%
                       dplyr::select(export_value_usd, import_value_usd) %>% summarise_all(sum, na.rm = TRUE), times = 10)

# Mean
microbenchmark(collapse = collap(us_trade, export_value_usd + import_value_usd ~ partner_iso + group_code + year, fmean),
               data.table = us_trade[, list(export_value_usd = mean(export_value_usd, na.rm = TRUE),
                                            import_value_usd = mean(import_value_usd, na.rm = TRUE)),
                                     by = c("partner_iso", "group_code", "year")],
               dplyr = group_by(us_trade, partner_iso, group_code, year) %>%
                 dplyr::select(export_value_usd, import_value_usd) %>% summarise_all(mean, na.rm = TRUE), times = 10)

# Replace with group-sum
microbenchmark(collapse = fgroup_by(us_trade, partner_iso, group_code, year) %>%
                 fselect(export_value_usd, import_value_usd) %>% fsum(TRA = "replace_fill"),
               data.table = us_trade[, `:=`(export_value_usd2 = sum(export_value_usd, na.rm = TRUE),
                                             import_value_usd2 = sum(import_value_usd, na.rm = TRUE)),
                                      by = c("partner_iso", "group_code", "year")],
               dplyr = group_by(us_trade, partner_iso, group_code, year) %>%
                 dplyr::select(export_value_usd, import_value_usd) %>% mutate_all(sum, na.rm = TRUE), times = 10)

# Centering, partner-product
microbenchmark(collapse = fgroup_by(us_trade, partner_iso, product_code) %>%
                 fselect(export_value_usd, import_value_usd) %>% fwithin,
               data.table = us_trade[, `:=`(export_value_usd2 = export_value_usd - mean(export_value_usd, na.rm = TRUE),
                                            import_value_usd2 = import_value_usd - mean(import_value_usd, na.rm = TRUE)),
                                     by = c("partner_iso", "group_code", "year")],
               dplyr = group_by(us_trade, partner_iso, group_code, year) %>%
                 dplyr::select(export_value_usd, import_value_usd) %>% mutate_all(function(x) x - mean(x, na.rm = TRUE)), times = 10)

# Lag
# Much better to sort data for dplyr
setorder(us_trade, partner_iso, product_code, year)
# We have an additional problem here: There are time-gaps within some partner-product pairs
tryCatch(L(us_trade, 1, export_value_usd + import_value_usd ~ partner_iso + product_code, ~ year), 
         error = function(e) e)
# The solution is that we create a unique id for each continuous partner-product sequence
settransform(us_trade, id = unattrib(seqid(year + unattrib(finteraction(partner_iso, product_code)) * 20L)))
# Notes: Normally id = seqid(year) would be enough on sorted data, but here we also have very different start and end dates...thus is some cases the next group starts the year after the former ends, checked with: table(with(us_trade, ffirst(fdiff(year), list(partner_iso, product_code))) == 1). unattrib removes all attributes, giving a plain integer id (just done for the benchmark here, otherwise collapse would have an advantage).

fNdistinct(us_trade$id) 
# Another nice comparison... fmode is similarly fast for categorical aggregation...
microbenchmark(fNdistinct(us_trade$id), n_distinct(us_trade$id))

# Here we go now: 
microbenchmark(collapse = L(us_trade, 1, export_value_usd + import_value_usd ~ id),
               collapse_ordered = L(us_trade, 1, export_value_usd + import_value_usd ~ id, ~ year),
               data.table = us_trade[, shift(.SD), keyby = id,
                                     .SDcols = c("export_value_usd","import_value_usd")],
               data.table_ordered = us_trade[order(year), shift(.SD), keyby = id,
                                     .SDcols = c("export_value_usd","import_value_usd")],
               dplyr = group_by(us_trade, id) %>% dplyr::select(export_value_usd, import_value_usd) %>%
                       mutate_all(lag), times = 10)

# Note: you can do ordered lags using mutate_all(lag, order_by = "year") for dplyr, but at computation times in excess of 90 seconds..

```

The benchmark shows that *collapse* is consistently very fast for the operations it provides (this also pertains to weighted computations which are only slightly slower than unweighted ones). But of course *collapse* cannot do a lot of things you can do with *dplyr* or *data.table* and vice-versa. It is and remains an advanced package, but I think it lives up to the high standards set forth by these packages. I am also highly indebted to *data.table* for inspiration and some vital bits of C-code. Feel free to get in touch for any suggestions or comments about *collapse*. I hope you will find it useful.

```{r, echo=FALSE}
options(oldopts)
```

### 2020-06-08 docs/blog/blogposts/Programming.Rmd
Add website.

---
title: "Programming"
author: "Sebastian Krantz"
date: "5 6 2020"
output: html_document
---

# Programming

```{r}
g <- GRP(wlddev, c("country", "decade"), return.order = TRUE)
g

head(add_vars(g[["groups"]], fmean(get_vars(wlddev, 9:12), g, use.g.names = FALSE)))
```

## Creating own grouped functions in C++

Here I show how one can use these tools

```{r}
str(g)
```

```{Rcpp, eval = FALSE}
// [[Rcpp::plugins(cpp11)]]
#include <Rcpp.h>
using namespace Rcpp;

template <int RTYPE>
LogicalVector ganyNACppImpl(Vector<RTYPE> x, int ng, IntegerVector g) {
  int l = x.size();
  if(l != g.size()) stop("length(x) must match length(g)");
  LogicalVector out(ng);

  if(RTYPE == REALSXP) { // numeric vector: all logical operations on NA/NaN evaluate to false, except != which is true.
    for(int i = 0; i != l; ++i) if(x[i] != x[i]) out[g[i]-1] = true;
  } else { // other vectors
    for(int i = 0; i != l; ++i) if(x[i] == Vector<RTYPE>::get_na()) out[g[i]-1] = true;
  }

  return out;
}

// disabling other types
template <>
LogicalVector ganyNACppImpl(Vector<CPLXSXP> x, int ng, IntegerVector) {
  stop("Not supported SEXP type!");
}

template <>
LogicalVector ganyNACppImpl(Vector<VECSXP> x, int ng, IntegerVector) {
  stop("Not supported SEXP type!");
}

template <>
LogicalVector ganyNACppImpl(Vector<RAWSXP> x, int ng, IntegerVector) {
  stop("Not supported SEXP type!");
}

template <>
LogicalVector ganyNACppImpl(Vector<EXPRSXP> x, int ng, IntegerVector) {
  stop("Not supported SEXP type!");
}

// [[Rcpp::export]]
LogicalVector ganyNACpp(const SEXP& x, int ng = 0, const IntegerVector& g = 0){
  RCPP_RETURN_VECTOR(ganyNACppImpl, x, ng, g);
}

```
On the R side things are then pretty simple: 
```{r, message=FALSE}
library(Rcpp)
sourceCpp("ganyNA.cpp")

ganyNA <- function(x, g, use.g.names = TRUE) {
  if(is.atomic(g) || !is.GRP(g)) g <- GRP(g, return.groups = use.g.names)
  if(use.g.names) setNames(ganyNACpp(x, g[[1L]], g[[2L]]), group_names.GRP(g, FALSE)) else
    ganyNACpp(x, g[[1L]], g[[2L]])
}

ganyNA(wlddev$PCGDP, wlddev$year)

# 10 million obs and 1 million groups, 1% of data missing
x <- na_insert(rnorm(1e7), prop = 0.01)
g <- sample.int(1e6, 1e7, TRUE)
system.time(ganyNA(x, g))
system.time(ganyNA(x, g, use.g.names = FALSE))

```

We could additionally add a `TRA` argument and then internally call the `TRA()` function to allow for replacing and sweeping out statistics, but this does not make much sense here.

### 2020-06-08 docs/blog/blogposts/examples.R

@@ -1,0 +1,98 @@
+library(tradestatistics)
+`%>%` <- magrittr::`%>%`
+us_trade <- ots_create_tidy_data(years = 2000:2018, reporters = "usa", table = "yrpc") %>%
+  ots_inflation_adjustment
+
+fdim(us_trade)
+
+fNdistinct(us_trade)
+qsu(with(us_trade, fNobs(year, list(partner_iso, product_code))))
+qsu(us_trade, by = export_value_usd + import_value_usd ~ year, pid = ~ partner_iso + year)
+us_trade_agg <- collap(us_trade, ~ partner_iso + group_code + year, fsum)
+qsu(with(us_trade_agg, fNobs(year, list(partner_iso, group_code))))
+
+L(us_trade_agg, 1, ~ partner_iso + group_code)
+
+settfm(us_trade_agg, id = finteraction(partner_iso, group_code))
+settfm(us_trade_agg, id = seqid(year, radixorder(partner_iso, group_code, year)))
+
+View(slt(us_trade_agg, id, year, t))
+
+View(L(us_trade_agg, 1, ~id, ~year))
+
+settransform(us_trade, t = seqid(year, radixorder(partner_iso, product_code, year)))
+View(slt(us_trade, year, t))
+
+us_trade <- ftransform(us_trade,
+             export_value_usd = replace_NA(export_value_usd, 0),
+             import_value_usd = replace_NA(import_value_usd, 0)) %>%
+             tfm(value_usd = export_value_usd + import_value_usd) %>%
+             tfm(partner_value_usd = fsum(value_usd, partner_iso, TRA = "replace_fill"))
+
+
+
+plot(fdiff(EuStockMarkets, 0:1))
+
+library(dplyr)
+library(data.table) # Running with 2 threads
+library(microbenchmark)
+suppressMessages(
+                microbenchmark(collapse = collap(us_trade, export_value_usd + import_value_usd ~ partner_iso + group_code + year, fsum),
+                               data.table = us_trade[, list(export_value_usd = sum(export_value_usd, na.rm = TRUE),
+                                                             import_value_usd = sum(import_value_usd, na.rm = TRUE)),
+                                                      by = c("partner_iso", "group_code", "year")],
+                               dplyr = group_by(us_trade, partner_iso, group_code, year) %>%
+                                       dplyr::select(export_value_usd, import_value_usd) %>% summarise_all(sum, na.rm = TRUE), times = 10))
+
+suppressMessages(
[…]

### 2020-08-10 docs/blog/blogposts/Programming.Rmd
Updating website for collapse 1.3.0.

@@
   if(is.atomic(g) || !is.GRP(g)) g <- GRP(g, return.groups = use.g.names)
-  if(use.g.names) setNames(ganyNACpp(x, g[[1L]], g[[2L]]), group_names.GRP(g, FALSE)) else
+  if(use.g.names) setNames(ganyNACpp(x, g[[1L]], g[[2L]]), GRPnames(g, FALSE)) else
     ganyNACpp(x, g[[1L]], g[[2L]])

### 2020-11-01 R/recode_replace.R
Updating for collapse 1.4.0.

@@ -11,7 +11,7 @@ recode_num <- function(X, ..., default = NULL, missing = NULL) {
   arglen <- length(args)
   missingl <- !is.null(missing)
   if(missingl && any(nam == missing))  warning(paste0("To improve performance missing values are replaced prior to recode, so this replaces all missing values with ",
-                                               missing, " and those are then again replaced with ", nam[nam == missing], ". If this is not desired, call replace_NA after recode with missing = NULL."))
+                                               missing, " and those are then again replaced with ", args[[which(nam == missing)]], ". If this is not desired, call replace_NA after recode with missing = NULL."))
   if(arglen == 1L) {
     args <- args[[1L]]
     if(is.null(default)) {

### 2020-11-01 R/zzz.R

@@ -1,7 +1,7 @@
 
 .onLoad <- function(libname, pkgname) {
 
-  res <- .Call(C_collapsedev2_init, "init.success")
+  res <- .Call(C_collapse_init, "init.success")
   if(!is.character(res) || res != "init.success") stop("collapse not succesfully loaded!")
 
   # https://stackoverflow.com/questions/12598242/global-variables-in-packages-in-r
@@ -43,16 +43,16 @@
 }
 
 .onAttach <- function(libname, pkgname) {
-  packageStartupMessage(paste0("collapsedev2 ",packageVersion("collapsedev2"),", see ?`collapse-package` or ?`collapse-documentation`\nNote: stats::D  ->  D.expression, D.call, D.name"))
+  packageStartupMessage(paste0("collapse ",packageVersion("collapse"),", see ?`collapse-package` or ?`collapse-documentation`\nNote: stats::D  ->  D.expression, D.call, D.name"))
 }
 
 .onUnload <- function (libpath) {
-  library.dynam.unload("collapsedev2", libpath)
+  library.dynam.unload("collapse", libpath)
 }
 
 # Note: To create local dev version of package change package name in DESCRIPTION, NAMESPACE, this file,
 # replace all instances of `_collapse_` in source files, and also rename `R_init_collapse` in ExportSymbols.cpp.
-
+# and in vignetter / Rd files replace library(collapse)
 release_questions <- function() {
   c(
     "Have you updated the version number in DESCRIPTION, NEWS.md, NEWS.Rd, cran.comments and .onAttach?",

### 2020-11-01 src/collapse.h

@@ -2,157 +2,157 @@
 using namespace Rcpp;
 
 // BWCpp
-RcppExport SEXP _collapsedev2_BWCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
+RcppExport SEXP _collapse_BWCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
 // BWmCpp
-RcppExport SEXP _collapsedev2_BWmCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
+RcppExport SEXP _collapse_BWmCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
 // BWlCpp
-RcppExport SEXP _collapsedev2_BWlCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
+RcppExport SEXP _collapse_BWlCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
 // TRACpp
-RcppExport SEXP _collapsedev2_TRACpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP);
[…]

### 2020-12-19 tests/testthat.R
Not running all code depending on suggested packages on CRAN, to avoid another fiasco as experienced with the archival of lfe.

@@ -1,7 +1,7 @@
 # rm(list = ls())
 # Sys.setenv("R_TESTS" = "")
 library(testthat)
-library(collapse)
+# library(collapse)
 
 test_check("collapse")
 

### 2020-12-19 tests/testthat/test-flm-fFtest.R

@@ -7,17 +7,21 @@ w <- mtcars$wt
 lmr <- lm(mpg ~ cyl + vs + am + carb + hp, mtcars)
 lmw <- lm(mpg ~ cyl + vs + am + carb + hp, weights = wt, mtcars)
 
+NCRAN <- identical(Sys.getenv("NCRAN"), "TRUE")
+
 test_that("flm works as intended", {
 
-  for(i in 1:6) expect_equal(drop(flm(y, x, add.icpt = TRUE, method = i)), coef(lmr))
-  for(i in 1:6) expect_equal(drop(flm(y, x, w, add.icpt = TRUE, method = i)), coef(lmw))
+  if(NCRAN) for(i in 1:6) expect_equal(drop(flm(y, x, add.icpt = TRUE, method = i)), coef(lmr))
+  if(NCRAN) for(i in 1:6) expect_equal(drop(flm(y, x, w, add.icpt = TRUE, method = i)), coef(lmw))
   expect_equal(flm(y, x, method = 1L, return.raw = TRUE), .lm.fit(x, y))
   expect_equal(flm(y, x, method = 2L, return.raw = TRUE), solve(crossprod(x), crossprod(x, y)))
-  expect_equal(flm(y, x, method = 3L, return.raw = TRUE), qr.coef(qr(x, LAPACK = TRUE), y))
-  expect_equal(flm(y, x, method = 4L, return.raw = TRUE), RcppArmadillo::fastLmPure(x, y))
+  expect_equal(flm(y, x, method = 3L, return.raw = TRUE), qr.coef(qr(x), y))
   expect_equal(flm(y, x, method = 5L, return.raw = TRUE), cinv(crossprod(x)) %*% crossprod(x, y))
-  expect_equal(flm(y, x, method = 6L, return.raw = TRUE), RcppEigen::fastLmPure(x, y, 3L))
-  for(i in 1:6) expect_visible(flm(y, x, w, method = i, return.raw = TRUE))
+  if(NCRAN) {
+    expect_equal(flm(y, x, method = 4L, return.raw = TRUE), RcppArmadillo::fastLmPure(x, y))
+    expect_equal(flm(y, x, method = 6L, return.raw = TRUE), RcppEigen::fastLmPure(x, y, 3L))
+  }
+  if(NCRAN) for(i in 1:6) expect_visible(flm(y, x, w, method = i, return.raw = TRUE))
[…]

### 2020-12-19 tests/testthat/test-roworder-colorder-rename.R

@@ -17,6 +17,8 @@ test_that("roworder works as intended", {
 
 })
 
+if(identical(Sys.getenv("NCRAN"), "TRUE")) {
+
 library(magrittr)
 test_that("colorder works as intended", {
   expect_identical(colorder(mtcars, vs, cyl:hp, am),
@@ -38,6 +40,8 @@ test_that("colorder works as intended", {
                    colorderv(mtcars, c("vs", "cyl|am"), regex = TRUE))
 
 })
+
+}
 
 test_that("frename works as intended", {
 

### 2021-06-03 R/BY.R
Function renaming: Eliminate '.' in function names and renaming `fNobs` -> `fnobs`, `fNdistinct` -> `fndistinct`, `pwNobs` -> `pwnobs`, `fHDwithin` -> `fhdwithin` `fHDbetween` -> `fhdbetween`.

@@ -10,8 +10,8 @@ BY.default <- function(x, g, FUN, ..., use.g.names = TRUE, sort = TRUE,
   if(!(is.function(FUN) || is.character(FUN))) stop("FUN needs to be a function")
   aplyfun <- if(parallel) function(...) mclapply(..., mc.cores = mc.cores) else lapply
   simplify <- switch(return[1L], same = 1L, vector = 2L, list = 3L, stop("BY.default only supports same, vector and list output!"))
-  if(!is.factor(g)) g <- if(is.GRP(g)) as.factor_GRP(g) else if(is.list(g))
-                         as.factor_GRP(GRP.default(g, sort = sort, call = FALSE)) else
+  if(!is.factor(g)) g <- if(is_GRP(g)) as_factor_GRP(g) else if(is.list(g))
+                         as_factor_GRP(GRP.default(g, sort = sort, call = FALSE)) else
                            qF(g, sort = sort, na.exclude = FALSE)
     res <- aplyfun(split(x, g), FUN, ...)
     if(simplify < 3L) {
@@ -51,8 +51,8 @@ BY.data.frame <- function(x, g, FUN, ..., use.g.names = TRUE, sort = TRUE,
   aplyfun <- if(parallel) function(...) mclapply(..., mc.cores = mc.cores) else lapply
   return <- switch(return[1L], same = 1L, matrix = 3L, data.frame = 2L, list = 0L,
                    stop("Unknown return option!"))
-  if(!is.factor(g)) g <- if(is.GRP(g)) as.factor_GRP(g) else if(is.list(g))
-                    as.factor_GRP(GRP.default(g, sort = sort, call = FALSE)) else
+  if(!is.factor(g)) g <- if(is_GRP(g)) as_factor_GRP(g) else if(is.list(g))
+                    as_factor_GRP(GRP.default(g, sort = sort, call = FALSE)) else
                       qF(g, sort = sort, na.exclude = FALSE)
   if(return != 0L) {
     ax <- attributes(x)
@@ -128,8 +128,8 @@ BY.matrix <- function(x, g, FUN, ..., use.g.names = TRUE, sort = TRUE,
   aplyfun <- if(parallel) function(...) parallel::mclapply(..., mc.cores = mc.cores) else lapply
   return <- switch(return[1L], same = 3L, matrix = 2L, data.frame = 1L, list = 0L,
                    stop("Unknown return option!"))
[…]

### 2021-06-03 R/TRA.R

@@ -15,7 +15,7 @@ TRA.default <- function(x, STATS, FUN = "-", g = NULL, ...) {
     }
     return(.Call(Cpp_TRA,x,STATS,g,TtI(FUN)))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   if(g[[1L]] != length(STATS)) stop("number of groups must match length(STATS)")
   .Call(Cpp_TRA,x,STATS,g[[2L]],TtI(FUN))
 }
@@ -32,7 +32,7 @@ TRA.matrix <- function(x, STATS, FUN = "-", g = NULL, ...) {
     }
     return(.Call(Cpp_TRAm,x,STATS,g,TtI(FUN)))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   if(g[[1L]] != nrow(STATS)) stop("number of groups must match nrow(STATS)")
   .Call(Cpp_TRAm,x,STATS,g[[2L]],TtI(FUN))
 }
@@ -49,7 +49,7 @@ TRA.data.frame <- function(x, STATS, FUN = "-", g = NULL, ...) {
     }
     return(.Call(Cpp_TRAl,x,STATS,g,TtI(FUN)))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   if(g[[1L]] != fnrow2(STATS)) stop("number of groups must match nrow(STATS)")
   .Call(Cpp_TRAl,x,STATS,g[[2L]],TtI(FUN))
 }

### 2021-06-03 R/collap.R

@@ -12,11 +12,13 @@ fmin_uw <- function(x, g, w, ...) fmin(x, g, ...)
 fmax_uw <- function(x, g, w, ...) fmax(x, g, ...)
 ffirst_uw <- function(x, g, w, ...) ffirst(x, g, ...)
 flast_uw <- function(x, g, w, ...) flast(x, g, ...)
-fNobs_uw <- function(x, g, w, ...) fNobs(x, g, ...)
-fNdistinct_uw <- function(x, g, w, ...) fNdistinct(x, g, ...)
+fnobs_uw <- function(x, g, w, ...) fnobs(x, g, ...)
+fndistinct_uw <- function(x, g, w, ...) fndistinct(x, g, ...)
+fNobs_uw <- function(x, g, w, ...) fnobs(x, g, ...)
+fNdistinct_uw <- function(x, g, w, ...) fndistinct(x, g, ...)
 
 FSF <- c("fmean","fmedian","fmode","fsum","fprod","fsd","fvar",
-         "fmin","fmax","fnth","ffirst","flast","fNobs","fNdistinct")
+         "fmin","fmax","fnth","ffirst","flast","fnobs","fndistinct", "fNobs","fNdistinct")
 
 .FAST_STAT_FUN_EXT <- c(FSF, paste0(FSF, "_uw"))
 
@@ -36,8 +38,10 @@ mymatchfun <- function(FUN) {
          fnth = fnth,
          ffirst = ffirst,
          flast = flast,
-         fNobs = fNobs,
-         fNdistinct = fNdistinct,
+         fnobs = fnobs,
+         fndistinct = fndistinct,
+         fNobs = fnobs,
+         fNdistinct = fndistinct,
          # cat(paste0(paste0(FSF, "_uw"), " = ", paste0(FSF, "_uw"), ",\n"))
          fmean_uw = fmean_uw,
          fmedian_uw = fmedian_uw,
@@ -51,8 +55,10 @@ mymatchfun <- function(FUN) {
          fnth_uw = fnth_uw,
          ffirst_uw = ffirst_uw,
          flast_uw = flast_uw,
-         fNobs_uw = fNobs_uw,
-         fNdistinct_uw = fNdistinct_uw,
+         fnobs_uw = fnobs_uw,
+         fndistinct_uw = fndistinct_uw,
+         fNobs_uw = fnobs_uw,
+         fNdistinct_uw = fndistinct_uw,
[…]

### 2021-06-03 R/descr.R

@@ -11,20 +11,20 @@ descr <- function(X, Ndistinct = TRUE, higher = TRUE, table = TRUE,
 
   dotsok <- if(!missing(...)) names(substitute(c(...))[-1L]) %!in% c('pid','g') else TRUE
 
-  numstats <- if(Ndistinct && dotsok) function(x, ...) armat(qsu.default(x, higher = higher, ...), fNdistinctCpp(x)) else function(x, ...) qsu.default(x, higher = higher, ...)
+  numstats <- if(Ndistinct && dotsok) function(x, ...) armat(qsu.default(x, higher = higher, ...), fndistinctCpp(x)) else function(x, ...) qsu.default(x, higher = higher, ...)
 
   descrnum <- if(is.numeric(Qprobs)) function(x, ...) list(Class = class(x), Label = attr(x, label.attr), Stats = numstats(x, ...),
                                                           Quant = quantile(x, probs = Qprobs, na.rm = TRUE)) else
                                          function(x, ...) list(Class = class(x), Label = attr(x, label.attr), Stats = numstats(x, ...))
   # Could make this more efficient ?
   descrcat <- function(x, tab = table) if(tab) list(Class = class(x), Label = attr(x, label.attr),
-                                                    Stats = if(Ndistinct) c(N = fNobsCpp(x), Ndist = fNdistinctCpp(x)) else `names<-`(fNobsCpp(x), 'Nobs'),
-                                                    Table = natrm(fNobs.default(x, x))) else # table(x). fNobs is a lot Faster, but includes NA as level !
+                                                    Stats = if(Ndistinct) c(N = fnobsCpp(x), Ndist = fndistinctCpp(x)) else `names<-`(fnobsCpp(x), 'Nobs'),
+                                                    Table = natrm(fnobs.default(x, x))) else # table(x). fnobs is a lot Faster, but includes NA as level !
                                                       list(Class = class(x), Label = attr(x, label.attr),
-                                                           Stats = if(Ndistinct) c(N = fNobsCpp(x), Ndist = fNdistinctCpp(x)) else `names<-`(fNobsCpp(x), 'Nobs'))
+                                                           Stats = if(Ndistinct) c(N = fnobsCpp(x), Ndist = fndistinctCpp(x)) else `names<-`(fnobsCpp(x), 'Nobs'))
 
[…]

### 2021-06-03 R/fHDbetween_fHDwithin.R

@@ -34,7 +34,7 @@ myModFrame <- function(f, data) {
 finteract <- function(x, facts, mf) { # x and facts are logical
   f <- which(x & facts)
   if(length(f) == 1L) mf[[f]] else if(length(f) == 2L) do.call(`:`, mf[f]) else
-    as.factor_GRP(GRP.default(mf[f], call = FALSE))
+    as_factor_GRP(GRP.default(mf[f], call = FALSE))
 }
 
 slinteract <- function(sl, facts, mf) { # sl and facts are  logical
@@ -266,10 +266,10 @@ flmres <- function(y, X, w = NULL, meth = "qr", resi = TRUE, ...) {
 
 
 
-fHDwithin <- function(x, ...) UseMethod("fHDwithin") # , x
+fhdwithin <- function(x, ...) UseMethod("fhdwithin") # , x
 
-fHDwithin.default <- function(x, fl, w = NULL, na.rm = TRUE, fill = FALSE, lm.method = "qr", ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(fHDwithin.matrix(x, fl, w, na.rm, fill, ...))
+fhdwithin.default <- function(x, fl, w = NULL, na.rm = TRUE, fill = FALSE, lm.method = "qr", ...) {
+  if(is.matrix(x) && !inherits(x, "matrix")) return(fhdwithin.matrix(x, fl, w, na.rm, fill, ...))
   ax <- attributes(x)
   if(na.rm) {
     cc <- complete.cases(x, fl, w) # gives error if lengths don't match, otherwise demeanlist and qr.resid give errors !!
@@ -322,7 +322,7 @@ fHDwithin.default <- function(x, fl, w = NULL, na.rm = TRUE, fill = FALSE, lm.me
     return(setAttributes(x, ax))
   } else return(setAttributes(demean(x, fl, w, ...), ax))
 }
-fHDwithin.pseries <- function(x, effect = 1:2, w = NULL, na.rm = TRUE, fill = TRUE, ...) {
+fhdwithin.pseries <- function(x, effect = 1:2, w = NULL, na.rm = TRUE, fill = TRUE, ...) {
   ix <- attr(x, "index")
   namix <- attr(ix, "names")
   effect <- cols2int(effect, ix, namix)
@@ -348,7 +348,7 @@ fHDwithin.pseries <- function(x, effect = 1:2, w = NULL, na.rm = TRUE, fill = TR
 }
 
[…]

### 2021-06-03 R/fNdistinct.R

@@ -3,98 +3,98 @@
 # For foundational changes to this code see fsum.R
 # Note: matrix method needs memory equal to size of the object, while data.frame method does not need any memory ?!
 
-fNdistinct <- function(x, ...) UseMethod("fNdistinct") # , x
+fndistinct <- function(x, ...) UseMethod("fndistinct") # , x
 
-fNdistinct.default <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = TRUE, ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(fNdistinct.matrix(x, g, TRA, na.rm, use.g.names, ...))
+fndistinct.default <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = TRUE, ...) {
+  if(is.matrix(x) && !inherits(x, "matrix")) return(fndistinct.matrix(x, g, TRA, na.rm, use.g.names, ...))
   if(!missing(...)) unused_arg_action(match.call(), ...)
   if(is.null(TRA)) {
-    if(is.null(g)) return(.Call(Cpp_fNdistinct,x,0L,0L,NULL,na.rm))
+    if(is.null(g)) return(.Call(Cpp_fndistinct,x,0L,0L,NULL,na.rm))
     if(is.atomic(g)) {
       if(use.g.names) {
         if(!is.nmfactor(g)) g <- qF(g, na.exclude = FALSE)
         lev <- attr(g, "levels")
-        return(`names<-`(.Call(Cpp_fNdistinct,x,length(lev),g,NULL,na.rm), lev))
+        return(`names<-`(.Call(Cpp_fndistinct,x,length(lev),g,NULL,na.rm), lev))
       }
-      if(is.nmfactor(g)) return(.Call(Cpp_fNdistinct,x,fnlevels(g),g,NULL,na.rm))
+      if(is.nmfactor(g)) return(.Call(Cpp_fndistinct,x,fnlevels(g),g,NULL,na.rm))
       g <- qG(g, na.exclude = FALSE)
[…]

### 2021-06-03 R/fbetween_fwithin.R

@@ -14,7 +14,7 @@ fwithin.default <- function(x, g = NULL, w = NULL, na.rm = TRUE, mean = 0, theta
     g <- qG(g, sort = FALSE, na.exclude = FALSE)
     return(.Call(Cpp_BW,x,attr(g,"N.groups"),g,NULL,w,na.rm,theta,ckm(mean),FALSE,FALSE))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_BW,x,g[[1L]],g[[2L]],g[[3L]],w,na.rm,theta,ckm(mean),FALSE,FALSE)
 }
 
@@ -32,7 +32,7 @@ fwithin.matrix <- function(x, g = NULL, w = NULL, na.rm = TRUE, mean = 0, theta 
     g <- qG(g, sort = FALSE, na.exclude = FALSE)
     return(.Call(Cpp_BWm,x,attr(g,"N.groups"),g,NULL,w,na.rm,theta,ckm(mean),FALSE,FALSE))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_BWm,x,g[[1L]],g[[2L]],g[[3L]],w,na.rm,theta,ckm(mean),FALSE,FALSE)
 }
 
@@ -44,7 +44,7 @@ fwithin.data.frame <- function(x, g = NULL, w = NULL, na.rm = TRUE, mean = 0, th
     g <- qG(g, sort = FALSE, na.exclude = FALSE)
     return(.Call(Cpp_BWl,x,attr(g,"N.groups"),g,NULL,w,na.rm,theta,ckm(mean),FALSE,FALSE))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_BWl,x,g[[1L]],g[[2L]],g[[3L]],w,na.rm,theta,ckm(mean),FALSE,FALSE)
 }
 
[…]

### 2021-06-03 R/fcumsum.R

@@ -12,7 +12,7 @@ fcumsum.default <- function(x, g = NULL, o = NULL, na.rm = TRUE, fill = FALSE, c
     }
     return(.Call(C_fcumsum,x,nl,g,o,na.rm,fill))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(C_fcumsum,x,g[[1L]],g[[2L]],o,na.rm,fill)
 }
 
@@ -37,7 +37,7 @@ fcumsum.matrix <- function(x, g = NULL, o = NULL, na.rm = TRUE, fill = FALSE, ch
     }
     return(.Call(C_fcumsumm,x,nl,g,o,na.rm,fill))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(C_fcumsumm,x,g[[1L]],g[[2L]],o,na.rm,fill)
 }
 
@@ -79,7 +79,7 @@ fcumsum.data.frame <- function(x, g = NULL, o = NULL, na.rm = TRUE, fill = FALSE
     }
     return(.Call(C_fcumsuml,x,nl,g,o,na.rm,fill))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(C_fcumsuml,x,g[[1L]],g[[2L]],o,na.rm,fill)
 }
 

### 2021-06-03 R/ffirst.R

@@ -18,7 +18,7 @@ ffirst.default <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = 
       g <- qG(g, na.exclude = FALSE)
       return(.Call(C_ffirst,x,attr(g,"N.groups"),g,na.rm))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
     if(use.g.names) return(`names<-`(.Call(C_ffirst,x,g[[1L]],g[[2L]],na.rm), GRPnames(g)))
     return(.Call(C_ffirst,x,g[[1L]],g[[2L]],na.rm))
   }
@@ -28,7 +28,7 @@ ffirst.default <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = 
     g <- qG(g, na.exclude = FALSE)
     return(.Call(Cpp_TRA,x,.Call(C_ffirst,x,attr(g,"N.groups"),g,na.rm),g,TtI(TRA)))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_TRA,x,.Call(C_ffirst,x,g[[1L]],g[[2L]],na.rm),g[[2L]],TtI(TRA))
 }
 
@@ -46,7 +46,7 @@ ffirst.matrix <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = T
       g <- qG(g, na.exclude = FALSE)
       return(.Call(C_ffirstm,x,attr(g,"N.groups"),g,na.rm,FALSE))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
     if(use.g.names) return(`dimnames<-`(.Call(C_ffirstm,x,g[[1L]],g[[2L]],na.rm,FALSE), list(GRPnames(g), dimnames(x)[[2L]])))
     return(.Call(C_ffirstm,x,g[[1L]],g[[2L]],na.rm,FALSE))
   }
[…]

### 2021-06-03 R/flag.R

@@ -12,7 +12,7 @@ flag.default <- function(x, n = 1, g = NULL, t = NULL, fill = NA, stubs = TRUE, 
     }
     return(.Call(Cpp_flaglead,x,n,fill,nl,g,G_t(t),stubs))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_flaglead,x,n,fill,g[[1L]],g[[2L]],G_t(t),stubs)
 }
 
@@ -35,7 +35,7 @@ flag.matrix <- function(x, n = 1, g = NULL, t = NULL, fill = NA, stubs = length(
     }
     return(.Call(Cpp_flagleadm,x,n,fill,nl,g,G_t(t),stubs))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_flagleadm,x,n,fill,g[[1L]],g[[2L]],G_t(t),stubs)
 }
 
@@ -74,7 +74,7 @@ flag.data.frame <- function(x, n = 1, g = NULL, t = NULL, fill = NA, stubs = len
     }
     return(.Call(Cpp_flagleadl,x,n,fill,nl,g,G_t(t),stubs))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_flagleadl,x,n,fill,g[[1L]],g[[2L]],G_t(t),stubs)
 }
 flag.list <- function(x, n = 1, g = NULL, t = NULL, fill = NA, stubs = length(n) > 1L, ...)
@@ -127,7 +127,7 @@ L.data.frame <- function(x, n = 1, by = NULL, t = NULL, cols = is.numeric,
     } else {
       gn <- NULL
       if(length(cols)) cols <- cols2int(cols, x, nam)
-      if(!is.GRP(by)) by <- if(is.null(by)) list(0L, 0L, NULL) else if(is.atomic(by)) # Necessary for if by is passed externally !
[…]

### 2021-06-03 R/flast.R

@@ -18,7 +18,7 @@ flast.default <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = T
       g <- qG(g, na.exclude = FALSE)
       return(.Call(C_flast,x,attr(g,"N.groups"),g,na.rm))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
     if(use.g.names) return(`names<-`(.Call(C_flast,x,g[[1L]],g[[2L]],na.rm), GRPnames(g)))
     return(.Call(C_flast,x,g[[1L]],g[[2L]],na.rm))
   }
@@ -28,7 +28,7 @@ flast.default <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = T
     g <- qG(g, na.exclude = FALSE)
     return(.Call(Cpp_TRA,x,.Call(C_flast,x,attr(g,"N.groups"),g,na.rm),g,TtI(TRA)))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_TRA,x,.Call(C_flast,x,g[[1L]],g[[2L]],na.rm),g[[2L]],TtI(TRA))
 }
 
@@ -46,7 +46,7 @@ flast.matrix <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = TR
       g <- qG(g, na.exclude = FALSE)
       return(.Call(C_flastm,x,attr(g,"N.groups"),g,na.rm,FALSE))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
     if(use.g.names) return(`dimnames<-`(.Call(C_flastm,x,g[[1L]],g[[2L]],na.rm,FALSE), list(GRPnames(g), dimnames(x)[[2L]])))
     return(.Call(C_flastm,x,g[[1L]],g[[2L]],na.rm,FALSE))
   }
[…]

### 2021-06-03 R/fmin_fmax.R

@@ -18,7 +18,7 @@ fmin.default <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = TR
       g <- qG(g, na.exclude = FALSE)
       return(.Call(Cpp_fminmax,x,attr(g,"N.groups"),g,na.rm,1L))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
     if(use.g.names) return(`names<-`(.Call(Cpp_fminmax,x,g[[1L]],g[[2L]],na.rm,1L), GRPnames(g)))
     return(.Call(Cpp_fminmax,x,g[[1L]],g[[2L]],na.rm,1L))
   }
@@ -28,7 +28,7 @@ fmin.default <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = TR
     g <- qG(g, na.exclude = FALSE)
     return(.Call(Cpp_TRA,x,.Call(Cpp_fminmax,x,attr(g,"N.groups"),g,na.rm,1L),g,TtI(TRA)))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_TRA,x,.Call(Cpp_fminmax,x,g[[1L]],g[[2L]],na.rm,1L),g[[2L]],TtI(TRA))
 }
 
@@ -46,7 +46,7 @@ fmin.matrix <- function(x, g = NULL, TRA = NULL, na.rm = TRUE, use.g.names = TRU
       g <- qG(g, na.exclude = FALSE)
       return(.Call(Cpp_fminmaxm,x,attr(g,"N.groups"),g,na.rm,FALSE,1L))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
[…]

### 2021-06-03 R/fnobs.R

@@ -1,98 +1,98 @@
 
 # For foundational changes to this code see fsum.R
 
-fNobs <- function(x, ...) UseMethod("fNobs") # , x
+fnobs <- function(x, ...) UseMethod("fnobs") # , x
 
-fNobs.default <- function(x, g = NULL, TRA = NULL, use.g.names = TRUE, ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(fNobs.matrix(x, g, TRA, use.g.names, ...))
+fnobs.default <- function(x, g = NULL, TRA = NULL, use.g.names = TRUE, ...) {
+  if(is.matrix(x) && !inherits(x, "matrix")) return(fnobs.matrix(x, g, TRA, use.g.names, ...))
   if(!missing(...)) unused_arg_action(match.call(), ...)
   if(is.null(TRA)) {
-    if(is.null(g)) return(.Call(Cpp_fNobs,x,0L,0L))
+    if(is.null(g)) return(.Call(Cpp_fnobs,x,0L,0L))
     if(is.atomic(g)) {
       if(use.g.names) {
         if(!is.nmfactor(g)) g <- qF(g, na.exclude = FALSE)
         lev <- attr(g, "levels")
-        return(`names<-`(.Call(Cpp_fNobs,x,length(lev),g), lev))
+        return(`names<-`(.Call(Cpp_fnobs,x,length(lev),g), lev))
       }
-      if(is.nmfactor(g)) return(.Call(Cpp_fNobs,x,fnlevels(g),g))
+      if(is.nmfactor(g)) return(.Call(Cpp_fnobs,x,fnlevels(g),g))
       g <- qG(g, na.exclude = FALSE)
-      return(.Call(Cpp_fNobs,x,attr(g,"N.groups"),g))
+      return(.Call(Cpp_fnobs,x,attr(g,"N.groups"),g))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
-    if(use.g.names) return(`names<-`(.Call(Cpp_fNobs,x,g[[1L]],g[[2L]]), GRPnames(g)))
[…]

### 2021-06-03 R/fnth_fmedian.R

@@ -18,7 +18,7 @@ fnth.default <- function(x, n = 0.5, g = NULL, w = NULL, TRA = NULL, na.rm = TRU
       g <- qG(g, na.exclude = FALSE)
       return(.Call(Cpp_fnth,x,n,attr(g,"N.groups"),g,NULL,w,na.rm,ret))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
     if(use.g.names) return(`names<-`(.Call(Cpp_fnth,x,n,g[[1L]],g[[2L]],g[[3L]],w,na.rm,ret), GRPnames(g)))
     return(.Call(Cpp_fnth,x,n,g[[1L]],g[[2L]],g[[3L]],w,na.rm,ret))
   }
@@ -28,7 +28,7 @@ fnth.default <- function(x, n = 0.5, g = NULL, w = NULL, TRA = NULL, na.rm = TRU
     g <- qG(g, na.exclude = FALSE)
     return(.Call(Cpp_TRA,x,.Call(Cpp_fnth,x,n,attr(g,"N.groups"),g,NULL,w,na.rm,ret),g,TtI(TRA)))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_TRA,x,.Call(Cpp_fnth,x,n,g[[1L]],g[[2L]],g[[3L]],w,na.rm,ret),g[[2L]],TtI(TRA))
 }
 
@@ -47,7 +47,7 @@ fnth.matrix <- function(x, n = 0.5, g = NULL, w = NULL, TRA = NULL, na.rm = TRUE
       g <- qG(g, na.exclude = FALSE)
       return(.Call(Cpp_fnthm,x,n,attr(g,"N.groups"),g,NULL,w,na.rm,FALSE,ret))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
[…]

### 2021-06-03 R/fprod.R

@@ -18,7 +18,7 @@ fprod.default <- function(x, g = NULL, w = NULL, TRA = NULL, na.rm = TRUE, use.g
       g <- qG(g, na.exclude = FALSE)
       return(.Call(Cpp_fprod,x,attr(g,"N.groups"),g,w,na.rm))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
     if(use.g.names) return(`names<-`(.Call(Cpp_fprod,x,g[[1L]],g[[2L]],w,na.rm), GRPnames(g)))
     return(.Call(Cpp_fprod,x,g[[1L]],g[[2L]],w,na.rm))
   }
@@ -28,7 +28,7 @@ fprod.default <- function(x, g = NULL, w = NULL, TRA = NULL, na.rm = TRUE, use.g
     g <- qG(g, na.exclude = FALSE)
     return(.Call(Cpp_TRA,x,.Call(Cpp_fprod,x,attr(g,"N.groups"),g,w,na.rm),g,TtI(TRA)))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_TRA,x,.Call(Cpp_fprod,x,g[[1L]],g[[2L]],w,na.rm),g[[2L]],TtI(TRA))
 }
 
@@ -46,7 +46,7 @@ fprod.matrix <- function(x, g = NULL, w = NULL, TRA = NULL, na.rm = TRUE, use.g.
       g <- qG(g, na.exclude = FALSE)
       return(.Call(Cpp_fprodm,x,attr(g,"N.groups"),g,w,na.rm,FALSE))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
     if(use.g.names) return(`dimnames<-`(.Call(Cpp_fprodm,x,g[[1L]],g[[2L]],w,na.rm,FALSE), list(GRPnames(g), dimnames(x)[[2L]])))
[…]

### 2021-06-03 R/fsummarise.R

@@ -6,10 +6,10 @@ fFUN_add_groups <- function(x) {
 
 othFUN_compute <- function(x) {
   if(length(x) == 2L) # No additional function arguments
-    return(substitute(copyMostAttrib(unlist(lapply(split(a, as.factor_GRP(.g_)), b), FALSE, FALSE), a),
+    return(substitute(copyMostAttrib(unlist(lapply(split(a, as_factor_GRP(.g_)), b), FALSE, FALSE), a),
                       list(a = x[[2L]], b = x[[1L]])))
   # With more arguments, things become more complex..
-  lapply_call <- as.call(c(list(quote(lapply), substitute(split(a, as.factor_GRP(.g_)), list(a = x[[2L]]))), as.list(x[-2L])))
+  lapply_call <- as.call(c(list(quote(lapply), substitute(split(a, as_factor_GRP(.g_)), list(a = x[[2L]]))), as.list(x[-2L])))
   substitute(copyMostAttrib(unlist(a, FALSE, FALSE), b),
              list(a = lapply_call, b = x[[2L]]))
 }

### 2021-06-03 R/fvar_fsd.R

@@ -19,7 +19,7 @@ fsd.default <- function(x, g = NULL, w = NULL, TRA = NULL, na.rm = TRUE, use.g.n
       g <- qG(g, na.exclude = FALSE)
       return(.Call(Cpp_fvarsd,x,attr(g,"N.groups"),g,NULL,w,na.rm,stable.algo,TRUE))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
     if(use.g.names) return(`names<-`(.Call(Cpp_fvarsd,x,g[[1L]],g[[2L]],g[[3L]],w,na.rm,stable.algo,TRUE), GRPnames(g)))
     return(.Call(Cpp_fvarsd,x,g[[1L]],g[[2L]],g[[3L]],w,na.rm,stable.algo,TRUE))
   }
@@ -29,7 +29,7 @@ fsd.default <- function(x, g = NULL, w = NULL, TRA = NULL, na.rm = TRUE, use.g.n
     g <- qG(g, na.exclude = FALSE)
     return(.Call(Cpp_TRA,x,.Call(Cpp_fvarsd,x,attr(g,"N.groups"),g,NULL,w,na.rm,stable.algo,TRUE),g,TtI(TRA)))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
   .Call(Cpp_TRA,x,.Call(Cpp_fvarsd,x,g[[1L]],g[[2L]],g[[3L]],w,na.rm,stable.algo,TRUE),g[[2L]],TtI(TRA))
 }
 
@@ -47,7 +47,7 @@ fsd.matrix <- function(x, g = NULL, w = NULL, TRA = NULL, na.rm = TRUE, use.g.na
       g <- qG(g, na.exclude = FALSE)
       return(.Call(Cpp_fvarsdm,x,attr(g,"N.groups"),g,NULL,w,na.rm,stable.algo,TRUE,FALSE))
     }
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names, call = FALSE)
[…]

### 2021-06-03 R/global_macros.R

@@ -20,22 +20,41 @@
 # allobj <- ls(getNamespace("collapse"), all.names=TRUE)
 
 
-.COLLAPSE_ALL <- sort(unique(c("%!in%", "%c-%", "%c*%", "%c/%", "%c+%", "%cr%", "%r-%", "%r*%", "%r/%", "%r+%", "%rr%", "add_stub", "add_vars", "add_vars<-", "all_identical", "all_obj_equal", "allNA", "as.character_factor", "as.factor_GRP", "as.factor_qG", "as.numeric_factor", "atomic_elem", "atomic_elem<-", "av", "av<-", "B", "BY", "cat_vars", "cat_vars<-", "char_vars", "char_vars<-", "cinv", "ckmatch", "collap", "collapg", "collapv", "colorder", "colorderv", "copyAttrib", "copyMostAttrib", "D", "dapply", "Date_vars", "Date_vars<-", "descr", "Dlog", "F", "fact_vars", "fact_vars<-", "fbetween", "fcompute", "fdiff", "fdim", "fdroplevels", "fdroplevels.factor", "ffirst", "fFtest", "fgroup_by", "fgroup_vars", "fgrowth", "fHDbetween", "fHDwithin", "finteraction", "flag", "flast", "flm", "fmax", "fmean", "fmedian", "fmin", "fmode", "fncol", "fNdistinct", "fnlevels", "fNobs", "fnrow", "fnth", "fprod", "frename", "fscale", "fsd", "fselect", "fselect<-", "fsubset", "fsum", "fsummarise", "ftransform", "ftransform<-", "ftransformv", "fungroup", "funique", "fvar", "fwithin", "G", "gby", "get_elem", "get_vars", "get_vars<-", "GGDC10S", "groupid", "GRP", "GRPnames", "gv", "gv<-", "gvr", "gvr<-", "has_elem", "HDB", "HDW", "irreg_elem", "is.categorical", "is.Date", "is.GRP", "is.qG", "is.regular", "is.unlistable", "L", "ldepth", "list_elem", "list_elem<-", "logi_vars", "logi_vars<-", "mctl", "missing_cases", "mrtl", "na_insert", "na_omit", "na_rm", "namlab", "num_vars", "num_vars<-", "nv", "nv<-", "print.pwcor", "print.pwcov", "print.qsu", "psacf", "psccf", "psmat", "pspacf", "pwcor", "pwcov", "pwNobs", "qDF", "qDT", "qF", "qG", "qM", "qsu",
[…]

### 2021-06-03 R/list_functions.R

@@ -4,7 +4,7 @@ rapply2d <- function(l, FUN, ..., classes = "data.frame") {
 }
 
 get_elem_FUN <- function(x, FUN, return = "sublist", keep_class = FALSE)
-  switch(return, sublist = if(keep_class) colsubsetFUN(x, FUN) else .subset(x, vapply(`attributes<-`(x, NULL), FUN, TRUE)),
+  switch(return, sublist = if(keep_class) fcolsubset(x, vapply(`attributes<-`(x, NULL), FUN, TRUE)) else .subset(x, vapply(`attributes<-`(x, NULL), FUN, TRUE)),
          names = attr(x, "names")[vapply(`attributes<-`(x, NULL), FUN, TRUE)],
          indices = which(vapply(`attributes<-`(x, NULL), FUN, TRUE)),
          named_indices = which(`names<-`(vapply(`attributes<-`(x, NULL), FUN, TRUE), attr(x, "names"))),
@@ -49,7 +49,7 @@ is.regular <- function(x) {
   is.list(x) || is.atomic(x)
 }
 
-is.unlistable <- function(l, DF.as.list = FALSE) if(DF.as.list) all(unlist(rapply(l, is.atomic, how = "list"), use.names = FALSE)) else
+is_unlistable <- function(l, DF.as.list = FALSE) if(DF.as.list) all(unlist(rapply(l, is.atomic, how = "list"), use.names = FALSE)) else
   all(unlist(rapply2d(l, is_regular), use.names = FALSE)) # fastest way?
 
 # If data.frame, search all, otherwise, make optional counting df or not, but don't search them.

### 2021-06-03 R/my_RcppExports.R

@@ -23,32 +23,32 @@ TRAlCpp <- function(x, xAG, g = 0L, ret = 1L) {
     .Call(Cpp_TRAl, x, xAG, g, ret)
 }
 
-fNdistinctCpp <- function(x, ng = 0L, g = 0L, gs = NULL, narm = TRUE) {
-    .Call(Cpp_fNdistinct, x, ng, g, gs, narm)
+fndistinctCpp <- function(x, ng = 0L, g = 0L, gs = NULL, narm = TRUE) {
+    .Call(Cpp_fndistinct, x, ng, g, gs, narm)
 }
 
-fNdistinctlCpp <- function(x, ng = 0L, g = 0L, gs = NULL, narm = TRUE, drop = TRUE) {
-    .Call(Cpp_fNdistinctl, x, ng, g, gs, narm, drop)
+fndistinctlCpp <- function(x, ng = 0L, g = 0L, gs = NULL, narm = TRUE, drop = TRUE) {
+    .Call(Cpp_fndistinctl, x, ng, g, gs, narm, drop)
 }
 
-fNdistinctmCpp <- function(x, ng = 0L, g = 0L, gs = NULL, narm = TRUE, drop = TRUE) {
-    .Call(Cpp_fNdistinctm, x, ng, g, gs, narm, drop)
+fndistinctmCpp <- function(x, ng = 0L, g = 0L, gs = NULL, narm = TRUE, drop = TRUE) {
+    .Call(Cpp_fndistinctm, x, ng, g, gs, narm, drop)
 }
 
-pwNobsmCpp <- function(x) {
-    .Call(Cpp_pwNobsm, x)
+pwnobsmCpp <- function(x) {
+    .Call(Cpp_pwnobsm, x)
 }
 
-fNobsCpp <- function(x, ng = 0L, g = 0L) {
-    .Call(Cpp_fNobs, x, ng, g)
+fnobsCpp <- function(x, ng = 0L, g = 0L) {
+    .Call(Cpp_fnobs, x, ng, g)
 }
 
-fNobsmCpp <- function(x, ng = 0L, g = 0L, drop = TRUE) {
-    .Call(Cpp_fNobsm, x, ng, g, drop)
+fnobsmCpp <- function(x, ng = 0L, g = 0L, drop = TRUE) {
+    .Call(Cpp_fnobsm, x, ng, g, drop)
 }
 
[…]

### 2021-06-03 R/psacf.R

@@ -24,7 +24,7 @@ psacf.default <- function(x, g, t = NULL, lag.max = NULL, type = c("correlation"
     if(is.null(lag.max)) lag.max <- round(2*sqrt(length(x)/ng))
     acf <- getacf(ng, g)
   } else {
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
     if(is.null(lag.max)) lag.max <- round(2*sqrt(length(x)/g[[1L]]))
     acf <- getacf(g[[1L]], g[[2L]])
   }
@@ -91,7 +91,7 @@ psacf.data.frame <- function(x, by, t = NULL, cols = is.numeric, lag.max = NULL,
     if(is.null(lag.max)) lag.max <- round(2*sqrt(nrx/ng))
     acf <- getacf(ng, by)
   } else {
-    if(!is.GRP(by)) by <- GRP.default(by, return.groups = FALSE, call = FALSE)
+    if(!is_GRP(by)) by <- GRP.default(by, return.groups = FALSE, call = FALSE)
     if(is.null(lag.max)) lag.max <- round(2*sqrt(nrx/by[[1L]]))
     acf <- getacf(by[[1L]], by[[2L]])
   }
@@ -237,7 +237,7 @@ psccf.default <- function(x, y, g, t = NULL, lag.max = NULL, type = c("correlati
     if(is.null(lag.max)) lag.max <- round(2*sqrt(lx/ng))
     acf <- getccf(ng, g)
   } else {
-    if(!is.GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
+    if(!is_GRP(g)) g <- GRP.default(g, return.groups = FALSE, call = FALSE)
     if(is.null(lag.max)) lag.max <- round(2*sqrt(lx/g[[1L]]))
     acf <- getccf(g[[1L]], g[[2L]])
   }

### 2021-06-03 R/psmat.R

@@ -10,14 +10,14 @@ psmat.default <- function(x, g, t = NULL, transpose = FALSE, ...) {
     matrix(x, nrow = round(g), byrow = TRUE,
     dimnames = list(paste0("GRP.",seq_len(g)), seq_len(length(x)/round(g))))
   } else {
-  if(!is.nmfactor(g)) if(is.atomic(g)) g <- qF(g, na.exclude = FALSE) else if(is.GRP(g))
-                    g <- as.factor_GRP(g) else g <- as.factor_GRP(GRP.default(g, call = FALSE))
+  if(!is.nmfactor(g)) if(is.atomic(g)) g <- qF(g, na.exclude = FALSE) else if(is_GRP(g))
+                    g <- as_factor_GRP(g) else g <- as_factor_GRP(GRP.default(g, call = FALSE))
   if(is.null(t)) {
     message("No timevar provided: Assuming Balanced Panel")
     return(.Call(Cpp_psmat,x, g, NULL, transpose))
   } else {
-    if(!is.nmfactor(t)) if(is.atomic(t)) t <- qF(t, na.exclude = FALSE) else if(is.GRP(t))
-                      t <- as.factor_GRP(t) else t <- as.factor_GRP(GRP.default(t, call = FALSE))
+    if(!is.nmfactor(t)) if(is.atomic(t)) t <- qF(t, na.exclude = FALSE) else if(is_GRP(t))
+                      t <- as_factor_GRP(t) else t <- as_factor_GRP(GRP.default(t, call = FALSE))
     return(.Call(Cpp_psmat,x, g, t, transpose))
     }
   }
@@ -56,14 +56,14 @@ psmat.data.frame <- function(x, by, t = NULL, cols = NULL, transpose = FALSE, ar
       x <- x[v]
     } else if(length(cols)) x <- x[cols2int(cols, x, names(x), FALSE)]
 
-    if(!is.nmfactor(by)) if(is.atomic(by)) by <- qF(by, na.exclude = FALSE) else if(is.GRP(by))
-                         by <- as.factor_GRP(by) else by <- as.factor_GRP(GRP.default(by, call = FALSE))
+    if(!is.nmfactor(by)) if(is.atomic(by)) by <- qF(by, na.exclude = FALSE) else if(is_GRP(by))
+                         by <- as_factor_GRP(by) else by <- as_factor_GRP(GRP.default(by, call = FALSE))
       if(is.null(t)) {
[…]

### 2021-06-03 R/pwcor_pwcov_pwNobs.R

@@ -1,10 +1,10 @@
 # sumcc <- function(x, y)  sum(complete.cases(x,y))
-# pwNobs <- function(x) qM(dapply(x, function(y) dapply(x, sumcc, y)))
+# pwnobs <- function(x) qM(dapply(x, function(y) dapply(x, sumcc, y)))
 
-pwNobs <- function(X) {
-  if(is.atomic(X) && is.matrix(X)) return(.Call(Cpp_pwNobsm, X)) # cn <- dimnames(X)[[2L]] # X <- mctl(X)
+pwnobs <- function(X) {
+  if(is.atomic(X) && is.matrix(X)) return(.Call(Cpp_pwnobsm, X)) # cn <- dimnames(X)[[2L]] # X <- mctl(X)
   if(!is.list(X)) stop("X must be a matrix or data.frame!") # -> if unequal length will warn below !!
-  dg <- fNobs.data.frame(X)
+  dg <- fnobs.data.frame(X)
   oldClass(X) <- NULL
   n <- length(X)
   nr <- length(X[[1L]])
@@ -40,7 +40,7 @@ corr.pmat <- function(cm, nm) {
   # p.mat
 }
 
-complpwNobs <- function(X) {
+complpwnobs <- function(X) {
   # if(is.list(X)) { # Not needed anymore because now always coercing to matrix...
   #   n <- length(unclass(X))
   #   coln <- attr(X, "names")
@@ -97,7 +97,7 @@ pwcor <- function(X, ..., w = NULL, N = FALSE, P = FALSE, array = TRUE, use = "p
     } else r <- switch(use, complete.obs = stop("no complete element pairs"), namat(X))
   }
   if(!(N || P)) return(`oldClass<-`(r, c("pwcor", "matrix")))
-  n <- if(lcc) nmat(lcc, X) else switch(use, pairwise.complete.obs = pwNobs(X), complpwNobs(X)) # TODO: what about weights paiwrise ? # what if using ... to supply y ???
+  n <- if(lcc) nmat(lcc, X) else switch(use, pairwise.complete.obs = pwnobs(X), complpwnobs(X)) # TODO: what about weights paiwrise ? # what if using ... to supply y ???
   if(N) {
     res <- if(P) list(r = r, N = n, P = corr.pmat(r, n)) else list(r = r, N = n)
   } else res <- list(r = r, P = corr.pmat(r, n))
[…]

### 2021-06-03 R/recode_replace.R

@@ -1,7 +1,7 @@
 
 
 # inspired by ?dplyr::recode
-# Think about adopting this code for as.numeric_factor and as.character_factor
+# Think about adopting this code for as_numeric_factor and as_character_factor
 recode_num <- function(X, ..., default = NULL, missing = NULL) {
   if(missing(...)) stop("recode_num requires arguments of the form: value = replacement")
   args <- list(...)

### 2021-06-03 R/select_replace_add_vars.R

@@ -107,8 +107,8 @@ fact_vars <- function(x, return = "data") get_vars_indl(x, vapply(`attributes<-`
 logi_vars <- function(x, return = "data") get_vars_indl(x, vapply(`attributes<-`(x, NULL), is.logical, TRUE), return)
 "logi_vars<-" <- function(x, value) `get_vars_ind<-`(x, vapply(`attributes<-`(x, NULL), is.logical, TRUE), value)
 
-Date_vars <- function(x, return = "data") get_vars_indl(x, vapply(`attributes<-`(x, NULL), is.Date, TRUE), return)
-"Date_vars<-" <- function(x, value) `get_vars_ind<-`(x, vapply(`attributes<-`(x, NULL), is.Date, TRUE), value)
+date_vars <- function(x, return = "data") get_vars_indl(x, vapply(`attributes<-`(x, NULL), is_date, TRUE), return)
+"date_vars<-" <- function(x, value) `get_vars_ind<-`(x, vapply(`attributes<-`(x, NULL), is_date, TRUE), value)
 
 cat_vars <- function(x, return = "data") get_vars_indl(x, !vapply(`attributes<-`(x, NULL), is.numeric, TRUE), return)
 "cat_vars<-" <- function(x, value) `get_vars_ind<-`(x, !vapply(`attributes<-`(x, NULL), is.numeric, TRUE), value)

### 2021-06-03 R/unlist2d.R

@@ -88,7 +88,7 @@ unlist2d <- function(l, idcols = ".id", row.names = FALSE, recursive = TRUE, id.
 # Examples:
 
 # # example:
-# l = lapply(split(iris[1:4], iris[5]), function(x)list(N = fNobs(x), mean = fmean(x), sd = fsd(x)))
+# l = lapply(split(iris[1:4], iris[5]), function(x)list(N = fnobs(x), mean = fmean(x), sd = fsd(x)))
 #
 # # neat example:
 # l = list(a = mtcars[1:8],b = list(c = mtcars[4:11], d = list(e = mtcars[2:10])))

### 2021-06-03 R/varying.R

@@ -15,7 +15,7 @@ varying.default <- function(x, g = NULL, any_group = TRUE, use.g.names = TRUE, .
     g <- qG(g, na.exclude = FALSE)
     return(.Call(Cpp_varying,x,attr(g,"N.groups"),g,any_group))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names && !any_group, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names && !any_group, call = FALSE)
   if(use.g.names && !any_group) return(`names<-`(.Call(Cpp_varying,x,g[[1L]],g[[2L]],any_group), GRPnames(g)))
   .Call(Cpp_varying,x,g[[1L]],g[[2L]],any_group)
 }
@@ -43,7 +43,7 @@ varying.matrix <- function(x, g = NULL, any_group = TRUE, use.g.names = TRUE, dr
     g <- qG(g, na.exclude = FALSE)
     return(.Call(Cpp_varyingm,x,attr(g,"N.groups"),g,any_group,drop))
   }
-  if(!is.GRP(g)) g <- GRP.default(g, return.groups = use.g.names && !any_group, call = FALSE)
+  if(!is_GRP(g)) g <- GRP.default(g, return.groups = use.g.names && !any_group, call = FALSE)
   if(use.g.names && !any_group) return(`dimnames<-`(.Call(Cpp_varyingm,x,g[[1L]],g[[2L]],any_group,FALSE), list(GRPnames(g), dimnames(x)[[2L]])))
   .Call(Cpp_varyingm,x,g[[1L]],g[[2L]],any_group,drop)
 }
@@ -75,7 +75,7 @@ varying.data.frame <- function(x, by = NULL, cols = NULL, any_group = TRUE, use.
     by <- qG(by, na.exclude = FALSE)
     return(.Call(Cpp_varyingl,x,attr(by,"N.groups"),by,any_group,drop))
   }
-  if(!is.GRP(by)) by <- GRP.default(by, return.groups = use.g.names && !any_group, call = FALSE)
+  if(!is_GRP(by)) by <- GRP.default(by, return.groups = use.g.names && !any_group, call = FALSE)
   if(use.g.names && !any_group && !inherits(x, "data.table") && length(groups <- GRPnames(by)))
[…]

### 2021-06-03 misc/collapse_package_examples.R

@@ -63,8 +63,8 @@ fscale(d, f, w, mean = "overall.mean",  # Full harmonization of group means and 
        sd = "within.sd")                # while preserving the level and scale of the data.
 
 get_vars(iris, 1:2)                      # Use get_vars for fast selecting data.frame columns, gv is a shortcut
-fHDbetween(gv(iris, 1:2), gv(iris, 3:5)) # Linear prediction with factors and continuous covariates
-fHDwithin(gv(iris, 1:2), gv(iris, 3:5))  # Linear partialling out factors and continuous covariates
+fhdbetween(gv(iris, 1:2), gv(iris, 3:5)) # Linear prediction with factors and continuous covariates
+fhdwithin(gv(iris, 1:2), gv(iris, 3:5))  # Linear partialling out factors and continuous covariates
 ss(iris, 1:10, 1:2)                      # Similarly fsubset/ss is super fast for subsetting rows of data.frame.
 
 # Simple Time-Computations..
@@ -123,10 +123,10 @@ head(wlddev) # World Bank World Development Data: 216 countries, 59 years, 4 ser
 
 # Starting with some discriptive tools...
 namlab(wlddev, class = TRUE)           # Show variable names, labels and classes
-fNobs(wlddev)                          # Observation count
-pwNobs(wlddev)                         # Pairwise observation count
-fNobs(wlddev, wlddev$country)          # Grouped observation count
-fNdistinct(wlddev)                     # Distinct values
+fnobs(wlddev)                          # Observation count
+pwnobs(wlddev)                         # Pairwise observation count
+fnobs(wlddev, wlddev$country)          # Grouped observation count
+fndistinct(wlddev)                     # Distinct values
 descr(wlddev)                          # Describe data
 varying(wlddev, ~ country)             # Show which variables vary within countries
 qsu(wlddev, pid = ~ country,           # Panel-summarize columns 9 though 12 of this data

### 2021-06-03 src/ExportSymbols.cpp

@@ -53,13 +53,13 @@ static const R_CallMethodDef CallEntries[] = {
   {"Cpp_TRA", (DL_FUNC) &_collapse_TRACpp, 4},
   {"Cpp_TRAm", (DL_FUNC) &_collapse_TRAmCpp, 4},
   {"Cpp_TRAl", (DL_FUNC) &_collapse_TRAlCpp, 4},
-  {"Cpp_fNdistinct", (DL_FUNC) &_collapse_fNdistinctCpp, 5},
-  {"Cpp_fNdistinctl", (DL_FUNC) &_collapse_fNdistinctlCpp, 6},
-  {"Cpp_fNdistinctm", (DL_FUNC) &_collapse_fNdistinctmCpp, 6},
-  {"Cpp_pwNobsm", (DL_FUNC) &_collapse_pwNobsmCpp, 1},
-  {"Cpp_fNobs", (DL_FUNC) &_collapse_fNobsCpp, 3},
-  {"Cpp_fNobsm", (DL_FUNC) &_collapse_fNobsmCpp, 4},
-  {"Cpp_fNobsl", (DL_FUNC) &_collapse_fNobslCpp, 4},
+  {"Cpp_fndistinct", (DL_FUNC) &_collapse_fndistinctCpp, 5},
+  {"Cpp_fndistinctl", (DL_FUNC) &_collapse_fndistinctlCpp, 6},
+  {"Cpp_fndistinctm", (DL_FUNC) &_collapse_fndistinctmCpp, 6},
+  {"Cpp_pwnobsm", (DL_FUNC) &_collapse_pwnobsmCpp, 1},
+  {"Cpp_fnobs", (DL_FUNC) &_collapse_fnobsCpp, 3},
+  {"Cpp_fnobsm", (DL_FUNC) &_collapse_fnobsmCpp, 4},
+  {"Cpp_fnobsl", (DL_FUNC) &_collapse_fnobslCpp, 4},
   {"Cpp_varying", (DL_FUNC) &_collapse_varyingCpp, 4},
   {"Cpp_varyingm", (DL_FUNC) &_collapse_varyingmCpp, 5},
   {"Cpp_varyingl", (DL_FUNC) &_collapse_varyinglCpp, 5},

### 2021-06-03 src/RcppExports.cpp

@@ -47,94 +47,6 @@ BEGIN_RCPP
     return rcpp_result_gen;
 END_RCPP
 }
-// fNdistinctCpp
-SEXP fNdistinctCpp(const SEXP& x, int ng, const IntegerVector& g, const SEXP& gs, bool narm);
-RcppExport SEXP _collapse_fNdistinctCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP narmSEXP) {
-BEGIN_RCPP
-    Rcpp::RObject rcpp_result_gen;
-    Rcpp::RNGScope rcpp_rngScope_gen;
-    Rcpp::traits::input_parameter< const SEXP& >::type x(xSEXP);
-    Rcpp::traits::input_parameter< int >::type ng(ngSEXP);
-    Rcpp::traits::input_parameter< const IntegerVector& >::type g(gSEXP);
-    Rcpp::traits::input_parameter< const SEXP& >::type gs(gsSEXP);
-    Rcpp::traits::input_parameter< bool >::type narm(narmSEXP);
-    rcpp_result_gen = Rcpp::wrap(fNdistinctCpp(x, ng, g, gs, narm));
-    return rcpp_result_gen;
-END_RCPP
-}
-// fNdistinctlCpp
-SEXP fNdistinctlCpp(const List& x, int ng, const IntegerVector& g, const SEXP& gs, bool narm, bool drop);
-RcppExport SEXP _collapse_fNdistinctlCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP narmSEXP, SEXP dropSEXP) {
-BEGIN_RCPP
-    Rcpp::RObject rcpp_result_gen;
-    Rcpp::RNGScope rcpp_rngScope_gen;
-    Rcpp::traits::input_parameter< const List& >::type x(xSEXP);
-    Rcpp::traits::input_parameter< int >::type ng(ngSEXP);
-    Rcpp::traits::input_parameter< const IntegerVector& >::type g(gSEXP);
-    Rcpp::traits::input_parameter< const SEXP& >::type gs(gsSEXP);
-    Rcpp::traits::input_parameter< bool >::type narm(narmSEXP);
-    Rcpp::traits::input_parameter< bool >::type drop(dropSEXP);
[…]

### 2021-06-03 src/TRA.cpp

@@ -110,7 +110,7 @@ SEXP ret1(const SEXP& x, const SEXP& xAG, const SEXP& g) {
   }
 
   // Attribute Handling - 4 Situations:
-  // 1 - x is classed (factor, date, time series), xAG is not classed. i.e. vector of fNobs, fmean etc.
+  // 1 - x is classed (factor, date, time series), xAG is not classed. i.e. vector of fnobs, fmean etc.
   //    -> Sallow replacing, removing class and levels attributes from x, discard attributes of xAG (if any)
   //    -> or (if type matches i.e. double for date or time series), copy attributes of x unless x is a factor
   // 2 - x is not classed, xAG is classed (factor, date, time series). - an unusual situation should not occurr - copy attributes of xAG, discard attributes of x

### 2021-06-03 src/collapse.h

@@ -13,20 +13,20 @@ RcppExport SEXP _collapse_TRACpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retS
 RcppExport SEXP _collapse_TRAmCpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP);
 // TRAlCpp
 RcppExport SEXP _collapse_TRAlCpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP);
-// fNdistinctCpp
-RcppExport SEXP _collapse_fNdistinctCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP narmSEXP);
-// fNdistinctlCpp
-RcppExport SEXP _collapse_fNdistinctlCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP narmSEXP, SEXP dropSEXP);
-// fNdistinctmCpp
-RcppExport SEXP _collapse_fNdistinctmCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP narmSEXP, SEXP dropSEXP);
-// pwNobsmCpp
-RcppExport SEXP _collapse_pwNobsmCpp(SEXP xSEXP);
-// fNobsCpp
-RcppExport SEXP _collapse_fNobsCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP);
-// fNobsmCpp
-RcppExport SEXP _collapse_fNobsmCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP dropSEXP);
-// fNobslCpp
-RcppExport SEXP _collapse_fNobslCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP dropSEXP);
+// fndistinctCpp
+RcppExport SEXP _collapse_fndistinctCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP narmSEXP);
+// fndistinctlCpp
+RcppExport SEXP _collapse_fndistinctlCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP narmSEXP, SEXP dropSEXP);
[…]

### 2021-06-03 src/fnobs.cpp

@@ -5,7 +5,7 @@ using namespace Rcpp;
 SEXP sym_label = Rf_install("label");
 
 template <int RTYPE>
-IntegerVector fNobsCppImpl(Vector<RTYPE> x, int ng, IntegerVector g) {
+IntegerVector fnobsCppImpl(Vector<RTYPE> x, int ng, IntegerVector g) {
 
   int l = x.size();
 
@@ -35,35 +35,35 @@ IntegerVector fNobsCppImpl(Vector<RTYPE> x, int ng, IntegerVector g) {
 }
 
 template <>
-IntegerVector fNobsCppImpl(Vector<CPLXSXP> x, int ng, IntegerVector) {
+IntegerVector fnobsCppImpl(Vector<CPLXSXP> x, int ng, IntegerVector) {
   stop("Not supported SEXP type!");
 }
 
 template <>
-IntegerVector fNobsCppImpl(Vector<VECSXP> x, int ng, IntegerVector) {
+IntegerVector fnobsCppImpl(Vector<VECSXP> x, int ng, IntegerVector) {
   stop("Not supported SEXP type!");
 }
 
 template <>
-IntegerVector fNobsCppImpl(Vector<RAWSXP> x, int ng, IntegerVector) {
+IntegerVector fnobsCppImpl(Vector<RAWSXP> x, int ng, IntegerVector) {
   stop("Not supported SEXP type!");
 }
 
 template <>
-IntegerVector fNobsCppImpl(Vector<EXPRSXP> x, int ng, IntegerVector) {
+IntegerVector fnobsCppImpl(Vector<EXPRSXP> x, int ng, IntegerVector) {
   stop("Not supported SEXP type!");
 }
 
 // [[Rcpp::export]]
-IntegerVector fNobsCpp(const SEXP& x, int ng = 0, const IntegerVector& g = 0){
-  RCPP_RETURN_VECTOR(fNobsCppImpl, x, ng, g);
+IntegerVector fnobsCpp(const SEXP& x, int ng = 0, const IntegerVector& g = 0){
+  RCPP_RETURN_VECTOR(fnobsCppImpl, x, ng, g);
 }
 
 
 
 
 template <int RTYPE>
-SEXP fNobsmCppImpl(Matrix<RTYPE> x, int ng, IntegerVector g, bool drop) {
+SEXP fnobsmCppImpl(Matrix<RTYPE> x, int ng, IntegerVector g, bool drop) {
[…]

### 2021-06-03 src/varying.cpp

@@ -12,7 +12,7 @@ LogicalVector varyingCppImpl(Vector<RTYPE> x, int ng, IntegerVector g, bool any_
   auto isnanT = (RTYPE == REALSXP) ? [](storage_t x) { return x != x; } :
     [](storage_t x) { return x == Vector<RTYPE>::get_na(); };
 
-    if(ng == 0) { // Note: Does not return NA if all NA... can be checked with fNobs ...
+    if(ng == 0) { // Note: Does not return NA if all NA... can be checked with fnobs ...
       int j = l-1;
       storage_t vi = x[j];
       while(isnanT(vi) && j!=0) vi = x[--j];

### 2021-06-03 tests/testthat/test-GRP.R

@@ -187,25 +187,25 @@ test_that("fgroup_vars works as intended", {
 
 test_that("GRP <> factor conversions run seamlessly", {
 
-  expect_identical(unclass(iris$Species), unclass(as.factor_GRP(GRP(iris$Species)))) # as.factor_GRP always adds class "na.included"
+  expect_identical(unclass(iris$Species), unclass(as_factor_GRP(GRP(iris$Species)))) # as_factor_GRP always adds class "na.included"
 
-  expect_identical(unclass(wlddev$iso3c[1:200]), unclass(as.factor_GRP(GRP(wlddev$iso3c[1:200])))) # as.factor_GRP always adds class "na.included"
-  expect_identical(unclass(fdroplevels(wlddev$iso3c[1:200])), unclass(as.factor_GRP(GRP(wlddev$iso3c[1:200], drop = TRUE)))) # as.factor_GRP always adds class "na.included"
+  expect_identical(unclass(wlddev$iso3c[1:200]), unclass(as_factor_GRP(GRP(wlddev$iso3c[1:200])))) # as_factor_GRP always adds class "na.included"
+  expect_identical(unclass(fdroplevels(wlddev$iso3c[1:200])), unclass(as_factor_GRP(GRP(wlddev$iso3c[1:200], drop = TRUE)))) # as_factor_GRP always adds class "na.included"
 
-  expect_identical(unclass(`vlabels<-`(wlddev2$iso3c, "label", NULL)), unclass(as.factor_GRP(GRP(wlddev2$iso3c))))
+  expect_identical(unclass(`vlabels<-`(wlddev2$iso3c, "label", NULL)), unclass(as_factor_GRP(GRP(wlddev2$iso3c))))
   int <- sample.int(10,100,TRUE)
-  expect_identical(unclass(qF(int)), unclass(as.factor_GRP(GRP(int))))
-  expect_identical(unclass(qF(int)), unclass(as.factor_GRP(GRP(qF(int)))))
+  expect_identical(unclass(qF(int)), unclass(as_factor_GRP(GRP(int))))
+  expect_identical(unclass(qF(int)), unclass(as_factor_GRP(GRP(qF(int)))))
   intNA <- int
   intNA[sample(100,20)] <- NA
-  expect_identical(unclass(qF(intNA, na.exclude = FALSE)), unclass(as.factor_GRP(GRP(intNA))))
[…]

### 2021-06-03 tests/testthat/test-TRA.R

@@ -94,47 +94,47 @@ test_that("TRA gives errors for wrong input", {
 
 test_that("TRA handles different data types as intended", {
   # Vector & Matrix: Simple
-  expect_true(is.integer(fNobs(letters, TRA = "replace_fill")))
-  expect_true(is.integer(fNobs(letters, TRA = "replace")))
-  for(i in c("-", "+", "*", "/", "%", "%%", "-%%"))  expect_error(fNobs(letters, TRA = i))
-  expect_true(is.integer(fNobs(AirPassengers, TRA = "replace_fill")))
-  expect_true(is.integer(fNobs(AirPassengers, TRA = "replace")))
-  for(i in c("-", "+", "*", "/", "%", "%%", "-%%"))  expect_true(is.numeric(fNobs(AirPassengers, TRA = i)))
-  expect_true(is.integer(fNobs(EuStockMarkets, TRA = "replace_fill")))
-  expect_true(is.integer(fNobs(EuStockMarkets, TRA = "replace")))
-  for(i in c("-", "+", "*", "/", "%", "%%", "-%%"))  expect_true(is.numeric(fNobs(EuStockMarkets, TRA = i)))
+  expect_true(is.integer(fnobs(letters, TRA = "replace_fill")))
+  expect_true(is.integer(fnobs(letters, TRA = "replace")))
+  for(i in c("-", "+", "*", "/", "%", "%%", "-%%"))  expect_error(fnobs(letters, TRA = i))
+  expect_true(is.integer(fnobs(AirPassengers, TRA = "replace_fill")))
+  expect_true(is.integer(fnobs(AirPassengers, TRA = "replace")))
+  for(i in c("-", "+", "*", "/", "%", "%%", "-%%"))  expect_true(is.numeric(fnobs(AirPassengers, TRA = i)))
+  expect_true(is.integer(fnobs(EuStockMarkets, TRA = "replace_fill")))
+  expect_true(is.integer(fnobs(EuStockMarkets, TRA = "replace")))
+  for(i in c("-", "+", "*", "/", "%", "%%", "-%%"))  expect_true(is.numeric(fnobs(EuStockMarkets, TRA = i)))
   # Vector & Matrix: Grouped
-  expect_true(is.integer(fNobs(letters, sample.int(3, length(letters), TRUE), TRA = "replace_fill")))
[…]

### 2021-06-03 tests/testthat/test-attribute-handling.R

@@ -15,7 +15,7 @@ f1 <- sample.int(5, length(AirPassengers), replace = TRUE)
 f2 <- sample.int(5, nrow(EuStockMarkets), replace = TRUE)
 
 numFUN <- setdiff(.FAST_STAT_FUN, c("fnth", "fmode", "ffirst", "flast"))
-countFUN <- c("fNobs", "fNdistinct")
+countFUN <- c("fnobs", "fndistinct")
 
 test_that("statistical functions handle attributes properly", {
 
@@ -102,7 +102,7 @@ test_that("statistical functions handle attributes properly", {
 
 })
 
-transFUN <- setdiff(c(.FAST_FUN, .OPERATOR_FUN), c(.FAST_STAT_FUN, "fHDbetween", "fHDwithin", "HDB", "HDW"))
+transFUN <- setdiff(c(.FAST_FUN, .OPERATOR_FUN), c(.FAST_STAT_FUN, "fhdbetween", "fhdwithin", "HDB", "HDW"))
 
 options(collapse_unused_arg_action = "none", warn = -1)
 
@@ -147,7 +147,7 @@ test_that("TRA attribute preservation works well", {
   expect_equal(attributes(TRA(AirPassengers, 1, "-")), attributes(AirPassengers))            # Double
   expect_equal(attributes(TRA(AirPassengers, 1L, "-")), attributes(AirPassengers))           # Integer -> Coerced to double in numeric operation
   f <- qF(sample.int(5L, length(AirPassengers), TRUE), na.exclude = FALSE)
-  num <- unclass(fmean(AirPassengers, f)); int <- fNobs(AirPassengers, f)
+  num <- unclass(fmean(AirPassengers, f)); int <- fnobs(AirPassengers, f)
   expect_equal(attributes(TRA(AirPassengers, num, "replace", f)), attributes(AirPassengers))      # Double
   expect_equal(attributes(TRA(AirPassengers, int, "replace", f))[[1]], tsp(AirPassengers))        # Integer -> Change of type !!
   expect_equal(attributes(TRA(AirPassengers, num, "replace_fill", f)), attributes(AirPassengers)) # Double
@@ -163,7 +163,7 @@ test_that("TRA attribute preservation works well", {
[…]

### 2021-06-03 tests/testthat/test-fHDbetween-fHDwithin-HDB-HDW.R

@@ -1,4 +1,4 @@
-context("fHDbetween / HDB and fHDwithin / HDW")
+context("fhdbetween / HDB and fhdwithin / HDW")
 
 # rm(list = ls())
 
@@ -49,65 +49,65 @@ basefitted <- function(y, X, na.rm = FALSE) {
   drop(qr.fitted(qr.default(X), y))
 }
 
-# fHDbetween and fHDwithin
+# fhdbetween and fhdwithin
 
-test_that("fHDbetween with one factor performs like fbetween", {
-  expect_equal(fHDbetween(x, f), fbetween(x, f))
-  expect_equal(fHDbetween(x, f, na.rm = FALSE), fbetween(x, f, na.rm = FALSE))
-  expect_equal(fHDbetween(xNA, f, na.rm = FALSE), fbetween(xNA, f, na.rm = FALSE))
-  expect_equal(`attributes<-`(fHDbetween(xNA, f, fill = TRUE), NULL), fbetween(xNA, f))
-  expect_equal(fHDbetween(m, g), fbetween(m, g))
-  expect_equal(fHDbetween(m, g, na.rm = FALSE), fbetween(m, g, na.rm = FALSE))
-  expect_equal(fHDbetween(mNA, g, na.rm = FALSE), fbetween(mNA, g, na.rm = FALSE))
-  # expect_equal(fHDbetween(mNA, g, fill = TRUE), fbetween(mNA, g)) # not matching, fHDbetween matrix is not variable.wise
-  expect_equal(fHDbetween(mtcars, g), fbetween(mtcars, g))
-  expect_equal(fHDbetween(mtcars, g, na.rm = FALSE), fbetween(mtcars, g, na.rm = FALSE))
-  expect_equal(fHDbetween(mtcNA, g, na.rm = FALSE), fbetween(mtcNA, g, na.rm = FALSE))
-  expect_equal(fHDbetween(mtcNA, g, variable.wise = TRUE), fbetween(mtcNA, g))
+test_that("fhdbetween with one factor performs like fbetween", {
+  expect_equal(fhdbetween(x, f), fbetween(x, f))
+  expect_equal(fhdbetween(x, f, na.rm = FALSE), fbetween(x, f, na.rm = FALSE))
+  expect_equal(fhdbetween(xNA, f, na.rm = FALSE), fbetween(xNA, f, na.rm = FALSE))
[…]

### 2021-06-03 tests/testthat/test-fNobs-fNdistinct.R

@@ -1,4 +1,4 @@
-context("fNobs and fNdistinct")
+context("fnobs and fndistinct")
 
 # rm(list = ls())
 
@@ -18,175 +18,175 @@ Ndistinct <- function(x, na.rm = FALSE) {
   return(length(unique(x)))
 }
 
-# fNobs
+# fnobs
 
-test_that("fNobs performs like Nobs (defined above)", {
-  expect_equal(fNobs(NA), as.double(Nobs(NA)))
-  expect_equal(fNobs(1), Nobs(1))
-  expect_equal(fNobs(1:3), Nobs(1:3))
-  expect_equal(fNobs(-1:1), Nobs(-1:1))
-  expect_equal(fNobs(x), Nobs(x))
-  expect_equal(fNobs(xNA), Nobs(xNA))
-  expect_equal(fNobs(data), fNobs(m))
-  expect_equal(fNobs(m), dapply(m, Nobs))
-  expect_equal(fNobs(mNA), dapply(mNA, Nobs))
-  expect_equal(fNobs(x, f), BY(x, f, Nobs))
-  expect_equal(fNobs(xNA, f), BY(xNA, f, Nobs))
-  expect_equal(fNobs(m, g), BY(m, g, Nobs))
-  expect_equal(fNobs(mNA, g), BY(mNA, g, Nobs))
-  expect_equal(dapply(fNobs(data, g), unattrib), BY(data, g, Nobs))
-  expect_equal(dapply(fNobs(dataNA, g), unattrib), BY(dataNA, g, Nobs))
+test_that("fnobs performs like Nobs (defined above)", {
+  expect_equal(fnobs(NA), as.double(Nobs(NA)))
+  expect_equal(fnobs(1), Nobs(1))
+  expect_equal(fnobs(1:3), Nobs(1:3))
+  expect_equal(fnobs(-1:1), Nobs(-1:1))
+  expect_equal(fnobs(x), Nobs(x))
+  expect_equal(fnobs(xNA), Nobs(xNA))
+  expect_equal(fnobs(data), fnobs(m))
+  expect_equal(fnobs(m), dapply(m, Nobs))
+  expect_equal(fnobs(mNA), dapply(mNA, Nobs))
+  expect_equal(fnobs(x, f), BY(x, f, Nobs))
+  expect_equal(fnobs(xNA, f), BY(xNA, f, Nobs))
+  expect_equal(fnobs(m, g), BY(m, g, Nobs))
[…]

### 2021-06-03 tests/testthat/test-flag-L-F.R

@@ -31,7 +31,7 @@ o = order(o)
 od = order(rnorm(length(td)))
 muo = m[od, ]
 datauo = data[od, ]
-guo = as.factor_GRP(g)[od]
+guo = as_factor_GRP(g)[od]
 tduo = td[od]
 t2duo = seq_along(od)[od]
 od = order(od)

### 2021-06-03 tests/testthat/test-fmedian.R

@@ -11,7 +11,7 @@ xNA[sample.int(100,20)] <- NA
 wNA[is.na(xNA)] <- NA  # only missing weights if x also missing
 f <- as.factor(sample.int(10, 100, TRUE))
 g <- GRP(mtcars, ~ cyl + vs + am)
-gf <- as.factor_GRP(g)
+gf <- as_factor_GRP(g)
 mtcNA <- na_insert(mtcars)
 mtcNA[27, 1] <- NA # single group NA !!
 m <- as.matrix(mtcars)
@@ -336,7 +336,7 @@ test_that("fmedian produces errors for wrong input", {
 # fnth
 
 g <- GRP(mtcars, ~ cyl)
-gf <- as.factor_GRP(g)
+gf <- as_factor_GRP(g)
 
 test_that("fnth gives a proper lower/upper/average weighted median on complete data", {
 

### 2021-06-03 tests/testthat/test-fmode.R

@@ -12,7 +12,7 @@ f <- as.factor(sample.int(10, 100, TRUE))
 data <- wlddev[wlddev$iso3c %in% c("BLZ","IND","USA","SRB","GRL"), ]
 l <- nrow(data)
 g <- GRP(droplevels(data$iso3c))
-gf <- as.factor_GRP(g)
+gf <- as_factor_GRP(g)
 dataNA <- na_insert(data)
 m <- as.matrix(num_vars(data)) # without num_vars also works for ties = "first"
 mNA <- as.matrix(num_vars(dataNA))
@@ -100,7 +100,7 @@ wMode <- function(x, w, na.rm = FALSE, ties = "first") {
   g <- GRP.default(x, call = FALSE)
   switch(ties,
          first = {
-           g <- as.factor_GRP(g)
+           g <- as_factor_GRP(g)
            o <- radixorder(unlist(split.default(seq_along(w), g), use.names = FALSE))
            sw <- unlist(lapply(split.default(w, g), cumsum), use.names = FALSE)[o]
            fsubset.default(x, which.max(sw))
@@ -118,7 +118,7 @@ wBY <- function(x, f, FUN, w, ...) {
   if(is.atomic(x)) return(dapply(x, function(xi) mapply(FUN, split(xi, f), wspl, ...)))
   dapply(dapply(x, function(xi) {
     r <- Map(FUN, split(xi, f), wspl, ...)
-    if(is.Date(xi)) do.call(c, r) else unlist(r)
+    if(is_date(xi)) do.call(c, r) else unlist(r)
     }), `names<-`, NULL)
 }
 

### 2021-06-03 tests/testthat/test-fsum.R

@@ -13,7 +13,7 @@ wNA[sample.int(100,20)] <- NA
 wdatNA[sample.int(32, 5)] <- NA
 f <- as.factor(sample.int(10, 100, TRUE))
 g <- GRP(mtcars, ~ cyl + vs + am)
-gf <- as.factor_GRP(g)
+gf <- as_factor_GRP(g)
 mtcNA <- na_insert(mtcars)
 mtcNA[27,1] <- NA # single group NA !!
 m <- as.matrix(mtcars)

### 2021-06-03 tests/testthat/test-list-processing.R

@@ -55,12 +55,12 @@ test_that("get_elem works well", { # Could still add more tests..
 })
 
 test_that("reg_elem and irreg_elem work well", {
-  expect_true(is.unlistable(reg_elem(l)))
-  expect_false(is.unlistable(irreg_elem(l)))
-  expect_true(is.unlistable(reg_elem(list(l), keep.tree = FALSE)))
-  expect_true(is.unlistable(reg_elem(list(l), keep.tree = TRUE)))
-  expect_false(is.unlistable(irreg_elem(list(l), keep.tree = FALSE)))
-  expect_false(is.unlistable(irreg_elem(list(l), keep.tree = TRUE)))
+  expect_true(is_unlistable(reg_elem(l)))
+  expect_false(is_unlistable(irreg_elem(l)))
+  expect_true(is_unlistable(reg_elem(list(l), keep.tree = FALSE)))
+  expect_true(is_unlistable(reg_elem(list(l), keep.tree = TRUE)))
+  expect_false(is_unlistable(irreg_elem(list(l), keep.tree = FALSE)))
+  expect_false(is_unlistable(irreg_elem(list(l), keep.tree = TRUE)))
 
 })
 
@@ -68,9 +68,9 @@ test_that("has_elem works well", {
   expect_true(has_elem(l, is.matrix))
   expect_true(has_elem(l, is.data.frame))
   expect_false(has_elem(l, is.data.frame, DF.as.list = TRUE))
-  expect_true(has_elem(l, is.categorical))
-  expect_false(has_elem(l, is.Date))
-  expect_false(has_elem(l, is.qG))
+  expect_true(has_elem(l, is_categorical))
+  expect_false(has_elem(l, is_date))
+  expect_false(has_elem(l, is_qG))
 
   expect_false(has_elem(l, "am", recursive = FALSE))
   expect_false(has_elem(l, "pivot", recursive = FALSE))

### 2021-06-03 tests/testthat/test-misc.R

@@ -4,7 +4,7 @@ context("Misc")
 
 m <- na_insert(qM(mtcars))
 
-test_that("descr, pwcor, pwcov, pwNobs", {
+test_that("descr, pwcor, pwcov, pwnobs", {
 
   expect_visible(descr(wlddev))
   expect_visible(as.data.frame(descr(wlddev)))
@@ -23,13 +23,13 @@ test_that("descr, pwcor, pwcov, pwNobs", {
   expect_output(print(pwcov(nv(wlddev), N = TRUE, P = TRUE)))
   expect_output(print(pwcov(nv(wlddev), N = TRUE, P = TRUE, use = "complete.obs")))
 
-  expect_visible(pwNobs(wlddev))
-  expect_visible(pwNobs(GGDC10S))
+  expect_visible(pwnobs(wlddev))
+  expect_visible(pwnobs(GGDC10S))
 
   expect_visible(descr(m))
   expect_visible(pwcor(m))
   expect_visible(pwcov(m))
-  expect_visible(pwNobs(m))
+  expect_visible(pwnobs(m))
 
 })
 
@@ -72,7 +72,7 @@ test_that("na_rm works well", {
   expect_equal(sapply(na_insert(wlddev), function(x) vlabels(na_rm(x))), vlabels(wlddev))
   expect_equal(sapply(na_insert(wlddev), function(x) vclasses(na_rm(x))), vclasses(wlddev))
   wldNA <- na_insert(wlddev)
-  expect_equal(lengths(lapply(wldNA, na_rm)), fNobs(wldNA))
+  expect_equal(lengths(lapply(wldNA, na_rm)), fnobs(wldNA))
   rm(wldNA)
 })
 
@@ -146,7 +146,7 @@ test_that("fast functions give same result using different grouping mechanisms",
    expect_equal(FUN(fselect(gmtc, -cyl), TRA = 1L), TRA(fselect(gmtc, -cyl), FUN(gmtc, keep.group_vars = FALSE), 1L))
  }
 
-  for(i in setdiff(.FAST_FUN, c(.FAST_STAT_FUN, "fHDbetween", "fHDwithin"))) {
+  for(i in setdiff(.FAST_FUN, c(.FAST_STAT_FUN, "fhdbetween", "fhdwithin"))) {
 
     FUN <- match.fun(i)
[…]

### 2021-06-03 tests/testthat/test-select-replace-vars.R

@@ -12,10 +12,10 @@ test_that("selecting vars works well", {
   expect_identical(get_vars(wlddev, is.factor), wlddev[sapply(wlddev, is.factor)])
 
   expect_identical(num_vars(wlddev), wlddev[sapply(wlddev, is.numeric)])
-  expect_identical(cat_vars(wlddev), wlddev[sapply(wlddev, is.categorical)])
+  expect_identical(cat_vars(wlddev), wlddev[sapply(wlddev, is_categorical)])
   expect_identical(char_vars(wlddev), wlddev[sapply(wlddev, is.character)])
   expect_identical(fact_vars(wlddev), wlddev[sapply(wlddev, is.factor)])
-  expect_identical(Date_vars(wlddev), wlddev[sapply(wlddev, is.Date)])
+  expect_identical(date_vars(wlddev), wlddev[sapply(wlddev, is_date)])
 })
 
 test_that("replacing vars works well", {
@@ -72,7 +72,7 @@ test_that("replacing vars works well", {
   expect_identical(wlddevold, wlddev)
 
   wlddevold <- wlddev
-  Date_vars(wlddev) <- Date_vars(wlddev)
+  date_vars(wlddev) <- date_vars(wlddev)
   expect_identical(wlddevold, wlddev)
 })
 
@@ -125,13 +125,23 @@ test_that("replacing with or adding atomic elements works well", {
 })
 
 
+test_that("empty selections work well", {
+  expect_equal(cat_vars(mtcars), mtcars[0L])
+  expect_equal(char_vars(mtcars), mtcars[0L])
+  expect_equal(fact_vars(mtcars), mtcars[0L])
+  expect_equal(logi_vars(mtcars), mtcars[0L])
+  expect_equal(get_vars(mtcars, is.character), mtcars[0L])
+  expect_equal(get_vars(mtcars, 0L), mtcars[0L])
+  expect_error(get_vars(mtcars, NULL))
+})
+
 
 test_that("select vars errors for wrong input", {
   expect_error(get_vars(wlddev, 13))
   expect_error(get_vars(wlddev, 1:13))
   expect_error(get_vars(wlddev, -13))
   expect_error(get_vars(wlddev, c("PCGDP","ODA3")))
-  expect_warning(get_vars(wlddev, "bla", regex = TRUE)) # Better give error
[…]

### 2021-06-27 docs/blog/blogposts/2020-06-04-welcome-to-collapse.Rmd

@@
 
-<img src='collapse logo small.png' width="150px" align="right" />
+<img src='collapse_logo_small.png' width="150px" align="right" />
 
@@
 I will start by briefly demonstrating the *Fast Statistical Functions*, which are a central feature of *collapse*.
-Currently there are 13 of them (`fmean`, `fmedian`, `fmode`, `fsum`, `fprod`, `fsd`, `fvar`, `fmin`, `fmax`, `ffirst`, `flast`, `fNobs` and `fNdistinct`), they are all S3 generic and support fast grouped computations on vectors, matrices, data.frame's, lists and grouped tibble's (class *grouped_df*).
+Currently there are 13 of them (`fmean`, `fmedian`, `fmode`, `fsum`, `fprod`, `fsd`, `fvar`, `fmin`, `fmax`, `ffirst`, `flast`, `fnobs` and `fndistinct`), they are all S3 generic and support fast grouped computations on vectors, matrices, data.frame's, lists and grouped tibble's (class *grouped_df*).
 Calling these functions on different objects yields simple column-wise statistical computations:
@@
 It's replacement version `get_vars<-` is 6x faster than `[<-.data.frame`.
-Apart from `fbetween` and `fwithin`, the functions `fHDbetween` and `fHDwithin` can average or center data on multiple groups, and they can also project out continuous variables alongside (i.e.
+Apart from `fbetween` and `fwithin`, the functions `fhdbetween` and `fhdwithin` can average or center data on multiple groups, and they can also project out continuous variables alongside (i.e.
 they provide fitted values or residuals from regression problems which may or may not involve one or more factors).
@@
 # Distinct value count
-fNdistinct(wlddev)
+fndistinct(wlddev)
 # Use descr(wlddev) for a detailed description of each variable
@@
 qDF(rbind(class = vclasses(us_trade), 
-          Ndistinct = fNdistinct(us_trade)))
+          Ndistinct = fndistinct(us_trade)))
 
@@
 
-fNdistinct(us_trade$id) 
+fndistinct(us_trade$id) 
 # Another nice comparison... fmode is similarly fast for categorical aggregation...
-microbenchmark(fNdistinct(us_trade$id), n_distinct(us_trade$id))
+microbenchmark(fndistinct(us_trade$id), n_distinct(us_trade$id))
 

### 2021-06-27 docs/blog/blogposts/Programming.Rmd

@@
 ganyNA <- function(x, g, use.g.names = TRUE) {
-  if(is.atomic(g) || !is.GRP(g)) g <- GRP(g, return.groups = use.g.names)
+  if(is.atomic(g) || !is_GRP(g)) g <- GRP(g, return.groups = use.g.names)
   if(use.g.names) setNames(ganyNACpp(x, g[[1L]], g[[2L]]), GRPnames(g, FALSE)) else

### 2021-07-09 misc/collapse cheat sheet/collapse cheat sheet.tex

\documentclass[10pt,a4paper,landscape]{article}
\usepackage[dvipsnames]{xcolor}
\usepackage{amsmath}
\usepackage{multicol,multirow}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage[landscape]{geometry}
\usepackage[colorlinks=true,citecolor=blue,linkcolor=blue]{hyperref}
\usepackage{booktabs}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage[scaled=1]{helvet}
\renewcommand\familydefault{\sfdefault} 
\usepackage[T1]{fontenc}

\usepackage{setspace} % \linespread{1.3}
\setlength\parindent{0pt}

\usepackage{graphicx}
\usepackage[section]{placeins}
\usepackage{adjustbox}

% \title{\textcolor{Gray}{\textbf{POLICY BREIF}}\\ \vspace{2mm} \textcolor{RoyalBlue}

\ifthenelse{\lengthtest { \paperwidth = 11in}}
    { \geometry{top=.2in,left=.3in,right=.3in,bottom=.3in} } %Normally all .5in (L,R=4 before)
	{\ifthenelse{ \lengthtest{ \paperwidth = 297mm}}
		{\geometry{top=0.8cm,left=1cm,right=1cm,bottom=1cm} }
		{\geometry{top=1cm,left=1cm,right=1cm,bottom=1cm} }
	}
\pagestyle{empty}
\makeatletter
\renewcommand{\section}{\@startsection{section}{1}{0mm}%
                                {-2ex plus -.5ex minus -.2ex}%before -1
                                {0.5ex plus .2ex}%x
                                {\normalfont\large\bfseries}}
\renewcommand{\subsection}{\@startsection{subsection}{2}{0mm}%
                                {-2explus -.5ex minus -.2ex}%before -1
                                {0.5ex plus .2ex}%
                                {\normalfont\normalsize\bfseries}}
\renewcommand{\subsubsection}{\@startsection{subsubsection}{3}{0mm}%
                                {-2ex plus -.5ex minus -.2ex}%set spaces around sections
                                {0.3ex plus .2ex}%normally 1
                                {\normalfont\small\bfseries}}
\makeatother

% For pmatrix  equal spacing:
\renewcommand*{\arraystretch}{1.1}
\renewcommand*{\arraycolsep}{1.2pt}

\setcounter{secnumdepth}{0}
\setlength{\parindent}{0pt}
%\setlength{\parskip}{0pt plus 0.5ex}
\setlength{\parskip}{0.02em}%Length of space between paragraphs

\setlength{\columnsep}{6mm}

\setlength{\itemsep}{0.2em}

% -----------------------------------------------------------------------

\title{My Title}

\begin{document}

\raggedright
\footnotesize

\begin{center}
     \Large{\textbf{My Title}}\\\small{by Sebastian Krantz} %\vspace{mm}
\end{center} 
\begin{multicols*}{4}
\setlength{\premulticols}{1pt}
\setlength{\postmulticols}{1pt}
\setlength{\multicolsep}{1pt}
\setlength{\columnsep}{2pt}

\section{\textcolor{Gray}{Basics}}
\hline

collapse is a powerful (C/C++ based) package supporting advanced (grouped, weighted, time series, panel data and recursive) operations in R. \newline

It also offers a class-agnostic approach to data manipulation - handling matrix and data frame based objects in a uniform \& non-destructive way. \newline

It is particuarly well integrated with data.table, dplyr ((grouped) tibbles) and plm classes for panel data, and can be programmed using pipes \%>\% (magrittr) or standard evaluation.

\section{Fast Statistical Functions}
bla bla

\end{multicols*}
\end{document}

### 2021-11-06 R/RcppExports.R
Pushing full development version 1.7 to Github.

@@ -2,194 +2,194 @@
 # Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
 
 TRACpp <- function(x, xAG, g = 0L, ret = 1L) {
-    .Call(`_collapse_TRACpp`, x, xAG, g, ret)
+    .Call(`_collapsedev17_TRACpp`, x, xAG, g, ret)
 }
 
 TRAlCpp <- function(x, xAG, g = 0L, ret = 1L) {
-    .Call(`_collapse_TRAlCpp`, x, xAG, g, ret)
+    .Call(`_collapsedev17_TRAlCpp`, x, xAG, g, ret)
 }
 
 TRAmCpp <- function(x, xAG, g = 0L, ret = 1L) {
-    .Call(`_collapse_TRAmCpp`, x, xAG, g, ret)
+    .Call(`_collapsedev17_TRAmCpp`, x, xAG, g, ret)
 }
 
 BWCpp <- function(x, ng = 0L, g = 0L, gs = NULL, w = NULL, narm = TRUE, theta = 1, set_mean = 0, B = FALSE, fill = FALSE) {
-    .Call(`_collapse_BWCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
+    .Call(`_collapsedev17_BWCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
 }
 
 BWmCpp <- function(x, ng = 0L, g = 0L, gs = NULL, w = NULL, narm = TRUE, theta = 1, set_mean = 0, B = FALSE, fill = FALSE) {
-    .Call(`_collapse_BWmCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
+    .Call(`_collapsedev17_BWmCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
 }
 
 BWlCpp <- function(x, ng = 0L, g = 0L, gs = NULL, w = NULL, narm = TRUE, theta = 1, set_mean = 0, B = FALSE, fill = FALSE) {
-    .Call(`_collapse_BWlCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
+    .Call(`_collapsedev17_BWlCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
 }
 
[…]

### 2021-11-06 R/zzz.R

@@ -1,7 +1,7 @@
 
 .onLoad <- function(libname, pkgname) {
 
-  res <- .Call(C_collapse_init, "init.success")
+  res <- .Call(C_collapsedev17_init, "init.success")
   if(!is.character(res) || res != "init.success") stop("collapse not succesfully loaded!")
 
   # https://stackoverflow.com/questions/12598242/global-variables-in-packages-in-r
@@ -46,15 +46,15 @@
 }
 
 .onAttach <- function(libname, pkgname) {
-  packageStartupMessage(paste0("collapse ",packageVersion("collapse"),", see ?`collapse-package` or ?`collapse-documentation`\nNote: stats::D  ->  D.expression, D.call, D.name"))
+  packageStartupMessage(paste0("collapse ",packageVersion("collapsedev17"),", see ?`collapse-package` or ?`collapse-documentation`\nNote: stats::D  ->  D.expression, D.call, D.name"))
 }
 
 .onUnload <- function (libpath) {
-  library.dynam.unload("collapse", libpath)
+  library.dynam.unload("collapsedev17", libpath)
 }
 
 # Note: To create local dev version of package change package name in DESCRIPTION, NAMESPACE, this file (including C_collapse_init),
-# replace all instances of `_collapse_` in source files, and also rename `R_init_collapse` in ExportSymbols.cpp.
+# replace all instances of `_collapse_` in source files (except for _collapse_DT_alloccol`), and also rename `R_init_collapse` in ExportSymbols.cpp.
 # and in vignetter / Rd files replace library(collapse)
 release_questions <- function() {
   c(

### 2021-11-06 src/ExportSymbols.cpp

@@ -33,6 +33,10 @@ RcppExport SEXP groups2GRP(SEXP x, SEXP lx, SEXP gs);
 RcppExport SEXP Cna_rm(SEXP x);
 RcppExport SEXP whichv(SEXP x, SEXP val, SEXP Rinvert);
 RcppExport SEXP multiassign(SEXP lhs, SEXP rhs, SEXP envir);
+RcppExport SEXP vlabels(SEXP x, SEXP attrn);
+RcppExport SEXP setvlabels(SEXP x, SEXP attrn, SEXP value, SEXP ind, SEXP sc);
+RcppExport SEXP setnames(SEXP x, SEXP nam);
+RcppExport SEXP groupVec(SEXP X, SEXP starts, SEXP sizes);
 // fnobs rewritten in C:
 RcppExport SEXP fnobsC(SEXP x, SEXP Rng, SEXP g);
 RcppExport SEXP fnobsmC(SEXP x, SEXP Rng, SEXP g, SEXP Rdrop);
@@ -66,34 +70,34 @@ static const R_CMethodDef CEntries[]  = {
 };
 
 static const R_CallMethodDef CallEntries[] = {
-  {"Cpp_BW", (DL_FUNC) &_collapse_BWCpp, 10},
-  {"Cpp_BWm", (DL_FUNC) &_collapse_BWmCpp, 10},
-  {"Cpp_BWl", (DL_FUNC) &_collapse_BWlCpp, 10},
-  {"Cpp_TRA", (DL_FUNC) &_collapse_TRACpp, 4},
-  {"Cpp_TRAm", (DL_FUNC) &_collapse_TRAmCpp, 4},
-  {"Cpp_TRAl", (DL_FUNC) &_collapse_TRAlCpp, 4},
-  {"Cpp_fndistinct", (DL_FUNC) &_collapse_fndistinctCpp, 5},
-  {"Cpp_fndistinctl", (DL_FUNC) &_collapse_fndistinctlCpp, 6},
-  {"Cpp_fndistinctm", (DL_FUNC) &_collapse_fndistinctmCpp, 6},
-  {"Cpp_pwnobsm", (DL_FUNC) &_collapse_pwnobsmCpp, 1},
+  {"Cpp_BW", (DL_FUNC) &_collapsedev17_BWCpp, 10},
+  {"Cpp_BWm", (DL_FUNC) &_collapsedev17_BWmCpp, 10},
[…]

### 2021-11-06 src/RcppExports.cpp

@@ -12,7 +12,7 @@ Rcpp::Rostream<false>& Rcpp::Rcerr = Rcpp::Rcpp_cerr_get();
 
 // TRACpp
 SEXP TRACpp(const SEXP& x, const SEXP& xAG, const IntegerVector& g, int ret);
-RcppExport SEXP _collapse_TRACpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
+RcppExport SEXP _collapsedev17_TRACpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
@@ -26,7 +26,7 @@ END_RCPP
 }
 // TRAlCpp
 List TRAlCpp(const List& x, const SEXP& xAG, const IntegerVector& g, int ret);
-RcppExport SEXP _collapse_TRAlCpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
+RcppExport SEXP _collapsedev17_TRAlCpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
@@ -40,7 +40,7 @@ END_RCPP
 }
 // TRAmCpp
 SEXP TRAmCpp(const SEXP& x, const SEXP& xAG, const IntegerVector& g, int ret);
-RcppExport SEXP _collapse_TRAmCpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
+RcppExport SEXP _collapsedev17_TRAmCpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
@@ -54,7 +54,7 @@ END_RCPP
 }
 // BWCpp
[…]

### 2021-11-06 src/collapse.h

@@ -2,153 +2,153 @@
 using namespace Rcpp;
 
 // BWCpp
-RcppExport SEXP _collapse_BWCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
+RcppExport SEXP _collapsedev17_BWCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
 // BWmCpp
-RcppExport SEXP _collapse_BWmCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
+RcppExport SEXP _collapsedev17_BWmCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
 // BWlCpp
-RcppExport SEXP _collapse_BWlCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
+RcppExport SEXP _collapsedev17_BWlCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
 // TRACpp
-RcppExport SEXP _collapse_TRACpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP);
[…]

### 2022-01-13 R/RcppExports.R
Change package name collapsedev17 -> collapse.

@@ -2,194 +2,194 @@
 # Generator token: 10BE3573-1514-4C36-9D1C-5A225CD40393
 
 TRACpp <- function(x, xAG, g = 0L, ret = 1L) {
-    .Call(`_collapsedev17_TRACpp`, x, xAG, g, ret)
+    .Call(`_collapse_TRACpp`, x, xAG, g, ret)
 }
 
 TRAlCpp <- function(x, xAG, g = 0L, ret = 1L) {
-    .Call(`_collapsedev17_TRAlCpp`, x, xAG, g, ret)
+    .Call(`_collapse_TRAlCpp`, x, xAG, g, ret)
 }
 
 TRAmCpp <- function(x, xAG, g = 0L, ret = 1L) {
-    .Call(`_collapsedev17_TRAmCpp`, x, xAG, g, ret)
+    .Call(`_collapse_TRAmCpp`, x, xAG, g, ret)
 }
 
 BWCpp <- function(x, ng = 0L, g = 0L, gs = NULL, w = NULL, narm = TRUE, theta = 1, set_mean = 0, B = FALSE, fill = FALSE) {
-    .Call(`_collapsedev17_BWCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
+    .Call(`_collapse_BWCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
 }
 
 BWmCpp <- function(x, ng = 0L, g = 0L, gs = NULL, w = NULL, narm = TRUE, theta = 1, set_mean = 0, B = FALSE, fill = FALSE) {
-    .Call(`_collapsedev17_BWmCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
+    .Call(`_collapse_BWmCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
 }
 
 BWlCpp <- function(x, ng = 0L, g = 0L, gs = NULL, w = NULL, narm = TRUE, theta = 1, set_mean = 0, B = FALSE, fill = FALSE) {
-    .Call(`_collapsedev17_BWlCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
+    .Call(`_collapse_BWlCpp`, x, ng, g, gs, w, narm, theta, set_mean, B, fill)
 }
 
[…]

### 2022-01-13 R/zzz.R

@@ -1,7 +1,7 @@
 
 .onLoad <- function(libname, pkgname) {
 
-  res <- .Call(C_collapsedev17_init, "init.success")
+  res <- .Call(C_collapse_init, "init.success")
   if(!is.character(res) || res != "init.success") stop("collapse not succesfully loaded!")
 
   # https://stackoverflow.com/questions/12598242/global-variables-in-packages-in-r
@@ -78,12 +78,12 @@
 
   # Experimental collapse_remove option: doesn't work because namespace exports not defined yet.
   # if(length(crem <- getOption("collapse_remove")) && is.character(crem)) {
-  #   # clpns <- getNamespace("collapsedev17")
+  #   # clpns <- getNamespace("collapse")
   #   exports <- getNamespaceInfo(clpns, "exports") # clpns[[".__NAMESPACE__."]][["exports"]] # .getNamespaceInfo(clpns, "exports")
   #   stop("length:", length(exports))
   #   remove(list = crem, envir = exports)
   #   # setNamespaceInfo(clpns, "exports", exports)
-  #   # detach("package:collapsedev17")
+  #   # detach("package:collapse")
   #   # attachNamespace(clpns)
   #   # clpns[[".__NAMESPACE__."]][["exports"]] <- exports
   # }
@@ -97,11 +97,11 @@
 }
 
 .onAttach <- function(libname, pkgname) {
-  packageStartupMessage(paste0("collapse ",packageVersion("collapsedev17"),", see ?`collapse-package` or ?`collapse-documentation`")) # \nNote: stats::D  ->  D.expression, D.call, D.name
+  packageStartupMessage(paste0("collapse ",packageVersion("collapse"),", see ?`collapse-package` or ?`collapse-documentation`")) # \nNote: stats::D  ->  D.expression, D.call, D.name
 }
 
 .onUnload <- function (libpath) {
-  library.dynam.unload("collapsedev17", libpath)
+  library.dynam.unload("collapse", libpath)
 }
 
 # Note: To create local dev version of package change package name in DESCRIPTION, NAMESPACE, this file (including C_collapse_init),

### 2022-01-13 src/ExportSymbols.cpp

@@ -79,34 +79,34 @@ static const R_CMethodDef CEntries[]  = {
 };
 
 static const R_CallMethodDef CallEntries[] = {
-  {"Cpp_BW", (DL_FUNC) &_collapsedev17_BWCpp, 10},
-  {"Cpp_BWm", (DL_FUNC) &_collapsedev17_BWmCpp, 10},
-  {"Cpp_BWl", (DL_FUNC) &_collapsedev17_BWlCpp, 10},
-  {"Cpp_TRA", (DL_FUNC) &_collapsedev17_TRACpp, 4},
-  {"Cpp_TRAm", (DL_FUNC) &_collapsedev17_TRAmCpp, 4},
-  {"Cpp_TRAl", (DL_FUNC) &_collapsedev17_TRAlCpp, 4},
-  {"Cpp_fndistinct", (DL_FUNC) &_collapsedev17_fndistinctCpp, 5},
-  {"Cpp_fndistinctl", (DL_FUNC) &_collapsedev17_fndistinctlCpp, 6},
-  {"Cpp_fndistinctm", (DL_FUNC) &_collapsedev17_fndistinctmCpp, 6},
-  {"Cpp_pwnobsm", (DL_FUNC) &_collapsedev17_pwnobsmCpp, 1},
+  {"Cpp_BW", (DL_FUNC) &_collapse_BWCpp, 10},
+  {"Cpp_BWm", (DL_FUNC) &_collapse_BWmCpp, 10},
+  {"Cpp_BWl", (DL_FUNC) &_collapse_BWlCpp, 10},
+  {"Cpp_TRA", (DL_FUNC) &_collapse_TRACpp, 4},
+  {"Cpp_TRAm", (DL_FUNC) &_collapse_TRAmCpp, 4},
+  {"Cpp_TRAl", (DL_FUNC) &_collapse_TRAlCpp, 4},
+  {"Cpp_fndistinct", (DL_FUNC) &_collapse_fndistinctCpp, 5},
+  {"Cpp_fndistinctl", (DL_FUNC) &_collapse_fndistinctlCpp, 6},
+  {"Cpp_fndistinctm", (DL_FUNC) &_collapse_fndistinctmCpp, 6},
+  {"Cpp_pwnobsm", (DL_FUNC) &_collapse_pwnobsmCpp, 1},
   {"C_fnobs", (DL_FUNC) &fnobsC, 3},
   {"C_fnobsm", (DL_FUNC) &fnobsmC, 4},
[…]

### 2022-01-13 src/RcppExports.cpp

@@ -12,7 +12,7 @@ Rcpp::Rostream<false>& Rcpp::Rcerr = Rcpp::Rcpp_cerr_get();
 
 // TRACpp
 SEXP TRACpp(const SEXP& x, const SEXP& xAG, const IntegerVector& g, int ret);
-RcppExport SEXP _collapsedev17_TRACpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
+RcppExport SEXP _collapse_TRACpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
@@ -26,7 +26,7 @@ END_RCPP
 }
 // TRAlCpp
 List TRAlCpp(const List& x, const SEXP& xAG, const IntegerVector& g, int ret);
-RcppExport SEXP _collapsedev17_TRAlCpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
+RcppExport SEXP _collapse_TRAlCpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
@@ -40,7 +40,7 @@ END_RCPP
 }
 // TRAmCpp
 SEXP TRAmCpp(const SEXP& x, const SEXP& xAG, const IntegerVector& g, int ret);
-RcppExport SEXP _collapsedev17_TRAmCpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
+RcppExport SEXP _collapse_TRAmCpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP) {
 BEGIN_RCPP
     Rcpp::RObject rcpp_result_gen;
     Rcpp::RNGScope rcpp_rngScope_gen;
@@ -54,7 +54,7 @@ END_RCPP
 }
 // BWCpp
[…]

### 2022-01-13 src/collapse.h

@@ -2,153 +2,153 @@
 using namespace Rcpp;
 
 // BWCpp
-RcppExport SEXP _collapsedev17_BWCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
+RcppExport SEXP _collapse_BWCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
 // BWmCpp
-RcppExport SEXP _collapsedev17_BWmCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
+RcppExport SEXP _collapse_BWmCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
 // BWlCpp
-RcppExport SEXP _collapsedev17_BWlCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
+RcppExport SEXP _collapse_BWlCpp(SEXP xSEXP, SEXP ngSEXP, SEXP gSEXP, SEXP gsSEXP, SEXP wSEXP, SEXP narmSEXP, SEXP thetaSEXP, SEXP set_meanSEXP, SEXP BSEXP, SEXP fillSEXP);
 // TRACpp
-RcppExport SEXP _collapsedev17_TRACpp(SEXP xSEXP, SEXP xAGSEXP, SEXP gSEXP, SEXP retSEXP);
[…]

### 2022-01-21 tests/testthat/test-fmutate.R
Using magrittr pipe everywhere to avoid test failures on old R releases.

@@ -1,5 +1,7 @@
 context("fsummarise and fmutate")
 
+library(magrittr)
+
 bmean <- base::mean
 bsum <- base::sum
 bsd <- stats::sd
@@ -41,11 +43,11 @@ test_that("fsummarise works like dplyr::summarise for tagged vector expressions"
               dplyr::summarise(gmtc, mpg = bmean(mpg), carb = bmax(carb), .groups = "drop"))
 
  expect_equal(fsummarise(gmtc, mpg = bmean(mpg), carb = fmax(carb), keep.group_vars = FALSE),
-              fsummarise(gmtc, mpg = bmean(mpg), carb = fmax(carb)) |> slt(-cyl,-vs,-am))
+              fsummarise(gmtc, mpg = bmean(mpg), carb = fmax(carb)) %>% slt(-cyl,-vs,-am))
 
  # Multi-return values
  expect_equal(smr(gmtc, mpg = quantile(mpg)),
-              dplyr::summarise(gmtc, mpg = quantile(mpg), .groups = "drop") |> tfm(mpg = unname(mpg)))
+              dplyr::summarise(gmtc, mpg = quantile(mpg), .groups = "drop") %>% tfm(mpg = unname(mpg)))
 
  # More complex expressions
  expect_equal(smr(gmtc, mpg = bmean(mpg) + 1),
@@ -54,13 +56,13 @@ test_that("fsummarise works like dplyr::summarise for tagged vector expressions"
   expect_equal(smr(gmtc, mpg = bmean(mpg) + q),
               dplyr::summarise(gmtc, mpg = bmean(mpg) + q, .groups = "drop"))
  expect_equal(smr(gmtc, mpg = quantile(mpg) + q),
-              dplyr::summarise(gmtc, mpg = quantile(mpg) + q, .groups = "drop") |> tfm(mpg = unname(mpg)))
+              dplyr::summarise(gmtc, mpg = quantile(mpg) + q, .groups = "drop") %>% tfm(mpg = unname(mpg)))
 
  expect_equal(smr(gmtc, mpg = bmean(mpg) + bmax(v)),
               dplyr::summarise(gmtc, mpg = bmean(mpg) + bmax(v), .groups = "drop"))
 
  expect_equal(smr(gmtc, mpg = quantile(mpg) + bmax(v)),
[…]

### 2022-08-20 misc/collapse cheat sheet/collapse_cheat_sheet_sweave.Rnw
Adding latex cheat sheet (2022): first draft version.

\documentclass[8pt,a4paper,landscape]{extarticle}

%% Language and font encodings
\usepackage[english]{babel}
\usepackage[utf8x]{inputenc}
% \usepackage[T1]{fontenc}

%% Sets page size and margins
\usepackage[landscape,top=0.8cm,bottom=0.5cm,left=0.5cm,right=0.5cm,marginparwidth=1.75cm]{geometry}

%% Useful packages
\usepackage{amsmath}
\usepackage{multicol} % # ,multirow
\setlength{\columnsep}{0.5cm}
\usepackage{graphicx}
\usepackage[dvipsnames]{xcolor}
% \usepackage[colorinlistoftodos]{todonotes}
\usepackage[colorlinks=true, allcolors=NavyBlue]{hyperref}
%\usepackage[section]{placeins}
\usepackage{booktabs}
\usepackage[scaled=1]{helvet}
\renewcommand\familydefault{\sfdefault}
\renewcommand\bfdefault{sb}
% \renewcommand\bfdefault{md}
%\usepackage{fontspec}
% \usepackage{enumitem}
\usepackage{scrextend} % https://tex.stackexchange.com/questions/37080/how-can-i-indent-a-block-of-text-for-a-specified-amount
\usepackage{anyfontsize} % \usepackage{moresize} % https://texblog.org/2012/08/29/changing-the-font-size-in-latex/

\usepackage{setspace} % \linespread{1.3}
\setcounter{secnumdepth}{0}
\setlength{\parindent}{0pt}
\setlength{\parskip}{0.02em}%Length of space between paragraphs
% \setlength{\columnsep}{6mm}
\setlength{\itemsep}{0.2em}
\pagestyle{empty}

\makeatletter
\renewcommand{\section}{\@startsection{section}{1}{0mm}%
                                {-2ex plus -.5ex minus -.2ex}%before -1
                                {0.5ex plus .2ex}%x
                                {\normalfont\large\bfseries\textcolor{WildStrawberry}}}
\renewcommand{\subsection}{\@startsection{subsection}{2}{0mm}%
                                {-2explus -.5ex minus -.2ex}%before -1
                                {0.5ex plus .2ex}%
                                {\normalfont\normalsize\bfseries\textcolor{WildStrawberry}}}
\renewcommand{\subsubsection}{\@startsection{subsubsection}{3}{0mm}%
                                {-2ex plus -.5ex minus -.2ex}%set spaces around sections
                                {0.3ex plus .2ex}%normally 1
                                {\normalfont\small\bfseries\textcolor{WildStrawberry}}}
\makeatother

% https://tex.stackexchange.com/questions/36030/how-to-make-a-single-word-look-as-some-code
%\definecolor{light-gray}{gray}{0.95}
%\newcommand{\code}[1]{\colorbox{light-gray}{\texttt{#1}}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\hrrule}{\vspace{2mm}\textcolor{gray}{\hrulefill}\vspace{-2.5mm}} % darkgray
\newcommand{\itxt}[1]{\textcolor{gray}{\textbf{#1}}}

% https://groups.google.com/g/knitr/c/mgtuTyBzkaA
\renewenvironment{knitrout}{\setlength{\topsep}{0mm}}{}

% -----------------------------------------------------------------------

\begin{document}

<<setup, echo=FALSE, message=FALSE, warning=FALSE>>=
knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = TRUE, size = "scriptsize")
library(collapse)
library(magrittr)
library(tibble)
library(microbenchmark)
options(width=80)

iris2 <- copyv(iris, NA, NA)
@

\raggedright
% \footnotesize
\small

{
     {\fontsize{22}{30}\selectfont \textcolor{Gray}{Advanced and Fast Data Transformation with \textit{collapse}}}{\Huge\ : : CHEAT SHEET} %\\%\small{by Sebastian Krantz} %
     \vspace{2mm}
}

\begin{multicols*}{4}
%\setlength{\premulticols}{1pt}
%\setlength{\postmulticols}{1pt}
%\setlength{\multicolsep}{1pt}
%\setlength{\columnsep}{2pt}

\section{\textcolor{WildStrawberry}{Introduction}} % Basics /
%\colorbox{gray}{
\textbf{\textit{collapse}} is a C/C++ based package supporting advanced (grouped, weighted, time series, panel data and recursive) statistical operations in R, with very efficient low-level vectorizations across both groups and columns. \newline

It also offers an efficient class-agnostic approach to data transformation in R: handling matrix and data frame based objects in a uniform, object preserving, way, while ensuring seamless compatibility with \textit{dplyr} / (grouped) \textit{tibble}, \textit{data.table}, \textit{xts}, \textit{sf} and \textit{plm} classes for panel data.  \newline

\textit{collapse} provides full flexibility and control to the user for statistical programming - with multiple ways to achieve the same outcome. The package-wide default is \code{na.rm = TRUE}, and implemented at very low cost at the algorithm level.  \newline

See the documentation at \code{help("collapse-documentation")} or \href{https://sebkrantz.github.io/collapse/reference/index.html}{online}. See also the \href{https://fastverse.github.io/fastverse/}{\textit{fastverse}} package/project for a recommended set of complimentary packages.
%}

\hrrule
\section{Row/Column Arithmetic (by Reference)}
\setstretch{1.5}
Column-wise sweeping out of vectors/matrices/DFs/lists\\
\quad \code{\%cr\%}, \code{\%c+\%}, \code{\%c-\%}, \code{\%c*\%}, \code{\%c/\%} e.g. \code{Z = X \%c/\% rowSums(X)}\\
Row-wise sweeping vectors from vectors/matrices/DFs/lists\\
\quad \code{\%rr\%}, \code{\%r+\%}, \code{\%r-\%}, \code{\%r*\%}, \code{\%r/\%} e.g. \code{Z = X \%r/\% colSums(X)}\\
Standard (column-wise) math by reference (returns invisibly)\\
\quad \code{\%+=\%}, \code{\%-=\%}, \code{\%*=\%}, \code{\%/=\%} \quad e.g. \quad \code{X \%-=\% rowSums(X)}\\
Same thing, also supporting row-wise operations by reference\\
\quad \code{setop(X, "/", rowSums(X))}\\
\setstretch{1}
\quad \code{setop(X, "/", colSums(X), rowwise = TRUE)}\\

\hrrule
\section{\fontsize{9}{11}\selectfont Transform Data by (Grouped) Replacing or
Sweeping out Statistics (by Reference)}
%\vspace{1mm}
\itxt{A generalisation of rowwise operations, that also supports sweeping by groups e.g. aggregate statistics}\newline

\qquad \code{TRA(x, STATS, FUN = "-", g = NULL, set = FALSE)}\\
\ \code{setTRA(x, STATS, FUN = "-", g = NULL)}\newline

\begin{addmargin}[1em]{0em}% 2em left, 2em right
\begin{itemize}
\item[\code{STATS}] statistics matching columns of \code{x}  (i.e. aggregated vector, matrix or data frame / list)
\item[\code{FUN}] integer/string indicating transformation to perform: \\\vspace{1mm}
\resizebox{0.2\textwidth}{!}{
\hspace{-13mm}
\begin{tabular}{lll}
    \emph{Int.} & \emph{String} & \emph{Description}  \\
      0 & \code{"replace\_NA"}   & replace missing values in \code{x} \\
      1 & \code{"replace\_fill"}   & replace data and missing values in \code{x} \\
      2 & \code{"replace"} & replace data but preserve missing values in \code{x} \\
      3 & \code{"-"}   & subtract: \code{x - STATS(g)} \\
      4 & \code{"-+"}  & \code{x - STATS(g) + fmean(STATS, w = GRPN)} \\
      5 & \code{"/"}   & divide: \code{x / STATS(g)} \\
      6 & \code{"\%"}  & compute percentages: \code{x * 100/STATS(g)} \\
      7 & \code{"+"} & add: \code{x + STATS(g)} \\
      8 & \code{"*"} & multiply: \code{x * STATS(g)}  \\
      9 & \code{"\%\%"} & modulus: \code{x \%\% STATS(g)}  \\
     10 & \code{"-\%\%"} & subtract modulus: \code{x - x \%\% STATS(g)}
\end{tabular}
}
\item[\code{g}] [optional] (list of) vectors / factors or \code{GRP()} object
\item[\code{set}] TRUE transforms \code{x} by reference. \code{setTRA} is equivalent to \code{invisible(TRA(..., set = TRUE))}
\end{itemize}
\end{addmargin}

% \vfill\null
\columnbreak
% \hrrule
\section{Fast Statistical Functions}
%\vspace{1mm}
\itxt{Fast functions to perform column–wise grouped and weighted computations on matrix-like objects:}
\newline

\quad \code{fmean, fmedian, fmode, fsum, fprod, fsd, fvar} \\
\quad \code{fmin, fmax, fnth, ffirst, flast, fnobs, fndistinct} \newline

\textbf{Syntax} \newline

\quad \code{FUN(x, g = NULL, [w = NULL], TRA = NULL,} \\
\quad\qquad \code{[na.rm = TRUE], use.g.names = TRUE, }\\
\quad\qquad \code{[drop = TRUE], [nthreads = 1L])} \newline

\begin{addmargin}[1em]{0em}% 2em left, 0em right
\begin{itemize}
\item[\code{x}] vector, matrix, or (grouped) data frame / list
\item[\code{g}] [optional] (list of) vectors / factors or GRP() object
\item[\code{w}] [optional] vector of (frequency) weights
\item[\code{TRA}] [optional] operation to transform data with computed statistics (see \code{FUN} argument to \code{TRA()} and Examples)
\item[\code{drop}] drop matrix / data frame dimensions. default \code{TRUE}
% \item[\code{na.rm}] default \code{TRUE}: algorithms have branches to skip \code{NA's}
\end{itemize}
\end{addmargin}

\vspace{2mm}
\textbf{Examples}
<<>>=
fmean(AirPassengers)  # Vector
fmean(AirPassengers, w = cycle(AirPassengers))  # Weighted mean
fmean(EuStockMarkets) # Matrix
fmean(EuStockMarkets, drop = FALSE) # Don't drop dimensions
fmean(airquality)     # Data Frame (can also use drop = FALSE)
fmean(iris[1:4], g = iris$Species) # Grouped
X = iris[1:4]; g = iris$Species; w <- abs(rnorm(nrow(X)))
fmean(X, g, w)  # Grouped and weighted (random weights)
## Transfomrations: here centering data on the weighted group median
TRA(X, fmedian(X, g, w), "-", g) |> head(3)
fmedian(X, g, w, TRA = "-") |> head(3) # Same thing: more compact
fmedian(X, g, w, "-", set = TRUE) # Modify in-place (same as setTRA())
head(iris, 3) # Changed iris too, as X = iris[1:4] did a shallow copy
@

% \begin{addmargin}[2em]{0em}
% \code{fmean(data[3:5], data\$grp1, data\$weights)\\
% data \%>\% fgroup\_by(grp1) \%>\% fmean(weights)\\
% TRA(mat, fmedian(mat, g), "-", g)\\
% fmedian(mat, g, TRA = "-")  \# same thing
% }
% \end{addmargin}

<<echo=FALSE, include=FALSE>>=
iris <- iris2
@
\vspace{-1mm}

\hrrule
\section{Basic Computing with R Functions}
% \vspace{1mm}
\itxt{Apply R functions to rows or columns (by groups)}\\
\setstretch{1.5}
\code{dapply(x, FUN, ..., MARGIN = 2)} - column/row apply\\
\code{BY(x, g, FUN, ...)} - split-apply-combine computing\\
\setstretch{1}

%\vfill\null
% \columnbreak

\section{Grouping and Ordering}
% \vspace{1mm}
\itxt{Optimized functions for grouping, ordering, unique values, and for creating and interacting factors}
\newline

\code{GRP()} - create a grouping object (class 'GRP'): pass to \code{g} arg. %\newline
<<>>=
g <- GRP(iris, ~ Species) # or GRP(iris$Species) or GRP(iris["Species"])
fndistinct(iris[1:4], g)  # Pure computation without grouping overhead
@

\code{fgroup\_by()} - attach 'GRP' object to data: a class-agnostic\\ \hphantom{\code{fgroup\_by()} -} grouped frame supporting fast computations

<<>>=
mtcars |> fgroup_by(cyl, vs, am) |> ss(1:2)
# Group Stats: [N. groups | mean (sd) min-max of group sizes]
# Fast Statistical Functions also have a grouped_df method
mtcars |> fgroup_by(cyl, vs, am) |> fmedian() |> head(3)
@
%\qquad {\scriptsize \textcolor{darkgray}{\emph{Group Stats:} N. groups $|$ Mean (Std. Dev.) Min-Max of group sizes}} \newline

\code{GRPN(), fgroup\_vars(), fungroup()} - get group count,\\ \qquad grouping columns/variables, and ungroup data\\ [0.5em]
\code{qF(), qG()} - quick \code{as.factor}, and vector grouping object\\ \qquad of class 'qG': a factor-light without levels attribute\\
\setstretch{1.5}
\code{group()} - (multivariate) group id ('qG') in appearance order\\
\code{groupid()} - run-length-type group id ('qG')\\
\code{seqid()} - group-id from integer-sequences ('qG')\\
% \code{timeid()} - time variable to integer by GCD (class 'qG')\\
\code{radixorder[v]()} - (multivariate) radix-based ordering\\
\code{finteraction()} - fast factor interactions (or return 'qG')\\
\code{fdroplevels()} - fast removal of unused factor levels\\
\code{f[n]unique()} - fast unique values / rows (by \code{cols})\\
\code{gsplit()} - fast  splitting vector based on 'GRP' objects\\
\code{greorder()} - efficiently reorder \code{y = unlist(gsplit(x, g))}\\ \qquad such that \code{identical(greorder(y, g), x)}
\setstretch{1}
\newline

\textsl{collapse} optimizes grouping using both factors / 'qG' objects and 'GRP' objects. 'GRP' objects contain most information and are thus most efficient for complex computations.

<<>>=
X <- iris[1:4]; v <- as.character(iris$Species)
f <- qF(v, na.exclude = FALSE) # Adds 'na.included' class: no NA checks
gv <- group(v) # 'qG' object: first appearance order, with 'na.included'
microbenchmark(fmode(X, v), fmode(X, f), fmode(X, gv), fmode(X, g))
@
\vspace{-2mm}

\hrrule
\section{Quick Conversions}
% \vspace{1mm}
\itxt{Fast and exact conversion of common data objects} \\ [0.5em]
\code{qM(), qDF(), qDT(), qTBL()} - convert vectors, arrays, data.frames or lists to matrix, data.frame, data.table or tibble\\ [0.5em]
\code{m[r|c]tl()} - matrix rows/cols to list, data.frame or data.table\\ [0.5em]
\code{qF(), as\_numeric\_factor(), as\_character\_factor()} - convert to/from factors or all factors in a list / data.frame

% \vfill\null
% \columnbreak
% \newpage

\section{Fast Data Manipulation}
\itxt{Efficient implementations of data manipulation functions}\newline
\setstretch{1.5}
\code{fselect[<-]()} - select/replace columns\\
\code{fsubset()} - subset data (rows and columns)\\
\code{ss()} - fast alternative to \code{[}, particularly for data frames\\
\code{[row|col]order[v]()} - reorder (sort) rows and columns\\
\code{fmutate(), fsummarise()} - dplyr-like, incl. \code{across()} feature\\
\code{[f|set]transform[v][<-]()} – transform cols (by reference)\\
\code{fcompute[v]()} - compute new cols dropping existing ones\\
\code{[f|set]rename()} - rename (any object with 'names' attribute)\\
\code{[set]relabel()} - assign/change variable labels ('label' attr.)\\
\code{get\_vars[<-]()} – select/replace columns (standard eval.)\\ [0.5em]
\setstretch{1}
\code{[num|cat|char|fact|logi|date]\_vars[<-]()} - select/\\ \qquad replace columns by data type or retrieve names/indices\\ [0.5em]
\code{add\_vars[<-]()} - add or column-bind columns \newline

\textbf{Examples}
<<>>=
mtcars |> fsubset(mpg > fnth(mpg, 0.95), disp:wt, cylinders = cyl)
mtcars |> colorder(cyl, vs, am, pos = 'after') |> head(2)
mtcars |> ftransformv(disp:drat, fsd, w = wt) |> ss(1:2, 1:5)
mtcars |> fcomputev(disp:drat, fsd, g = cyl, w = wt, TRA = 1,
                    keep = .c(cyl, wt)) |> head(2)
mtcars |> fgroup_by(cyl) |> fsummarise(across(disp:drat, fsd, w = wt))
qTBL(mtcars) |> fgroup_by(cyl) |> fsummarise(across(disp:drat,
     function(x) list(Models = list(lm(disp ~., x))), .apply = FALSE))
add_vars(iris) <- num_vars(iris) |> fsum(TRA = '%') |> add_stub("pc_")
@
<<echo=FALSE, include=FALSE>>=
iris <- iris2
@

\vspace{-1mm}

\hrrule
\section{Multi-Type Aggregation}
\itxt{Convenient interface to complex multi-type aggregations}\\ \vspace{1mm}
\code{collap(data, by, FUN = fmean, catFUN = fmode, \\
\quad\qquad\ cols = NULL, w = NULL, wFUN = fsum,\\
\quad\qquad\ custom = NULL, keep.col.order = TRUE, ...)}\newline

%\textbf{Example} \\

<<>>=
# Population weighted aggregates: mean (numeric) and mode (categorical)
collap(wlddev, country + PCGDP + LIFEEX ~ income, w = ~ POP)
@

\newpage

% \hrrule
\section{Advanced Transformations}
\itxt{Common transformations in (panel data) econometrics} \newline

Scaling, Centering and Averaging\\
\code{fscale(x, g, w, na.rm = TRUE, mean = 0, sd = 1, ...)}\\
\code{fwithin(x, g, w, na.rm = TRUE, mean = 0, theta = 1,)}\\
\code{fbetween(x, g, w, na.rm = TRUE, fill = FALSE, ...)}\newline

Higher-Dimensional Centering/Avg. and Linear Prediction\\
\code{fhdwithin(x, fl, w, na.rm = TRUE,\\
\hspace{11mm}\ \      variable.wise = FALSE, ...)}\\
\code{fhdbetween(x, fl, w, na.rm = TRUE, fill = FALSE,\\
\hspace{13mm} variable.wise = FALSE, ...)}\newline

Operators (function shortcuts with extra features) \\

 \code{STD(), W(), B(), HDW(), HDB()}\newline

\textbf{Examples}

<<>>=
# Grouped scaling
iris |> fgroup_by(Species) |> fscale() |> head(3)
STD(iris, ~ Species, stub = FALSE) |> head(3) # Same thing
# Fitting country-level cubic polynomials for PCGDP, LIFEEX and POP
HDW(wlddev, PCGDP + LIFEEX + POP ~ iso3c * poly(year, 3)) |> head(3)
@

% Such as Formula input:
% STD(mtcars, mpg + carb ~ cyl, w = ~ wt)
%
% Examples:
% lm(mpg ~ hp + B(hp, cyl), mtcars) – FE lm
% STD(mtcars, by = ~ cyl, w = ~wt, stub = FALSE)
% HDW(wlddev, PCGDP ~ qF(country)*poly(year, 3))
%
% Fast grouped  bivariate regression slopes:
% mtcars |> fgroup_by(cyl) |>
%  fmutate(dmx = W(mpg)) |>
%  fsummarise(beta = fsum(dmx, hp) %/=%
%                    fsum(dmx^2))

\hrrule
\section{Linear Models}
Fast (barebones) linear model fitting with 6 different solvers\\
\code{flm(y, X, w = NULL, add.icpt = FALSE, method = "lm")}\\ [0.5em]
Fast $R^2$-based F-test of exclusion restrictions for lm's (with FE)\\
\code{fFtest(y, exc, X = NULL, w = NULL, ...)}\\ [0.5em]
Both functions also have formula interfaces:
<<>>=
flm(cbind(mpg, disp) ~ hp + carb, weights = wt, mtcars)
# Test the exclusion of cyl-dummies and hp.
fFtest(mpg ~ qF(cyl) + hp | carb + qF(am), weights = wt, mtcars)
@

<<echo=FALSE>>=
vlabels(wlddev) <- NULL
options(width = 65)
@

\hrrule
\section{Time Series and Panel Series}
\itxt{Fast and flexible indexed series and data frames}\newline

Turn DF into an 'indexed\_frame' using id and/or time vars\\
\code{data\_ix = findex\_by(data, id1, ..., time)}\\ [0.5em]
\code{data\_ix\$indexed\_series} - contains 'indexed\_series' \\ [0.5em]
\code{index\_df = findex(data\_ix)} - retrieve 'index\_df': DF of id's \\ [0.5em]
\code{with(data\_ix, findex(indexed\_series))} - deep indexation: 'index\_df' can always be retrieved from 'indexed\_series'\\ [0.5em]
\code{is\_irregular(data\_ix)} - check for irregularity in ts/panel\\ [0.5em]
\code{data = unindex(data\_ix)} - remove index\\ [0.5em]
\code{reindex(data, index = index\_df)} - reindexing\\ [0.5em] % (e.g. after \code{readRDS})\\
'indexed\_series' can be 1-or-2D atomic objects. All vectors / time series / matrices can be indexed: \\
\code{reindex(vec/mat, index = vec/index\_df)}\\ [0.5em]
\textbf{Example} \\
% , drop.index.levels = "none"
<<>>=
wldi <- wlddev |> findex_by(iso3c, year) # Balanced: 216 countries
fsubset(wldi, 1:2, iso3c, year, PCGDP:POP)
# Index stats: [N. ids] | [N. periods (tot.N. periods: (max-min)/GCD)]
LIFEEXi = wldi$LIFEEX # Indexed series
str(LIFEEXi, strict.width = "cut")
LIFEEXi[1:7] # Subsetting indexed series
c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5])) # Irregular?
@

\vspace{2mm}
\itxt{Fast functions to perform time-based computations on (irregular) time series and (unbalanced) panel data}\newline

Lags/Leads, Differences, Growth Rates and Cumulative Sums\\
\code{flag(x, n = 1, g = NULL, t = NULL, fill = NA, ...)}\\
\code{fdiff(x, n = 1, diff = 1, ..., log = FALSE, rho = 1)}\\
\code{fgrowth(x, n = 1, diff = 1,, scale = 100, power = 1)}\\
\code{fcumsum(x, g, o, na.rm = TRUE, fill = FALSE)}\newline

Operators:   \code{L(), F(), D(), Dlog(), G()} \newline

\textbf{Example: Computing Growth Rates} \\
<<echo=FALSE>>=
options(width = 70)
@

<<>>=
# Ad-hoc use: note that G() supports formulas which fgrowth() doesn't
fgrowth(AirPassengers) |> head()
G(wlddev, c(1, 10), by = PCGDP ~ iso3c, t = ~ year) |> ss(11:12)
wlddev |> fgroup_by(iso3c) |> fselect(iso3c, year, PCGDP, LIFEEX) |>
  fmutate(PCGDP_growth = fgrowth(PCGDP, t = year)) |> head(2)
settransform(wlddev, PCGDP_growth = G(PCGDP, g = iso3c, t = year))
# Note: can omit t -> requires consecutive observations and groups
# Usage with indexed series / frames:
G(wldi) |> head(2) # default: compute growth of num_vars(), keep ids
settransform(wldi, PCGDP_growth = fgrowth(PCGDP))
lm(G(PCGDP) ~ L(G(LIFEEX), 0:3), wldi) |> summary() |> coef() |> round(3)
@

% flag(vec/mat/DF, n = -1:3, g = id, t = time)
% L(DF, n = -1:2, by = ~ id1 + id2, t = ~ time)
% Use with indexed data or grouped data:
% data_ix |> L(); data_ix |> fmutate(L_var = L(var))
% lm(G(var1) ~ L(G(var2), 0:2), data_ix)
% data |> fgroup_by(id) |> flag(t = time)
% data |> gby(id) |> fmutate(Lv = L(var, t = time))
%
% Cumulative Sums:
% fcumsum(x, g = NULL, o = NULL, fill = FALSE, )
\vspace{2mm}
\code{psacf(), pspacf(), psccf()} - panel series ACF/PACF/CCF\\ [0.5em]
\code{psmat()} - panel data to array conversion/reshaping\\ % \newpage
% \vspace{-5mm}
% <<fig.height=5>>=
% psacf(G(LIFEEXi), main = NA) # ACF of growth rate of life expectancy
% @
%
% <<fig.height=5>>=
% psmat(LIFEEXi) |> plot(colours = TRUE)
% @
%

\hrrule
\section{Summary Statistics}
\code{qsu()} - fast (grouped, weighted, panel-decomposed)\\
\qquad summary statistics for cross-sectional and panel data\\ [0.5em]
\code{qtab()} - faster table() function, incl. weights\\ [0.5em]
\code{descr()} - detailed statistical description of data.frame\\ [0.5em]
\code{varying()} - check variation within groups (panel-ids)\\ [0.5em]
\code{pwcor(), pwcov(), pwnobs()} - pairwise correlations,\\ \qquad covariance and obs. (with P-value and pretty printing)

\hrrule
\section{List-Processing}
%\vspace{1mm}
\itxt{Functions to process (nested) lists (of data objects)}\\ [0.5em]
%\setstretch{1.5}
\code{ldepth()} - level of nesting of list\\ [0.5em]
\code{is\_unlistable()} - is list composed of atomic objects\\ [0.5em]
\code{has\_elem()} - search if list contains certain elements\\ [0.5em]
\code{get\_elem()} - pull out elements from list / subset list\\ [0.5em]
\code{atomic\_elem[<-](), list\_elem[<-]()} - get list with atomic /\\ \qquad sub-list elements, examining only first level of list\\ [0.5em]
\code{reg\_elem(), irreg\_elem()} - get full list tree leading to atomic\\ \qquad ('regular') or non-atomic ('irregular') elements\\ [0.5em]
\code{rsplit()} - efficient (recursive) splitting\\ [0.5em]
\code{t\_list()} - efficient list transpose (transpose lists of lists)\\ [0.5em]
\code{rapply2d()} - recursive apply to lists of data objects\\ [0.5em]
\code{unlist2d()} - recursive row-binding to data.frame\\
%\setstretch{1}

\hrrule
\section{Recode and Replace Values}
\code{recode\_num(), recode\_char()} - recode numeric / character values (+ regex recoding) in matrix-like objects
\code{replace\_NA(), replace\_Inf(), replace\_outliers()} - replace special values   \code{pad()} -  add observations / rows.

\hrrule
\section{(Memory) Efficient Programming}
\itxt{Functions to make R programming more (memory) efficient}\\
\code{anyv}, \code{allv}, \code{allNA}, \code{whichv}, \code{whichNA}, \code{\%==\%},
\code{\%!=\%}, \code{copyv}, \code{setv}, \code{alloc}, \code{setop}, \code{\%+=\%}, \code{\%-=\%}, \code{\%*=\%}, \code{\%/=\%}, \code{missing\_cases}, \code{na\_insert}, \code{na\_rm}, \code{na\_omit}, \code{vlengths}, \code{vtypes}, \code{vgcd}, \code{frange}, \code{fnlevels}, \code{fnrow}, \code{fncol}, \code{fdim}, \code{seq\_row}, \code{seq\_col}, \code{cinv}

\hrrule
\section{Small (Helper) Functions}
\itxt{Functions for (meta-)programming and dealing with attributes}\\
\code{massign}, \code{\%=\%}, \code{.c}, \code{vlabels(<-)}, \code{setLabels}, \code{vclasses}, \code{namlab}, \code{add\_stub}, \code{rm\_stub}, \code{\%!in\%}, \code{ckmatch}, \code{all\_identical}, \code{all\_obj\_equal}, \code{all\_funs}, \code{setDimnames}, \code{setRownames}, \code{setColnames}, \code{unattrib}, \code{setAttrib}, \code{copyAttrib}, \code{copyMostAttrib}, \code{is\_categorical}, \code{is\_date}

\hrrule
\section{Alternative APIs and Global Options}
\itxt{Shorthands for frequently used functions}\\ [0.5em]
\code{fselect -> slt, fsubset -> sbt, fmutate -> mtt, [f/set]transform[v] -> [set]tfm[v], fsummarise -> smr,
across -> acr, fgroup\_by -> gby, findex\_by -> iby, findex -> ix, frename -> rnm, get\_vars -> gv, num\_vars -> nv,
add\_vars -> av} \newline

\itxt{Namespace Masking}\\ [0.5em]
Can set \code{option(collpse\_mask = ...)} with functions starting with f-, masking any like-named base R or \emph{dplyr} functions. A few keywords exist to mask multiple functions: \code{"manip", "helper", "fast-fun", "fast-stat-fun", "fast-tfrm-fun"} and \code{"all"}. This allows you to write cleaner code with \emph{collapse}, but poses additional namespace challenges:

<<eval = FALSE>>=
options(collapse_mask = "all")
library(collapse)
# The folowing is 100% collapse code, apart from the base pipe

wlddev |>
  subset(year >= 1990) |>
  group_by(year) |>
  summarise(across(PCGDP:GINI, mean, w = POP),
            n = n())

sum(mtcars)
diff(EuStockMarkets)
droplevels(wlddev)
mean(nv(iris), g = iris$Species)
scale(nv(GGDC10S), g = GGDC10S$Variable)
unique(GGDC10S, cols = c("Variable", "Country"))
range(wlddev$date)

wlddev |>
  index_by(iso3c, year) |>
  mutate(PCGDP_lag = lag(PCGDP),
         PCGDP_diff = PCGDP - PCGDP_lag,
         PCGDP_growth = growth(PCGDP)) |>
  unindex()

@
The best way to set this option is inside an \code{.Rprofile} file placed in the user or project directory.\newline

For more information see \code{help("collapse-options")}.

\end{multicols*}
\end{document}

### 2022-08-21 misc/collapse cheat sheet/collapse_cheat_sheet_sweave.Rnw
Further progress. Finished page 1.

@@
 \textbf{\textit{collapse}} is a C/C++ based package supporting advanced (grouped, weighted, time series, panel data and recursive) statistical operations in R, with very efficient low-level vectorizations across both groups and columns.
-\newline
+\\ [0.8em]
 
-It also offers an efficient class-agnostic approach to data transformation in R: handling matrix and data frame based objects in a uniform, object preserving, way, while ensuring seamless compatibility with \textit{dplyr} / (grouped) \textit{tibble}, \textit{data.table}, \textit{xts}, \textit{sf} and \textit{plm} classes for panel data.
-\newline
+It also offers a flexible, class-agnostic, approach to data transformation in R: handling matrix and data frame based objects in a uniform, attribute preserving, way, and ensuring seamless compatibility with \textit{dplyr} / (grouped) \textit{tibble}, \textit{data.table}, \textit{xts}, \textit{sf} and \textit{plm} classes for panel data ('pseries', 'pdata.frame').
+\\ [0.8em]
 
-\textit{collapse} provides full flexibility and control to the user for statistical programming - with multiple ways to achieve the same outcome.
-The package-wide default is \code{na.rm = TRUE}, and implemented at very low cost at the algorithm level.
-\newline
+\textit{collapse} provides full control to the user for statistical programming - with several ways to reach the same outcome and rich optimization possibilities.
+The default is \code{na.rm = TRUE}, and implemented at low cost at the algorithm level.
+\\ [0.8em]
 
-See the documentation at \code{help("collapse-documentation")} or \href{https://sebkrantz.github.io/collapse/reference/index.html}{online}.
-See also the \href{https://fastverse.github.io/fastverse/}{\textit{fastverse}} package/project for a recommended set of complimentary packages.
+Calling \code{help("collapse-documentation")} brings up a detailed documentation, which is also available \href{https://sebkrantz.github.io/collapse/reference/index.html}{online}.
+See also the \href{https://fastverse.github.io/fastverse/}{\textit{fastverse}} package/project for a recommended set of complimentary packages and easy package management.
 %}
@@
 \quad \code{\%+=\%}, \code{\%-=\%}, \code{\%*=\%}, \code{\%/=\%} \quad e.g. \quad \code{X \%-=\% rowSums(X)}\\
-Same thing, also supporting row-wise operations by reference\\
+Same thing, also supports row-wise operations by reference\\
 \quad \code{setop(X, "/", rowSums(X))}\\
@@
 \begin{itemize}
-\item[\code{STATS}] statistics matching columns of \code{x}  (i.e. aggregated vector, matrix or data frame / list)
+\item[\code{x}] vector, matrix, or (grouped) data frame / list
+\item[\code{STATS}] statistics matching (columns of) \code{x}  (i.e. aggregated vector, matrix or data frame / list)
 \item[\code{FUN}] integer/string indicating transformation to perform: \\\vspace{1mm}
@@
 %\vspace{1mm}
-\itxt{Fast functions to perform column–wise grouped and weighted computations on matrix-like objects:}
+\itxt{Fast functions to perform column–wise grouped and weighted computations on matrix-like objects}
 \newline
@@
 % \vspace{1mm}
-\itxt{Optimized functions for grouping, ordering, unique values, and for creating and interacting factors}
+\itxt{Optimized functions for grouping, ordering, unique values, splitting \& recombining, and dealing with factors}
 \newline
@@
 g <- GRP(iris, ~ Species) # or GRP(iris$Species) or GRP(iris["Species"])
-fndistinct(iris[1:4], g)  # Pure computation without grouping overhead
+fndistinct(iris[1:4], g)  # Computation without grouping overhead
 @
@@
 # Group Stats: [N. groups | mean (sd) min-max of group sizes]
-# Fast Statistical Functions also have a grouped_df method
-mtcars |> fgroup_by(cyl, vs, am) |> fmedian() |> head(3)
+# Fast Functions also have a grouped_df method: here wt-weighted medians
+mtcars |> fgroup_by(cyl, vs, am) |> fmedian(wt) |> head(3)
 @
@@
 \code{fdroplevels()} - fast removal of unused factor levels\\
-\code{f[n]unique()} - fast unique values / rows (by \code{cols})\\
+\code{f[n]unique()} - fast unique values / rows (by columns)\\
 \code{gsplit()} - fast  splitting vector based on 'GRP' objects\\
@@
 \section{Fast Data Manipulation}
-\itxt{Efficient implementations of data manipulation functions}\newline
+\vspace{-1.4mm}
+\itxt{Efficient implementations of data manipulation functions} % \newline
 \setstretch{1.5}
@@
 \code{fmutate(), fsummarise()} - dplyr-like, incl. \code{across()} feature\\
-\code{[f|set]transform[v][<-]()} – transform cols (by reference)\\
+\code{[f|set]transform[v][<-]()} - transform cols (by reference)\\
 \code{fcompute[v]()} - compute new cols dropping existing ones\\
@@
 \code{[set]relabel()} - assign/change variable labels ('label' attr.)\\
-\code{get\_vars[<-]()} – select/replace columns (standard eval.)\\ [0.5em]
+\code{get\_vars[<-]()} - select/replace columns (standard eval.)\\ [0.5em]
 \setstretch{1}
@@
 mtcars |> colorder(cyl, vs, am, pos = 'after') |> head(2)
-mtcars |> ftransformv(disp:drat, fsd, w = wt) |> ss(1:2, 1:5)
-mtcars |> fcomputev(disp:drat, fsd, g = cyl, w = wt, TRA = 1,
-                    keep = .c(cyl, wt)) |> head(2)
-mtcars |> fgroup_by(cyl) |> fsummarise(across(disp:drat, fsd, w = wt))
-qTBL(mtcars) |> fgroup_by(cyl) |> fsummarise(across(disp:drat,
-     function(x) list(Models = list(lm(disp ~., x))), .apply = FALSE))
-add_vars(iris) <- num_vars(iris) |> fsum(TRA = '%') |> add_stub("pc_")
+i <- base::invisible # These are equivalent, the second option is faster:
+mtcars |> fgroup_by(cyl, vs, am) |> fmutate(sum_mpg = fsum(mpg)) |> i()
+mtcars |> fmutate(sum_mpg = fsum(mpg, list(cyl, vs, am), TRA = 1)) |> i()
+# These are also equivalent (weighted means), again the second is faster
+mtcars |> fgroup_by(cyl) |> fmutate(across(disp:drat, fmean, wt)) |> i()
+mtcars |> ftransformv(disp:drat, fmean, cyl, wt, 1, apply = FALSE) |> i()
+# ftransform()/fcompute() support list input and ignore attached groupings
+mtcars %>% fgroup_by(cyl) %>% ftransform(fselect(., hp:qsec) %>%
+           fmedian(TRA = 1) %>% fungroup() %>% fsum(TRA = "/")) |> i()
+# Again a faster equivalent: note the use of 'set' to avoid a deep copy
+mtcars %>% ftransform(fselect(., hp:qsec) %>% fmedian(cyl, TRA = 1) %>%
+                      fsum(TRA = "/", set = TRUE)) %>% i()
+# Aggregation: weighted standard deviations
+mtcars |> fgroup_by(vs) |> fsummarise(across(disp:drat, fsd, w = wt))
+# Grouped linear models: .apply = FALSE applies functions to DF subset
+qTBL(mtcars) |> fgroup_by(vs) |> fsummarise(across(disp:drat,
+     function(x) list(models = list(lm(disp ~., x))), .apply = FALSE))
+# Adding some columns. Use ftransform<- to also replace existing ones
+add_vars(iris) <- num_vars(iris) |> fsum(TRA = '%') |> add_stub("perc_")
 @
@@
 @
-
-\vspace{-1mm}
+\vspace{-2mm}
 
@@
 \quad\qquad\ cols = NULL, w = NULL, wFUN = fsum,\\
-\quad\qquad\ custom = NULL, keep.col.order = TRUE, ...)}\newline
+\quad\qquad\ custom = NULL, keep.col.order = TRUE, ...)}\\ [0.3em]
 
@@
 <<>>=
-# Population weighted aggregates: mean (numeric) and mode (categorical)
+# Population weighted mean (PCGDP, LIFEEX) & mode (country), and sum(POP)
 collap(wlddev, country + PCGDP + LIFEEX ~ income, w = ~ POP)
 @
+%\vspace{-10mm}
 
@@
 
+% ------------------------------------------------------------------
+% Second Page
+% ------------------------------------------------------------------
+
 % \hrrule
 \section{Advanced Transformations}
-\itxt{Common transformations in (panel data) econometrics} \newline
+\itxt{Common transformations (in econometrics)} \newline
 
 Scaling, Centering and Averaging\\
-\code{fscale(x, g, w, na.rm = TRUE, mean = 0, sd = 1, ...)}\\
-\code{fwithin(x, g, w, na.rm = TRUE, mean = 0, theta = 1,)}\\
-\code{fbetween(x, g, w, na.rm = TRUE, fill = FALSE, ...)}\newline
+\code{fscale(x, g = NULL, w = NULL, na.rm = TRUE,\\
+\hphantom{fscaleb}mean = 0, sd = 1, ...)}\\
+\code{fwithin(x, g = NULL, w = NULL, na.rm = TRUE,
+\hphantom{fwithinb}mean = 0, theta = 1, ...)}\\
+\code{fbetween(x, g = NULL, w = NULL, na.rm = TRUE,\\
+\hphantom{fbetweenb}fill = FALSE, ...)} \\ [0.5em]
 
 Higher-Dimensional Centering/Avg. and Linear Prediction\\
-\code{fhdwithin(x, fl, w, na.rm = TRUE,\\
-\hspace{11mm}\ \      variable.wise = FALSE, ...)}\\
-\code{fhdbetween(x, fl, w, na.rm = TRUE, fill = FALSE,\\
-\hspace{13mm} variable.wise = FALSE, ...)}\newline
+\code{fhdwithin(x, fl, w = NULL, na.rm = TRUE,\\
+\hphantom{fhdwithinb}fill = FALSE, lm.method = "qr", ...)}\\
+\code{fhdbetween()} - same arguments as \code{fhdwithin()} \\ [0.7em]
 
-Operators (function shortcuts with extra features) \\
+Statistical Operators (function shorthands with extra features) \\
 
- \code{STD(), W(), B(), HDW(), HDB()}\newline
+ \code{STD(), W(), B(), HDW(), HDB()} \newline
 
@@
 # Grouped scaling
-iris |> fgroup_by(Species) |> fscale() |> head(3)
-STD(iris, ~ Species, stub = FALSE) |> head(3) # Same thing
-# Fitting country-level cubic polynomials for PCGDP, LIFEEX and POP
-HDW(wlddev, PCGDP + LIFEEX + POP ~ iso3c * poly(year, 3)) |> head(3)
+iris |> fgroup_by(Species) |> fscale() |> head(2)
+STD(iris, ~ Species, stub = FALSE) |> invisible() # Same thing: faster
+# Grouped and weighted scaling. Operators support formulas and keep ids
+STD(mtcars, mpg + carb ~ cyl, w = ~ wt) |> head(2)
+# Much shorter than fsubset(mpg > fmean(mpg, cyl, TRA = 1))
+mtcars |> fsubset(mpg > B(mpg, cyl)) |> head(2)
+# Regression with cyl fixed effects - a la Mundlak (1978)
+lm(mpg ~ carb + B(carb, cyl), data = mtcars) |> coef()
+# Fast grouped (vs) bivariate regression slopes: mpg ~ carb
+mtcars |> fgroup_by(vs) |> fmutate(dm_carb = W(carb)) |>
+  fsummarise(beta = fsum(mpg, dm_carb) %/=% fsum(dm_carb^2))
+# Residuals from regressing on 'Petal' vars and 'Species' FE
+fhdwithin(iris[1:2], iris[3:5]) |> head(2)
+# Detrending with country-level cubic polynomials
+HDW(wlddev, PCGDP + LIFEEX + POP ~ iso3c * poly(year, 3)) |> head(2)
+# Note: higher-dimensional centering and polynomials requires fixest
 @
-
-% Such as Formula input:
-% STD(mtcars, mpg + carb ~ cyl, w = ~ wt)
-%
-% Examples:
-% lm(mpg ~ hp + B(hp, cyl), mtcars) – FE lm
-% STD(mtcars, by = ~ cyl, w = ~wt, stub = FALSE)
-% HDW(wlddev, PCGDP ~ qF(country)*poly(year, 3))
-%
-% Fast grouped  bivariate regression slopes:
-% mtcars |> fgroup_by(cyl) |>
-%  fmutate(dmx = W(mpg)) |>
-%  fsummarise(beta = fsum(dmx, hp) %/=%
-%                    fsum(dmx^2))
+ % \emph{Note}: for higher-dimensional centerig / averaging, \emph{collapse} imports \emph{fixest}'s C++ demeaning algorithm, if available.
+\vspace{-2mm}
 
@@
 
-\hrrule
+% \hrrule
 \section{Time Series and Panel Series}
@@
 % , drop.index.levels = "none"
-<<>>=
+<<cache=FALSE>>=
 wldi <- wlddev |> findex_by(iso3c, year) # Balanced: 216 countries
@@
 \textbf{Example: Computing Growth Rates} \\
-<<echo=FALSE>>=
+<<echo=FALSE, include=FALSE, cache=FALSE>>=
 options(width = 70)
+wldi = reindex(wldi)
 @
 
-<<>>=
+<<cache=FALSE>>=
 # Ad-hoc use: note that G() supports formulas which fgrowth() doesn't
@@
 
+% lm(G(PCGDP) ~ L(G(LIFEEX), 0:3), wldi) |> summary() |> coef() |> round(3)
 % flag(vec/mat/DF, n = -1:3, g = id, t = time)
@@
 \code{anyv}, \code{allv}, \code{allNA}, \code{whichv}, \code{whichNA}, \code{\%==\%},
-\code{\%!=\%}, \code{copyv}, \code{setv}, \code{alloc}, \code{setop}, \code{\%+=\%}, \code{\%-=\%}, \code{\%*=\%}, \code{\%/=\%}, \code{missing\_cases}, \code{na\_insert}, \code{na\_rm}, \code{na\_omit}, \code{vlengths}, \code{vtypes}, \code{vgcd}, \code{frange}, \code{fnlevels}, \code{fnrow}, \code{fncol}, \code{fdim}, \code{seq\_row}, \code{seq\_col}, \code{cinv}
+\code{\%!=\%}, \code{copyv}, \code{setv}, \code{alloc}, \code{missing\_cases}, \code{na\_insert}, \code{na\_rm}, \code{na\_omit}, \code{vlengths}, \code{vtypes}, \code{vgcd}, \code{frange}, \code{fnlevels}, \code{fnrow}, \code{fncol}, \code{fdim}, \code{seq\_row}, \code{seq\_col}, \code{cinv}
+
+<<>>=
+x <- na_insert(rnorm(1e7))
+microbenchmark(na_rm(x), x[!is.na(x)])
+microbenchmark(na.omit(wlddev), na_omit(wlddev, na.attr = TRUE))
+microbenchmark(fsubset(wlddev, year == 2010), fsubset(wlddev, year %==% 2010))
+# Efficient subassignment by reference
+with(wlddev, setv(POP, year %==% 2010, fmax(POP)))
+@
 

### 2022-08-22 misc/collapse cheat sheet/collapse_cheat_sheet_sweave.Rnw
First complete latex cheatsheet.

@@
 %% Sets page size and margins
-\usepackage[landscape,top=0.8cm,bottom=0.5cm,left=0.5cm,right=0.5cm,marginparwidth=1.75cm]{geometry}
+\usepackage[landscape,top=0.5cm,bottom=0.6cm,left=0.5cm,right=0.5cm,marginparwidth=1.75cm]{geometry}
 
@@
 \newcommand{\code}[1]{\texttt{#1}}
-\newcommand{\hrrule}{\vspace{2mm}\textcolor{gray}{\hrulefill}\vspace{-2.5mm}} % darkgray
+\newcommand{\hrrule}{\vspace{1mm}\textcolor{gray}{\hrulefill}\vspace{-2.5mm}} % darkgray
 \newcommand{\itxt}[1]{\textcolor{gray}{\textbf{#1}}}
@@
 <<setup, echo=FALSE, message=FALSE, warning=FALSE>>=
-knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = TRUE, size = "scriptsize")
+knitr::opts_chunk$set(warning = FALSE, message = FALSE, cache = FALSE, size = "scriptsize")
 library(collapse)
@@
 {
-     {\fontsize{22}{30}\selectfont \textcolor{Gray}{Advanced and Fast Data Transformation with \textit{collapse}}}{\Huge\ : : CHEAT SHEET} %\\%\small{by Sebastian Krantz} %
+     {\fontsize{22}{30}\selectfont \textcolor{Gray}{Advanced and Fast Data Transformation with \emph{collapse}}}{\Huge\ : : CHEAT SHEET} %\\%\small{by Sebastian Krantz} %
      \vspace{2mm}
@@
 %\colorbox{gray}{
-\textbf{\textit{collapse}} is a C/C++ based package supporting advanced (grouped, weighted, time series, panel data and recursive) statistical operations in R, with very efficient low-level vectorizations across both groups and columns.
+\textbf{\emph{collapse}} is a C/C++ based package supporting advanced (grouped, weighted, time series, panel data and recursive) statistical operations in R, with very efficient low-level vectorizations across both groups and columns.
 \\ [0.8em]
 
-It also offers a flexible, class-agnostic, approach to data transformation in R: handling matrix and data frame based objects in a uniform, attribute preserving, way, and ensuring seamless compatibility with \textit{dplyr} / (grouped) \textit{tibble}, \textit{data.table}, \textit{xts}, \textit{sf} and \textit{plm} classes for panel data ('pseries', 'pdata.frame').
+It also offers a flexible, class-agnostic, approach to data transformation in R: handling matrix and data frame based objects in a uniform, attribute preserving, way, and ensuring seamless compatibility with \emph{dplyr} / (grouped) \emph{tibble}, \emph{data.table}, \emph{xts}, \emph{sf} and \emph{plm} classes for panel data ('pseries', 'pdata.frame').
 \\ [0.8em]
 
-\textit{collapse} provides full control to the user for statistical programming - with several ways to reach the same outcome and rich optimization possibilities.
+\emph{collapse} provides full control to the user for statistical programming - with several ways to reach the same outcome and rich optimization possibilities.
 The default is \code{na.rm = TRUE}, and implemented at low cost at the algorithm level.
@@
 Calling \code{help("collapse-documentation")} brings up a detailed documentation, which is also available \href{https://sebkrantz.github.io/collapse/reference/index.html}{online}.
-See also the \href{https://fastverse.github.io/fastverse/}{\textit{fastverse}} package/project for a recommended set of complimentary packages and easy package management.
+See also the \href{https://fastverse.github.io/fastverse/}{\emph{fastverse}} package/project for a recommended set of complimentary packages and easy package management.
 %}
@@
 % \vfill\null
-\columnbreak
+% \columnbreak
 % \hrrule
+
 \section{Fast Statistical Functions}
@@
 @
-\vspace{-1mm}
+% \vspace{-1mm}
 
@@
 
-\textsl{collapse} optimizes grouping using both factors / 'qG' objects and 'GRP' objects. 'GRP' objects contain most information and are thus most efficient for complex computations.
+\emph{collapse} optimizes grouping using both factors / 'qG' objects and 'GRP' objects. 'GRP' objects contain most information and are thus most efficient for complex computations.
 
-<<>>=
+<<cache=TRUE>>=
 X <- iris[1:4]; v <- as.character(iris$Species)
@@
 @
-\vspace{-2mm}
+% \vspace{-2mm}
 
@@
 \code{[row|col]order[v]()} - reorder (sort) rows and columns\\
-\code{fmutate(), fsummarise()} - dplyr-like, incl. \code{across()} feature\\
+\code{fmutate(), fsummarise()} - \emph{dplyr}-like, incl. \code{across()} feature\\
 \code{[f|set]transform[v][<-]()} - transform cols (by reference)\\
@@
 
-\newpage
+% \newpage
 
@@
 \section{Advanced Transformations}
-\itxt{Common transformations (in econometrics)} \newline
-
+\itxt{Common transformations (in econometrics)} \\ [0.7em]
 Scaling, Centering and Averaging\\
@@
 iris |> fgroup_by(Species) |> fscale() |> head(2)
-STD(iris, ~ Species, stub = FALSE) |> invisible() # Same thing: faster
+STD(iris, ~ Species, stub = FALSE) |> invisible() # Same thing + faster
 # Grouped and weighted scaling. Operators support formulas and keep ids
 STD(mtcars, mpg + carb ~ cyl, w = ~ wt) |> head(2)
-# Much shorter than fsubset(mpg > fmean(mpg, cyl, TRA = 1))
+# Much shorter than fsubset(mpg > fmean(mpg, cyl, TRA = "replace"))
 mtcars |> fsubset(mpg > B(mpg, cyl)) |> head(2)
@@
 HDW(wlddev, PCGDP + LIFEEX + POP ~ iso3c * poly(year, 3)) |> head(2)
-# Note: higher-dimensional centering and polynomials requires fixest
+# Note: HD centering/prediction and polynomials requires package 'fixest'
 @
  % \emph{Note}: for higher-dimensional centerig / averaging, \emph{collapse} imports \emph{fixest}'s C++ demeaning algorithm, if available.
-\vspace{-2mm}
+% \vspace{-2mm}
 
@@
 \section{Linear Models}
+\vspace{1mm}
 Fast (barebones) linear model fitting with 6 different solvers\\
@@
 Fast $R^2$-based F-test of exclusion restrictions for lm's (with FE)\\
-\code{fFtest(y, exc, X = NULL, w = NULL, ...)}\\ [0.5em]
+\code{fFtest(y, exc, X = NULL, w = NULL, full.df = TRUE)}\newline
+
 Both functions also have formula interfaces:
@@
 <<echo=FALSE>>=
+wlddev2 <- data.table::copy(wlddev)
 vlabels(wlddev) <- NULL
-options(width = 65)
+options(width = 70)
 @
@@
 \section{Time Series and Panel Series}
-\itxt{Fast and flexible indexed series and data frames}\newline
+\itxt{Fast and flexible indexed series and data frames: a modern upgrade of \emph{plm}'s 'pseries' and 'pdata.frame'}\newline
 
@@
 \code{data\_ix = findex\_by(data, id1, ..., time)}\\ [0.5em]
-\code{data\_ix\$indexed\_series} - contains 'indexed\_series' \\ [0.5em]
-\code{index\_df = findex(data\_ix)} - retrieve 'index\_df': DF of id's \\ [0.5em]
-\code{with(data\_ix, findex(indexed\_series))} - deep indexation: 'index\_df' can always be retrieved from 'indexed\_series'\\ [0.5em]
-\code{is\_irregular(data\_ix)} - check for irregularity in ts/panel\\ [0.5em]
-\code{data = unindex(data\_ix)} - remove index\\ [0.5em]
-\code{reindex(data, index = index\_df)} - reindexing\\ [0.5em] % (e.g. after \code{readRDS})\\
-'indexed\_series' can be 1-or-2D atomic objects. All vectors / time series / matrices can be indexed: \\
+\code{data\_ix\$indexed\_series} - columns are 'indexed\_series' \\ [0.5em]
+\code{index\_df = findex(data\_ix)} - retrieve 'index\_df': DF of ids \\ [0.5em]
+\code{index\_df = with(data\_ix, findex(indexed\_series))} - can fetch 'index\_df' from 'indexed\_series' in any caller environment \\ [0.5em]
+\code{data = unindex(data\_ix)} - unindex (also 'indexed\_series') \\ [0.5em]
+\code{reindex(data, index = index\_df)} - reindex / new pointers\\ [0.5em] % (e.g. after \code{readRDS})\\
+'indexed\_series' can be 1-or-2D atomic objects. Vectors / time series / matrices can also be indexed directly using: \\
 \code{reindex(vec/mat, index = vec/index\_df)}\\ [0.5em]
-\textbf{Example} \\
+\code{is\_irregular()} - irregularity in any index[ed] obj. or time vec\newline
+
+\textbf{Example: Indexing Panel Data} \\
 % , drop.index.levels = "none"
-<<cache=FALSE>>=
+<<>>=
 wldi <- wlddev |> findex_by(iso3c, year) # Balanced: 216 countries
@@
 LIFEEXi[1:7] # Subsetting indexed series
-c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5])) # Irregular?
+c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5])) # Is irregular?
 @
+\emph{Note}: 'indexed\_series' and frames are supported via existing 'pseries'/'pdata.frame' methods for time series/panel functions. \newline
 
-\vspace{2mm}
+% \vspace{2mm}
 \itxt{Fast functions to perform time-based computations on (irregular) time series and (unbalanced) panel data}\newline
@@
 \code{flag(x, n = 1, g = NULL, t = NULL, fill = NA, ...)}\\
-\code{fdiff(x, n = 1, diff = 1, ..., log = FALSE, rho = 1)}\\
-\code{fgrowth(x, n = 1, diff = 1,, scale = 100, power = 1)}\\
-\code{fcumsum(x, g, o, na.rm = TRUE, fill = FALSE)}\newline
+\code{fdiff(x, n = 1, diff = 1, g = NULL, t = NULL,\\
+\hphantom{fdiffd}fill = NA, log = FALSE, rho = 1, ...)}\\
+\code{fgrowth(x, n = 1, diff = 1, g = NULL, t = NULL, fill\\ \ = NA, logdiff = FALSE, scale = 100, power = 1, ...)}\\
+\code{fcumsum(x, g = NULL, o = NULL, na.rm = TRUE,\\
+\hphantom{fcumsumd}fill = FALSE, check.o = TRUE, ...)}\\ [0.7em]
 
-Operators:   \code{L(), F(), D(), Dlog(), G()} \newline
+Statistical Operators:   \code{L(), F(), D(), Dlog(), G()} \newline
 
 \textbf{Example: Computing Growth Rates} \\
-<<echo=FALSE, include=FALSE, cache=FALSE>>=
-options(width = 70)
+<<echo=FALSE, include=FALSE>>=
+# options(width = 70)
 wldi = reindex(wldi)
 @
-
-<<cache=FALSE>>=
+<<>>=
 # Ad-hoc use: note that G() supports formulas which fgrowth() doesn't
@@
 # Usage with indexed series / frames:
+@
+<<>>=
 G(wldi) |> head(2) # default: compute growth of num_vars(), keep ids
 settransform(wldi, PCGDP_growth = fgrowth(PCGDP))
-lm(G(PCGDP) ~ L(G(LIFEEX), 0:3), wldi) |> summary() |> coef() |> round(3)
+lm(G(PCGDP) ~ L(G(LIFEEX), 0:2), wldi) |> summary() |> coef() |> round(3)
 @
@@
 % fcumsum(x, g = NULL, o = NULL, fill = FALSE, )
-\vspace{2mm}
-\code{psacf(), pspacf(), psccf()} - panel series ACF/PACF/CCF\\ [0.5em]
+% \vspace{2mm}
+\code{psacf(), pspacf(), psccf()} - panel series ACF/PACF/CCF\\ [0.3em]
 \code{psmat()} - panel data to array conversion/reshaping\\ % \newpage
@@
 %
-
+% \vspace{-1mm}
 \hrrule
@@
 \code{qsu()} - fast (grouped, weighted, panel-decomposed)\\
-\qquad summary statistics for cross-sectional and panel data\\ [0.5em]
-\code{qtab()} - faster table() function, incl. weights\\ [0.5em]
+\qquad summary statistics for cross-sectional and panel data\\
+<<>>=
+# Panel data statistics: overall, on group-means and group-centered data
+qsu(iris, pid = Sepal.Length ~ Species, higher = TRUE)
+@
+\code{qtab()} - faster \code{table()} function, incl. weights \& custom funs\\ [0.5em]
 \code{descr()} - detailed statistical description of data.frame\\ [0.5em]
@@
 \code{pwcor(), pwcov(), pwnobs()} - pairwise correlations,\\ \qquad covariance and obs. (with P-value and pretty printing)
+% \vspace{-1mm}
 
@@
 \code{get\_elem()} - pull out elements from list / subset list\\ [0.5em]
-\code{atomic\_elem[<-](), list\_elem[<-]()} - get list with atomic /\\ \qquad sub-list elements, examining only first level of list\\ [0.5em]
+\code{atomic\_elem[<-](), list\_elem[<-]()} - get list with atomic /\\ \qquad sub-list elements, examining only first level of list\\ [0.3em]
 \code{reg\_elem(), irreg\_elem()} - get full list tree leading to atomic\\ \qquad ('regular') or non-atomic ('irregular') elements\\ [0.5em]
@@
 \code{rapply2d()} - recursive apply to lists of data objects\\ [0.5em]
-\code{unlist2d()} - recursive row-binding to data.frame\\
+\code{unlist2d()} - recursive row-binding to data.frame\\ [0.5em]
 %\setstretch{1}
 
-\hrrule
+\textbf{Example: Nested Linear Models} \\ %\vspace{-1mm}
+<<>>=
+(dl <- mtcars |> rsplit(mpg + hp + carb ~ vs + am)) |> str(max.level = 2)
+nest_lm <- dl |> rapply2d(lm, formula = mpg ~ .)
+(nest_coef <- nest_lm |> rapply2d(summary, classes = "lm") |>
+     get_elem("coefficients")) |> str(give.attr = FALSE, strict = "cut")
+nest_coef |> unlist2d(c("vs", "am"), row.names = "variable") |> head(2)
+@
+% \vspace{-2mm}
+% \columnbreak
+
+% \hrrule
 \section{Recode and Replace Values}
-\code{recode\_num(), recode\_char()} - recode numeric / character values (+ regex recoding) in matrix-like objects
-\code{replace\_NA(), replace\_Inf(), replace\_outliers()} - replace special values   \code{pad()} -  add observations / rows.
+\code{recode\_num(), recode\_char()} - recode numeric / character values (+ regex recoding) in matrix-like objects\\ [0.5em]
+\code{replace\_[NA|Inf|outliers]()} - replace special values\\ [0.5em]
+\code{pad()} -  add (missing) observations / rows i.e. expand objects
+% \vspace{-1mm}
 
@@
 \section{(Memory) Efficient Programming}
-\itxt{Functions to make R programming more (memory) efficient}\\
-\code{anyv}, \code{allv}, \code{allNA}, \code{whichv}, \code{whichNA}, \code{\%==\%},
-\code{\%!=\%}, \code{copyv}, \code{setv}, \code{alloc}, \code{missing\_cases}, \code{na\_insert}, \code{na\_rm}, \code{na\_omit}, \code{vlengths}, \code{vtypes}, \code{vgcd}, \code{frange}, \code{fnlevels}, \code{fnrow}, \code{fncol}, \code{fdim}, \code{seq\_row}, \code{seq\_col}, \code{cinv}
+\itxt{Functions for (memory) efficient R programming}\\ [0.5em]
 
-<<>>=
-x <- na_insert(rnorm(1e7))
-microbenchmark(na_rm(x), x[!is.na(x)])
-microbenchmark(na.omit(wlddev), na_omit(wlddev, na.attr = TRUE))
-microbenchmark(fsubset(wlddev, year == 2010), fsubset(wlddev, year %==% 2010))
-# Efficient subassignment by reference
-with(wlddev, setv(POP, year %==% 2010, fmax(POP)))
+\code{any|all[v|NA]}, \code{which[v|NA]}, \code{\%[=|!]=\%}, \code{copyv}, \code{setv}, \code{alloc} \code{missing\_cases}, \code{na\_[insert|rm|omit]}, \code{vlengths}, \code{vtypes}, \code{vgcd}, \code{frange}, \code{fnlevels}, \code{fn[row|col]}, \code{fdim}, \code{seq\_[row|col]}\\
+<<eval = FALSE>>=
+fsubset(wlddev, year %==% 2010) # 2x faster fsubset(wlddev, year == 2010)
+attach(mtcars) # Efficient sub-assignment by reference, various options...
+setv(am, 0, vs); setv(am, 1:10, vs); setv(am, 1:10, vs[10:20])
 @
+% x <- na_insert(rnorm(1e7))
+% microbenchmark(na_rm(x), x[!is.na(x)])
+% microbenchmark(na.omit(wlddev), na_omit(wlddev, na.attr = TRUE))
+% # Efficient subassignment by reference
+% with(wlddev, setv(POP, year %==% 2010, fmax(POP)))
+% microbenchmark(A = fsubset(wlddev, year == 2010), B = fsubset(wlddev, year %==% 2010))
+% x <- rnorm(1e7); xNA <- na_insert(x)
+% microbenchmark(setv(xNA, NA, x), )
+% with(wlddev, setv(POP, year %==% 2010, fmax(POP)))
+% \vspace{-1mm}
 
@@
 \section{Small (Helper) Functions}
-\itxt{Functions for (meta-)programming and dealing with attributes}\\
-\code{massign}, \code{\%=\%}, \code{.c}, \code{vlabels(<-)}, \code{setLabels}, \code{vclasses}, \code{namlab}, \code{add\_stub}, \code{rm\_stub}, \code{\%!in\%}, \code{ckmatch}, \code{all\_identical}, \code{all\_obj\_equal}, \code{all\_funs}, \code{setDimnames}, \code{setRownames}, \code{setColnames}, \code{unattrib}, \code{setAttrib}, \code{copyAttrib}, \code{copyMostAttrib}, \code{is\_categorical}, \code{is\_date}
+\itxt{Functions for (meta-)programming and attributes}\\ [0.5em]
+\code{.c}, \code{massign}, \code{\%=\%}, \code{vlabels[<-]}, \code{setLabels}, \code{vclasses}, \code{namlab}, \code{[add|rm]\_stub}, \code{\%!in\%}, \code{ckmatch}, \code{all\_identical}, \code{all\_obj\_equal}, \code{all\_funs}, \code{set[Dim|Row|Col]names}, \code{unattrib}, \code{setAttrib}, \code{copyAttrib}, \code{copyMostAttrib} %, \code{is\_categorical}, \code{is\_date}
+
+<<include=FALSE, echo=FALSE>>=
+wlddev <- wlddev2
+@
+<<>>=
+.c(var1, var2, var3) # Non-standard concatenation
+.c(values, vectors) %=% eigen(cov(mtcars)) # Multiple Assignment
+# Variable labels: vlabels[<-], [set]relabel() etc. namlab() shows summary
+namlab(wlddev[c(2, 9)], N = TRUE, Ndist = TRUE, class = TRUE)
+@
+% \vspace{-1mm}
 
 \hrrule
-\section{Alternative APIs and Global Options}
+\section{API Extensions}
 \itxt{Shorthands for frequently used functions}\\ [0.5em]
 \code{fselect -> slt, fsubset -> sbt, fmutate -> mtt, [f/set]transform[v] -> [set]tfm[v], fsummarise -> smr,
-across -> acr, fgroup\_by -> gby, findex\_by -> iby, findex -> ix, frename -> rnm, get\_vars -> gv, num\_vars -> nv,
+across -> acr, fgroup\_by -> gby, finteraction -> itn, findex\_by -> iby, findex -> ix, frename -> rnm, get\_vars -> gv, num\_vars -> nv,
 add\_vars -> av} \newline
@@
 \itxt{Namespace Masking}\\ [0.5em]
-Can set \code{option(collpse\_mask = ...)} with functions starting with f-, masking any like-named base R or \emph{dplyr} functions.
-A few keywords exist to mask multiple functions: \code{"manip", "helper", "fast-fun", "fast-stat-fun", "fast-tfrm-fun"} and \code{"all"}.
-This allows you to write cleaner code with \emph{collapse}, but poses additional namespace challenges:
+Can set \code{option(collpse\_mask = c(...))} with a vector of functions starting with f-, to export versions without f-, masking base R or \emph{dplyr}.
+A few keywords exist to mask multiple functions, see \code{help("collapse-options")}.
+This allows clean \& fast code, but poses additional namespace challenges:
 
 <<eval = FALSE>>=
+# Masking all f- functions and specials n = GRPN and table = qtab
 options(collapse_mask = "all")
@@
   group_by(year) |>
-  summarise(across(PCGDP:GINI, mean, w = POP),
-            n = n())
+  summarise(n = n(), across(PCGDP:GINI, mean, w = POP))
 
+with(mtcars, table(cyl, vs, am))
 sum(mtcars)
@@
          PCGDP_diff = PCGDP - PCGDP_lag,
-         PCGDP_growth = growth(PCGDP)) |>
-  unindex()
+         PCGDP_growth = growth(PCGDP)) |> unindex()
 
 @
-The best way to set this option is inside an \code{.Rprofile} file placed in the user or project directory.\newline
-
-For more information see \code{help("collapse-options")}.
+The best way to set this option is inside an \code{.Rprofile} file placed in the user or project directory. Use it carefully.
 

### 2022-08-23 misc/collapse cheat sheet/collapse_cheat_sheet.Rnw
Adding footer to cheatsheet, and reorganizing folder a bit.

--- 2022-08-22 misc/collapse cheat sheet/collapse_cheat_sheet_sweave.Rnw
@@
 %% Sets page size and margins
-\usepackage[landscape,top=0.5cm,bottom=0.6cm,left=0.5cm,right=0.5cm,marginparwidth=1.75cm]{geometry}
+\usepackage[landscape,top=0.5cm,bottom=-1cm,left=0.5cm,right=0.5cm,marginparwidth=1.75cm]{geometry}
 
@@
 \pagestyle{empty}
+% \usepackage{fancyhdr}
+% \rfoot{Page \thepage}
 
@@
 
-\begin{multicols*}{4}
+%\begin{adjustbox}{totalheight=0.5\textheight} % -2\baselineskip
+% \resizebox*{!}{\textheight}{%
+\begin{multicols}{4}
 %\setlength{\premulticols}{1pt}
@@
 \emph{collapse} provides full control to the user for statistical programming - with several ways to reach the same outcome and rich optimization possibilities.
-The default is \code{na.rm = TRUE}, and implemented at low cost at the algorithm level.
+Its default is \code{na.rm = TRUE}, and implemented at very low cost at the algorithm level.
 \\ [0.8em]
@@
 @
-%\vspace{-10mm}
+\end{multicols} % \vspace{-20mm}
 
-% \newpage
+% \end{adjustbox}
+%}
+% \hrrule
+\vspace{-5mm}
+\textcolor{lightgray}{\hrulefill}\\
+{\scriptsize \vspace{-0.5mm}
+ Page 1 of 2 \hfill \textcopyright\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://sebkrantz.github.io/collapse/}{sebkrantz.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 1.8.8\ \textbullet\ Updated: 2022-08
+}
+
+\newpage
 
@@
 % ------------------------------------------------------------------
+\begin{multicols}{4}
 
@@
 \hrrule
-\section{List-Processing}
+\section{List Processing}
 %\vspace{1mm}
@@
 
-\end{multicols*}
+\end{multicols}
+
+\vspace{-5mm}
+\textcolor{lightgray}{\hrulefill}\\
+{\scriptsize \vspace{-0.5mm}
+ Page 2 of 2 \hfill \textcopyright\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://sebkrantz.github.io/collapse/}{sebkrantz.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 1.8.8\ \textbullet\ Updated: 2022-08
+}
+
 \end{document}

### 2023-10-14 misc/collapse cheat sheet/collapse_cheat_sheet.Rnw
Adding background pic.
Update cheatsheet page 1 for v2.0.

--- 2022-08-22 misc/collapse cheat sheet/collapse_cheat_sheet_sweave.Rnw
@@
 
+% https://tex.stackexchange.com/questions/46280/how-to-create-a-background-image-on-title-page-with-latex
+\usepackage{eso-pic}
+\newcommand\BackgroundPic{%
+\put(0,0){%
+\parbox[b][\paperheight]{\paperwidth}{%
+\vspace{-3mm} % \vfill
+\flushright
+\includegraphics[width=0.3\paperwidth,%
+keepaspectratio]{"background_cropped.png"}%
+\vfill
+}}}
+
+\newcommand\Logo{%
+\put(0,0){%
+\parbox[b][\paperheight]{\paperwidth}{%
+\vspace{-1mm} % \vfill
+\flushright
+\includegraphics[width=0.08\paperwidth,%
+keepaspectratio]{"collapse_logo_vsmall.png"} \hspace{3mm} %
+\vfill
+}}}
+
 \begin{document}
+
+% \makebox[0pt][r]{%
+%   \raisebox{-\totalheight}[0pt][0pt]{%
+%     \includegraphics[width=4in]{background}}}%
+\AddToShipoutPicture*{\BackgroundPic}
+\AddToShipoutPicture*{\Logo}
 
@@
 {
-     {\fontsize{22}{30}\selectfont \textcolor{Gray}{Advanced and Fast Data Transformation with \emph{collapse}}}{\Huge\ : : CHEAT SHEET} %\\%\small{by Sebastian Krantz} %
-     \vspace{2mm}
+     {\fontsize{22}{30}\selectfont \textcolor{Gray}{Advanced and Fast Data Transformation with \emph{collapse}}}{\Huge\ \textcolor{darkgray}{: : CHEAT SHEET}} %\\%\small{by Sebastian Krantz} %
+     % \vspace{2mm}
 }
 
-\begin{multicols*}{4}
+%\begin{adjustbox}{totalheight=0.5\textheight} % -2\baselineskip
+% \resizebox*{!}{\textheight}{%
+\begin{multicols}{4}
 %\setlength{\premulticols}{1pt}
@@
 
-It also offers a flexible, class-agnostic, approach to data transformation in R: handling matrix and data frame based objects in a uniform, attribute preserving, way, and ensuring seamless compatibility with \emph{dplyr} / (grouped) \emph{tibble}, \emph{data.table}, \emph{xts}, \emph{sf} and \emph{plm} classes for panel data ('pseries', 'pdata.frame').
+It also offers a flexible, class-agnostic, approach to data transformation in R: handling matrix and data frame based objects in a uniform, attribute preserving, way, and ensuring seamless compatibility with base R, \emph{dplyr} / (grouped) \emph{tibble}, \emph{data.table}, \emph{xts/zoo}, \emph{sf}, and \emph{plm} classes for panel data.
 \\ [0.8em]
@@
 \emph{collapse} provides full control to the user for statistical programming - with several ways to reach the same outcome and rich optimization possibilities.
-The default is \code{na.rm = TRUE}, and implemented at low cost at the algorithm level.
+It is globally configurable using \code{set\_collapse()} which includes algorithm defaults, multithreading, and the exported namespace (see below).
 \\ [0.8em]
@@
 
-\quad \code{fmean, fmedian, fmode, fsum, fprod, fsd, fvar} \\
+\quad \code{fmean, fmedian, fmode, fsum, fprod, fsd, fvar,} \\
 \quad \code{fmin, fmax, fnth, ffirst, flast, fnobs, fndistinct} \newline
@@
 fmean(EuStockMarkets) # Matrix
-fmean(EuStockMarkets, drop = FALSE) # Don't drop dimensions
-fmean(airquality)     # Data Frame (can also use drop = FALSE)
+fmean(airquality)     # Data Frame (use drop = FALSE to keep frame)
 fmean(iris[1:4], g = iris$Species) # Grouped
@@
 ## Transfomrations: here centering data on the weighted group median
-TRA(X, fmedian(X, g, w), "-", g) |> head(3)
-fmedian(X, g, w, TRA = "-") |> head(3) # Same thing: more compact
+TRA(X, fmedian(X, g, w), "-", g) |> head(2)
+fmedian(X, g, w, TRA = "-") |> head(2) # Same thing: more compact
 fmedian(X, g, w, "-", set = TRUE) # Modify in-place (same as setTRA())
-head(iris, 3) # Changed iris too, as X = iris[1:4] did a shallow copy
 @
-
 % \begin{addmargin}[2em]{0em}
@@
 % \end{addmargin}
+\vspace{-1mm}
+\hrrule
+\section{Other Statistical Functions}
+% \vspace{1mm}
+\itxt{Fast (weighted) sample quantiles, range, and distances}\\
+\setstretch{1.5}
+\code{fquantile(x, probs, w, o, na.rm = TRUE, type = 7)}\\
+\code{frange(x, na.rm = TRUE)} \\
+\code{fdist(x, v, method = "euclidean", nthreads = 1)}\\
+\setstretch{1}
 
@@
 % \vspace{1mm}
-\itxt{Optimized functions for grouping, ordering, unique values, splitting \& recombining, and dealing with factors}
+\itxt{Optimized functions for grouping, ordering, unique values, matching, splitting, and dealing with factors}
 \newline
@@
 # Fast Functions also have a grouped_df method: here wt-weighted medians
-mtcars |> fgroup_by(cyl, vs, am) |> fmedian(wt) |> head(3)
+mtcars |> fgroup_by(cyl, vs, am) |> fmedian(wt) |> head(2)
 @
@@
 
-\code{GRPN(), fgroup\_vars(), fungroup()} - get group count,\\ \qquad grouping columns/variables, and ungroup data\\ [0.5em]
+\code{GRPN(), fcount[v](), fgroup\_vars(), fungroup()} - get group count, grouping columns, and ungroup data\\ [0.5em]
 \code{qF(), qG()} - quick \code{as.factor}, and vector grouping object\\ \qquad of class 'qG': a factor-light without levels attribute\\
@@
 \code{fdroplevels()} - fast removal of unused factor levels\\
-\code{f[n]unique()} - fast unique values / rows (by columns)\\
+\code{f[n]unique(), fduplicated()} - fast unique values / rows\\
+\code{fmatch(), \%[!][i]in\%} - fast matching of values / rows\\
 \code{gsplit()} - fast  splitting vector based on 'GRP' objects\\
@@
 @
-% \vspace{-2mm}
+\vspace{-1mm}
 
 \hrrule
+\vspace{-2mm}
 \section{Quick Conversions}
-% \vspace{1mm}
 \itxt{Fast and exact conversion of common data objects} \\ [0.5em]
@@
 \vspace{-1.4mm}
-\itxt{Efficient implementations of data manipulation functions} % \newline
+\itxt{Minimal overhead implementations} \newline
 \setstretch{1.5}
@@
 \code{[num|cat|char|fact|logi|date]\_vars[<-]()} - select/\\ \qquad replace columns by data type or retrieve names/indices\\ [0.5em]
-\code{add\_vars[<-]()} - add or column-bind columns \newline
+\code{add\_vars[<-]()} - add or column-bind columns\\
+\code{rowbind()} - row-bind lists / data frame-like objects\\ [0.5em]
+\code{join(), pivot()} - join and reshape data frame-like objects \newline
 
@@
 mtcars |> fgroup_by(vs) |> fsummarise(across(disp:drat, fsd, w = wt))
-# Grouped linear models: .apply = FALSE applies functions to DF subset
-qTBL(mtcars) |> fgroup_by(vs) |> fsummarise(across(disp:drat,
-     function(x) list(models = list(lm(disp ~., x))), .apply = FALSE))
+# Grouped linear models (one way of doing it)
+qTBL(mtcars) |> fgroup_by(vs) |> fsummarise(reg = list(lm(disp ~ drat)))
 # Adding some columns. Use ftransform<- to also replace existing ones
@@
 @
-% \vspace{-2mm}
+\vspace{-2mm}
 
@@
 @
-%\vspace{-10mm}
+%\vspace{-2mm}
+\end{multicols} % \vspace{-20mm}
+%\vspace{20mm}
 
-% \newpage
+% \end{adjustbox}
+%}
+% \hrrule
+\vspace{-5mm}
+\textcolor{lightgray}{\hrulefill}\\
+{\scriptsize \vspace{-0.5mm}
+ Page 1 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://sebkrantz.github.io/collapse/}{sebkrantz.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.0\ \textbullet\ Updated: 2023-10
+}
+
+\newpage
 
@@
 
-\itxt{Namespace Masking}\\ [0.5em]
+\itxt{Namespace masking}\\ [0.5em]
 Can set \code{option(collpse\_mask = c(...))} with a vector of functions starting with f-, to export versions without f-, masking base R or \emph{dplyr}.
@@
 
-\end{multicols*}
+\end{multicols}
+
+\vspace{-5.5mm}
+\textcolor{lightgray}{\hrulefill}\\
+{\scriptsize \vspace{-0.5mm}
+ Page 2 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://sebkrantz.github.io/collapse/}{sebkrantz.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 1.8.8\ \textbullet\ Updated: 2022-08
+}
+
 \end{document}

### 2024-01-01 misc/JSS article/jss-article-rnw/article.Rnw
Adding JSS article.

\documentclass[article]{jss}

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern}

%% another package (only for this demo article)
\usepackage{framed}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, cache=TRUE}
% \renewenvironment{knitrout}{\setlength{\topsep}{0mm}}{}

<<preliminaries, echo=FALSE, message=FALSE, warning=FALSE, results=hide>>=
# knitr::opts_chunk$set(cache = FALSE, size = "small")
options(prompt = "R> ", continue = "+  ", width = 80, digits = 4, useFancyQuotes = FALSE, warn = 1)
library(data.table)
library(magrittr)
library(microbenchmark)
library(collapse)
@

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation (and optionally ORCID link)
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Sebastian Krantz~\orcidlink{0000-0001-6212-5229}\\Kiel Institute for the World Economy}
\Plainauthor{Sebastian Krantz}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}}
\Plaintitle{collapse: Advanced and Fast Statistical Computing and Data Transformation in R}
\Shorttitle{\proglang{collapse}: Advanced and Fast Data Transformation in \proglang{R}}

%% - \Abstract{} almost as usual
\Abstract{
\pkg{collapse} is a large C/C++-based infrastructure package for \proglang{R} facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency. It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g. \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling seamless integration with large parts of the \proglang{R} ecosystem. This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples. A small benchmark demonstrates its computational performance.
%  This short article illustrates how to write a manuscript for the
%  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
%  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
%  it is recommended to keep the {\LaTeX} code as simple as possible,
%  i.e., avoid inclusion of packages/commands that are not necessary.
%  For outlining the typical structure of a JSS article some brief text snippets
%  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
%  discussing count data regression in \proglang{R}. Editorial comments and
%  instructions are marked by vertical bars.
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{statistical computing, vectorization, data manipulation and transformation, summary statistics, class-agnostic programming, \proglang{R}}
\Plainkeywords{statistical computing, vectorization, data transformation and manipulation, summary statistics, class-agnostic programming, R}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Sebastian Krantz\\
  Kiel Institute for the World Economy\\
  Haus Welt-Club\\
  D\"usternbrooker Weg 148\\
  24105 Kiel, Germany\\
  E-mail: \email{sebastian.krantz@ifw-kiel.de}
}

\begin{document}

%% -- Introduction -------------------------------------------------------------

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
%% - If such markup is in (sub)section titles, a plain text version has to be
%%   added as well.
%% - All software mentioned should be properly \cite-d.
%% - All abbreviations should be introduced.
%% - Unless the expansions of abbreviations are proper names (like "Journal
%%   of Statistical Software" above) they should be in sentence case (like
%%   "generalized linear models" below).

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions. Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently. It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools. This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures and powerful data manipulation functions. The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors. \pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g. \emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}). It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc. with much faster \pkg{collapse} equivalents.} \newline

What is the purpose of combining all of this in a package? The short answer is to make computations in \proglang{R} as flexible and powerful as possible. The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.}; (3) increase the memory efficiency and flexibility of \proglang{R} programs\footnote{E.g. by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.}; and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline

R already has a large and tested data manipulation and statistical computing ecosystem. Notably, the \pkg{tidyverse} \citep{rtidyverse} provides a consistent toolkit for data manipulation in R, centered around the \class{tibble} \citep{rtibble} object and tidy data principles \citep{rtidydata}. \pkg{data.table} \citep{rdatatable} provides an enhanced high-performance data frame with parsimonious data manipulation syntax. \pkg{sf} \citep{rsf} provides a data frame for spatial data and supporting functionality. \pkg{tsibble} \citep{rtsibble} and \pkg{xts} \citep{rxts} provide classes and operations for time series data, the former via an enhanced \class{tibble}, the latter through an efficient matrix-based class. Econometric packages like \pkg{plm} \citep{rplm} and \pkg{fixest} \citep{rfixest} also provide solutions to deal with panel data and irregularity in the time dimension. Packages like \pkg{matrixStats} \citep{rmatrixstats} and \pkg{Rfast} \citep{rfast} offer fast statistical calculations along the rows and columns of matrices and faster basic statistical procedures. \pkg{DescTools} \citep{rdesctools} provides a wide variety of descriptive statistics, including weighted versions. \pkg{survey} \citep{rsurvey} allows statistical computations on complex survey data. \pkg{labelled} \citep{rlabelled} provides tools to deal with labelled data. Packages like \pkg{tidyr} \citep{rtidyr}, \pkg{purrr} \citep{rpurrr} and \pkg{rrapply} \citep{rrapply} provide some functions to deal with nested data and messy structures. \newline

\pkg{collapse} relates to and integrates key elements from these projects. It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object. It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest} and \pkg{xts}. It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}. Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and C-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal. It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}. It provides general (recursive) tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr} and \pkg{rrapply}. Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes and requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics. In summary, \pkg{collapse} is a foundation package for statistical computing and data manipulation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient. A significant benefit is that, rather than piecing together a somewhat fragmented ecosystem oriented at different classes and tasks, core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages. This tends to result in \proglang{R} scripts that are shorter, more efficient, and more lightweight in dependencies. \newline
% TODO: Need examples of this

Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars} and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy} or \pkg{StatsBase.jl} \citep{jlstatsbase}. % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recursive operations, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable. In short, it is very utile for complex statistical workflows, rich datasets (e.g. surveys), and for integrating with different parts of the \proglang{R} ecosystem. On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive). There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.}. Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function. However, as demonstrated below, the possibility of combining vectorized statistical functions also permits calculating more complex statistics in a vectorized way. \newline

The package has a built-in structured \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{documentation} facilitating its use. This documentation includes a central \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and supplementary topic pages which briefly describe related functionality. The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:
%
<<collapse_topics>>=
.COLLAPSE_TOPICS
help("collapse-documentation")
@
%
\pkg{collapse} is too large and complex to fully present it in a single article or even to present selected topics in depth. The following sections therefore briefly introduce its key components: (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) joins and reshaping; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options (configurability). Section (\ref{sec:bench}) provides a small benchmark, Section (\ref{sec:conclusion}) concludes. For deeper engagement with \pkg{collapse}, the short vignette summarizing available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{Documentation and Resources} is an excellent starting point.
%
\section{Fast statistical functions} \label{sec:fast_stat_fun}
%
The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{Fast Statistical Functions}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs} and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R. Specifically, operations such as calculating the mean via the S3 generic \code{fmean()} function are vectorized across columns and groups. They may also involve weights or transformations of the original data. The basic syntax of these functions is
\begin{Code}
FUN(x, g = NULL, [w = NULL,] TRA = NULL, [na.rm = TRUE,]
    use.g.names = TRUE, drop = TRUE, [nthreads = 1L,] ...)
\end{Code}
with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e. simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation). The use of Single Instruction Multiple Data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.}. The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there, not here for parsimony) demonstrate their basic usage.
\begin{Code}
fmean(mtcars$mpg)                                  # Vector
fmean(EuStockMarkets)                              # Matrix
fmean(mtcars)                                      # Data Frame
fmean(mtcars$mpg, w = mtcars$wt)                   # Weighted mean
fmean(mtcars$mpg, g = mtcars$cyl)                  # Grouped mean
fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)   # Weighted group mean
fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt) # Of data frame
fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")    # Replace data by g. mean
\end{Code}
\subsection{Transformations}
The \code{TRA} argument, more specifically, invokes the \fct{TRA} function for column-wise (grouped) replacing and sweeping operations (by reference). Its syntax is
\begin{Code}
TRA(x, STATS, FUN = "-", g = NULL, set = FALSE, ...)
\end{Code}
where \code{STATS} is a vector/matrix/data.frame of statistics used to transform \code{x}. The \code{FUN} argument supports 11 different operations indicated using either an integer or string.
%
\begin{table}[h!]
\resizebox{\textwidth}{!}{
  \begin{tabular}{lllll}
  \emph{Int} && \emph{String} && \emph{Description}  \\
  0 && \code{"replace\_na"/"na"}   && replace missing values in \code{x} \\
  1 && \code{"replace\_fill"/"fill"}   && replace data and missing values in \code{x} \\
  2 && \code{"replace"} && replace data but preserve missing values in \code{x} \\
  3 && \code{"-"}   && subtract (i.e. center) \\
  4 && \code{"-+"}  && center on overall average statistic \\
  5 && \code{"/"}   && divide (i.e. scale) \\
  6 && \code{"\%"}     && compute percentages (i.e. divide and multiply by 100) \\
  7 && \code{"+"} && add \\
  8 && \code{"*"} && multiply \\
  9 && \code{"\%\%"} && modulus (i.e. remainder from division by  \code{STATS}) \\
  10 && \code{"-\%\%"} && subtract modulus (i.e. make data divisible by \code{STATS})
  \end{tabular}
}
\end{table}
%
\code{TRA()} is called internally in the \emph{Fast Statistical Functions}, the \code{TRA} argument is passed to \code{FUN}. Thus \code{fmean(x, g, w, TRA = "-")} is equivalent to \code{TRA(x, fmean(x, g, w), "-", g)}. The \code{set} argument can also be passed to \emph{Fast Statistical Functions} to toggle transformation by reference. The following examples demonstrate how this design allows flexible ad-hoc transformations using \proglang{R}'s built-in \code{airquality} dataset with daily measurements for 5 months.
%
<<aq>>=
fnobs(airquality)
@
This imputes \code{Ozone} and \code{Solar.R} by reference using the month median.
<<aq_imp>>=
fmedian(airquality[1:2], airquality$Month, TRA = "replace_na", set = TRUE)
@
This performs different grouped and/or weighted transformations at once.
<<aq_trans>>=
airquality |> fmutate(
  rad_day = fsum(as.double(Solar.R), Day, TRA = "/"),
  ozone_deg = Ozone / Temp,
  ozone_amed = Ozone > fmedian(Ozone, Month, TRA = "fill"),
  ozone_resid = fmean(Ozone, list(Month, ozone_amed), ozone_deg, "-")
) |> head(3)
@
%
\subsection{Grouping objects and optimization}
%
Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, the cost of grouping can be optimized by using factors or, even better, \class{GRP} objects, which contain all information for \pkg{collapse}'s vectorized functions to operate across groups efficiently. These objects can be created with \code{GRP()}. The syntax is
 \begin{Code}
 GRP(X, by = NULL, sort = TRUE, decreasing = FALSE, na.last = TRUE,
     return.groups = TRUE, return.order = sort, method = "auto", ...)
 \end{Code}
 The example below creates and displays a \class{GRP} object from 3 columns in \code{mtcars}. The \code{by} argument also supports column names or indices, and \code{X} could also be an atomic vector.
%
<<GRP>>=
str(g <- GRP(mtcars, ~ cyl + vs + am))
@
%
\class{GRP} objects make grouped statistical computations in \pkg{collapse} fully programmable. Below, the object is used with the \emph{Fast Statistical Functions} and some utility functions to aggregate the data (with optional weights) efficiently.
%
<<gcomp>>=
dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt
add_vars(g$groups,
  fmean(dat, g, w, use.g.names = FALSE) |> add_stub("w_mean_"),
  fsd(dat, g, w, use.g.names = FALSE) |> add_stub("w_sd_")) |> head(2)
@
%
Similar, data can be transformed, here also using the S3 generic \fct{fscale} function.
%
<<gtrans>>=
mtcars |> add_vars(fmean(dat, g, w, "-") |> add_stub("w_demean_"),
                   fscale(dat, g, w) |> add_stub("w_scale_")) |> head(2)
@
%
This programming access can become very useful. For example, the useR 2022 presentation \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{Slides 18-19} aggregates the \href{https://worldmrio.com/}{EORA Global Supply Chain Database} from the country to the world region level. After defining a single grouping object, a list of value-added shares matrices (VB) and outputs for years 1990-2021, is aggregated with no grouping cost using a single line of code like \code{lapply(VB_list, function(x) x$VB |> fsum(g) |> t() |> fmean(g, x$O) |> t())}. On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another example from my recent work involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}. Thus each iteration of the numerical routine raises the vector \textbf{x} to a different power ($a$), sums it in 1 million groups ($j$) to generate \textbf{y}, and computes the Euclidean distance to \textbf{z} (using \code{collapse::fdist}). Without grouping objects and vectorization, this would have been difficult to handle within reasonable computing times (a few seconds on the M1).}
%
\section{Integration with data manipulation functions} \label{sec:integration}
%
\pkg{collapse} also provides a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{fast data manipulation functions} familiar to \proglang{R} and \pkg{tidyverse} users, including \fct{fselect}, \fct{fsubset}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{ftransform}, \fct{fmutate}, \fct{across}, \fct{frename}, \fct{fcount}, etc. These are integrated with the \emph{Fast Statistical Functions} to enable vectorized statistical operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow. For example, the following code calculates
%
<<fsummarise>>=
mtcars |>
  fsubset(mpg > 11) |>
  fgroup_by(cyl, vs, am) |>
  fsummarise(across(c(mpg, carb, hp), fmean),
             qsec_w_med = fmean(qsec, wt)) |> head(2)
@
%
the mean of columns \code{mpg}, \code{carb} and \code{hp}, and the weighted mean of \code{qsec}, after subsetting and grouping the data. This code is very fast (especially with many groups) because data does not need to be split by groups at all. There is also no need to call \code{lapply()} inside the \code{across()} statement: \code{fmean.data.frame()} is applied to a subset of the data containing the three columns.\footnote{Internally, the \code{g} argument of the statistical functions is set as a keyword argument by \code{fsummarise/across} and the function is evaluated on a suitable subset of columns. Thus \code{w} becomes the second positional argument...} The \emph{Fast Statistical Functions} also have a method for grouped data, so \code{fsummarise} is not always needed. The following code calculates weighted group means. By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.
%
<<fsummarise_2>>=
mtcars |>
  fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
  fgroup_by(cyl, vs, am) |>
  fmean(wt) |> head(2)
@
%
\subsection{Vectorizations for advanced tasks}
%
\fct{fsummarise} and \fct{fmutate} can also evaluate other statistical functions in the classical way (split-apply-combine) and handle more complex expressions involving multiple columns and/or functions. However, using any \emph{Fast Statistical Function} causes the whole expression to be vectorized, i.e., evaluated only once and not for every group. This eager vectorization approach is helpful for efficient grouped calculation of more complex statistics. The example below calculates grouped (\code{vs}) bivariate regression slopes (\code{mpg ~ carb}) in a vectorized way.
%
<<flinreg>>=
mtcars |>
 fgroup_by(vs) |>
 fmutate(dm_carb = fmean(carb, TRA = "-")) |>
 fsummarise(slope = fsum(mpg, dm_carb) %/=% fsum(dm_carb^2))
@
%
Apart from vectorization, this code avoids 3 intermediate copies: (1) \code{fmean(carb, TRA = "-")} avoids an expanded vector of group means, (2) \code{fsum(mpg, dm\_carb)} uses the weights (\code{w}) argument to \fct{fsum} to avoid materializing a multiplication (as in \code{fsum(mpg * dm\_carb)}), and (3) division by reference (\code{\%/=\%}) avoids allocating an additional vector for the final result. Under the hood, the expression boils down to an (expensive) grouping step, 5 allocations (of which 2 full length), and 6 loops in C to calculate the result. Any modern laptop can calculate 1 million regression slopes in less than 1 second like this. \newline

\pkg{collapse} also supplies advanced statistics, such as weighted medians and modes. The following example calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}\footnote{\pkg{collapse} calculates weighted quantiles by replacing the sample size with the sum of weights and 1 with the minimum non-zero weight in the respective quantile definition. See \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for more details.}, and a weighted maximum mode.\footnote{The weighted maximum mode is the largest element with the maximum sum of weights.}
%
<<wstats>>=
mtcars |>
    fgroup_by(cyl, vs, am) |>
    fmutate(o = radixorder(GRPid(), mpg)) |>
    fsummarise(mpg_min = fmin(mpg),
               mpg_Q1 = fnth(mpg, 0.25, wt, o = o, ties = "q8"),
               mpg_mean = fmean(mpg, wt),
               mpg_median = fmedian(mpg, wt, o = o, ties = "q8"),
               mpg_mode = fmode(mpg, wt, ties = "max"),
               mpg_Q3 = fnth(mpg, 0.75, wt, o = o, ties = "q8"),
               mpg_max = fmax(mpg)) |> head(3)
@
%
Both weighted mode and quantiles have a sub-column parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \code{fnth/fmedian/fmode} (default 1).} and, as in this case, can also harness an (optional) optimization by computing an overall ordering vector and passing it to the functions to avoids repeated sorting (using quickselect) of the same elements within each group. For advanced data aggregation, \pkg{collapse} also provides a convenience function, \fct{collap}, which (by default) uses \code{fmean} for numeric and \code{fmode} for non-numeric columns. Below, it aggregates GDP per Capita, Life Expectancy, and country name by World Bank Income Group, with population weights.\footnote{\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}} is a dataset supplied by \pkg{collapse}, extracted from the World Bank World Development Indicators.} This yields population-weighted statistics, the largest country, and each income group's total population (sum of weights) for each year, preserving (default \code{keep.col.order = TRUE}) the order of columns.
%
<<collap>>=
collap(wlddev, country + PCGDP + LIFEEX ~ year + income, w = ~ POP) |>
  head(4)
@
%
\section{Time series and panel data} \label{sec:ts_ps}
%
\pkg{collapse} provides a flexible high-performance architecture to perform (\emph{time aware}) computations on time series and panel series. In particular, the user enjoys great flexibility in deciding the desired degree of indexation and mode of computation. It is possible to apply time series and panel data transformations without any indexation by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation. The following list summarizes \pkg{collapse}'s time series and panel data functions.
%
\begin{table}[h]
\begin{tabular}{p{\textwidth}}
\emph{Classes, Constructors and Utilities} \\
\code{findex\_by(), findex(), unindex(), reindex(), timeid(), is\_irregular(), to\_plm()} $+$ rich set of S3 methods for 'indexed\_frame', 'indexed\_series' and 'index\_df'. \\\\
\emph{Core Time-Based Functions} \\
\code{flag(), fdiff(), fgrowth(), fcumsum(), psmat()} [panel data to array conversions] \code{psacf(), pspacf(), psccf()} [autocorrelation functions for panel data] \\\\
\emph{Data Transformation Functions with Supporting Methods} \\
\code{f[hd]between(), f[hd]within(), fscale()} [scaling and (higher-dimensional) centering] \\\\
\emph{Data Manipulation Functions with Supporting Methods} \\
\code{fsubset(), funique(), roworder[v]()} [internal], and \code{na\_omit()} [internal] \\\\
\emph{Summary Functions with Supporting Methods} \\
\code{qsu(), varying()} [panel-variance decomposed statistics] \\
\end{tabular}
\end{table}
%
\subsection{Ad-hoc computations}
%
Time series functions such as \fct{fgrowth} (to compute growth rates) are S3 generic and can be applied to most time series classes. In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation. If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years.}. If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id. For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}. % The following provides a basic example:
%
<<ts_example>>=
fgrowth(airmiles) |> round(2)
@
The following code creates an irregular series by removing the 3rd and 15th observation and shows how indexation with the \code{t} argument accounts for this.
%
<<ts_example_ctd>>=
am_ir <- airmiles[-c(3, 15)]
t <- time(airmiles)[-c(3, 15)]
fgrowth(am_ir, t = t) |> round(2)
fgrowth(am_ir, -1:3, t = t) |> head(4)
@
%
For these functions, there also exists shorthands in the form of statistical operators, e.g., \code{L()/D()/G()} are shorthands for \code{flag()/fdiff()/fgrowth()}, which facilitate their use inside formulas and also provide enhanced data frame interfaces for convenient ad-hoc computations. With panel data, \code{t} can be omitted, but this requires sorted data with consecutive groups.\footnote{This is because a group-lag is computed in a single pass, requiring all group elements to be consecutive.}
%
<<example_growth>>=
G(wlddev, c(1, 10), by = POP + LIFEEX ~ iso3c, t = ~ year) |> head(3)
settransform(wlddev, POP_growth = G(POP, g = iso3c, t = year))
@
%
These functions and operators are also integrated with \fct{fsummarise} and \fct{fmutate} for vectorized grouped computations.
%
<<example_growth_continued>>=
wlddev |> fgroup_by(iso3c) |> fselect(iso3c, year, POP, LIFEEX) |>
  fmutate(across(c(POP, LIFEEX), G, t = year)) |> head(2)
@
%
Similarly, functions to scale, center, and average data have groups (\code{g}) and also weights (\code{w}) arguments, and corresponding operators \code{STD(),[HD]W(),[HD]B()} to facilitate ad-hoc transformations. Below, two ways to perform grouped scaling are demonstrated. The operator version is slightly faster and renames the transformed columns by default (\code{stub = TRUE}).
%
<<example_scale_cener>>=
iris |> fgroup_by(Species) |> fscale() |> head(2)
STD(iris, ~ Species) |> head(2)
@
The following example demonstrates a fixed-effects regression a la \citet{mundlak1978pooling}.
<<example_reg>>=
lm(mpg ~ carb + B(carb, cyl), data = mtcars) |> coef()
@
\pkg{collapse} also offers higher-dimensional between and within transformations, powered by C++ code conditionally imported (and accessed directly) from \pkg{fixest}. The following detrends GDP per Capita and Life Expectancy at Birth using country-specific cubic polynomials.
<<example_HD>>=
HDW(wlddev, PCGDP + LIFEEX ~ iso3c * poly(year, 3), stub = F) |> head(2)
@
%
\subsection{Indexed series and frames}
%
For more complex use cases, indexation is very convenient. \pkg{collapse} supports \pkg{plm}'s \class{pseries} and \class{pdata.frame} classes through dedicated methods. Flexibility and performance considerations lead
to the creation of new classes \class{indexes\_series} and \class{indexed\_frame} which inherit from the former. Any data frame-like object can be an \class{indexed\_frame} with any number of individual and/or time identifiers (e.g. an indexed \class{data.table} is fully functional for other operations). The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}. The basic syntax is:
%
\begin{Code}
data_ix = findex_by(data, id1, ..., time)
data_ix$indexed_series
with(data, indexed_series)
index_df = findex(data_ix)
\end{Code}
%
Data can be indexed using one or more indexing variables. Unlike \class{pdata.frame}, an \\ \class{indexed\_frame} is a deeply indexed structure, i.e., every series inside the frame is already an \class{indexes\_series} and contains, in its \class{index\_df} attribute, an external pointer to the \class{index\_df} attribute of the frame (to avoid duplication in memory). A comprehensive set of \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \pkg{flag/L}, ensure that these objects behave in a time-/panel-aware manor in any caller environment (created by \code{with}, \code{lm} etc.). % A basic demonstration with World Bank panel data showcases the flexibility of these classes.
Indexation can be undone using \code{unindex()} and redone with \code{reindex()} and a suitable \class{index\_df}. \class{indexes\_series} can be atomic vectors or matrices (including objects such as \class{ts} or \class{xts}) and can also be created directly using \code{reindex()}.
%
\begin{Code}
data = unindex(data_ix)
data_ix = reindex(data, index = index_df)
indexed_series = reindex(vec/mat, index = vec/index_df)
\end{Code}
%
An example using the \code{wlddev} data follows:
<<indexing>>=
wldi <- wlddev |> findex_by(iso3c, year)
wldi |> fsubset(-3, iso3c, year, PCGDP:POP) |> G() |> head(4)
@
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. This creates an \class{indexes\_series} of life expectancy and demonstrates its properties.
<<indexing_2>>=
LIFEEXi = wldi$LIFEEX
str(LIFEEXi, width = 70, strict = "cut")
c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5]))
G(LIFEEXi[c(1:5, 7:10)])
@
The transformation and estimation below demonstrate the deep indexation of \\ \class{indexed\_frame}'s, allowing correct computations in arbitrary data masking environments.
<<indexing_3>>=
settransform(wldi, PCGDP_ld = Dlog(PCGDP))
lm(D(LIFEEX) ~ L(PCGDP_ld, 0:5) + B(PCGDP_ld), wldi) |>
  summary() |> coef() |> round(3)
@
The above example could also have been executed in one line as \code{lm(D(LIFEEX) ~ } \\ \code{L(Dlog(PCGDP), 0:5) + B(Dlog(PCGDP)), wldi)}, log-differencing \code{PCGDP} twice. \newline

In comparison with existing solutions, the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.}. Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures for time series operations or native handling of irregularity in basic operations. \class{indexed\_series} and \class{indexed\_frame}, on the other hand, work anywhere and can be superimposed on any suitable object (such as \class{sf} to create a spatiotemporal panel), as long as \pkg{collapse}'s functions (\code{flag()/L()} etc.) are used to perform the time-based computations. The \class{index\_df} attached to these objects can be used with other general tools such as \code{collapse::BY()} to perform grouped computations on these objects with 3rd-party functions. An example of calculating a 5-year rolling average is given below. Last but not least, the performance of these classes is second to none, as demonstrated in the useR 2022 presentation \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{on slide 40}.
%
<<rollmean>>=
BY(LIFEEXi, findex(LIFEEXi)$iso3c, data.table::frollmean, 5) |> head(10)
@
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
While \pkg{collapse} has a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{Data Manipulation Functions}, its implementations of table joins and pivots is particularly noteworthy since they offer several new features such as various verbose options for table joins, pivots supporting variable labels, and 'recast' pivots.
%
\subsection{Joins}
%
Compared to commercial environments such as \proglang{STATA}, the implementation of joins in most open-source software, including \proglang{R}, is non-verbose, i.e., provides no information on how many and which records were joined from both tables. This is somewhat unsatisfying and often provokes manual efforts to validate the join operation. \code{collapse::join} provides a rich set of options to make table join operations intelligible. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
     multiple = FALSE, sort = FALSE, keep.col.order = TRUE,
     drop.dup.cols = FALSE, verbose = 1, column = NULL, attr = NULL, ...)
\end{Code}
By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. Users can request the generation of a '.join' column, akin to \proglang{STATA}'s '\_merge' column, indicating the origin of records in the joined table, as shown below.
%
<<joins>>=
df1 <- data.frame(id1 = c(1, 1, 2, 3),
                  id2 = c("a", "b", "b", "c"),
                  name = c("John", "Jane", "Bob", "Carl"),
                  age = c(35, 28, 42, 50))
df2 <- data.frame(id1 = c(1, 2, 3, 3),
                  id2 = c("a", "b", "c", "e"),
                  salary = c(60000, 55000, 70000, 80000),
                  dept = c("IT", "Marketing", "Sales", "IT"))

join(df1, df2, on = c("id1", "id2"), how = "full", column = TRUE)
@
%
An alternative to the join column is to request an attribute that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}). Users can also invoke the \code{validate} argument to check the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.
%
<<joins_2>>=
join(df1, df2, on = c("id1", "id2"), validate = "1:1", attr = "join") |>
  attr("join") |> str(width = 70, strict = "cut")
@
%
A few further particularities are worth highlighting. First, \code{collapse::join} is also class-agnostic and preserves the attributes of \code{x}. It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"} or \code{"anti"}) and default to \code{"left"}, so the default behavior simply adds columns to \code{x}. By default (\code{sort = FALSE}), the order of rows in \code{x} is also preserved. Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join (using \fct{fmatch}) followed by sorting, particularly if the data is already sorted on the keys. } Additionally, by default (\code{multiple = FALSE}), only the first matches from both tables are joined to avoid silent cartesian duplication of records. In multi-match settings, this will be reflected by few records from \code{y} being used. \fct{fmatch} also has a built-in overidentification check, which issues a warning if more key columns than necessary to identify the records are provided.
%
<<join_3>>=
df2$name = df1$name
join(df1, df2) |> capture.output(type="m") |> strwrap(77) |> cat(sep="\n")
@
%
% This warning can be silenced by passing \code{overid = 2} or restricting the join columns using \code{on}.
A final noteworthy feature is the handling of duplicate non-id columns in both tables.
%
<<join_4>>=
join(df1, df2, on = c("id1", "id2"))
@
%
By default (\code{suffix = NULL}), \fct{join} extracts the name of the \code{y} table and appends \code{y}-columns with it. \code{x}-columns are not renamed. This is congruent to the principle of adding columns to \code{x} and altering this table as little as possible. Another option, \code{drop.dup.cols = "x"/"y"} can be used to simply drop duplicate columns from \code{x} or \code{y} before the join operation.
%
\subsection{Pivots}
%
The reshaping/pivoting functionality of commercial and open source software is also unsatisfying for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data. Such datasets can only be reshaped by losing these labels or additional manual efforts to retain them. Modern \proglang{R} packages also offer different functions for different reshaping operations, such as \code{data.table::melt/ tidyr::pivot_longer} to combine columns and \code{data.table::dcast/tidyr::pivot_wider} to expand them, requiring users to learn both. Finally, since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is no modern replacement for \code{reshape2::recast()}. This requires \proglang{R} users to consecutively call two reshaping functions, incurring a high cost in terms of syntax and memory for such frequently required data transpositions. \newline

\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, can handle complex labelled data without loss of information, and offers best-in-\proglang{R} computational performance and memory efficiency. Its syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
      how = "longer", na.rm = FALSE, factor = c("names", "labels"),
      check.dups = FALSE, nthreads = 1, fill = NULL, drop = TRUE,
      sort = FALSE, transpose = FALSE)
\end{Code}
%
Below, its use is demonstrated with a generated dataset about fruits. We could also think about a survey with households and individuals, or sectors and firms. Variable labels are stored in \code{attr(column, "label")}. The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples with real data.
%
<<pivots>>=
data <- data.frame(type = rep(c("A", "B"), each = 2),
            type_name = rep(c("Apples", "Bananas"), each = 2),
            id = rep(1:2, 2), r = abs(rnorm(4)), h = abs(rnorm(4)*2))
setrelabel(data, id = "Fruit Id", r = "Fruit Radius", h = "Fruit Height")
print(data)
vlabels(data)
@
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = c(...))}. If \code{labels = "lab_name"} is specified, variable labels are saved to an additional column named 'lab\_name'. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to assign alternative names to the 'variable' and 'value' columns, respectively.
%
<<pivot_longer>>=
(dl <- pivot(data, ids = c("type", "type_name", "id"), labels = "label"))
vlabels(dl)
@
%
In general, \fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by \code{pivot(data, values = c("r", "h"), labels = "label")}. An exact reverse operation can also be performed by specifying as little as \code{pivot(dl, labels = "label", how = "w")}. \newline

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} \emph{Note} how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation, i.e., with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
%
<<pivot_wider>>=
(dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
namlab(dw)
@
%
For the recast pivot (\code{how = "recast"}), unless a column named 'variable' exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is optional, and omitting either the list's 'from' or 'to' elements will omit the respective operations.
%
<<pivot_recast>>=
(dr <- pivot(data, ids = "id", names = list(from = "type"),
             labels = list(from = "type_name", to = "label"), how = "r"))
vlabels(dr)
@
%
As with the other pivots, this operation does not incur any loss of information. A suitable reverse operation also exists: \code{pivot(dr, "id", names = list(to = "type"), labels = list(from = "label", to = "type_name"), how = "r")}. More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it is also possible to perform longer and recast pivots without id variables. The recast pivot without IDs resembles a generalization of \code{data.table::transpose}, albeit slightly less efficient.
%
\section{List processing} \label{sec:list_proc}
%
Often, in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model object, performance statistics, and predictions) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices, e.g., in response to nested user inputs in shiny apps. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{List Processing Functions} attempt to provide a basic toolkit. \newline

To create nested data, \fct{rsplit} generalizes \fct{split} and (recursively) splits up data frame-like objects into a (nested) list.
%
<<list_proc>>=
(dl <- mtcars |> rsplit(mpg + hp + carb ~ vs + am)) |> str(max.level = 2)
@
%
If a nested structure is not wanted, argument \code{flatten = TRUE} lets \fct{rsplit} operate like a faster version of \fct{split}. With a single column on the LHS of the formula, the default (\code{simplify = TRUE}) returns a nested list of atomic vectors. Having created a nested list, \fct{rapply2d} is used to fit a linear model on each frame,\footnote{\fct{rapply2d} is just a recursive wrapper around \fct{lapply}, with different defaults than \fct{rapply}. Notably, by default, it excludes data frames from being considered as sub-lists and does not simplify the result.} followed by \fct{get\_elem} to obtain the coefficient matrices. \fct{get\_elem} offers several options for filtering lists but, by default, simplifies the list tree as much as possible while maintaining existing hierarchies. In this case, it returns the same nested list with coefficient matrices in all final nodes.
%
<<list_proc_2>>=
nest_lm_coef <- dl |>
  rapply2d(lm, formula = mpg ~ .) |>
  rapply2d(summary, classes = "lm") |>
  get_elem("coefficients")

nest_lm_coef |> str(give.attr = FALSE, strict = "cut")
@
%
At last, \fct{unlist2d} is applied to unlist the nested list to a data frame. This function can create a data frame (or \class{data.table}) representation of any nested list containing data using recursive row-binding and coercion operations while generating (optional) id variables representing the list tree and (optionally) saving row names of matrices or data frames.
%
<<list_proc_3>>=
nest_lm_coef |> unlist2d(c("vs", "am"), row.names = "variable") |> head(2)
@
%
This example does not represent an optimal workflow for this specific task\footnote{A better way of achieving the same result would be \code{mtcars |> fgroup\_by(vs, am) |> fsummarise(qDF(lmtest::coeftest(lm(mpg ~ hp + carb)), "variable"))}.} but exemplifies the power of these tools to create, query, and combine nested data in very general ways.
  % I use it extensively to fetch and compare different statistics from (sometimes large) nested lists with different statistical model outputs.
\pkg{collapse}'s list processing toolkit provides further useful functions such as \fct{t\_list} to turn lists of lists inside out, \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, and \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes. A non-recursive and class-agnostic \fct{rowbind} function also exists to efficiently bind lists of data frame-like objects (like \code{data.table::rbindlist}).
%
\section{Summary statistics} \label{sec:summ_stat}
%
\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{Summary Statistics Functions} provide a parsimonious and powerful toolset to examine complex datasets. A particular focus has been on providing tools for examining longitudinal (panel) data. Recall the indexed world development panel (\code{wldi}) from Section (\ref{sec:ts_ps}). The function \fct{varying} indicates which of these variables are time-varying:
%
<<prep, include=FALSE, echo=FALSE>>=
wldi <- wldi[1:13]
options(width = 77)
@
<<varying_1>>=
varying(wldi)
varying(wldi, any_group = FALSE) |> head(3)
@
%
Country-variance can be examined using \code{varying(wldi, effect = "year")}. For non-indexed data, \fct{varying} also has a \code{g} argument. A related exercise is to decompose the variance of a panel series into a component due to variation between countries and one capturing variance within countries over time. Using the \code{W()/B()} operators and the \code{LIFEEXi} \class{indexed\_series} from Section (\ref{sec:ts_ps}), this is easily demonstrated
%
<<pdec>>=
all.equal(fvar(W(LIFEEXi)) + fvar(B(LIFEEXi)), fvar(LIFEEXi))
@
%
The function \fct{qsu} (quick-summary) provides an efficient method to approximately compute this decomposition, considering the group-means instead of the between transformation\footnote{This is more efficient and equal to using the between transformation if the panel is balanced.} and adding the mean back to the within transformation to preserve the scale of the data.
%
<<qsu_1>>=
qsu(LIFEEXi)
@
%
This decomposition shows more variation in life expectancy between countries than within countries over time. It can also be computed for different subgroups, such as OECD members and non-members, and with sampling weights, such as population. \fct{qsu} can also return Pearson's measures of higher-order statistics.
%
<<qsu_2>>=
qsu(LIFEEXi, g = wlddev$OECD, w = wlddev$POP, higher = TRUE) |> aperm()
@
%
The output shows that the variation in life expectancy is significantly larger for non-OECD countries and that for these countries, the between- and within-country variation is approximately equal in magnitude.\footnote{\fct{qsu} also has a convenient formula interface to perform these transformations in an ad-hoc fashion, e.g., the above can be obtained using \code{qsu(wlddev, LIFEEX ~ OECD, ~ iso3c, ~POP, higher = TRUE)}, without prior indexation.} For more detailed (grouped, weighted) statistics, \fct{descr} provides a rich statistical description of variables in a dataset.
%
<<descr>>=
descr(wlddev, LIFEEX ~ OECD, w = ~ replace_na(POP))
@
<<set_width, echo=FALSE>>=
options(width = 80)
@

%
While \fct{descr} does not support panel-variance decompositions like \fct{qsu}, it also provides detailed (grouped, weighted) frequency tables for categorical data and is thus very useful for complex survey data. A \code{stepwise} argument toggles describing one variable at a time, allowing users to naturally 'click-through' a large dataset rather than printing a massive output to the console. More details and examples are in the \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation}. Both \fct{qsu} and \fct{descr} provide an \code{as.data.frame()} method for efficient tidying and further analysis. \newline

A final noteworthy function from \pkg{collapse}'s descriptive statistics toolkit is \fct{qtab}, an enhanced drop-in replacement for \code{base::table}. It is enhanced both in a statistical and computational sense, providing a remarkable performance boost, an option (\code{sort = FALSE}) to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights - vectorized when using \emph{Fast Statistical Functions}, as demonstrated below.
%
<<qtab>>=
library(magrittr) # World after 2015 (latest country data)
wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
wlda15 %$% qtab(OECD, income)
@
This shows the total population (latest post-2015 estimates) in millions.
<<qtab_2>>=
wlda15 %$% qtab(OECD, income, w = POP) %>% divide_by(1e6)
@
This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e. \fct{fmean} is only invoked once).
<<qtab_3>>=
wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean) %>% replace_na(0)
@
Finally, this calculates a population-weighted average of life expectancy in each group.
<<qtab_4>>=
wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean,
                wFUN.args = list(w = POP)) %>% replace_na(0)
@
%
\class{qtab} objects inherit the \class{table} class, thus all \class{table} methods apply. Apart from the above functions, \pkg{collapse} also provides functions \pkg{pwcor}, \pkg{pwcov}, \pkg{pwnobs} for convenient (pairwise, weighted) correlations, covariances, and observations counts, and also functions \pkg{psacf}, \pkg{pspacf} and \pkg{psccf} for auto- and cross-covariance and correlation function estimation on panel series.
%
\section{Global options} \label{sec:glob_opt}
%
\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function. These options are saved in an internal environment called \code{.op} (for safety and performance reasons) visible in the \href{https://sebkrantz.github.io/collapse/reference/fmean.html}{documentation of some functions}. Its contents can be accessed using \fct{get\_collapse}. \newline

The current set of options comprises the default behavior for missing values (\code{na.rm} arguments in all statistical functions and algorithms), sorted grouping (\code{sort}), multithreading and algorithmic optimizations (\code{nthreads}, \code{stable.algo}), presentational settings (\code{stub}, \code{digits}, \code{verbose}), and, surpassing all else, the package namespace itself (\code{mask}, \code{remove}). \newline

As evident from previous sections, \pkg{collapse} provides performance-improved or otherwise enhanced versions of functionality already present in base R (like the \emph{Fast Statistical Functions}, \fct{funique}, \fct{fmatch}, \fct{fsubset}, \fct{ftransform}, etc.) and other packages (esp. \pkg{dplyr} \citep{rdplyr}: \fct{fselect}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, etc.). The objective of being namespace compatible warrants such a naming convention, but this has a syntactical cost, particularly when \pkg{collapse} is used as the primary data manipulation framework. \newline

To reduce this cost, \pkg{collapse}'s \code{mask} option allows masking existing \proglang{R} functions with the faster \pkg{collapse} versions by creating additional functions in the namespace and instantly exporting them. All \pkg{collapse} functions starting with 'f' can be passed to the option (with or without the 'f'), e.g., \code{set\_collapse(mask = c("subset", "transform"))} creates \code{subset <- fsubset} and \code{transform <- ftransform} and exports them. Special functions are \code{"n"} and \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} (for use in \code{(f)summarise}/\code{(f)mutate}), \code{table <- qtab}, and replace \code{\%in\%} with a fast version using \code{fmatch}, respectively. There are also a couple of \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask groups of functions}. The most powerful of these is \code{"all"}, which masks all f-functions $+$ specials, as shown below.
%
\begin{Code}
set_collapse(mask = "all", na.rm = FALSE, sort = FALSE, nthreads = 4)
wlddev |>
  subset(year >= 1990 & is.finite(GINI)) |>
  group_by(year) |>
  summarise(n = n(), across(PCGDP:GINI, mean, w = POP))
with(mtcars, table(cyl, vs, am))
sum(mtcars)
diff(EuStockMarkets)
mean(num_vars(iris), g = iris$Species)
unique(wlddev, cols = c("iso3c", "year"))
range(wlddev$date)
wlddev |>
  index_by(iso3c, year) |>
  mutate(PCGDP_lag = lag(PCGDP),
         PCGDP_diff = PCGDP - PCGDP_lag,
         PCGDP_growth = growth(PCGDP)) |> unindex()
\end{Code}
%
The above is now 100\% \pkg{collapse} code. Similarly, using this option, all code in this article could have been written without f-prefixes. Thus, \pkg{collapse}, together with namespace masking, provides a fast experience of \proglang{R} - within GNU \proglang{R} - without the need to even restart the session. Masking is completely interactive and reversible within the active session: calling \code{set\_collapse(mask = NULL)} instantly removes the additional functions. Option \code{remove} can further be used to remove any \pkg{collapse} function from the list of exported functions, allowing manual conflict management. Function \code{fastverse::fastverse\_conflicts()} from the related \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse} project}\footnote{Website: https://fastverse.github.io/fastverse/} can be used to display namespace conflicts with \pkg{collapse}. Invoking either \code{mask} or \code{remove} detaches \pkg{collapse} and reattaches it at the top of the search path, letting its namespace to take precedence over other packages.
%
\section{Benchmark} \label{sec:bench}
%
This section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks. They are executed on an Apple M1 MacBook Pro with 16 GB unified memory. All packages are compiled and used in single-threaded mode to avoid artifacts of different parallel setups and resource utilization. The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets.
%
<<bench, cache=TRUE>>=
set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 1)
set.seed(101)
m <- matrix(rnorm(1e7), ncol = 1000)
data <- qDT(replicate(100, rnorm(1e5), simplify = FALSE))
g <- sample.int(1e4, 1e5, TRUE)

microbenchmark(R = colMeans(m),
               Rfast = Rfast::colmeans(m),
               collapse = fmean(m))
microbenchmark(R = rowsum(data, g, reorder = FALSE),
               data.table = data[, lapply(.SD, sum), by = g],
               collapse = fsum(data, g))
add_vars(data) <- g
microbenchmark(data.table = data[, lapply(.SD, median), by = g],
               collapse = data |> fgroup_by(g) |> fmedian())
d <- data.table(g = unique(g), x = 1, y = 2, z = 3)
microbenchmark(data.table = d[data, on = "g"],
               collapse = join(data, d, on = "g", verbose = 0))
microbenchmark(data.table = melt(data, "g"),
               collapse = pivot(data, "g"))
settransform(data, id = rowid(g))
cols = grep("^V", names(data), value = TRUE)
microbenchmark(data.table = dcast(data, g ~ id, value.var = cols),
          collapse = pivot(data, ids = "g", names = "id", how = "w"))
@
%
The benchmark below further shows that \pkg{collapse} provides faster algorithms for basic computationally intensive operations such as unique values and matching. These algorithms power much of its functionality, such as efficient factor generation with \fct{qF}, cross-tabulation with \fct{qtab}, \fct{join}'s, \fct{pivot}'s, etc.
%
<<bench_2, cache=TRUE>>=
set.seed(101)
g_int <- sample.int(1e3, 1e7, replace = TRUE)
char <- c(letters, LETTERS, month.abb, month.name)
char <- outer(char, char, paste0)
g_char <- sample(char, 1e7, replace = TRUE)
microbenchmark(base_int = unique(g_int), collapse_int = funique(g_int),
            base_char = unique(g_char), collapse_char = funique(g_char))
microbenchmark(base_int = match(g_int, 1:1000),
               collapse_int = fmatch(g_int, 1:1000),
               base_char = match(g_char, char),
               data.table_char = chmatch(g_char, char),
               collapse_char = fmatch(g_char, char), times = 10)
@
%
Apart from the raw algorithmic efficiency demonstrated here, \pkg{collapse} is often more efficient than other solutions by simply doing less. For example, if grouping columns are factor variables, \pkg{collapse}'s algorithms in \code{funique()}, \code{group()} or \code{fmatch()}, etc., use the values as hashes without checking for collisions. Similarly, if data is already sorted/unique, it is directly returned by functions like \code{roworder()}/\code{funique()}.

\newpage
%% -- Summary/conclusions/discussion -------------------------------------------

\section{Conclusion} \label{sec:conclusion}

It is coming close to 4 years since the first CRAN release of \pkg{collapse} in March 2020, and since then, the package has grown and matured considerably. At the time of writing this article, it has been downloaded $>$1.5 million times off CRAN and is known and used by thousands of \proglang{R} users. In this article, I have articulated key ideas and design principles and demonstrated some core features of the package. In summary, my work with \proglang{R} as an applied economist has led me to believe that there should be a new foundation package for statistical computing and data manipulation in \proglang{R} that is statistically advanced, class-agnostic, flexible, fast, lightweight, stable, and able to manipulate complex scientific data with ease. \pkg{collapse} is my attempt at providing such a package, and the feedback I have received over the years, particularly from users in academia, government, and international organizations, is a strong indication that I have responded to a need felt in larger parts of the \proglang{R} community. As mentioned, a single article cannot summarize \pkg{collapse} in 2024, but there is an excellent \href{https://sebkrantz.github.io/collapse/index.html}{website} with comprehensive \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{Documentation Resources}.

%% -- Optional special unnumbered sections -------------------------------------

\section*{Computational details}
The results in this paper were obtained using \proglang{R} \citep{R} 4.3.0 with \pkg{collapse} 2.0.8, \pkg{data.table} 1.14.10, \pkg{Rfast} 2.1.0, \pkg{fixest} 0.11.2, \pkg{magrittr} \citep{rmagrittr} 2.0.3 and \pkg{microbenchmark} \citep{rmicrobenchmark} 1.4.10. All packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at https://CRAN.R-project.org/. The benchmark was run on an Apple M1 MacBook Pro (2020) with 16GB unified memory and serially compiled CRAN binaries for Mac.

\section*{Acknowledgments}

The source code of \pkg{collapse} has been heavily inspired by (and partly copied from) \pkg{data.table} (Matt Dowle and Arun Srinivasan), \proglang{R}'s source code (R Core Team and contributors worldwide), the \pkg{kit} package (Morgan Jacob), and \pkg{Rcpp} (Dirk Eddelbuettel). Packages \pkg{plm} (Yves Croissant, Giovanni Millo, and Kevin Tappe) and \pkg{fixest} (Laurent Berge) have also provided a lot of inspiration (and a port to its demeaning algorithm in the case of \pkg{fixest}). I also thank many people from diverse fields for helpful answers on Stackoverflow and many other people for encouragement, feature requests, and helpful issues and suggestions.
% \begin{leftbar}
% All acknowledgments (note the AE spelling) should be collected in this
% unnumbered section before the references. It may contain the usual information
% about funding and feedback from colleagues/reviewers/etc. Furthermore,
% information such as relative contributions of the authors may be added here
% (if any).
% \end{leftbar}

%% -- Bibliography -------------------------------------------------------------
%% - References need to be provided in a .bib BibTeX database.
%% - All references should be made with \cite, \citet, \citep, \citealp etc.
%%   (and never hard-coded). See the FAQ for details.
%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
%% - Titles in the .bib should be in title case.
%% - DOIs should be included where available.

\newpage

\bibliography{refs}

%% -- Appendix (if any) --------------------------------------------------------
%% - After the bibliography with page break.
%% - With proper section titles and _not_ just "Appendix".

\newpage
%
% \begin{appendix}
%
% \section{More technical details} \label{app:technical}
%
% \begin{leftbar}
% Appendices can be included after the bibliography (with a page break). Each
% section within the appendix should have a proper section title (rather than
% just \emph{Appendix}).
%
% For more technical style details, please check out JSS's style FAQ at
% \url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
% which includes the following topics:
% \begin{itemize}
%   \item Title vs.\ sentence case.
%   \item Graphics formatting.
%   \item Naming conventions.
%   \item Turning JSS manuscripts into \proglang{R} package vignettes.
%   \item Trouble shooting.
%   \item Many other potentially helpful details\dots
% \end{itemize}
% \end{leftbar}
%
%
% \section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
%
% \begin{leftbar}
% References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All
% references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|,
% \verb|\citealp| etc.\ (and never hard-coded). This commands yield different
% formats of author-year citations and allow to include additional details (e.g.,
% pages, chapters, \dots) in brackets. In case you are not familiar with these
% commands see the JSS style FAQ for details.
%
% Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially
% when acquiring the entries automatically from mixed online sources. However,
% it is important that informations are complete and presented in a consistent
% style to avoid confusions. JSS requires the following format.
% \begin{itemize}
%   \item JSS-specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
%     be used in the references.
%   \item Titles should be in title case.
%   \item Journal titles should not be abbreviated and in title case.
%   \item DOIs should be included where available.
%   \item Software should be properly cited as well. For \proglang{R} packages
%     \code{citation("pkgname")} typically provides a good starting point.
% \end{itemize}
% \end{leftbar}
%
% \end{appendix}

%% -----------------------------------------------------------------------------

\end{document}

### 2024-01-01 misc/JSS article/jss-article-rnw/article.tex
Adding JSS article.

--- 2024-01-01 misc/JSS article/jss-article-rnw/article.Rnw
@@
 %% need no \usepackage{Sweave}
-\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, cache=TRUE}
+
 % \renewenvironment{knitrout}{\setlength{\topsep}{0mm}}{}
-
-<<preliminaries, echo=FALSE, message=FALSE, warning=FALSE, results=hide>>=
-# knitr::opts_chunk$set(cache = FALSE, size = "small")
-options(prompt = "R> ", continue = "+  ", width = 80, digits = 4, useFancyQuotes = FALSE, warn = 1)
-library(data.table)
-library(magrittr)
-library(microbenchmark)
-library(collapse)
-@
 
@@
 %
-\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.
+\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++ based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.
 Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
@@
 %
-<<collapse_topics>>=
-.COLLAPSE_TOPICS
-help("collapse-documentation")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> .COLLAPSE_TOPICS
+\end{Sinput}
+\begin{Soutput}
+ [1] "collapse-documentation"     "fast-statistical-functions"
+ [3] "fast-grouping-ordering"     "fast-data-manipulation"    
+ [5] "quick-conversion"           "advanced-aggregation"      
+ [7] "data-transformations"       "time-series-panel-series"  
+ [9] "list-processing"            "summary-statistics"        
+[11] "recode-replace"             "efficient-programming"     
+[13] "small-helpers"              "collapse-options"          
+\end{Soutput}
+\begin{Sinput}
+R> help("collapse-documentation")
+\end{Sinput}
+\end{Schunk}
 %
@@
 %
-<<aq>>=
-fnobs(airquality)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fnobs(airquality)
+\end{Sinput}
+\begin{Soutput}
+  Ozone Solar.R    Wind    Temp   Month     Day 
+    116     146     153     153     153     153 
+\end{Soutput}
+\end{Schunk}
 This imputes \code{Ozone} and \code{Solar.R} by reference using the month median.
-<<aq_imp>>=
-fmedian(airquality[1:2], airquality$Month, TRA = "replace_na", set = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fmedian(airquality[1:2], airquality$Month, TRA = "replace_na", set = TRUE)
+\end{Sinput}
+\end{Schunk}
 This performs different grouped and/or weighted transformations at once.
-<<aq_trans>>=
-airquality |> fmutate(
-  rad_day = fsum(as.double(Solar.R), Day, TRA = "/"),
-  ozone_deg = Ozone / Temp,
-  ozone_amed = Ozone > fmedian(Ozone, Month, TRA = "fill"),
-  ozone_resid = fmean(Ozone, list(Month, ozone_amed), ozone_deg, "-")
-) |> head(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> airquality |> fmutate(
++    rad_day = fsum(as.double(Solar.R), Day, TRA = "/"),
++    ozone_deg = Ozone / Temp,
++    ozone_amed = Ozone > fmedian(Ozone, Month, TRA = "fill"),
++    ozone_resid = fmean(Ozone, list(Month, ozone_amed), ozone_deg, "-")
++  ) |> head(3)
+\end{Sinput}
+\begin{Soutput}
+  Ozone Solar.R Wind Temp Month Day rad_day ozone_deg ozone_amed ozone_resid
+1    41     190  7.4   67     5   1   0.191    0.6119       TRUE     -10.279
+2    36     118  8.0   72     5   2   0.135    0.5000       TRUE     -15.279
+3    12     149 12.6   74     5   3   0.168    0.1622      FALSE      -3.035
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<GRP>>=
-str(g <- GRP(mtcars, ~ cyl + vs + am))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> str(g <- GRP(mtcars, ~ cyl + vs + am))
+\end{Sinput}
+\begin{Soutput}
+Class 'GRP'  hidden list of 9
+ $ N.groups    : int 7
+ $ group.id    : int [1:32] 4 4 3 5 6 5 6 2 2 5 ...
+ $ group.sizes : int [1:7] 1 3 7 3 4 12 2
+ $ groups      :'data.frame':	7 obs. of  3 variables:
+  ..$ cyl: num [1:7] 4 4 4 6 6 8 8
+  ..$ vs : num [1:7] 0 1 1 0 1 0 0
+  ..$ am : num [1:7] 1 0 1 1 0 0 1
+ $ group.vars  : chr [1:3] "cyl" "vs" "am"
+ $ ordered     : Named logi [1:2] TRUE FALSE
+  ..- attr(*, "names")= chr [1:2] "ordered" "sorted"
+ $ order       : int [1:32] 27 8 9 21 3 18 19 20 26 28 ...
+  ..- attr(*, "starts")= int [1:7] 1 2 5 12 15 19 31
+  ..- attr(*, "maxgrpn")= int 12
+  ..- attr(*, "sorted")= logi FALSE
+ $ group.starts: int [1:7] 27 8 3 1 4 5 29
+ $ call        : language GRP.default(X = mtcars, by = ~cyl + vs + am)
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<gcomp>>=
-dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt
-add_vars(g$groups,
-  fmean(dat, g, w, use.g.names = FALSE) |> add_stub("w_mean_"),
-  fsd(dat, g, w, use.g.names = FALSE) |> add_stub("w_sd_")) |> head(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt
+R> add_vars(g$groups,
++    fmean(dat, g, w, use.g.names = FALSE) |> add_stub("w_mean_"),
++    fsd(dat, g, w, use.g.names = FALSE) |> add_stub("w_sd_")) |> head(2)
+\end{Sinput}
+\begin{Soutput}
+  cyl vs am w_mean_mpg w_mean_disp w_sd_mpg w_sd_disp
+1   4  0  1      26.00       120.3    0.000       0.0
+2   4  1  0      23.02       137.1    1.236      11.6
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<gtrans>>=
-mtcars |> add_vars(fmean(dat, g, w, "-") |> add_stub("w_demean_"),
-                   fscale(dat, g, w) |> add_stub("w_scale_")) |> head(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> mtcars |> add_vars(fmean(dat, g, w, "-") |> add_stub("w_demean_"),
++                     fscale(dat, g, w) |> add_stub("w_scale_")) |> head(2)
+\end{Sinput}
+\begin{Soutput}
+              mpg cyl disp  hp drat    wt  qsec vs am gear carb w_demean_mpg
+Mazda RX4      21   6  160 110  3.9 2.620 16.46  0  1    4    4       0.4357
+Mazda RX4 Wag  21   6  160 110  3.9 2.875 17.02  0  1    4    4       0.4357
+              w_demean_disp w_scale_mpg w_scale_disp
+Mazda RX4             5.027      0.6657       0.6657
+Mazda RX4 Wag         5.027      0.6657       0.6657
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<fsummarise>>=
-mtcars |>
-  fsubset(mpg > 11) |>
-  fgroup_by(cyl, vs, am) |>
-  fsummarise(across(c(mpg, carb, hp), fmean),
-             qsec_w_med = fmean(qsec, wt)) |> head(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> mtcars |>
++    fsubset(mpg > 11) |>
++    fgroup_by(cyl, vs, am) |>
++    fsummarise(across(c(mpg, carb, hp), fmean),
++               qsec_w_med = fmean(qsec, wt)) |> head(2)
+\end{Sinput}
+\begin{Soutput}
+  cyl vs am  mpg  carb    hp qsec_w_med
+1   4  0  1 26.0 2.000 91.00      16.70
+2   4  1  0 22.9 1.667 84.67      21.04
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<fsummarise_2>>=
-mtcars |>
-  fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
-  fgroup_by(cyl, vs, am) |>
-  fmean(wt) |> head(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> mtcars |>
++    fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
++    fgroup_by(cyl, vs, am) |>
++    fmean(wt) |> head(2)
+\end{Sinput}
+\begin{Soutput}
+  cyl vs am sum.wt   mpg carb   hp
+1   4  0  1  2.140 26.00 2.00 91.0
+2   4  1  0  8.805 23.02 1.72 83.6
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<flinreg>>=
-mtcars |>
- fgroup_by(vs) |>
- fmutate(dm_carb = fmean(carb, TRA = "-")) |>
- fsummarise(slope = fsum(mpg, dm_carb) %/=% fsum(dm_carb^2))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> mtcars |>
++   fgroup_by(vs) |>
++   fmutate(dm_carb = fmean(carb, TRA = "-")) |>
++   fsummarise(slope = fsum(mpg, dm_carb) %/=% fsum(dm_carb^2))
+\end{Sinput}
+\begin{Soutput}
+  vs   slope
+1  0 -0.5557
+2  1 -2.0706
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<wstats>>=
-mtcars |>
-    fgroup_by(cyl, vs, am) |>
-    fmutate(o = radixorder(GRPid(), mpg)) |>
-    fsummarise(mpg_min = fmin(mpg),
-               mpg_Q1 = fnth(mpg, 0.25, wt, o = o, ties = "q8"),
-               mpg_mean = fmean(mpg, wt),
-               mpg_median = fmedian(mpg, wt, o = o, ties = "q8"),
-               mpg_mode = fmode(mpg, wt, ties = "max"),
-               mpg_Q3 = fnth(mpg, 0.75, wt, o = o, ties = "q8"),
-               mpg_max = fmax(mpg)) |> head(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> mtcars |>
++      fgroup_by(cyl, vs, am) |>
++      fmutate(o = radixorder(GRPid(), mpg)) |>
++      fsummarise(mpg_min = fmin(mpg),
++                 mpg_Q1 = fnth(mpg, 0.25, wt, o = o, ties = "q8"),
++                 mpg_mean = fmean(mpg, wt),
++                 mpg_median = fmedian(mpg, wt, o = o, ties = "q8"),
++                 mpg_mode = fmode(mpg, wt, ties = "max"),
++                 mpg_Q3 = fnth(mpg, 0.75, wt, o = o, ties = "q8"),
++                 mpg_max = fmax(mpg)) |> head(3)
+\end{Sinput}
+\begin{Soutput}
+  cyl vs am mpg_min mpg_Q1 mpg_mean mpg_median mpg_mode mpg_Q3 mpg_max
+1   4  0  1    26.0  26.00    26.00      26.00     26.0  26.00    26.0
+2   4  1  0    21.5  22.10    23.02      23.17     24.4  24.38    24.4
+3   4  1  1    21.4  22.29    27.74      27.85     30.4  31.79    33.9
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<collap>>=
-collap(wlddev, country + PCGDP + LIFEEX ~ year + income, w = ~ POP) |>
-  head(4)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> collap(wlddev, country + PCGDP + LIFEEX ~ year + income, w = ~ POP) |>
++    head(4)
+\end{Sinput}
+\begin{Soutput}
+        country year              income   PCGDP LIFEEX       POP
+1 United States 1960         High income 12768.7  68.59 7.495e+08
+2      Ethiopia 1960          Low income   658.5  38.33 1.474e+08
+3         India 1960 Lower middle income   500.8  45.27 9.280e+08
+4         China 1960 Upper middle income  1166.1  49.86 1.184e+09
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<ts_example>>=
-fgrowth(airmiles) |> round(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fgrowth(airmiles) |> round(2)
+\end{Sinput}
+\begin{Soutput}
+Time Series:
+Start = 1937 
+End = 1960 
+Frequency = 1 
+ [1]    NA 16.50 42.29 54.03 31.65  2.38 15.23 33.29 54.36 76.92  2.71 -2.10
+[13] 12.91 18.51 32.03 18.57 17.82 13.61 18.19 12.83 13.32  0.01 15.49  4.25
+\end{Soutput}
+\end{Schunk}
 The following code creates an irregular series by removing the 3rd and 15th observation and shows how indexation with the \code{t} argument accounts for this.
 %
-<<ts_example_ctd>>=
-am_ir <- airmiles[-c(3, 15)]
-t <- time(airmiles)[-c(3, 15)]
-fgrowth(am_ir, t = t) |> round(2)
-fgrowth(am_ir, -1:3, t = t) |> head(4)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> am_ir <- airmiles[-c(3, 15)]
+R> t <- time(airmiles)[-c(3, 15)]
+R> fgrowth(am_ir, t = t) |> round(2)
+\end{Sinput}
+\begin{Soutput}
+ [1]    NA 16.50    NA 31.65  2.38 15.23 33.29 54.36 76.92  2.71 -2.10 12.91
+[13] 18.51    NA 17.82 13.61 18.19 12.83 13.32  0.01 15.49  4.25
+\end{Soutput}
+\begin{Sinput}
+R> fgrowth(am_ir, -1:3, t = t) |> head(4)
+\end{Sinput}
+\begin{Soutput}
+         FG1   --    G1  L2G1  L3G1
+[1,] -14.167  412    NA    NA    NA
+[2,]      NA  480 16.50    NA    NA
+[3,] -24.043 1052    NA 119.2 155.3
+[4,]  -2.327 1385 31.65    NA 188.5
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<example_growth>>=
-G(wlddev, c(1, 10), by = POP + LIFEEX ~ iso3c, t = ~ year) |> head(3)
-settransform(wlddev, POP_growth = G(POP, g = iso3c, t = year))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> G(wlddev, c(1, 10), by = POP + LIFEEX ~ iso3c, t = ~ year) |> head(3)
+\end{Sinput}
+\begin{Soutput}
+  iso3c year G1.POP L10G1.POP G1.LIFEEX L10G1.LIFEEX
+1   AFG 1960     NA        NA        NA           NA
+2   AFG 1961  1.917        NA     1.590           NA
+3   AFG 1962  1.985        NA     1.544           NA
+\end{Soutput}
+\begin{Sinput}
+R> settransform(wlddev, POP_growth = G(POP, g = iso3c, t = year))
+\end{Sinput}
+\end{Schunk}
 %
@@
 %
-<<example_growth_continued>>=
-wlddev |> fgroup_by(iso3c) |> fselect(iso3c, year, POP, LIFEEX) |>
-  fmutate(across(c(POP, LIFEEX), G, t = year)) |> head(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlddev |> fgroup_by(iso3c) |> fselect(iso3c, year, POP, LIFEEX) |>
++    fmutate(across(c(POP, LIFEEX), G, t = year)) |> head(2)
+\end{Sinput}
+\begin{Soutput}
+  iso3c year     POP LIFEEX G1.POP G1.LIFEEX
+1   AFG 1960 8996973  32.45     NA        NA
+2   AFG 1961 9169410  32.96  1.917      1.59
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<example_scale_cener>>=
-iris |> fgroup_by(Species) |> fscale() |> head(2)
-STD(iris, ~ Species) |> head(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> iris |> fgroup_by(Species) |> fscale() |> head(2)
+\end{Sinput}
+\begin{Soutput}
+  Species Sepal.Length Sepal.Width Petal.Length Petal.Width
+1  setosa       0.2667      0.1899       -0.357     -0.4365
+2  setosa      -0.3007     -1.1291       -0.357     -0.4365
+\end{Soutput}
+\begin{Sinput}
+R> STD(iris, ~ Species) |> head(2)
+\end{Sinput}
+\begin{Soutput}
+  Species STD.Sepal.Length STD.Sepal.Width STD.Petal.Length STD.Petal.Width
+1  setosa           0.2667          0.1899           -0.357         -0.4365
+2  setosa          -0.3007         -1.1291           -0.357         -0.4365
+\end{Soutput}
+\end{Schunk}
 The following example demonstrates a fixed-effects regression a la \citet{mundlak1978pooling}.
-<<example_reg>>=
-lm(mpg ~ carb + B(carb, cyl), data = mtcars) |> coef()
-@
+\begin{Schunk}
+\begin{Sinput}
+R> lm(mpg ~ carb + B(carb, cyl), data = mtcars) |> coef()
+\end{Sinput}
+\begin{Soutput}
+ (Intercept)         carb B(carb, cyl) 
+     34.8297      -0.4655      -4.7750 
+\end{Soutput}
+\end{Schunk}
 \pkg{collapse} also offers higher-dimensional between and within transformations, powered by C++ code conditionally imported (and accessed directly) from \pkg{fixest}.
 The following detrends GDP per Capita and Life Expectancy at Birth using country-specific cubic polynomials.
-<<example_HD>>=
-HDW(wlddev, PCGDP + LIFEEX ~ iso3c * poly(year, 3), stub = F) |> head(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> HDW(wlddev, PCGDP + LIFEEX ~ iso3c * poly(year, 3), stub = F) |> head(2)
+\end{Sinput}
+\begin{Soutput}
+   PCGDP   LIFEEX
+1  9.964 0.023670
+2 14.045 0.006743
+\end{Soutput}
+\end{Schunk}
 %
@@
 An example using the \code{wlddev} data follows:
-<<indexing>>=
-wldi <- wlddev |> findex_by(iso3c, year)
-wldi |> fsubset(-3, iso3c, year, PCGDP:POP) |> G() |> head(4)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wldi <- wlddev |> findex_by(iso3c, year)
+R> wldi |> fsubset(-3, iso3c, year, PCGDP:POP) |> G() |> head(4)
+\end{Sinput}
+\begin{Soutput}
+  iso3c year G1.PCGDP G1.LIFEEX G1.GINI G1.ODA G1.POP
+1   AFG 1960       NA        NA      NA     NA     NA
+2   AFG 1961       NA     1.590      NA  98.75  1.917
+3   AFG 1963       NA        NA      NA     NA     NA
+4   AFG 1964       NA     1.448      NA  24.48  2.112
+
+Indexed by:  iso3c [1] | year [4 (61)] 
+\end{Soutput}
+\end{Schunk}
 The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. This creates an \class{indexes\_series} of life expectancy and demonstrates its properties.
-<<indexing_2>>=
-LIFEEXi = wldi$LIFEEX
-str(LIFEEXi, width = 70, strict = "cut")
-c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5]))
-G(LIFEEXi[c(1:5, 7:10)])
-@
+\begin{Schunk}
+\begin{Sinput}
+R> LIFEEXi = wldi$LIFEEX
+R> str(LIFEEXi, width = 70, strict = "cut")
+\end{Sinput}
+\begin{Soutput}
+ 'indexed_series' num [1:13176] 32.4 33 33.5 34 34.5 ...
+ - attr(*, "label")= chr "Life expectancy at birth, total (years)"
+ - attr(*, "index_df")=Classes 'index_df', 'pindex' and 'data.frame'..
+  ..$ iso3c: Factor w/ 216 levels "ABW","AFG","AGO",..: 2 2 2 2 2 2 ..
+  .. ..- attr(*, "label")= chr "Country Code"
+  ..$ year : Ord.factor w/ 61 levels "1960"<"1961"<..: 1 2 3 4 5 6 7..
+  .. ..- attr(*, "label")= chr "Year"
+\end{Soutput}
+\begin{Sinput}
+R> c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5]))
+\end{Sinput}
+\begin{Soutput}
+[1] FALSE  TRUE
+\end{Soutput}
+\begin{Sinput}
+R> G(LIFEEXi[c(1:5, 7:10)])
+\end{Sinput}
+\begin{Soutput}
+[1]    NA 1.590 1.544 1.494 1.448    NA 1.366 1.362 1.365
+
+Indexed by:  iso3c [1] | year [9 (61)] 
+\end{Soutput}
+\end{Schunk}
 The transformation and estimation below demonstrate the deep indexation of \\ \class{indexed\_frame}'s, allowing correct computations in arbitrary data masking environments.
-<<indexing_3>>=
-settransform(wldi, PCGDP_ld = Dlog(PCGDP))
-lm(D(LIFEEX) ~ L(PCGDP_ld, 0:5) + B(PCGDP_ld), wldi) |>
-  summary() |> coef() |> round(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> settransform(wldi, PCGDP_ld = Dlog(PCGDP))
+R> lm(D(LIFEEX) ~ L(PCGDP_ld, 0:5) + B(PCGDP_ld), wldi) |>
++    summary() |> coef() |> round(3)
+\end{Sinput}
+\begin{Soutput}
+                   Estimate Std. Error t value Pr(>|t|)
+(Intercept)           0.299      0.007  44.412    0.000
+L(PCGDP_ld, 0:5)--    0.300      0.080   3.735    0.000
+L(PCGDP_ld, 0:5)L1    0.269      0.081   3.332    0.001
+L(PCGDP_ld, 0:5)L2    0.227      0.079   2.854    0.004
+L(PCGDP_ld, 0:5)L3    0.200      0.078   2.563    0.010
+L(PCGDP_ld, 0:5)L4    0.143      0.076   1.871    0.061
+L(PCGDP_ld, 0:5)L5    0.095      0.073   1.301    0.193
+B(PCGDP_ld)          -1.021      0.316  -3.234    0.001
+\end{Soutput}
+\end{Schunk}
 The above example could also have been executed in one line as \code{lm(D(LIFEEX) ~ } \\ \code{L(Dlog(PCGDP), 0:5) + B(Dlog(PCGDP)), wldi)}, log-differencing \code{PCGDP} twice. \newline
@@
 %
-<<rollmean>>=
-BY(LIFEEXi, findex(LIFEEXi)$iso3c, data.table::frollmean, 5) |> head(10)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> BY(LIFEEXi, findex(LIFEEXi)$iso3c, data.table::frollmean, 5) |> head(10)
+\end{Sinput}
+\begin{Soutput}
+ [1]    NA    NA    NA    NA 33.46 33.96 34.46 34.95 35.43 35.92
+
+Indexed by:  iso3c [1] | year [10 (61)] 
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<joins>>=
-df1 <- data.frame(id1 = c(1, 1, 2, 3),
-                  id2 = c("a", "b", "b", "c"),
-                  name = c("John", "Jane", "Bob", "Carl"),
-                  age = c(35, 28, 42, 50))
-df2 <- data.frame(id1 = c(1, 2, 3, 3),
-                  id2 = c("a", "b", "c", "e"),
-                  salary = c(60000, 55000, 70000, 80000),
-                  dept = c("IT", "Marketing", "Sales", "IT"))
-
-join(df1, df2, on = c("id1", "id2"), how = "full", column = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> df1 <- data.frame(id1 = c(1, 1, 2, 3),
++                    id2 = c("a", "b", "b", "c"),
++                    name = c("John", "Jane", "Bob", "Carl"),
++                    age = c(35, 28, 42, 50))
+R> df2 <- data.frame(id1 = c(1, 2, 3, 3),
++                    id2 = c("a", "b", "c", "e"),
++                    salary = c(60000, 55000, 70000, 80000),
++                    dept = c("IT", "Marketing", "Sales", "IT"))
+R> join(df1, df2, on = c("id1", "id2"), how = "full", column = TRUE)
+\end{Sinput}
+\begin{Soutput}
+full join: df1[id1, id2] 3/4 (75%) <m:m> df2[id1, id2] 3/4 (75%)
+  id1 id2 name age salary      dept   .join
+1   1   a John  35  60000        IT matched
+2   1   b Jane  28     NA      <NA>     df1
+3   2   b  Bob  42  55000 Marketing matched
+4   3   c Carl  50  70000     Sales matched
+5   3   e <NA>  NA  80000        IT     df2
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<joins_2>>=
-join(df1, df2, on = c("id1", "id2"), validate = "1:1", attr = "join") |>
-  attr("join") |> str(width = 70, strict = "cut")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> join(df1, df2, on = c("id1", "id2"), validate = "1:1", attr = "join") |>
++    attr("join") |> str(width = 70, strict = "cut")
+\end{Sinput}
+\begin{Soutput}
+left join: df1[id1, id2] 3/4 (75%) <1:1> df2[id1, id2] 3/4 (75%)
+List of 3
+ $ call   : language join(x = df1, y = df2, on = c("id1", "id2"), v"..
+ $ on.cols:List of 2
+  ..$ x: chr [1:2] "id1" "id2"
+  ..$ y: chr [1:2] "id1" "id2"
+ $ match  : 'qG' int [1:4] 1 NA 2 3
+  ..- attr(*, "N.nomatch")= int 1
+  ..- attr(*, "N.groups")= int 4
+  ..- attr(*, "N.distinct")= int 3
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<join_3>>=
-df2$name = df1$name
-join(df1, df2) |> capture.output(type="m") |> strwrap(77) |> cat(sep="\n")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> df2$name = df1$name
+R> join(df1, df2) |> capture.output(type="m") |> strwrap(77) |> cat(sep="\n")
+\end{Sinput}
+\begin{Soutput}
+left join: df1[id1, id2, name] 1/4 (25%) <m:m> df2[id1, id2, name] 1/4 (25%)
+  id1 id2 name age salary dept
+1   1   a John  35  60000   IT
+2   1   b Jane  28     NA <NA>
+3   2   b  Bob  42     NA <NA>
+4   3   c Carl  50     NA <NA>
+Warning in fmatch(x[ixon], y[iyon], nomatch = NA_integer_, count = count, :
+Overidentified match/join: the first 2 of 3 columns uniquely match the
+records. With overid > 0, fmatch() continues to match columns. Consider
+removing columns or setting overid = 0 to terminate the algorithm after 2
+columns (the results may differ, see ?fmatch). Alternatively set overid = 2
+to silence this warning.
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<join_4>>=
-join(df1, df2, on = c("id1", "id2"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> join(df1, df2, on = c("id1", "id2"))
+\end{Sinput}
+\begin{Soutput}
+left join: df1[id1, id2] 3/4 (75%) <m:m> df2[id1, id2] 3/4 (75%)
+duplicate columns: name => renamed using suffix '_df2' for y
+  id1 id2 name age salary      dept name_df2
+1   1   a John  35  60000        IT     John
+2   1   b Jane  28     NA      <NA>     <NA>
+3   2   b  Bob  42  55000 Marketing     Jane
+4   3   c Carl  50  70000     Sales      Bob
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<pivots>>=
-data <- data.frame(type = rep(c("A", "B"), each = 2),
-            type_name = rep(c("Apples", "Bananas"), each = 2),
-            id = rep(1:2, 2), r = abs(rnorm(4)), h = abs(rnorm(4)*2))
-setrelabel(data, id = "Fruit Id", r = "Fruit Radius", h = "Fruit Height")
-print(data)
-vlabels(data)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> data <- data.frame(type = rep(c("A", "B"), each = 2),
++              type_name = rep(c("Apples", "Bananas"), each = 2),
++              id = rep(1:2, 2), r = abs(rnorm(4)), h = abs(rnorm(4)*2))
+R> setrelabel(data, id = "Fruit Id", r = "Fruit Radius", h = "Fruit Height")
+R> print(data)
+\end{Sinput}
+\begin{Soutput}
+  type type_name id      r      h
+1    A    Apples  1 0.1054 3.5624
+2    A    Apples  2 0.7924 3.9380
+3    B   Bananas  1 0.1717 1.1828
+4    B   Bananas  2 0.3461 0.3884
+\end{Soutput}
+\begin{Sinput}
+R> vlabels(data)
+\end{Sinput}
+\begin{Soutput}
+          type      type_name             id              r              h 
+            NA             NA     "Fruit Id" "Fruit Radius" "Fruit Height" 
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<pivot_longer>>=
-(dl <- pivot(data, ids = c("type", "type_name", "id"), labels = "label"))
-vlabels(dl)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> (dl <- pivot(data, ids = c("type", "type_name", "id"), labels = "label"))
+\end{Sinput}
+\begin{Soutput}
+  type type_name id variable        label  value
+1    A    Apples  1        r Fruit Radius 0.1054
+2    A    Apples  2        r Fruit Radius 0.7924
+3    B   Bananas  1        r Fruit Radius 0.1717
+4    B   Bananas  2        r Fruit Radius 0.3461
+5    A    Apples  1        h Fruit Height 3.5624
+6    A    Apples  2        h Fruit Height 3.9380
+7    B   Bananas  1        h Fruit Height 1.1828
+8    B   Bananas  2        h Fruit Height 0.3884
+\end{Soutput}
+\begin{Sinput}
+R> vlabels(dl)
+\end{Sinput}
+\begin{Soutput}
+      type  type_name         id   variable      label      value 
+        NA         NA "Fruit Id"         NA         NA         NA 
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<pivot_wider>>=
-(dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
-namlab(dw)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> (dw <- pivot(data, "id", names = "type", labels = "type_name", how = "w"))
+\end{Sinput}
+\begin{Soutput}
+  id    r_A    r_B   h_A    h_B
+1  1 0.1054 0.1717 3.562 1.1828
+2  2 0.7924 0.3461 3.938 0.3884
+\end{Soutput}
+\begin{Sinput}
+R> namlab(dw)
+\end{Sinput}
+\begin{Soutput}
+  Variable                  Label
+1       id               Fruit Id
+2      r_A  Fruit Radius - Apples
+3      r_B Fruit Radius - Bananas
+4      h_A  Fruit Height - Apples
+5      h_B Fruit Height - Bananas
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<pivot_recast>>=
-(dr <- pivot(data, ids = "id", names = list(from = "type"),
-             labels = list(from = "type_name", to = "label"), how = "r"))
-vlabels(dr)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> (dr <- pivot(data, ids = "id", names = list(from = "type"),
++               labels = list(from = "type_name", to = "label"), how = "r"))
+\end{Sinput}
+\begin{Soutput}
+  id variable        label      A      B
+1  1        r Fruit Radius 0.1054 0.1717
+2  2        r Fruit Radius 0.7924 0.3461
+3  1        h Fruit Height 3.5624 1.1828
+4  2        h Fruit Height 3.9380 0.3884
+\end{Soutput}
+\begin{Sinput}
+R> vlabels(dr)
+\end{Sinput}
+\begin{Soutput}
+        id   variable      label          A          B 
+"Fruit Id"         NA         NA   "Apples"  "Bananas" 
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<list_proc>>=
-(dl <- mtcars |> rsplit(mpg + hp + carb ~ vs + am)) |> str(max.level = 2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> (dl <- mtcars |> rsplit(mpg + hp + carb ~ vs + am)) |> str(max.level = 2)
+\end{Sinput}
+\begin{Soutput}
+List of 2
+ $ 0:List of 2
+  ..$ 0:'data.frame':	12 obs. of  3 variables:
+  ..$ 1:'data.frame':	6 obs. of  3 variables:
+ $ 1:List of 2
+  ..$ 0:'data.frame':	7 obs. of  3 variables:
+  ..$ 1:'data.frame':	7 obs. of  3 variables:
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<list_proc_2>>=
-nest_lm_coef <- dl |>
-  rapply2d(lm, formula = mpg ~ .) |>
-  rapply2d(summary, classes = "lm") |>
-  get_elem("coefficients")
-
-nest_lm_coef |> str(give.attr = FALSE, strict = "cut")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> nest_lm_coef <- dl |>
++    rapply2d(lm, formula = mpg ~ .) |>
++    rapply2d(summary, classes = "lm") |>
++    get_elem("coefficients")
+R> nest_lm_coef |> str(give.attr = FALSE, strict = "cut")
+\end{Sinput}
+\begin{Soutput}
+List of 2
+ $ 0:List of 2
+  ..$ 0: num [1:3, 1:4] 15.8791 0.0683 -4.5715 3.655 0.0345 ...
+  ..$ 1: num [1:3, 1:4] 26.9556 -0.0319 -0.308 2.293 0.0149 ...
+ $ 1:List of 2
+  ..$ 0: num [1:3, 1:4] 30.896903 -0.099403 -0.000332 3.346033 0.03587 ...
+  ..$ 1: num [1:3, 1:4] 37.0012 -0.1155 0.4762 7.3316 0.0894 ...
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<list_proc_3>>=
-nest_lm_coef |> unlist2d(c("vs", "am"), row.names = "variable") |> head(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> nest_lm_coef |> unlist2d(c("vs", "am"), row.names = "variable") |> head(2)
+\end{Sinput}
+\begin{Soutput}
+  vs am    variable Estimate Std. Error t value Pr(>|t|)
+1  0  0 (Intercept) 15.87915    3.65495   4.345 0.001865
+2  0  0          hp  0.06832    0.03449   1.981 0.078938
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<prep, include=FALSE, echo=FALSE>>=
-wldi <- wldi[1:13]
-options(width = 77)
-@
-<<varying_1>>=
-varying(wldi)
-varying(wldi, any_group = FALSE) |> head(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> varying(wldi)
+\end{Sinput}
+\begin{Soutput}
+country    date    year  decade  region  income    OECD   PCGDP  LIFEEX 
+  FALSE    TRUE    TRUE    TRUE   FALSE   FALSE   FALSE    TRUE    TRUE 
+   GINI     ODA     POP 
+   TRUE    TRUE    TRUE 
+\end{Soutput}
+\begin{Sinput}
+R> varying(wldi, any_group = FALSE) |> head(3)
+\end{Sinput}
+\begin{Soutput}
+    country date year decade region income  OECD PCGDP LIFEEX GINI  ODA  POP
+ABW   FALSE TRUE TRUE   TRUE  FALSE  FALSE FALSE  TRUE   TRUE   NA TRUE TRUE
+AFG   FALSE TRUE TRUE   TRUE  FALSE  FALSE FALSE  TRUE   TRUE   NA TRUE TRUE
+AGO   FALSE TRUE TRUE   TRUE  FALSE  FALSE FALSE  TRUE   TRUE TRUE TRUE TRUE
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<pdec>>=
-all.equal(fvar(W(LIFEEXi)) + fvar(B(LIFEEXi)), fvar(LIFEEXi))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> all.equal(fvar(W(LIFEEXi)) + fvar(B(LIFEEXi)), fvar(LIFEEXi))
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<qsu_1>>=
-qsu(LIFEEXi)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> qsu(LIFEEXi)
+\end{Sinput}
+\begin{Soutput}
+             N/T     Mean       SD      Min      Max
+Overall    11670  64.2963  11.4764   18.907  85.4171
+Between      207  64.9537   9.8936  40.9663  85.4171
+Within   56.3768  64.2963   6.0842  32.9068  84.4198
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<qsu_2>>=
-qsu(LIFEEXi, g = wlddev$OECD, w = wlddev$POP, higher = TRUE) |> aperm()
-@
+\begin{Schunk}
+\begin{Sinput}
+R> qsu(LIFEEXi, g = wlddev$OECD, w = wlddev$POP, higher = TRUE) |> aperm()
+\end{Sinput}
+\begin{Soutput}
+, , FALSE
+
+             N/T     Mean      SD      Min      Max     Skew    Kurt
+Overall     9503  63.5476  9.2368   18.907  85.4171  -0.7394  2.7961
+Between      171  63.5476  6.0788  43.0905  85.4171  -0.8041   3.082
+Within   55.5731  65.8807  6.9545  30.3388  82.8832  -1.0323  4.0998
+
+, , TRUE
+
+             N/T     Mean      SD      Min      Max     Skew    Kurt
+Overall     2156  74.9749  5.3627   45.369  84.3563  -1.2966  6.5505
+Between       36  74.9749  2.9256  66.2983  78.6733  -1.3534  4.5999
+Within   59.8889  65.8807  4.4944  44.9513  77.2733   -0.627  3.9839
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<descr>>=
-descr(wlddev, LIFEEX ~ OECD, w = ~ replace_na(POP))
-@
-<<set_width, echo=FALSE>>=
-options(width = 80)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> descr(wlddev, LIFEEX ~ OECD, w = ~ replace_na(POP))
+\end{Sinput}
+\begin{Soutput}
+Dataset: wlddev, 1 Variables, N = 13176, WeightSum = 313233706778
+Grouped by: OECD [2]
+           N   Perc       WeightSum  Perc
+FALSE  10980  83.33  2.49344474e+11  79.6
+TRUE    2196  16.67  6.38892329e+10  20.4
+-----------------------------------------------------------------------------
+LIFEEX (numeric): Life expectancy at birth, total (years)
+Statistics (N = 11659, 11.51% NAs)
+          N   Perc  Ndist   Mean    SD    Min    Max   Skew  Kurt
+FALSE  9503  81.51   8665  63.55  9.24  18.91  85.42  -0.74   2.8
+TRUE   2156  18.49   2016  74.97  5.36  45.37  84.36   -1.3  6.55
+
+Quantiles
+          1%     5%    10%    25%    50%    75%    90%    95%    99%
+FALSE  41.32  37.63  48.98   57.5  65.87  69.68   74.1  32.39  76.18
+TRUE   56.67  65.65  69.69  71.84  75.32  78.61  81.26  81.23  83.59
+-----------------------------------------------------------------------------
+\end{Soutput}
+\end{Schunk}
 
@@
 %
-<<qtab>>=
-library(magrittr) # World after 2015 (latest country data)
-wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
-wlda15 %$% qtab(OECD, income)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> library(magrittr) # World after 2015 (latest country data)
+R> wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
+R> wlda15 %$% qtab(OECD, income)
+\end{Sinput}
+\begin{Soutput}
+       income
+OECD    High income Low income Lower middle income Upper middle income
+  FALSE          45         30                  47                  58
+  TRUE           34          0                   0                   2
+\end{Soutput}
+\end{Schunk}
 This shows the total population (latest post-2015 estimates) in millions.
-<<qtab_2>>=
-wlda15 %$% qtab(OECD, income, w = POP) %>% divide_by(1e6)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlda15 %$% qtab(OECD, income, w = POP) %>% divide_by(1e6)
+\end{Sinput}
+\begin{Soutput}
+       income
+OECD    High income Low income Lower middle income Upper middle income
+  FALSE       93.01     694.89             3063.54             2459.71
+  TRUE      1098.75       0.00                0.00              211.01
+\end{Soutput}
+\end{Schunk}
 This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e. \fct{fmean} is only invoked once).
-<<qtab_3>>=
-wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean) %>% replace_na(0)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean) %>% replace_na(0)
+\end{Sinput}
+\begin{Soutput}
+       income
+OECD    High income Low income Lower middle income Upper middle income
+  FALSE       78.75      62.81               68.30               73.81
+  TRUE        81.09       0.00                0.00               76.37
+\end{Soutput}
+\end{Schunk}
 Finally, this calculates a population-weighted average of life expectancy in each group.
-<<qtab_4>>=
-wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean,
-                wFUN.args = list(w = POP)) %>% replace_na(0)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean,
++                  wFUN.args = list(w = POP)) %>% replace_na(0)
+\end{Sinput}
+\begin{Soutput}
+       income
+OECD    High income Low income Lower middle income Upper middle income
+  FALSE       77.91      63.81               68.76               75.93
+  TRUE        81.13       0.00                0.00               76.10
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench, cache=TRUE>>=
-set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 1)
-set.seed(101)
-m <- matrix(rnorm(1e7), ncol = 1000)
-data <- qDT(replicate(100, rnorm(1e5), simplify = FALSE))
-g <- sample.int(1e4, 1e5, TRUE)
-
-microbenchmark(R = colMeans(m),
-               Rfast = Rfast::colmeans(m),
-               collapse = fmean(m))
-microbenchmark(R = rowsum(data, g, reorder = FALSE),
-               data.table = data[, lapply(.SD, sum), by = g],
-               collapse = fsum(data, g))
-add_vars(data) <- g
-microbenchmark(data.table = data[, lapply(.SD, median), by = g],
-               collapse = data |> fgroup_by(g) |> fmedian())
-d <- data.table(g = unique(g), x = 1, y = 2, z = 3)
-microbenchmark(data.table = d[data, on = "g"],
-               collapse = join(data, d, on = "g", verbose = 0))
-microbenchmark(data.table = melt(data, "g"),
-               collapse = pivot(data, "g"))
-settransform(data, id = rowid(g))
-cols = grep("^V", names(data), value = TRUE)
-microbenchmark(data.table = dcast(data, g ~ id, value.var = cols),
-          collapse = pivot(data, ids = "g", names = "id", how = "w"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 1)
+R> set.seed(101)
+R> m <- matrix(rnorm(1e7), ncol = 1000)
+R> data <- qDT(replicate(100, rnorm(1e5), simplify = FALSE))
+R> g <- sample.int(1e4, 1e5, TRUE)
+R> microbenchmark(R = colMeans(m),
++                 Rfast = Rfast::colmeans(m),
++                 collapse = fmean(m))
+\end{Sinput}
+\begin{Soutput}
+Unit: milliseconds
+     expr   min    lq  mean median    uq    max neval
+        R 9.489 9.538 9.622  9.587 9.625 11.013   100
+    Rfast 4.732 4.757 4.974  4.779 4.826 22.022   100
+ collapse 1.356 1.459 1.685  1.564 1.737  7.917   100
+\end{Soutput}
+\begin{Sinput}
+R> microbenchmark(R = rowsum(data, g, reorder = FALSE),
++                 data.table = data[, lapply(.SD, sum), by = g],
++                 collapse = fsum(data, g))
+\end{Sinput}
+\begin{Soutput}
+Unit: milliseconds
+       expr    min     lq  mean median     uq   max neval
+          R 10.721 11.036 12.69 11.323 11.901 51.54   100
+ data.table 16.200 16.695 18.56 16.940 18.126 60.54   100
+   collapse  5.109  5.512  6.08  5.617  5.952 16.85   100
+\end{Soutput}
+\begin{Sinput}
+R> add_vars(data) <- g
+R> microbenchmark(data.table = data[, lapply(.SD, median), by = g],
++                 collapse = data |> fgroup_by(g) |> fmedian())
+\end{Sinput}
+\begin{Soutput}
+Unit: milliseconds
+       expr   min    lq  mean median    uq   max neval
+ data.table 130.1 130.9 132.5  131.8 133.1 150.5   100
+   collapse 117.1 118.1 119.5  118.6 119.6 130.3   100
+\end{Soutput}
+\begin{Sinput}
+R> d <- data.table(g = unique(g), x = 1, y = 2, z = 3)
+R> microbenchmark(data.table = d[data, on = "g"],
++                 collapse = join(data, d, on = "g", verbose = 0))
+\end{Sinput}
+\begin{Soutput}
+Unit: milliseconds
+       expr    min     lq   mean median     uq    max neval
+ data.table 21.136 23.775 34.691 25.463 45.039 72.758   100
+   collapse  1.178  1.323  1.348  1.356  1.377  1.509   100
+\end{Soutput}
+\begin{Sinput}
+R> microbenchmark(data.table = melt(data, "g"),
++                 collapse = pivot(data, "g"))
+\end{Sinput}
+\begin{Soutput}
+Unit: milliseconds
+       expr   min    lq  mean median    uq   max neval
+ data.table 11.87 14.86 19.92  16.07 17.60 56.57   100
+   collapse 11.67 12.69 19.28  15.83 17.27 60.33   100
+\end{Soutput}
+\begin{Sinput}
+R> settransform(data, id = rowid(g))
+R> cols = grep("^V", names(data), value = TRUE)
+R> microbenchmark(data.table = dcast(data, g ~ id, value.var = cols),
++            collapse = pivot(data, ids = "g", names = "id", how = "w"))
+\end{Sinput}
+\begin{Soutput}
+Unit: milliseconds
+       expr   min    lq   mean median     uq    max neval
+ data.table 63.45 102.5 104.88 104.89 108.03 140.38   100
+   collapse 33.23  71.4  74.14  72.77  75.69  93.31   100
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_2, cache=TRUE>>=
-set.seed(101)
-g_int <- sample.int(1e3, 1e7, replace = TRUE)
-char <- c(letters, LETTERS, month.abb, month.name)
-char <- outer(char, char, paste0)
-g_char <- sample(char, 1e7, replace = TRUE)
-microbenchmark(base_int = unique(g_int), collapse_int = funique(g_int),
-            base_char = unique(g_char), collapse_char = funique(g_char))
-microbenchmark(base_int = match(g_int, 1:1000),
-               collapse_int = fmatch(g_int, 1:1000),
-               base_char = match(g_char, char),
-               data.table_char = chmatch(g_char, char),
-               collapse_char = fmatch(g_char, char), times = 10)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(101)
+R> g_int <- sample.int(1e3, 1e7, replace = TRUE)
+R> char <- c(letters, LETTERS, month.abb, month.name)
+R> char <- outer(char, char, paste0)
+R> g_char <- sample(char, 1e7, replace = TRUE)
+R> microbenchmark(base_int = unique(g_int), collapse_int = funique(g_int),
++              base_char = unique(g_char), collapse_char = funique(g_char))
+\end{Sinput}
+\begin{Soutput}
+Unit: milliseconds
+          expr    min     lq  mean median    uq    max neval
+      base_int 60.165 62.240 64.93 64.143 65.25 113.92   100
+  collapse_int  8.518  9.183 11.16  9.374 12.51  54.94   100
+     base_char 92.549 95.915 98.02 97.444 99.13 124.71   100
+ collapse_char 22.018 23.435 24.55 23.829 24.36  34.54   100
+\end{Soutput}
+\begin{Sinput}
+R> microbenchmark(base_int = match(g_int, 1:1000),
++                 collapse_int = fmatch(g_int, 1:1000),
++                 base_char = match(g_char, char),
++                 data.table_char = chmatch(g_char, char),
++                 collapse_char = fmatch(g_char, char), times = 10)
+\end{Sinput}
+\begin{Soutput}
+Unit: milliseconds
+            expr    min     lq   mean median     uq   max neval
+        base_int 27.159 27.309 29.033 28.026 30.473 32.37    10
+    collapse_int  8.592  8.626  9.294  8.808  9.155 13.23    10
+       base_char 76.203 76.389 80.089 77.340 82.390 88.61    10
+ data.table_char 41.366 41.470 42.315 41.750 42.040 47.57    10
+   collapse_char 26.039 26.153 26.871 26.369 26.630 31.65    10
+\end{Soutput}
+\end{Schunk}
 %

### 2024-01-01 misc/JSS article/jss-article-rnw/article_replication.R

@@ -1,0 +1,404 @@
+#####################################################
+### Replication Script for the JSS Article:
+### collapse: Advanced and Fast Statistical Computing
+###           and Data Transformation in R
+### -------------------------------------------------
+### By: Sebastian Krantz, IfW Kiel
+### E-Mail: sebastian.krantz@ifw-kiel.de
+#####################################################
+
+###################################################
+### code chunk number 1: Preliminaries
+###################################################
+
+options(prompt = "R> ", continue = "+  ", width = 80, digits = 4, useFancyQuotes = FALSE, warn = 1)
+library(data.table)     # v1.14.10
+library(magrittr)       # v2.0.3
+library(microbenchmark) # v1.4.10
+library(collapse)       # v2.0.8
+# Also used: {Rfast} v2.1.0 and {fixest} v0.11.2
+
+###################################################
+### code chunk number 2: collapse Topics and Documentation
+###################################################
+.COLLAPSE_TOPICS
+help("collapse-documentation")
+
+
+###################################################
+### code chunk number 3: Airquality Dataset
+###################################################
+fnobs(airquality)
+
+
+###################################################
+### code chunk number 4: Imputation by Reference
+###################################################
+fmedian(airquality[1:2], airquality$Month, TRA = "replace_na", set = TRUE)
+
+
+###################################################
+### code chunk number 5: Transformation Example
+###################################################
+airquality |> fmutate(
+  rad_day = fsum(as.double(Solar.R), Day, TRA = "/"),
+  ozone_deg = Ozone / Temp,
+  ozone_amed = Ozone > fmedian(Ozone, Month, TRA = "fill"),
+  ozone_resid = fmean(Ozone, list(Month, ozone_amed), ozone_deg, "-")
+) |> head(3)
+
+
+###################################################
+### code chunk number 6: GRP Objects
+###################################################
+str(g <- GRP(mtcars, ~ cyl + vs + am))
+
+
+###################################################
+### code chunk number 7: Aggregation with GRP Objects
+###################################################
+dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt
+add_vars(g$groups,
+  fmean(dat, g, w, use.g.names = FALSE) |> add_stub("w_mean_"),
+  fsd(dat, g, w, use.g.names = FALSE) |> add_stub("w_sd_")) |> head(2)
+
+
+###################################################
[…]

### 2024-02-13 misc/JSS article/jss-article-rnw/article.Rnw
Update article (resubmission).

@@
 options(prompt = "R> ", continue = "+  ", width = 80, digits = 4, useFancyQuotes = FALSE, warn = 1)
-library(data.table)
-library(magrittr)
-library(microbenchmark)
-library(collapse)
+# Loading libraries and installing if unavailable
+if(!requireNamespace("fastverse", quietly = TRUE)) install.packages("fastverse")
+library(fastverse) # loads data.table, collapse, magrittr and kit (not used)
+fastverse_extend(microbenchmark, Rfast, fixest, install = TRUE) # loads and installs if unavailable
+# Package versions used in the article:
+# fastverse 0.3.2, collapse 2.0.10, data.table 1.15.0, magrittr 2.0.3,
+# microbenchmark 1.4.10, Rfast 2.1.0, and fixest 0.11.3
 @
@@
 \Abstract{
-\pkg{collapse} is a large C/C++-based infrastructure package for \proglang{R} facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency.
+\pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency.
 It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g.
-\class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling seamless integration with large parts of the \proglang{R} ecosystem.
+\class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem.
 This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples.
@@
 They are executed on an Apple M1 MacBook Pro with 16 GB unified memory.
-All packages are compiled and used in single-threaded mode to avoid artifacts of different parallel setups and resource utilization.
-The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets.
+The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets, using a large server with many (slow) cores.
 %
 <<bench, cache=TRUE>>=
-set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 1)
+setDTthreads(4)
+set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)
 set.seed(101)
@@
 microbenchmark(R = colMeans(m),
-               Rfast = Rfast::colmeans(m),
+               Rfast = Rfast::colmeans(m, parallel = TRUE, cores = 4),
                collapse = fmean(m))
@@
 \pkg{collapse} is my attempt at providing such a package, and the feedback I have received over the years, particularly from users in academia, government, and international organizations, is a strong indication that I have responded to a need felt in larger parts of the \proglang{R} community.
-As mentioned, a single article cannot summarize \pkg{collapse} in 2024, but there is an excellent \href{https://sebkrantz.github.io/collapse/index.html}{website} with comprehensive \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{Documentation Resources}.
+As mentioned, a single article cannot summarize \pkg{collapse}, but there is an excellent \href{https://sebkrantz.github.io/collapse/index.html}{website} with comprehensive \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{Documentation Resources}.
 
@@
 \section*{Computational details}
-The results in this paper were obtained using \proglang{R} \citep{R} 4.3.0 with \pkg{collapse} 2.0.8, \pkg{data.table} 1.14.10, \pkg{Rfast} 2.1.0, \pkg{fixest} 0.11.2, \pkg{magrittr} \citep{rmagrittr} 2.0.3 and \pkg{microbenchmark} \citep{rmicrobenchmark} 1.4.10.
+The results in this paper were obtained using \proglang{R} \citep{R} 4.3.0 with \pkg{collapse} 2.0.10, \pkg{data.table} 1.15.0, \pkg{Rfast} 2.1.0, \pkg{fixest} 0.11.3, \pkg{magrittr} \citep{rmagrittr} 2.0.3 and \pkg{microbenchmark} \citep{rmicrobenchmark} 1.4.10.
 All packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at https://CRAN.R-project.org/.
-The benchmark was run on an Apple M1 MacBook Pro (2020) with 16GB unified memory and serially compiled CRAN binaries for Mac.
+The benchmark was run on an Apple M1 MacBook Pro (2020) with 16GB unified memory.
+Packages were compiled from source using Homebrew Clang version 16.0.4 with OpenMP enabled and the -O3 optimization flag.
 

### 2024-02-13 misc/JSS article/jss-article-rnw/article_replication.R

@@ -12,11 +12,14 @@
 ###################################################
 
 options(prompt = "R> ", continue = "+  ", width = 80, digits = 4, useFancyQuotes = FALSE, warn = 1)
-library(data.table)     # v1.14.10
-library(magrittr)       # v2.0.3
-library(microbenchmark) # v1.4.10
-library(collapse)       # v2.0.8
-# Also used: {Rfast} v2.1.0 and {fixest} v0.11.2
+
+# Loading libraries and installing if unavailable
+if(!requireNamespace("fastverse", quietly = TRUE)) install.packages("fastverse")
+library(fastverse) # loads data.table, collapse, magrittr and kit (not used)
+fastverse_extend(microbenchmark, Rfast, fixest, install = TRUE) # loads and installs if unavailable
+# Package versions used in the article:
+# fastverse 0.3.2, collapse 2.0.10, data.table 1.15.0, magrittr 2.0.3,
+# microbenchmark 1.4.10, Rfast 2.1.0, and fixest 0.11.3
 
 ###################################################
 ### code chunk number 2: collapse Topics and Documentation
@@ -355,14 +358,15 @@ wlda15 %$% qtab(OECD, income, w = LIFEEX, wFUN = fmean,
 ###################################################
 ### code chunk number 46: Benchmark: Statistics and Data Manipulation
 ###################################################
-set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 1)
+setDTthreads(4)
+set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)
 set.seed(101)
 m <- matrix(rnorm(1e7), ncol = 1000)
 data <- qDT(replicate(100, rnorm(1e5), simplify = FALSE))
 g <- sample.int(1e4, 1e5, TRUE)
 
 microbenchmark(R = colMeans(m),
-               Rfast = Rfast::colmeans(m),
+               Rfast = Rfast::colmeans(m, parallel = TRUE, cores = 4),
                collapse = fmean(m))
 microbenchmark(R = rowsum(data, g, reorder = FALSE),
                data.table = data[, lapply(.SD, sum), by = g],
@@ -376,7 +380,7 @@ microbenchmark(data.table = d[data, on = "g"],
 microbenchmark(data.table = melt(data, "g"),
                collapse = pivot(data, "g"))
 settransform(data, id = rowid(g))
-cols = grep("^V", names(data), value = TRUE)
[…]

### 2024-02-17 misc/JSS article/jss-article-rnw/article.tex

@@
 %
-\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++ based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.
+\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.
 Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
@@
 1   4  0  1    26.0  26.00    26.00      26.00     26.0  26.00    26.0
-2   4  1  0    21.5  22.10    23.02      23.17     24.4  24.38    24.4
-3   4  1  1    21.4  22.29    27.74      27.85     30.4  31.79    33.9
+2   4  1  0    21.5  21.90    23.02      23.16     24.4  24.38    24.4
+3   4  1  1    21.4  22.37    27.74      28.28     30.4  31.51    33.9
 \end{Soutput}
@@
    PCGDP   LIFEEX
-1  9.964 0.023670
-2 14.045 0.006743
+1  8.885 0.023614
+2 13.685 0.006724
 \end{Soutput}
@@
   type type_name id      r      h
-1    A    Apples  1 0.1054 3.5624
-2    A    Apples  2 0.7924 3.9380
-3    B   Bananas  1 0.1717 1.1828
-4    B   Bananas  2 0.3461 0.3884
+1    A    Apples  1 0.9322 0.4315
+2    A    Apples  2 0.2230 0.7225
+3    B   Bananas  1 0.8147 0.4918
+4    B   Bananas  2 0.5642 2.4131
 \end{Soutput}
@@
   type type_name id variable        label  value
-1    A    Apples  1        r Fruit Radius 0.1054
-2    A    Apples  2        r Fruit Radius 0.7924
-3    B   Bananas  1        r Fruit Radius 0.1717
-4    B   Bananas  2        r Fruit Radius 0.3461
-5    A    Apples  1        h Fruit Height 3.5624
-6    A    Apples  2        h Fruit Height 3.9380
-7    B   Bananas  1        h Fruit Height 1.1828
-8    B   Bananas  2        h Fruit Height 0.3884
+1    A    Apples  1        r Fruit Radius 0.9322
+2    A    Apples  2        r Fruit Radius 0.2230
+3    B   Bananas  1        r Fruit Radius 0.8147
+4    B   Bananas  2        r Fruit Radius 0.5642
+5    A    Apples  1        h Fruit Height 0.4315
+6    A    Apples  2        h Fruit Height 0.7225
+7    B   Bananas  1        h Fruit Height 0.4918
+8    B   Bananas  2        h Fruit Height 2.4131
 \end{Soutput}
@@
   id    r_A    r_B    h_A    h_B
-1  1 0.1054 0.1717 3.562 1.1828
-2  2 0.7924 0.3461 3.938 0.3884
+1  1 0.9322 0.8147 0.4315 0.4918
+2  2 0.2230 0.5642 0.7225 2.4131
 \end{Soutput}
@@
   id variable        label      A      B
-1  1        r Fruit Radius 0.1054 0.1717
-2  2        r Fruit Radius 0.7924 0.3461
-3  1        h Fruit Height 3.5624 1.1828
-4  2        h Fruit Height 3.9380 0.3884
+1  1        r Fruit Radius 0.9322 0.8147
+2  2        r Fruit Radius 0.2230 0.5642
+3  1        h Fruit Height 0.4315 0.4918
+4  2        h Fruit Height 0.7225 2.4131
 \end{Soutput}
@@
           1%     5%    10%    25%    50%    75%    90%    95%    99%
-FALSE  41.32  37.63  48.98   57.5  65.87  69.68   74.1  32.39  76.18
-TRUE   56.67  65.65  69.69  71.84  75.32  78.61  81.26  81.23  83.59
+FALSE  41.39  45.78  49.08  57.51  65.98  70.14  74.12  75.63  76.91
+TRUE   56.65  65.98   69.7  71.85  75.38  78.64  81.26  82.43   83.6
 -----------------------------------------------------------------------------
@@
 They are executed on an Apple M1 MacBook Pro with 16 GB unified memory.
-All packages are compiled and used in single-threaded mode to avoid artifacts of different parallel setups and resource utilization.
-The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets.
-%
-\begin{Schunk}
-\begin{Sinput}
-R> set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 1)
+The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets, using a large server with many (slow) cores.
+%
+\begin{Schunk}
+\begin{Sinput}
+R> setDTthreads(4)
+R> set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)
 R> set.seed(101)
@@
 R> microbenchmark(R = colMeans(m),
-+                 Rfast = Rfast::colmeans(m),
++                 Rfast = Rfast::colmeans(m, parallel = TRUE, cores = 4),
 +                 collapse = fmean(m))
@@
      expr   min    lq  mean median    uq   max neval
-        R 9.489 9.538 9.622  9.587 9.625 11.013   100
-    Rfast 4.732 4.757 4.974  4.779 4.826 22.022   100
- collapse 1.356 1.459 1.685  1.564 1.737  7.917   100
+        R 9.632 9.837 9.959  9.852 9.940 12.70   100
+    Rfast 1.315 1.805 2.257  1.840 1.917 11.92   100
+ collapse 1.321 1.432 1.847  1.506 1.603 16.33   100
 \end{Soutput}
@@
        expr    min    lq   mean median    uq   max neval
-          R 10.721 11.036 12.69 11.323 11.901 51.54   100
- data.table 16.200 16.695 18.56 16.940 18.126 60.54   100
-   collapse  5.109  5.512  6.08  5.617  5.952 16.85   100
+          R 10.715 11.19 11.550 11.295 11.61 16.89   100
+ data.table  9.154 18.70 19.743 19.229 20.19 30.22   100
+   collapse  2.330  2.89  3.619  3.078  3.71 15.94   100
 \end{Soutput}
@@
        expr    min     lq   mean median     uq   max neval
- data.table 130.1 130.9 132.5  131.8 133.1 150.5   100
-   collapse 117.1 118.1 119.5  118.6 119.6 130.3   100
+ data.table 136.36 137.50 140.23 138.54 140.62 173.7   100
+   collapse  68.42  76.85  82.25  80.28  83.36 163.3   100
 \end{Soutput}
@@
        expr   min     lq   mean median     uq    max neval
- data.table 21.136 23.775 34.691 25.463 45.039 72.758   100
-   collapse  1.178  1.323  1.348  1.356  1.377  1.509   100
+ data.table 8.915 10.881 16.398 13.166 16.903 73.045   100
+   collapse 1.228  1.379  1.452  1.413  1.459  2.337   100
 \end{Soutput}
@@
        expr   min    lq  mean median    uq   max neval
- data.table 11.87 14.86 19.92  16.07 17.60 56.57   100
-   collapse 11.67 12.69 19.28  15.83 17.27 60.33   100
+ data.table 11.79 12.84 17.99  15.87 17.68 73.75   100
+   collapse 11.73 13.77 20.37  15.69 18.17 76.75   100
 \end{Soutput}
@@
        expr   min    lq  mean median     uq   max neval
- data.table 63.45 102.5 104.88 104.89 108.03 140.38   100
-   collapse 33.23  71.4  74.14  72.77  75.69  93.31   100
+ data.table 60.01 69.99 90.12  74.28 111.73 157.7   100
+   collapse 32.59 41.80 63.15  48.87  63.98 177.0   100
 \end{Soutput}
@@
           expr   min     lq   mean median     uq    max neval
-      base_int 60.165 62.240 64.93 64.143 65.25 113.92   100
-  collapse_int  8.518  9.183 11.16  9.374 12.51  54.94   100
-     base_char 92.549 95.915 98.02 97.444 99.13 124.71   100
- collapse_char 22.018 23.435 24.55 23.829 24.36  34.54   100
+      base_int 60.07 61.852  67.56 65.613  68.55 133.05   100
+  collapse_int  6.69  9.222  10.34  9.581  10.18  17.56   100
+     base_char 91.79 94.570 102.38 99.497 103.65 170.41   100
+ collapse_char 22.29 23.879  26.83 24.577  27.58  80.46   100
 \end{Soutput}
@@
             expr    min     lq   mean median     uq   max neval
-        base_int 27.159 27.309 29.033 28.026 30.473 32.37    10
-    collapse_int  8.592  8.626  9.294  8.808  9.155 13.23    10
-       base_char 76.203 76.389 80.089 77.340 82.390 88.61    10
- data.table_char 41.366 41.470 42.315 41.750 42.040 47.57    10
-   collapse_char 26.039 26.153 26.871 26.369 26.630 31.65    10
+        base_int 26.873 26.993 28.261 27.094 27.306 33.25    10
+    collapse_int  8.603  8.633  9.352  8.662  8.723 15.53    10
+       base_char 84.091 84.726 86.127 85.100 86.008 91.23    10
+ data.table_char 40.970 41.169 42.597 41.307 41.407 48.49    10
+   collapse_char 30.482 30.541 30.774 30.622 30.803 31.47    10
 \end{Soutput}

### 2024-03-04 misc/JSS article/jss-article-rnw/article.Rnw
Implementing JSS form comments: commas after e.g. and i.e.

@@
 \pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency.
-It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g.
-\class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem.
+It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem.
 This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples.
@@
 The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors.
-\pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g.
-\emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}).
+\pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g., \emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}).
 It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc.
@@
 The short answer is to make computations in \proglang{R} as flexible and powerful as possible.
-The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.}; (3) increase the memory efficiency and flexibility of \proglang{R} programs\footnote{E.g.
-by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.}; and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline
+The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.}; (3) increase the memory efficiency and flexibility of \proglang{R} programs\footnote{E.g., by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.}; and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline
 
@@
 In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recursive operations, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable.
-In short, it is very utile for complex statistical workflows, rich datasets (e.g.
-surveys), and for integrating with different parts of the \proglang{R} ecosystem.
+In short, it is very utile for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem.
 On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive).
@@
 \end{Code}
-with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e.
-simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation).
+with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation).
 The use of Single Instruction Multiple Data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.}.
@@
 \begin{Code}
-fmean(mtcars$mpg)                                  # Vector
-fmean(EuStockMarkets)                              # Matrix
-fmean(mtcars)                                      # Data Frame
-fmean(mtcars$mpg, w = mtcars$wt)                   # Weighted mean
-fmean(mtcars$mpg, g = mtcars$cyl)                  # Grouped mean
-fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)   # Weighted group mean
-fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt) # Of data frame
-fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")    # Replace data by g. mean
+fmean(mtcars$mpg)
+fmean(EuStockMarkets)
+fmean(mtcars)
+fmean(mtcars$mpg, w = mtcars$wt)
+fmean(mtcars$mpg, g = mtcars$cyl)
+fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)
+fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt)
+fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")
 \end{Code}
@@
   2 && \code{"replace"} && replace data but preserve missing values in \code{x} \\
-  3 && \code{"-"}   && subtract (i.e. center) \\
+  3 && \code{"-"}   && subtract (i.e., center) \\
   4 && \code{"-+"}  && center on overall average statistic \\
-  5 && \code{"/"}   && divide (i.e. scale) \\
-  6 && \code{"\%"}     && compute percentages (i.e. divide and multiply by 100) \\
+  5 && \code{"/"}   && divide (i.e., scale) \\
+  6 && \code{"\%"}     && compute percentages (i.e., divide and multiply by 100) \\
   7 && \code{"+"} && add \\
   8 && \code{"*"} && multiply \\
-  9 && \code{"\%\%"} && modulus (i.e. remainder from division by  \code{STATS}) \\
-  10 && \code{"-\%\%"} && subtract modulus (i.e. make data divisible by \code{STATS})
+  9 && \code{"\%\%"} && modulus (i.e., remainder from division by  \code{STATS}) \\
+  10 && \code{"-\%\%"} && subtract modulus (i.e., make data divisible by \code{STATS})
   \end{tabular}
@@
 to the creation of new classes \class{indexes\_series} and \class{indexed\_frame} which inherit from the former.
-Any data frame-like object can be an \class{indexed\_frame} with any number of individual and/or time identifiers (e.g.
-an indexed \class{data.table} is fully functional for other operations).
+Any data frame-like object can be an \class{indexed\_frame} with any number of individual and/or time identifiers (e.g., an indexed \class{data.table} is fully functional for other operations).
 The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}.
@@
 <<qtab>>=
-library(magrittr) # World after 2015 (latest country data)
+library("magrittr") # World after 2015 (latest country data)
 wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
@@
 @
-This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e. \fct{fmean} is only invoked once).
+This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only invoked once).
 <<qtab_3>>=

### 2024-03-04 misc/JSS article/jss-article-rnw/article.Rnw
Implementing further JSS comments.

@@
 %% need no \usepackage{Sweave}
+%% need no \usepackage{Sweave.sty}
 \SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, cache=TRUE}
@@
 # knitr::opts_chunk$set(cache = FALSE, size = "small")
-options(prompt = "R> ", continue = "+  ", width = 80, digits = 4, useFancyQuotes = FALSE, warn = 1)
+options(prompt = "R> ", continue = "+  ", width = 77, digits = 4, useFancyQuotes = FALSE, warn = 1)
 # Loading libraries and installing if unavailable
@@
 \begin{Code}
-FUN(x, g = NULL, [w = NULL,] TRA = NULL, [na.rm = TRUE,]
-    use.g.names = TRUE, drop = TRUE, [nthreads = 1L,] ...)
+FUN(x, g = NULL, [w = NULL], TRA = NULL, [na.rm = TRUE],
+    use.g.names = TRUE, drop = TRUE, [nthreads = 1L], ...)
 \end{Code}
 with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation).
-The use of Single Instruction Multiple Data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.}.
-The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there, not here for parsimony) demonstrate their basic usage.
+The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.}.
+The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there) demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects.
+As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette} on object handling, these statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix} and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes.
+Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods.
+Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome.
+For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix.
+%, and attributes are preserved as much as possible ()
 \begin{Code}
@@
 \subsection{Transformations}
-The \code{TRA} argument, more specifically, invokes the \fct{TRA} function for column-wise (grouped) replacing and sweeping operations (by reference). Its syntax is
+The final example invoking \code{TRA} expands the mean vector to full length, like \code{stats::ave(mtcars$mpg, mtcars$cyl)}, but much faster.
+The \code{TRA} argument invokes the \fct{TRA} function for column-wise (grouped) replacing and sweeping operations (by reference).
+Its syntax is
 \begin{Code}
@@
 \end{Code}
-where \code{STATS} is a vector/matrix/data.frame of statistics used to transform \code{x}. The \code{FUN} argument supports 11 different operations indicated using either an integer or string.
+where \code{STATS} is a vector/matrix/data.frame of statistics used to transform \code{x}. Table~\ref{tab:TRA} lists the 11 possible \code{FUN} operations, toggled using either an integer or string.
 %
@@
 }
+\caption{\label{tab:TRA} Arguments to \code{TRA()}.}
 \end{table}
@@
 %
-Similar, data can be transformed, here also using the S3 generic \fct{fscale} function.
+Similar, data can be transformed, here using the S3 generic \fct{fscale} function.
 %
@@
 Any modern laptop can calculate 1 million regression slopes in less than 1 second like this.
+Another neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse_is_sick/sick.html}{blog post}, involved vectorizing an expression to compute the p-value across 300k groups for a simulation study, yielding a 70x performance increase.
 \newline
@@
 It is possible to apply time series and panel data transformations without any indexation by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation.
-The following list summarizes \pkg{collapse}'s time series and panel data functions.
+Table~\ref{tab:TSfun} summarizes \pkg{collapse}'s time series and panel data architecture.
 %
@@
 \end{tabular}
+\caption{\label{tab:TSfun} Time series and panel data architecture.}
 \end{table}
@@
 @
-The following example demonstrates a fixed-effects regression a la \citet{mundlak1978pooling}.
+The following example demonstrates a fixed-effects regression \`a la \citet{mundlak1978pooling}.
 <<example_reg>>=
@@
 Unlike \class{pdata.frame}, an \\ \class{indexed\_frame} is a deeply indexed structure, i.e., every series inside the frame is already an \class{indexes\_series} and contains, in its \class{index\_df} attribute, an external pointer to the \class{index\_df} attribute of the frame (to avoid duplication in memory).
-A comprehensive set of \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \pkg{flag/L}, ensure that these objects behave in a time-/panel-aware manor in any caller environment (created by \code{with}, \code{lm} etc.).
+A comprehensive set of \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()}, ensure that these objects behave in a time-/panel-aware manor in any caller environment (created by \fct{with}, \fct{lm} etc.).
 % A basic demonstration with World Bank panel data showcases the flexibility of these classes.
@@
 @
-The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. This creates an \class{indexes\_series} of life expectancy and demonstrates its properties.
+The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. This creates an \class{indexes\_series} of life expectancy and demonstrates its properties:
 <<indexing_2>>=
@@
 \class{indexed\_series} and \class{indexed\_frame}, on the other hand, work anywhere and can be superimposed on any suitable object (such as \class{sf} to create a spatiotemporal panel), as long as \pkg{collapse}'s functions (\code{flag()/L()} etc.) are used to perform the time-based computations.
-The \class{index\_df} attached to these objects can be used with other general tools such as \code{collapse::BY()} to perform grouped computations on these objects with 3rd-party functions.
+The \class{index\_df} attached to these objects can be used with other general tools such as \code{collapse::BY()} to perform grouped computations using 3rd-party functions.
 An example of calculating a 5-year rolling average is given below.
@@
 %
-While \pkg{collapse} has a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{Data Manipulation Functions}, its implementations of table joins and pivots is particularly noteworthy since they offer several new features such as various verbose options for table joins, pivots supporting variable labels, and 'recast' pivots.
+While \pkg{collapse} has a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, its implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} is particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and 'recast' pivots.
+Both implementations also provide outstanding computational performance and memory efficiency.
 %
@@
 By default (\code{verbose = 1}), it prints information about the join operation and number of records joined.
-Users can request the generation of a '.join' column, akin to \proglang{STATA}'s '\_merge' column, indicating the origin of records in the joined table, as shown below.
+Users can request the generation of a '.join' column (\code{column = TRUE}), akin to \proglang{STATA}'s '\_merge' column, indicating the origin of records in the joined table, as shown below.
 %
@@
 %
-An alternative to the join column is to request an attribute that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}).
+An alternative to the join column is to request an attribute that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}, implementing a vectorized hash join).
 Users can also invoke the \code{validate} argument to check the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.
@@
 Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join (using \fct{fmatch}) followed by sorting, particularly if the data is already sorted on the keys.
-} Additionally, by default (\code{multiple = FALSE}), only the first matches from both tables are joined to avoid silent cartesian duplication of records.
+} Additionally, by default (\code{multiple = FALSE}), only the first matches from \code{y} are joined to avoid silent cartesian duplication of records.
 In multi-match settings, this will be reflected by few records from \code{y} being used.
@@
 df2$name = df1$name
-join(df1, df2) |> capture.output(type="m") |> strwrap(77) |> cat(sep="\n")
+join(df1, df2) |> capture.output(type = "m") |>
+  strwrap(77) |> cat(sep = "\n")
 @
@@
 Finally, since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is no modern replacement for \code{reshape2::recast()}.
-This requires \proglang{R} users to consecutively call two reshaping functions, incurring a high cost in terms of syntax and memory for such frequently required data transpositions.
+This requires \proglang{R} users to consecutively call two reshaping functions, incurring a high cost in terms of syntax and memory.
 \newline
 
-\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, can handle complex labelled data without loss of information, and offers best-in-\proglang{R} computational performance and memory efficiency.
+\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, and supports complex labelled data without loss of information.
 Its syntax is:
@@
 %
-Below, its use is demonstrated with a generated dataset about fruits.
+Below, I demonstrate it with a generated dataset about fruits.
 We could also think about a survey with households and individuals, or sectors and firms.
@@
 Notably, it is also possible to perform longer and recast pivots without id variables.
-The recast pivot without IDs resembles a generalization of \code{data.table::transpose}, albeit slightly less efficient.
+The recast pivot without ids resembles a generalization of \code{data.table::transpose()}, albeit slightly less efficient.
 %
@@
 wldi <- wldi[1:13]
-options(width = 77)
+# options(width = 77)
 @
@@
 <<set_width, echo=FALSE>>=
-options(width = 80)
+# options(width = 80)
 @
@@
 %
-This section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks.
+This Section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks.
 They are executed on an Apple M1 MacBook Pro with 16 GB unified memory.

### 2024-03-04 misc/JSS article/jss-article-rnw/article.Rnw
Minors + large footnote on benchmarking.

@@
 \pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency.
-It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem.
+It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling seamless integration with large parts of the \proglang{R} ecosystem.
 This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples.
@@
 It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools.
-This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures and powerful data manipulation functions.
-The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors.
-\pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g., \emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}).
+This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures, and powerful data manipulation functions.
+The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference, and aversion of logical vectors.
+\pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g., \class{tibble}, \class{data.table}, \class{sf}, \class{xts}, \class{pdata.frame}).
 It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc.
-with much faster \pkg{collapse} equivalents.} \newline
+with much faster \pkg{collapse} equivalents.
+See Section~\ref{sec:glob_opt}.} \newline
 
@@
 The short answer is to make computations in \proglang{R} as flexible and powerful as possible.
-The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.}; (3) increase the memory efficiency and flexibility of \proglang{R} programs\footnote{E.g., by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.}; and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline
+The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition;\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.} (3) increase the memory efficiency and flexibility of \proglang{R} programs;\footnote{E.g., by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.} and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline
 
@@
 It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object.
-It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest} and \pkg{xts}.
+It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest}, and \pkg{xts}.
 It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}.
@@
 It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}.
-It provides general (recursive) tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr} and \pkg{rrapply}.
-Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes and requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics.
+It provides general (recursive) tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}.
+Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics.
 In summary, \pkg{collapse} is a foundation package for statistical computing and data manipulation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient.
@@
 
-Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars} and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy} or \pkg{StatsBase.jl} \citep{jlstatsbase}.
+Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars}, and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy}, or \pkg{StatsBase.jl} \citep{jlstatsbase}.
 % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
@@
 In short, it is very utile for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem.
-On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive).
-There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.}.
-Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function.
-However, as demonstrated below, the possibility of combining vectorized statistical functions also permits calculating more complex statistics in a vectorized way.
+On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive).
+There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function.
+However, as demonstrated in Section~\ref{ssec:vfat}, vectorized statistical functions can be combined to calculate more complex statistics in a vectorized way.
 \newline
@@
 %
-\pkg{collapse} is too large and complex to fully present it in a single article or even to present selected topics in depth.
-The following sections therefore briefly introduce its key components: (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) joins and reshaping; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options (configurability).
-Section (\ref{sec:bench}) provides a small benchmark, Section (\ref{sec:conclusion}) concludes.
-For deeper engagement with \pkg{collapse}, the short vignette summarizing available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{Documentation and Resources} is an excellent starting point.
+\pkg{collapse} is too large and complex to fully present it in a single article, or even to present selected topics in depth.
+The following sections therefore briefly introduce its key components: (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) joins and reshaping; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options.
+Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
+For deeper engagement with \pkg{collapse}, the short vignette summarizing available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} is an excellent starting point.
 %
@@
 %
-The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{Fast Statistical Functions}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs} and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.
+The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs} and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.
 Specifically, operations such as calculating the mean via the S3 generic \code{fmean()} function are vectorized across columns and groups.
@@
 \end{Code}
-with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation).
-The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.}.
-The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there) demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects.
-As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette} on object handling, these statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix} and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes.
+with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading.\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation).
+The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there) demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects.
+As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes.
 Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods.
@@
 The \code{set} argument can also be passed to \emph{Fast Statistical Functions} to toggle transformation by reference.
-The following examples demonstrate how this design allows flexible ad-hoc transformations using \proglang{R}'s built-in \code{airquality} dataset with daily measurements for 5 months.
+The following examples demonstrate how this design allows flexible ad-hoc transformations using \proglang{R}'s built-in \code{airquality} dataset with daily measurements in New York from May to September 1973.
 %
@@
 %
-Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, the cost of grouping can be optimized by using factors or, even better, \class{GRP} objects, which contain all information for \pkg{collapse}'s vectorized functions to operate across groups efficiently.
+Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, the cost of grouping can be optimized by using factors or, even better, \class{GRP} objects, which readily contain all information \pkg{collapse}'s vectorized statistical functions might require to operate across groups.
 These objects can be created with \code{GRP()}.
@@
 \class{GRP} objects make grouped statistical computations in \pkg{collapse} fully programmable.
-Below, the object is used with the \emph{Fast Statistical Functions} and some utility functions to aggregate the data (with optional weights) efficiently.
+Below, the object is used with the \emph{Fast Statistical Functions} and some utility functions to efficiently aggregate data (with optional frequency weights).
 %
@@
 Thus each iteration of the numerical routine raises the vector \textbf{x} to a different power ($a$), sums it in 1 million groups ($j$) to generate \textbf{y}, and computes the Euclidean distance to \textbf{z} (using \code{collapse::fdist}).
-Without grouping objects and vectorization, this would have been difficult to handle within reasonable computing times (a few seconds on the M1).}
+Without grouping objects and vectorization, this would have been difficult to handle within reasonable computing times (of a few seconds on the M1).}
 %
@@
 The following code calculates weighted group means.
-By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.
+By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::fgroup\_by}.
+The latter requires an additional C routine to convert the grouping object to a \class{GRP} object, and is thus much less efficient.}
 %
@@
 %
-\subsection{Vectorizations for advanced tasks}
+\subsection{Vectorizations for advanced tasks} \label{ssec:vfat}
 %
@@
 Any modern laptop can calculate 1 million regression slopes in less than 1 second like this.
-Another neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse_is_sick/sick.html}{blog post}, involved vectorizing an expression to compute the p-value across 300k groups for a simulation study, yielding a 70x performance increase.
+Another neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse_is_sick/sick.html}{blog post}, vectorizes an expression to compute the p-value across 300k groups for a simulation study, yielding a 70x performance increase.
 \newline
 
-\pkg{collapse} also supplies advanced statistics, such as weighted medians and modes.
-The following example calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}\footnote{\pkg{collapse} calculates weighted quantiles by replacing the sample size with the sum of weights and 1 with the minimum non-zero weight in the respective quantile definition.
-See \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for more details.}, and a weighted maximum mode.\footnote{The weighted maximum mode is the largest element with the maximum sum of weights.}
+\pkg{collapse} also vectorizes advanced statistics, such as weighted medians and modes.
+The following example calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} calculates weighted quantiles by replacing the sample size with the sum of weights and 1 with the minimum non-zero weight in the respective quantile definition.
+See \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for more details.} and a weighted maximum mode.\footnote{The weighted maximum mode is the largest element with the maximum sum of weights.}
 %
@@
 %
-\pkg{collapse} provides a flexible high-performance architecture to perform (\emph{time aware}) computations on time series and panel series.
+\pkg{collapse} also provides a flexible high-performance architecture to perform (\emph{time aware}) computations on time series and panel series.
 In particular, the user enjoys great flexibility in deciding the desired degree of indexation and mode of computation.
-It is possible to apply time series and panel data transformations without any indexation by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation.
+It is possible to apply time series and panel data transformations without any indexation by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion, or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation.
 Table~\ref{tab:TSfun} summarizes \pkg{collapse}'s time series and panel data architecture.
@@
 \begin{tabular}{p{\textwidth}}
-\emph{Classes, Constructors and Utilities} \\
+\emph{Classes, constructors and utilities} \\
 \code{findex\_by(), findex(), unindex(), reindex(), timeid(), is\_irregular(), to\_plm()} $+$ rich set of S3 methods for 'indexed\_frame', 'indexed\_series' and 'index\_df'. \\\\
-\emph{Core Time-Based Functions} \\
+\emph{Core time-based functions} \\
 \code{flag(), fdiff(), fgrowth(), fcumsum(), psmat()} [panel data to array conversions] \code{psacf(), pspacf(), psccf()} [autocorrelation functions for panel data] \\\\
-\emph{Data Transformation Functions with Supporting Methods} \\
+\emph{Data transformation functions with supporting methods} \\
 \code{f[hd]between(), f[hd]within(), fscale()} [scaling and (higher-dimensional) centering] \\\\
-\emph{Data Manipulation Functions with Supporting Methods} \\
+\emph{Data manipulation functions with supporting methods} \\
 \code{fsubset(), funique(), roworder[v]()} [internal], and \code{na\_omit()} [internal] \\\\
@@
 In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation.
-If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years.}.
-If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id.
+If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years.} If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id.
 For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}.
@@
 
-In comparison with existing solutions, the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.}.
-Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures for time series operations or native handling of irregularity in basic operations.
+In comparison with existing solutions, the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions.\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.} Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures for time series operations or native handling of irregularity in basic operations.
 \class{indexed\_series} and \class{indexed\_frame}, on the other hand, work anywhere and can be superimposed on any suitable object (such as \class{sf} to create a spatiotemporal panel), as long as \pkg{collapse}'s functions (\code{flag()/L()} etc.) are used to perform the time-based computations.
@@
 By default (\code{verbose = 1}), it prints information about the join operation and number of records joined.
-Users can request the generation of a '.join' column (\code{column = TRUE}), akin to \proglang{STATA}'s '\_merge' column, indicating the origin of records in the joined table, as shown below.
+Users can request the generation of a \code{.join} column (\code{column = TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, indicating the origin of records in the joined table.
 %
@@
 %
-An alternative to the join column is to request an attribute that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}, implementing a vectorized hash join).
+An alternative to the join column is to request an attribute that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}).
 Users can also invoke the \code{validate} argument to check the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.
@@
 First, \code{collapse::join} is also class-agnostic and preserves the attributes of \code{x}.
-It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"} or \code{"anti"}) and default to \code{"left"}, so the default behavior simply adds columns to \code{x}.
+It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"}, or \code{"anti"}) and defaults to \code{"left"}, so the default behavior simply adds columns to \code{x}.
 By default (\code{sort = FALSE}), the order of rows in \code{x} is also preserved.
@@
 This is congruent to the principle of adding columns to \code{x} and altering this table as little as possible.
-Another option, \code{drop.dup.cols = "x"/"y"} can be used to simply drop duplicate columns from \code{x} or \code{y} before the join operation.
+Another option, \code{drop.dup.cols = "x"/"y"}, can be used to simply drop duplicate columns from \code{x} or \code{y} before the join operation.
 %
@@
 The reshaping/pivoting functionality of commercial and open source software is also unsatisfying for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data.
-Such datasets can only be reshaped by losing these labels or additional manual efforts to retain them.
-Modern \proglang{R} packages also offer different functions for different reshaping operations, such as \code{data.table::melt/ tidyr::pivot_longer} to combine columns and \code{data.table::dcast/tidyr::pivot_wider} to expand them, requiring users to learn both.
-Finally, since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is no modern replacement for \code{reshape2::recast()}.
-This requires \proglang{R} users to consecutively call two reshaping functions, incurring a high cost in terms of syntax and memory.
+Such datasets can presently only be reshaped by losing these labels or additional manual efforts to retain them.
+Modern \proglang{R} packages also offer different functions for different reshaping operations, such as \code{data.table::melt/tidyr::pivot_longer} to combine columns and \code{data.table::dcast/tidyr::pivot_wider} to expand them, requiring users to learn both.
+Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two reshaping functions and incurring a high cost in terms of syntax and memory.
 \newline
@@
 %
-Below, I demonstrate it with a generated dataset about fruits.
-We could also think about a survey with households and individuals, or sectors and firms.
+The demonstration below employs a generated dataset about fruits.
+We could equivalently think about a survey with households and individuals, or sectors and firms.
 Variable labels are stored in \code{attr(column, "label")}.
-The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples with real data.
+The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples.
 %
@@
 To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = c(...))}.
-If \code{labels = "lab_name"} is specified, variable labels are saved to an additional column named 'lab\_name'.
-In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to assign alternative names to the 'variable' and 'value' columns, respectively.
+If \code{labels = "lab_name"} is specified, variable labels are saved to an additional column named \code{lab\_name}.
+In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to assign alternative names to the \code{variable} and \code{value} columns, respectively.
 %
@@
 %
-In general, \fct{pivot} only requires essential information and intelligently guesses the rest.
+\fct{pivot} only requires essential information and intelligently guesses the rest.
 For example, the same result could have been obtained by \code{pivot(data, values = c("r", "h"), labels = "label")}.
@@
 %
-For the recast pivot (\code{how = "recast"}), unless a column named 'variable' exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}.
-Again, taking along labels is optional, and omitting either the list's 'from' or 'to' elements will omit the respective operations.
+For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}.
+Again, taking along labels is optional, and omitting either the list's \code{from} or \code{to} elements will omit the respective operations.
 %
@@
 %
-Often, in programming, nested structures are needed.
+Often in programming, nested structures are needed.
 A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model object, performance statistics, and predictions) in a list.
@@
 A typical use case in development involves serving different data according to user choices, e.g., in response to nested user inputs in shiny apps.
-Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data.
-\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{List Processing Functions} attempt to provide a basic toolkit.
+Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data.
+\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit.
 \newline
@@
 \pkg{collapse}'s list processing toolkit provides further useful functions such as \fct{t\_list} to turn lists of lists inside out, \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, and \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes.
-A non-recursive and class-agnostic \fct{rowbind} function also exists to efficiently bind lists of data frame-like objects (like \code{data.table::rbindlist}).
+A non-recursive and class-agnostic \fct{rowbind} function also exists to efficiently bind lists of data frame-like objects (like \code{data.table::rbindlist()}).
 %
@@
 %
-\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{Summary Statistics Functions} provide a parsimonious and powerful toolset to examine complex datasets.
+\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious and powerful toolset to examine complex datasets.
 A particular focus has been on providing tools for examining longitudinal (panel) data.
-Recall the indexed world development panel (\code{wldi}) from Section (\ref{sec:ts_ps}).
+Recall the indexed world development panel (\code{wldi}) from Section~\ref{sec:ts_ps}.
 The function \fct{varying} indicates which of these variables are time-varying:
@@
 A related exercise is to decompose the variance of a panel series into a component due to variation between countries and one capturing variance within countries over time.
-Using the \code{W()/B()} operators and the \code{LIFEEXi} \class{indexed\_series} from Section (\ref{sec:ts_ps}), this is easily demonstrated
+Using the \code{W()/B()} operators and the \code{LIFEEXi} \class{indexed\_series} from Section~\ref{sec:ts_ps}, this is easily demonstrated
 %
@@
 %
-While \fct{descr} does not support panel-variance decompositions like \fct{qsu}, it also provides detailed (grouped, weighted) frequency tables for categorical data and is thus very useful for complex survey data.
+While \fct{descr} does not support panel-variance decompositions like \fct{qsu}, it also computes detailed (grouped, weighted) frequency tables for categorical data and is thus very utile with complex surveys.
 A \code{stepwise} argument toggles describing one variable at a time, allowing users to naturally 'click-through' a large dataset rather than printing a massive output to the console.
@@
 <<qtab>>=
-library("magrittr") # World after 2015 (latest country data)
+library("magrittr")
 wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
@@
 @
-This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only invoked once).
+This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only called once).
 <<qtab_3>>=
@@
 \pkg{dplyr} \citep{rdplyr}: \fct{fselect}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, etc.).
-The objective of being namespace compatible warrants such a naming convention, but this has a syntactical cost, particularly when \pkg{collapse} is used as the primary data manipulation framework.
+The objective of being namespace compatible warrants such a naming convention, but this has a syntactical cost, particularly when \pkg{collapse} is the primary data manipulation package.
 \newline
@@
 All \pkg{collapse} functions starting with 'f' can be passed to the option (with or without the 'f'), e.g., \code{set\_collapse(mask = c("subset", "transform"))} creates \code{subset <- fsubset} and \code{transform <- ftransform} and exports them.
-Special functions are \code{"n"} and \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} (for use in \code{(f)summarise}/\code{(f)mutate}), \code{table <- qtab}, and replace \code{\%in\%} with a fast version using \code{fmatch}, respectively.
-There are also a couple of \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask groups of functions}.
+Special functions are \code{"n"}, \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} (for use in \code{(f)summarise}/\code{(f)mutate}), \code{table <- qtab}, and replace \code{\%in\%} with a fast version using \code{fmatch}, respectively.
+There are also several \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}.
 The most powerful of these is \code{"all"}, which masks all f-functions $+$ specials, as shown below.
@@
 Similarly, using this option, all code in this article could have been written without f-prefixes.
-Thus, \pkg{collapse}, together with namespace masking, provides a fast experience of \proglang{R} - within GNU \proglang{R} - without the need to even restart the session.
+Thus, \pkg{collapse}, together with namespace masking, is able to provide a fast any syntactically clean experience of \proglang{R} - within GNU \proglang{R} - without the need to even restart the session.
 Masking is completely interactive and reversible within the active session: calling \code{set\_collapse(mask = NULL)} instantly removes the additional functions.
@@
 They are executed on an Apple M1 MacBook Pro with 16 GB unified memory.
-The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets, using a large server with many (slow) cores.
+The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets, using a large server with many (slow) cores.\footnote{A lot may be said about benchmarking \pkg{collapse}, which would be beyond the scope of this article.
+Users should note, however, that its defaults (\code{na.rm = TRUE, sort = TRUE, stable.algo = TRUE, nthreads = 1}) cater to convenience rather than maximum performance.
+For maximum performance, set these 3 settings to \code{FALSE} and increase the number of threads.
+To also provide a minimalistic guide for \proglang{R} users seeking to understand the relative performance of \pkg{collapse} and \pkg{data.table}, reflecting current (spring 2024) developments: \pkg{collapse} has highly efficient algorithms for grouping and computing statistics, but presently does not provide sub-column level parallel grouping architecture.
+Simple statistics like \code{fmean()} are parallelized across columns and perform grouped computations in a single pass.
+More complex ones \code{fmedian(), fmode()} have group-level parallelism.
+\pkg{data.table}, on the other hand, has sub-column parallel grouping and also group-level parallel implementations for simple statistics such as \code{mean()}, but no parallelism for complex statistics such as \code{median()}.
+\pkg{data.table}'s \href{https://rdatatable.gitlab.io/data.table/reference/datatable-optimize.html}{GForce optimization} also only applies to simple statistics, not complex expressions or weighted statistics - as can be vectorized using \emph{Fast Statistical Functions} in \pkg{collapse}.
+Thus, if your data is moderately sized ($\leq$100mio obs), you have more than 1 column to compute on, you want to do complex statistical things, or if your processor is very fast (high single core speed), \pkg{collapse} is a great choice.
+On the other hand, if your data is really long ($>$100mio obs), you have only a few columns to compute on, you are computing simple statistics that \pkg{data.table} optimizes, and you have massive parallel compute, then \pkg{data.table} is a great choice.
+Let me finally note that packages like \pkg{polars} use optimized memory formats based on \href{https://arrow.apache.org/}{Apache Arrow}, \href{https://pola.rs/posts/i-wrote-one-of-the-fastest-dataframe-libraries/}{multithreaded hash-based grouping, SIMD instructions and multithreading at the group-level}, and a \href{https://pola.rs/posts/polars_birds_eye_view/}{query optimizer} - all implemented in \proglang{Rust}, a thread-safe programming language.
+Nobody will ever develop something similar off \proglang{R}'s 30-year-old C API.
+But then, \proglang{R} is a statistical language not intended to work like an optimized database.
+\pkg{collapse} seamlessly integrates with \proglang{R}, while \pkg{polars}, at present, has very little to do with it.}
 %
@@
 \pkg{collapse} is my attempt at providing such a package, and the feedback I have received over the years, particularly from users in academia, government, and international organizations, is a strong indication that I have responded to a need felt in larger parts of the \proglang{R} community.
-As mentioned, a single article cannot summarize \pkg{collapse}, but there is an excellent \href{https://sebkrantz.github.io/collapse/index.html}{website} with comprehensive \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{Documentation Resources}.
+As mentioned, a single article cannot summarize \pkg{collapse}, but there is an excellent \href{https://sebkrantz.github.io/collapse/index.html}{website} with comprehensive \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}.
 

### 2024-03-04 misc/JSS article/jss-article-rnw/article.tex
Update Tex.

@@
 %% need no \usepackage{Sweave}
+%% need no \usepackage{Sweave.sty}
 
@@
 \pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency.
-It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g.
-\class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem.
+It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling seamless integration with large parts of the \proglang{R} ecosystem.
 This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples.
@@
 It also offers vectorizations for advanced operations such as weighted statistics (including mode and quantiles), functions and classes for fully indexed (time-aware) computations on time series and panel data, recursive (list-processing) tools to deal with nested data and advanced descriptive statistical tools.
-This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures and powerful data manipulation functions.
-The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference and aversion of logical vectors.
-\pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g.
-\emph{tibble}, \emph{data.table}, \emph{sf}, \emph{xts}, \emph{pdata.frame}).
+This functionality is supported by efficient algorithms for intensive operations like grouping, unique values, matching, ordering, etc., tailored to \proglang{R}'s data structures, and powerful data manipulation functions.
+The package also supplies many features for memory efficient \proglang{R} programming, such as data transformation and math by reference, and aversion of logical vectors.
+\pkg{collapse} is class-agnostic, i.e., it provides most statistical operations for atomic vectors, matrices, and data frames/lists, and seamlessly supports key variants of these objects used in the \proglang{R} ecosystem (e.g., \class{tibble}, \class{data.table}, \class{sf}, \class{xts}, \class{pdata.frame}).
 It is globally and interactively configurable, which includes setting different defaults for key function arguments (such as \code{na.rm} arguments to statistical functions, default \code{TRUE}), and modifying the package namespace itself.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse}, but can be interactively modified to overwrite key functions like \code{unique}, \code{match}, \code{\%in\%}, \code{table}, \code{subset}, \code{mutate}, \code{summarise} etc.
-with much faster \pkg{collapse} equivalents.} \newline
+with much faster \pkg{collapse} equivalents.
+See Section~\ref{sec:glob_opt}.} \newline
 
@@
 The short answer is to make computations in \proglang{R} as flexible and powerful as possible.
-The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.}; (3) increase the memory efficiency and flexibility of \proglang{R} programs\footnote{E.g.
-by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.}; and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline
+The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase performance and parsimony by avoiding \proglang{R}-level repetition;\footnote{Such as applying \proglang{R} functions across columns or split-apply-combine computing to apply functions across groups or other divisions of data.} (3) increase the memory efficiency and flexibility of \proglang{R} programs;\footnote{E.g., by avoiding object conversions and the need for certain classes to do certain things, such as converting to data frame or \class{data.table} to do something "by groups" and then convert back to matrix to continue with linear algebra, and in general to reduce the need for metaprogramming.} and (4) to create a new foundation package for statistics and data manipulation in \proglang{R} that implements successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, including some new ideas, in a stable, high performance, and broadly compatible manor.\footnote{Examples of such ideas are \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), reshaping labelled data (myself) etc...} \newline
 
@@
 
-Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars} and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy} or \pkg{StatsBase.jl} \citep{jlstatsbase}.
+Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars}, and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy}, or \pkg{StatsBase.jl} \citep{jlstatsbase}.
 % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
 In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recursive operations, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable.
-In short, it is very utile for complex statistical workflows, rich datasets (e.g.
-surveys), and for integrating with different parts of the \proglang{R} ecosystem.
-On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive).
-There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.}.
-Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function.
-However, as demonstrated below, the possibility of combining vectorized statistical functions also permits calculating more complex statistics in a vectorized way.
+In short, it is very utile for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem.
+On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive).
+There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function.
+However, as demonstrated in Section~\ref{ssec:vfat}, vectorized statistical functions can be combined to calculate more complex statistics in a vectorized way.
 \newline
@@
 \begin{Code}
-FUN(x, g = NULL, [w = NULL,] TRA = NULL, [na.rm = TRUE,]
-    use.g.names = TRUE, drop = TRUE, [nthreads = 1L,] ...)
+FUN(x, g = NULL, [w = NULL], TRA = NULL, [na.rm = TRUE],
+    use.g.names = TRUE, drop = TRUE, [nthreads = 1L], ...)
 \end{Code}
-with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e.
-simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation).
-The use of Single Instruction Multiple Data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.}.
-The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there, not here for parsimony) demonstrate their basic usage.
+with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading.\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation).
+The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there) demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects.
+As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes.
+Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods.
+Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome.
+For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix.
+%, and attributes are preserved as much as possible ()
 \begin{Code}
-fmean(mtcars$mpg)                                  # Vector
-fmean(EuStockMarkets)                              # Matrix
-fmean(mtcars)                                      # Data Frame
-fmean(mtcars$mpg, w = mtcars$wt)                   # Weighted mean
-fmean(mtcars$mpg, g = mtcars$cyl)                  # Grouped mean
-fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)   # Weighted group mean
-fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt) # Of data frame
-fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")    # Replace data by g. mean
+fmean(mtcars$mpg)
+fmean(EuStockMarkets)
+fmean(mtcars)
+fmean(mtcars$mpg, w = mtcars$wt)
+fmean(mtcars$mpg, g = mtcars$cyl)
+fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)
+fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt)
+fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")
 \end{Code}
 \subsection{Transformations}
-The \code{TRA} argument, more specifically, invokes the \fct{TRA} function for column-wise (grouped) replacing and sweeping operations (by reference). Its syntax is
+The final example invoking \code{TRA} expands the mean vector to full length, like \code{stats::ave(mtcars$mpg, mtcars$cyl)}, but much faster.
+The \code{TRA} argument invokes the \fct{TRA} function for column-wise (grouped) replacing and sweeping operations (by reference).
+Its syntax is
 \begin{Code}
@@
   2 && \code{"replace"} && replace data but preserve missing values in \code{x} \\
-  3 && \code{"-"}   && subtract (i.e. center) \\
+  3 && \code{"-"}   && subtract (i.e., center) \\
   4 && \code{"-+"}  && center on overall average statistic \\
-  5 && \code{"/"}   && divide (i.e. scale) \\
-  6 && \code{"\%"}     && compute percentages (i.e. divide and multiply by 100) \\
+  5 && \code{"/"}   && divide (i.e., scale) \\
+  6 && \code{"\%"}     && compute percentages (i.e., divide and multiply by 100) \\
   7 && \code{"+"} && add \\
   8 && \code{"*"} && multiply \\
-  9 && \code{"\%\%"} && modulus (i.e. remainder from division by  \code{STATS}) \\
-  10 && \code{"-\%\%"} && subtract modulus (i.e. make data divisible by \code{STATS})
+  9 && \code{"\%\%"} && modulus (i.e., remainder from division by  \code{STATS}) \\
+  10 && \code{"-\%\%"} && subtract modulus (i.e., make data divisible by \code{STATS})
   \end{tabular}
 }
+\caption{\label{tab:TRA} Arguments to \code{TRA()}.}
 \end{table}
@@
 Any modern laptop can calculate 1 million regression slopes in less than 1 second like this.
+Another neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse_is_sick/sick.html}{blog post}, vectorizes an expression to compute the p-value across 300k groups for a simulation study, yielding a 70x performance increase.
 \newline
 
-\pkg{collapse} also supplies advanced statistics, such as weighted medians and modes.
-The following example calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}\footnote{\pkg{collapse} calculates weighted quantiles by replacing the sample size with the sum of weights and 1 with the minimum non-zero weight in the respective quantile definition.
-See \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for more details.}, and a weighted maximum mode.\footnote{The weighted maximum mode is the largest element with the maximum sum of weights.}
+\pkg{collapse} also vectorizes advanced statistics, such as weighted medians and modes.
+The following example calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} calculates weighted quantiles by replacing the sample size with the sum of weights and 1 with the minimum non-zero weight in the respective quantile definition.
+See \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for more details.} and a weighted maximum mode.\footnote{The weighted maximum mode is the largest element with the maximum sum of weights.}
 %
@@
 %
-\pkg{collapse} provides a flexible high-performance architecture to perform (\emph{time aware}) computations on time series and panel series.
+\pkg{collapse} also provides a flexible high-performance architecture to perform (\emph{time aware}) computations on time series and panel series.
 In particular, the user enjoys great flexibility in deciding the desired degree of indexation and mode of computation.
-It is possible to apply time series and panel data transformations without any indexation by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation.
-The following list summarizes \pkg{collapse}'s time series and panel data functions.
+It is possible to apply time series and panel data transformations without any indexation by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion, or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation.
+Table~\ref{tab:TSfun} summarizes \pkg{collapse}'s time series and panel data architecture.
 %
@@
 \end{Schunk}
-The following example demonstrates a fixed-effects regression a la \citet{mundlak1978pooling}.
+The following example demonstrates a fixed-effects regression \`a la \citet{mundlak1978pooling}.
 \begin{Schunk}
@@
 \end{Schunk}
-The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. This creates an \class{indexes\_series} of life expectancy and demonstrates its properties.
+The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. This creates an \class{indexes\_series} of life expectancy and demonstrates its properties:
 \begin{Schunk}
@@
 
-In comparison with existing solutions, the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.}.
-Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures for time series operations or native handling of irregularity in basic operations.
+In comparison with existing solutions, the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions.\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.} Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures for time series operations or native handling of irregularity in basic operations.
 \class{indexed\_series} and \class{indexed\_frame}, on the other hand, work anywhere and can be superimposed on any suitable object (such as \class{sf} to create a spatiotemporal panel), as long as \pkg{collapse}'s functions (\code{flag()/L()} etc.) are used to perform the time-based computations.
-The \class{index\_df} attached to these objects can be used with other general tools such as \code{collapse::BY()} to perform grouped computations on these objects with 3rd-party functions.
+The \class{index\_df} attached to these objects can be used with other general tools such as \code{collapse::BY()} to perform grouped computations using 3rd-party functions.
 An example of calculating a 5-year rolling average is given below.
@@
 By default (\code{verbose = 1}), it prints information about the join operation and number of records joined.
-Users can request the generation of a '.join' column, akin to \proglang{STATA}'s '\_merge' column, indicating the origin of records in the joined table, as shown below.
+Users can request the generation of a \code{.join} column (\code{column = TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, indicating the origin of records in the joined table.
 %
@@
 First, \code{collapse::join} is also class-agnostic and preserves the attributes of \code{x}.
-It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"} or \code{"anti"}) and default to \code{"left"}, so the default behavior simply adds columns to \code{x}.
+It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"}, or \code{"anti"}) and defaults to \code{"left"}, so the default behavior simply adds columns to \code{x}.
 By default (\code{sort = FALSE}), the order of rows in \code{x} is also preserved.
 Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join (using \fct{fmatch}) followed by sorting, particularly if the data is already sorted on the keys.
-} Additionally, by default (\code{multiple = FALSE}), only the first matches from both tables are joined to avoid silent cartesian duplication of records.
+} Additionally, by default (\code{multiple = FALSE}), only the first matches from \code{y} are joined to avoid silent cartesian duplication of records.
 In multi-match settings, this will be reflected by few records from \code{y} being used.
@@
 R> df2$name = df1$name
-R> join(df1, df2) |> capture.output(type="m") |> strwrap(77) |> cat(sep="\n")
+R> join(df1, df2) |> capture.output(type = "m") |>
++    strwrap(77) |> cat(sep = "\n")
 \end{Sinput}
@@
 The reshaping/pivoting functionality of commercial and open source software is also unsatisfying for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data.
-Such datasets can only be reshaped by losing these labels or additional manual efforts to retain them.
-Modern \proglang{R} packages also offer different functions for different reshaping operations, such as \code{data.table::melt/ tidyr::pivot_longer} to combine columns and \code{data.table::dcast/tidyr::pivot_wider} to expand them, requiring users to learn both.
-Finally, since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is no modern replacement for \code{reshape2::recast()}.
-This requires \proglang{R} users to consecutively call two reshaping functions, incurring a high cost in terms of syntax and memory for such frequently required data transpositions.
+Such datasets can presently only be reshaped by losing these labels or additional manual efforts to retain them.
+Modern \proglang{R} packages also offer different functions for different reshaping operations, such as \code{data.table::melt/tidyr::pivot_longer} to combine columns and \code{data.table::dcast/tidyr::pivot_wider} to expand them, requiring users to learn both.
+Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two reshaping functions and incurring a high cost in terms of syntax and memory.
 \newline
 
-\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, can handle complex labelled data without loss of information, and offers best-in-\proglang{R} computational performance and memory efficiency.
+\code{collapse::pivot} provides a modern class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots, and supports complex labelled data without loss of information.
 Its syntax is:
@@
 %
-Below, its use is demonstrated with a generated dataset about fruits.
-We could also think about a survey with households and individuals, or sectors and firms.
+The demonstration below employs a generated dataset about fruits.
+We could equivalently think about a survey with households and individuals, or sectors and firms.
 Variable labels are stored in \code{attr(column, "label")}.
-The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples with real data.
+The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation} provides more elaborate examples.
 %
@@
   type type_name id       r      h
-1    A    Apples  1 0.9322 0.4315
-2    A    Apples  2 0.2230 0.7225
-3    B   Bananas  1 0.8147 0.4918
-4    B   Bananas  2 0.5642 2.4131
+1    A    Apples  1 1.42361 1.7860
+2    A    Apples  2 1.30998 0.0831
+3    B   Bananas  1 0.04423 1.6729
+4    B   Bananas  2 0.08639 1.2877
 \end{Soutput}
@@
   type type_name id variable        label   value
-1    A    Apples  1        r Fruit Radius 0.9322
-2    A    Apples  2        r Fruit Radius 0.2230
-3    B   Bananas  1        r Fruit Radius 0.8147
-4    B   Bananas  2        r Fruit Radius 0.5642
-5    A    Apples  1        h Fruit Height 0.4315
-6    A    Apples  2        h Fruit Height 0.7225
-7    B   Bananas  1        h Fruit Height 0.4918
-8    B   Bananas  2        h Fruit Height 2.4131
+1    A    Apples  1        r Fruit Radius 1.42361
+2    A    Apples  2        r Fruit Radius 1.30998
+3    B   Bananas  1        r Fruit Radius 0.04423
+4    B   Bananas  2        r Fruit Radius 0.08639
+5    A    Apples  1        h Fruit Height 1.78599
+6    A    Apples  2        h Fruit Height 0.08310
+7    B   Bananas  1        h Fruit Height 1.67294
+8    B   Bananas  2        h Fruit Height 1.28770
 \end{Soutput}
@@
   id   r_A     r_B    h_A   h_B
-1  1 0.9322 0.8147 0.4315 0.4918
-2  2 0.2230 0.5642 0.7225 2.4131
+1  1 1.424 0.04423 1.7860 1.673
+2  2 1.310 0.08639 0.0831 1.288
 \end{Soutput}
@@
   id variable        label      A       B
-1  1        r Fruit Radius 0.9322 0.8147
-2  2        r Fruit Radius 0.2230 0.5642
-3  1        h Fruit Height 0.4315 0.4918
-4  2        h Fruit Height 0.7225 2.4131
+1  1        r Fruit Radius 1.4236 0.04423
+2  2        r Fruit Radius 1.3100 0.08639
+3  1        h Fruit Height 1.7860 1.67294
+4  2        h Fruit Height 0.0831 1.28770
 \end{Soutput}
@@
 \begin{Sinput}
-R> library(magrittr) # World after 2015 (latest country data)
+R> library("magrittr")
 R> wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
@@
 \end{Schunk}
-This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e. \fct{fmean} is only invoked once).
+This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only called once).
 \begin{Schunk}
@@
 %
-This section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks.
+This Section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks.
 They are executed on an Apple M1 MacBook Pro with 16 GB unified memory.
-The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets, using a large server with many (slow) cores.
+The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare more software packages on larger datasets, using a large server with many (slow) cores.\footnote{A lot may be said about benchmarking \pkg{collapse}, which would be beyond the scope of this article.
+Users should note, however, that its defaults (\code{na.rm = TRUE, sort = TRUE, stable.algo = TRUE, nthreads = 1}) cater to convenience rather than maximum performance.
+For maximum performance, set these 3 settings to \code{FALSE} and increase the number of threads.
+To also provide a minimalistic guide for \proglang{R} users seeking to understand the relative performance of \pkg{collapse} and \pkg{data.table}, reflecting current (spring 2024) developments: \pkg{collapse} has highly efficient algorithms for grouping and computing statistics, but presently does not provide sub-column level parallel grouping architecture.
+Simple statistics like \code{fmean()} are parallelized across columns and perform grouped computations in a single pass.
+More complex ones \code{fmedian(), fmode()} have group-level parallelism.
+\pkg{data.table}, on the other hand, has sub-column parallel grouping and also group-level parallel implementations for simple statistics such as \code{mean()}, but no parallelism for complex statistics such as \code{median()}.
+\pkg{data.table}'s \href{https://rdatatable.gitlab.io/data.table/reference/datatable-optimize.html}{GForce optimization} also only applies to simple statistics, not complex expressions or weighted statistics - as can be vectorized using \emph{Fast Statistical Functions} in \pkg{collapse}.
+Thus, if your data is moderately sized ($\leq$100mio obs), you have more than 1 column to compute on, you want to do complex statistical things, or if your processor is very fast (high single core speed), \pkg{collapse} is a great choice.
+On the other hand, if your data is really long ($>$100mio obs), you have only a few columns to compute on; you are computing simple statistics that \pkg{data.table} optimizes, and you have massive parallel compute, then \pkg{data.table} is a great choice.
+Let me finally note that packages like \pkg{polars} use optimized memory formats based on \href{https://arrow.apache.org/}{Apache Arrow}, \href{https://pola.rs/posts/i-wrote-one-of-the-fastest-dataframe-libraries/}{multithreaded hash-based grouping, SIMD instructions and multithreading at the group-level}, and a \href{https://pola.rs/posts/polars_birds_eye_view/}{query optimizer} - all implemented in \proglang{Rust}, a thread-safe programming language.
+Nobody will ever develop something similar off \proglang{R}'s 30-year-old C API.
+But then, \proglang{R} is a statistical language not intended to work like an optimized database.
+\pkg{collapse} seamlessly integrates with \proglang{R}, while \pkg{polars}, at present, has very little to do with it.}
 %
@@
      expr   min    lq   mean median    uq    max neval
-        R 9.632 9.837 9.959  9.852 9.940 12.70   100
-    Rfast 1.315 1.805 2.257  1.840 1.917 11.92   100
- collapse 1.321 1.432 1.847  1.506 1.603 16.33   100
+        R 9.653 9.837 10.127  9.850 9.896 21.768   100
+    Rfast 1.307 1.789  1.914  1.835 1.893  6.005   100
+ collapse 1.322 1.431  1.566  1.474 1.577  4.987   100
 \end{Soutput}
@@
        expr   min     lq   mean median     uq    max neval
-          R 10.715 11.19 11.550 11.295 11.61 16.89   100
- data.table  9.154 18.70 19.743 19.229 20.19 30.22   100
-   collapse  2.330  2.89  3.619  3.078  3.71 15.94   100
+          R 10.75 11.213 11.706 11.281 11.472 16.423   100
+ data.table 11.45 18.526 19.976 19.278 20.465 57.433   100
+   collapse  1.92  2.713  3.303  2.959  3.552  6.251   100
 \end{Soutput}
@@
        expr    min     lq   mean median     uq   max neval
- data.table 136.36 137.50 140.23 138.54 140.62 173.7   100
-   collapse  68.42  76.85  82.25  80.28  83.36 163.3   100
+ data.table 136.21 137.08 138.09 137.64 138.34 143.8   100
+   collapse  65.79  75.85  78.94  79.23  81.74 116.9   100
 \end{Soutput}
@@
        expr   min     lq  mean median     uq   max neval
- data.table 8.915 10.881 16.398 13.166 16.903 73.045   100
-   collapse 1.228  1.379  1.452  1.413  1.459  2.337   100
+ data.table 8.438 11.882 24.01 13.756 18.897 62.78   100
+   collapse 1.221  1.382  1.48  1.408  1.466  2.94   100
 \end{Soutput}
@@
        expr   min    lq  mean median    uq   max neval
- data.table 11.79 12.84 17.99  15.87 17.68 73.75   100
-   collapse 11.73 13.77 20.37  15.69 18.17 76.75   100
+ data.table 11.32 14.70 21.36  16.00 17.63 58.90   100
+   collapse 11.52 14.33 21.07  15.66 16.89 61.89   100
 \end{Soutput}
@@
        expr   min     lq   mean median     uq   max neval
- data.table 60.01 69.99 90.12  74.28 111.73 157.7   100
-   collapse 32.59 41.80 63.15  48.87  63.98 177.0   100
+ data.table 74.29 115.48 118.73 118.87 122.64 135.1   100
+   collapse 55.69  87.16  91.14  90.35  94.12 118.8   100
 \end{Soutput}
@@
           expr    min     lq  mean median    uq    max neval
-      base_int 60.07 61.852  67.56 65.613  68.55 133.05   100
-  collapse_int  6.69  9.222  10.34  9.581  10.18  17.56   100
-     base_char 91.79 94.570 102.38 99.497 103.65 170.41   100
- collapse_char 22.29 23.879  26.83 24.577  27.58  80.46   100
+      base_int 60.216 63.192 64.81 64.170 65.81 108.20   100
+  collapse_int  8.624  9.192 10.44  9.384 10.35  18.11   100
+     base_char 92.110 94.566 98.56 97.757 99.91 160.08   100
+ collapse_char 20.464 23.445 25.37 23.744 24.71 113.24   100
 \end{Soutput}
@@
             expr    min     lq   mean median     uq   max neval
-        base_int 26.873 26.993 28.261 27.094 27.306 33.25    10
-    collapse_int  8.603  8.633  9.352  8.662  8.723 15.53    10
-       base_char 84.091 84.726 86.127 85.100 86.008 91.23    10
- data.table_char 40.970 41.169 42.597 41.307 41.407 48.49    10
-   collapse_char 30.482 30.541 30.774 30.622 30.803 31.47    10
+        base_int 26.877 27.027 28.455 27.132 30.216 31.18    10
+    collapse_int  8.637  8.701  9.269  8.722  8.973 13.14    10
+       base_char 86.277 86.299 90.211 87.652 92.643 98.55    10
+ data.table_char 41.150 41.615 42.225 41.806 41.906 47.05    10
+   collapse_char 32.333 32.548 33.099 32.620 32.819 37.39    10
 \end{Soutput}

### 2024-03-06 misc/JSS article/jss-article-rnw/article.Rnw

@@
 \pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency.
-It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling seamless integration with large parts of the \proglang{R} ecosystem.
+It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem.
 This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples.
@@
 In summary, \pkg{collapse} is a foundation package for statistical computing and data manipulation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient.
-A significant benefit is that, rather than piecing together a somewhat fragmented ecosystem oriented at different classes and tasks, core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages.
+A significant benefit is that, rather than piecing together a fragmented ecosystem oriented at different classes and tasks, many core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages.
 This tends to result in \proglang{R} scripts that are shorter, more efficient, and more lightweight in dependencies.
@@
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, the short vignette summarizing available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} is an excellent starting point.
+For deeper engagement with \pkg{collapse}, a short vignette summarizing available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} is an excellent starting point.
 %
@@
 with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - weights, \code{TRA} - transformation, \code{na.rm} - missing values, \code{use.g.names} -  attach group names upon aggregation (if \code{g} is used), \code{drop} - drop dimensions (i.e., simplify to atomic vector if \code{is.null(g)} and \code{x} is matrix or data frame-like), \code{nthreads} - multithreading.\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions (detailed in the documentation).
-The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette (and executed there) demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects.
+The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects.
 As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes.
@@
 %, and attributes are preserved as much as possible ()
-\begin{Code}
+<<faststatfun>>=
 fmean(mtcars$mpg)
 fmean(EuStockMarkets)
-fmean(mtcars)
+fmean(mtcars[5:10])
 fmean(mtcars$mpg, w = mtcars$wt)
@@
 fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt)
-fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")
-\end{Code}
+fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill") |> head(20)
+@
+%
 \subsection{Transformations}
@@
 }
-\caption{\label{tab:TRA} Arguments to \code{TRA()}.}
+\caption{\label{tab:TRA} Available \code{FUN} choices in \code{TRA()}.}
 \end{table}
@@
 The \code{set} argument can also be passed to \emph{Fast Statistical Functions} to toggle transformation by reference.
-The following examples demonstrate how this design allows flexible ad-hoc transformations using \proglang{R}'s built-in \code{airquality} dataset with daily measurements in New York from May to September 1973.
+The following examples demonstrate how this design allows flexible ad-hoc transformations using \proglang{R}'s built-in \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/airquality}{\code{airquality} dataset} with daily measurements in New York from May to September 1973.
 %
@@
 @
-This imputes \code{Ozone} and \code{Solar.R} by reference using the month median.
+This imputes columns \code{Ozone} and \code{Solar.R} by reference using the month median.
 <<aq_imp>>=
@@
 <<aq_trans>>=
-airquality |> fmutate(
+airquality |> fmutate(ozone_deg = Ozone / Temp,
   rad_day = fsum(as.double(Solar.R), Day, TRA = "/"),
-  ozone_deg = Ozone / Temp,
   ozone_amed = Ozone > fmedian(Ozone, Month, TRA = "fill"),
@@
 These objects can be created with \code{GRP()}.
-The syntax is
+Its syntax is
 \begin{Code}
@@
 \end{Code}
- The example below creates and displays a \class{GRP} object from 3 columns in \code{mtcars}. The \code{by} argument also supports column names or indices, and \code{X} could also be an atomic vector.
+ The example below creates and displays a \class{GRP} object from 3 columns in \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/mtcars}{\code{mtcars}}.
+The \code{by} argument also supports column names or indices, and \code{X} could also be an atomic vector.
 %
@@
 <<gcomp>>=
-dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt
+dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt;
 add_vars(g$groups,
@@
 %
-Similar, data can be transformed, here using the S3 generic \fct{fscale} function.
+Similarly, data can be transformed, here using the S3 generic \fct{fscale} function.
 %
@@
 For example, the useR 2022 presentation \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{Slides 18-19} aggregates the \href{https://worldmrio.com/}{EORA Global Supply Chain Database} from the country to the world region level.
-After defining a single grouping object, a list of value-added shares matrices (VB) and outputs for years 1990-2021, is aggregated with no grouping cost using a single line of code like \code{lapply(VB_list, function(x) x$VB |> fsum(g) |> t() |> fmean(g, x$O) |> t())}.
-On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another example from my recent work involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}.
+After defining a single grouping object, a list of value-added shares matrices (VB) and outputs (O) for years 1990-2021, is aggregated with no grouping cost using a single line of code like \code{lapply(VB_list, function(x) x$VB |> fsum(g) |> t() |> fmean(g, x$O) |> t())}.
+On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another recent example involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}.
 Thus each iteration of the numerical routine raises the vector \textbf{x} to a different power ($a$), sums it in 1 million groups ($j$) to generate \textbf{y}, and computes the Euclidean distance to \textbf{z} (using \code{collapse::fdist}).
@@
 These are integrated with the \emph{Fast Statistical Functions} to enable vectorized statistical operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow.
-For example, the following code calculates
+For example, the following code calculates the mean of columns
 %
 <<fsummarise>>=
-mtcars |>
-  fsubset(mpg > 11) |>
-  fgroup_by(cyl, vs, am) |>
+mtcars |> fsubset(mpg > 11) |> fgroup_by(cyl, vs, am) |>
   fsummarise(across(c(mpg, carb, hp), fmean),
@@
 %
-the mean of columns \code{mpg}, \code{carb} and \code{hp}, and the weighted mean of \code{qsec}, after subsetting and grouping the data.
+ \code{mpg}, \code{carb} and \code{hp}, and the weighted mean of \code{qsec}, after subsetting and grouping the data.
 This code is very fast (especially with many groups) because data does not need to be split by groups at all.
@@
 Thus \code{w} becomes the second positional argument...} The \emph{Fast Statistical Functions} also have a method for grouped data, so \code{fsummarise} is not always needed.
-The following code calculates weighted group means.
-By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::fgroup\_by}.
-The latter requires an additional C routine to convert the grouping object to a \class{GRP} object, and is thus much less efficient.}
+The following example calculates weighted group means.
+By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::group\_by}.
+The latter requires an additional C routine to convert the \pkg{dplyr} grouping object to a \class{GRP} object, and is thus less efficient.}
 %
 <<fsummarise_2>>=
-mtcars |>
-  fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
-  fgroup_by(cyl, vs, am) |>
-  fmean(wt) |> head(2)
+mtcars |> fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
+  fgroup_by(cyl, vs, am) |> fmean(wt) |> head(2)
 @
@@
 %
-\fct{fsummarise} and \fct{fmutate} can also evaluate other statistical functions in the classical way (split-apply-combine) and handle more complex expressions involving multiple columns and/or functions.
+\fct{fsummarise} and \fct{fmutate} can also evaluate arbitrary statistical functions in the classical way (split-apply-combine) and handle more complex expressions involving multiple columns and/or functions.
 However, using any \emph{Fast Statistical Function} causes the whole expression to be vectorized, i.e., evaluated only once and not for every group.
-This eager vectorization approach is helpful for efficient grouped calculation of more complex statistics.
+This eager vectorization approach enables efficient grouped calculation of more complex statistics.
 The example below calculates grouped (\code{vs}) bivariate regression slopes (\code{mpg ~ carb}) in a vectorized way.
@@
 <<flinreg>>=
-mtcars |>
- fgroup_by(vs) |>
+mtcars |> fgroup_by(vs) |>
  fmutate(dm_carb = fmean(carb, TRA = "-")) |>
@@
 Any modern laptop can calculate 1 million regression slopes in less than 1 second like this.
-Another neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse_is_sick/sick.html}{blog post}, vectorizes an expression to compute the p-value across 300k groups for a simulation study, yielding a 70x performance increase.
+Another very neat example, shared by Andrew Ghazi in a recent \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post},\footnote{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html} vectorizes an expression to compute the $p$~value across 300k groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}.
 \newline
@@
 <<wstats>>=
-mtcars |>
-    fgroup_by(cyl, vs, am) |>
+mtcars |> fgroup_by(cyl, vs, am) |>
   fmutate(o = radixorder(GRPid(), mpg)) |>
@@
 %
-Both weighted mode and quantiles have a sub-column parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \code{fnth/fmedian/fmode} (default 1).} and, as in this case, can also harness an (optional) optimization by computing an overall ordering vector and passing it to the functions to avoids repeated sorting (using quickselect) of the same elements within each group.
+Both weighted mode and quantiles have a sub-column parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \code{fnth/fmedian/fmode} (default 1).} and, as shown above, can also harness an (optional) optimization by computing an overall ordering vector and passing it to each quantile function to avoid repeated partial sorting (using quickselect) of the same elements within groups.
 For advanced data aggregation, \pkg{collapse} also provides a convenience function, \fct{collap}, which (by default) uses \code{fmean} for numeric and \code{fmode} for non-numeric columns.
-Below, it aggregates GDP per Capita, Life Expectancy, and country name by World Bank Income Group, with population weights.\footnote{\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}} is a dataset supplied by \pkg{collapse}, extracted from the World Bank World Development Indicators.} This yields population-weighted statistics, the largest country, and each income group's total population (sum of weights) for each year, preserving (default \code{keep.col.order = TRUE}) the order of columns.
+Below, it aggregates GDP per capita, life expectancy, and country name by World Bank income group, with population weights.\footnote{\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}} is a dataset supplied by \pkg{collapse}, extracted from the World Bank World Development Indicators.} This yields population-weighted statistics, the largest country, and each income group's total population (sum of weights) for each year, preserving (default \code{keep.col.order = TRUE}) the order of columns.
 %
@@
 %
-\pkg{collapse} also provides a flexible high-performance architecture to perform (\emph{time aware}) computations on time series and panel series.
+\pkg{collapse} also provides a flexible high-performance architecture to perform (time aware) computations on time series and panel series.
 In particular, the user enjoys great flexibility in deciding the desired degree of indexation and mode of computation.
@@
 \emph{Classes, constructors and utilities} \\
-\code{findex\_by(), findex(), unindex(), reindex(), timeid(), is\_irregular(), to\_plm()} $+$ rich set of S3 methods for 'indexed\_frame', 'indexed\_series' and 'index\_df'. \\\\
+\code{findex\_by(), findex(), unindex(), reindex(), timeid(), is\_irregular(), to\_plm()} $+$ S3 methods for \class{indexed\_frame}, \class{indexed\_series} and \class{index\_df} \\\\
 \emph{Core time-based functions} \\
-\code{flag(), fdiff(), fgrowth(), fcumsum(), psmat()} [panel data to array conversions] \code{psacf(), pspacf(), psccf()} [autocorrelation functions for panel data] \\\\
+\code{flag(), fdiff(), fgrowth(), fcumsum(), psmat()} \\ \code{psacf(), pspacf(), psccf()} \\\\
 \emph{Data transformation functions with supporting methods} \\
-\code{f[hd]between(), f[hd]within(), fscale()} [scaling and (higher-dimensional) centering] \\\\
+\code{fscale(), f[hd]between(), f[hd]within()} \\\\
 \emph{Data manipulation functions with supporting methods} \\
-\code{fsubset(), funique(), roworder[v]()} [internal], and \code{na\_omit()} [internal] \\\\
-\emph{Summary Functions with Supporting Methods} \\
-\code{qsu(), varying()} [panel-variance decomposed statistics] \\
+\code{fsubset(), funique(), roworder[v]()} (internal), \code{na\_omit()} (internal) \\\\
+\emph{Summary functions with supporting methods} \\
+\code{varying(), qsu()} \\
 \end{tabular}
@@
 In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation.
-If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years.} If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id.
+If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years.
+Users need to manually call \code{timeid()} on plain numeric vectors with decimals to yield an appropriate integer representation.} If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id.
 For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}.
@@
 \begin{Code}
-data_ix = findex_by(data, id1, ..., time)
-data_ix$indexed_series
-with(data, indexed_series)
-index_df = findex(data_ix)
+data_ix <- findex_by(data, id1, ..., time)
+data_ix$indexed_series; with(data, indexed_series)
+index_df <- findex(data_ix)
 \end{Code}
@@
 \begin{Code}
-data = unindex(data_ix)
-data_ix = reindex(data, index = index_df)
-indexed_series = reindex(vec/mat, index = vec/index_df)
+data <- unindex(data_ix)
+data_ix <- reindex(data, index = index_df)
+indexed_series <- reindex(vec/mat, index = vec/index_df)
 \end{Code}
@@
 <<indexing_2>>=
-LIFEEXi = wldi$LIFEEX
-str(LIFEEXi, width = 70, strict = "cut")
+LIFEEXi <- wldi$LIFEEX; str(LIFEEXi, width = 70, strict = "cut")
 c(is_irregular(LIFEEXi), is_irregular(LIFEEXi[-5]))
@@
 By default (\code{verbose = 1}), it prints information about the join operation and number of records joined.
-Users can request the generation of a \code{.join} column (\code{column = TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, indicating the origin of records in the joined table.
+Users can request the generation of a \code{.join} column (\code{column = "name"/TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, indicating the origin of records in the joined table.
 %
 <<joins>>=
-df1 <- data.frame(id1 = c(1, 1, 2, 3),
-                  id2 = c("a", "b", "b", "c"),
-                  name = c("John", "Jane", "Bob", "Carl"),
-                  age = c(35, 28, 42, 50))
-df2 <- data.frame(id1 = c(1, 2, 3, 3),
-                  id2 = c("a", "b", "c", "e"),
+df1 <- data.frame(id1 = c(1, 1, 2, 3), id2 = c("a", "b", "b", "c"),
+  name = c("John", "Jane", "Bob", "Carl"), age = c(35, 28, 42, 50))
+df2 <- data.frame(id1 = c(1, 2, 3, 3), id2 = c("a", "b", "c", "e"),
   salary = c(60000, 55000, 70000, 80000),
   dept = c("IT", "Marketing", "Sales", "IT"))
-
 join(df1, df2, on = c("id1", "id2"), how = "full", column = TRUE)
@@
 %
-An alternative to the join column is to request an attribute that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}).
+An alternative to the join column is to request an attribute (\code{attr = "name"/TRUE}) that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}).
 Users can also invoke the \code{validate} argument to check the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.
@@
 In multi-match settings, this will be reflected by few records from \code{y} being used.
-\fct{fmatch} also has a built-in overidentification check, which issues a warning if more key columns than necessary to identify the records are provided.
+\fct{fmatch} also has a built-in overidentification check, which issues a warning if more key columns than necessary to identify the records are provided:
 %
@@
 % This warning can be silenced by passing \code{overid = 2} or restricting the join columns using \code{on}.
-A final noteworthy feature is the handling of duplicate non-id columns in both tables.
+A final noteworthy feature is the handling of duplicate non-id columns in both tables:
 %
@@
 %
-The reshaping/pivoting functionality of commercial and open source software is also unsatisfying for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data.
+The reshaping/pivoting functionality of both commercial and open source software is also unsatisfying for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data.
 Such datasets can presently only be reshaped by losing these labels or additional manual efforts to retain them.
 Modern \proglang{R} packages also offer different functions for different reshaping operations, such as \code{data.table::melt/tidyr::pivot_longer} to combine columns and \code{data.table::dcast/tidyr::pivot_wider} to expand them, requiring users to learn both.
-Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two reshaping functions and incurring a high cost in terms of syntax and memory.
+Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two reshaping functions, incurring a high cost in terms of syntax and memory.
 \newline
@@
 The second option is a wider pivot with \code{how = "wider"}.
-Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} \emph{Note} how the labels are combined with existing labels such that also this operation is without loss of information.
+Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{multiple columns with names and labels could be selected, which would be combined using \code{"\_"} for names and \code{" - "} for labels.} Note how the labels are combined with existing labels such that also this operation is without loss of information.
 It is, however, a destructive operation, i.e., with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it.
@@
 <<list_proc_2>>=
-nest_lm_coef <- dl |>
-  rapply2d(lm, formula = mpg ~ .) |>
-  rapply2d(summary, classes = "lm") |>
-  get_elem("coefficients")
-
+nest_lm_coef <- dl |> rapply2d(lm, formula = mpg ~ .) |>
+  rapply2d(summary, classes = "lm") |> get_elem("coefficients")
 nest_lm_coef |> str(give.attr = FALSE, strict = "cut")
@@
 A related exercise is to decompose the variance of a panel series into a component due to variation between countries and one capturing variance within countries over time.
-Using the \code{W()/B()} operators and the \code{LIFEEXi} \class{indexed\_series} from Section~\ref{sec:ts_ps}, this is easily demonstrated
+Using the \code{W()/B()} operators and the \code{LIFEEXi} \class{indexed\_series} from Section~\ref{sec:ts_ps}, this is easily demonstrated:
 %
@@
 set_collapse(mask = "all", na.rm = FALSE, sort = FALSE, nthreads = 4)
-wlddev |>
-  subset(year >= 1990 & is.finite(GINI)) |>
+wlddev |> subset(year >= 1990 & is.finite(GINI)) |>
   group_by(year) |>
@@
 range(wlddev$date)
-wlddev |>
-  index_by(iso3c, year) |>
+wlddev |> index_by(iso3c, year) |>
   mutate(PCGDP_lag = lag(PCGDP),
@@
 Similarly, using this option, all code in this article could have been written without f-prefixes.
-Thus, \pkg{collapse}, together with namespace masking, is able to provide a fast any syntactically clean experience of \proglang{R} - within GNU \proglang{R} - without the need to even restart the session.
+Thus, \pkg{collapse}, together with namespace masking, is able to provide a fast and syntactically clean experience of \proglang{R} - without the need to even restart the session.
 Masking is completely interactive and reversible within the active session: calling \code{set\_collapse(mask = NULL)} instantly removes the additional functions.
@@
 %
-This Section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks.
+This section offers a small benchmark to demonstrate that \pkg{collapse} provides best-in-\proglang{R} performance for many basic statistical and data manipulation tasks.
 They are executed on an Apple M1 MacBook Pro with 16 GB unified memory.
@@
 \pkg{data.table}'s \href{https://rdatatable.gitlab.io/data.table/reference/datatable-optimize.html}{GForce optimization} also only applies to simple statistics, not complex expressions or weighted statistics - as can be vectorized using \emph{Fast Statistical Functions} in \pkg{collapse}.
-Thus, if your data is moderately sized ($\leq$100mio obs), you have more than 1 column to compute on, you want to do complex statistical things, or if your processor is very fast (high single core speed), \pkg{collapse} is a great choice.
-On the other hand, if your data is really long ($>$100mio obs), you have only a few columns to compute on, you are computing simple statistics that \pkg{data.table} optimizes, and you have massive parallel compute, then \pkg{data.table} is a great choice.
-Let me finally note that packages like \pkg{polars} use optimized memory formats based on \href{https://arrow.apache.org/}{Apache Arrow}, \href{https://pola.rs/posts/i-wrote-one-of-the-fastest-dataframe-libraries/}{multithreaded hash-based grouping, SIMD instructions and multithreading at the group-level}, and a \href{https://pola.rs/posts/polars_birds_eye_view/}{query optimizer} - all implemented in \proglang{Rust}, a thread-safe programming language.
-Nobody will ever develop something similar off \proglang{R}'s 30-year-old C API.
-But then, \proglang{R} is a statistical language not intended to work like an optimized database.
-\pkg{collapse} seamlessly integrates with \proglang{R}, while \pkg{polars}, at present, has very little to do with it.}
+Thus, if your data is moderately sized ($\leq$100mio.
+obs.), you have more than 1 column to compute on, you want to do complex statistical things, or if your processor is very fast (high single core speed), \pkg{collapse} is a great choice.
+On the other hand, if your data is really long ($>$100mio.
+obs.), you have only a few columns to compute on, you are computing simple statistics that \pkg{data.table} optimizes, and you have massive parallel compute, then \pkg{data.table} is a great choice.
+My recommendation: use both, just need to call \code{library(fastverse)}.
+Finally, let me note that \pkg{polars} uses optimized memory buffers based on \href{https://arrow.apache.org/}{Apache Arrow}, \href{https://pola.rs/posts/i-wrote-one-of-the-fastest-dataframe-libraries/}{multithreaded hash-based grouping, SIMD instructions and multithreading at the group-level}, and a \href{https://pola.rs/posts/polars_birds_eye_view/}{query optimizer} - all implemented in \proglang{Rust}, a thread-safe programming language.
+While some of these parallel algorithms could be ported to \pkg{collapse}, this is more challenging since C, and particularly R's C API, is not thread safe - and it would still be lacking the benefits of Arrow memory buffers.
+At core, \proglang{R} is a 30-year old statistical language and not intended to work like an optimized database.
+\pkg{collapse} seamlessly integrates with \proglang{R}'s data structures; \pkg{polars}, at present, has nothing to do with them (and is therefore also not part of this benchmark).}
 %
@@
 It is coming close to 4 years since the first CRAN release of \pkg{collapse} in March 2020, and since then, the package has grown and matured considerably.
-At the time of writing this article, it has been downloaded $>$1.5 million times off CRAN and is known and used by thousands of \proglang{R} users.
+At the time of writing this article in early 2024, it has been downloaded $>$1.5 million times off CRAN.
 In this article, I have articulated key ideas and design principles and demonstrated some core features of the package.
@@
 \pkg{collapse} is my attempt at providing such a package, and the feedback I have received over the years, particularly from users in academia, government, and international organizations, is a strong indication that I have responded to a need felt in larger parts of the \proglang{R} community.
-As mentioned, a single article cannot summarize \pkg{collapse}, but there is an excellent \href{https://sebkrantz.github.io/collapse/index.html}{website} with comprehensive \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}.
+As mentioned, a single article cannot comprehensively introduce \pkg{collapse}, but there is a modern \href{https://sebkrantz.github.io/collapse/index.html}{website} with comprehensive \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}.
 

### 2024-03-06 misc/JSS article/jss-article-rnw/article.tex

@@
 %, and attributes are preserved as much as possible ()
-\begin{Code}
-fmean(mtcars$mpg)
-fmean(EuStockMarkets)
-fmean(mtcars)
-fmean(mtcars$mpg, w = mtcars$wt)
-fmean(mtcars$mpg, g = mtcars$cyl)
-fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)
-fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt)
-fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill")
-\end{Code}
+\begin{Schunk}
+\begin{Sinput}
+R> fmean(mtcars$mpg)
+\end{Sinput}
+\begin{Soutput}
+[1] 20.09
+\end{Soutput}
+\begin{Sinput}
+R> fmean(EuStockMarkets)
+\end{Sinput}
+\begin{Soutput}
+ DAX  SMI  CAC FTSE 
+2531 3376 2228 3566 
+\end{Soutput}
+\begin{Sinput}
+R> fmean(mtcars[5:10])
+\end{Sinput}
+\begin{Soutput}
+   drat      wt    qsec      vs      am    gear 
+ 3.5966  3.2173 17.8488  0.4375  0.4062  3.6875 
+\end{Soutput}
+\begin{Sinput}
+R> fmean(mtcars$mpg, w = mtcars$wt)
+\end{Sinput}
+\begin{Soutput}
+[1] 18.55
+\end{Soutput}
+\begin{Sinput}
+R> fmean(mtcars$mpg, g = mtcars$cyl)
+\end{Sinput}
+\begin{Soutput}
+    4     6     8 
+26.66 19.74 15.10 
+\end{Soutput}
+\begin{Sinput}
+R> fmean(mtcars$mpg, g = mtcars$cyl, w = mtcars$wt)
+\end{Sinput}
+\begin{Soutput}
+    4     6     8 
+25.94 19.65 14.81 
+\end{Soutput}
+\begin{Sinput}
+R> fmean(mtcars[5:10], g = mtcars$cyl, w = mtcars$wt)
+\end{Sinput}
+\begin{Soutput}
+   drat    wt  qsec     vs     am  gear
+4 4.031 2.415 19.38 0.9149 0.6498 4.047
+6 3.569 3.152 18.12 0.6212 0.3788 3.821
+8 3.206 4.133 16.89 0.0000 0.1204 3.241
+\end{Soutput}
+\begin{Sinput}
+R> fmean(mtcars$mpg, g = mtcars$cyl, TRA = "fill") |> head(20)
+\end{Sinput}
+\begin{Soutput}
+ [1] 19.74 19.74 26.66 19.74 15.10 19.74 15.10 26.66 26.66 19.74 19.74 15.10
+[13] 15.10 15.10 15.10 15.10 15.10 26.66 26.66 26.66
+\end{Soutput}
+\end{Schunk}
+%
 \subsection{Transformations}
@@
 \end{Schunk}
-This imputes \code{Ozone} and \code{Solar.R} by reference using the month median.
+This imputes columns \code{Ozone} and \code{Solar.R} by reference using the month median.
 \begin{Schunk}
@@
 \begin{Sinput}
-R> airquality |> fmutate(
+R> airquality |> fmutate(ozone_deg = Ozone / Temp,
 +    rad_day = fsum(as.double(Solar.R), Day, TRA = "/"),
-+    ozone_deg = Ozone / Temp,
 +    ozone_amed = Ozone > fmedian(Ozone, Month, TRA = "fill"),
@@
 \begin{Soutput}
-  Ozone Solar.R Wind Temp Month Day rad_day ozone_deg ozone_amed ozone_resid
-1    41     190  7.4   67     5   1   0.191    0.6119       TRUE     -10.279
-2    36     118  8.0   72     5   2   0.135    0.5000       TRUE     -15.279
-3    12     149 12.6   74     5   3   0.168    0.1622      FALSE      -3.035
+  Ozone Solar.R Wind Temp Month Day ozone_deg rad_day ozone_amed ozone_resid
+1    41     190  7.4   67     5   1    0.6119   0.191       TRUE     -10.279
+2    36     118  8.0   72     5   2    0.5000   0.135       TRUE     -15.279
+3    12     149 12.6   74     5   3    0.1622   0.168      FALSE      -3.035
 \end{Soutput}
@@
 \begin{Sinput}
-R> dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt
+R> dat <- get_vars(mtcars, c("mpg", "disp")); w <- mtcars$wt;
 R> add_vars(g$groups,
@@
 These are integrated with the \emph{Fast Statistical Functions} to enable vectorized statistical operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow.
-For example, the following code calculates
-%
-\begin{Schunk}
-\begin{Sinput}
-R> mtcars |>
-+    fsubset(mpg > 11) |>
-+    fgroup_by(cyl, vs, am) |>
+For example, the following code calculates the mean of columns
+%
+\begin{Schunk}
+\begin{Sinput}
+R> mtcars |> fsubset(mpg > 11) |> fgroup_by(cyl, vs, am) |>
 +    fsummarise(across(c(mpg, carb, hp), fmean),
@@
 Thus \code{w} becomes the second positional argument...} The \emph{Fast Statistical Functions} also have a method for grouped data, so \code{fsummarise} is not always needed.
-The following code calculates weighted group means.
-By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::fgroup\_by}.
-The latter requires an additional C routine to convert the grouping object to a \class{GRP} object, and is thus much less efficient.}
-%
-\begin{Schunk}
-\begin{Sinput}
-R> mtcars |>
-+    fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
-+    fgroup_by(cyl, vs, am) |>
-+    fmean(wt) |> head(2)
+The following example calculates weighted group means.
+By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::group\_by}.
+The latter requires an additional C routine to convert the \pkg{dplyr} grouping object to a \class{GRP} object, and is thus less efficient.}
+%
+\begin{Schunk}
+\begin{Sinput}
+R> mtcars |> fsubset(mpg > 11, cyl, vs, am, mpg, carb, hp, wt) |>
++    fgroup_by(cyl, vs, am) |> fmean(wt) |> head(2)
 \end{Sinput}
@@
 \begin{Sinput}
-R> mtcars |>
-+   fgroup_by(vs) |>
+R> mtcars |> fgroup_by(vs) |>
 +   fmutate(dm_carb = fmean(carb, TRA = "-")) |>
@@
 \begin{Sinput}
-R> mtcars |>
-+      fgroup_by(cyl, vs, am) |>
+R> mtcars |> fgroup_by(cyl, vs, am) |>
 +    fmutate(o = radixorder(GRPid(), mpg)) |>
@@
 \begin{Sinput}
-R> LIFEEXi = wldi$LIFEEX
-R> str(LIFEEXi, width = 70, strict = "cut")
+R> LIFEEXi <- wldi$LIFEEX; str(LIFEEXi, width = 70, strict = "cut")
 \end{Sinput}
@@
 By default (\code{verbose = 1}), it prints information about the join operation and number of records joined.
-Users can request the generation of a \code{.join} column (\code{column = TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, indicating the origin of records in the joined table.
-%
-\begin{Schunk}
-\begin{Sinput}
-R> df1 <- data.frame(id1 = c(1, 1, 2, 3),
-+                    id2 = c("a", "b", "b", "c"),
-+                    name = c("John", "Jane", "Bob", "Carl"),
-+                    age = c(35, 28, 42, 50))
-R> df2 <- data.frame(id1 = c(1, 2, 3, 3),
-+                    id2 = c("a", "b", "c", "e"),
+Users can request the generation of a \code{.join} column (\code{column = "name"/TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, indicating the origin of records in the joined table.
+%
+\begin{Schunk}
+\begin{Sinput}
+R> df1 <- data.frame(id1 = c(1, 1, 2, 3), id2 = c("a", "b", "b", "c"),
++    name = c("John", "Jane", "Bob", "Carl"), age = c(35, 28, 42, 50))
+R> df2 <- data.frame(id1 = c(1, 2, 3, 3), id2 = c("a", "b", "c", "e"),
 +    salary = c(60000, 55000, 70000, 80000),
@@
   type type_name id      r     h
-1    A    Apples  1 1.42361 1.7860
-2    A    Apples  2 1.30998 0.0831
-3    B   Bananas  1 0.04423 1.6729
-4    B   Bananas  2 0.08639 1.2877
+1    A    Apples  1 0.3916 1.845
+2    A    Apples  2 1.0761 1.667
+3    B   Bananas  1 0.4958 1.976
+4    B   Bananas  2 1.2679 2.974
 \end{Soutput}
@@
   type type_name id variable        label  value
-1    A    Apples  1        r Fruit Radius 1.42361
-2    A    Apples  2        r Fruit Radius 1.30998
-3    B   Bananas  1        r Fruit Radius 0.04423
-4    B   Bananas  2        r Fruit Radius 0.08639
-5    A    Apples  1        h Fruit Height 1.78599
-6    A    Apples  2        h Fruit Height 0.08310
-7    B   Bananas  1        h Fruit Height 1.67294
-8    B   Bananas  2        h Fruit Height 1.28770
+1    A    Apples  1        r Fruit Radius 0.3916
+2    A    Apples  2        r Fruit Radius 1.0761
+3    B   Bananas  1        r Fruit Radius 0.4958
+4    B   Bananas  2        r Fruit Radius 1.2679
+5    A    Apples  1        h Fruit Height 1.8454
+6    A    Apples  2        h Fruit Height 1.6666
+7    B   Bananas  1        h Fruit Height 1.9762
+8    B   Bananas  2        h Fruit Height 2.9737
 \end{Soutput}
@@
   id    r_A    r_B   h_A   h_B
-1  1 1.424 0.04423 1.7860 1.673
-2  2 1.310 0.08639 0.0831 1.288
+1  1 0.3916 0.4958 1.845 1.976
+2  2 1.0761 1.2679 1.667 2.974
 \end{Soutput}
@@
   id variable        label      A      B
-1  1        r Fruit Radius 1.4236 0.04423
-2  2        r Fruit Radius 1.3100 0.08639
-3  1        h Fruit Height 1.7860 1.67294
-4  2        h Fruit Height 0.0831 1.28770
+1  1        r Fruit Radius 0.3916 0.4958
+2  2        r Fruit Radius 1.0761 1.2679
+3  1        h Fruit Height 1.8454 1.9762
+4  2        h Fruit Height 1.6666 2.9737
 \end{Soutput}
@@
 \begin{Sinput}
-R> nest_lm_coef <- dl |>
-+    rapply2d(lm, formula = mpg ~ .) |>
-+    rapply2d(summary, classes = "lm") |>
-+    get_elem("coefficients")
+R> nest_lm_coef <- dl |> rapply2d(lm, formula = mpg ~ .) |>
++    rapply2d(summary, classes = "lm") |> get_elem("coefficients")
 R> nest_lm_coef |> str(give.attr = FALSE, strict = "cut")
@@
 \pkg{data.table}'s \href{https://rdatatable.gitlab.io/data.table/reference/datatable-optimize.html}{GForce optimization} also only applies to simple statistics, not complex expressions or weighted statistics - as can be vectorized using \emph{Fast Statistical Functions} in \pkg{collapse}.
-Thus, if your data is moderately sized ($\leq$100mio obs), you have more than 1 column to compute on, you want to do complex statistical things, or if your processor is very fast (high single core speed), \pkg{collapse} is a great choice.
-On the other hand, if your data is really long ($>$100mio obs), you have only a few columns to compute on; you are computing simple statistics that \pkg{data.table} optimizes, and you have massive parallel compute, then \pkg{data.table} is a great choice.
-Let me finally note that packages like \pkg{polars} use optimized memory formats based on \href{https://arrow.apache.org/}{Apache Arrow}, \href{https://pola.rs/posts/i-wrote-one-of-the-fastest-dataframe-libraries/}{multithreaded hash-based grouping, SIMD instructions and multithreading at the group-level}, and a \href{https://pola.rs/posts/polars_birds_eye_view/}{query optimizer} - all implemented in \proglang{Rust}, a thread-safe programming language.
-Nobody will ever develop something similar off \proglang{R}'s 30-year-old C API.
-But then, \proglang{R} is a statistical language not intended to work like an optimized database.
-\pkg{collapse} seamlessly integrates with \proglang{R}, while \pkg{polars}, at present, has very little to do with it.}
+Thus, if your data is moderately sized ($\leq$100mio.
+obs.), you have more than 1 column to compute on, you want to do complex statistical things, or if your processor is very fast (high single core speed), \pkg{collapse} is a great choice.
+On the other hand, if your data is really long ($>$100mio.
+obs.), you have only a few columns to compute on, you are computing simple statistics that \pkg{data.table} optimizes, and you have massive parallel compute, then \pkg{data.table} is a great choice.
+My recommendation: use both, just need to call \code{library(fastverse)}.
+Finally, let me note that \pkg{polars} uses optimized memory buffers based on \href{https://arrow.apache.org/}{Apache Arrow}, \href{https://pola.rs/posts/i-wrote-one-of-the-fastest-dataframe-libraries/}{multithreaded hash-based grouping, SIMD instructions and multithreading at the group-level}, and a \href{https://pola.rs/posts/polars_birds_eye_view/}{query optimizer} - all implemented in \proglang{Rust}, a thread-safe programming language.
+While some of these parallel algorithms could be ported to \pkg{collapse}, this is more challenging since C, and particularly R's C API, is not thread safe - and it would still be lacking the benefits of Arrow memory buffers.
+At core, \proglang{R} is a 30-year old statistical language and not intended to work like an optimized database.
+\pkg{collapse} seamlessly integrates with \proglang{R}'s data structures; \pkg{polars}, at present, has nothing to do with them (and is therefore also not part of this benchmark).}
 %
@@
      expr   min    lq   mean median     uq   max neval
-        R 9.653 9.837 10.127  9.850 9.896 21.768   100
-    Rfast 1.307 1.789  1.914  1.835 1.893  6.005   100
- collapse 1.322 1.431  1.566  1.474 1.577  4.987   100
+        R 9.827 9.846 10.453  9.872 10.069 21.22   100
+    Rfast 1.316 1.792  2.352  1.857  2.110 11.40   100
+ collapse 1.324 1.449  1.965  1.520  1.772 11.31   100
 \end{Soutput}
@@
        expr    min     lq   mean median     uq     max neval
-          R 10.75 11.213 11.706 11.281 11.472 16.423   100
- data.table 11.45 18.526 19.976 19.278 20.465 57.433   100
-   collapse  1.92  2.713  3.303  2.959  3.552  6.251   100
+          R 10.921 11.244 11.810 11.315 11.673  16.293   100
+ data.table  8.183 18.755 21.582 19.753 21.508 113.610   100
+   collapse  1.913  2.819  3.393  3.051  3.445   7.233   100
 \end{Soutput}
@@
        expr    min     lq   mean median     uq    max neval
- data.table 136.21 137.08 138.09 137.64 138.34 143.8   100
-   collapse  65.79  75.85  78.94  79.23  81.74 116.9   100
+ data.table 135.72 137.46 139.66 138.75 140.71 158.89   100
+   collapse  63.92  74.84  79.86  79.75  85.08  96.31   100
 \end{Soutput}
@@
        expr   min     lq   mean median     uq     max neval
- data.table 8.438 11.882 24.01 13.756 18.897 62.78   100
-   collapse 1.221  1.382  1.48  1.408  1.466  2.94   100
+ data.table 8.930 13.689 29.964 16.518 57.245 125.224   100
+   collapse 1.281  1.419  1.553  1.485  1.585   2.295   100
 \end{Soutput}
@@
        expr   min    lq  mean median    uq   max neval
- data.table 11.32 14.70 21.36  16.00 17.63 58.90   100
-   collapse 11.52 14.33 21.07  15.66 16.89 61.89   100
+ data.table 11.65 15.19 22.12  16.39 18.31 62.51   100
+   collapse 11.67 15.02 21.86  16.05 17.61 63.43   100
 \end{Soutput}
@@
        expr   min     lq   mean median    uq   max neval
- data.table 74.29 115.48 118.73 118.87 122.64 135.1   100
-   collapse 55.69  87.16  91.14  90.35  94.12 118.8   100
+ data.table 69.41 116.29 120.08 119.79 123.5 143.8   100
+   collapse 57.53  88.93  92.62  91.98  96.6 125.0   100
 \end{Soutput}
@@
           expr    min     lq  mean median     uq    max neval
-      base_int 60.216 63.192 64.81 64.170 65.81 108.20   100
-  collapse_int  8.624  9.192 10.44  9.384 10.35  18.11   100
-     base_char 92.110 94.566 98.56 97.757 99.91 160.08   100
- collapse_char 20.464 23.445 25.37 23.744 24.71 113.24   100
+      base_int 59.641 62.592 64.06 63.488 64.871 111.96   100
+  collapse_int  8.413  9.133 10.11  9.291  9.694  15.32   100
+     base_char 91.712 93.905 96.68 95.830 97.707 141.16   100
+ collapse_char 21.339 22.828 24.25 23.053 23.813  71.40   100
 \end{Soutput}
@@
             expr    min     lq   mean median    uq   max neval
-        base_int 26.877 27.027 28.455 27.132 30.216 31.18    10
-    collapse_int  8.637  8.701  9.269  8.722  8.973 13.14    10
-       base_char 86.277 86.299 90.211 87.652 92.643 98.55    10
- data.table_char 41.150 41.615 42.225 41.806 41.906 47.05    10
-   collapse_char 32.333 32.548 33.099 32.620 32.819 37.39    10
+        base_int 26.730 26.743 28.290  26.89 30.47 31.18    10
+    collapse_int  8.584  8.609  9.103   8.66  8.69 13.19    10
+       base_char 86.599 87.263 90.344  88.16 91.71 98.66    10
+ data.table_char 41.636 41.751 42.335  41.83 41.88 47.11    10
+   collapse_char 32.563 32.793 33.322  32.88 32.94 37.40    10
 \end{Soutput}
@@
 \pkg{collapse} is my attempt at providing such a package, and the feedback I have received over the years, particularly from users in academia, government, and international organizations, is a strong indication that I have responded to a need felt in larger parts of the \proglang{R} community.
-As mentioned, a single article cannot summarize \pkg{collapse}, but there is an excellent \href{https://sebkrantz.github.io/collapse/index.html}{website} with comprehensive \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{Documentation Resources}.
+As mentioned, a single article cannot comprehensively introduce \pkg{collapse}, but there is a modern \href{https://sebkrantz.github.io/collapse/index.html}{website} with comprehensive \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}.
 

### 2024-03-07 misc/JSS article/jss-article-rnw/article.Rnw
Adding \proglang for C/C++ etc.

@@
 \Abstract{
-\pkg{collapse} is a large C/C++-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency.
+\pkg{collapse} is a large \proglang{C/C++}-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency.
 It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem.
@@
 %
-\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large C/C++-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.
+\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}}\footnote{Website: https://sebkrantz.github.io/collapse/} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.
 Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
@@
 It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}.
-Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and C-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal.
+Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and \proglang{C}-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal.
 It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}.
@@
 On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes (except for joins where it remains competitive).
-There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any Julia function.
+There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any \proglang{Julia} function.
 However, as demonstrated in Section~\ref{ssec:vfat}, vectorized statistical functions can be combined to calculate more complex statistics in a vectorized way.
@@
 The use of single instruction multiple data (SIMD) parallelism in single-threaded mode also implies limited gains from multithreading for simple operations such as \fct{fsum}.} The following examples, taken from the \href{https://sebkrantz.github.io/collapse/articles/collapse_for_tidyverse_users.html#using-the-fast-statistical-functions}{\pkg{collapse} for \pkg{tidyverse} Users} vignette demonstrate their basic usage to calculate (column-wise, grouped, weighted) statistics on different objects.
-As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding C implementations that intelligently preserve object attributes.
+As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes.
 Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods.
@@
 By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::group\_by}.
-The latter requires an additional C routine to convert the \pkg{dplyr} grouping object to a \class{GRP} object, and is thus less efficient.}
+The latter requires an additional \proglang{C} routine to convert the \pkg{dplyr} grouping object to a \class{GRP} object, and is thus less efficient.}
 %
@@
 Apart from vectorization, this code avoids 3 intermediate copies: (1) \code{fmean(carb, TRA = "-")} avoids an expanded vector of group means, (2) \code{fsum(mpg, dm\_carb)} uses the weights (\code{w}) argument to \fct{fsum} to avoid materializing a multiplication (as in \code{fsum(mpg * dm\_carb)}), and (3) division by reference (\code{\%/=\%}) avoids allocating an additional vector for the final result.
-Under the hood, the expression boils down to an (expensive) grouping step, 5 allocations (of which 2 full length), and 6 loops in C to calculate the result.
+Under the hood, the expression boils down to an (expensive) grouping step, 5 allocations (of which 2 full length), and 6 loops in \proglang{C} to calculate the result.
 Any modern laptop can calculate 1 million regression slopes in less than 1 second like this.
@@
 @
-\pkg{collapse} also offers higher-dimensional between and within transformations, powered by C++ code conditionally imported (and accessed directly) from \pkg{fixest}.
+\pkg{collapse} also offers higher-dimensional between and within transformations, powered by \proglang{C++} code conditionally imported (and accessed directly) from \pkg{fixest}.
 The following detrends GDP per Capita and Life Expectancy at Birth using country-specific cubic polynomials.
@@
 Finally, let me note that \pkg{polars} uses optimized memory buffers based on \href{https://arrow.apache.org/}{Apache Arrow}, \href{https://pola.rs/posts/i-wrote-one-of-the-fastest-dataframe-libraries/}{multithreaded hash-based grouping, SIMD instructions and multithreading at the group-level}, and a \href{https://pola.rs/posts/polars_birds_eye_view/}{query optimizer} - all implemented in \proglang{Rust}, a thread-safe programming language.
-While some of these parallel algorithms could be ported to \pkg{collapse}, this is more challenging since C, and particularly R's C API, is not thread safe - and it would still be lacking the benefits of Arrow memory buffers.
+While some of these parallel algorithms could be ported to \pkg{collapse}, this is more challenging since \proglang{C}, and particularly \proglang{R}'s \proglang{C} API, is not thread safe - and it would still be lacking the benefits of Arrow memory buffers.
 At core, \proglang{R} is a 30-year old statistical language and not intended to work like an optimized database.

### 2024-03-07 misc/JSS article/jss-article-rnw/article.tex
Adding \proglang for C/C++ etc.

@@
 \end{Schunk}
-\pkg{collapse} also offers higher-dimensional between and within transformations, powered by C++ code conditionally imported (and accessed directly) from \pkg{fixest}.
+\pkg{collapse} also offers higher-dimensional between and within transformations, powered by \proglang{C++} code conditionally imported (and accessed directly) from \pkg{fixest}.
 The following detrends GDP per Capita and Life Expectancy at Birth using country-specific cubic polynomials.
@@
   type type_name id      r      h
-1    A    Apples  1 0.3916 1.845
-2    A    Apples  2 1.0761 1.667
-3    B   Bananas  1 0.4958 1.976
-4    B   Bananas  2 1.2679 2.974
+1    A    Apples  1 0.0356 2.4303
+2    A    Apples  2 1.8575 0.5427
+3    B   Bananas  1 0.5966 1.8641
+4    B   Bananas  2 0.8115 0.7461
 \end{Soutput}
@@
   type type_name id variable        label  value
-1    A    Apples  1        r Fruit Radius 0.3916
-2    A    Apples  2        r Fruit Radius 1.0761
-3    B   Bananas  1        r Fruit Radius 0.4958
-4    B   Bananas  2        r Fruit Radius 1.2679
-5    A    Apples  1        h Fruit Height 1.8454
-6    A    Apples  2        h Fruit Height 1.6666
-7    B   Bananas  1        h Fruit Height 1.9762
-8    B   Bananas  2        h Fruit Height 2.9737
+1    A    Apples  1        r Fruit Radius 0.0356
+2    A    Apples  2        r Fruit Radius 1.8575
+3    B   Bananas  1        r Fruit Radius 0.5966
+4    B   Bananas  2        r Fruit Radius 0.8115
+5    A    Apples  1        h Fruit Height 2.4303
+6    A    Apples  2        h Fruit Height 0.5427
+7    B   Bananas  1        h Fruit Height 1.8641
+8    B   Bananas  2        h Fruit Height 0.7461
 \end{Soutput}
@@
   id    r_A    r_B    h_A    h_B
-1  1 0.3916 0.4958 1.845 1.976
-2  2 1.0761 1.2679 1.667 2.974
+1  1 0.0356 0.5966 2.4303 1.8641
+2  2 1.8575 0.8115 0.5427 0.7461
 \end{Soutput}
@@
   id variable        label      A      B
-1  1        r Fruit Radius 0.3916 0.4958
-2  2        r Fruit Radius 1.0761 1.2679
-3  1        h Fruit Height 1.8454 1.9762
-4  2        h Fruit Height 1.6666 2.9737
+1  1        r Fruit Radius 0.0356 0.5966
+2  2        r Fruit Radius 1.8575 0.8115
+3  1        h Fruit Height 2.4303 1.8641
+4  2        h Fruit Height 0.5427 0.7461
 \end{Soutput}
@@
      expr   min    lq   mean median     uq   max neval
-        R 9.827 9.846 10.453  9.872 10.069 21.22   100
-    Rfast 1.316 1.792  2.352  1.857  2.110 11.40   100
- collapse 1.324 1.449  1.965  1.520  1.772 11.31   100
+        R 9.755 9.871 10.279  9.978 10.422 14.36   100
+    Rfast 1.374 1.823  2.460  1.974  2.376 15.22   100
+ collapse 1.415 1.488  2.088  1.535  1.702 12.40   100
 \end{Soutput}
@@
        expr    min     lq   mean median     uq   max neval
-          R 10.921 11.244 11.810 11.315 11.673  16.293   100
- data.table  8.183 18.755 21.582 19.753 21.508 113.610   100
-   collapse  1.913  2.819  3.393  3.051  3.445   7.233   100
+          R 10.948 11.254 13.085 11.404 11.749 96.20   100
+ data.table 13.031 19.024 22.010 19.743 23.276 62.05   100
+   collapse  2.213  2.874  3.607  3.084  3.567 13.94   100
 \end{Soutput}
@@
        expr    min     lq   mean median     uq    max neval
- data.table 135.72 137.46 139.66 138.75 140.71 158.89   100
-   collapse  63.92  74.84  79.86  79.75  85.08  96.31   100
+ data.table 136.63 137.82 139.30 138.51 140.47 149.33   100
+   collapse  73.28  79.03  82.12  81.74  84.15  99.39   100
 \end{Soutput}
@@
        expr   min     lq   mean median     uq    max neval
- data.table 8.930 13.689 29.964 16.518 57.245 125.224   100
-   collapse 1.281  1.419  1.553  1.485  1.585   2.295   100
+ data.table 8.525 12.288 24.414 14.395 20.293 73.406   100
+   collapse 1.272  1.386  1.477  1.424  1.459  2.843   100
 \end{Soutput}
@@
        expr   min    lq  mean median    uq   max neval
- data.table 11.65 15.19 22.12  16.39 18.31 62.51   100
-   collapse 11.67 15.02 21.86  16.05 17.61 63.43   100
+ data.table 11.83 15.16 22.61  16.60 18.44 67.87   100
+   collapse 11.75 14.87 22.46  16.33 18.52 71.10   100
 \end{Soutput}
@@
        expr   min     lq   mean median     uq   max neval
- data.table 69.41 116.29 120.08 119.79 123.5 143.8   100
-   collapse 57.53  88.93  92.62  91.98  96.6 125.0   100
+ data.table 73.64 113.04 117.94 115.83 121.94 181.0   100
+   collapse 54.87  72.95  83.95  79.95  92.68 134.4   100
 \end{Soutput}
@@
           expr   min     lq  mean median     uq    max neval
-      base_int 59.641 62.592 64.06 63.488 64.871 111.96   100
-  collapse_int  8.413  9.133 10.11  9.291  9.694  15.32   100
-     base_char 91.712 93.905 96.68 95.830 97.707 141.16   100
- collapse_char 21.339 22.828 24.25 23.053 23.813  71.40   100
+      base_int 60.68 63.462 66.31 64.833  66.42 129.54   100
+  collapse_int  8.50  9.193 10.48  9.374  10.23  19.99   100
+     base_char 92.71 95.737 99.69 98.380 100.45 149.15   100
+ collapse_char 22.25 23.781 25.86 24.143  25.14  79.24   100
 \end{Soutput}
@@
             expr    min     lq   mean median     uq   max neval
-        base_int 26.730 26.743 28.290  26.89 30.47 31.18    10
-    collapse_int  8.584  8.609  9.103   8.66  8.69 13.19    10
-       base_char 86.599 87.263 90.344  88.16 91.71 98.66    10
- data.table_char 41.636 41.751 42.335  41.83 41.88 47.11    10
-   collapse_char 32.563 32.793 33.322  32.88 32.94 37.40    10
+        base_int 27.156 27.313 29.131 27.956 31.399 32.45    10
+    collapse_int  8.766  8.838  9.395  8.862  8.901 13.63    10
+       base_char 84.300 84.414 88.440 85.985 89.799 98.14    10
+ data.table_char 42.195 42.372 43.311 42.621 43.390 48.58    10
+   collapse_char 29.627 29.786 30.585 30.038 30.367 34.70    10
 \end{Soutput}

### 2024-03-08 misc/JSS article/jss-article-rnw/article.tex
Minors + using nojss class.

@@
-\documentclass[article]{jss}
+\documentclass[nojss]{jss} % article
 
@@
 While \pkg{collapse} has a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, its implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} is particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and 'recast' pivots.
-Both implementations also provide outstanding computational performance and memory efficiency.
+Both implementations provide outstanding computational performance and memory efficiency.
 %
@@
   type type_name id      r     h
-1    A    Apples  1 0.0356 2.4303
-2    A    Apples  2 1.8575 0.5427
-3    B   Bananas  1 0.5966 1.8641
-4    B   Bananas  2 0.8115 0.7461
+1    A    Apples  1 0.3139 2.983
+2    A    Apples  2 0.4095 1.572
+3    B   Bananas  1 0.4302 2.220
+4    B   Bananas  2 0.1851 2.380
 \end{Soutput}
@@
   type type_name id variable        label  value
-1    A    Apples  1        r Fruit Radius 0.0356
-2    A    Apples  2        r Fruit Radius 1.8575
-3    B   Bananas  1        r Fruit Radius 0.5966
-4    B   Bananas  2        r Fruit Radius 0.8115
-5    A    Apples  1        h Fruit Height 2.4303
-6    A    Apples  2        h Fruit Height 0.5427
-7    B   Bananas  1        h Fruit Height 1.8641
-8    B   Bananas  2        h Fruit Height 0.7461
+1    A    Apples  1        r Fruit Radius 0.3139
+2    A    Apples  2        r Fruit Radius 0.4095
+3    B   Bananas  1        r Fruit Radius 0.4302
+4    B   Bananas  2        r Fruit Radius 0.1851
+5    A    Apples  1        h Fruit Height 2.9829
+6    A    Apples  2        h Fruit Height 1.5722
+7    B   Bananas  1        h Fruit Height 2.2203
+8    B   Bananas  2        h Fruit Height 2.3797
 \end{Soutput}
@@
   id    r_A    r_B   h_A  h_B
-1  1 0.0356 0.5966 2.4303 1.8641
-2  2 1.8575 0.8115 0.5427 0.7461
+1  1 0.3139 0.4302 2.983 2.22
+2  2 0.4095 0.1851 1.572 2.38
 \end{Soutput}
@@
   id variable        label      A      B
-1  1        r Fruit Radius 0.0356 0.5966
-2  2        r Fruit Radius 1.8575 0.8115
-3  1        h Fruit Height 2.4303 1.8641
-4  2        h Fruit Height 0.5427 0.7461
+1  1        r Fruit Radius 0.3139 0.4302
+2  2        r Fruit Radius 0.4095 0.1851
+3  1        h Fruit Height 2.9829 2.2203
+4  2        h Fruit Height 1.5722 2.3797
 \end{Soutput}
@@
      expr   min    lq   mean median    uq    max neval
-        R 9.755 9.871 10.279  9.978 10.422 14.36   100
-    Rfast 1.374 1.823  2.460  1.974  2.376 15.22   100
- collapse 1.415 1.488  2.088  1.535  1.702 12.40   100
+        R 9.827 9.858 10.061  9.885 9.970 16.701   100
+    Rfast 1.349 1.819  2.135  1.877 2.009 11.236   100
+ collapse 1.321 1.475  1.728  1.521 1.701  5.502   100
 \end{Soutput}
@@
        expr    min     lq   mean median     uq    max neval
-          R 10.948 11.254 13.085 11.404 11.749 96.20   100
- data.table 13.031 19.024 22.010 19.743 23.276 62.05   100
-   collapse  2.213  2.874  3.607  3.084  3.567 13.94   100
+          R 11.002 11.331 12.255 11.516 11.946 24.269   100
+ data.table 13.918 19.036 21.956 19.746 21.676 64.273   100
+   collapse  1.911  2.743  3.408  3.136  3.702  6.514   100
 \end{Soutput}
@@
        expr    min     lq   mean median     uq   max neval
- data.table 136.63 137.82 139.30 138.51 140.47 149.33   100
-   collapse  73.28  79.03  82.12  81.74  84.15  99.39   100
+ data.table 136.94 137.72 141.04 138.48 140.01 265.1   100
+   collapse  70.48  76.56  81.03  79.72  82.17 191.9   100
 \end{Soutput}
@@
        expr   min     lq   mean median     uq    max neval
- data.table 8.525 12.288 24.414 14.395 20.293 73.406   100
-   collapse 1.272  1.386  1.477  1.424  1.459  2.843   100
+ data.table 9.153 12.140 24.382 14.118 20.849 66.701   100
+   collapse 1.220  1.391  1.472  1.424  1.466  2.948   100
 \end{Soutput}
@@
        expr   min    lq mean median    uq   max neval
- data.table 11.83 15.16 22.61  16.60 18.44 67.87   100
-   collapse 11.75 14.87 22.46  16.33 18.52 71.10   100
+ data.table 11.96 15.34 22.6  16.63 18.41 75.20   100
+   collapse 11.74 15.01 22.0  16.32 18.28 63.01   100
 \end{Soutput}
@@
        expr   min     lq   mean median     uq   max neval
- data.table 73.64 113.04 117.94 115.83 121.94 181.0   100
-   collapse 54.87  72.95  83.95  79.95  92.68 134.4   100
+ data.table 72.48 116.74 119.65  119.7 123.68 142.7   100
+   collapse 55.83  87.97  92.83   90.6  93.19 200.7   100
 \end{Soutput}
@@
           expr    min     lq  mean median     uq    max neval
-      base_int 60.68 63.462 66.31 64.833  66.42 129.54   100
-  collapse_int  8.50  9.193 10.48  9.374  10.23  19.99   100
-     base_char 92.71 95.737 99.69 98.380 100.45 149.15   100
- collapse_char 22.25 23.781 25.86 24.143  25.14  79.24   100
+      base_int 60.294 63.508 65.73 65.251  66.62 110.97   100
+  collapse_int  8.677  9.347 10.49  9.595  10.91  15.83   100
+     base_char 92.569 95.032 99.15 98.483 100.67 141.98   100
+ collapse_char 21.756 23.128 24.91 23.691  24.46  88.37   100
 \end{Soutput}
@@
             expr   min     lq   mean median      uq    max neval
-        base_int 27.156 27.313 29.131 27.956 31.399 32.45    10
-    collapse_int  8.766  8.838  9.395  8.862  8.901 13.63    10
-       base_char 84.300 84.414 88.440 85.985 89.799 98.14    10
- data.table_char 42.195 42.372 43.311 42.621 43.390 48.58    10
-   collapse_char 29.627 29.786 30.585 30.038 30.367 34.70    10
+        base_int 27.24 27.502 29.085 28.093  30.789  32.41    10
+    collapse_int  8.78  8.867  9.429  8.891   9.185  13.29    10
+       base_char 94.72 96.474 99.857 97.048 101.644 110.06    10
+ data.table_char 42.21 42.412 44.063 42.758  43.287  51.68    10
+   collapse_char 36.61 36.769 37.592 36.959  37.667  41.62    10
 \end{Soutput}

### 2024-03-11 misc/JSS article/jss-article-rnw/article.Rnw
Minors + using nojss class.

@@
 
-\pkg{collapse} also vectorizes advanced statistics, such as weighted medians and modes.
+\pkg{collapse} also vectorizes advanced statistics, such as weighted quantiles and modes.
 The following example calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} calculates weighted quantiles by replacing the sample size with the sum of weights and 1 with the minimum non-zero weight in the respective quantile definition.

### 2024-03-20 misc/JSS article/jss-article-rnw/article.Rnw
Using memory efficiency.

@@
 \Abstract{
-\pkg{collapse} is a large \proglang{C/C++}-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and programming efficiency.
+\pkg{collapse} is a large \proglang{C/C++}-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R} - at outstanding levels of performance and memory efficiency.
 It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular variants (e.g., \class{factor}, \class{ts}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}), enabling its seamless integration with large parts of the \proglang{R} ecosystem.

### 2024-04-13 R/BY.R
Adding explicit support for xts/zoo and units.

@@ -7,7 +7,7 @@ BY.default <- function(x, g, FUN, ..., use.g.names = TRUE, sort = .op[["sort"]],
                        return = c("same", "vector", "list")) {
 
   # If matrix, dispatch to matrix method
-  if(is.matrix(x) && !inherits(x, "matrix")) return(UseMethod("BY", unclass(x)))
+  # if(is.matrix(x) && !inherits(x, "matrix")) return(UseMethod("BY", unclass(x)))
   if(!is.function(FUN)) FUN <- match.fun(FUN)
 
   aplyfun <- if(parallel) function(...) mclapply(..., mc.cores = mc.cores) else lapply
@@ -373,4 +373,5 @@ BY.grouped_df <- function(x, FUN, ..., reorder = TRUE, keep.group_vars = TRUE, u
   condalcSA(c(g[[4L]], res), ar, any(ar$class == "data.table"))
 }
 
-
+BY.zoo <- function(x, ...) if(is.matrix(x)) BY.matrix(x, ...) else BY.default(x, ...)
+BY.units <- BY.zoo # return = "same" preserves attributes by default

### 2024-04-13 R/ffirst.R

@@ -4,7 +4,7 @@
 ffirst <- function(x, ...) UseMethod("ffirst") # , x
 
 ffirst.default <- function(x, g = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(ffirst.matrix(x, g, TRA, na.rm, use.g.names, ...))
+  # if(is.matrix(x) && !inherits(x, "matrix")) return(ffirst.matrix(x, g, TRA, na.rm, use.g.names, ...))
   if(is.null(TRA)) {
     if(!missing(...)) unused_arg_action(match.call(), ...)
     if(is.null(g)) return(.Call(C_ffirst,x,0L,0L,NULL,na.rm))
@@ -49,6 +49,9 @@ ffirst.matrix <- function(x, g = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g
   g <- G_guo(g)
   TRAmC(x,.Call(C_ffirstm,x,g[[1L]],g[[2L]],g$group.starts,na.rm,FALSE),g[[2L]],TRA, ...)
 }
+
+ffirst.zoo <- function(x, ...) if(is.matrix(x)) ffirst.matrix(x, ...) else ffirst.default(x, ...)
+ffirst.units <- function(x, ...) if(is.matrix(x)) copyMostAttrib(ffirst.matrix(x, ...), x) else ffirst.default(x, ...)
 
 ffirst.data.frame <- function(x, g = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, drop = TRUE, ...) {
   if(is.null(TRA)) {

### 2024-04-13 R/fhdbetween_fhdwithin.R

@@ -276,7 +276,7 @@ flmres <- function(y, X, w = NULL, meth = "qr", resi = TRUE, ...) {
 fhdwithin <- function(x, ...) UseMethod("fhdwithin") # , x
 
 fhdwithin.default <- function(x, fl, w = NULL, na.rm = .op[["na.rm"]], fill = FALSE, lm.method = "qr", ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(fhdwithin.matrix(x, fl, w, na.rm, fill, ...))
+  # if(is.matrix(x) && !inherits(x, "matrix")) return(fhdwithin.matrix(x, fl, w, na.rm, fill, ...))
   ax <- attributes(x)
   if(na.rm) {
     cc <- complete.cases(x, fl, w) # gives error if lengths don't match, otherwise demeanlist and qr.resid give errors !!
@@ -418,6 +418,9 @@ fhdwithin.matrix <- function(x, fl, w = NULL, na.rm = .op[["na.rm"]], fill = FAL
   } else return(setAttributes(demean(x, fl, w, ...), ax))
 }
 
+fhdwithin.zoo <- function(x, ...) if(is.matrix(x)) fhdwithin.matrix(x, ...) else fhdwithin.default(x, ...)
+fhdwithin.units <- fhdwithin.zoo
+
 # x = collapse:::colsubset(pwlddev, is.numeric)
 fhdwithin.pdata.frame <- function(x, effect = "all", w = NULL, na.rm = .op[["na.rm"]], fill = TRUE, variable.wise = TRUE, ...) {
   ix <- findex(x)
@@ -539,7 +542,7 @@ fhdwithin.list <- function(x, ...) fhdwithin.data.frame(x, ...)
 HDW <- function(x, ...) UseMethod("HDW") # , x
 
 HDW.default <- function(x, fl, w = NULL, na.rm = .op[["na.rm"]], fill = FALSE, lm.method = "qr", ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(HDW.matrix(x, fl, w, na.rm, fill, lm.method, ...))
+  # if(is.matrix(x) && !inherits(x, "matrix")) return(HDW.matrix(x, fl, w, na.rm, fill, lm.method, ...))
   fhdwithin.default(x, fl, w, na.rm, fill, lm.method, ...)
 }
 
@@ -551,6 +554,9 @@ HDW.matrix <- function(x, fl, w = NULL, na.rm = .op[["na.rm"]], fill = FALSE, st
[…]

### 2024-04-13 R/fmean.R

@@ -4,7 +4,7 @@
 fmean <- function(x, ...) UseMethod("fmean") # , x
 
 fmean.default <- function(x, g = NULL, w = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, nthreads = .op[["nthreads"]], ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(fmean.matrix(x, g, w, TRA, na.rm, use.g.names, nthreads = nthreads, ...))
+  # if(is.matrix(x) && !inherits(x, "matrix")) return(fmean.matrix(x, g, w, TRA, na.rm, use.g.names, nthreads = nthreads, ...))
   if(is.null(TRA)) {
     if(!missing(...)) unused_arg_action(match.call(), ...)
     if(is.null(g)) return(.Call(C_fmean,x,0L,0L,NULL,w,na.rm,nthreads))
@@ -49,6 +49,9 @@ fmean.matrix <- function(x, g = NULL, w = NULL, TRA = NULL, na.rm = .op[["na.rm"
   g <- G_guo(g)
   TRAmC(x,.Call(C_fmeanm,x,g[[1L]],g[[2L]],g[[3L]],w,na.rm,drop,nthreads),g[[2L]],TRA, ...)
 }
+
+fmean.zoo <- function(x, ...) if(is.matrix(x)) fmean.matrix(x, ...) else fmean.default(x, ...)
+fmean.units <- function(x, ...) if(is.matrix(x)) copyMostAttrib(fmean.matrix(x, ...), x) else fmean.default(x, ...)
 
 fmean.data.frame <- function(x, g = NULL, w = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, drop = TRUE, nthreads = .op[["nthreads"]], ...) {
   if(is.null(TRA)) {

### 2024-04-13 R/fmin_fmax.R

@@ -4,7 +4,7 @@
 fmin <- function(x, ...) UseMethod("fmin") # , x
 
 fmin.default <- function(x, g = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(fmin.matrix(x, g, TRA, na.rm, use.g.names, ...))
+  # if(is.matrix(x) && !inherits(x, "matrix")) return(fmin.matrix(x, g, TRA, na.rm, use.g.names, ...))
   if(is.null(TRA)) {
     if(!missing(...)) unused_arg_action(match.call(), ...)
     if(is.null(g)) return(.Call(C_fmin,x,0L,0L,na.rm))
@@ -49,6 +49,9 @@ fmin.matrix <- function(x, g = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.n
   g <- G_guo(g)
   TRAmC(x,.Call(C_fminm,x,g[[1L]],g[[2L]],na.rm,FALSE),g[[2L]],TRA, ...)
 }
+
+fmin.zoo <- function(x, ...) if(is.matrix(x)) fmin.matrix(x, ...) else fmin.default(x, ...)
+fmin.units <- function(x, ...) if(is.matrix(x)) copyMostAttrib(fmin.matrix(x, ...), x) else fmin.default(x, ...)
 
 fmin.data.frame <- function(x, g = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, drop = TRUE, ...) {
   if(is.null(TRA)) {
@@ -115,7 +118,7 @@ fmin.grouped_df <- function(x, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names =
 fmax <- function(x, ...) UseMethod("fmax") # , x
 
 fmax.default <- function(x, g = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(fmax.matrix(x, g, TRA, na.rm, use.g.names, ...))
+  # if(is.matrix(x) && !inherits(x, "matrix")) return(fmax.matrix(x, g, TRA, na.rm, use.g.names, ...))
   if(is.null(TRA)) {
     if(!missing(...)) unused_arg_action(match.call(), ...)
[…]

### 2024-04-13 R/fndistinct.R

@@ -2,7 +2,7 @@
 fndistinct <- function(x, ...) UseMethod("fndistinct") # , x
 
 fndistinct.default <- function(x, g = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, nthreads = .op[["nthreads"]], ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(fndistinct.matrix(x, g, TRA, na.rm, use.g.names, nthreads = nthreads, ...))
+  # if(is.matrix(x) && !inherits(x, "matrix")) return(fndistinct.matrix(x, g, TRA, na.rm, use.g.names, nthreads = nthreads, ...))
   if(!is.null(g)) g <- GRP(g, return.groups = use.g.names && is.null(TRA), call = FALSE) # sort = FALSE for TRA: not faster here...
   res <- .Call(C_fndistinct,x,g,na.rm,nthreads)
   if(is.null(TRA)) {
@@ -25,6 +25,9 @@ fndistinct.matrix <- function(x, g = NULL, TRA = NULL, na.rm = .op[["na.rm"]], u
   }
   TRAmC(x,res,g[[2L]],TRA, ...)
 }
+
+fndistinct.zoo <- function(x, ...) if(is.matrix(x)) fndistinct.matrix(x, ...) else fndistinct.default(x, ...)
+fndistinct.units <- fndistinct.zoo
 
 fndistinct.data.frame <- function(x, g = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, drop = TRUE, nthreads = .op[["nthreads"]], ...) {
   if(!is.null(g)) g <- GRP(g, return.groups = use.g.names && is.null(TRA), call = FALSE) # sort = FALSE for TRA: not faster here...
@@ -82,6 +85,7 @@ fNdistinct <- function(x, ...) {
   UseMethod("fndistinct")
 }
 fNdistinct.default <- function(x, ...) {
+  if(is.matrix(x) && !inherits(x, "matrix")) return(fndistinct.matrix(x, ...))
   # .Deprecated(msg = "This method belongs to a renamed function and will be removed end of 2022, see help('collapse-renamed')")
   fndistinct.default(x, ...)
 }
@@ -93,3 +97,4 @@ fNdistinct.data.frame <- function(x, ...) {
   # .Deprecated(msg = "This method belongs to a renamed function and will be removed end of 2022, see help('collapse-renamed')")
   fndistinct.data.frame(x, ...)
 }
+

### 2024-04-13 R/fnth_fmedian.R

@@ -3,7 +3,7 @@
 fnth <- function(x, n = 0.5, ...) UseMethod("fnth") # , x
 
 fnth.default <- function(x, n = 0.5, g = NULL, w = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, ties = "q7", nthreads = .op[["nthreads"]], o = NULL, check.o = is.null(attr(o, "sorted")), ...) {
-  if(is.matrix(x) && !inherits(x, "matrix")) return(fnth.matrix(x, n, g, w, TRA, na.rm, use.g.names, ties = ties, nthreads = nthreads, ...))
+  # if(is.matrix(x) && !inherits(x, "matrix")) return(fnth.matrix(x, n, g, w, TRA, na.rm, use.g.names, ties = ties, nthreads = nthreads, ...))
   if(!is.null(g)) g <- GRP(g, return.groups = use.g.names && is.null(TRA), call = FALSE) # sort = FALSE for TRA: not faster here...
   res <- .Call(C_fnth, x, n, g, w, na.rm, ties, nthreads, o, check.o)
   if(is.null(TRA)) {
@@ -26,6 +26,9 @@ fnth.matrix <- function(x, n = 0.5, g = NULL, w = NULL, TRA = NULL, na.rm = .op[
   }
   TRAmC(x,res,g[[2L]],TRA, ...)
 }
+
+fnth.zoo <- function(x, ...) if(is.matrix(x)) fnth.matrix(x, ...) else fnth.default(x, ...)
+fnth.units <- function(x, ...) if(is.matrix(x)) copyMostAttrib(fnth.matrix(x, ...), x) else fnth.default(x, ...)
 
 fnth.data.frame <- function(x, n = 0.5, g = NULL, w = NULL, TRA = NULL, na.rm = .op[["na.rm"]], use.g.names = TRUE, drop = TRUE, ties = "q7", nthreads = .op[["nthreads"]], ...) {
   if(!is.null(g)) g <- GRP(g, return.groups = use.g.names && is.null(TRA), call = FALSE) # sort = FALSE for TRA: not faster here...
@@ -103,6 +106,9 @@ fmedian.default <- function(x, ..., ties = "mean")
 
 fmedian.matrix <- function(x, ..., ties = "mean")
   fnth.matrix(x, 0.5, ..., ties = ties)
+
[…]

### 2024-04-13 tests/testthat/test-misc.R

@@ -101,10 +101,9 @@ test_that("adding and removing stubs works", {
   expect_identical(rm_stub(names(iris), "Width", regex = TRUE)[4], "Petal.")
 })
 
-test_that("deep matrix dispatch works well", {
+test_that("zoo dispatch works well", {
 
-  tsm <- EuStockMarkets
-  class(tsm) <- setdiff(class(tsm), "matrix")
+  tsm <- zoo::as.zoo(EuStockMarkets)
   set.seed(101)
   f <- qF(sample.int(5, nrow(tsm), TRUE))
   NCOL2 <- function(x) if(length(d <- dim(x)) > 1L) d[2L] else length(x)
@@ -118,6 +117,32 @@ test_that("deep matrix dispatch works well", {
 
   for(i in c("flag", "L", "fdiff", "D", "Dlog", "fgrowth", "G"))
       expect_true(all(is.na(match.fun(i)(tsm)[1L, ])))
+
+})
+
+test_that("units support works well", {
+
+  v = abs(rnorm(5))
+  m = abs(matrix(rnorm(25), 5))
+  g = qF(c(1,1,2,3,3))
+  attributes(v) <- list(units = structure(list(numerator = "m", denominator = character(0)), class = "symbolic_units"), class = "units")
+  attributes(m) <- list(dim = c(5L, 5L), units = structure(list(numerator = "m", denominator = character(0)), class = "symbolic_units"), class = "units")
+
+  for (f in setdiff(c(.FAST_FUN, .OPERATOR_FUN), c("fnobs", "fndistinct", "F"))) {
+    # print(f)
+    FUN = match.fun(f)
+    if (!startsWith(f, "fhd") && !startsWith(f, "HD")) {
+      expect_true(inherits(FUN(v), "units"))
+      expect_true(inherits(FUN(m), "units"))
+    }
+    if (f %in% c("fnth","flag","L","fdiff","D","Dlog", "fgrowth","G")) {
+      expect_true(inherits(FUN(v, g = g), "units"))
+      expect_true(inherits(FUN(m, g = g), "units"))
+    } else {
+      expect_true(inherits(FUN(v, g), "units"))
+      expect_true(inherits(FUN(m, g), "units"))
+    }
+  }
 
 })
 

### 2025-02-19 misc/JSS article/R&R/article.Rnw

\documentclass[nojss]{jss} % article

%% -- LaTeX packages and custom commands ---------------------------------------

%% recommended packages
\usepackage{orcidlink,thumbpdf,lmodern,float}

%% another package (only for this demo article)
\usepackage{framed}

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% For Sweave-based articles about R packages:
%% need no \usepackage{Sweave}
%% need no \usepackage{Sweave.sty}
\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, cache=FALSE}
% \renewenvironment{knitrout}{\setlength{\topsep}{0mm}}{}

<<preliminaries, echo=FALSE, message=FALSE, warning=FALSE, results=hide>>=
# knitr::opts_chunk$set(cache = FALSE, size = "small")
rm(list = ls()); gc()
options(prompt = "R> ", continue = "+  ", width = 77, digits = 4, useFancyQuotes = FALSE, warn = 1)

# Loading libraries and installing if unavailable
if(!requireNamespace("fastverse", quietly = TRUE)) install.packages("fastverse")
library(fastverse) # loads data.table, collapse, magrittr and kit (not used)
# Package versions used in the article:
# fastverse 0.3.4, collapse 2.0.19, data.table 1.16.4,
# bench 1.1.3, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.0.0

# Reset collapse options (if set)
set_collapse(nthreads = 1L, remove = NULL, stable.algo = TRUE, sort = TRUE,
             digits = 2L, stub = TRUE, verbose = 1L, mask = NULL, na.rm = TRUE)

# For presentation: removing whitespace around labels
vlabels(GGDC10S) <- trimws(vlabels(GGDC10S))

# This is the benchmark function (janitor 2.2.0 is used)
bmark <- function(...) {
  bench::mark(..., min_time = 2, check = FALSE) |> janitor::clean_names() |>
  fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
  fmutate(expression = names(expression)) |> dapply(as.character) |> qDF()
}
@

%% -- Article metainformation (author, title, ...) -----------------------------

%% - \author{} with primary affiliation (and optionally ORCID link)
%% - \Plainauthor{} without affiliations
%% - Separate authors by \And or \AND (in \author) or by comma (in \Plainauthor).
%% - \AND starts a new line, \And does not.
\author{Sebastian Krantz~\orcidlink{0000-0001-6212-5229}\\Kiel Institute for the World Economy}
\Plainauthor{Sebastian Krantz}

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{\proglang{collapse}: Advanced and Fast Statistical Computing and Data Transformation in \proglang{R}} % \hphantom{aa}
\Plaintitle{collapse: Advanced and Fast Statistical Computing and Data Transformation in R}
\Shorttitle{\proglang{collapse}: Advanced and Fast Data Transformation in \proglang{R}}

%% - \Abstract{} almost as usual
\Abstract{
\pkg{collapse} is a large \proglang{C/C++}-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R}---at outstanding levels of performance and memory efficiency. It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular extensions (\class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, \class{pdata.frame}), enabling its seamless integration with large parts of the \proglang{R} ecosystem. This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples. A small benchmark demonstrates its computational performance.
%  This short article illustrates how to write a manuscript for the
%  \emph{Journal of Statistical Software} (JSS) using its {\LaTeX} style files.
%  Generally, we ask to follow JSS's style guide and FAQs precisely. Also,
%  it is recommended to keep the {\LaTeX} code as simple as possible,
%  i.e., avoid inclusion of packages/commands that are not necessary.
%  For outlining the typical structure of a JSS article some brief text snippets
%  are employed that have been inspired by \cite{Zeileis+Kleiber+Jackman:2008},
%  discussing count data regression in \proglang{R}. Editorial comments and
%  instructions are marked by vertical bars.
}

%% - \Keywords{} with LaTeX markup, at least one required
%% - \Plainkeywords{} without LaTeX markup (if necessary)
%% - Should be comma-separated and in sentence case.
\Keywords{statistical computing, vectorization, data manipulation and transformation, class-agnostic programming, summary statistics, \proglang{C/C++}, \proglang{R}}
\Plainkeywords{statistical computing, vectorization, data transformation and manipulation, class-agnostic programming, summary statistics, C/C++, R}

%% - \Address{} of at least one author
%% - May contain multiple affiliations for each author
%%   (in extra lines, separated by \emph{and}\\).
%% - May contain multiple authors for the same affiliation
%%   (in the same first line, separated by comma).
\Address{
  Sebastian Krantz\\
  Kiel Institute for the World Economy\\
  Haus Welt-Club\\
  D\"usternbrooker Weg 148\\
  24105 Kiel, Germany\\
  E-mail: \email{sebastian.krantz@ifw-kiel.de}
}

\begin{document}

%% -- Introduction -------------------------------------------------------------

%% - In principle "as usual".
%% - But should typically have some discussion of both _software_ and _methods_.
%% - Use \proglang{}, \pkg{}, and \code{} markup throughout the manuscript.
%% - If such markup is in (sub)section titles, a plain text version has to be
%%   added as well.
%% - All software mentioned should be properly \cite-d.
%% - All abbreviations should be introduced.
%% - Unless the expansions of abbreviations are proper names (like "Journal
%%   of Statistical Software" above) they should be in sentence case (like
%%   "generalized linear models" below).

\section[Introduction]{Introduction} \label{sec:intro}
%
\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/. Linecount (v2.0.19): R: 13,646, C: 18,594, C++: 9,844. \hphantom{aaa.}Exported namespace: 391 objects, of which 237 functions (excl. methods and shorthands), and 2 datasets.}
% Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
Core functionality includes a rich set of S3 generic (grouped and weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}). It also provides functions and classes for fully indexed (time-aware) computations on time series and panel data, advanced descriptive statistical tools, recursive tools to deal with nested data, and powerful data manipulation functions---such as vectorized and verbose hash-joins or fast aggregation and recast pivots. This functionality is supported by efficient algorithms for grouping, ordering, deduplication, and matching callable at \proglang{R} and \proglang{C} levels. The package also provides efficient data object conversions, functions for memory efficient R programming, such as (grouped) transformation and math by reference, and helpers to effectively deal with variable labels, attributes, and missing data. \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem---notably, \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}. It is globally and interactively configurable, which includes setting different defaults for key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the package namespace to mask equivalent but slower base \proglang{R} or \pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents. See Section~\ref{sec:glob_opt}.\vspace{-5mm}} \newline

Why combine all of these features in a package? The short answer is to make computations in \proglang{R} as flexible and powerful as possible. The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase the performance and memory efficiency of \proglang{R} programs;\footnote{Principally by avoiding \proglang{R}-level repetition such as applying \proglang{R} functions across columns/groups using a split-apply-combine logic, but also by avoiding object conversions and the need for certain classes to do certain things, such as converting matrices to \class{data.frame} or \class{data.table} just to compute statistics by groups.} and (3) to create a new foundation package for statistics and data transformation in \proglang{R} that implements many successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, in a stable, high performance, and broadly compatible manner.\footnote{Such ideas include \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), variable labels (\proglang{STATA}), recast pivots (\pkg{reshape(2)}), etc...} \newline

R already has a large and tested data manipulation and statistical computing ecosystem. Notably, the \pkg{tidyverse} \citep{rtidyverse} provides a consistent toolkit for data manipulation in R, centered around the \class{tibble} \citep{rtibble} object and tidy data principles \citep{rtidydata}. \pkg{data.table} \citep{rdatatable} provides an enhanced high-performance data frame with parsimonious data manipulation syntax. \pkg{sf} \citep{rsf} provides a data frame for spatial data and supporting functionality. \pkg{tsibble} \citep{rtsibble} and \pkg{xts} \citep{rxts} provide classes and operations for time series data, the former via an enhanced \class{tibble}, the latter through an efficient matrix-based class. Econometric packages like \pkg{plm} \citep{rplm} and \pkg{fixest} \citep{rfixest} also provide solutions to deal with panel data and irregularity in the time dimension. Packages like \pkg{matrixStats} \citep{rmatrixstats} and \pkg{Rfast} \citep{rfast} offer fast statistical calculations along the rows and columns of matrices as well as faster basic statistical procedures. \pkg{DescTools} \citep{rdesctools} provides a wide variety of descriptive statistics, including weighted versions. \pkg{survey} \citep{rsurvey} allows statistical computations on complex survey data. \pkg{labelled} \citep{rlabelled} provides tools to deal with labelled data. Packages like \pkg{tidyr} \citep{rtidyr}, \pkg{purrr} \citep{rpurrr} and \pkg{rrapply} \citep{rrapply} provide some functions to deal with nested data and messy structures. \newline

\pkg{collapse} relates to and integrates key elements from these projects. It offers \pkg{tidyverse}-like data manipulation at the speed and stability of \pkg{data.table} for any data frame-like object. It can turn any vector/matrix/data frame into a time-aware indexed series or frame and perform operations such as lagging, differencing, scaling or centering, encompassing and enhancing core manipulation functionality of \pkg{plm}, \pkg{fixest}, and \pkg{xts}. It also performs fast (grouped, weighted) statistical computations along the columns of matrix-like objects, complementing and enhancing \pkg{matrixStats} and \pkg{Rfast}. Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and \proglang{C}-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal. It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}. It provides novel recursive tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}. Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics.

In summary, \pkg{collapse} is a foundation package for statistical computing and data transformation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient. A significant benefit is that, rather than piecing together a fragmented ecosystem oriented at different classes and tasks, many core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages. This tends to result in \proglang{R} scripts that are shorter, more efficient, and more lightweight in dependencies. \newline
% TODO: Need examples of this

Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars}, and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy}, or \pkg{StatsBase.jl} \citep{jlstatsbase}. % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recast pivots and recursive operations on lists, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable. In short, it is very utile for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem. On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes. There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation (in most cases) is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any \proglang{Julia} function. However, as demonstrated in Section~\ref{ssec:vfat}, vectorized statistical functions can be combined to calculate more complex statistics in a vectorized way. \newline

The package has a built-in structured \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{documentation} facilitating its use. This documentation includes a central \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and supplementary topic pages which briefly describe related functionality. The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:
%
<<collapse_topics>>=
.COLLAPSE_TOPICS
help("collapse-documentation")
@
%
This article does not fully present \pkg{collapse}, but the following sections introduce its key components, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options. Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes. For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} (\href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}).
%
\section{Fast statistical functions} \label{sec:fast_stat_fun}
%
The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R} means that these operations are implemented using compiled C/C++ code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data. The functions basic syntax is
\begin{Code}
FUN(x, g = NULL, w = NULL, TRA = NULL, na.rm = TRUE, use.g.names = TRUE, ...)
\end{Code}
with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - sampling weights (only some functions), and \code{TRA} - transformation of \code{x}. The following examples using \fct{fmean} demonstrate their basic usage on the familiar \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/iris}{\code{iris}} dataset, recording 50 measurements of 4 variables for 3 species of iris flowers. All examples support weights (\code{w}), and \fct{fmean} can also be multithreaded across columns (\code{nthreads}).\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions, as detailed in the respective function documentation. The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes. Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods. % Users can also directly call the basis methods in case S3 dispatch does not yield the intended outcome. For example, \code{fmean.default(EuStockMarkets)} computes the mean of the entire matrix. %, and attributes are preserved as much as possible ()
<<faststatfun>>=
fmean(iris$Sepal.Length)
fmean(iris[1:4])
identical(fmean(iris[1:4]), fmean(as.matrix(iris[1:4])))
fmean(iris$Sepal.Length, g = iris$Species)
fmean(iris[1:4], g = iris$Species, nthreads = 4)
fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill")[1:10]
@
%
\subsection{Transformations}
The final example expands the mean vector to full length---like \code{ave(x, g)} but much faster. The \code{TRA} argument toggles (grouped) replacing and sweeping operations (by reference), generalizing \code{sweep(x, 2, STATS = fmean(x))}.\footnote{The \code{TRA} argument internally calls \fct{TRA}: \code{TRA(x, STATS, FUN = "-", g = NULL, set = FALSE, ...)}.} %Its syntax is
%\begin{Code}
%TRA(x, STATS, FUN = "-", g = NULL, set = FALSE, ...)
%\end{Code}
%where \code{STATS} is a vector/matrix/data.frame of statistics used to transform \code{x}.
Table~\ref{tab:TRA} lists the 11 possible \code{TRA} operations. % to transform \code{x} with the statistics.
%
\begin{table}[H]
\resizebox{\textwidth}{!}{
  \begin{tabular}{lll}
  \emph{String} && \emph{Description}  \\
  \code{"replace\_na"/"na"}   && replace missing values in \code{x} by \code{STATS} \\
  \code{"replace\_fill"/"fill"}   && replace data and missing values in \code{x} by \code{STATS} \\
  \code{"replace"} && replace data by \code{STATS} but preserve missing values in \code{x} \\
  \code{"-"}   && subtract \code{STATS} (center) \\
  \code{"-+"}  && subtract \code{STATS} and add overall average statistic \\
  \code{"/"}   && divide by \code{STATS} (scale) \\
  \code{"\%"}     && compute percentages (divide and multiply by 100) \\
  \code{"+"} && add \code{STATS} \\
  \code{"*"} && multiply by \code{STATS} \\
  \code{"\%\%"} && modulus (remainder from division by \code{STATS}) \\
  \code{"-\%\%"} && subtract modulus (make data divisible by \code{STATS})
  \end{tabular}
}
\caption{\label{tab:TRA} Available \code{TRA} argument choices.}
\end{table}
%
%\code{TRA()} is called internally in the \emph{Fast Statistical Functions}, the \code{TRA} argument is passed to \code{FUN}. Thus \code{fmean(x, g, w, TRA = "-")} is equivalent to \code{TRA(x, fmean(x, g, w), "-", g)}.
Additionally, a \code{set} argument can be passed to \emph{Fast Statistical Functions} to toggle transformation by reference, e.g., \code{fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill", set = TRUE)} would modify \code{Sepal.Length} in-place and return the result invisibly. \newline

Having grouping and data transformation functionality directly built into generic statistical functions facilitates and speeds up many common operations. Take for example this generated sector-level trade dataset of export values (v) by country (c), sector (s), and year (y).\footnote{\label{fn:tfm}\fct{tfm} abbreviates \fct{ftransform}---a faster equivalent of \fct{transform}. \fct{ss} replaces \code{[.data.frame}.}

<<exports>>=
set.seed(101)
exports <- expand.grid(y=1:10, c=paste0("c",1:10), s=paste0("s",1:10)) |>
  tfm(v = abs(rnorm(1e3))) |> colorder(c, s) |> ss(-sample.int(1e3, 500))
@
Like any real trade dataset, it is unbalanced---\code{ss(-sample.int(1e3, 500))} randomly removes 500 rows. Suppose we wanted to extract the latest trade within the last two years.
<<exports_2>>=
latest <- fsubset(exports, y >= 8 & y == fmax(y, list(c, s), "fill"), -y)
@
Below, I compute how many products different countries have exported in the last two years
<<prod>>=
with(latest, fndistinct(s, c))
@
and this computes \citet{balassa1965tariff}'s index of Revealed Comparative Advantage, defined as the share of a sector in country exports divided by the share of the sector in world exports.
<<RCA>>=
with(latest, fsum(v, c, TRA = "/") / fsum(v, s, TRA = "/"))[1:10]
@
More complex use cases are frequent in my work. For example, I recently combined multiple spatial datasets on points of interest (POIs). In the face of significant duplicates and problems matching POIs directly across datasets, I decided to keep the richest source for each location and POI type. After creating POI confidence, location, and type indicators comparable across sources (datasets), my deduplication expression was \code{fsubset(data\_long, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (i.e., richest) source by location and type.

\subsection{Grouping objects and optimization} \label{ssec:gopt}
%
Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, for repeated operations the cost of grouping can be minimized by using factors (see \code{?qF}) or \class{GRP} objects as inputs. The latter contain all information \pkg{collapse}'s statistical functions may require to operate across groups. They can be created with \code{GRP()}. Its basic syntax is
\begin{Code}
GRP(X, by = NULL, sort = TRUE, return.groups = TRUE, method = "auto", ...)
\end{Code}
 Below, I create a \class{GRP} object from the included World Development Dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}). The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.
%
<<GRP>>=
str(g <- GRP(wlddev, ~ income + OECD))
@
%
\class{GRP} objects make grouped statistical computations in \pkg{collapse} fully programmable, e.g., I can employ the object with the \emph{Fast Statistical Functions} and some utilities\footnote{\fct{add\_vars} is a fast \fct{cbind.data.frame} which also has an assignment method, and \fct{get\_vars} enables fast and secure extraction of data frame columns.} to efficiently aggregate GDP per capita, life expectancy, and country name, with population weights.
%
<<gcomp>>=
add_vars(g$groups,
 get_vars(wlddev, "country") |> fmode(g, wlddev$POP, use = FALSE),
 get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = F))
@
For advanced data aggregation, \pkg{collapse} also provides a convenience function, \fct{collap}, which, by default, uses \fct{fmean} for numeric, \fct{fmode} for categorical, and \fct{fsum} for weight columns, and preserves their order. The equivalent expression using this function would be
%
<<collap>>=
collap(wlddev, country + PCGDP + LIFEEX ~ income + OECD, w = ~ POP)
@
%
Similarly, data can be transformed, here using \fct{fmean} to center the data by country to level differences in average economic status, adding back the overall mean across countries.\footnote{\fct{add\_stub} adds a prefix (or suffix if \code{pre = FALSE}) to columns ($\to$ \code{center\_PCGDP} and \code{center\_LIFEEX}).}
%
<<gtrans>>=
add_vars(wlddev) <- get_vars(wlddev, c("PCGDP", "LIFEEX")) |>
    fmean(wlddev$iso3c, TRA = "-+") |> add_stub("center_")
@
%
For (higher-dimensional) centering, \pkg{collapse} also has specialized function(s) \fct{f[hd]within} with additional options, and \fct{fscale} supports various scaling and centering operations.

Exempting \fct{collap}, these examples may seem bulky for quick analysis, but a robust low-level API is very useful for package development, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}. I also wrote a blog post on \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{aggregating survey data using \pkg{collapse}}, which showcases more aspects of the \fct{collap} function using real census data. Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices. For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} I aggregate global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}.\footnote{A recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case) - see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
%On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another recent example involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}. Thus each iteration of the numerical routine raises the vector \textbf{x} to a different power ($a$), sums it in 1 million groups ($j$) to generate \textbf{y}, and computes the Euclidean distance to \textbf{z} (using \code{collapse::fdist}). Without grouping objects and vectorization, this would have been difficult to handle within reasonable computing times (of a few seconds on the M1).}
%
\section{Integration with data manipulation functions} \label{sec:integration}
%
\pkg{collapse} also provides a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{fast data manipulation functions} akin to base \proglang{R} and \pkg{tidyverse} functions, including \fct{fselect}, \fct{fsubset}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{ftransform}, \fct{fmutate}, \fct{across}, \fct{frename}, \fct{fcount}, etc. These are integrated with the \emph{Fast Statistical Functions} to enable vectorized statistical operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow. For example, the following code aggregates the \code{wlddev} data by income group for years post 2015 (to smooth volatility), with population weights.
%
<<fsummarise>>=
wlddev |> fsubset(year >= 2015) |> fgroup_by(income) |>
  fsummarise(country = fmode(country, POP),
             across(c(PCGDP, LIFEEX, GINI), fmean, POP))
@
%
This code is very fast because data does not need to be split by groups. Under the hood it is principally a syntax translation to the low-level API introduced above.\footnote{\fct{fgroup\_by} creates a \class{GRP} object from the \code{income} column, attaches it as an attribute, and \fct{fsummarise}/\fct{across} fetches it and passes it to the \code{g} arguments of the \emph{Fast Statistical Functions} set as a keyword argument (and sets \code{use.g.names = FALSE}). Thus, \code{w} becomes the second positional argument. Since \fct{fmean} is S3 generic, \fct{across} directly invokes \fct{fmean.data.frame} on the subset of columns. \vspace{-10mm}} \emph{Fast Statistical Functions} also have a method for grouped data, so \fct{fsummarise} is not always needed. %The following example calculates weighted group means. By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::group\_by}. The latter requires an additional \proglang{C} routine to convert the \pkg{dplyr} grouping object to a \class{GRP} object, and is thus less efficient.}
%
<<fsummarise_2>>=
wlddev |> fsubset(year >= 2015, income, PCGDP:GINI, POP) |>
  fgroup_by(income) |> fmean(POP, keep.w = FALSE)
@
%
\subsection{Vectorizations for advanced tasks} \label{ssec:vfat}
%
\fct{fsummarise} and \fct{fmutate} can also evaluate arbitrary functions in the classical (split-apply-combine) way and handle more complex expressions involving multiple columns and/or functions. However, using any \emph{Fast Statistical Function} causes the whole expression to be vectorized, i.e., evaluated only once and not for every group. This eager vectorization approach enables efficient grouped calculation of more complex statistics. The example below forecasts the exports for each country-sector via linear regression (\code{v ~ y}) in a vectorized way.
%
<<glinreg>>=
exports |> fgroup_by(c, s) |> fmutate(dmy = fwithin(y)) |>
  fsummarise(v_10 = flast(v), beta = fsum(v, dmy) %/=% fsum(dmy, dmy)) |>
  fmutate(v_11 = v_10 + beta, v_12 = v_11 + beta, beta = NULL) |> head(4)
@
%
The expression \code{fsum(v, dmy) \%/=\% fsum(dmy, dmy)} amounts to \code{cov(v, y)/var(y)}, but is both vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation for the final result. I do not recommend forecasting trade in this way, rather, this example is inspired by a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{bog post} where I forecasted high-resolution (1km2) population estimates for South Africa. The data, taken from WorldPop, was available for the years 2014-2020, and I needed estimates for 2021-2022. Linear regression was sensible, and using the above expression I was able to run 1.6 million regressions and obtain 2 forecasts in 0.26 seconds on a laptop. Another neat example from the community, shared by Andrew Ghazi in a \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post}, vectorizes an expression to compute the $p$~value, \code{2 * pt(abs(fmean(x) * sqrt(6) / fsd(x)), 5, lower.tail = FALSE)}, across 300k groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}. The eager vectorization approach of \pkg{collapse} here replaces \code{fmean(x)} and \code{fsd(x)} by their grouped versions and evaluates the entire expression once rather than 300k times as in \pkg{dplyr}. \newline

\pkg{collapse} also vectorizes advanced statistics. The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} calculates weighted quantiles in a theoretically consistent way by applying the probability measure to the sum of weights to create a target sum and cumulatively summing weights to find order statistics which are then combined following \citet{hyndman1996sample}. See \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for details.}
%
<<wstats>>=
wlddev |> fsubset(is.finite(POP)) |> fgroup_by(income, OECD) |>
  fmutate(o = radixorder(GRPid(), LIFEEX)) |>
  fsummarise(min = fmin(LIFEEX),
             Q1 = fnth(LIFEEX, 0.25, POP, o = o, ties = "q8"),
             mean = fmean(LIFEEX, POP),
             median = fmedian(LIFEEX, POP, o = o, ties = "q8"),
             Q3 = fnth(LIFEEX, 0.75, POP, o = o, ties = "q8"),
             max = fmax(LIFEEX))
@
%
Weighted quantiles have a sub-column parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \fct{fnth}/\fct{fmedian}/\fct{fmode} (default 1).} and, as shown above, can also harness an (optional) optimization utilizing an overall ordering vector (combining groups and the data column) to avoid repeated partial sorting of the same elements within groups.
%
\section{Time series and panel data} \label{sec:ts_ps}
%
\pkg{collapse} also provides a flexible high-performance architecture to perform (time aware) computations on time series and panel series. In particular, the user can either apply time series and panel data transformations without any classes by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion, or use \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation for worry-free application in many contexts. Table~\ref{tab:TSfun} compactly summarizes \pkg{collapse}'s architecture for time series and panel data.
%
\begin{table}[h]
\begin{tabular}{p{\textwidth}}
\emph{Classes, constructors and utilities} \\
\code{findex\_by(), findex(), unindex(), reindex(), timeid(), is\_irregular(), to\_plm()} $+$ S3 methods for \class{indexed\_frame}, \class{indexed\_series} and \class{index\_df} \\\\
\emph{Core time-based functions} \\
\code{flag(), fdiff(), fgrowth(), fcumsum(), psmat()} \\ \code{psacf(), pspacf(), psccf()} \\\\
\emph{Data transformation functions with supporting methods} \\
\code{fscale(), f[hd]between(), f[hd]within()} \\\\
\emph{Data manipulation functions with supporting methods} \\
\code{fsubset(), funique(), roworder[v]()} (internal), \code{na\_omit()} (internal) \\\\
\emph{Summary functions with supporting methods} \\
\code{varying(), qsu()} \\
\end{tabular}
\caption{\label{tab:TSfun} Time series and panel data architecture.}
\end{table}
%
\subsection{Ad-hoc computations}
%
Time series functions such as \fct{fgrowth} (to compute growth rates) are S3 generic and can be applied to most time series classes. In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation. If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years. Users need to manually call \code{timeid()} on plain numeric vectors with decimals to yield an appropriate integer representation. If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id. For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}.} But first, a basic example:
%
<<ts_example>>=
fgrowth(airmiles) |> round(2)
@
The following extracts one sector from the \code{exports} dataset generated above, creating an irregular time series missing the 3rd and 6th period.\footnote{\code{\%=\%} is an infix operator for the \fct{massign} function in \pkg{collapse} which is a multivariate version of \fct{assign}.} Indexation using the \code{t} argument allows for correct (time-aware) computations on this context without 'expanding' the data.
%
<<ts_example_2>>=
.c(y, v) %=% fsubset(exports, c == "c1" & s == "s7", -c, -s)
print(y)
fgrowth(v, t = y) |> round(2)
fgrowth(v, -1:3, t = y) |> head(4)
@
%
Functions \code{flag()/fdiff()/fgrowth()} also have shorthands \code{L()/D()/G()} which both facilitate their use inside formulas and provide an enhanced data frame interface for convenient ad-hoc computations. With panel data, \code{t} can be omitted, but this requires sorted data with consecutive groups.\footnote{This is because a group-lag is computed in a single pass, requiring all group elements to be consecutive.} Below, I demonstrate two ways to compute a sequence of lagged growth rates using both the \code{G()} operator and the \code{tfm()} function - a shorthand for \code{ftransform()}.\footnote{See also Footnote~\ref{fn:tfm}. A number of key functions in \pkg{collapse} have syntactic shorthands. The \code{list(v = v)} is needed here to prevent \fct{fgrowth} from creating a matrix with the growth rates---the \class{list} method applies.}
%
<<example_growth>>=
G(exports, -1:2, by = v ~ c + s, t = ~ y) |> head(3)
tfm(exports, fgrowth(list(v = v), -1:2, g = list(c, s), t = y)) |> head(3)
@
%
These functions and operators are also integrated with \fct{fgroup\_by} and \fct{fmutate} for vectorized computations. However, using ad-hoc grouping is always more efficient.
%
<<example_growth_continued>>=
A <- exports |> fgroup_by(c, s) |> fmutate(gv = G(v, t = y)) |> fungroup()
head(B <- exports |> fmutate(gv = G(v, g = list(c, s), t = y)), 4)
identical(A, B)
@
%
% Similarly, functions to scale, center, and average data have groups (\code{g}) and also weights (\code{w}) arguments, and corresponding operators \code{STD(),[HD]W(),[HD]B()} to facilitate ad-hoc transformations. Below, two ways to perform grouped scaling are demonstrated. The operator version is slightly faster and renames the transformed columns by default (\code{stub = TRUE}).
% %
% <<example_scale_cener>>=
% iris |> fgroup_by(Species) |> fscale() |> head(2)
% STD(iris, ~ Species) |> head(2)
% @
% The following example demonstrates a fixed-effects regression \`a la \citet{mundlak1978pooling}.
% <<example_reg>>=
% lm(mpg ~ carb + B(carb, cyl), data = mtcars) |> coef()
% @
% \pkg{collapse} also offers higher-dimensional between and within transformations, powered by \proglang{C++} code conditionally imported (and accessed directly) from \pkg{fixest}. The following detrends GDP per Capita and Life Expectancy at Birth using country-specific cubic polynomials.
% <<example_HD>>=
% HDW(wlddev, PCGDP + LIFEEX ~ iso3c * poly(year, 3), stub = F) |> head(2)
% @
%
\subsection{Indexed series and frames}
%
For more complex use cases, indexation is convenient. \pkg{collapse} supports \pkg{plm}'s \class{pseries} and \class{pdata.frame} classes through dedicated methods. Flexibility and performance considerations lead
to the creation of new classes \class{indexes\_series} and \class{indexed\_frame} which inherit from the former. Any data frame-like object can become an \class{indexed\_frame} and function as usual for other operations. The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}. Their basic syntax is:
%
\begin{Code}
data_ix <- findex_by(data, id1, ..., time)
data_ix$indexed_series; with(data, indexed_series)
index_df <- findex(data_ix)
\end{Code}
%
Data can be indexed using one or more indexing variables. Unlike \class{pdata.frame}, an \\ \class{indexed\_frame} is a deeply indexed structure---every series inside the frame is already an \class{indexes\_series}. A comprehensive set of \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()}, ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.). % A basic demonstration with World Bank panel data showcases the flexibility of these classes.
Indexation can be undone using \code{unindex()} and redone with \code{reindex()} and a suitable \class{index\_df}. \class{indexes\_series} can be atomic vectors or matrices (including objects such as \class{ts} or \class{xts}) and can be created directly using \code{reindex()}.
%
\begin{Code}
data <- unindex(data_ix)
data_ix <- reindex(data, index = index_df)
indexed_series <- reindex(vec/mat, index = vec/index_df)
\end{Code}
%
An example using the \code{exports} data follows:
<<indexing>>=
exportsi <- exports |> findex_by(c, s, y)
exportsi |> G() |> print(max = 15)
exportsi |> findex() |> print(2)
@
The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
<<indexing_2>>=
vi <- exportsi$v; str(vi, width = 70, strict = "cut")
is_irregular(vi)
vi |> psmat() |> head(3)
fdiff(vi) |> psmat() |> head(3)
@
\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work in arbitrary data masking environments.
<<indexing_3>>=
settransform(exportsi, v_ld = Dlog(v))
lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
@
It is worth highlighting that the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions.\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.} Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures or native handling of irregularity in basic operations. \class{indexed\_series} and \class{indexed\_frame}, on the other hand, work 'anywhere', and can be superimposed on any suitable object, as long as \pkg{collapse}'s functions (\code{flag()/L()} etc.) are used to perform time-based computations. \newline

Indexed series/frames also support transformation such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}. Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series. The \class{index\_df} attached to these objects can also be used with other general tools such as \code{collapse::BY()} to perform grouped computations using 3rd-party functions. An example of calculating a 5-year rolling average is given below (\fct{ix} abbreviates \fct{findex}).
%
<<rollmean>>=
BY(vi, ix(vi)$c.s, data.table::frollmean, 5) |> head(10)
@
% Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
%
\section{Table joins and pivots} \label{sec:join_pivot}
%
Among all \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and 'recast' pivots. Both implementations provide outstanding computational performance, syntax, and memory efficiency.
%
\subsection{Joins}
%
Compared to commercial software such as \proglang{STATA}, the implementation of joins in most open-source software, including \proglang{R}, provides no information on how many records were joined from both tables. This often provokes manual efforts to validate the join operation. \code{collapse::join} provides a rich set of options to understand table join operations. Its syntax is:
\begin{Code}
join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
  multiple = FALSE, sort = FALSE, keep.col.order = TRUE,
  drop.dup.cols = FALSE, verbose = 1, column = NULL, attr = NULL, ...)
\end{Code}
It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}), i.e., it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases. By default (\code{verbose = 1}), it prints information about the join operation and number of records joined. To demonstrate \fct{join}, I generate a small database for a bachelor in economics curriculum. It has a \code{teacher} table of 4 teachers (\code{id}: PK) and a linked (\code{id}: FK) \code{course} table of 5 courses.
%
<<joins>>=
teacher <- data.frame(id = 1:4, names = c("John", "Jane", "Bob", "Carl"),
  age = c(35, 32, 42, 67), subject = c("Math", "Econ", "Stats", "Trade"))
course <- data.frame(id = c(1, 2, 2, 3, 5), semester = c(1, 1, 2, 1, 2),
  course = c("Math I", "Microecon", "Macroecon", "Stats I", "History"))
join(teacher, course, on = "id")
@
%
Users can request the generation of a \code{.join} column (\code{column = "name"/TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, to indicate the origin of records in the joined table---useful on a full join.
%
<<joins_2>>=
join(teacher, course, how = "full", multiple = TRUE, column = TRUE)
@
%
An alternative is to request an attribute (\code{attr = "name"/TRUE}) that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}).
\begin{Code}
R> join(teacher, course, multiple = TRUE, attr = "jn") |> attr("jn") |> str()
\end{Code}
<<joins_3, echo=FALSE>>=
join(teacher, course, multiple = TRUE, attr = "jn") |> attr("jn") |> str(strict.width = "cut", width = 70)
@
Users can also invoke the \code{validate} argument to examine the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.
%
<<joins_4>>=
join(teacher, course, on = "id", validate = "1:1") |>
  tryCatch(error = function(e) strwrap(e) |> cat(sep = "\n"))
@
%
A few further particularities are worth highlighting. First, \fct{join} is class-agnostic and preserves the attributes of \code{x} (any list-based object). It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"}, or \code{"anti"} join). This demonstrates the latter two:
<<joins_5>>=
for (h in c("semi", "anti")) join(teacher, course, how = h) |> print()
@
By default (\code{sort = FALSE}), the order of rows in \code{x} is preserved. Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join (using \fct{fmatch}) followed by sorting, particularly if the data is already sorted on the keys. } The join relationship is indicated inside the \code{<>} as the number of records joined from each table divided by the number of unique matches. % In multi-match settings, this will be reflected by few records from \code{y} being used.
<<joins_6>>=
course$names <- teacher$names[course$id]
join(teacher, course, on = "id", how = "inner", multiple = TRUE)
@
As shown above, \fct{join}'s handling of duplicate columns in both tables is rather special. By default (\code{suffix = NULL}), \fct{join} extracts the name of the \code{y} table and appends \code{y}-columns with it. \code{x}-columns are not renamed. This is congruent to the principle of adding columns to \code{x} and altering this table as little as possible. Alternatively, option \code{drop.dup.cols = "x"/"y"} can be used to simply remove duplicate columns from \code{x} or \code{y} before the join operation.
%
<<joins_7>>=
join(teacher, course, on = "id", multiple = TRUE, drop.dup.cols = "y")
@
%
A final noteworthy feature is that \fct{fmatch} has a built-in overidentification check, which warns if more key columns than necessary to identify the records are provided. This check only triggers with 3+ id columns as for efficiency reasons the first two ids are jointly hashed.

\fct{join} is thus a highly efficient, versatile, and verbose implementation of table joins for \proglang{R}.
%
\subsection{Pivots}
%
The reshaping/pivoting functionality of both commercial and open-source software is also presently unsatisfactory for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data. Such datasets can presently only be reshaped by losing these labels or additional manual efforts to retain them. Modern \proglang{R} packages also offer different reshaping functions, such as \fct{data.table::melt}/\fct{tidyr::pivot\_longer} to combine columns and \fct{data.table::dcast}/\fct{tidyr::pivot\_wider} to expand them, requiring users to learn both. Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two reshaping functions, incurring a high cost in terms of both syntax and memory efficiency. \newline

\code{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast'\newpage

pivots, and supports complex labelled data without loss of information. Its basic syntax is:
%
\begin{Code}
pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
  how = "longer", na.rm = FALSE, check.dups = FALSE, ...)
\end{Code}
%
The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Databas (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}), providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America. While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes. \fct{namlab} provides a compact overview of variable names and labels stored in \code{attr(column, "label")}, with (optional) additional information/statistics.}
%
<<pivots>>=
data <- GGDC10S |>
  fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
  fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
@
%
To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}. If \code{labels = "name"} is specified, variable labels (stored in \code{attr(column, "label")}) are saved in an additional column. In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
%
<<pivot_longer>>=
head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
                 labels = "Sector", how = "longer"))
@
%
\fct{pivot} only requires essential information and intelligently guesses the rest. For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}. An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", "wider")}, with \code{dl} the long data. \newline

The second option is a wider pivot with \code{how = "wider"}. Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{If multiple columns are selected, they are combined using \code{"\_"} for names and \code{" - "} for labels.} Note below how the labels are combined with existing labels such that also this operation is without loss of information. It is, however, a destructive operation---with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it. Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
%
<<pivot_wider>>=
head(dw <- pivot(data, c("Country", "Year"), names = "Variable",
                 labels = "Label", how = "w"))
namlab(dw)
@
%
For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}. Again, taking along labels is entirely optional---omitting either the labels-list's \code{from} or \code{to} element will omit the respective operation.
%
<<pivot_recast>>=
head(dr <- pivot(data, c("Country", "Year"),
             names = list(from = "Variable", to = "Sectorcode"),
             labels = list(from = "Label", to = "Sector"), how = "r"))
vlabels(dr)[3:6]
@
%
This (\code{dr}) is the tidy format \citep{rtidydata} where each variable is a separate column. It is analytically more useful, e.g., to compute labor productivity as \code{settransform(dr, LP = VA / EMP)}, or to estimate a panel-regression with sector fixed-effects. The recast pivot is thus a natural operation to change data representations. As with the other pivots, it preserves all information and can be reversed by simply swapping the contents of the \code{from} and \code{to} keywords. \newline

\fct{pivot} also supports fast aggregation pivots, the default being \code{FUN = "last"}, which simply overwrites values in appearance order if the combination of \code{ids} and \code{names} does not fully identify the data. The latter can be checked with \code{check.dups = TRUE}. There are a small number of internal functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}. These carry out computations 'on the fly' and are thus extremely fast. \fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but require a deep copy of the columns aggregated which is avoided using the internal functions. The following example performs aggregation across years with the internal mean function during a recast pivot.
<<pivot_agg>>=
head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"), how = "r",
     names = list(from = "Variable", to = "Sectorcode"),
     labels = list(from = "Label", to = "Sector"), FUN = "mean"))
@
More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}. Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.
%
\section{List processing} \label{sec:list_proc}
%
Often in programming, nested structures are needed. A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model predictions and performance statistics) in a list. Nested data is also often the result of web scraping or web APIs. A typical use case in development involves serving different data according to user choices. Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data. \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit. \newline

To create nested data, \fct{rsplit} generalizes \fct{split} and (recursively) splits up data frame-like objects into (nested) lists. For example, we can split the \code{GGDC10S} data by country and\newpage variable, such that, e.g., agricultural employment in Argentina can be accessed as:\footnote{If a nested structure is not needed, \code{flatten = TRUE} lets \fct{rsplit} operate like a faster version of \fct{split}.}
%
<<list_proc>>=
dl <- GGDC10S |> rsplit( ~ Country + Variable)
dl$ARG$EMP$AGR[1:12]
@
%
This is a convenient data representation for \emph{Shiny Apps} where we can let the user choose data (e.g., \code{dl[[input$country]][[input$variable]][[input$sector]]}) without expensive subsetting operations. As mentioned, such data representation can also be the result of an API call parsing JSON or a nested loop or \fct{lapply} call. Below, I write a nested loop running a regression of agriculture on mining and manuacturing output/employment:
<<list_proc_2>>=
result <- list()
for (country in c("ARG", "BRA", "CHL")) {
  for (variable in c("EMP", "VA")) {
    m <- lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year,
            data = dl[[country]][[variable]])
    result[[country]][[variable]] <- list(model = m, BIC = BIC(m),
                                          summary = summary(m))
  }
}
@
This programming may not be ideal for this particular use case as I could have used data.frame-based tools and saved the result in a column.\footnote{E.g., \code{GGDC10S |> fgroup\_by(Country, Variable) |> fsummarise(result = my\_fun(lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year)))} with \code{my\_fun <- function(m) list(list(m, BIC(m), summary(m)))}.}
However, there are limits to data.frame-based workflows. For example, I recently trained a complex ML model for different variables and parameters, which involved loading a different dataset at each iteration. Loops are useful in such cases, and lists a natural vehicle to structure complex outputs. The main issue with nested lists is that they are complex to query. What if we want to know just the $R^2$ of these 6 models? We would need to use, e.g., \code{result$ARG$EMP$summary$r.squared} for each model.\newline

This nested list-access problem was the main reason for creating \fct{get\_elem}: an efficient recursive list-filtering function which, by default, simplifies the list tree as much as possible.
%
<<list_proc_3>>=
str(r_sq_l <- result |> get_elem("r.squared"))
rowbind(r_sq_l, idcol = "Country", return = "data.frame")
@
%
% Thus \code{get\_elem("R2")} returns the same nested list with the $R^2$ in all final nodes. Then \code{rowbind()} could be used to create a data frame:
Note how the \code{"summary"} branch was eliminated since it is common to all final nodes; \code{result |> get\_elem("r.squared", keep.tree = TRUE)} could have been used to retain it. \fct{rowbind} then efficiently combines lists of lists. We can also apply \fct{t\_list} to turn the list inside-out.
<<list_proc_3.5>>=
r_sq_l |> t_list() |> rowbind(idcol = "Variable", return = "data.frame")
@
\fct{rowbind} is limited if \fct{get\_elem} returns a more nested or asymmetric list, potentially with vectors/arrays in the final nodes. Suppose we want to extract the coefficient matrices:
<<list_proc_4>>=
result$ARG$EMP$summary$coefficients
@
For such cases, I created \fct{unlist2d} as a complete recursive generalization of \fct{unlist}. It creates a \class{data.frame} (or \class{data.table}) representation of any nested list using recursive row-binding and coercion operations while generating (optional) id variables representing the list tree and (optionally) saving row names of matrices or data frames. In the present example
%
<<list_proc_5>>=
result |> get_elem("coefficients") |> get_elem(is.matrix) |>
  unlist2d(idcols = c("Country", "Variable"),
           row.names = "Covariate") |> head(3)
@
%
where \code{get\_elem(is.matrix)} is needed because the models also contain \code{"coefficients"}. \newline

This exemplifies the power of these tools to create, query, and combine nested data in very general ways, and with many applications.
  % I use it extensively to fetch and compare different statistics from (sometimes large) nested lists with different statistical model outputs.
Further useful functions include \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, and \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes.
%
\section{Summary statistics} \label{sec:summ_stat}
%
\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious and powerful toolkit to examine complex datasets. A particular focus has been on providing tools for examining longitudinal (panel) data. Recall the indexed World Development Panel (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{ssec:gopt}. The function \fct{varying} can be used to examine which of these variables are time-varying.
%
<<varying>>=
varying(wlddev, ~ iso3c)
@
%
A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time. Using the (de-)meaning functions supporting \class{indexed\_series} from Table~\ref{tab:TSfun}, this is easily demonstrated.
%
<<pdec>>=
LIFEEXi <- reindex(wlddev$LIFEEX, wlddev$iso3c)
all.equal(fvar(LIFEEXi), fvar(fbetween(LIFEEXi)) + fvar(fwithin(LIFEEXi)))
@
%
The \fct{qsu} (quick-summary) function provides an efficient method to (approximately) compute this decomposition, considering the group-means instead of the between transformation\footnote{This is more efficient and equal to using the between transformation if the panel is balanced.} and adding the mean back to the within transformation to preserve the scale of the data.
%
<<qsu_1>>=
qsu(LIFEEXi)
@
%
The decomposition above implies more variation in life expectancy between countries than within countries over time. It can also be computed for different subgroups, such as OECD members and non-members, and with sampling weights, such as population.
%
<<qsu_2>>=
qsu(LIFEEXi, g = wlddev$OECD, w = wlddev$POP) |> aperm()
@
%
The output shows that the variation in life expectancy is significantly larger for non-OECD countries. For the latter the between- and within-country variation is approximately equal.\footnote{\fct{qsu} also has a convenient formula interface to perform these transformations in an ad-hoc fashion, e.g., the above can be obtained using \code{qsu(wlddev, LIFEEX $\sim$ OECD, $\sim$ iso3c, $\sim$ POP)}, without prior indexation.} For greater detail, \fct{descr} provides a rich (grouped, weighted) statistical description.
%
<<descr>>=
wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
wlda15 |> descr(income + LIFEEX ~ OECD)
@

%
While \fct{descr} does not support panel-variance decompositions like \fct{qsu}, it also computes detailed (grouped, weighted) frequency tables for categorical data and is thus very utile with mixed-type data. A \code{stepwise} argument toggles describing one variable at a time, allowing users to naturally 'click-through' a large dataset rather than printing a massive output to the console. The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples. Both \fct{qsu} and \fct{descr} provide an \code{as.data.frame()} method for efficient tidying and further analysis. \newline

A final noteworthy function from \pkg{collapse}'s descriptive statistics toolkit is \fct{qtab}, an enhanced drop-in replacement for \fct{table}. It is enhanced both in a statistical and a computational sense, providing a remarkable performance boost, an option (\code{sort = FALSE}) to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights---vectorized when using \emph{Fast Statistical Functions}, as demonstrated below.
%
<<qtab>>=
wlda15 |> with(qtab(OECD, income))
@
This shows the total population (latest post-2015 estimates) in millions.
<<qtab_2>>=
wlda15 |> with(qtab(OECD, income, w = POP) / 1e6)
@
This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only called once).
<<qtab_3>>=
wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean))
@
Finally, this calculates a population-weighted average of life expectancy in each group.
<<qtab_4>>=
wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean,
                    wFUN.args = list(w = POP)))
@
%
\class{qtab} objects inherit the \class{table} class, thus all \class{table} methods apply. Apart from the above functions, \pkg{collapse} also provides functions \fct{pwcor}, \fct{pwcov}, and \fct{pwnobs} for convenient (pairwise, weighted) correlations, covariances, and observations counts, respectively.
%
\section{Global options} \label{sec:glob_opt}
%
\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function. These options are saved in an internal environment called \code{.op}. Its contents can be accessed using \fct{get\_collapse}. \newline

The current set of options comprises the default behavior for missing values (\code{na.rm} arguments in all statistical functions and algorithms), sorted grouping (\code{sort}), multithreading and algorithmic optimizations (\code{nthreads}, \code{stable.algo}), presentational settings (\code{stub}, \code{digits}, \code{verbose}), and, surpassing all else, the package namespace itself (\code{mask}, \code{remove}). \newline

As evident from previous sections, \pkg{collapse} provides performance-improved or otherwise enhanced versions of functionality already present in base R (like the \emph{Fast Statistical Functions}, \fct{funique}, \fct{fmatch}, \fct{fsubset}, \fct{ftransform}, etc.) and other packages (esp. \pkg{dplyr} \citep{rdplyr}: \fct{fselect}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, etc.). The objective of being namespace compatible warrants such a naming convention, but this has a syntactical cost, particularly when \pkg{collapse} is the primary data manipulation package. \newline

To reduce this cost, \pkg{collapse}'s \code{mask} option allows masking existing \proglang{R} functions with the faster \pkg{collapse} versions by creating additional functions in the namespace and instantly exporting them. All \pkg{collapse} functions starting with 'f' can be passed to the option (with or without the 'f'), e.g., \code{set\_collapse(mask = c("subset", "transform"))} creates \code{subset <- fsubset} and \code{transform <- ftransform} and exports them. Special functions are \code{"n"}, \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} (for use in \code{(f)summarise}/\code{(f)mutate}), \code{table <- qtab}, and replace \code{\%in\%} with a fast version using \fct{fmatch}, respectively. There are also several \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}. The most powerful of these is \code{"all"}, which masks all f-functions and special functions, as shown below.
%
\begin{Code}
set_collapse(mask = "all")
wlddev |> subset(year >= 1990 & is.finite(GINI)) |>
  group_by(year) |>
  summarise(n = n(), across(PCGDP:GINI, mean, w = POP))
with(mtcars, table(cyl, vs, am))
sum(mtcars)
diff(EuStockMarkets)
mean(num_vars(iris), g = iris$Species)
unique(wlddev, cols = c("iso3c", "year"))
range(wlddev$date)
wlddev |> index_by(iso3c, year) |>
  mutate(PCGDP_lag = lag(PCGDP),
         PCGDP_diff = PCGDP - PCGDP_lag,
         PCGDP_growth = growth(PCGDP)) |> unindex()
\end{Code}
%
The above is now 100\% \pkg{collapse} code. Similarly, using this option, all code in this article could have been written without f-prefixes. Thus, \pkg{collapse} is able to offer a fast and syntactically clean experience of \proglang{R} - without the need to even restart the session. Masking is completely and interactively reversible: calling \code{set\_collapse(mask = NULL)} instantly removes the additional functions. Option \code{remove} can further be used to remove (un-export) any \pkg{collapse} function, allowing manual conflict management. Function \code{fastverse::fastverse\_conflicts()} from the related \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse} project} \citep{rfastverse} can be used to display namespace conflicts with \pkg{collapse}. Invoking either \code{mask} or \code{remove} detaches \pkg{collapse} and re-attaches it at the top of the search path, letting its namespace to take precedence over other packages.
%
\section{Benchmark} \label{sec:bench}
%
This section provides several simple benchmarks to show that \pkg{collapse} provides best-in-\proglang{R} performance for statistics and data manipulation on moderately sized datasets. They are executed on an Apple M1 MacBook Pro with 16 GB unified memory. It also discusses results from \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{3rd party benchmarks} involving \pkg{collapse}. The first set of benchmarks show that \pkg{collapse} provides faster computationally intensive operations like unique values and matching on large integer and character vectors. It creates integer/character vectors of 10 million obs, with 1000 unique integers and 5776 unique strings, respectively, which are deduplicated/matched in the benchmark. These fast basic operations impact many critical components of the package. % These are serially implemented and critical in supporting more complex functionality. %These algorithms power much of its functionality, such as efficient factor generation with \fct{qF}, cross-tabulation with \fct{qtab}, \fct{join}'s, \fct{pivot}'s, etc.
%
<<bench_1>>=
set.seed(101)
int <- 1:1000; g_int <- sample.int(1000, 1e7, replace = TRUE)
char <- c(letters, LETTERS, month.abb, month.name)
g_char <- sample(char <- outer(char, char, paste0), 1e7, TRUE)
bmark(base_int = unique(g_int), collapse_int = funique(g_int))
bmark(base_char = unique(g_char), collapse_char = funique(g_char))
bmark(base_int = match(g_int, int), collapse_int = fmatch(g_int, int))
bmark(base_char = match(g_char, char), data.table_char =
      chmatch(g_char, char), collapse_char = fmatch(g_char, char))
@
%
The second set below shows that \pkg{collapse}'s statistical functions are very efficient on aggregating a numeric matrix with 10,000 rows and 1000 columns. They are faster than base \proglang{R} even without multithreading, but using 4 threads in this case induces a sizeable difference.
<<bench_2, cache=TRUE>>=
set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)
m <- matrix(rnorm(1e7), ncol = 1000)
bmark(R = colSums(m), collapse = fsum(m))
bmark(R = colMeans(m), collapse = fmean(m))
bmark(MS = matrixStats::colMedians(m), collapse = fmedian(m))
@
%
Below I also show a grouped version summing the columns within 1000 random groups.
%
<<bench_2.1, cache=TRUE>>=
g <- sample.int(1e3, 1e4, TRUE)
bmark(R = rowsum(m, g), collapse = fsum(m, g))
@
%
I now turn to basic operations on a medium sized real-world database recording all flights from New York City (EWR, JFK, and LGA) in 2023---provided by the \pkg{nycflights23} package. The \code{flights} table has 435k flights, and grouping it by day and route yields 76k unique trips.
%
<<bench_flights_setup, cache=TRUE>>=
fastverse_extend(nycflights23, dplyr, data.table); setDTthreads(4)
list(flights, airports, airlines, planes, weather) |> sapply(nrow)
flights |> fselect(month, day, origin, dest) |> fnunique()
@
%
In the following, I select 6 numeric variables and sum them across the 76k trips using \pkg{dplyr}, \pkg{data.table}, and \pkg{collapse}. Ostensibly, despite \fct{sum} being 'primitive' (implemented in \proglang{C}), there is a factor 100 between \pkg{dplyr}'s split-apply-combine and \pkg{collapse}'s fully vectorized execution.
%
<<bench_flights_summarise, cache=TRUE>>=
vars <- .c(dep_delay, arr_delay, air_time, distance, hour, minute)
bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
                summarise(across(all_of(vars), sum), .groups = "drop"),
      data.table = qDT(flights)[, lapply(.SD, sum), .SDcols = vars,
                                by = .(month, day, origin, dest)],
      collapse = flights |> fgroup_by(month, day, origin, dest) |>
                   get_vars(vars) |> fsum())
@
%
Below, I also benchmark the mean and median functions in the same way. It is evident that with non-primitive \proglang{R} functions the split-apply-combine logic is even more costly.
%
<<bench_flights_summarise_2, cache=TRUE, echo=FALSE>>=
bmark(dplyr_mean = flights |> group_by(month, day, origin, dest) |>
              summarise(across(all_of(vars), mean), .groups = "drop"),
      data.table_mean = qDT(flights)[, lapply(.SD, mean), .SDcols = vars,
                                by = .(month, day, origin, dest)],
      collapse_mean = flights |> fgroup_by(month, day, origin, dest) |>
                 get_vars(vars) |> fmean())
bmark(dplyr_median = flights |> group_by(month, day, origin, dest) |>
              summarise(across(all_of(vars), median), .groups = "drop"),
      data.table_median = qDT(flights)[, lapply(.SD, median), .SDcols = vars,
                                by = .(month, day, origin, dest)],
      collapse_median = flights |> fgroup_by(month, day, origin, dest) |>
                 get_vars(vars) |> fmedian())
@
%
So far, \pkg{data.table}, by virtue of it's internal vectorizations (also via dedicated grouped \proglang{C} implementations of simple functions), is competitive.\footnote{Much longer data will likely also favor \pkg{data.table} over \pkg{collapse} due to its sub-column-level parallel grouping and implementation of simple functions like \fct{sum} and \fct{mean}, see, e.g., the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}.} Below, I compute the range of one column (\code{x}) using \code{max(x) - min(x)}. As elucidated in Section~\ref{ssec:vfat}, this expression is also vectorized in \pkg{collapse}, where it amounts to \code{fmax(x, g) - fmin(x, g)}, but not in \pkg{data.table}.
%
<<bench_flights_summarise_3, cache=TRUE>>=
bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
    summarise(rng = max(arr_delay) - min(arr_delay), .groups = "drop"),
  data.table = qDT(flights)[, .(rng = max(arr_delay) - min(arr_delay)),
                            by = .(month, day, origin, dest)],
  collapse = flights |> fgroup_by(month, day, origin, dest) |>
    fsummarise(rng = fmax(arr_delay) - fmin(arr_delay)))
@
%
I also benchmark table joins and pivots. The following demonstrates how all tables can be joined together using \pkg{collapse} and its default first-match left-join, which preserves \code{flights}.
%
\begin{Code}
R> flights |> join(weather, on = c("origin", "time_hour")) |>
+    join(planes, on = "tailnum") |> join(airports, on = c(dest = "faa")) |>
+    join(airlines, on = "carrier") |> dim()
\end{Code}
<<bench_flights_join_setup, cache=TRUE, echo=FALSE>>=
flights |> join(weather, on = c("origin", "time_hour")) |>
  join(planes, on = "tailnum") |> join(airports, on = c(dest = "faa")) |>
  join(airlines, on = "carrier") |> dim() |>
  capture.output() |> substr(1, 76) |> cat(sep = "\n")
@
%
The verbosity of \fct{join} is essential to understanding what has happened here---how many records from each table were matched and which duplicate non-id columns were suffixed with the (default) \code{y}-table name. Normally, I would set \code{drop.dup.cols = "y"} as it seems not useful to preserve them here, but the other packages don't have this option. For the benchmark, I set \code{verbose = 0} in \pkg{collapse} and employ the fastest syntax for \pkg{dplyr} and \pkg{data.table}:\footnote{\code{left\_join(..., multiple = "first")} for \pkg{dplyr} and \code{y[x, on = ids, mult = "first"]} for \pkg{data.table}.}
%
<<bench_flights_join, cache=TRUE, echo=FALSE>>=
bmark(
  dplyr_joins = flights |>
    left_join(weather, by = c("origin", "time_hour"), multiple = "first") |>
    left_join(planes, by = "tailnum", multiple = "first") |>
    left_join(airports, by = c(dest = "faa"), multiple = "first") |>
    left_join(airlines, by = "carrier", multiple = "first"),

  data.table_joins = qDT(airlines)[qDT(airports)[qDT(planes)[qDT(weather)[qDT(flights), on = c("origin", "time_hour"), mult = "first"], on = "tailnum", mult = "first"], on = c("faa" = "dest"), mult = "first"], on = "carrier", mult = "first"],

  collapse_joins = flights |>
    join(weather, on = c("origin", "time_hour"), verbose = 0) |>
    join(planes, on = "tailnum", verbose = 0) |>
    join(airports, on = c("dest" = "faa"), verbose = 0) |>
    join(airlines, on = "carrier", verbose = 0)
)
@
%
Evidently, the vectorized hash join provided by \pkg{collapse} is 10x faster than \pkg{data.table} on this database, at a substantially lower memory footprint. It remains competitive on \href{https://duckdblabs.github.io/db-benchmark/}{big data}.\footnote{\pkg{data.table} joins utilize multithreaded radix-ordering---a very different logic more useful for big data.}

% \footnote{A lot may be said about benchmarking \pkg{collapse}, which would be beyond the scope of this article. Users should note, however, that its defaults (\code{na.rm = TRUE, sort = TRUE, stable.algo = TRUE, nthreads = 1}) cater to convenience rather than maximum performance. For maximum performance, set these 3 settings to \code{FALSE} and increase the number of threads. To also provide a minimalistic guide for \proglang{R} users seeking to understand the relative performance of \pkg{collapse} and \pkg{data.table}, reflecting current (spring 2024) developments: \pkg{collapse} has highly efficient algorithms for grouping and computing statistics, but presently does not provide sub-column level parallel grouping architecture. Simple statistics like \code{fmean()} are parallelized across columns and perform grouped computations in a single pass. More complex ones \code{fmedian(), fmode()} have group-level parallelism. \pkg{data.table}, on the other hand, has sub-column parallel grouping and also group-level parallel implementations for simple statistics such as \code{mean()}, but no parallelism for complex statistics such as \code{median()}. \pkg{data.table}'s \href{https://rdatatable.gitlab.io/data.table/reference/datatable-optimize.html}{GForce optimization} also only applies to simple statistics, not complex expressions or weighted statistics - as can be vectorized using \emph{Fast Statistical Functions} in \pkg{collapse}.}
% Thus, if your data is moderately sized ($\leq$100mio. obs.), you have more than 1 column to compute on, you want to do complex statistical things, or if your processor is very fast (high single core speed), \pkg{collapse} is a great choice. On the other hand, if your data is really long ($>$100mio. obs.), you have only a few columns to compute on, you are computing simple statistics that \pkg{data.table} optimizes, and you have massive parallel compute, then \pkg{data.table} is a great choice. My recommendation: use both, just need to call \code{library(fastverse)}. Finally, let me note that \pkg{polars} uses optimized memory buffers based on \href{https://arrow.apache.org/}{Apache Arrow}, \href{https://pola.rs/posts/i-wrote-one-of-the-fastest-dataframe-libraries/}{multithreaded hash-based grouping, SIMD instructions and multithreading at the group-level}, and a \href{https://pola.rs/posts/polars_birds_eye_view/}{query optimizer} - all implemented in \proglang{Rust}, a thread-safe programming language. While some of these parallel algorithms could be ported to \pkg{collapse}, this is more challenging since \proglang{C}, and particularly \proglang{R}'s \proglang{C} API, is not thread safe - and it would still be lacking the benefits of Arrow memory buffers. At core, \proglang{R} is a 30-year old statistical language and not intended to work like an optimized database. \pkg{collapse} seamlessly integrates with \proglang{R}'s data structures; \pkg{polars}, at present, has nothing to do with them (and is therefore also not part of this benchmark).

Last but not least, I benchmark pivots, starting a with long pivot that simply melt the 6 columns aggregated beforehand into one column, duplicating all other columns 6 times:
%
<<bench_pivot_long, cache=TRUE, max.print=50>>=
bmark(tidyr = tidyr::pivot_longer(flights, cols = vars),
      data.table = qDT(flights) |> melt(measure = vars),
      collapse = pivot(flights, values = vars))
@
%
Memory-wise, \pkg{collapse} and \pkg{data.table} are equally efficient, but \pkg{collapse} is faster, presumably due to more extensive use of \code{memset()} to copy values in \proglang{C}, or smaller \proglang{R}-level overheads.

To complete the picture, I also also perform a wide pivot where the 6 columns are summed (for efficiency) across the 3 origin airports and expanded to create 18 airport-value columns.
%
<<bench_pivot_wide, cache=TRUE, max.print=50>>=
bmark(tidyr = tidyr::pivot_wider(flights, id_cols = .c(month, day, dest),
          names_from = "origin", values_from = vars, values_fn = sum),
      data.table = dcast(qDT(flights), month + day + dest ~ origin,
                         value.var = vars, fun = sum),
      collapse_fsum = pivot(flights, .c(month, day, dest), vars,
                            "origin", how = "wider", FUN = fsum),
      collapse_itnl = pivot(flights, .c(month, day, dest), vars,
                            "origin", how = "wider", FUN = "sum"))
@
%
Again, \pkg{collapse} is fastest, as it offers full vectorization, either via \fct{fsum}, which translates to \code{fsum(x, g, TRA = "fill")} before pivoting and thus entails a full deep copy of the \code{vars} columns, or via an optimized internal sum function which sums values 'on the fly' during the reshaping process. \pkg{data.table} is not vectorized here but at least memory efficient. %\newline

In summary, these benchmarks show that \pkg{collapse} provides outstanding performance and memory efficiency on a typical medium-sized real-world database popular in the \proglang{R} community.
% %
% <<bench, cache=TRUE>>=
% setDTthreads(4)
% set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)
% set.seed(101)
% m <- matrix(rnorm(1e7), ncol = 1000)
% data <- qDT(replicate(100, rnorm(1e5), simplify = FALSE))
% g <- sample.int(1e4, 1e5, TRUE)
%
% microbenchmark(R = colMeans(m), collapse = fmean(m))
% microbenchmark(R = rowsum(data, g, reorder = FALSE),
%                data.table = data[, lapply(.SD, sum), by = g],
%                collapse = fsum(data, g))
% add_vars(data) <- g
% microbenchmark(data.table = data[, lapply(.SD, median), by = g],
%                collapse = data |> fgroup_by(g) |> fmedian())
% d <- data.table(g = unique(g), x = 1, y = 2, z = 3)
% microbenchmark(data.table = d[data, on = "g"],
%                collapse = join(data, d, on = "g", verbose = 0))
% microbenchmark(data.table = melt(data, "g"),
%                collapse = pivot(data, "g"))
% settransform(data, id = rowid(g))
% cols = grep("^V", names(data), value = TRUE)
% microbenchmark(data.table = dcast(data, g ~ id, value.var = cols),
%           collapse = pivot(data, ids = "g", names = "id", how = "w"))
% @
%
%
% Apart from the raw algorithmic efficiency demonstrated here, \pkg{collapse} is often more efficient than other solutions by simply doing less. For example, if grouping columns are factor variables, \pkg{collapse}'s algorithms in \code{funique()}, \code{group()} or \code{fmatch()}, etc., use the values as hashes without checking for collisions. Similarly, if data is already sorted/unique, it is directly returned by functions like \code{roworder()}/\code{funique()}. \newline
%
\subsection{Other benchmarks}
%
The \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks} compare many software packages for database-like operations using large datasets (big data) on a linux server. The January 2025 run distinguishes 6 packages that consistently achieve outstanding performances: \pkg{DuckDB}, \pkg{Polars}, \pkg{ClickHouse}, Apache \pkg{Datafusion}, \pkg{data.table}, and \pkg{collapse}. Of these, \pkg{DuckDB}, \pkg{ClickHouse}, and \pkg{Datafusion} are vectorized database (SQL) engines, and \pkg{Polars} is a Python/Rust DataFrame library and SQL engine. These four are supported by (semi-)commercial entities, leaving \pkg{data.table} as the only fully community-led project, and \pkg{collapse} as the only project that is single-authored and without financial support. The benchmarks show that \pkg{collapse} achieves the highest relative performance on 'smaller' datasets (10-100 million rows) and performing advanced operations. % \newline

Since June 2024, there is also an independent \href{https://github.com/AdrianAntico/Benchmarks?tab=readme-ov-file#background}{database-like operations benchmark by Adrian Antico} using a windows server and executing scripts inside IDEs (VScode, Rstudio), on which \pkg{collapse} achieved the overall fastest runtimes. I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.github.io/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages. These benchmarks align in showing that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and on windows systems.\footnote{Reasons for the particularly strong performance of \pkg{collapse} on Windows may be that it is largely written in C and has limited multithreading in favor or highly efficient serial algorithms---there appear to be persistent obstacles to efficient (low-overhead) multithreading on Windows, implying that multithreaded query engines do not develop their full thrust on medium-sized ($\leq$100 million row) datasets.}

\subsection{Limitations and outlook}

\pkg{collapse} maximizes three principal objectives: being class-agnostic/fully compatible with the \proglang{R} ecosystem (supporting statistical operations on vector, matrix and data.frame-like objects), being statistically advanced, and being fast. This warranted some design choices away from maximum performance for large data manipulation.\footnote{Which nowadays would demand creating a multithreaded, vectorized query engine with optimized memory buffers/vector types to take full advantage of SIMD processing as in \pkg{DuckDB} or \pkg{Polars}. Such an architecture is very difficult to square with \proglang{R} vectors and \proglang{R}'s 30-year old C API.} Its limited use of multithreading and SIMD instructions, partly by design constraints and by R's C API, and the use of standard types for internal indexing, imposes hard-limits---the maximum integer in \proglang{R} is 2,147,483,647 $\to$ the maximum vector length \pkg{collapse} supports. It is and will remain an in-memory tool. \newline

Despite these constraints, \pkg{collapse} provides very respectable performance even on very large datasets by virtue of its algorithmic and memory efficiency. It is, together with the popular \pkg{data.table} package offering more sub-column-level parallel architecture for basic operations, well-positioned to remain a premier tool for in-memory statistics and data manipulation. % \newline

%% -- Summary/conclusions/discussion -------------------------------------------

\section{Conclusion} \label{sec:conclusion}

\pkg{collapse} was first released to CRAN in March 2020, and has grown and matured considerably over the course of 5 years. It has become a new foundation package for statistical computing and data transformation in \proglang{R}---one that is statistically advanced, class-agnostic, flexible, fast, lightweight, stable, and able to manipulate complex scientific data with ease. As such, it opens up new possibilities for statistics, data manipulation, and package development in \proglang{R}. \newline % It also sets an ambitious benchmark for high-quality software engineering on the language. \newline

This article provided a quick guide to the package, articulating its key ideas and design principles and demonstrating all core features. At this point the API is stable---it has changed very little over the 5 years and no further changes are planned. Compatibility with \proglang{R} version 3.4.0 will be maintained for as long as possible. Minor new features are currently planned. \newline

For deeper engagement with \pkg{collapse}, visit its \href{https://sebkrantz.github.io/collapse/index.html}{website} or start with the vignette summarizing all available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}. Users can also follow \pkg{collapse} on \href{https://x.com/collapse_R}{Twitter/X} and \href{https://bsky.app/profile/rcollapse.bsky.social}{Bluesky} to be notified about major updates and participate in community discussions.

\newpage
%% -- Optional special unnumbered sections -------------------------------------

\section*{Computational details}
The results in this paper were obtained using \proglang{R} \citep{R} 4.3.0 with \pkg{collapse} 2.0.19, \pkg{data.table} 1.16.4, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.0.0, \pkg{fastverse} 0.3.4, and \pkg{bench} 1.1.3 \citep{rbench}. All packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at https://CRAN.R-project.org/. The benchmark was run on an Apple M1 MacBook Pro (2020) with 16GB unified memory. Packages were compiled from source using Homebrew Clang version 16.0.4 with OpenMP enabled and the -O2 flag.

\section*{Acknowledgments}

The source code of \pkg{collapse} has been heavily inspired by (and partly copied from) \pkg{data.table} (Matt Dowle and Arun Srinivasan), \proglang{R}'s source code (R Core Team and contributors worldwide), the \pkg{kit} package (Morgan Jacob), and \pkg{Rcpp} (Dirk Eddelbuettel). Packages \pkg{plm} (Yves Croissant, Giovanni Millo, and Kevin Tappe) and \pkg{fixest} (Laurent Berge) have also provided a lot of inspiration (and a port to its demeaning algorithm in the case of \pkg{fixest}). I also thank many people from diverse fields for helpful answers on Stackoverflow and many other people for encouragement, feature requests, and helpful issues and suggestions.
% \begin{leftbar}
% All acknowledgments (note the AE spelling) should be collected in this
% unnumbered section before the references. It may contain the usual information
% about funding and feedback from colleagues/reviewers/etc. Furthermore,
% information such as relative contributions of the authors may be added here
% (if any).
% \end{leftbar}

%% -- Bibliography -------------------------------------------------------------
%% - References need to be provided in a .bib BibTeX database.
%% - All references should be made with \cite, \citet, \citep, \citealp etc.
%%   (and never hard-coded). See the FAQ for details.
%% - JSS-specific markup (\proglang, \pkg, \code) should be used in the .bib.
%% - Titles in the .bib should be in title case.
%% - DOIs should be included where available.

\newpage

\bibliography{refs}

%% -- Appendix (if any) --------------------------------------------------------
%% - After the bibliography with page break.
%% - With proper section titles and _not_ just "Appendix".

\newpage
%
% \begin{appendix}
%
% \section{More technical details} \label{app:technical}
%
% \begin{leftbar}
% Appendices can be included after the bibliography (with a page break). Each
% section within the appendix should have a proper section title (rather than
% just \emph{Appendix}).
%
% For more technical style details, please check out JSS's style FAQ at
% \url{https://www.jstatsoft.org/pages/view/style#frequently-asked-questions}
% which includes the following topics:
% \begin{itemize}
%   \item Title vs.\ sentence case.
%   \item Graphics formatting.
%   \item Naming conventions.
%   \item Turning JSS manuscripts into \proglang{R} package vignettes.
%   \item Trouble shooting.
%   \item Many other potentially helpful details\dots
% \end{itemize}
% \end{leftbar}
%
%
% \section[Using BibTeX]{Using \textsc{Bib}{\TeX}} \label{app:bibtex}
%
% \begin{leftbar}
% References need to be provided in a \textsc{Bib}{\TeX} file (\code{.bib}). All
% references should be made with \verb|\cite|, \verb|\citet|, \verb|\citep|,
% \verb|\citealp| etc.\ (and never hard-coded). This commands yield different
% formats of author-year citations and allow to include additional details (e.g.,
% pages, chapters, \dots) in brackets. In case you are not familiar with these
% commands see the JSS style FAQ for details.
%
% Cleaning up \textsc{Bib}{\TeX} files is a somewhat tedious task -- especially
% when acquiring the entries automatically from mixed online sources. However,
% it is important that informations are complete and presented in a consistent
% style to avoid confusions. JSS requires the following format.
% \begin{itemize}
%   \item JSS-specific markup (\verb|\proglang|, \verb|\pkg|, \verb|\code|) should
%     be used in the references.
%   \item Titles should be in title case.
%   \item Journal titles should not be abbreviated and in title case.
%   \item DOIs should be included where available.
%   \item Software should be properly cited as well. For \proglang{R} packages
%     \code{citation("pkgname")} typically provides a good starting point.
% \end{itemize}
% \end{leftbar}
%
% \end{appendix}

%% -----------------------------------------------------------------------------

\end{document}

### 2025-02-19 misc/JSS article/R&R/article.tex

--- 2025-02-19 misc/JSS article/R&R/article.Rnw
@@
 %% need no \usepackage{Sweave.sty}
-\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, cache=FALSE}
+
 % \renewenvironment{knitrout}{\setlength{\topsep}{0mm}}{}
-
-<<preliminaries, echo=FALSE, message=FALSE, warning=FALSE, results=hide>>=
-# knitr::opts_chunk$set(cache = FALSE, size = "small")
-rm(list = ls()); gc()
-options(prompt = "R> ", continue = "+  ", width = 77, digits = 4, useFancyQuotes = FALSE, warn = 1)
-
-# Loading libraries and installing if unavailable
-if(!requireNamespace("fastverse", quietly = TRUE)) install.packages("fastverse")
-library(fastverse) # loads data.table, collapse, magrittr and kit (not used)
-# Package versions used in the article:
-# fastverse 0.3.4, collapse 2.0.19, data.table 1.16.4,
-# bench 1.1.3, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.0.0
-
-# Reset collapse options (if set)
-set_collapse(nthreads = 1L, remove = NULL, stable.algo = TRUE, sort = TRUE,
-             digits = 2L, stub = TRUE, verbose = 1L, mask = NULL, na.rm = TRUE)
-
-# For presentation: removing whitespace around labels
-vlabels(GGDC10S) <- trimws(vlabels(GGDC10S))
-
-# This is the benchmark function (janitor 2.2.0 is used)
-bmark <- function(...) {
-  bench::mark(..., min_time = 2, check = FALSE) |> janitor::clean_names() |>
-  fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
-  fmutate(expression = names(expression)) |> dapply(as.character) |> qDF()
-}
-@
 
@@
 %, and attributes are preserved as much as possible ()
-<<faststatfun>>=
-fmean(iris$Sepal.Length)
-fmean(iris[1:4])
-identical(fmean(iris[1:4]), fmean(as.matrix(iris[1:4])))
-fmean(iris$Sepal.Length, g = iris$Species)
-fmean(iris[1:4], g = iris$Species, nthreads = 4)
-fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill")[1:10]
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fmean(iris$Sepal.Length)
+\end{Sinput}
+\begin{Soutput}
+[1] 5.843
+\end{Soutput}
+\begin{Sinput}
+R> fmean(iris[1:4])
+\end{Sinput}
+\begin{Soutput}
+Sepal.Length  Sepal.Width Petal.Length  Petal.Width 
+       5.843        3.057        3.758        1.199 
+\end{Soutput}
+\begin{Sinput}
+R> identical(fmean(iris[1:4]), fmean(as.matrix(iris[1:4])))
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\begin{Sinput}
+R> fmean(iris$Sepal.Length, g = iris$Species)
+\end{Sinput}
+\begin{Soutput}
+    setosa versicolor  virginica 
+     5.006      5.936      6.588 
+\end{Soutput}
+\begin{Sinput}
+R> fmean(iris[1:4], g = iris$Species, nthreads = 4)
+\end{Sinput}
+\begin{Soutput}
+           Sepal.Length Sepal.Width Petal.Length Petal.Width
+setosa            5.006       3.428        1.462       0.246
+versicolor        5.936       2.770        4.260       1.326
+virginica         6.588       2.974        5.552       2.026
+\end{Soutput}
+\begin{Sinput}
+R> fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill")[1:10]
+\end{Sinput}
+\begin{Soutput}
+ [1] 5.006 5.006 5.006 5.006 5.006 5.006 5.006 5.006 5.006 5.006
+\end{Soutput}
+\end{Schunk}
 %
@@
 
-<<exports>>=
-set.seed(101)
-exports <- expand.grid(y=1:10, c=paste0("c",1:10), s=paste0("s",1:10)) |>
-  tfm(v = abs(rnorm(1e3))) |> colorder(c, s) |> ss(-sample.int(1e3, 500))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(101)
+R> exports <- expand.grid(y=1:10, c=paste0("c",1:10), s=paste0("s",1:10)) |>
++    tfm(v = abs(rnorm(1e3))) |> colorder(c, s) |> ss(-sample.int(1e3, 500))
+\end{Sinput}
+\end{Schunk}
 Like any real trade dataset, it is unbalanced---\code{ss(-sample.int(1e3, 500))} randomly removes 500 rows. Suppose we wanted to extract the latest trade within the last two years.
-<<exports_2>>=
-latest <- fsubset(exports, y >= 8 & y == fmax(y, list(c, s), "fill"), -y)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> latest <- fsubset(exports, y >= 8 & y == fmax(y, list(c, s), "fill"), -y)
+\end{Sinput}
+\end{Schunk}
 Below, I compute how many products different countries have exported in the last two years
-<<prod>>=
-with(latest, fndistinct(s, c))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> with(latest, fndistinct(s, c))
+\end{Sinput}
+\begin{Soutput}
+ c1  c2  c3  c4  c5  c6  c7  c8  c9 c10 
+  9   9   9  10   9   9   8  10   9  10 
+\end{Soutput}
+\end{Schunk}
 and this computes \citet{balassa1965tariff}'s index of Revealed Comparative Advantage, defined as the share of a sector in country exports divided by the share of the sector in world exports.
-<<RCA>>=
-with(latest, fsum(v, c, TRA = "/") / fsum(v, s, TRA = "/"))[1:10]
-@
+\begin{Schunk}
+\begin{Sinput}
+R> with(latest, fsum(v, c, TRA = "/") / fsum(v, s, TRA = "/"))[1:10]
+\end{Sinput}
+\begin{Soutput}
+ [1] 0.6674 0.5673 0.6403 0.8162 0.7989 0.9122 0.8561 0.5611 0.7847 1.1557
+\end{Soutput}
+\end{Schunk}
 More complex use cases are frequent in my work.
@@
 %
-<<GRP>>=
-str(g <- GRP(wlddev, ~ income + OECD))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> str(g <- GRP(wlddev, ~ income + OECD))
+\end{Sinput}
+\begin{Soutput}
+Class 'GRP'  hidden list of 9
+ $ N.groups    : int 6
+ $ group.id    : int [1:13176] 3 3 3 3 3 3 3 3 3 3 ...
+ $ group.sizes : int [1:6] 2745 2074 1830 2867 3538 122
+ $ groups      :'data.frame':	6 obs. of  2 variables:
+  ..$ income: Factor w/ 4 levels "High income",..: 1 1 2 3 4 4
+  .. ..- attr(*, "label")= chr "Income Level"
+  ..$ OECD  : logi [1:6] FALSE TRUE FALSE FALSE FALSE TRUE
+  .. ..- attr(*, "label")= chr "Is OECD Member Country?"
+ $ group.vars  : chr [1:2] "income" "OECD"
+ $ ordered     : Named logi [1:2] TRUE FALSE
+  ..- attr(*, "names")= chr [1:2] "ordered" "sorted"
+ $ order       : int [1:13176] 245 246 247 248 249 250 251 252 253 254 ...
+  ..- attr(*, "starts")= int [1:6] 1 2746 4820 6650 9517 13055
+  ..- attr(*, "maxgrpn")= int 3538
+  ..- attr(*, "sorted")= logi FALSE
+ $ group.starts: int [1:6] 245 611 1 306 62 7687
+ $ call        : language GRP.default(X = wlddev, by = ~income + OECD)
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<gcomp>>=
-add_vars(g$groups,
- get_vars(wlddev, "country") |> fmode(g, wlddev$POP, use = FALSE),
- get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = F))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> add_vars(g$groups,
++   get_vars(wlddev, "country") |> fmode(g, wlddev$POP, use = FALSE),
++   get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = F))
+\end{Sinput}
+\begin{Soutput}
+               income  OECD       country   PCGDP LIFEEX
+1         High income FALSE  Saudi Arabia 22426.7  73.00
+2         High income  TRUE United States 31749.6  75.84
+3          Low income FALSE      Ethiopia   557.1  53.51
+4 Lower middle income FALSE         India  1238.8  60.59
+5 Upper middle income FALSE         China  3820.6  68.21
+6 Upper middle income  TRUE        Mexico  8311.2  69.06
+\end{Soutput}
+\end{Schunk}
 For advanced data aggregation, \pkg{collapse} also provides a convenience function, \fct{collap}, which, by default, uses \fct{fmean} for numeric, \fct{fmode} for categorical, and \fct{fsum} for weight columns, and preserves their order.
@@
 %
-<<collap>>=
-collap(wlddev, country + PCGDP + LIFEEX ~ income + OECD, w = ~ POP)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> collap(wlddev, country + PCGDP + LIFEEX ~ income + OECD, w = ~ POP)
+\end{Sinput}
+\begin{Soutput}
+        country              income  OECD   PCGDP LIFEEX       POP
+1  Saudi Arabia         High income FALSE 22426.7  73.00 3.114e+09
+2 United States         High income  TRUE 31749.6  75.84 5.573e+10
+3      Ethiopia          Low income FALSE   557.1  53.51 2.095e+10
+4         India Lower middle income FALSE  1238.8  60.59 1.138e+11
+5         China Upper middle income FALSE  3820.6  68.21 1.114e+11
+6        Mexico Upper middle income  TRUE  8311.2  69.06 8.162e+09
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<gtrans>>=
-add_vars(wlddev) <- get_vars(wlddev, c("PCGDP", "LIFEEX")) |>
-    fmean(wlddev$iso3c, TRA = "-+") |> add_stub("center_")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> add_vars(wlddev) <- get_vars(wlddev, c("PCGDP", "LIFEEX")) |>
++      fmean(wlddev$iso3c, TRA = "-+") |> add_stub("center_")
+\end{Sinput}
+\end{Schunk}
 %
@@
 %
-<<fsummarise>>=
-wlddev |> fsubset(year >= 2015) |> fgroup_by(income) |>
-  fsummarise(country = fmode(country, POP),
-             across(c(PCGDP, LIFEEX, GINI), fmean, POP))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlddev |> fsubset(year >= 2015) |> fgroup_by(income) |>
++    fsummarise(country = fmode(country, POP),
++               across(c(PCGDP, LIFEEX, GINI), fmean, POP))
+\end{Sinput}
+\begin{Soutput}
+               income       country PCGDP LIFEEX  GINI
+1         High income United States 43340  80.70 36.14
+2          Low income      Ethiopia   663  63.05 39.13
+3 Lower middle income         India  2177  68.31 36.48
+4 Upper middle income         China  8168  75.51 41.68
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<fsummarise_2>>=
-wlddev |> fsubset(year >= 2015, income, PCGDP:GINI, POP) |>
-  fgroup_by(income) |> fmean(POP, keep.w = FALSE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlddev |> fsubset(year >= 2015, income, PCGDP:GINI, POP) |>
++    fgroup_by(income) |> fmean(POP, keep.w = FALSE)
+\end{Sinput}
+\begin{Soutput}
+               income PCGDP LIFEEX  GINI
+1         High income 43340  80.70 36.14
+2          Low income   663  63.05 39.13
+3 Lower middle income  2177  68.31 36.48
+4 Upper middle income  8168  75.51 41.68
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<glinreg>>=
-exports |> fgroup_by(c, s) |> fmutate(dmy = fwithin(y)) |>
-  fsummarise(v_10 = flast(v), beta = fsum(v, dmy) %/=% fsum(dmy, dmy)) |>
-  fmutate(v_11 = v_10 + beta, v_12 = v_11 + beta, beta = NULL) |> head(4)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> exports |> fgroup_by(c, s) |> fmutate(dmy = fwithin(y)) |>
++    fsummarise(v_10 = flast(v), beta = fsum(v, dmy) %/=% fsum(dmy, dmy)) |>
++    fmutate(v_11 = v_10 + beta, v_12 = v_11 + beta, beta = NULL) |> head(4)
+\end{Sinput}
+\begin{Soutput}
+   c  s   v_10   v_11     v_12
+1 c1 s1 0.2233 0.2152  0.20707
+2 c1 s2 1.6860 1.7235  1.76112
+3 c1 s3 0.2971 0.1364 -0.02433
+4 c1 s4 1.0617 1.3544  1.64707
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<wstats>>=
-wlddev |> fsubset(is.finite(POP)) |> fgroup_by(income, OECD) |>
-  fmutate(o = radixorder(GRPid(), LIFEEX)) |>
-  fsummarise(min = fmin(LIFEEX),
-             Q1 = fnth(LIFEEX, 0.25, POP, o = o, ties = "q8"),
-             mean = fmean(LIFEEX, POP),
-             median = fmedian(LIFEEX, POP, o = o, ties = "q8"),
-             Q3 = fnth(LIFEEX, 0.75, POP, o = o, ties = "q8"),
-             max = fmax(LIFEEX))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlddev |> fsubset(is.finite(POP)) |> fgroup_by(income, OECD) |>
++    fmutate(o = radixorder(GRPid(), LIFEEX)) |>
++    fsummarise(min = fmin(LIFEEX),
++               Q1 = fnth(LIFEEX, 0.25, POP, o = o, ties = "q8"),
++               mean = fmean(LIFEEX, POP),
++               median = fmedian(LIFEEX, POP, o = o, ties = "q8"),
++               Q3 = fnth(LIFEEX, 0.75, POP, o = o, ties = "q8"),
++               max = fmax(LIFEEX))
+\end{Sinput}
+\begin{Soutput}
+               income  OECD   min    Q1  mean median    Q3   max
+1         High income FALSE 42.67 70.58 73.00  73.80 76.52 85.42
+2         High income  TRUE 55.42 72.61 75.84  76.24 78.81 84.36
+3          Low income FALSE 26.17 46.95 53.51  53.01 60.37 74.43
+4 Lower middle income FALSE 18.91 54.69 60.59  62.31 67.55 76.70
+5 Upper middle income FALSE 36.53 65.86 68.21  69.50 73.27 80.28
+6 Upper middle income  TRUE 45.37 64.56 69.06  71.96 74.96 77.69
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<ts_example>>=
-fgrowth(airmiles) |> round(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fgrowth(airmiles) |> round(2)
+\end{Sinput}
+\begin{Soutput}
+Time Series:
+Start = 1937 
+End = 1960 
+Frequency = 1 
+ [1]    NA 16.50 42.29 54.03 31.65  2.38 15.23 33.29 54.36 76.92  2.71 -2.10
+[13] 12.91 18.51 32.03 18.57 17.82 13.61 18.19 12.83 13.32  0.01 15.49  4.25
+\end{Soutput}
+\end{Schunk}
 The following extracts one sector from the \code{exports} dataset generated above, creating an irregular time series missing the 3rd and 6th period.\footnote{\code{\%=\%} is an infix operator for the \fct{massign} function in \pkg{collapse} which is a multivariate version of \fct{assign}.} Indexation using the \code{t} argument allows for correct (time-aware) computations on this context without 'expanding' the data.
 %
-<<ts_example_2>>=
-.c(y, v) %=% fsubset(exports, c == "c1" & s == "s7", -c, -s)
-print(y)
-fgrowth(v, t = y) |> round(2)
-fgrowth(v, -1:3, t = y) |> head(4)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> .c(y, v) %=% fsubset(exports, c == "c1" & s == "s7", -c, -s)
+R> print(y)
+\end{Sinput}
+\begin{Soutput}
+[1]  1  2  4  5  7  8  9 10
+\end{Soutput}
+\begin{Sinput}
+R> fgrowth(v, t = y) |> round(2)
+\end{Sinput}
+\begin{Soutput}
+[1]     NA 175.52     NA -22.37     NA 624.27 -79.01 534.56
+\end{Soutput}
+\begin{Sinput}
+R> fgrowth(v, -1:3, t = y) |> head(4)
+\end{Sinput}
+\begin{Soutput}
+        FG1     --     G1   L2G1   L3G1
+[1,] -63.71 0.3893     NA     NA     NA
+[2,]     NA 1.0726 175.52     NA     NA
+[3,]  28.82 0.8450     NA -21.22 117.05
+[4,]     NA 0.6559 -22.37     NA -38.85
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<example_growth>>=
-G(exports, -1:2, by = v ~ c + s, t = ~ y) |> head(3)
-tfm(exports, fgrowth(list(v = v), -1:2, g = list(c, s), t = y)) |> head(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> G(exports, -1:2, by = v ~ c + s, t = ~ y) |> head(3)
+\end{Sinput}
+\begin{Soutput}
+   c  s y  FG1.v      v   G1.v L2G1.v
+1 c1 s1 2 -18.15 0.5525     NA     NA
+2 c1 s1 3 214.87 0.6749  22.17     NA
+3 c1 s1 4 -31.02 0.2144 -68.24  -61.2
+\end{Soutput}
+\begin{Sinput}
+R> tfm(exports, fgrowth(list(v = v), -1:2, g = list(c, s), t = y)) |> head(3)
+\end{Sinput}
+\begin{Soutput}
+   c  s y      v  FG1.v   G1.v L2G1.v
+1 c1 s1 2 0.5525 -18.15     NA     NA
+2 c1 s1 3 0.6749 214.87  22.17     NA
+3 c1 s1 4 0.2144 -31.02 -68.24  -61.2
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<example_growth_continued>>=
-A <- exports |> fgroup_by(c, s) |> fmutate(gv = G(v, t = y)) |> fungroup()
-head(B <- exports |> fmutate(gv = G(v, g = list(c, s), t = y)), 4)
-identical(A, B)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> A <- exports |> fgroup_by(c, s) |> fmutate(gv = G(v, t = y)) |> fungroup()
+R> head(B <- exports |> fmutate(gv = G(v, g = list(c, s), t = y)), 4)
+\end{Sinput}
+\begin{Soutput}
+   c  s y      v     gv
+1 c1 s1 2 0.5525     NA
+2 c1 s1 3 0.6749  22.17
+3 c1 s1 4 0.2144 -68.24
+4 c1 s1 5 0.3108  44.98
+\end{Soutput}
+\begin{Sinput}
+R> identical(A, B)
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\end{Schunk}
 %
@@
 An example using the \code{exports} data follows:
-<<indexing>>=
-exportsi <- exports |> findex_by(c, s, y)
-exportsi |> G() |> print(max = 15)
-exportsi |> findex() |> print(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> exportsi <- exports |> findex_by(c, s, y)
+R> exportsi |> G() |> print(max = 15)
+\end{Sinput}
+\begin{Soutput}
+   c  s y   G1.v
+1 c1 s1 2     NA
+2 c1 s1 3  22.17
+3 c1 s1 4 -68.24
+ [ reached 'max' / getOption("max.print") -- omitted 497 rows ]
+
+Indexed by:  c.s [100] | y [10] 
+\end{Soutput}
+\begin{Sinput}
+R> exportsi |> findex() |> print(2)
+\end{Sinput}
+\begin{Soutput}
+    c.s y
+1 c1.s1 2
+2 c1.s1 3
+---             
+499 c10.s10 7
+500 c10.s10 9
+
+c.s [100] | y [10]
+\end{Soutput}
+\end{Schunk}
 The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
-<<indexing_2>>=
-vi <- exportsi$v; str(vi, width = 70, strict = "cut")
-is_irregular(vi)
-vi |> psmat() |> head(3)
-fdiff(vi) |> psmat() |> head(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> vi <- exportsi$v; str(vi, width = 70, strict = "cut")
+\end{Sinput}
+\begin{Soutput}
+ 'indexed_series' num [1:500] 0.552 0.675 0.214 0.311 1.174 ...
+ - attr(*, "index_df")=Classes 'index_df', 'pindex' and 'data.frame'..
+  ..$ c.s: Factor w/ 100 levels "c1.s1","c2.s1",..: 1 1 1 1 1 1 1 1 ..
+  ..$ y  : Ord.factor w/ 10 levels "1"<"2"<"3"<"4"<..: 2 3 4 5 6 7 8..
+  ..- attr(*, "nam")= chr [1:3] "c" "s" "y"
+\end{Soutput}
+\begin{Sinput}
+R> is_irregular(vi)
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\begin{Sinput}
+R> vi |> psmat() |> head(3)
+\end{Sinput}
+\begin{Soutput}
+       1     2     3     4     5    6     7      8     9    10
+c1.s1 NA 0.552 0.675 0.214 0.311 1.17 0.619 0.1127 0.917 0.223
+c2.s1 NA 0.795    NA    NA 0.237   NA    NA 0.0585 0.818    NA
+c3.s1 NA 0.709 0.268 1.464    NA   NA 0.467 0.1193 0.467    NA
+\end{Soutput}
+\begin{Sinput}
+R> fdiff(vi) |> psmat() |> head(3)
+\end{Sinput}
+\begin{Soutput}
+       1  2      3      4      5     6      7      8     9     10
+c1.s1 NA NA  0.122 -0.461 0.0964 0.863 -0.555 -0.506 0.804 -0.694
+c2.s1 NA NA     NA     NA     NA    NA     NA     NA 0.759     NA
+c3.s1 NA NA -0.441  1.196     NA    NA     NA -0.348 0.348     NA
+\end{Soutput}
+\end{Schunk}
 \fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work in arbitrary data masking environments.
-<<indexing_3>>=
-settransform(exportsi, v_ld = Dlog(v))
-lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> settransform(exportsi, v_ld = Dlog(v))
+R> lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
+\end{Sinput}
+\begin{Soutput}
+               Estimate Std. Error t value Pr(>|t|)
+(Intercept)      -0.008      0.141  -0.058    0.954
+L(v_ld, 1:2)L1   -0.349      0.115  -3.042    0.004
+L(v_ld, 1:2)L2   -0.033      0.154  -0.215    0.831
+\end{Soutput}
+\end{Schunk}
 It is worth highlighting that the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions.\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.} Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures or native handling of irregularity in basic operations.
@@
 %
-<<rollmean>>=
-BY(vi, ix(vi)$c.s, data.table::frollmean, 5) |> head(10)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> BY(vi, ix(vi)$c.s, data.table::frollmean, 5) |> head(10)
+\end{Sinput}
+\begin{Soutput}
+ [1]     NA     NA     NA     NA 0.5853 0.5986 0.4861 0.6267 0.6092     NA
+
+Indexed by:  c.s [2] | y [9 (10)] 
+\end{Soutput}
+\end{Schunk}
 % Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
@@
 %
-<<joins>>=
-teacher <- data.frame(id = 1:4, names = c("John", "Jane", "Bob", "Carl"),
-  age = c(35, 32, 42, 67), subject = c("Math", "Econ", "Stats", "Trade"))
-course <- data.frame(id = c(1, 2, 2, 3, 5), semester = c(1, 1, 2, 1, 2),
-  course = c("Math I", "Microecon", "Macroecon", "Stats I", "History"))
-join(teacher, course, on = "id")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> teacher <- data.frame(id = 1:4, names = c("John", "Jane", "Bob", "Carl"),
++    age = c(35, 32, 42, 67), subject = c("Math", "Econ", "Stats", "Trade"))
+R> course <- data.frame(id = c(1, 2, 2, 3, 5), semester = c(1, 1, 2, 1, 2),
++    course = c("Math I", "Microecon", "Macroecon", "Stats I", "History"))
+R> join(teacher, course, on = "id")
+\end{Sinput}
+\begin{Soutput}
+left join: teacher[id] 3/4 (75%) <1:1st> course[id] 3/5 (60%)
+  id names age subject semester    course
+1  1  John  35    Math        1    Math I
+2  2  Jane  32    Econ        1 Microecon
+3  3   Bob  42   Stats        1   Stats I
+4  4  Carl  67   Trade       NA      <NA>
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<joins_2>>=
-join(teacher, course, how = "full", multiple = TRUE, column = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> join(teacher, course, how = "full", multiple = TRUE, column = TRUE)
+\end{Sinput}
+\begin{Soutput}
+full join: teacher[id] 3/4 (75%) <1:1.33> course[id] 4/5 (80%)
+  id names age subject semester    course   .join
+1  1  John  35    Math        1    Math I matched
+2  2  Jane  32    Econ        1 Microecon matched
+3  2  Jane  32    Econ        2 Macroecon matched
+4  3   Bob  42   Stats        1   Stats I matched
+5  4  Carl  67   Trade       NA      <NA> teacher
+6  5  <NA>  NA    <NA>        2   History  course
+\end{Soutput}
+\end{Schunk}
 %
@@
 \end{Code}
-<<joins_3, echo=FALSE>>=
-join(teacher, course, multiple = TRUE, attr = "jn") |> attr("jn") |> str(strict.width = "cut", width = 70)
-@
+\begin{Schunk}
+\begin{Soutput}
+left join: teacher[id] 3/4 (75%) <1:1.33> course[id] 4/5 (80%)
+List of 3
+ $ call   : language join(x = teacher, y = course, multiple = TRUE,"..
+ $ on.cols:List of 2
+  ..$ x: chr "id"
+  ..$ y: chr "id"
+ $ match  : 'qG' int [1:5] 1 2 3 4 NA
+  ..- attr(*, "N.nomatch")= int 1
+  ..- attr(*, "N.groups")= int 5
+  ..- attr(*, "N.distinct")= int 4
+\end{Soutput}
+\end{Schunk}
 Users can also invoke the \code{validate} argument to examine the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.
 %
-<<joins_4>>=
-join(teacher, course, on = "id", validate = "1:1") |>
-  tryCatch(error = function(e) strwrap(e) |> cat(sep = "\n"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> join(teacher, course, on = "id", validate = "1:1") |>
++    tryCatch(error = function(e) strwrap(e) |> cat(sep = "\n"))
+\end{Sinput}
+\begin{Soutput}
+Error in join(teacher, course, on = "id", validate = "1:1"): Join is
+not 1:1: teacher (x) is unique on the join columns; course (y) is
+not unique on the join columns
+\end{Soutput}
+\end{Schunk}
 %
@@
 This demonstrates the latter two:
-<<joins_5>>=
-for (h in c("semi", "anti")) join(teacher, course, how = h) |> print()
-@
+\begin{Schunk}
+\begin{Sinput}
+R> for (h in c("semi", "anti")) join(teacher, course, how = h) |> print()
+\end{Sinput}
+\begin{Soutput}
+semi join: teacher[id] 3/4 (75%) <1:1st> course[id] 3/5 (60%)
+  id names age subject
+1  1  John  35    Math
+2  2  Jane  32    Econ
+3  3   Bob  42   Stats
+anti join: teacher[id] 3/4 (75%) <1:1st> course[id] 3/5 (60%)
+  id names age subject
+1  4  Carl  67   Trade
+\end{Soutput}
+\end{Schunk}
 By default (\code{sort = FALSE}), the order of rows in \code{x} is preserved.
@@
 % In multi-match settings, this will be reflected by few records from \code{y} being used.
-<<joins_6>>=
-course$names <- teacher$names[course$id]
-join(teacher, course, on = "id", how = "inner", multiple = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> course$names <- teacher$names[course$id]
+R> join(teacher, course, on = "id", how = "inner", multiple = TRUE)
+\end{Sinput}
+\begin{Soutput}
+inner join: teacher[id] 3/4 (75%) <1:1.33> course[id] 4/5 (80%)
+duplicate columns: names => renamed using suffix '_course' for y
+  id names age subject semester    course names_course
+1  1  John  35    Math        1    Math I         John
+2  2  Jane  32    Econ        1 Microecon         Jane
+3  2  Jane  32    Econ        2 Macroecon         Jane
+4  3   Bob  42   Stats        1   Stats I          Bob
+\end{Soutput}
+\end{Schunk}
 As shown above, \fct{join}'s handling of duplicate columns in both tables is rather special.
@@
 %
-<<joins_7>>=
-join(teacher, course, on = "id", multiple = TRUE, drop.dup.cols = "y")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> join(teacher, course, on = "id", multiple = TRUE, drop.dup.cols = "y")
+\end{Sinput}
+\begin{Soutput}
+left join: teacher[id] 3/4 (75%) <1:1.33> course[id] 4/5 (80%)
+duplicate columns: names => dropped from y
+  id names age subject semester    course
+1  1  John  35    Math        1    Math I
+2  2  Jane  32    Econ        1 Microecon
+3  2  Jane  32    Econ        2 Macroecon
+4  3   Bob  42   Stats        1   Stats I
+5  4  Carl  67   Trade       NA      <NA>
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<pivots>>=
-data <- GGDC10S |>
-  fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
-  fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
-namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> data <- GGDC10S |>
++    fmutate(Label = ifelse(Variable == "VA", "Value Added", "Employment")) |>
++    fsubset(is.finite(AGR), Country, Variable, Label, Year, AGR:MAN)
+R> namlab(data, N = TRUE, Ndistinct = TRUE, class = TRUE)
+\end{Sinput}
+\begin{Soutput}
+  Variable     Class    N Ndist         Label
+1  Country character 4364    43       Country
+2 Variable character 4364     2      Variable
+3    Label character 4364     2          <NA>
+4     Year   numeric 4364    67          Year
+5      AGR   numeric 4364  4353   Agriculture
+6      MIN   numeric 4355  4224        Mining
+7      MAN   numeric 4355  4353 Manufacturing
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<pivot_longer>>=
-head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
-                 labels = "Sector", how = "longer"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> head(dl <- pivot(data, ids = 1:4, names = list("Sectorcode", "Value"),
++                   labels = "Sector", how = "longer"))
+\end{Sinput}
+\begin{Soutput}
+  Country Variable       Label Year Sectorcode      Sector Value
+1     BWA       VA Value Added 1964        AGR Agriculture 16.30
+2     BWA       VA Value Added 1965        AGR Agriculture 15.73
+3     BWA       VA Value Added 1966        AGR Agriculture 17.68
+4     BWA       VA Value Added 1967        AGR Agriculture 19.15
+5     BWA       VA Value Added 1968        AGR Agriculture 21.10
+6     BWA       VA Value Added 1969        AGR Agriculture 21.86
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<pivot_wider>>=
-head(dw <- pivot(data, c("Country", "Year"), names = "Variable",
-                 labels = "Label", how = "w"))
-namlab(dw)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> head(dw <- pivot(data, c("Country", "Year"), names = "Variable",
++                   labels = "Label", how = "w"))
+\end{Sinput}
+\begin{Soutput}
+  Country Year AGR_VA AGR_EMP MIN_VA MIN_EMP MAN_VA MAN_EMP
+1     BWA 1964  16.30   152.1  3.494  1.9400 0.7366   2.420
+2     BWA 1965  15.73   153.3  2.496  1.3263 1.0182   2.330
+3     BWA 1966  17.68   153.9  1.970  1.0022 0.8038   1.282
+4     BWA 1967  19.15   155.1  2.299  1.1192 0.9378   1.042
+5     BWA 1968  21.10   156.2  1.839  0.7855 0.7503   1.069
+6     BWA 1969  21.86   157.4  5.245  2.0314 2.1396   2.124
+\end{Soutput}
+\begin{Sinput}
+R> namlab(dw)
+\end{Sinput}
+\begin{Soutput}
+  Variable                       Label
+1  Country                     Country
+2     Year                        Year
+3   AGR_VA   Agriculture - Value Added
+4  AGR_EMP    Agriculture - Employment
+5   MIN_VA        Mining - Value Added
+6  MIN_EMP         Mining - Employment
+7   MAN_VA Manufacturing - Value Added
+8  MAN_EMP  Manufacturing - Employment
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<pivot_recast>>=
-head(dr <- pivot(data, c("Country", "Year"),
-             names = list(from = "Variable", to = "Sectorcode"),
-             labels = list(from = "Label", to = "Sector"), how = "r"))
-vlabels(dr)[3:6]
-@
+\begin{Schunk}
+\begin{Sinput}
+R> head(dr <- pivot(data, c("Country", "Year"),
++               names = list(from = "Variable", to = "Sectorcode"),
++               labels = list(from = "Label", to = "Sector"), how = "r"))
+\end{Sinput}
+\begin{Soutput}
+  Country Year Sectorcode      Sector    VA   EMP
+1     BWA 1964        AGR Agriculture 16.30 152.1
+2     BWA 1965        AGR Agriculture 15.73 153.3
+3     BWA 1966        AGR Agriculture 17.68 153.9
+4     BWA 1967        AGR Agriculture 19.15 155.1
+5     BWA 1968        AGR Agriculture 21.10 156.2
+6     BWA 1969        AGR Agriculture 21.86 157.4
+\end{Soutput}
+\begin{Sinput}
+R> vlabels(dr)[3:6]
+\end{Sinput}
+\begin{Soutput}
+   Sectorcode        Sector            VA           EMP 
+           NA            NA "Value Added"  "Employment" 
+\end{Soutput}
+\end{Schunk}
 %
@@
 The following example performs aggregation across years with the internal mean function during a recast pivot.
-<<pivot_agg>>=
-head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"), how = "r",
-     names = list(from = "Variable", to = "Sectorcode"),
-     labels = list(from = "Label", to = "Sector"), FUN = "mean"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"), how = "r",
++       names = list(from = "Variable", to = "Sectorcode"),
++       labels = list(from = "Label", to = "Sector"), FUN = "mean"))
+\end{Sinput}
+\begin{Soutput}
+  Country Sectorcode      Sector       VA      EMP
+1     BWA        AGR Agriculture    462.2   188.06
+2     ETH        AGR Agriculture  34389.9 17624.34
+3     GHA        AGR Agriculture   1549.4  3016.04
+4     KEN        AGR Agriculture 139705.9  5348.91
+5     MWI        AGR Agriculture  28512.6  2762.62
+6     MUS        AGR Agriculture   3819.6    59.34
+\end{Soutput}
+\end{Schunk}
 More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}.
@@
 %
-<<list_proc>>=
-dl <- GGDC10S |> rsplit( ~ Country + Variable)
-dl$ARG$EMP$AGR[1:12]
-@
+\begin{Schunk}
+\begin{Sinput}
+R> dl <- GGDC10S |> rsplit( ~ Country + Variable)
+R> dl$ARG$EMP$AGR[1:12]
+\end{Sinput}
+\begin{Soutput}
+ [1] 1800 1835 1731 2030 1889 1843 1789 1724 1678 1725 1650 1553
+\end{Soutput}
+\end{Schunk}
 %
@@
 Below, I write a nested loop running a regression of agriculture on mining and manuacturing output/employment:
-<<list_proc_2>>=
-result <- list()
-for (country in c("ARG", "BRA", "CHL")) {
-  for (variable in c("EMP", "VA")) {
-    m <- lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year,
-            data = dl[[country]][[variable]])
-    result[[country]][[variable]] <- list(model = m, BIC = BIC(m),
-                                          summary = summary(m))
-  }
-}
-@
+\begin{Schunk}
+\begin{Sinput}
+R> result <- list()
+R> for (country in c("ARG", "BRA", "CHL")) {
++    for (variable in c("EMP", "VA")) {
++      m <- lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year,
++              data = dl[[country]][[variable]])
++      result[[country]][[variable]] <- list(model = m, BIC = BIC(m),
++                                            summary = summary(m))
++    }
++  }
+\end{Sinput}
+\end{Schunk}
 This programming may not be ideal for this particular use case as I could have used data.frame-based tools and saved the result in a column.\footnote{E.g., \code{GGDC10S |> fgroup\_by(Country, Variable) |> fsummarise(result = my\_fun(lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year)))} with \code{my\_fun <- function(m) list(list(m, BIC(m), summary(m)))}.}
@@
 %
-<<list_proc_3>>=
-str(r_sq_l <- result |> get_elem("r.squared"))
-rowbind(r_sq_l, idcol = "Country", return = "data.frame")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> str(r_sq_l <- result |> get_elem("r.squared"))
+\end{Sinput}
+\begin{Soutput}
+List of 3
+ $ ARG:List of 2
+  ..$ EMP: num 0.907
+  ..$ VA : num 1
+ $ BRA:List of 2
+  ..$ EMP: num 0.789
+  ..$ VA : num 0.999
+ $ CHL:List of 2
+  ..$ EMP: num 0.106
+  ..$ VA : num 0.999
+\end{Soutput}
+\begin{Sinput}
+R> rowbind(r_sq_l, idcol = "Country", return = "data.frame")
+\end{Sinput}
+\begin{Soutput}
+  Country    EMP     VA
+1     ARG 0.9068 0.9996
+2     BRA 0.7888 0.9988
+3     CHL 0.1058 0.9991
+\end{Soutput}
+\end{Schunk}
 %
@@
 We can also apply \fct{t\_list} to turn the list inside-out.
-<<list_proc_3.5>>=
-r_sq_l |> t_list() |> rowbind(idcol = "Variable", return = "data.frame")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> r_sq_l |> t_list() |> rowbind(idcol = "Variable", return = "data.frame")
+\end{Sinput}
+\begin{Soutput}
+  Variable    ARG    BRA    CHL
+1      EMP 0.9068 0.7888 0.1058
+2       VA 0.9996 0.9988 0.9991
+\end{Soutput}
+\end{Schunk}
 \fct{rowbind} is limited if \fct{get\_elem} returns a more nested or asymmetric list, potentially with vectors/arrays in the final nodes. Suppose we want to extract the coefficient matrices:
-<<list_proc_4>>=
-result$ARG$EMP$summary$coefficients
-@
+\begin{Schunk}
+\begin{Sinput}
+R> result$ARG$EMP$summary$coefficients
+\end{Sinput}
+\begin{Soutput}
+              Estimate Std. Error  t value  Pr(>|t|)
+(Intercept)  26.583617  1.2832583  20.7157 1.747e-28
+log(MIN + 1)  0.083168  0.0352493   2.3594 2.169e-02
+log(MAN + 1) -0.064413  0.0767614  -0.8391 4.048e-01
+Year         -0.009683  0.0005556 -17.4278 1.003e-24
+\end{Soutput}
+\end{Schunk}
 For such cases, I created \fct{unlist2d} as a complete recursive generalization of \fct{unlist}.
@@
 %
-<<list_proc_5>>=
-result |> get_elem("coefficients") |> get_elem(is.matrix) |>
-  unlist2d(idcols = c("Country", "Variable"),
-           row.names = "Covariate") |> head(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> result |> get_elem("coefficients") |> get_elem(is.matrix) |>
++    unlist2d(idcols = c("Country", "Variable"),
++             row.names = "Covariate") |> head(3)
+\end{Sinput}
+\begin{Soutput}
+  Country Variable    Covariate Estimate Std. Error t value  Pr(>|t|)
+1     ARG      EMP  (Intercept) 26.58362    1.28326 20.7157 1.747e-28
+2     ARG      EMP log(MIN + 1)  0.08317    0.03525  2.3594 2.169e-02
+3     ARG      EMP log(MAN + 1) -0.06441    0.07676 -0.8391 4.048e-01
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<varying>>=
-varying(wlddev, ~ iso3c)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> varying(wlddev, ~ iso3c)
+\end{Sinput}
+\begin{Soutput}
+      country          date          year        decade        region 
+        FALSE          TRUE          TRUE          TRUE         FALSE 
+       income          OECD         PCGDP        LIFEEX          GINI 
+        FALSE         FALSE          TRUE          TRUE          TRUE 
+          ODA           POP  center_PCGDP center_LIFEEX 
+         TRUE          TRUE          TRUE          TRUE 
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<pdec>>=
-LIFEEXi <- reindex(wlddev$LIFEEX, wlddev$iso3c)
-all.equal(fvar(LIFEEXi), fvar(fbetween(LIFEEXi)) + fvar(fwithin(LIFEEXi)))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> LIFEEXi <- reindex(wlddev$LIFEEX, wlddev$iso3c)
+R> all.equal(fvar(LIFEEXi), fvar(fbetween(LIFEEXi)) + fvar(fwithin(LIFEEXi)))
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<qsu_2>>=
-qsu(LIFEEXi, g = wlddev$OECD, w = wlddev$POP) |> aperm()
-@
+\begin{Schunk}
+\begin{Sinput}
+R> qsu(LIFEEXi, g = wlddev$OECD, w = wlddev$POP) |> aperm()
+\end{Sinput}
+\begin{Soutput}
+, , FALSE
+
+             N/T       WeightSum     Mean      SD      Min      Max
+Overall     9503  2.48998382e+11  63.5476  9.2368   18.907  85.4171
+Between      171             171  63.5476  6.0788  43.0905  85.4171
+Within   55.5731  1.45613089e+09  65.8807  6.9545  30.3388  82.8832
+
+, , TRUE
+
+             N/T       WeightSum     Mean      SD      Min      Max
+Overall     2156  6.38797019e+10  74.9749  5.3627   45.369  84.3563
+Between       36              36  74.9749  2.9256  66.2983  78.6733
+Within   59.8889  1.77443616e+09  65.8807  4.4944  44.9513  77.2733
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<descr>>=
-wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
-wlda15 |> descr(income + LIFEEX ~ OECD)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
+R> wlda15 |> descr(income + LIFEEX ~ OECD)
+\end{Sinput}
+\begin{Soutput}
+Dataset: wlda15, 2 Variables, N = 216
+Grouped by: OECD [2]
+         N   Perc
+FALSE  180  83.33
+TRUE    36  16.67
+-----------------------------------------------------------------------------
+income (factor): Income Level
+Statistics (N = 216)
+         N   Perc  Ndist
+FALSE  180  83.33      4
+TRUE    36  16.67      2
+
+Table (Freq Perc)
+                       FALSE     TRUE    Total
+High income          45 25.0  34 94.4  79 36.6
+Upper middle income  58 32.2   2  5.6  60 27.8
+Lower middle income  47 26.1   0  0.0  47 21.8
+Low income           30 16.7   0  0.0  30 13.9
+-----------------------------------------------------------------------------
+LIFEEX (numeric): Life expectancy at birth, total (years)
+Statistics (N = 200, 7.41% NAs)
+         N  Perc  Ndist   Mean    SD    Min    Max   Skew  Kurt
+FALSE  164    82    164  71.25  7.06  53.28  85.08   -0.5  2.61
+TRUE    36    18     36  80.83  2.55  75.05  84.36  -0.92  2.72
+
+Quantiles
+          1%     5%    10%    25%    50%    75%    90%    95%    99%
+FALSE   54.3  58.38  61.25  66.43  72.56  76.68  78.93  80.88  83.77
+TRUE   75.12  75.83   76.8  79.04  81.77  82.63  83.21  83.54  84.13
+-----------------------------------------------------------------------------
+\end{Soutput}
+\end{Schunk}
 
@@
 %
-<<qtab>>=
-wlda15 |> with(qtab(OECD, income))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlda15 |> with(qtab(OECD, income))
+\end{Sinput}
+\begin{Soutput}
+       income
+OECD    High income Low income Lower middle income Upper middle income
+  FALSE          45         30                  47                  58
+  TRUE           34          0                   0                   2
+\end{Soutput}
+\end{Schunk}
 This shows the total population (latest post-2015 estimates) in millions.
-<<qtab_2>>=
-wlda15 |> with(qtab(OECD, income, w = POP) / 1e6)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlda15 |> with(qtab(OECD, income, w = POP) / 1e6)
+\end{Sinput}
+\begin{Soutput}
+       income
+OECD    High income Low income Lower middle income Upper middle income
+  FALSE       93.01     694.89             3063.54             2459.71
+  TRUE      1098.75       0.00                0.00              211.01
+\end{Soutput}
+\end{Schunk}
 This shows the average life expectancy in years. The use of \fct{fmean} toggles an efficient vectorized computation of the table entries (i.e., \fct{fmean} is only called once).
-<<qtab_3>>=
-wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean))
+\end{Sinput}
+\begin{Soutput}
+       income
+OECD    High income Low income Lower middle income Upper middle income
+  FALSE       78.75      62.81               68.30               73.81
+  TRUE        81.09                                              76.37
+\end{Soutput}
+\end{Schunk}
 Finally, this calculates a population-weighted average of life expectancy in each group.
-<<qtab_4>>=
-wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean,
-                    wFUN.args = list(w = POP)))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlda15 |> with(qtab(OECD, income, w = LIFEEX, wFUN = fmean,
++                      wFUN.args = list(w = POP)))
+\end{Sinput}
+\begin{Soutput}
+       income
+OECD    High income Low income Lower middle income Upper middle income
+  FALSE       77.91      63.81               68.76               75.93
+  TRUE        81.13                                              76.10
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_1>>=
-set.seed(101)
-int <- 1:1000; g_int <- sample.int(1000, 1e7, replace = TRUE)
-char <- c(letters, LETTERS, month.abb, month.name)
-g_char <- sample(char <- outer(char, char, paste0), 1e7, TRUE)
-bmark(base_int = unique(g_int), collapse_int = funique(g_int))
-bmark(base_char = unique(g_char), collapse_char = funique(g_char))
-bmark(base_int = match(g_int, int), collapse_int = fmatch(g_int, int))
-bmark(base_char = match(g_char, char), data.table_char =
-      chmatch(g_char, char), collapse_char = fmatch(g_char, char))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(101)
+R> int <- 1:1000; g_int <- sample.int(1000, 1e7, replace = TRUE)
+R> char <- c(letters, LETTERS, month.abb, month.name)
+R> g_char <- sample(char <- outer(char, char, paste0), 1e7, TRUE)
+R> bmark(base_int = unique(g_int), collapse_int = funique(g_int))
+\end{Sinput}
+\begin{Soutput}
+    expression     min  median mem_alloc n_itr n_gc total_time
+1     base_int 63.27ms 65.41ms   166.2MB    29   29      2.05s
+2 collapse_int  8.36ms  8.62ms    38.2MB   217   44         2s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(base_char = unique(g_char), collapse_char = funique(g_char))
+\end{Sinput}
+\begin{Soutput}
+     expression     min  median mem_alloc n_itr n_gc total_time
+1     base_char 101.5ms 101.5ms   166.2MB     1   19    101.5ms
+2 collapse_char  22.4ms  23.5ms    38.2MB    69   12       1.7s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(base_int = match(g_int, int), collapse_int = fmatch(g_int, int))
+\end{Sinput}
+\begin{Soutput}
+    expression     min  median mem_alloc n_itr n_gc total_time
+1     base_int  26.6ms 26.96ms    76.3MB    33   33   908.78ms
+2 collapse_int  7.99ms  8.29ms    38.2MB   182   37      1.51s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(base_char = match(g_char, char), data.table_char =
++        chmatch(g_char, char), collapse_char = fmatch(g_char, char))
+\end{Sinput}
+\begin{Soutput}
+       expression    min median mem_alloc n_itr n_gc total_time
+1       base_char 57.1ms 57.9ms   114.5MB     9   22   520.29ms
+2 data.table_char 40.5ms 41.5ms    38.1MB    40    8      1.66s
+3   collapse_char 11.9ms 12.3ms    38.1MB   126   26      1.55s
+\end{Soutput}
+\end{Schunk}
 %
@@
 They are faster than base \proglang{R} even without multithreading, but using 4 threads in this case induces a sizeable difference.
-<<bench_2, cache=TRUE>>=
-set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)
-m <- matrix(rnorm(1e7), ncol = 1000)
-bmark(R = colSums(m), collapse = fsum(m))
-bmark(R = colMeans(m), collapse = fmean(m))
-bmark(MS = matrixStats::colMedians(m), collapse = fmedian(m))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)
+R> m <- matrix(rnorm(1e7), ncol = 1000)
+R> bmark(R = colSums(m), collapse = fsum(m))
+\end{Sinput}
+\begin{Soutput}
+  expression    min median mem_alloc n_itr n_gc total_time
+1          R 9.19ms 9.25ms    7.86KB   215    0      2.01s
+2   collapse  1.3ms 1.34ms    7.86KB  1480    0         2s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(R = colMeans(m), collapse = fmean(m))
+\end{Sinput}
+\begin{Soutput}
+  expression    min median mem_alloc n_itr n_gc total_time
+1          R  9.2ms 9.25ms    7.86KB   215    0         2s
+2   collapse 1.29ms 1.34ms    7.86KB  1477    1         2s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(MS = matrixStats::colMedians(m), collapse = fmedian(m))
+\end{Sinput}
+\begin{Soutput}
+  expression   min  median mem_alloc n_itr n_gc total_time
+1         MS 106ms 106.3ms   86.04KB    19    0      2.04s
+2   collapse  25ms  25.3ms    7.86KB    79    0      2.01s
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_2.1, cache=TRUE>>=
-g <- sample.int(1e3, 1e4, TRUE)
-bmark(R = rowsum(m, g), collapse = fsum(m, g))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> g <- sample.int(1e3, 1e4, TRUE)
+R> bmark(R = rowsum(m, g), collapse = fsum(m, g))
+\end{Sinput}
+\begin{Soutput}
+  expression    min median mem_alloc n_itr n_gc total_time
+1          R 7.06ms 7.39ms    7.85MB   245   12      1.85s
+2   collapse 1.76ms    2ms    7.67MB   799   38      1.69s
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_flights_setup, cache=TRUE>>=
-fastverse_extend(nycflights23, dplyr, data.table); setDTthreads(4)
-list(flights, airports, airlines, planes, weather) |> sapply(nrow)
-flights |> fselect(month, day, origin, dest) |> fnunique()
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fastverse_extend(nycflights23, dplyr, data.table); setDTthreads(4)
+R> list(flights, airports, airlines, planes, weather) |> sapply(nrow)
+\end{Sinput}
+\begin{Soutput}
+[1] 435352   1251     14   4840  26204
+\end{Soutput}
+\begin{Sinput}
+R> flights |> fselect(month, day, origin, dest) |> fnunique()
+\end{Sinput}
+\begin{Soutput}
+[1] 75899
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_flights_summarise, cache=TRUE>>=
-vars <- .c(dep_delay, arr_delay, air_time, distance, hour, minute)
-bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
-                summarise(across(all_of(vars), sum), .groups = "drop"),
-      data.table = qDT(flights)[, lapply(.SD, sum), .SDcols = vars,
-                                by = .(month, day, origin, dest)],
-      collapse = flights |> fgroup_by(month, day, origin, dest) |>
-                   get_vars(vars) |> fsum())
-@
+\begin{Schunk}
+\begin{Sinput}
+R> vars <- .c(dep_delay, arr_delay, air_time, distance, hour, minute)
+R> bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
++                  summarise(across(all_of(vars), sum), .groups = "drop"),
++        data.table = qDT(flights)[, lapply(.SD, sum), .SDcols = vars,
++                                  by = .(month, day, origin, dest)],
++        collapse = flights |> fgroup_by(month, day, origin, dest) |>
++                     get_vars(vars) |> fsum())
+\end{Sinput}
+\begin{Soutput}
+  expression     min   median mem_alloc n_itr n_gc total_time
+1      dplyr 464.4ms 543.81ms   51.46MB     4   32      2.19s
+2 data.table  10.2ms   11.1ms   18.93MB   161   23      2.01s
+3   collapse   4.6ms   5.05ms    9.11MB   355   20         2s
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_flights_summarise_2, cache=TRUE, echo=FALSE>>=
-bmark(dplyr_mean = flights |> group_by(month, day, origin, dest) |>
-              summarise(across(all_of(vars), mean), .groups = "drop"),
-      data.table_mean = qDT(flights)[, lapply(.SD, mean), .SDcols = vars,
-                                by = .(month, day, origin, dest)],
-      collapse_mean = flights |> fgroup_by(month, day, origin, dest) |>
-                 get_vars(vars) |> fmean())
-bmark(dplyr_median = flights |> group_by(month, day, origin, dest) |>
-              summarise(across(all_of(vars), median), .groups = "drop"),
-      data.table_median = qDT(flights)[, lapply(.SD, median), .SDcols = vars,
-                                by = .(month, day, origin, dest)],
-      collapse_median = flights |> fgroup_by(month, day, origin, dest) |>
-                 get_vars(vars) |> fmedian())
-@
+\begin{Schunk}
+\begin{Soutput}
+       expression     min  median mem_alloc n_itr n_gc total_time
+1      dplyr_mean   1.48s   1.53s   51.46MB     2   42      3.07s
+2 data.table_mean 10.41ms 11.46ms   18.93MB   152   20         2s
+3   collapse_mean  4.77ms  5.12ms    9.11MB   349   20         2s
+\end{Soutput}
+\begin{Soutput}
+         expression     min  median mem_alloc n_itr n_gc total_time
+1      dplyr_median   5.62s   5.62s    55.7MB     1   91      5.62s
+2 data.table_median 29.19ms 30.27ms    18.9MB    65    5      2.02s
+3   collapse_median  13.6ms  14.6ms    11.1MB   133   10      2.01s
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_flights_summarise_3, cache=TRUE>>=
-bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
-    summarise(rng = max(arr_delay) - min(arr_delay), .groups = "drop"),
-  data.table = qDT(flights)[, .(rng = max(arr_delay) - min(arr_delay)),
-                            by = .(month, day, origin, dest)],
-  collapse = flights |> fgroup_by(month, day, origin, dest) |>
-    fsummarise(rng = fmax(arr_delay) - fmin(arr_delay)))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
++      summarise(rng = max(arr_delay) - min(arr_delay), .groups = "drop"),
++    data.table = qDT(flights)[, .(rng = max(arr_delay) - min(arr_delay)),
++                              by = .(month, day, origin, dest)],
++    collapse = flights |> fgroup_by(month, day, origin, dest) |>
++      fsummarise(rng = fmax(arr_delay) - fmin(arr_delay)))
+\end{Sinput}
+\begin{Soutput}
+  expression      min   median mem_alloc n_itr n_gc total_time
+1      dplyr 121.05ms 134.78ms   23.08MB    13   29      2.03s
+2 data.table   77.8ms  82.68ms    5.77MB    24   15      2.01s
+3   collapse   6.73ms   7.25ms     6.8MB   264    7         2s
+\end{Soutput}
+\end{Schunk}
 %
@@
 \end{Code}
-<<bench_flights_join_setup, cache=TRUE, echo=FALSE>>=
-flights |> join(weather, on = c("origin", "time_hour")) |>
-  join(planes, on = "tailnum") |> join(airports, on = c(dest = "faa")) |>
-  join(airlines, on = "carrier") |> dim() |>
-  capture.output() |> substr(1, 76) |> cat(sep = "\n")
-@
+\begin{Schunk}
+\begin{Soutput}
+left join: flights[origin, time_hour] 434462/435352 (99.8%) <21.94:1st> weat
+duplicate columns: year, month, day, hour => renamed using suffix '_weather'
+left join: x[tailnum] 424068/435352 (97.4%) <87.62:1st> planes[tailnum] 4840
+duplicate columns: year => renamed using suffix '_planes' for y
+left join: x[dest] 427868/435352 (98.3%) <3753.23:1st> airports[faa] 114/125
+left join: x[carrier] 435352/435352 (100%) <31096.57:1st> airlines[carrier] 
+duplicate columns: name => renamed using suffix '_airlines' for y
+[1] 435352     48
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_flights_join, cache=TRUE, echo=FALSE>>=
-bmark(
-  dplyr_joins = flights |>
-    left_join(weather, by = c("origin", "time_hour"), multiple = "first") |>
-    left_join(planes, by = "tailnum", multiple = "first") |>
-    left_join(airports, by = c(dest = "faa"), multiple = "first") |>
-    left_join(airlines, by = "carrier", multiple = "first"),
-
-  data.table_joins = qDT(airlines)[qDT(airports)[qDT(planes)[qDT(weather)[qDT(flights), on = c("origin", "time_hour"), mult = "first"], on = "tailnum", mult = "first"], on = c("faa" = "dest"), mult = "first"], on = "carrier", mult = "first"],
-
-  collapse_joins = flights |>
-    join(weather, on = c("origin", "time_hour"), verbose = 0) |>
-    join(planes, on = "tailnum", verbose = 0) |>
-    join(airports, on = c("dest" = "faa"), verbose = 0) |>
-    join(airlines, on = "carrier", verbose = 0)
-)
-@
+\begin{Schunk}
+\begin{Soutput}
+        expression     min  median mem_alloc n_itr n_gc total_time
+1      dplyr_joins 183.2ms 220.5ms   558.9MB     9   19      2.07s
+2 data.table_joins 130.8ms 184.1ms   490.4MB    12   27       2.2s
+3   collapse_joins  12.5ms  15.5ms    89.7MB   116   33      2.01s
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_pivot_long, cache=TRUE, max.print=50>>=
-bmark(tidyr = tidyr::pivot_longer(flights, cols = vars),
-      data.table = qDT(flights) |> melt(measure = vars),
-      collapse = pivot(flights, values = vars))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> bmark(tidyr = tidyr::pivot_longer(flights, cols = vars),
++        data.table = qDT(flights) |> melt(measure = vars),
++        collapse = pivot(flights, values = vars))
+\end{Sinput}
+\begin{Soutput}
+  expression    min median mem_alloc n_itr n_gc total_time
+1      tidyr 82.6ms 82.7ms     251MB     5   23      416ms
+2 data.table 50.9ms   52ms     209MB    11   18      579ms
+3   collapse 17.2ms 18.5ms     209MB    24   43      456ms
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_pivot_wide, cache=TRUE, max.print=50>>=
-bmark(tidyr = tidyr::pivot_wider(flights, id_cols = .c(month, day, dest),
-          names_from = "origin", values_from = vars, values_fn = sum),
-      data.table = dcast(qDT(flights), month + day + dest ~ origin,
-                         value.var = vars, fun = sum),
-      collapse_fsum = pivot(flights, .c(month, day, dest), vars,
-                            "origin", how = "wider", FUN = fsum),
-      collapse_itnl = pivot(flights, .c(month, day, dest), vars,
-                            "origin", how = "wider", FUN = "sum"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> bmark(tidyr = tidyr::pivot_wider(flights, id_cols = .c(month, day, dest),
++            names_from = "origin", values_from = vars, values_fn = sum),
++        data.table = dcast(qDT(flights), month + day + dest ~ origin,
++                           value.var = vars, fun = sum),
++        collapse_fsum = pivot(flights, .c(month, day, dest), vars,
++                              "origin", how = "wider", FUN = fsum),
++        collapse_itnl = pivot(flights, .c(month, day, dest), vars,
++                              "origin", how = "wider", FUN = "sum"))
+\end{Sinput}
+\begin{Soutput}
+     expression      min   median mem_alloc n_itr n_gc total_time
+1         tidyr 452.49ms 496.93ms   142.4MB     4   25      2.03s
+2    data.table 299.13ms 304.01ms      21MB     7   25      2.19s
+3 collapse_fsum   8.65ms  10.23ms    39.1MB   169   12         2s
+4 collapse_itnl   7.06ms   7.69ms    12.4MB   249    5         2s
+\end{Soutput}
+\end{Schunk}
 %
@@
 
-In summary, these benchmarks show that \pkg{collapse} provides outstanding performance and memory efficiency on a typical medium-sized real-world database popular in the \proglang{R} community.
+In summary, these benchmarks shows that \pkg{collapse} provides outstanding performance and memory efficiency on a typical medium-sized real-world database popular in the \proglang{R} community.
 % %

### 2025-04-08 misc/JSS article/R&R/article.Rnw
Fix computation of RCA indices.

@@
 <<RCA>>=
-with(latest, fsum(v, c, TRA = "/") / fsum(v, s, TRA = "/"))[1:10]
+with(latest, fsum(v, c, TRA = "/") / fsum(fsum(v, TRA = "/"), s, TRA = "fill"))[1:10]
 @
@@
 
-Last but not least, I benchmark pivots, starting a with long pivot that simply melt the 6 columns aggregated beforehand into one column, duplicating all other columns 6 times:
+Last but not least, I benchmark pivots, starting a with long pivot that simply melts the 6 columns aggregated beforehand into one column, duplicating all other columns 6 times:
 %

### 2025-05-02 misc/JSS article/R&R/article.Rnw
Minors.

@@
 In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recast pivots and recursive operations on lists, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable.
-In short, it is very utile for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem.
+In short, it is very useful for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem.
 On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes.
@@
 The expression \code{fsum(v, dmy) \%/=\% fsum(dmy, dmy)} amounts to \code{cov(v, y)/var(y)}, but is both vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation for the final result.
-I do not recommend forecasting trade in this way, rather, this example is inspired by a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{bog post} where I forecasted high-resolution (1km2) population estimates for South Africa.
+I do not recommend forecasting trade in this way, rather, this example is inspired by a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post} where I forecasted high-resolution (1$km^2$) population estimates for South Africa.
 The data, taken from WorldPop, was available for the years 2014-2020, and I needed estimates for 2021-2022.

### 2025-05-02 misc/JSS article/R&R/article_replication.R

@@ -78,7 +78,7 @@ with(latest, fndistinct(s, c))
 ###################################################
 ### code chunk number 7: RCA
 ###################################################
-with(latest, fsum(v, c, TRA = "/") / fsum(v, s, TRA = "/"))[1:10]
+with(latest, fsum(v, c, TRA = "/") / fsum(fsum(v, TRA = "/"), s, TRA = "fill"))[1:10]
 
 
 ###################################################

### 2025-05-05 misc/JSS article/final/article.Rnw
Add final article revision (hopefully).

--- 2025-05-02 misc/JSS article/R&R/article.Rnw
@@
 # Package versions used in the article:
-# fastverse 0.3.4, collapse 2.0.19, data.table 1.16.4,
+# fastverse 0.3.4, collapse 2.1.2, data.table 1.16.4,
 # bench 1.1.3, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.0.0
@@
 bmark <- function(...) {
-  bench::mark(..., min_time = 2, check = FALSE) |> janitor::clean_names() |>
+  bench::mark(..., min_time = 0.01, check = FALSE) |> janitor::clean_names() |>
   fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
@@
 }
+
 @
@@
 \href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/.
-Linecount (v2.0.19): R: 13,646, C: 18,594, C++: 9,844.
-\hphantom{aaa.}Exported namespace: 391 objects, of which 237 functions (excl.
+Linecount (v2.1.2): R: 13,785, C: 19,013, C++: 9,844.
+\hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl.
 methods and shorthands), and 2 datasets.}
 % Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
-Core functionality includes a rich set of S3 generic (grouped and weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}).
-It also provides functions and classes for fully indexed (time-aware) computations on time series and panel data, advanced descriptive statistical tools, recursive tools to deal with nested data, and powerful data manipulation functions---such as vectorized and verbose hash-joins or fast aggregation and recast pivots.
-This functionality is supported by efficient algorithms for grouping, ordering, deduplication, and matching callable at \proglang{R} and \proglang{C} levels.
-The package also provides efficient data object conversions, functions for memory efficient R programming, such as (grouped) transformation and math by reference, and helpers to effectively deal with variable labels, attributes, and missing data.
-\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem---notably, \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
-It is globally and interactively configurable, which includes setting different defaults for key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the package namespace to mask equivalent but slower base \proglang{R} or \pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
-See Section~\ref{sec:glob_opt}.\vspace{-5mm}} \newline
+Its core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}).
+It also provides powerful data manipulation functions---such as vectorized and verbose hash-joins or fast (aggregation, recast) pivots---functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools.
+This functionality is supported by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels.
+Lastly, it includes efficient object converters, functions for memory efficient R programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
+\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations for vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem---notably, \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
+It is also globally and interactively configurable, allowing changing defaults of key function arguments, such as the \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the namespace to mask equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
+See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline
 
@@
 The short answer is to make computations in \proglang{R} as flexible and powerful as possible.
-The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase the performance and memory efficiency of \proglang{R} programs;\footnote{Principally by avoiding \proglang{R}-level repetition such as applying \proglang{R} functions across columns/groups using a split-apply-combine logic, but also by avoiding object conversions and the need for certain classes to do certain things, such as converting matrices to \class{data.frame} or \class{data.table} just to compute statistics by groups.} and (3) to create a new foundation package for statistics and data transformation in \proglang{R} that implements many successful ideas developed in the \proglang{R} ecosystem and other programming environments such as \proglang{Python} or \proglang{STATA} \citep{STATA}, in a stable, high performance, and broadly compatible manner.\footnote{Such ideas include \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), variable labels (\proglang{STATA}), recast pivots (\pkg{reshape(2)}), etc...} \newline
+The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase the performance and memory efficiency of \proglang{R} programs; % \footnote{Principally by avoiding \proglang{R}-level repetition such as applying \proglang{R} functions across columns/groups using a split-apply-combine logic, but also by avoiding object conversions and the need for certain classes to do certain things, such as converting matrices to \class{data.frame} or \class{data.table} just to compute statistics by groups.}
+and (3) create a new foundation package for statistics and data transformation that implements successful ideas in the \proglang{R} ecosystem and other programming environments in a stable, high performance, and broadly compatible manner.\footnote{Such ideas include \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}: \citet{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), variable labels (\proglang{STATA}), recast pivots (\pkg{reshape2}), etc...} \newline
 
@@
 Notably, the \pkg{tidyverse} \citep{rtidyverse} provides a consistent toolkit for data manipulation in R, centered around the \class{tibble} \citep{rtibble} object and tidy data principles \citep{rtidydata}.
-\pkg{data.table} \citep{rdatatable} provides an enhanced high-performance data frame with parsimonious data manipulation syntax.
+\pkg{data.table} \citep{rdatatable} provides an enhanced high-performance data frame with parsimonious syntax.
 \pkg{sf} \citep{rsf} provides a data frame for spatial data and supporting functionality.
@@
 Econometric packages like \pkg{plm} \citep{rplm} and \pkg{fixest} \citep{rfixest} also provide solutions to deal with panel data and irregularity in the time dimension.
-Packages like \pkg{matrixStats} \citep{rmatrixstats} and \pkg{Rfast} \citep{rfast} offer fast statistical calculations along the rows and columns of matrices as well as faster basic statistical procedures.
+Packages like \pkg{matrixStats} \citep{rmatrixstats} and \pkg{Rfast} \citep{rfast} offer fast statistical calculations along the rows and columns of matrices as well as faster statistical procedures.
 \pkg{DescTools} \citep{rdesctools} provides a wide variety of descriptive statistics, including weighted versions.
-\pkg{survey} \citep{rsurvey} allows statistical computations on complex survey data.
+\pkg{survey} \citep{rsurvey} offers computations on complex surveys.
 \pkg{labelled} \citep{rlabelled} provides tools to deal with labelled data.
-Packages like \pkg{tidyr} \citep{rtidyr}, \pkg{purrr} \citep{rpurrr} and \pkg{rrapply} \citep{rrapply} provide some functions to deal with nested data and messy structures.
+Packages like \pkg{tidyr} \citep{rtidyr}, \pkg{purrr} \citep{rpurrr} and \pkg{rrapply} \citep{rrapply} provide functions to deal with nested or messy data.
 \newline
@@
 It provides novel recursive tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}.
-Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specific statistics.
-
-In summary, \pkg{collapse} is a foundation package for statistical computing and data transformation in \proglang{R} that enhances and integrates seamlessly with the \proglang{R} ecosystem while being outstandingly computationally efficient.
-A significant benefit is that, rather than piecing together a fragmented ecosystem oriented at different classes and tasks, many core computational tasks can be done with \pkg{collapse}, and easily extended by more specialized packages.
-This tends to result in \proglang{R} scripts that are shorter, more efficient, and more lightweight in dependencies.
+Finally, it provides a small but consistent and powerful set of descriptive statistical tools, yielding sufficient detail for most data exploration purposes, requiring users to invoke packages like \pkg{DescTools} or \pkg{survey} only for specialized statistics.
+\newline
+
+In summary, \pkg{collapse} is a new foundation package for advanced statistical computing and data transformation in \proglang{R} that integrates seamlessly with the ecosystem and offers outstanding speed and memory efficiency.
+Thus, many core tasks can be done with \pkg{collapse}, and easily extended by specialized packages, yielding more lightweight, faster, and shorter \proglang{R} programs.
 \newline
@@
 % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
-In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including various quantile and mode estimators, support for fully time-aware computations on irregular series/panels, higher order centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values.} supports recast pivots and recursive operations on lists, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable.
+In comparison with these, \pkg{collapse} offers a class-agnostic approach bridging the divide between data frames and atomic structures, has more advanced statistical capabilities,\footnote{Such as weighted statistics, including several (weighted) quantile and mode estimators, support for fully time-aware computations on irregular series/panels, scaling and centering, advanced (grouped, weighted, panel-decomposed) descriptive statistics etc., all supporting missing values and vectors/matrices/data frames.} supports recast pivots and recursive operations on lists, variable labels, verbosity for critical operations such as joins, and is extensively globally configurable.
 In short, it is very useful for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem.
 On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes.
-There may be performance improvements for "long data" in the future, but, at present, the treatment of columns as fundamental units of computation (in most cases) is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any \proglang{Julia} function.
-However, as demonstrated in Section~\ref{ssec:vfat}, vectorized statistical functions can be combined to calculate more complex statistics in a vectorized way.
-\newline
-
-The package has a built-in structured \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{documentation} facilitating its use.
+There may be performance improvements for very "long data" in the future, but, at present, the treatment of columns as fundamental units of computation (in most cases) is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any \proglang{Julia} function.
+However, as demonstrated in Section~\ref{sec:integration}, vectorized statistical functions can be combined to calculate more complex statistics in a vectorized way.
+\newline
+
+The package has a built-in structured \href{https://sebkrantz.github.io/collapse/reference/index.html}{documentation} facilitating its use.
 This documentation includes a central \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and supplementary topic pages which briefly describe related functionality.
@@
 %
-This article does not fully present \pkg{collapse}, but the following sections introduce its key components, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options.
+This article cannot fully present \pkg{collapse}, but the following sections introduce its key components, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options.
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} (\href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}).
+For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}, including the \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} $+$ \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
 %
@@
 %
-The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R} means that these operations are implemented using compiled C/C++ code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
-The functions basic syntax is
+The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R}/\pkg{collapse} means that these operations are implemented using compiled C/C++ code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
+Their basic syntax is
 \begin{Code}
@@
 with arguments \code{x} - data (vector, matrix or data frame-like), \code{g} - groups (atomic vector, list of vectors, or \class{GRP} object), \code{w} - sampling weights (only some functions), and \code{TRA} - transformation of \code{x}.
-The following examples using \fct{fmean} demonstrate their basic usage on the familiar \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/iris}{\code{iris}} dataset, recording 50 measurements of 4 variables for 3 species of iris flowers.
-All examples support weights (\code{w}), and \fct{fmean} can also be multithreaded across columns (\code{nthreads}).\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions, as detailed in the respective function documentation.
+The following examples with \fct{fmean} demonstrate their basic usage on the familiar \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/iris}{\code{iris}} dataset recording 50 measurements of 4 variables for 3 species of iris flowers.\footnote{\fct{num\_vars} returns the numeric variables in a data frame-like object; \fct{cat\_vars} the categorical ones.} All examples support weights (\code{w}), and \fct{fmean} can also be multithreaded across columns (\code{nthreads}).\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions, as detailed in the respective function documentation.
 The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes.
@@
 fmean(iris$Sepal.Length)
-fmean(iris[1:4])
-identical(fmean(iris[1:4]), fmean(as.matrix(iris[1:4])))
+fmean(num_vars(iris))
+identical(fmean(num_vars(iris)), fmean(as.matrix(num_vars(iris))))
 fmean(iris$Sepal.Length, g = iris$Species)
-fmean(iris[1:4], g = iris$Species, nthreads = 4)
-fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill")[1:10]
+fmean(num_vars(iris), g = iris$Species, nthreads = 4)
 @
@@
 \subsection{Transformations}
-The final example expands the mean vector to full length---like \code{ave(x, g)} but much faster.
-The \code{TRA} argument toggles (grouped) replacing and sweeping operations (by reference), generalizing \code{sweep(x, 2, STATS = fmean(x))}.\footnote{The \code{TRA} argument internally calls \fct{TRA}: \code{TRA(x, STATS, FUN = "-", g = NULL, set = FALSE, ...)}.} %Its syntax is
+The \code{TRA} argument toggles (grouped) replacing and sweeping operations (by reference), generalizing \code{sweep(x, 2, STATS = fmean(x))}.\footnote{The \code{TRA} argument internally calls \fct{TRA}: \code{TRA(x, STATS, FUN = "-", g = NULL, set = FALSE, ...)}.
+\vspace{-12mm}} %Its syntax is
 %\begin{Code}
@@
 %
+For example, option \code{TRA = "fill"} replaces elements with the corresponding statistics:
+<<TRA>>=
+fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill")[c(1:5, 51:55)]
+@
 %\code{TRA()} is called internally in the \emph{Fast Statistical Functions}, the \code{TRA} argument is passed to \code{FUN}.
 Thus \code{fmean(x, g, w, TRA = "-")} is equivalent to \code{TRA(x, fmean(x, g, w), "-", g)}.
-Additionally, a \code{set} argument can be passed to \emph{Fast Statistical Functions} to toggle transformation by reference, e.g., \code{fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill", set = TRUE)} would modify \code{Sepal.Length} in-place and return the result invisibly.
+Additionally, a \code{set} argument can be passed to \emph{Fast Statistical Functions} to toggle transformation by reference.
+For example \code{fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill", set = TRUE)} would modify \code{Sepal.Length} in-place and return the result invisibly.
 \newline
@@
 Having grouping and data transformation functionality directly built into generic statistical functions facilitates and speeds up many common operations.
-Take for example this generated sector-level trade dataset of export values (v) by country (c), sector (s), and year (y).\footnote{\label{fn:tfm}\fct{tfm} abbreviates \fct{ftransform}---a faster equivalent of \fct{transform}.
-\fct{ss} replaces \code{[.data.frame}.}
-
-<<exports>>=
-set.seed(101)
-exports <- expand.grid(y=1:10, c=paste0("c",1:10), s=paste0("s",1:10)) |>
-  tfm(v = abs(rnorm(1e3))) |> colorder(c, s) |> ss(-sample.int(1e3, 500))
-@
-Like any real trade dataset, it is unbalanced---\code{ss(-sample.int(1e3, 500))} randomly removes 500 rows. Suppose we wanted to extract the latest trade within the last two years.
-<<exports_2>>=
-latest <- fsubset(exports, y >= 8 & y == fmax(y, list(c, s), "fill"), -y)
-@
-Below, I compute how many products different countries have exported in the last two years
-<<prod>>=
-with(latest, fndistinct(s, c))
-@
-and this computes \citet{balassa1965tariff}'s index of Revealed Comparative Advantage, defined as the share of a sector in country exports divided by the share of the sector in world exports.
-<<RCA>>=
-with(latest, fsum(v, c, TRA = "/") / fsum(fsum(v, TRA = "/"), s, TRA = "fill"))[1:10]
-@
-More complex use cases are frequent in my work.
-For example, I recently combined multiple spatial datasets on points of interest (POIs).
-In the face of significant duplicates and problems matching POIs directly across datasets, I decided to keep the richest source for each location and POI type.
-After creating POI confidence, location, and type indicators comparable across sources (datasets), my deduplication expression was \code{fsubset(data\_long, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (i.e., richest) source by location and type.
-
-\subsection{Grouping objects and optimization} \label{ssec:gopt}
-%
-Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, for repeated operations the cost of grouping can be minimized by using factors (see \code{?qF}) or \class{GRP} objects as inputs.
-The latter contain all information \pkg{collapse}'s statistical functions may require to operate across groups.
-They can be created with \code{GRP()}.
-Its basic syntax is
-\begin{Code}
-GRP(X, by = NULL, sort = TRUE, return.groups = TRUE, method = "auto", ...)
-\end{Code}
- Below, I create a \class{GRP} object from the included World Development Dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}).
-The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.
-%
-<<GRP>>=
-str(g <- GRP(wlddev, ~ income + OECD))
-@
-%
-\class{GRP} objects make grouped statistical computations in \pkg{collapse} fully programmable, e.g., I can employ the object with the \emph{Fast Statistical Functions} and some utilities\footnote{\fct{add\_vars} is a fast \fct{cbind.data.frame} which also has an assignment method, and \fct{get\_vars} enables fast and secure extraction of data frame columns.} to efficiently aggregate GDP per capita, life expectancy, and country name, with population weights.
-%
-<<gcomp>>=
-add_vars(g$groups,
- get_vars(wlddev, "country") |> fmode(g, wlddev$POP, use = FALSE),
- get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = F))
-@
-For advanced data aggregation, \pkg{collapse} also provides a convenience function, \fct{collap}, which, by default, uses \fct{fmean} for numeric, \fct{fmode} for categorical, and \fct{fsum} for weight columns, and preserves their order.
-The equivalent expression using this function would be
-%
-<<collap>>=
-collap(wlddev, country + PCGDP + LIFEEX ~ income + OECD, w = ~ POP)
-@
-%
-Similarly, data can be transformed, here using \fct{fmean} to center the data by country to level differences in average economic status, adding back the overall mean across countries.\footnote{\fct{add\_stub} adds a prefix (or suffix if \code{pre = FALSE}) to columns ($\to$ \code{center\_PCGDP} and \code{center\_LIFEEX}).}
-%
-<<gtrans>>=
-add_vars(wlddev) <- get_vars(wlddev, c("PCGDP", "LIFEEX")) |>
-    fmean(wlddev$iso3c, TRA = "-+") |> add_stub("center_")
-@
-%
-For (higher-dimensional) centering, \pkg{collapse} also has specialized function(s) \fct{f[hd]within} with additional options, and \fct{fscale} supports various scaling and centering operations.
-
-Exempting \fct{collap}, these examples may seem bulky for quick analysis, but a robust low-level API is very useful for package development, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
-I also wrote a blog post on \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{aggregating survey data using \pkg{collapse}}, which showcases more aspects of the \fct{collap} function using real census data.
-Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices.
-For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} I aggregate global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}.\footnote{A recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case) - see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
-%On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another recent example involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}.
-Thus each iteration of the numerical routine raises the vector \textbf{x} to a different power ($a$), sums it in 1 million groups ($j$) to generate \textbf{y}, and computes the Euclidean distance to \textbf{z} (using \code{collapse::fdist}).
-Without grouping objects and vectorization, this would have been difficult to handle within reasonable computing times (of a few seconds on the M1).}
+It also avoids the need to convert atomic objects to data frames for grouped aggregations or transformations.
+The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://sebkrantz.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
+<<BY>>=
+BY(num_vars(iris), g = iris$Species, FUN = mad) |> head(1)
+@
+<<fscale>>=
+fscale(iris$Sepal.Length, g = iris$Species) |> fsd(iris$Species)
+@
 %
@@
 %
-\pkg{collapse} also provides a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{fast data manipulation functions} akin to base \proglang{R} and \pkg{tidyverse} functions, including \fct{fselect}, \fct{fsubset}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{ftransform}, \fct{fmutate}, \fct{across}, \fct{frename}, \fct{fcount}, etc.
-These are integrated with the \emph{Fast Statistical Functions} to enable vectorized statistical operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow.
-For example, the following code aggregates the \code{wlddev} data by income group for years post 2015 (to smooth volatility), with population weights.
+\pkg{collapse} also provides a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc.
+as optimized analogues to base \proglang{R}/\pkg{tidyverse} functions.
+These are integrated with the \emph{Fast Statistical and Transformation Functions} to enable vectorized operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow.
+I illustrate this using the included World Development Dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.
+<<wlddev_distinct>>=
+fndistinct(wlddev)
+@
+Below, I track changes in Life Expectancy at the country level, computing the range, average change per year, and correlation with GDP per Capita in log-differences\footnote{\fct{Dlog} is a shorthand for \code{fdiff(log(x))} in \pkg{collapse}, \fct{pwcor} wraps \code{cor(..., use = "pairwise.complete.obs")}.} while also saving the latest measurements.\footnote{Note that \fct{fsummarise} never re-groups data, making a call to \fct{fungroup} redundant.} The \fct{collap} function called at the end to aggregate the country-level statistics across income groups is a convenience function for mixed-type aggregation.\footnote{See also \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{this blog post on aggregating survey data using \pkg{collapse}} which showcases more aspects of the \fct{collap} function using real census data.} By default, it uses \fct{fmean} for numeric data and \fct{fmode} for categorical data, here also applying population weights such that the most populous country in each group is kept.
 %
 <<fsummarise>>=
-wlddev |> fsubset(year >= 2015) |> fgroup_by(income) |>
-  fsummarise(country = fmode(country, POP),
-             across(c(PCGDP, LIFEEX, GINI), fmean, POP))
-@
-%
-This code is very fast because data does not need to be split by groups.
-Under the hood it is principally a syntax translation to the low-level API introduced above.\footnote{\fct{fgroup\_by} creates a \class{GRP} object from the \code{income} column, attaches it as an attribute, and \fct{fsummarise}/\fct{across} fetches it and passes it to the \code{g} arguments of the \emph{Fast Statistical Functions} set as a keyword argument (and sets \code{use.g.names = FALSE}).
-Thus, \code{w} becomes the second positional argument.
-Since \fct{fmean} is S3 generic, \fct{across} directly invokes \fct{fmean.data.frame} on the subset of columns.
-\vspace{-10mm}} \emph{Fast Statistical Functions} also have a method for grouped data, so \fct{fsummarise} is not always needed.
-%The following example calculates weighted group means.
-By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::group\_by}.
-The latter requires an additional \proglang{C} routine to convert the \pkg{dplyr} grouping object to a \class{GRP} object, and is thus less efficient.}
-%
+wlddev |> fgroup_by(country, income) |>
+  fsummarise(n = fnobs(LIFEEX),
+             diff = fmax(LIFEEX) %-=% fmin(LIFEEX),
+             mean_diff = fmean(fdiff(LIFEEX)),
+             cor_PCGDP = pwcor(Dlog(LIFEEX), Dlog(PCGDP)),
+             across(c(LIFEEX, PCGDP, POP), flast)) |>
+  collap( ~ income, w = ~ POP, keep.w = FALSE) |> print(digits = 2)
+@
+With the exception of \code{pwcor(Dlog(LIFEEX), Dlog(PCGDP))}, which is evaluated for every country, all other expressions in \fct{fsummarise} and also the final \fct{collap} call are fully vectorized, i.e., evaluated only once with grouping information passed directly to the \code{g} arguments of \emph{Fast Statistical and Transformation Functions}.
+Using subtraction by reference in the range is thus memory efficient as the executed expression is equivalent to \code{fmax(LIFFEX, country) \%-=\% fmin(LIFFEX, country)}.
+\fct{across} also directly invokes \fct{flast.data.frame} on the subset of columns.
+%
+\emph{Fast Statistical Functions} also have a method for grouped data, hence \fct{fsummarise} is not always needed.
+Below, I average the data with population weights.\footnote{Functions like \fct{fmean}, when called on a grouped data frame with weights, by default (\code{keep.w = TRUE}) sum the weights column and retain it after the grouping columns to permit further weighted operations.
+\vspace{-10mm}}
 <<fsummarise_2>>=
-wlddev |> fsubset(year >= 2015, income, PCGDP:GINI, POP) |>
-  fgroup_by(income) |> fmean(POP, keep.w = FALSE)
-@
+wlddev |> fsubset(year >= 2015, income, PCGDP:POP) |>
+  fgroup_by(income) |> fmean(POP)
+@
+Using the \emph{Fast Statistical Functions} directly often results in more efficient code.
+For example, computing population shares for each year as \code{fmutate(wlddev, POP\_share = fsum(POP, year, TRA = "/"))} is considerably more efficient than \code{wlddev |> fgroup_by(year) |> fmutate(POP\_share = fsum(POP, TRA = "/")) |> fungroup()}, which in turn is more efficient than \code{POP / fsum(POP)} (extra allocation) or \code{proportions(POP)} (split-apply-combine).
+\newline
+
+\pkg{collapse} thus offers several ways to reach the same outcome.
+This may be confusing, but all options can be distinguished in efficiency terms.
+It is thus useful to be mindful of computer resources and attempt to craft minimalistic solutions.
+For example, I recently combined multiple spatial datasets on points of interest (POIs) and needed to deduplicate them.
+I decided to keep the richest source for each location and POI type.
+After creating POI confidence, location, and type indicators comparable across sources (datasets), my deduplication expression ended up being a single line of the form \code{fsubset(data, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (i.e., richest) source by location and type.
+This is very efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}.
 %
@@
 %
-\fct{fsummarise} and \fct{fmutate} can also evaluate arbitrary functions in the classical (split-apply-combine) way and handle more complex expressions involving multiple columns and/or functions.
-However, using any \emph{Fast Statistical Function} causes the whole expression to be vectorized, i.e., evaluated only once and not for every group.
-This eager vectorization approach enables efficient grouped calculation of more complex statistics.
-The example below forecasts the exports for each country-sector via linear regression (\code{v ~ y}) in a vectorized way.
+\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, implying that using any \emph{Fast Statistical Function} in an expression causes the whole expression to be vectorized.
+This only applies to visible expressions, \pkg{collapse} cannot read the contents of custom functions, thus such functions are always evaluated split-apply-combine.
+The eager vectorization approach enables efficient grouped calculation of more complex statistics.
+Below, I forecast the population of each region via linear regression (\code{POP ~ year}) in a fully vectorized way.
 %
 <<glinreg>>=
-exports |> fgroup_by(c, s) |> fmutate(dmy = fwithin(y)) |>
-  fsummarise(v_10 = flast(v), beta = fsum(v, dmy) %/=% fsum(dmy, dmy)) |>
-  fmutate(v_11 = v_10 + beta, v_12 = v_11 + beta, beta = NULL) |> head(4)
-@
-%
-The expression \code{fsum(v, dmy) \%/=\% fsum(dmy, dmy)} amounts to \code{cov(v, y)/var(y)}, but is both vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation for the final result.
-I do not recommend forecasting trade in this way, rather, this example is inspired by a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post} where I forecasted high-resolution (1$km^2$) population estimates for South Africa.
-The data, taken from WorldPop, was available for the years 2014-2020, and I needed estimates for 2021-2022.
-Linear regression was sensible, and using the above expression I was able to run 1.6 million regressions and obtain 2 forecasts in 0.26 seconds on a laptop.
-Another neat example from the community, shared by Andrew Ghazi in a \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post}, vectorizes an expression to compute the $p$~value, \code{2 * pt(abs(fmean(x) * sqrt(6) / fsd(x)), 5, lower.tail = FALSE)}, across 300k groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}.
-The eager vectorization approach of \pkg{collapse} here replaces \code{fmean(x)} and \code{fsd(x)} by their grouped versions and evaluates the entire expression once rather than 300k times as in \pkg{dplyr}.
+wlddev |> collap(POP ~ region + year, FUN = fsum) |>
+  fmutate(POP = POP / 1e6) |> fgroup_by(region) |>
+  fmutate(dmy = fmean(year, TRA = "-")) |>
+  fsummarise(beta = fsum(POP, dmy) %/=% fsum(dmy, dmy),
+             POP20 = flast(POP)) |>
+  fmutate(POP21 = POP20 + beta, POP22 = POP21 + beta,
+          POP23 = POP22 + beta, POP24 = POP23 + beta)
+@
+%
+When \fct{fsummarise} evaluates an expression involving \emph{Fast Statistical Functions}, it sets their \code{g} argument with a grouping (\class{GRP}) object that is directly handed to C/C++, and also sets \code{use.g.names = FALSE}.
+Hence, weights (\code{w}) becomes the second positional argument.
+Similarly, \fct{fmutate} sets \code{g} and \code{TRA = "fill"}, which can be overwritten by the user (here with \code{TRA = "-"}).
+The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)}, amounting to \code{cov(x, y)/var(y)}, is thus vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation.
+In a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post}, I forecasted high-resolution population estimates for South Africa like this.
+Using 1 $km^2$ \href{https://www.worldpop.org/}{WorldPop} data available for years 2014-2020, running 1.6 million cell-regressions and obtaining 2 forecasts for 2021 and 2022 took only 0.3 seconds on my M1 Mac.
+Another neat example from the community, shared by Andrew Ghazi in a \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post}, vectorizes an expression to compute the $p$~value, \code{2 * pt(abs(fmean(x) * sqrt(6) / fsd(x)), 5, lower.tail = FALSE)}, across 300,000 groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}.
+\newline
+% The eager vectorization approach of \pkg{collapse} here replaces \code{fmean(x)} and \code{fsd(x)} by their grouped versions and evaluates the entire expression once rather than 300k times as in \pkg{dplyr}.
 \newline
@@
 \pkg{collapse} also vectorizes advanced statistics.
-The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} calculates weighted quantiles in a theoretically consistent way by applying the probability measure to the sum of weights to create a target sum and cumulatively summing weights to find order statistics which are then combined following \citet{hyndman1996sample}.
-See \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for details.}
+The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} computes weighted quantiles in a theoretically consistent way, see \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for details.}
 %
 <<wstats>>=
-wlddev |> fsubset(is.finite(POP)) |> fgroup_by(income, OECD) |>
+wlddev |> fsubset(is.finite(POP)) |> fgroup_by(region) |>
   fmutate(o = radixorder(GRPid(), LIFEEX)) |>
@@
              mean = fmean(LIFEEX, POP),
-             median = fmedian(LIFEEX, POP, o = o, ties = "q8"),
+             median = fmedian(LIFEEX, POP, o = o),
              Q3 = fnth(LIFEEX, 0.75, POP, o = o, ties = "q8"),
@@
 Weighted quantiles have a sub-column parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \fct{fnth}/\fct{fmedian}/\fct{fmode} (default 1).} and, as shown above, can also harness an (optional) optimization utilizing an overall ordering vector (combining groups and the data column) to avoid repeated partial sorting of the same elements within groups.
-%
-\section{Time series and panel data} \label{sec:ts_ps}
-%
-\pkg{collapse} also provides a flexible high-performance architecture to perform (time aware) computations on time series and panel series.
-In particular, the user can either apply time series and panel data transformations without any classes by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion, or use \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation for worry-free application in many contexts.
+% \newline
+%
+% This code is very fast because data does not need to be split by groups.
+Under the hood it is principally a syntax translation to the low-level API introduced above.\footnote{\fct{fgroup\_by} creates a \class{GRP} object from the \code{income} column, attaches it as an attribute, and \fct{fsummarise}/\fct{across} fetches it and passes it to the \code{g} arguments of the \emph{Fast Statistical Functions} set as a keyword argument (and sets \code{use.g.names = FALSE}).
+%
+% Thus, \code{w} becomes the second positional argument.
+Since \fct{fmean} is S3 generic, \fct{across} directly invokes \fct{fmean.data.frame} on the subset of columns.
+\vspace{-10mm}}  %The following example calculates weighted group means.
+By default (\code{keep.w = TRUE}) \code{fmean.grouped\_df} also sums the weights in each group.\footnote{\class{grouped\_df} methods in \pkg{collapse} support grouped data created with either \fct{fgroup\_by} or \fct{dplyr::group\_by}.
+The latter requires an additional \proglang{C} routine to convert the \pkg{dplyr} grouping object to a \class{GRP} object, and is thus less efficient.}
+%
+\subsection{Grouping objects and lower-level API} \label{ssec:gopt}
+%
+Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, for repeated operations the cost of grouping can be minimized by using factors (see \code{?qF} for efficient factor generation) or \class{GRP} objects as inputs.
+The latter contain all information \pkg{collapse}'s statistical functions may require to operate across groups and are thus passed to internal C/C++ code without checks.
+They can be created with \code{GRP()}.
+Its basic syntax is:
+\begin{Code}
+GRP(X, by = NULL, sort = TRUE, return.groups = TRUE, method = "auto", ...)
+\end{Code}
+ Below, I create a \class{GRP} object from two columns in the World Development Dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}).
+The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.
+%
+<<GRP>>=
+str(g <- GRP(wlddev, ~ income + OECD))
+@
+%
+\class{GRP} objects make grouped statistical computations in \pkg{collapse} fully programmable.
+I can employ the object with the \emph{Fast Statistical Functions} and some utilities\footnote{\fct{add\_vars} is a fast \fct{cbind.data.frame} which also has an assignment method, and \fct{get\_vars} enables fast and secure extraction of data frame columns.} to efficiently aggregate GDP per capita, life expectancy, and country name, with population weights.
+%
+<<gcomp>>=
+add_vars(g$groups,
+ get_vars(wlddev, "country") |> fmode(g, wlddev$POP, use = FALSE),
+ get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = F),
+ get_vars(wlddev, "POP") |> fsum(g, use = FALSE))
+@
+The above is equivalent to \code{collap(wlddev, country + PCGDP + LIFEEX ~ income + OECD, w = ~ POP)}, which internally toggles many of the same function calls. \newline
+
+% For advanced data aggregation, \pkg{collapse} also provides a convenience function, \fct{collap}, which, by default, uses \fct{fmean} for numeric, \fct{fmode} for categorical, and \fct{fsum} for weight columns, and preserves their order.
+The equivalent expression using this function would be
+%
+% <<collap>>=
+% ollap(wlddev, country + PCGDP + LIFEEX ~ income + OECD, w = ~ POP)
+% @
+%
+Similarly, data can be transformed, here using \fct{fwithin} to level average differences in economic status, adding back the overall mean after subtracting out group means:\footnote{\fct{add\_stub} adds a prefix (or suffix if \code{pre = FALSE}) to columns $\to$ \code{center\_PCGDP} and \code{center\_LIFEEX}.}
+%
+<<gtrans>>=
+add_vars(wlddev) <- get_vars(wlddev, c("PCGDP", "LIFEEX")) |>
+    fwithin(g, mean = "overall.mean") |> add_stub("center_")
+@
+%
+% For (higher-dimensional) centering, \pkg{collapse} also has specialized function(s) \fct{f[hd]within} with additional options, and \fct{fscale} supports various scaling and centering operations.\newline
+
+The low-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
+Users should note that \pkg{collapse} does not provide metaprogramming capabilities in its non-standard evaluation functions---such as \href{https://rlang.r-lib.org/reference/topic-quosure.html}{quosures} or \href{https://dplyr.tidyverse.org/articles/programming.html}{indirection} familiar to \pkg{tidyverse} users.
+Instead, there exist standard-evaluation equivalents to some of these functions which typically end with a \code{v} (for 'variables'), such as \fct{collapv}, \fct{fslicev}, \fct{ftransformv}, etc.
+In other cases, such as those typically handled by \fct{fsummarise} or \fct{fmutate}, users may need to use the low-level API for programming, or resort to \fct{substitute} and friends.
+The main reason for not providing higher-level metaprogramming capabilities is to keep all functions as simple as possible.
+It also compels users to think deeper about their programs and devise more efficient solutions.
+\newline
+
+These metaprogramming restrictions are also vindicated by internally complex data manipulation functions producing greater overheads that can noticeably add up in larger scripts:\footnote{\fct{bmark} is a slim wrapper around \fct{bench::mark}.
+See the Computational details section.}
+<<fsumm_bench>>=
+ bmark(collapse = collapse::fsummarise(mtcars, sum = sum(mpg)),
+       dplyr = dplyr::summarise(mtcars, sum = sum(mpg)))
+@
+Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices.
+In the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case) - see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
+%On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another recent example involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}.
+Thus each iteration of the numerical routine raises the vector \textbf{x} to a different power ($a$), sums it in 1 million groups ($j$) to generate \textbf{y}, and computes the Euclidean distance to \textbf{z} (using \code{collapse::fdist}).
+Without grouping objects and vectorization, this would have been difficult to handle within reasonable computing times (of a few seconds on the M1).}
+%
+%
+\section{Time series and panel series} \label{sec:ts_ps}
+%
+\pkg{collapse} also provides a flexible high-performance architecture to perform time aware computations on time series and panel series.
+In particular, users can either apply time series and panel data transformations by passing individual and/or time identifiers to the respective functions in an ad-hoc fashion, or by using \class{indexed\_frame} and \class{indexes\_series} classes, which implement full and deep indexation for worry-free application in many contexts.
 Table~\ref{tab:TSfun} compactly summarizes \pkg{collapse}'s architecture for time series and panel data.
@@
 %
-Time series functions such as \fct{fgrowth} (to compute growth rates) are S3 generic and can be applied to most time series classes.
+Time series functions such as \fct{fgrowth} to compute growth rates are S3 generic and can be applied to most time series classes.
 In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation.
-If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a numeric variable denoting calendar years.
+If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a plain numeric vector representing calendar years.
 Users need to manually call \code{timeid()} on plain numeric vectors with decimals to yield an appropriate integer representation.
@@
 @
-The following extracts one sector from the \code{exports} dataset generated above, creating an irregular time series missing the 3rd and 6th period.\footnote{\code{\%=\%} is an infix operator for the \fct{massign} function in \pkg{collapse} which is a multivariate version of \fct{assign}.} Indexation using the \code{t} argument allows for correct (time-aware) computations on this context without 'expanding' the data.
+To illustrate the full capabilities of these time series functions, I generate a sector-level trade dataset of export values (v) by country (c), sector (s), and year (y).
+Like many detailed trade datasets, it is unbalanced---not all sectors/products are exported by country c in all years.
+<<exports>>=
+set.seed(101)
+exports <- expand.grid(y = 2001:2010, c = paste0("c", 1:10),
+  s = paste0("s", 1:10)) |> fmutate(v = abs(rnorm(1e3))) |>
+  colorder(c, s) |> fsubset(-sample.int(1e3, 500))
+@
+The following extracts one country-sector series from the \code{exports} dataset.
+It is irregular, missing years 2003 and 2006.\footnote{\code{\%=\%} is an infix operator for the \fct{massign} function in \pkg{collapse} which is a multivariate version of \fct{assign}.} Indexation using the \code{t} argument then allows for correct (time-aware) computations in this context without the need to 'expand' the data/fill gaps.
 %
@@
 %
-Functions \code{flag()/fdiff()/fgrowth()} also have shorthands \code{L()/D()/G()} which both facilitate their use inside formulas and provide an enhanced data frame interface for convenient ad-hoc computations.
-With panel data, \code{t} can be omitted, but this requires sorted data with consecutive groups.\footnote{This is because a group-lag is computed in a single pass, requiring all group elements to be consecutive.} Below, I demonstrate two ways to compute a sequence of lagged growth rates using both the \code{G()} operator and the \code{tfm()} function - a shorthand for \code{ftransform()}.\footnote{See also Footnote~\ref{fn:tfm}.
-A number of key functions in \pkg{collapse} have syntactic shorthands.
-The \code{list(v = v)} is needed here to prevent \fct{fgrowth} from creating a matrix with the growth rates---the \class{list} method applies.}
+Functions \code{flag()/fdiff()/fgrowth()} also have associated 'operators' \code{L()/D()/G()} to facilitate their use inside formulas and provide an enhanced data frame interface for convenient ad-hoc computations.
+With panel data, \code{t} can be omitted, but this requires sorted data with consecutive groups.
+%\footnote{This is because a group-lag is computed in a single pass, requiring all group elements to be consecutive.}
+Below, I demonstrate two ways to compute a sequence of lagged growth rates using either \code{G()} or \fct{fgrowth} and \fct{tfm}---a shorthand for \fct{ftransform}.\footnote{Several key functions in \pkg{collapse} have syntactic shorthands.
+The \code{list(v = v)} is needed here to prevent \fct{fgrowth} from creating a matrix with the growth rates, ensuring that the \class{list} method applies.}
 %
@@
 %
-These functions and operators are also integrated with \fct{fgroup\_by} and \fct{fmutate} for vectorized computations. However, using ad-hoc grouping is always more efficient.
+These functions and operators are also integrated with \fct{fgroup\_by} and \fct{fmutate} for vectorized computations. But, as mentioned earlier, ad-hoc grouping is always more efficient.
 %
@@
 to the creation of new classes \class{indexes\_series} and \class{indexed\_frame} which inherit from the former.
-Any data frame-like object can become an \class{indexed\_frame} and function as usual for other operations.
+Any data frame-like object can become an \class{indexed\_frame} and behave as usual for other operations.
 The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}.
@@
 Unlike \class{pdata.frame}, an \\ \class{indexed\_frame} is a deeply indexed structure---every series inside the frame is already an \class{indexes\_series}.
-A comprehensive set of \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()}, ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.).
+A comprehensive set of \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()} ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.).
 % A basic demonstration with World Bank panel data showcases the flexibility of these classes.
@@
 %
-An example using the \code{exports} data follows:
+It is worth highlighting that the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions.\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.} Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures or native handling of irregularity in basic operations.
+\class{indexed\_series} and \class{indexed\_frame}, on the other hand, work 'anywhere', and can be superimposed on any suitable object, as long as \pkg{collapse}'s functions (\code{flag()/L()}, etc.) are used to perform the time-based computations.
+\newline
+
+An example using the \code{exports} data follows. Note that data need not be sorted for indexing.
 <<indexing>>=
 exportsi <- exports |> findex_by(c, s, y)
-exportsi |> G() |> print(max = 15)
+exportsi |> G() |> head(5)
 exportsi |> findex() |> print(2)
@@
 @
-\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work in arbitrary data masking environments.
+\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work in arbitrary data masking environments:
 <<indexing_3>>=
@@
 @
-It is worth highlighting that the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions.\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.} Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures or native handling of irregularity in basic operations.
-\class{indexed\_series} and \class{indexed\_frame}, on the other hand, work 'anywhere', and can be superimposed on any suitable object, as long as \pkg{collapse}'s functions (\code{flag()/L()} etc.) are used to perform time-based computations.
-\newline
-
 Indexed series/frames also support transformation such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}.
@@
 %
-Among all \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and 'recast' pivots.
+Among all \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots.
 Both implementations provide outstanding computational performance, syntax, and memory efficiency.
@@
 %
-Compared to commercial software such as \proglang{STATA}, the implementation of joins in most open-source software, including \proglang{R}, provides no information on how many records were joined from both tables.
+Compared to commercial software such as \proglang{STATA} \citep{STATA}, the implementation of joins in most open-source software provides no information on how many records were joined from both tables.
 This often provokes manual efforts to validate the join operation.
-\code{collapse::join} provides a rich set of options to understand table join operations.
+\fct{collapse::join} provides many options to understand table join operations.
 Its syntax is:
@@
 \end{Code}
-It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}), i.e., it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases.
+It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}), thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases.
 By default (\code{verbose = 1}), it prints information about the join operation and number of records joined.
+\newline
+
 To demonstrate \fct{join}, I generate a small database for a bachelor in economics curriculum.
@@
 %
-Users can request the generation of a \code{.join} column (\code{column = "name"/TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, to indicate the origin of records in the joined table---useful on a full join.
+Users can request the generation of a \code{.join} column (\code{column = "name"/TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, to indicate the origin of records in the joined table---useful with a full join:
 %
@@
 %
-An alternative is to request an attribute (\code{attr = "name"/TRUE}) that also summarizes the join operation, including the output of \fct{fmatch} (the workhorse of \fct{join} if \code{sort = FALSE}).
+An alternative is to request an attribute (\code{attr = "name"/TRUE}) that also summarizes the join operation, including the output of \fct{fmatch}---the workhorse of \fct{join} if \code{sort = FALSE}.
 \begin{Code}
@@
 @
-Users can also invoke the \code{validate} argument to examine the uniqueness of the join keys in either table: passing a '1' for a non-unique key produces an error.
+Users can also invoke the \code{validate} argument to examine the uniqueness of the join keys in either table: passing a '1' produces an error if the key is not unique.
 %
@@
 %
+Similarly, the \code{require} argument allows users to demand a minimum matching success rate.
+%
+<<joins_4.5>>=
+join(teacher, course, on = "id", require = list(x = 0.8)) |>
+  tryCatch(error = function(e) substr(e, 102, 200) |> cat())
+@
+%
 A few further particularities are worth highlighting.
@@
 It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"}, or \code{"anti"} join).
-This demonstrates the latter two:
+This demonstrates the latter:
 <<joins_5>>=
@@
 This is congruent to the principle of adding columns to \code{x} and altering this table as little as possible.
-Alternatively, option \code{drop.dup.cols = "x"/"y"} can be used to simply remove duplicate columns from \code{x} or \code{y} before the join operation.
+Alternatively, option \code{drop.dup.cols = "x"/"y"} can be used to remove duplicate columns from either \code{x} or \code{y} before the join operation.
 %
@@
 A final noteworthy feature is that \fct{fmatch} has a built-in overidentification check, which warns if more key columns than necessary to identify the records are provided.
-This check only triggers with 3+ id columns as for efficiency reasons the first two ids are jointly hashed.
+This check only triggers with 3+ id columns as for efficiency reasons the first two ids are hashed jointly.
 
@@
 Modern \proglang{R} packages also offer different reshaping functions, such as \fct{data.table::melt}/\fct{tidyr::pivot\_longer} to combine columns and \fct{data.table::dcast}/\fct{tidyr::pivot\_wider} to expand them, requiring users to learn both.
-Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two reshaping functions, incurring a high cost in terms of both syntax and memory efficiency.
-\newline
-
-\code{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings: it has a single intuitive syntax to perform 'longer', 'wider', and 'recast'\newpage
-
-pivots, and supports complex labelled data without loss of information. Its basic syntax is:
+Since the depreciation of \pkg{reshape(2)} \citep{rreshape2}, there is also no modern replacement for \code{reshape2::recast()}, requiring \proglang{R} users to consecutively call two functions.
+\newline
+
+\code{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings.
+It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information.
+Its basic syntax is:
 %
@@
 %
-The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Databas (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}), providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America.
+The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America.
 While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes.
@@
 To reshape this dataset into a longer format, it suffices to call \code{pivot(data, ids = 1:4)}.
-If \code{labels = "name"} is specified, variable labels (stored in \code{attr(column, "label")}) are saved in an additional column.
+If \code{labels = "name"} is specified, variable labels stored in \code{attr(column, "label")} are saved to an additional column.
 In addition, \code{names = list(variable = "var_name", value = "val_name")} can be passed to set alternative names for the \code{variable} and \code{value} columns.
@@
 It is, however, a destructive operation---with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it.
-Further arguments like \code{na.rm}, \code{fill}, \code{sort}, and \code{transpose} can be used to control the casting process.
+Further arguments like \code{na.rm}, \code{fill}, \code{drop}, \code{sort}, and \code{transpose} can be used to control the casting process/output.
 %
@@
 %
-For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names need to be specified using a list passed to \code{names}, and similarly for \code{labels}.
+For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names needs to be specified using a list passed to \code{names}, and similarly for \code{labels}.
 Again, taking along labels is entirely optional---omitting either the labels-list's \code{from} or \code{to} element will omit the respective operation.
@@
 This (\code{dr}) is the tidy format \citep{rtidydata} where each variable is a separate column.
-It is analytically more useful, e.g., to compute labor productivity as \code{settransform(dr, LP = VA / EMP)}, or to estimate a panel-regression with sector fixed-effects.
-The recast pivot is thus a natural operation to change data representations.
+It is analytically more useful, e.g., to compute labor productivity as \code{settransform(dr, LP = VA / EMP)} or to estimate a panel-regression with sector fixed-effects.
+The recast pivot is thus a natural operation to change data representation.
 As with the other pivots, it preserves all information and can be reversed by simply swapping the contents of the \code{from} and \code{to} keywords.
@@
 @
-More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}.
+The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
 Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.
@@
 To create nested data, \fct{rsplit} generalizes \fct{split} and (recursively) splits up data frame-like objects into (nested) lists.
-For example, we can split the \code{GGDC10S} data by country and\newpage variable, such that, e.g., agricultural employment in Argentina can be accessed as:\footnote{If a nested structure is not needed, \code{flatten = TRUE} lets \fct{rsplit} operate like a faster version of \fct{split}.}
+For example, we can split the \code{GGDC10S} data by country and variable, such that, for example, agricultural employment in Argentina can be accessed as:\footnote{If a nested structure is not needed, \code{flatten = TRUE} lets \fct{rsplit} operate like a faster version of \fct{split}.}
 %
@@
 As mentioned, such data representation can also be the result of an API call parsing JSON or a nested loop or \fct{lapply} call.
-Below, I write a nested loop running a regression of agriculture on mining and manuacturing output/employment:
+Below, I write a nested loop running a regression of agriculture on mining and manuacturing output and employment.
 <<list_proc_2>>=
@@
 @
-This programming may not be ideal for this particular use case as I could have used data.frame-based tools and saved the result in a column.\footnote{E.g., \code{GGDC10S |> fgroup\_by(Country, Variable) |> fsummarise(result = my\_fun(lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year)))} with \code{my\_fun <- function(m) list(list(m, BIC(m), summary(m)))}.}
-However, there are limits to data.frame-based workflows.
+This programming may not be ideal for this particular use case as I could have used data frame-based tools and saved the result in a column.\footnote{E.g., \code{GGDC10S |> fgroup\_by(Country, Variable) |> fsummarise(result = my\_fun(lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year)))} with \code{my\_fun <- function(m) list(list(m, BIC(m), summary(m)))}.}
+However, there are limits to such workflows.
 For example, I recently trained a complex ML model for different variables and parameters, which involved loading a different dataset at each iteration.
@@
 % Thus \code{get\_elem("R2")} returns the same nested list with the $R^2$ in all final nodes. Then \code{rowbind()} could be used to create a data frame:
-Note how the \code{"summary"} branch was eliminated since it is common to all final nodes; \code{result |> get\_elem("r.squared", keep.tree = TRUE)} could have been used to retain it.
+Note how the \code{"summary"} branch was eliminated since it is common to all final nodes; \code{result |> get\_elem("r.squared", keep.tree = TRUE)} could have been used to keep it.
 \fct{rowbind} then efficiently combines lists of lists.
-We can also apply \fct{t\_list} to turn the list inside-out.
+We can also apply \fct{t\_list} to turn the list inside-out:
 <<list_proc_3.5>>=
@@
 @
-For such cases, I created \fct{unlist2d} as a complete recursive generalization of \fct{unlist}.
-It creates a \class{data.frame} (or \class{data.table}) representation of any nested list using recursive row-binding and coercion operations while generating (optional) id variables representing the list tree and (optionally) saving row names of matrices or data frames.
+For such cases, \fct{unlist2d} provides a complete recursive generalization of \fct{unlist}.
+It creates a \class{data.frame} or \class{data.table} representation of any nested list using recursive row-binding and coercion operations, while generating (optional) id variables representing the list tree and (optionally) saving row names of matrices or data frames.
 In the present example
@@
 
-This exemplifies the power of these tools to create, query, and combine nested data in very general ways, and with many applications.
+This exemplifies the power of these tools to create, query, and combine nested data in very general ways.
   % I use it extensively to fetch and compare different statistics from (sometimes large) nested lists with different statistical model outputs.
-Further useful functions include \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, and \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes.
+Further useful functions include \fct{has\_elem} to check for the existence of elements, \fct{ldepth} to return the maximum level of recursion, \fct{is\_unlistable} to check whether a list has atomic elements in all final nodes, \fct{[ir]reg\_elem} to recursively extract the (non-)atomic elements, and \fct{rapply2d} to apply functions to nested lists of data objects.
 %
@@
 %
-\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious and powerful toolkit to examine complex datasets.
+\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets.
 A particular focus has been on providing tools for examining longitudinal (panel) data.
-Recall the indexed World Development Panel (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{ssec:gopt}.
+Recall the indexed World Development panel dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}).
 The function \fct{varying} can be used to examine which of these variables are time-varying.
@@
 A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time.
-Using the (de-)meaning functions supporting \class{indexed\_series} from Table~\ref{tab:TSfun}, this is easily demonstrated.
+Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (see Table~\ref{tab:TSfun}), this is easily demonstrated.
 %
@@
 %
-The \fct{qsu} (quick-summary) function provides an efficient method to (approximately) compute this decomposition, considering the group-means instead of the between transformation\footnote{This is more efficient and equal to using the between transformation if the panel is balanced.} and adding the mean back to the within transformation to preserve the scale of the data.
+The \fct{qsu} (quick-summary) function provides an efficient method to compute this decomposition, considering the group-means instead of the between transformation\footnote{This is more efficient and equal to using the between transformation if the panel is balanced.} and adding the overall mean back to the within transformation to preserve the scale of the data.
 %
@@
 The decomposition above implies more variation in life expectancy between countries than within countries over time.
-It can also be computed for different subgroups, such as OECD members and non-members, and with sampling weights, such as population.
+It can also be computed for different subgroups and with sampling weights.
+\fct{qsu} also has a data frame method, and by default computes simple statistics:\footnote{The \code{pid} argument to \fct{qsu} can also be used to pass identifiers for panel-decomposition, e.g., \code{pid = ~ iso3c}.
+With indexed data, it is automatically set to the first column in the index (\code{effect = 1}).}
 %
 <<qsu_2>>=
-qsu(LIFEEXi, g = wlddev$OECD, w = wlddev$POP) |> aperm()
-@
-%
-The output shows that the variation in life expectancy is significantly larger for non-OECD countries.
-For the latter the between- and within-country variation is approximately equal.\footnote{\fct{qsu} also has a convenient formula interface to perform these transformations in an ad-hoc fashion, e.g., the above can be obtained using \code{qsu(wlddev, LIFEEX $\sim$ OECD, $\sim$ iso3c, $\sim$ POP)}, without prior indexation.} For greater detail, \fct{descr} provides a rich (grouped, weighted) statistical description.
+qsu(wlddev, by = LIFEEX ~ income, w = ~ POP)
+@
+%
+% The output shows that the variation in life expectancy is significantly larger for non-OECD countries.
+For the latter the between- and within-country variation is approximately equal.\footnote{\fct{qsu} also has a convenient formula interface to perform these transformations in an ad-hoc fashion, e.g., the above can be obtained using \code{qsu(wlddev, LIFEEX $\sim$ OECD, $\sim$ iso3c, $\sim$ POP)}, without prior indexation.}
+For greater detail, \fct{descr} provides a rich (grouped, weighted) statistical description.
+It does not support panel-variance decompositions like \fct{qsu}, but also computes detailed (grouped, weighted) frequency tables for categorical data and is thus very useful with mixed-type data.
 %
@@
 wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
-wlda15 |> descr(income + LIFEEX ~ OECD)
-@
-
-%
-While \fct{descr} does not support panel-variance decompositions like \fct{qsu}, it also computes detailed (grouped, weighted) frequency tables for categorical data and is thus very utile with mixed-type data.
-A \code{stepwise} argument toggles describing one variable at a time, allowing users to naturally 'click-through' a large dataset rather than printing a massive output to the console.
+wlda15 |> descr(by = income + LIFEEX ~ OECD)
+@
+
+%
+ It also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'click-through' the variables in a large dataset while spreading the computational burden.
 The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples.
-Both \fct{qsu} and \fct{descr} provide an \code{as.data.frame()} method for efficient tidying and further analysis.
+Both \fct{qsu} and \fct{descr} come with an \code{as.data.frame()} method for efficient tidying and easy further analysis.
 \newline
@@
 A final noteworthy function from \pkg{collapse}'s descriptive statistics toolkit is \fct{qtab}, an enhanced drop-in replacement for \fct{table}.
-It is enhanced both in a statistical and a computational sense, providing a remarkable performance boost, an option (\code{sort = FALSE}) to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights---vectorized when using \emph{Fast Statistical Functions}, as demonstrated below.
+It is enhanced both in a statistical and a computational sense, providing a remarkable performance boost, an option \code{sort = FALSE} to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights---vectorized when using \emph{Fast Statistical Functions}, as demonstrated below.
 %
@@
 %
-\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function.
+\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default values of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function.
 These options are saved in an internal environment called \code{.op}.
@@
   mutate(PCGDP_lag = lag(PCGDP),
-         PCGDP_diff = PCGDP - PCGDP_lag,
          PCGDP_growth = growth(PCGDP)) |> unindex()
@@
 Similarly, using this option, all code in this article could have been written without f-prefixes.
-Thus, \pkg{collapse} is able to offer a fast and syntactically clean experience of \proglang{R} - without the need to even restart the session.
+Thus, \pkg{collapse} is able to offer a fast and syntactically clean experience of \proglang{R}---without the need to even restart the session.
 Masking is completely and interactively reversible: calling \code{set\_collapse(mask = NULL)} instantly removes the additional functions.
@@
 Invoking either \code{mask} or \code{remove} detaches \pkg{collapse} and re-attaches it at the top of the search path, letting its namespace to take precedence over other packages.
+\newline
+
+Such global powers confer responsibilities upon package developers, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html#some-notes-on-global-options}{vignette on developing with \pkg{collapse}}.
+As a general rule, options \code{mask} and \code{remove} should be off-limits inside packages, and other options need to be reset immediately using \fct{on.exit}.
 %
@@
 This section provides several simple benchmarks to show that \pkg{collapse} provides best-in-\proglang{R} performance for statistics and data manipulation on moderately sized datasets.
-They are executed on an Apple M1 MacBook Pro with 16 GB unified memory.
+They are executed on a 2024 Apple MacBook Pro with 48GB M4 Pro chip.
 It also discusses results from \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{3rd party benchmarks} involving \pkg{collapse}.
@@
 %
-<<bench_1>>=
+<<bench_1, eval=TRUE>>=
 set.seed(101)
@@
 They are faster than base \proglang{R} even without multithreading, but using 4 threads in this case induces a sizeable difference.
-<<bench_2, cache=TRUE>>=
+<<bench_2, eval=TRUE, cache=TRUE>>=
 set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)
@@
 %
-Below I also show a grouped version summing the columns within 1000 random groups.
-%
-<<bench_2.1, cache=TRUE>>=
+Below, I also show a grouped version summing the columns within 1000 random groups.
+%
+<<bench_2.1, eval=TRUE, cache=TRUE>>=
 g <- sample.int(1e3, 1e4, TRUE)
@@
 %
-<<bench_flights_setup, cache=TRUE>>=
+<<bench_flights_setup, eval=TRUE, cache=TRUE>>=
 fastverse_extend(nycflights23, dplyr, data.table); setDTthreads(4)
@@
 %
-<<bench_flights_summarise, cache=TRUE>>=
+<<bench_flights_summarise, eval=TRUE, cache=TRUE>>=
 vars <- .c(dep_delay, arr_delay, air_time, distance, hour, minute)
@@
 %
-Below, I also benchmark the mean and median functions in the same way. It is evident that with non-primitive \proglang{R} functions the split-apply-combine logic is even more costly.
-%
-<<bench_flights_summarise_2, cache=TRUE, echo=FALSE>>=
+Below, I also benchmark the mean and median functions in the same way. It is evident that with non-primitive \proglang{R} functions the split-apply-combine logic becomes even more costly.
+%
+<<bench_flights_summarise_2, eval=TRUE, cache=TRUE, echo=FALSE>>=
 bmark(dplyr_mean = flights |> group_by(month, day, origin, dest) |>
@@
 So far, \pkg{data.table}, by virtue of it's internal vectorizations (also via dedicated grouped \proglang{C} implementations of simple functions), is competitive.\footnote{Much longer data will likely also favor \pkg{data.table} over \pkg{collapse} due to its sub-column-level parallel grouping and implementation of simple functions like \fct{sum} and \fct{mean}, see, e.g., the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}.} Below, I compute the range of one column (\code{x}) using \code{max(x) - min(x)}.
-As elucidated in Section~\ref{ssec:vfat}, this expression is also vectorized in \pkg{collapse}, where it amounts to \code{fmax(x, g) - fmin(x, g)}, but not in \pkg{data.table}.
-%
-<<bench_flights_summarise_3, cache=TRUE>>=
+As elucidated in Section~\ref{sec:integration}, this expression is also vectorized in \pkg{collapse}, where it amounts to \code{fmax(x, g) - fmin(x, g)}, but not in \pkg{data.table}.
+%
+<<bench_flights_summarise_3, eval=TRUE, cache=TRUE>>=
 bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
@@
 \end{Code}
-<<bench_flights_join_setup, cache=TRUE, echo=FALSE>>=
+<<bench_flights_join_setup, eval=TRUE, cache=TRUE, echo=FALSE>>=
 flights |> join(weather, on = c("origin", "time_hour")) |>
@@
 %
-<<bench_flights_join, cache=TRUE, echo=FALSE>>=
+<<bench_flights_join, eval=TRUE, cache=TRUE, echo=FALSE>>=
 bmark(
@@
 
-Last but not least, I benchmark pivots, starting a with long pivot that simply melts the 6 columns aggregated beforehand into one column, duplicating all other columns 6 times:
-%
-<<bench_pivot_long, cache=TRUE, max.print=50>>=
+Last but not least, I benchmark pivots, starting with a long pivot that simply melts the 6 columns aggregated beforehand into one column, duplicating all other columns 6 times:
+%
+<<bench_pivot_long, eval=TRUE, cache=TRUE, max.print=50>>=
 bmark(tidyr = tidyr::pivot_longer(flights, cols = vars),
@@
 %
-Memory-wise, \pkg{collapse} and \pkg{data.table} are equally efficient, but \pkg{collapse} is faster, presumably due to more extensive use of \code{memset()} to copy values in \proglang{C}, or smaller \proglang{R}-level overheads.
+Memory-wise, \pkg{collapse} and \pkg{data.table} are equally efficient, but \pkg{collapse} is faster, presumably due to more extensive use of \code{memset()} to copy values in \proglang{C} or smaller \proglang{R}-level overheads.
 
@@
 %
-<<bench_pivot_wide, cache=TRUE, max.print=50>>=
+<<bench_pivot_wide, eval=TRUE, cache=TRUE, max.print=50>>=
 bmark(tidyr = tidyr::pivot_wider(flights, id_cols = .c(month, day, dest),
@@
 I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.github.io/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages.
-These benchmarks align in showing that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and on windows systems.\footnote{Reasons for the particularly strong performance of \pkg{collapse} on Windows may be that it is largely written in C and has limited multithreading in favor or highly efficient serial algorithms---there appear to be persistent obstacles to efficient (low-overhead) multithreading on Windows, implying that multithreaded query engines do not develop their full thrust on medium-sized ($\leq$100 million row) datasets.}
+These benchmarks align in showing that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and on personal computers.
+% \footnote{Reasons for the particularly strong performance of \pkg{collapse} on Windows may be that it is largely written in C and has limited multithreading in favor or highly efficient serial algorithms---there appear to be persistent obstacles to efficient (low-overhead) multithreading on Windows, implying that multithreaded query engines do not develop their full thrust on medium-sized ($\leq$100 million row) datasets.}
 
@@
 At this point the API is stable---it has changed very little over the 5 years and no further changes are planned.
-Compatibility with \proglang{R} version 3.4.0 will be maintained for as long as possible.
+Compatibility with \proglang{R} version 3.5.0 will be maintained for as long as possible.
 Minor new features are currently planned.
@@
 Users can also follow \pkg{collapse} on \href{https://x.com/collapse_R}{Twitter/X} and \href{https://bsky.app/profile/rcollapse.bsky.social}{Bluesky} to be notified about major updates and participate in community discussions.
+\newline
+
+Last but not least, \pkg{collapse} users are encouraged to also familiarize themselves with the \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse}},\footnote{Website: https://fastverse.github.io/fastverse/} a suite of complementary high-performance packages for statistical computing and data manipulation in R which offer more advanced tools in several statistical computing domains.
+The \pkg{fastverse} metapackage also provides a lightweight framework to jointly load and manage these packages, as well as to build customized and fully separate package verses.
 
@@
 \section*{Computational details}
-The results in this paper were obtained using \proglang{R} \citep{R} 4.3.0 with \pkg{collapse} 2.0.19, \pkg{data.table} 1.16.4, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.0.0, \pkg{fastverse} 0.3.4, and \pkg{bench} 1.1.3 \citep{rbench}.
+The results in this paper were obtained using \proglang{R} \citep{R} 4.3.0 with \pkg{collapse} 2.1.2, \pkg{data.table} 1.16.4, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.0.0, \pkg{fastverse} 0.3.4, and \pkg{bench} 1.1.3 \citep{rbench}.
 All packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at https://CRAN.R-project.org/.
-The benchmark was run on an Apple M1 MacBook Pro (2020) with 16GB unified memory.
+The benchmark was run on an Apple MacBook Pro (2024) with 48GB M4 Pro chip.
 Packages were compiled from source using Homebrew Clang version 16.0.4 with OpenMP enabled and the -O2 flag.
+\newline
+
+The \fct{bmark} function used for benchmarking is defined as follows:
+\begin{Code}
+bmark <- function(...) {
+  bench::mark(..., min_time = 2, check = FALSE) |>
+    janitor::clean_names() |>
+    fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
+    fmutate(expression = names(expression)) |>
+    dapply(as.character) |> qDF()
+}
+\end{Code}
 

### 2025-05-05 misc/JSS article/final/article.tex
Run benchmark on M4 Mac.

--- 2025-05-05 misc/JSS article/final/article.Rnw
@@
 %% need no \usepackage{Sweave.sty}
-\SweaveOpts{engine=R, eps=FALSE, keep.source = TRUE, cache=FALSE}
+
 % \renewenvironment{knitrout}{\setlength{\topsep}{0mm}}{}
-
-<<preliminaries, echo=FALSE, message=FALSE, warning=FALSE, results=hide>>=
-# knitr::opts_chunk$set(cache = FALSE, size = "small")
-rm(list = ls()); gc()
-options(prompt = "R> ", continue = "+  ", width = 77, digits = 4, useFancyQuotes = FALSE, warn = 1)
-
-# Loading libraries and installing if unavailable
-if(!requireNamespace("fastverse", quietly = TRUE)) install.packages("fastverse")
-library(fastverse) # loads data.table, collapse, magrittr and kit (not used)
-# Package versions used in the article:
-# fastverse 0.3.4, collapse 2.1.2, data.table 1.16.4,
-# bench 1.1.3, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.0.0
-
-# Reset collapse options (if set)
-set_collapse(nthreads = 1L, remove = NULL, stable.algo = TRUE, sort = TRUE,
-             digits = 2L, stub = TRUE, verbose = 1L, mask = NULL, na.rm = TRUE)
-
-# For presentation: removing whitespace around labels
-vlabels(GGDC10S) <- trimws(vlabels(GGDC10S))
-
-# This is the benchmark function (janitor 2.2.0 is used)
-bmark <- function(...) {
-  bench::mark(..., min_time = 0.01, check = FALSE) |> janitor::clean_names() |>
-  fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
-  fmutate(expression = names(expression)) |> dapply(as.character) |> qDF()
-}
-
-@
 
@@
 %, and attributes are preserved as much as possible ()
-<<faststatfun>>=
-fmean(iris$Sepal.Length)
-fmean(num_vars(iris))
-identical(fmean(num_vars(iris)), fmean(as.matrix(num_vars(iris))))
-fmean(iris$Sepal.Length, g = iris$Species)
-fmean(num_vars(iris), g = iris$Species, nthreads = 4)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fmean(iris$Sepal.Length)
+\end{Sinput}
+\begin{Soutput}
+[1] 5.843
+\end{Soutput}
+\begin{Sinput}
+R> fmean(num_vars(iris))
+\end{Sinput}
+\begin{Soutput}
+Sepal.Length  Sepal.Width Petal.Length  Petal.Width 
+       5.843        3.057        3.758        1.199 
+\end{Soutput}
+\begin{Sinput}
+R> identical(fmean(num_vars(iris)), fmean(as.matrix(num_vars(iris))))
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\begin{Sinput}
+R> fmean(iris$Sepal.Length, g = iris$Species)
+\end{Sinput}
+\begin{Soutput}
+    setosa versicolor  virginica 
+     5.006      5.936      6.588 
+\end{Soutput}
+\begin{Sinput}
+R> fmean(num_vars(iris), g = iris$Species, nthreads = 4)
+\end{Sinput}
+\begin{Soutput}
+           Sepal.Length Sepal.Width Petal.Length Petal.Width
+setosa            5.006       3.428        1.462       0.246
+versicolor        5.936       2.770        4.260       1.326
+virginica         6.588       2.974        5.552       2.026
+\end{Soutput}
+\end{Schunk}
 %
@@
 For example, option \code{TRA = "fill"} replaces elements with the corresponding statistics:
-<<TRA>>=
-fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill")[c(1:5, 51:55)]
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill")[c(1:5, 51:55)]
+\end{Sinput}
+\begin{Soutput}
+ [1] 5.006 5.006 5.006 5.006 5.006 5.936 5.936 5.936 5.936 5.936
+\end{Soutput}
+\end{Schunk}
 %\code{TRA()} is called internally in the \emph{Fast Statistical Functions}, the \code{TRA} argument is passed to \code{FUN}.
@@
 The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://sebkrantz.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
-<<BY>>=
-BY(num_vars(iris), g = iris$Species, FUN = mad) |> head(1)
-@
-<<fscale>>=
-fscale(iris$Sepal.Length, g = iris$Species) |> fsd(iris$Species)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> BY(num_vars(iris), g = iris$Species, FUN = mad) |> head(1)
+\end{Sinput}
+\begin{Soutput}
+       Sepal.Length Sepal.Width Petal.Length Petal.Width
+setosa       0.2965      0.3706       0.1483           0
+\end{Soutput}
+\end{Schunk}
+\begin{Schunk}
+\begin{Sinput}
+R> fscale(iris$Sepal.Length, g = iris$Species) |> fsd(iris$Species)
+\end{Sinput}
+\begin{Soutput}
+    setosa versicolor  virginica 
+         1          1          1 
+\end{Soutput}
+\end{Schunk}
 %
@@
 I illustrate this using the included World Development Dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.
-<<wlddev_distinct>>=
-fndistinct(wlddev)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fndistinct(wlddev)
+\end{Sinput}
+\begin{Soutput}
+country   iso3c    date    year  decade  region  income    OECD   PCGDP 
+    216     216      61      61       7       7       4       2    9470 
+ LIFEEX    GINI     ODA     POP 
+  10548     368    7832   12877 
+\end{Soutput}
+\end{Schunk}
 Below, I track changes in Life Expectancy at the country level, computing the range, average change per year, and correlation with GDP per Capita in log-differences\footnote{\fct{Dlog} is a shorthand for \code{fdiff(log(x))} in \pkg{collapse}, \fct{pwcor} wraps \code{cor(..., use = "pairwise.complete.obs")}.} while also saving the latest measurements.\footnote{Note that \fct{fsummarise} never re-groups data, making a call to \fct{fungroup} redundant.} The \fct{collap} function called at the end to aggregate the country-level statistics across income groups is a convenience function for mixed-type aggregation.\footnote{See also \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{this blog post on aggregating survey data using \pkg{collapse}} which showcases more aspects of the \fct{collap} function using real census data.} By default, it uses \fct{fmean} for numeric data and \fct{fmode} for categorical data, here also applying population weights such that the most populous country in each group is kept.
 %
-<<fsummarise>>=
-wlddev |> fgroup_by(country, income) |>
-  fsummarise(n = fnobs(LIFEEX),
-             diff = fmax(LIFEEX) %-=% fmin(LIFEEX),
-             mean_diff = fmean(fdiff(LIFEEX)),
-             cor_PCGDP = pwcor(Dlog(LIFEEX), Dlog(PCGDP)),
-             across(c(LIFEEX, PCGDP, POP), flast)) |>
-  collap( ~ income, w = ~ POP, keep.w = FALSE) |> print(digits = 2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlddev |> fgroup_by(country, income) |>
++    fsummarise(n = fnobs(LIFEEX),
++               diff = fmax(LIFEEX) %-=% fmin(LIFEEX),
++               mean_diff = fmean(fdiff(LIFEEX)),
++               cor_PCGDP = pwcor(Dlog(LIFEEX), Dlog(PCGDP)),
++               across(c(LIFEEX, PCGDP, POP), flast)) |>
++    collap( ~ income, w = ~ POP, keep.w = FALSE) |> print(digits = 2)
+\end{Sinput}
+\begin{Soutput}
+        country              income  n diff mean_diff cor_PCGDP LIFEEX PCGDP
+1 United States         High income 60   13      0.23  -0.00828     81 44599
+2      Ethiopia          Low income 60   26      0.43   0.27153     64   691
+3         India Lower middle income 60   24      0.41  -0.15366     69  2324
+4         China Upper middle income 60   27      0.45   0.00077     76  8749
+\end{Soutput}
+\end{Schunk}
 With the exception of \code{pwcor(Dlog(LIFEEX), Dlog(PCGDP))}, which is evaluated for every country, all other expressions in \fct{fsummarise} and also the final \fct{collap} call are fully vectorized, i.e., evaluated only once with grouping information passed directly to the \code{g} arguments of \emph{Fast Statistical and Transformation Functions}.
@@
 \vspace{-10mm}}
-<<fsummarise_2>>=
-wlddev |> fsubset(year >= 2015, income, PCGDP:POP) |>
-  fgroup_by(income) |> fmean(POP)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlddev |> fsubset(year >= 2015, income, PCGDP:POP) |>
++    fgroup_by(income) |> fmean(POP)
+\end{Sinput}
+\begin{Soutput}
+               income   sum.POP PCGDP LIFEEX  GINI        ODA
+1         High income 5.901e+09 43340  80.70 36.14   81398948
+2          Low income 3.296e+09   663  63.05 39.13 2466732035
+3 Lower middle income 1.490e+10  2177  68.31 36.48 2234540565
+4 Upper middle income 1.318e+10  8168  75.51 41.68  -86630117
+\end{Soutput}
+\end{Schunk}
 Using the \emph{Fast Statistical Functions} directly often results in more efficient code.
@@
 %
-<<glinreg>>=
-wlddev |> collap(POP ~ region + year, FUN = fsum) |>
-  fmutate(POP = POP / 1e6) |> fgroup_by(region) |>
-  fmutate(dmy = fmean(year, TRA = "-")) |>
-  fsummarise(beta = fsum(POP, dmy) %/=% fsum(dmy, dmy),
-             POP20 = flast(POP)) |>
-  fmutate(POP21 = POP20 + beta, POP22 = POP21 + beta,
-          POP23 = POP22 + beta, POP24 = POP23 + beta)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlddev |> collap(POP ~ region + year, FUN = fsum) |>
++    fmutate(POP = POP / 1e6) |> fgroup_by(region) |>
++    fmutate(dmy = fmean(year, TRA = "-")) |>
++    fsummarise(beta = fsum(POP, dmy) %/=% fsum(dmy, dmy),
++               POP20 = flast(POP)) |>
++    fmutate(POP21 = POP20 + beta, POP22 = POP21 + beta,
++            POP23 = POP22 + beta, POP24 = POP23 + beta)
+\end{Sinput}
+\begin{Soutput}
+                      region   beta  POP20  POP21  POP22  POP23  POP24
+1        East Asia & Pacific 18.731 2291.4 2310.2 2328.9 2347.6 2366.4
+2      Europe & Central Asia  2.463  921.2  923.7  926.1  928.6  931.0
+3  Latin America & Caribbean  6.460  646.4  652.9  659.4  665.8  672.3
+4 Middle East & North Africa  5.409  456.7  462.1  467.5  472.9  478.3
+5              North America  2.301  365.9  368.2  370.5  372.8  375.1
+6                 South Asia 19.640 1835.8 1855.4 1875.1 1894.7 1914.3
+7         Sub-Saharan Africa 12.852 1103.5 1116.3 1129.2 1142.0 1154.9
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<wstats>>=
-wlddev |> fsubset(is.finite(POP)) |> fgroup_by(region) |>
-  fmutate(o = radixorder(GRPid(), LIFEEX)) |>
-  fsummarise(min = fmin(LIFEEX),
-             Q1 = fnth(LIFEEX, 0.25, POP, o = o, ties = "q8"),
-             mean = fmean(LIFEEX, POP),
-             median = fmedian(LIFEEX, POP, o = o),
-             Q3 = fnth(LIFEEX, 0.75, POP, o = o, ties = "q8"),
-             max = fmax(LIFEEX))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlddev |> fsubset(is.finite(POP)) |> fgroup_by(region) |>
++    fmutate(o = radixorder(GRPid(), LIFEEX)) |>
++    fsummarise(min = fmin(LIFEEX),
++               Q1 = fnth(LIFEEX, 0.25, POP, o = o, ties = "q8"),
++               mean = fmean(LIFEEX, POP),
++               median = fmedian(LIFEEX, POP, o = o),
++               Q3 = fnth(LIFEEX, 0.75, POP, o = o, ties = "q8"),
++               max = fmax(LIFEEX))
+\end{Sinput}
+\begin{Soutput}
+                      region   min    Q1  mean median    Q3   max
+1        East Asia & Pacific 18.91 65.28 68.45  69.67 73.86 85.08
+2      Europe & Central Asia 45.37 68.68 72.30  71.58 76.67 85.42
+3  Latin America & Caribbean 41.76 65.17 69.16  70.87 74.48 82.19
+4 Middle East & North Africa 29.92 61.96 66.65  69.12 72.64 82.80
+5              North America 68.90 73.57 75.54  75.62 78.38 82.05
+6                 South Asia 32.45 55.08 60.19  62.00 66.67 78.92
+7         Sub-Saharan Africa 26.17 46.51 52.53  52.23 58.32 74.51
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<gcomp>>=
-add_vars(g$groups,
- get_vars(wlddev, "country") |> fmode(g, wlddev$POP, use = FALSE),
- get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = F),
- get_vars(wlddev, "POP") |> fsum(g, use = FALSE))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> add_vars(g$groups,
++   get_vars(wlddev, "country") |> fmode(g, wlddev$POP, use = FALSE),
++   get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = F),
++   get_vars(wlddev, "POP") |> fsum(g, use = FALSE))
+\end{Sinput}
+\begin{Soutput}
+               income  OECD       country   PCGDP LIFEEX       POP
+1         High income FALSE  Saudi Arabia 22426.7  73.00 3.114e+09
+2         High income  TRUE United States 31749.6  75.84 5.573e+10
+3          Low income FALSE      Ethiopia   557.1  53.51 2.095e+10
+4 Lower middle income FALSE         India  1238.8  60.59 1.138e+11
+5 Upper middle income FALSE         China  3820.6  68.21 1.114e+11
+6 Upper middle income  TRUE        Mexico  8311.2  69.06 8.162e+09
+\end{Soutput}
+\end{Schunk}
 The above is equivalent to \code{collap(wlddev, country + PCGDP + LIFEEX ~ income + OECD, w = ~ POP)}, which internally toggles many of the same function calls. \newline
@@
 %
-<<gtrans>>=
-add_vars(wlddev) <- get_vars(wlddev, c("PCGDP", "LIFEEX")) |>
-    fwithin(g, mean = "overall.mean") |> add_stub("center_")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> add_vars(wlddev) <- get_vars(wlddev, c("PCGDP", "LIFEEX")) |>
++      fwithin(g, mean = "overall.mean") |> add_stub("center_")
+\end{Sinput}
+\end{Schunk}
 %
@@
 See the Computational details section.}
-<<fsumm_bench>>=
- bmark(collapse = collapse::fsummarise(mtcars, sum = sum(mpg)),
-       dplyr = dplyr::summarise(mtcars, sum = sum(mpg)))
-@
+\begin{Schunk}
+\begin{Sinput}
+R>  bmark(collapse = collapse::fsummarise(mtcars, sum = sum(mpg)),
++         dplyr = dplyr::summarise(mtcars, sum = sum(mpg)))
+\end{Sinput}
+\begin{Soutput}
+  expression      min   median mem_alloc n_itr n_gc total_time
+1   collapse   3.61µs   4.26µs   14.33KB  9999    1    43.81ms
+2      dplyr 211.27µs 240.26µs    3.92MB  7651   52      1.86s
+\end{Soutput}
+\end{Schunk}
 Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices.
@@
 %
-<<ts_example>>=
-fgrowth(airmiles) |> round(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fgrowth(airmiles) |> round(2)
+\end{Sinput}
+\begin{Soutput}
+Time Series:
+Start = 1937 
+End = 1960 
+Frequency = 1 
+ [1]    NA 16.50 42.29 54.03 31.65  2.38 15.23 33.29 54.36 76.92  2.71 -2.10
+[13] 12.91 18.51 32.03 18.57 17.82 13.61 18.19 12.83 13.32  0.01 15.49  4.25
+\end{Soutput}
+\end{Schunk}
 To illustrate the full capabilities of these time series functions, I generate a sector-level trade dataset of export values (v) by country (c), sector (s), and year (y).
 Like many detailed trade datasets, it is unbalanced---not all sectors/products are exported by country c in all years.
-<<exports>>=
-set.seed(101)
-exports <- expand.grid(y = 2001:2010, c = paste0("c", 1:10),
-  s = paste0("s", 1:10)) |> fmutate(v = abs(rnorm(1e3))) |>
-  colorder(c, s) |> fsubset(-sample.int(1e3, 500))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(101)
+R> exports <- expand.grid(y = 2001:2010, c = paste0("c", 1:10),
++    s = paste0("s", 1:10)) |> fmutate(v = abs(rnorm(1e3))) |>
++    colorder(c, s) |> fsubset(-sample.int(1e3, 500))
+\end{Sinput}
+\end{Schunk}
 The following extracts one country-sector series from the \code{exports} dataset.
@@
 %
-<<ts_example_2>>=
-.c(y, v) %=% fsubset(exports, c == "c1" & s == "s7", -c, -s)
-print(y)
-fgrowth(v, t = y) |> round(2)
-fgrowth(v, -1:3, t = y) |> head(4)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> .c(y, v) %=% fsubset(exports, c == "c1" & s == "s7", -c, -s)
+R> print(y)
+\end{Sinput}
+\begin{Soutput}
+[1] 2001 2002 2004 2005 2007 2008 2009 2010
+\end{Soutput}
+\begin{Sinput}
+R> fgrowth(v, t = y) |> round(2)
+\end{Sinput}
+\begin{Soutput}
+[1]     NA 175.52     NA -22.37     NA 624.27 -79.01 534.56
+\end{Soutput}
+\begin{Sinput}
+R> fgrowth(v, -1:3, t = y) |> head(4)
+\end{Sinput}
+\begin{Soutput}
+        FG1     --     G1   L2G1   L3G1
+[1,] -63.71 0.3893     NA     NA     NA
+[2,]     NA 1.0726 175.52     NA     NA
+[3,]  28.82 0.8450     NA -21.22 117.05
+[4,]     NA 0.6559 -22.37     NA -38.85
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<example_growth>>=
-G(exports, -1:2, by = v ~ c + s, t = ~ y) |> head(3)
-tfm(exports, fgrowth(list(v = v), -1:2, g = list(c, s), t = y)) |> head(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> G(exports, -1:2, by = v ~ c + s, t = ~ y) |> head(3)
+\end{Sinput}
+\begin{Soutput}
+   c  s    y  FG1.v      v   G1.v L2G1.v
+1 c1 s1 2002 -18.15 0.5525     NA     NA
+2 c1 s1 2003 214.87 0.6749  22.17     NA
+3 c1 s1 2004 -31.02 0.2144 -68.24  -61.2
+\end{Soutput}
+\begin{Sinput}
+R> tfm(exports, fgrowth(list(v = v), -1:2, g = list(c, s), t = y)) |> head(3)
+\end{Sinput}
+\begin{Soutput}
+   c  s    y      v  FG1.v   G1.v L2G1.v
+1 c1 s1 2002 0.5525 -18.15     NA     NA
+2 c1 s1 2003 0.6749 214.87  22.17     NA
+3 c1 s1 2004 0.2144 -31.02 -68.24  -61.2
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<example_growth_continued>>=
-A <- exports |> fgroup_by(c, s) |> fmutate(gv = G(v, t = y)) |> fungroup()
-head(B <- exports |> fmutate(gv = G(v, g = list(c, s), t = y)), 4)
-identical(A, B)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> A <- exports |> fgroup_by(c, s) |> fmutate(gv = G(v, t = y)) |> fungroup()
+R> head(B <- exports |> fmutate(gv = G(v, g = list(c, s), t = y)), 4)
+\end{Sinput}
+\begin{Soutput}
+   c  s    y      v     gv
+1 c1 s1 2002 0.5525     NA
+2 c1 s1 2003 0.6749  22.17
+3 c1 s1 2004 0.2144 -68.24
+4 c1 s1 2005 0.3108  44.98
+\end{Soutput}
+\begin{Sinput}
+R> identical(A, B)
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\end{Schunk}
 %
@@
 An example using the \code{exports} data follows. Note that data need not be sorted for indexing.
-<<indexing>>=
-exportsi <- exports |> findex_by(c, s, y)
-exportsi |> G() |> head(5)
-exportsi |> findex() |> print(2)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> exportsi <- exports |> findex_by(c, s, y)
+R> exportsi |> G() |> head(5)
+\end{Sinput}
+\begin{Soutput}
+   c  s    y   G1.v
+1 c1 s1 2002     NA
+2 c1 s1 2003  22.17
+3 c1 s1 2004 -68.24
+4 c1 s1 2005  44.98
+5 c1 s1 2006 277.76
+
+Indexed by:  c.s [1] | y [5 (10)] 
+\end{Soutput}
+\begin{Sinput}
+R> exportsi |> findex() |> print(2)
+\end{Sinput}
+\begin{Soutput}
+    c.s    y
+1 c1.s1 2002
+2 c1.s1 2003
+---                
+499 c10.s10 2007
+500 c10.s10 2009
+
+c.s [100] | y [10]
+\end{Soutput}
+\end{Schunk}
 The index statistics are: \code{[N. ids] | [N. periods (total periods: (max-min)/GCD)]}. % This extracts an \class{indexes\_series} and demonstrates its properties.
-<<indexing_2>>=
-vi <- exportsi$v; str(vi, width = 70, strict = "cut")
-is_irregular(vi)
-vi |> psmat() |> head(3)
-fdiff(vi) |> psmat() |> head(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> vi <- exportsi$v; str(vi, width = 70, strict = "cut")
+\end{Sinput}
+\begin{Soutput}
+ 'indexed_series' num [1:500] 0.552 0.675 0.214 0.311 1.174 ...
+ - attr(*, "index_df")=Classes 'index_df', 'pindex' and 'data.frame'..
+  ..$ c.s: Factor w/ 100 levels "c1.s1","c2.s1",..: 1 1 1 1 1 1 1 1 ..
+  ..$ y  : Ord.factor w/ 10 levels "2001"<"2002"<..: 2 3 4 5 6 7 8 9..
+  ..- attr(*, "nam")= chr [1:3] "c" "s" "y"
+\end{Soutput}
+\begin{Sinput}
+R> is_irregular(vi)
+\end{Sinput}
+\begin{Soutput}
+[1] TRUE
+\end{Soutput}
+\begin{Sinput}
+R> vi |> psmat() |> head(3)
+\end{Sinput}
+\begin{Soutput}
+      2001  2002  2003  2004  2005 2006  2007   2008  2009  2010
+c1.s1   NA 0.552 0.675 0.214 0.311 1.17 0.619 0.1127 0.917 0.223
+c2.s1   NA 0.795    NA    NA 0.237   NA    NA 0.0585 0.818    NA
+c3.s1   NA 0.709 0.268 1.464    NA   NA 0.467 0.1193 0.467    NA
+\end{Soutput}
+\begin{Sinput}
+R> fdiff(vi) |> psmat() |> head(3)
+\end{Sinput}
+\begin{Soutput}
+      2001 2002   2003   2004   2005  2006   2007   2008  2009   2010
+c1.s1   NA   NA  0.122 -0.461 0.0964 0.863 -0.555 -0.506 0.804 -0.694
+c2.s1   NA   NA     NA     NA     NA    NA     NA     NA 0.759     NA
+c3.s1   NA   NA -0.441  1.196     NA    NA     NA -0.348 0.348     NA
+\end{Soutput}
+\end{Schunk}
 \fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work in arbitrary data masking environments:
-<<indexing_3>>=
-settransform(exportsi, v_ld = Dlog(v))
-lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> settransform(exportsi, v_ld = Dlog(v))
+R> lm(v_ld ~ L(v_ld, 1:2), exportsi) |> summary() |> coef() |> round(3)
+\end{Sinput}
+\begin{Soutput}
+               Estimate Std. Error t value Pr(>|t|)
+(Intercept)      -0.008      0.141  -0.058    0.954
+L(v_ld, 1:2)L1   -0.349      0.115  -3.042    0.004
+L(v_ld, 1:2)L2   -0.033      0.154  -0.215    0.831
+\end{Soutput}
+\end{Schunk}
 Indexed series/frames also support transformation such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}.
@@
 \end{Code}
-<<joins_3, echo=FALSE>>=
-join(teacher, course, multiple = TRUE, attr = "jn") |> attr("jn") |> str(strict.width = "cut", width = 70)
-@
+\begin{Schunk}
+\begin{Soutput}
+left join: teacher[id] 3/4 (75%) <1:1.33> course[id] 4/5 (80%)
+List of 3
+ $ call   : language join(x = teacher, y = course, multiple = TRUE,"..
+ $ on.cols:List of 2
+  ..$ x: chr "id"
+  ..$ y: chr "id"
+ $ match  : 'qG' int [1:5] 1 2 3 4 NA
+  ..- attr(*, "N.nomatch")= int 1
+  ..- attr(*, "N.groups")= int 5
+  ..- attr(*, "N.distinct")= int 4
+\end{Soutput}
+\end{Schunk}
 Users can also invoke the \code{validate} argument to examine the uniqueness of the join keys in either table: passing a '1' produces an error if the key is not unique.
 %
-<<joins_4>>=
-join(teacher, course, on = "id", validate = "1:1") |>
-  tryCatch(error = function(e) strwrap(e) |> cat(sep = "\n"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> join(teacher, course, on = "id", validate = "1:1") |>
++    tryCatch(error = function(e) strwrap(e) |> cat(sep = "\n"))
+\end{Sinput}
+\begin{Soutput}
+Error in join(teacher, course, on = "id", validate = "1:1"): Join is
+not 1:1: teacher (x) is unique on the join columns; course (y) is
+not unique on the join columns
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<joins_4.5>>=
-join(teacher, course, on = "id", require = list(x = 0.8)) |>
-  tryCatch(error = function(e) substr(e, 102, 200) |> cat())
-@
+\begin{Schunk}
+\begin{Sinput}
+R> join(teacher, course, on = "id", require = list(x = 0.8)) |>
++    tryCatch(error = function(e) substr(e, 102, 200) |> cat())
+\end{Sinput}
+\begin{Soutput}
+Matched 75.0% of records in table teacher (x), but 80.0% is required
+\end{Soutput}
+\end{Schunk}
 %
@@
 This demonstrates the latter:
-<<joins_5>>=
-for (h in c("semi", "anti")) join(teacher, course, how = h) |> print()
-@
+\begin{Schunk}
+\begin{Sinput}
+R> for (h in c("semi", "anti")) join(teacher, course, how = h) |> print()
+\end{Sinput}
+\begin{Soutput}
+semi join: teacher[id] 3/4 (75%) <1:1st> course[id] 3/5 (60%)
+  id names age subject
+1  1  John  35    Math
+2  2  Jane  32    Econ
+3  3   Bob  42   Stats
+anti join: teacher[id] 3/4 (75%) <1:1st> course[id] 3/5 (60%)
+  id names age subject
+1  4  Carl  67   Trade
+\end{Soutput}
+\end{Schunk}
 By default (\code{sort = FALSE}), the order of rows in \code{x} is preserved.
@@
 The following example performs aggregation across years with the internal mean function during a recast pivot.
-<<pivot_agg>>=
-head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"), how = "r",
-     names = list(from = "Variable", to = "Sectorcode"),
-     labels = list(from = "Label", to = "Sector"), FUN = "mean"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"), how = "r",
++       names = list(from = "Variable", to = "Sectorcode"),
++       labels = list(from = "Label", to = "Sector"), FUN = "mean"))
+\end{Sinput}
+\begin{Soutput}
+  Country Sectorcode      Sector       VA      EMP
+1     BWA        AGR Agriculture    462.2   188.06
+2     ETH        AGR Agriculture  34389.9 17624.34
+3     GHA        AGR Agriculture   1549.4  3016.04
+4     KEN        AGR Agriculture 139705.9  5348.91
+5     MWI        AGR Agriculture  28512.6  2762.62
+6     MUS        AGR Agriculture   3819.6    59.34
+\end{Soutput}
+\end{Schunk}
 The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
@@
 Below, I write a nested loop running a regression of agriculture on mining and manuacturing output and employment.
-<<list_proc_2>>=
-result <- list()
-for (country in c("ARG", "BRA", "CHL")) {
-  for (variable in c("EMP", "VA")) {
-    m <- lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year,
-            data = dl[[country]][[variable]])
-    result[[country]][[variable]] <- list(model = m, BIC = BIC(m),
-                                          summary = summary(m))
-  }
-}
-@
+\begin{Schunk}
+\begin{Sinput}
+R> result <- list()
+R> for (country in c("ARG", "BRA", "CHL")) {
++    for (variable in c("EMP", "VA")) {
++      m <- lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year,
++              data = dl[[country]][[variable]])
++      result[[country]][[variable]] <- list(model = m, BIC = BIC(m),
++                                            summary = summary(m))
++    }
++  }
+\end{Sinput}
+\end{Schunk}
 This programming may not be ideal for this particular use case as I could have used data frame-based tools and saved the result in a column.\footnote{E.g., \code{GGDC10S |> fgroup\_by(Country, Variable) |> fsummarise(result = my\_fun(lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year)))} with \code{my\_fun <- function(m) list(list(m, BIC(m), summary(m)))}.}
@@
 We can also apply \fct{t\_list} to turn the list inside-out:
-<<list_proc_3.5>>=
-r_sq_l |> t_list() |> rowbind(idcol = "Variable", return = "data.frame")
-@
+\begin{Schunk}
+\begin{Sinput}
+R> r_sq_l |> t_list() |> rowbind(idcol = "Variable", return = "data.frame")
+\end{Sinput}
+\begin{Soutput}
+  Variable    ARG    BRA    CHL
+1      EMP 0.9068 0.7888 0.1058
+2       VA 0.9996 0.9988 0.9991
+\end{Soutput}
+\end{Schunk}
 \fct{rowbind} is limited if \fct{get\_elem} returns a more nested or asymmetric list, potentially with vectors/arrays in the final nodes. Suppose we want to extract the coefficient matrices:
-<<list_proc_4>>=
-result$ARG$EMP$summary$coefficients
-@
+\begin{Schunk}
+\begin{Sinput}
+R> result$ARG$EMP$summary$coefficients
+\end{Sinput}
+\begin{Soutput}
+              Estimate Std. Error  t value  Pr(>|t|)
+(Intercept)  26.583617  1.2832583  20.7157 1.747e-28
+log(MIN + 1)  0.083168  0.0352493   2.3594 2.169e-02
+log(MAN + 1) -0.064413  0.0767614  -0.8391 4.048e-01
+Year         -0.009683  0.0005556 -17.4278 1.003e-24
+\end{Soutput}
+\end{Schunk}
 For such cases, \fct{unlist2d} provides a complete recursive generalization of \fct{unlist}.
@@
 %
-<<qsu_2>>=
-qsu(wlddev, by = LIFEEX ~ income, w = ~ POP)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> qsu(wlddev, by = LIFEEX ~ income, w = ~ POP)
+\end{Sinput}
+\begin{Soutput}
+                        N       WeightSum     Mean      SD     Min      Max
+High income          3828  5.87959699e+10  75.6926  4.5346  42.672  85.4171
+Low income           1792  2.09491614e+10  53.5061  8.8691  26.172    74.43
+Lower middle income  2790  1.13837685e+11  60.5865  8.3607  18.907   76.699
+Upper middle income  3249  1.19295269e+11  68.2698  7.1889  36.535   80.279
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<descr>>=
-wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
-wlda15 |> descr(by = income + LIFEEX ~ OECD)
-@
+\begin{Schunk}
+\begin{Sinput}
+R> wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
+R> wlda15 |> descr(by = income + LIFEEX ~ OECD)
+\end{Sinput}
+\begin{Soutput}
+Dataset: wlda15, 2 Variables, N = 216
+Grouped by: OECD [2]
+         N   Perc
+FALSE  180  83.33
+TRUE    36  16.67
+-----------------------------------------------------------------------------
+income (factor): Income Level
+Statistics (N = 216)
+         N   Perc  Ndist
+FALSE  180  83.33      4
+TRUE    36  16.67      2
+
+Table (Freq Perc)
+                       FALSE     TRUE    Total
+High income          45 25.0  34 94.4  79 36.6
+Upper middle income  58 32.2   2  5.6  60 27.8
+Lower middle income  47 26.1   0  0.0  47 21.8
+Low income           30 16.7   0  0.0  30 13.9
+-----------------------------------------------------------------------------
+LIFEEX (numeric): Life expectancy at birth, total (years)
+Statistics (N = 200, 7.41% NAs)
+         N  Perc  Ndist   Mean    SD    Min    Max   Skew  Kurt
+FALSE  164    82    164  71.25  7.06  53.28  85.08   -0.5  2.61
+TRUE    36    18     36  80.83  2.55  75.05  84.36  -0.92  2.72
+
+Quantiles
+          1%     5%    10%    25%    50%    75%    90%    95%    99%
+FALSE   54.3  58.38  61.25  66.43  72.56  76.68  78.93  80.88  83.77
+TRUE   75.12  75.83   76.8  79.04  81.77  82.63  83.21  83.54  84.13
+-----------------------------------------------------------------------------
+\end{Soutput}
+\end{Schunk}
 
@@
 %
-<<bench_1, eval=TRUE>>=
-set.seed(101)
-int <- 1:1000; g_int <- sample.int(1000, 1e7, replace = TRUE)
-char <- c(letters, LETTERS, month.abb, month.name)
-g_char <- sample(char <- outer(char, char, paste0), 1e7, TRUE)
-bmark(base_int = unique(g_int), collapse_int = funique(g_int))
-bmark(base_char = unique(g_char), collapse_char = funique(g_char))
-bmark(base_int = match(g_int, int), collapse_int = fmatch(g_int, int))
-bmark(base_char = match(g_char, char), data.table_char =
-      chmatch(g_char, char), collapse_char = fmatch(g_char, char))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set.seed(101)
+R> int <- 1:1000; g_int <- sample.int(1000, 1e7, replace = TRUE)
+R> char <- c(letters, LETTERS, month.abb, month.name)
+R> g_char <- sample(char <- outer(char, char, paste0), 1e7, TRUE)
+R> bmark(base_int = unique(g_int), collapse_int = funique(g_int))
+\end{Sinput}
+\begin{Soutput}
+    expression     min  median mem_alloc n_itr n_gc total_time
+1     base_int 41.49ms  43.2ms   166.2MB    47   47      2.04s
+2 collapse_int  5.47ms  9.28ms    38.2MB   220   44      2.01s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(base_char = unique(g_char), collapse_char = funique(g_char))
+\end{Sinput}
+\begin{Soutput}
+     expression    min median mem_alloc n_itr n_gc total_time
+1     base_char 68.7ms   70ms   166.2MB    28   28      2.01s
+2 collapse_char 13.8ms 15.7ms    38.2MB   123   21      2.01s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(base_int = match(g_int, int), collapse_int = fmatch(g_int, int))
+\end{Sinput}
+\begin{Soutput}
+    expression     min  median mem_alloc n_itr n_gc total_time
+1     base_int 15.13ms 15.58ms    76.3MB    38   67   608.03ms
+2 collapse_int  5.47ms  7.74ms    38.2MB   196   49      1.43s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(base_char = match(g_char, char), data.table_char =
++        chmatch(g_char, char), collapse_char = fmatch(g_char, char))
+\end{Sinput}
+\begin{Soutput}
+       expression    min median mem_alloc n_itr n_gc total_time
+1       base_char 60.8ms 62.4ms   114.5MB    32   32      2.07s
+2 data.table_char 31.9ms   34ms    38.1MB    58   18      2.02s
+3   collapse_char 19.6ms 21.8ms    38.1MB    90   18      2.01s
+\end{Soutput}
+\end{Schunk}
 %
@@
 They are faster than base \proglang{R} even without multithreading, but using 4 threads in this case induces a sizeable difference.
-<<bench_2, eval=TRUE, cache=TRUE>>=
-set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)
-m <- matrix(rnorm(1e7), ncol = 1000)
-bmark(R = colSums(m), collapse = fsum(m))
-bmark(R = colMeans(m), collapse = fmean(m))
-bmark(MS = matrixStats::colMedians(m), collapse = fmedian(m))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> set_collapse(na.rm = FALSE, sort = FALSE, nthreads = 4)
+R> m <- matrix(rnorm(1e7), ncol = 1000)
+R> bmark(R = colSums(m), collapse = fsum(m))
+\end{Sinput}
+\begin{Soutput}
+  expression      min   median mem_alloc n_itr n_gc total_time
+1          R   5.67ms   6.21ms    7.86KB   324    0         2s
+2   collapse 280.93µs 321.64µs    7.86KB  6143    1         2s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(R = colMeans(m), collapse = fmean(m))
+\end{Sinput}
+\begin{Soutput}
+  expression      min   median mem_alloc n_itr n_gc total_time
+1          R   6.01ms   6.21ms   32.09KB   324    0         2s
+2   collapse 278.92µs 322.14µs    7.86KB  6164    0         2s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(MS = matrixStats::colMedians(m), collapse = fmedian(m))
+\end{Sinput}
+\begin{Soutput}
+  expression    min median mem_alloc n_itr n_gc total_time
+1         MS 78.7ms 79.7ms   238.9KB    26    0      2.07s
+2   collapse 19.4ms 19.6ms    27.2KB   101    0      2.02s
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_2.1, eval=TRUE, cache=TRUE>>=
-g <- sample.int(1e3, 1e4, TRUE)
-bmark(R = rowsum(m, g), collapse = fsum(m, g))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> g <- sample.int(1e3, 1e4, TRUE)
+R> bmark(R = rowsum(m, g), collapse = fsum(m, g))
+\end{Sinput}
+\begin{Soutput}
+  expression      min median mem_alloc n_itr n_gc total_time
+1          R   5.36ms 5.59ms    7.87MB   319   20      1.79s
+2   collapse 923.24µs 1.14ms    7.71MB  1279   60      1.58s
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_flights_setup, eval=TRUE, cache=TRUE>>=
-fastverse_extend(nycflights23, dplyr, data.table); setDTthreads(4)
-list(flights, airports, airlines, planes, weather) |> sapply(nrow)
-flights |> fselect(month, day, origin, dest) |> fnunique()
-@
+\begin{Schunk}
+\begin{Sinput}
+R> fastverse_extend(nycflights23, dplyr, data.table); setDTthreads(4)
+R> list(flights, airports, airlines, planes, weather) |> sapply(nrow)
+\end{Sinput}
+\begin{Soutput}
+[1] 435352   1255     14   4840  26207
+\end{Soutput}
+\begin{Sinput}
+R> flights |> fselect(month, day, origin, dest) |> fnunique()
+\end{Sinput}
+\begin{Soutput}
+[1] 75899
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_flights_summarise, eval=TRUE, cache=TRUE>>=
-vars <- .c(dep_delay, arr_delay, air_time, distance, hour, minute)
-bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
-                summarise(across(all_of(vars), sum), .groups = "drop"),
-      data.table = qDT(flights)[, lapply(.SD, sum), .SDcols = vars,
-                                by = .(month, day, origin, dest)],
-      collapse = flights |> fgroup_by(month, day, origin, dest) |>
-                   get_vars(vars) |> fsum())
-@
+\begin{Schunk}
+\begin{Sinput}
+R> vars <- .c(dep_delay, arr_delay, air_time, distance, hour, minute)
+R> bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
++                  summarise(across(all_of(vars), sum), .groups = "drop"),
++        data.table = qDT(flights)[, lapply(.SD, sum), .SDcols = vars,
++                                  by = .(month, day, origin, dest)],
++        collapse = flights |> fgroup_by(month, day, origin, dest) |>
++                     get_vars(vars) |> fsum())
+\end{Sinput}
+\begin{Soutput}
+  expression      min   median mem_alloc n_itr n_gc total_time
+1      dplyr 437.44ms 471.82ms    50.5MB     5   50      2.35s
+2 data.table   7.56ms   8.59ms    20.8MB   208   18         2s
+3   collapse      3ms   3.75ms     9.2MB   467   21         2s
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_flights_summarise_2, eval=TRUE, cache=TRUE, echo=FALSE>>=
-bmark(dplyr_mean = flights |> group_by(month, day, origin, dest) |>
-              summarise(across(all_of(vars), mean), .groups = "drop"),
-      data.table_mean = qDT(flights)[, lapply(.SD, mean), .SDcols = vars,
-                                by = .(month, day, origin, dest)],
-      collapse_mean = flights |> fgroup_by(month, day, origin, dest) |>
-                 get_vars(vars) |> fmean())
-bmark(dplyr_median = flights |> group_by(month, day, origin, dest) |>
-              summarise(across(all_of(vars), median), .groups = "drop"),
-      data.table_median = qDT(flights)[, lapply(.SD, median), .SDcols = vars,
-                                by = .(month, day, origin, dest)],
-      collapse_median = flights |> fgroup_by(month, day, origin, dest) |>
-                 get_vars(vars) |> fmedian())
-@
+\begin{Schunk}
+\begin{Soutput}
+       expression    min median mem_alloc n_itr n_gc total_time
+1      dplyr_mean  1.17s  1.18s   48.57MB     2   42      2.36s
+2 data.table_mean 7.64ms 8.79ms   18.93MB   190   18         2s
+3   collapse_mean 3.28ms 3.77ms    9.11MB   470   17      2.01s
+\end{Soutput}
+\begin{Soutput}
+         expression     min  median mem_alloc n_itr n_gc total_time
+1      dplyr_median   4.05s   4.05s    52.8MB     1   86      4.05s
+2 data.table_median 21.45ms 22.51ms    18.9MB    86    8      2.01s
+3   collapse_median   9.7ms 10.34ms    11.1MB   171   11         2s
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_flights_summarise_3, eval=TRUE, cache=TRUE>>=
-bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
-    summarise(rng = max(arr_delay) - min(arr_delay), .groups = "drop"),
-  data.table = qDT(flights)[, .(rng = max(arr_delay) - min(arr_delay)),
-                            by = .(month, day, origin, dest)],
-  collapse = flights |> fgroup_by(month, day, origin, dest) |>
-    fsummarise(rng = fmax(arr_delay) - fmin(arr_delay)))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> bmark(dplyr = flights |> group_by(month, day, origin, dest) |>
++      summarise(rng = max(arr_delay) - min(arr_delay), .groups = "drop"),
++    data.table = qDT(flights)[, .(rng = max(arr_delay) - min(arr_delay)),
++                              by = .(month, day, origin, dest)],
++    collapse = flights |> fgroup_by(month, day, origin, dest) |>
++      fsummarise(rng = fmax(arr_delay) - fmin(arr_delay)))
+\end{Sinput}
+\begin{Soutput}
+  expression     min  median mem_alloc n_itr n_gc total_time
+1      dplyr 90.25ms 99.01ms   20.18MB    20   44      2.09s
+2 data.table 59.54ms 63.14ms    5.77MB    32   27      2.04s
+3   collapse  5.32ms  5.88ms     6.8MB   303   11         2s
+\end{Soutput}
+\end{Schunk}
 %
@@
 \end{Code}
-<<bench_flights_join_setup, eval=TRUE, cache=TRUE, echo=FALSE>>=
-flights |> join(weather, on = c("origin", "time_hour")) |>
-  join(planes, on = "tailnum") |> join(airports, on = c(dest = "faa")) |>
-  join(airlines, on = "carrier") |> dim() |>
-  capture.output() |> substr(1, 76) |> cat(sep = "\n")
-@
+\begin{Schunk}
+\begin{Soutput}
+left join: flights[origin, time_hour] 434526/435352 (99.8%) <21.94:1st> weat
+duplicate columns: year, month, day, hour => renamed using suffix '_weather'
+left join: x[tailnum] 424068/435352 (97.4%) <87.62:1st> planes[tailnum] 4840
+duplicate columns: year => renamed using suffix '_planes' for y
+left join: x[dest] 435352/435352 (100%) <3689.42:1st> airports[faa] 118/1255
+left join: x[carrier] 435352/435352 (100%) <31096.57:1st> airlines[carrier] 
+duplicate columns: name => renamed using suffix '_airlines' for y
+[1] 435352     48
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_flights_join, eval=TRUE, cache=TRUE, echo=FALSE>>=
-bmark(
-  dplyr_joins = flights |>
-    left_join(weather, by = c("origin", "time_hour"), multiple = "first") |>
-    left_join(planes, by = "tailnum", multiple = "first") |>
-    left_join(airports, by = c(dest = "faa"), multiple = "first") |>
-    left_join(airlines, by = "carrier", multiple = "first"),
-
-  data.table_joins = qDT(airlines)[qDT(airports)[qDT(planes)[qDT(weather)[qDT(flights), on = c("origin", "time_hour"), mult = "first"], on = "tailnum", mult = "first"], on = c("faa" = "dest"), mult = "first"], on = "carrier", mult = "first"],
-
-  collapse_joins = flights |>
-    join(weather, on = c("origin", "time_hour"), verbose = 0) |>
-    join(planes, on = "tailnum", verbose = 0) |>
-    join(airports, on = c("dest" = "faa"), verbose = 0) |>
-    join(airlines, on = "carrier", verbose = 0)
-)
-@
+\begin{Schunk}
+\begin{Soutput}
+        expression      min  median mem_alloc n_itr n_gc total_time
+1      dplyr_joins 140.85ms 173.7ms   559.4MB    12   28      2.12s
+2 data.table_joins  99.45ms 132.9ms     491MB    15   36      2.11s
+3   collapse_joins   9.15ms  11.2ms    89.7MB   142   40      2.01s
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_pivot_long, eval=TRUE, cache=TRUE, max.print=50>>=
-bmark(tidyr = tidyr::pivot_longer(flights, cols = vars),
-      data.table = qDT(flights) |> melt(measure = vars),
-      collapse = pivot(flights, values = vars))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> bmark(tidyr = tidyr::pivot_longer(flights, cols = vars),
++        data.table = qDT(flights) |> melt(measure = vars),
++        collapse = pivot(flights, values = vars))
+\end{Sinput}
+\begin{Soutput}
+  expression    min median mem_alloc n_itr n_gc total_time
+1      tidyr 58.2ms 58.2ms     255MB     1   27     58.2ms
+2 data.table 41.9ms 43.9ms     209MB    14   22    613.6ms
+3   collapse   13ms 15.3ms     209MB    34   57    512.9ms
+\end{Soutput}
+\end{Schunk}
 %
@@
 %
-<<bench_pivot_wide, eval=TRUE, cache=TRUE, max.print=50>>=
-bmark(tidyr = tidyr::pivot_wider(flights, id_cols = .c(month, day, dest),
-          names_from = "origin", values_from = vars, values_fn = sum),
-      data.table = dcast(qDT(flights), month + day + dest ~ origin,
-                         value.var = vars, fun = sum),
-      collapse_fsum = pivot(flights, .c(month, day, dest), vars,
-                            "origin", how = "wider", FUN = fsum),
-      collapse_itnl = pivot(flights, .c(month, day, dest), vars,
-                            "origin", how = "wider", FUN = "sum"))
-@
+\begin{Schunk}
+\begin{Sinput}
+R> bmark(tidyr = tidyr::pivot_wider(flights, id_cols = .c(month, day, dest),
++            names_from = "origin", values_from = vars, values_fn = sum),
++        data.table = dcast(qDT(flights), month + day + dest ~ origin,
++                           value.var = vars, fun = sum),
++        collapse_fsum = pivot(flights, .c(month, day, dest), vars,
++                              "origin", how = "wider", FUN = fsum),
++        collapse_itnl = pivot(flights, .c(month, day, dest), vars,
++                              "origin", how = "wider", FUN = "sum"))
+\end{Sinput}
+\begin{Soutput}
+     expression      min   median mem_alloc n_itr n_gc total_time
+1         tidyr 353.82ms 383.22ms     143MB     5   37         2s
+2    data.table 222.83ms 230.79ms    21.7MB     9   40      2.08s
+3 collapse_fsum   5.04ms   7.01ms    39.1MB   233   23      2.07s
+4 collapse_itnl   4.12ms   4.51ms    12.4MB   392    8      2.01s
+\end{Soutput}
+\end{Schunk}
 %

### 2025-05-05 misc/JSS article/final/article.Rnw
Minor updates.

@@
 # Package versions used in the article:
-# fastverse 0.3.4, collapse 2.1.2, data.table 1.16.4,
-# bench 1.1.3, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.0.0
+# fastverse 0.3.4, collapse 2.1.2, data.table 1.17.0,
+# bench 1.1.4, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.5.0
+pkg <- c("bench", "dplyr", "tidyr", "matrixStats", "janitor")
+if(!all(avail <- is_installed(pkg))) install.packages(pkg[!avail])
 
@@
 bmark <- function(...) {
-  bench::mark(..., min_time = 0.01, check = FALSE) |> janitor::clean_names() |>
+  bench::mark(..., min_time = 2, check = FALSE) |> janitor::clean_names() |>
   fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
@@
 I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.github.io/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages.
-These benchmarks align in showing that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and on personal computers.
+These benchmarks align in showing that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and personal computers.
 % \footnote{Reasons for the particularly strong performance of \pkg{collapse} on Windows may be that it is largely written in C and has limited multithreading in favor or highly efficient serial algorithms---there appear to be persistent obstacles to efficient (low-overhead) multithreading on Windows, implying that multithreaded query engines do not develop their full thrust on medium-sized ($\leq$100 million row) datasets.}
@@
 \section*{Computational details}
-The results in this paper were obtained using \proglang{R} \citep{R} 4.3.0 with \pkg{collapse} 2.1.2, \pkg{data.table} 1.16.4, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.0.0, \pkg{fastverse} 0.3.4, and \pkg{bench} 1.1.3 \citep{rbench}.
+The results in this paper were obtained using \proglang{R} \citep{R} 4.4.3 with \pkg{collapse} 2.1.2, \pkg{data.table} 1.17.0, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.5.0, \pkg{fastverse} 0.3.4, and \pkg{bench} 1.1.4 \citep{rbench}.
 All packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at https://CRAN.R-project.org/.
 The benchmark was run on an Apple MacBook Pro (2024) with 48GB M4 Pro chip.
-Packages were compiled from source using Homebrew Clang version 16.0.4 with OpenMP enabled and the -O2 flag.
+Packages were compiled from source using Apple Clang version 17.0.0 with OpenMP enabled and the -O2 flag.
 \newline

### 2025-05-05 misc/JSS article/final/article.Rnw
Minors.

@@
 # fastverse 0.3.4, collapse 2.1.2, data.table 1.17.0,
-# bench 1.1.4, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.5.0
+# bench 1.1.4, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.5.0, janitor 2.2.0
 pkg <- c("bench", "dplyr", "tidyr", "matrixStats", "janitor")

### 2025-05-05 misc/JSS article/final/article.Rnw
Resetting global options.

@@
 rm(list = ls()); gc()
-options(prompt = "R> ", continue = "+  ", width = 77, digits = 4, useFancyQuotes = FALSE, warn = 1)
+r_opts <- options(prompt = "R> ", continue = "+  ", width = 77, digits = 4, useFancyQuotes = FALSE, warn = 1)
 
@@
 # Reset collapse options (if set)
-set_collapse(nthreads = 1L, remove = NULL, stable.algo = TRUE, sort = TRUE,
+oldopts <- set_collapse(nthreads = 1L, remove = NULL, stable.algo = TRUE, sort = TRUE,
              digits = 2L, stub = TRUE, verbose = 1L, mask = NULL, na.rm = TRUE)
@@
 
+<<reset_options, echo=FALSE>>=
+options(r_opts)
+set_collapse(oldopts)
+@
+
 \section*{Acknowledgments}

### 2025-05-05 misc/JSS article/final/article_replication.R

@@ -12,7 +12,7 @@
 ###################################################
 
 rm(list = ls()); gc()
-options(prompt = "R> ", continue = "+  ", width = 77, digits = 4, useFancyQuotes = FALSE, warn = 1)
+r_opts <- options(prompt = "R> ", continue = "+  ", width = 77, digits = 4, useFancyQuotes = FALSE, warn = 1)
 
 # Loading libraries and installing if unavailable
 if(!requireNamespace("fastverse", quietly = TRUE)) install.packages("fastverse")
@@ -25,7 +25,7 @@ pkg <- c("bench", "dplyr", "tidyr", "matrixStats", "janitor")
 if(!all(avail <- is_installed(pkg))) install.packages(pkg[!avail])
 
 # Reset collapse options (if set)
-set_collapse(nthreads = 1L, remove = NULL, stable.algo = TRUE, sort = TRUE,
+oldopts <- set_collapse(nthreads = 1L, remove = NULL, stable.algo = TRUE, sort = TRUE,
              digits = 2L, stub = TRUE, verbose = 1L, mask = NULL, na.rm = TRUE)
 
 # For presentation: removing whitespace around labels
@@ -550,6 +550,12 @@ bmark(tidyr = tidyr::pivot_wider(flights, id_cols = .c(month, day, dest),
                             "origin", how = "wider", FUN = fsum),
       collapse_itnl = pivot(flights, .c(month, day, dest), vars,
                             "origin", how = "wider", FUN = "sum"))
+
+###################################################
+### code chunk number 64: reset_options
+###################################################
+options(r_opts)
+set_collapse(oldopts)
 
 ###################################################
 ### Print Session Information

### 2025-05-05 misc/JSS article/final/article.Rnw
Add nycflights23.

@@
 \section*{Computational details}
-The results in this paper were obtained using \proglang{R} \citep{R} 4.4.3 with \pkg{collapse} 2.1.2, \pkg{data.table} 1.17.0, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.5.0, \pkg{fastverse} 0.3.4, and \pkg{bench} 1.1.4 \citep{rbench}.
+The results in this paper were obtained using \proglang{R} \citep{R} 4.4.3 with \pkg{collapse} 2.1.2, \pkg{data.table} 1.17.0, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.5.0, \pkg{fastverse} 0.3.4, \pkg{nycflights23} 0.2.0, and \pkg{bench} 1.1.4 \citep{rbench}.
 All packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at https://CRAN.R-project.org/.

### 2025-05-05 misc/JSS article/final/article_replication.R

@@ -19,9 +19,9 @@ if(!requireNamespace("fastverse", quietly = TRUE)) install.packages("fastverse")
 options(fastverse.styling = FALSE)
 library(fastverse) # loads data.table, collapse, magrittr and kit (not used)
 # Package versions used in the article:
-# fastverse 0.3.4, collapse 2.1.2, data.table 1.17.0,
+# fastverse 0.3.4, collapse 2.1.2, data.table 1.17.0, nycflights23 0.2.0,
 # bench 1.1.4, dplyr 1.1.4, tidyr 1.3.1, matrixStats 1.5.0, janitor 2.2.0
-pkg <- c("bench", "dplyr", "tidyr", "matrixStats", "janitor")
+pkg <- c("bench", "dplyr", "tidyr", "matrixStats", "janitor", "nycflights23")
 if(!all(avail <- is_installed(pkg))) install.packages(pkg[!avail])
 
 # Reset collapse options (if set)

### 2025-05-05 misc/JSS article/final/article.Rnw

--- 2025-05-05 misc/JSS article/final/article.Rnw
@@
 \section*{Computational details}
-The results in this paper were obtained using \proglang{R} \citep{R} 4.4.3 with \pkg{collapse} 2.1.2, \pkg{data.table} 1.17.0, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.5.0, \pkg{fastverse} 0.3.4, and \pkg{bench} 1.1.4 \citep{rbench}.
+The results in this paper were obtained using \proglang{R} \citep{R} 4.4.3 with \pkg{collapse} 2.1.2, \pkg{data.table} 1.17.0, \pkg{dplyr} 1.1.4, \pkg{tidyr} 1.3.1, \pkg{matrixStats} 1.5.0, \pkg{fastverse} 0.3.4, \pkg{nycflights23} 0.2.0 \citep{rnycflights23}, and \pkg{bench} 1.1.4 \citep{rbench}.
 All packages used are available from the Comprehensive \proglang{R} Archive Network (CRAN) at https://CRAN.R-project.org/.
-The benchmark was run on an Apple MacBook Pro (2024) with 48GB M4 Pro chip.
+The benchmark was run on an Apple MacBook Pro (2024) with a 48GB M4 Pro processor (single core speed $\sim$4.4 GHz).
 Packages were compiled from source using Apple Clang version 17.0.0 with OpenMP enabled and the -O2 flag.

### 2025-05-05 misc/JSS article/final/article.Rnw
Remaining R -> \proglang{R}.

@@
 This functionality is supported by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels.
-Lastly, it includes efficient object converters, functions for memory efficient R programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
+Lastly, it includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
 \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations for vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem---notably, \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
@@
 
-R already has a large and tested data manipulation and statistical computing ecosystem.
+\proglang{R} already has a large and tested data manipulation and statistical computing ecosystem.
 Notably, the \pkg{tidyverse} \citep{rtidyverse} provides a consistent toolkit for data manipulation in R, centered around the \class{tibble} \citep{rtibble} object and tidy data principles \citep{rtidydata}.
@@
 
-As evident from previous sections, \pkg{collapse} provides performance-improved or otherwise enhanced versions of functionality already present in base R (like the \emph{Fast Statistical Functions}, \fct{funique}, \fct{fmatch}, \fct{fsubset}, \fct{ftransform}, etc.) and other packages (esp.
+As evident from previous sections, \pkg{collapse} provides performance-improved or otherwise enhanced versions of functionality already present in base \proglang{R} (like the \emph{Fast Statistical Functions}, \fct{funique}, \fct{fmatch}, \fct{fsubset}, \fct{ftransform}, etc.) and other packages (esp.
 \pkg{dplyr} \citep{rdplyr}: \fct{fselect}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, etc.).
@@
 
-Last but not least, \pkg{collapse} users are encouraged to also familiarize themselves with the \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse}},\footnote{Website: https://fastverse.github.io/fastverse/} a suite of complementary high-performance packages for statistical computing and data manipulation in R which offer more advanced tools in several statistical computing domains.
+Last but not least, \pkg{collapse} users are encouraged to also familiarize themselves with the \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse}},\footnote{Website: https://fastverse.github.io/fastverse/} a suite of complementary high-performance packages for statistical computing and data manipulation in \proglang{R} which offer more advanced tools in several statistical computing domains.
 The \pkg{fastverse} metapackage also provides a lightweight framework to jointly load and manage these packages, as well as to build customized and fully separate package verses.

### 2025-05-05 misc/JSS article/final/article.Rnw
Also replacing C/C++ -> \proglang{C/C++}.

@@
 \href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/.
-Linecount (v2.1.2): R: 13,785, C: 19,013, C++: 9,844.
+Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844.
 \hphantom{aaa.}Exported namespace: 392 objects, of which 238 functions (excl.
@@
 %
-The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R}/\pkg{collapse} means that these operations are implemented using compiled C/C++ code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
+The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R}/\pkg{collapse} means that these operations are implemented using compiled \proglang{C/C++} code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
 Their basic syntax is
@@
 %
-When \fct{fsummarise} evaluates an expression involving \emph{Fast Statistical Functions}, it sets their \code{g} argument with a grouping (\class{GRP}) object that is directly handed to C/C++, and also sets \code{use.g.names = FALSE}.
+When \fct{fsummarise} evaluates an expression involving \emph{Fast Statistical Functions}, it sets their \code{g} argument with a grouping (\class{GRP}) object that is directly handed to \proglang{C/C++}, and also sets \code{use.g.names = FALSE}.
 Hence, weights (\code{w}) becomes the second positional argument.
@@
 Whereas the \code{g} argument supports ad-hoc grouping with vectors and lists/data frames, for repeated operations the cost of grouping can be minimized by using factors (see \code{?qF} for efficient factor generation) or \class{GRP} objects as inputs.
-The latter contain all information \pkg{collapse}'s statistical functions may require to operate across groups and are thus passed to internal C/C++ code without checks.
+The latter contain all information \pkg{collapse}'s statistical functions may require to operate across groups and are thus passed to internal \proglang{C/C++} code without checks.
 They can be created with \code{GRP()}.
@@
 This warranted some design choices away from maximum performance for large data manipulation.\footnote{Which nowadays would demand creating a multithreaded, vectorized query engine with optimized memory buffers/vector types to take full advantage of SIMD processing as in \pkg{DuckDB} or \pkg{Polars}.
-Such an architecture is very difficult to square with \proglang{R} vectors and \proglang{R}'s 30-year old C API.} Its limited use of multithreading and SIMD instructions, partly by design constraints and by R's C API, and the use of standard types for internal indexing, imposes hard-limits---the maximum integer in \proglang{R} is 2,147,483,647 $\to$ the maximum vector length \pkg{collapse} supports.
+Such an architecture is very difficult to square with \proglang{R} vectors and \proglang{R}'s 30-year old \proglang{C} API.} Its limited use of multithreading and SIMD instructions, partly by design constraints and by \proglang{R}'s \proglang{C} API, and the use of standard types for internal indexing, imposes hard-limits---the maximum integer in \proglang{R} is 2,147,483,647 $\to$ the maximum vector length \pkg{collapse} supports.
 It is and will remain an in-memory tool.

### 2025-05-05 misc/JSS article/final/article.Rnw
Further fixes to join()/pivot() section.

@@
 join(x, y, on = NULL, how = "left", suffix = NULL, validate = "m:m",
-  multiple = FALSE, sort = FALSE, keep.col.order = TRUE,
-  drop.dup.cols = FALSE, verbose = 1, column = NULL, attr = NULL, ...)
+  multiple = FALSE, sort = FALSE, keep.col.order = TRUE, verbose = 1,
+  drop.dup.cols = FALSE, require = NULL, column = NULL, attr = NULL, ...)
 \end{Code}
@@
 pivot(data, ids = NULL, values = NULL, names = NULL, labels = NULL,
-  how = "longer", na.rm = FALSE, check.dups = FALSE, ...)
+  how = "longer", na.rm = FALSE, check.dups = FALSE, FUN = "last", ...)
 \end{Code}
@@
 head(dw <- pivot(data, c("Country", "Year"), names = "Variable",
-                 labels = "Label", how = "w"))
+                 labels = "Label", how = "wider"))
 namlab(dw)
@@
   names = list(from = "Variable", to = "Sectorcode"),
-             labels = list(from = "Label", to = "Sector"), how = "r"))
+  labels = list(from = "Label", to = "Sector"), how = "recast"))
 vlabels(dr)[3:6]
@@
 <<pivot_agg>>=
-head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"), how = "r",
-     names = list(from = "Variable", to = "Sectorcode"),
+head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"),
+  how = "recast", names = list(from = "Variable", to = "Sectorcode"),
   labels = list(from = "Label", to = "Sector"), FUN = "mean"))

### 2025-05-05 misc/JSS article/final/article.tex
Further fixes to join()/pivot() section.

@@
 R> head(dw <- pivot(data, c("Country", "Year"), names = "Variable",
-+                   labels = "Label", how = "w"))
++                   labels = "Label", how = "wider"))
 \end{Sinput}
@@
 +    names = list(from = "Variable", to = "Sectorcode"),
-+               labels = list(from = "Label", to = "Sector"), how = "r"))
++    labels = list(from = "Label", to = "Sector"), how = "recast"))
 \end{Sinput}
@@
 \begin{Sinput}
-R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"), how = "r",
-+       names = list(from = "Variable", to = "Sectorcode"),
+R> head(dr_agg <- pivot(data, "Country", c("AGR", "MIN", "MAN"),
++    how = "recast", names = list(from = "Variable", to = "Sectorcode"),
 +    labels = list(from = "Label", to = "Sector"), FUN = "mean"))

### 2025-05-06 misc/JSS article/final/article.Rnw
Improve langauge.

@@
 Its core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}).
-It also provides powerful data manipulation functions---such as vectorized and verbose hash-joins or fast (aggregation, recast) pivots---functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools.
+It also provides powerful data manipulation functions, such as vectorized and verbose hash-joins or fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools.
 This functionality is supported by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels.
-Lastly, it includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
-\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations for vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem---notably, \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
-It is also globally and interactively configurable, allowing changing defaults of key function arguments, such as the \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the namespace to mask equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
+It also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
+\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations for vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
+It is also globally and interactively configurable, allowing changing defaults of key function arguments, such as the \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the package namespace to mask equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
 See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline
@@
 A particular focus has been on providing tools for examining longitudinal (panel) data.
-Recall the indexed World Development panel dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}).
+Recall the World Development panel dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{sec:integration}.
 The function \fct{varying} can be used to examine which of these variables are time-varying.
@@
 A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time.
-Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (see Table~\ref{tab:TSfun}), this is easily demonstrated.
+Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (see Table~\ref{tab:TSfun}), this is easily demonstrated:
 %
 <<pdec>>=
-LIFEEXi <- reindex(wlddev$LIFEEX, wlddev$iso3c)
+LIFEEXi <- reindex(wlddev$LIFEEX, index = wlddev$iso3c)
 all.equal(fvar(LIFEEXi), fvar(fbetween(LIFEEXi)) + fvar(fwithin(LIFEEXi)))
@@
 %
- It also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'click-through' the variables in a large dataset while spreading the computational burden.
+ It also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden.
 The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples.

### 2025-05-07 misc/JSS article/final/article.Rnw
More time.

@@
 bmark <- function(...) {
-  bench::mark(..., min_time = 2, check = FALSE) |> janitor::clean_names() |>
+  bench::mark(..., min_time = 3, check = FALSE) |> janitor::clean_names() |>
   fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
@@
 bmark <- function(...) {
-  bench::mark(..., min_time = 2, check = FALSE) |>
+  bench::mark(..., min_time = 3, check = FALSE) |>
     janitor::clean_names() |>

### 2025-05-07 misc/JSS article/final/article.tex
Update benchmarks.

@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1   collapse   3.61µs   4.26µs   14.33KB  9999    1    43.81ms
-2      dplyr 211.27µs 240.26µs    3.92MB  7651   52      1.86s
+1   collapse   3.32µs   4.18µs   14.33KB  9999    1    43.08ms
+2      dplyr    200µs 238.91µs    1.53MB  9934   66      2.48s
 \end{Soutput}
@@
 A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time.
-Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (see Table~\ref{tab:TSfun}), this is easily demonstrated.
-%
-\begin{Schunk}
-\begin{Sinput}
-R> LIFEEXi <- reindex(wlddev$LIFEEX, wlddev$iso3c)
+Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (see Table~\ref{tab:TSfun}), this is easily demonstrated:
+%
+\begin{Schunk}
+\begin{Sinput}
+R> LIFEEXi <- reindex(wlddev$LIFEEX, index = wlddev$iso3c)
 R> all.equal(fvar(LIFEEXi), fvar(fbetween(LIFEEXi)) + fvar(fwithin(LIFEEXi)))
@@
     expression     min  median mem_alloc n_itr n_gc total_time
-1     base_int 41.49ms  43.2ms   166.2MB    47   47      2.04s
-2 collapse_int  5.47ms  9.28ms    38.2MB   220   44      2.01s
+1     base_int 43.78ms  45.3ms   166.2MB    65   65         3s
+2 collapse_int  6.25ms  8.24ms    38.2MB   353   71         3s
 \end{Soutput}
@@
      expression    min median mem_alloc n_itr n_gc total_time
-1     base_char 68.7ms   70ms   166.2MB    28   28      2.01s
-2 collapse_char 13.8ms 15.7ms    38.2MB   123   21      2.01s
+1     base_char 69.9ms 71.3ms   166.2MB    43   43      3.07s
+2 collapse_char 13.9ms 15.6ms    38.2MB   180   36      3.01s
 \end{Soutput}
@@
     expression     min  median mem_alloc n_itr n_gc total_time
-1     base_int 15.13ms 15.58ms    76.3MB    38   67   608.03ms
-2 collapse_int  5.47ms  7.74ms    38.2MB   196   49      1.43s
+1     base_int 16.92ms  17.4ms    76.3MB    78   78      1.37s
+2 collapse_int  6.95ms  7.68ms    38.2MB   270   75      2.08s
 \end{Soutput}
@@
        expression    min median mem_alloc n_itr n_gc total_time
-1       base_char 60.8ms 62.4ms   114.5MB    32   32      2.07s
-2 data.table_char 31.9ms   34ms    38.1MB    58   18      2.02s
-3   collapse_char 19.6ms 21.8ms    38.1MB    90   18      2.01s
+1       base_char 58.8ms 58.8ms   114.5MB     1   48    58.78ms
+2 data.table_char 32.2ms 33.7ms    38.1MB    66   19      2.22s
+3   collapse_char 19.5ms 20.4ms    38.1MB   113   28      2.31s
 \end{Soutput}
@@
   expression     min  median mem_alloc n_itr n_gc total_time
-1          R   5.67ms   6.21ms    7.86KB   324    0         2s
-2   collapse 280.93µs 321.64µs    7.86KB  6143    1         2s
+1          R   5.4ms   6.2ms    7.86KB   493    0         3s
+2   collapse 282.9µs 324.5µs    7.86KB  9150    1         3s
 \end{Soutput}
@@
   expression      min  median mem_alloc n_itr n_gc total_time
-1          R   6.01ms   6.21ms   32.09KB   324    0         2s
-2   collapse 278.92µs 322.14µs    7.86KB  6164    0         2s
+1          R   5.38ms   6.2ms   32.09KB   491    0         3s
+2   collapse 280.93µs 327.2µs    7.86KB  8975    0         3s
 \end{Soutput}
@@
   expression    min median mem_alloc n_itr n_gc total_time
-1         MS 78.7ms 79.7ms   238.9KB    26    0      2.07s
-2   collapse 19.4ms 19.6ms    27.2KB   101    0      2.02s
+1         MS 77.8ms 79.4ms    89.9KB    38    0      3.02s
+2   collapse 19.1ms 19.5ms    27.2KB   154    0      3.01s
 \end{Soutput}
@@
   expression     min median mem_alloc n_itr n_gc total_time
-1          R   5.36ms 5.59ms    7.87MB   319   20      1.79s
-2   collapse 923.24µs 1.14ms    7.71MB  1279   60      1.58s
+1          R   5.4ms 5.62ms    7.87MB   459   35      2.61s
+2   collapse 900.2µs 1.13ms    7.71MB  1876  104      2.19s
 \end{Soutput}
@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1      dplyr 437.44ms 471.82ms    50.5MB     5   50      2.35s
-2 data.table   7.56ms   8.59ms    20.8MB   208   18         2s
-3   collapse      3ms   3.75ms     9.2MB   467   21         2s
+1      dplyr 363.83ms 479.56ms    50.5MB     7   65      3.15s
+2 data.table   7.22ms   7.99ms    20.8MB   339   33         3s
+3   collapse   2.96ms    3.4ms     9.2MB   769   33         3s
 \end{Soutput}
@@
        expression    min median mem_alloc n_itr n_gc total_time
-1      dplyr_mean  1.17s  1.18s   48.57MB     2   42      2.36s
-2 data.table_mean 7.64ms 8.79ms   18.93MB   190   18         2s
-3   collapse_mean 3.28ms 3.77ms    9.11MB   470   17      2.01s
+1      dplyr_mean  1.16s  1.17s   48.57MB     3   78      3.52s
+2 data.table_mean 7.43ms 8.31ms   18.93MB   283   28         3s
+3   collapse_mean  3.1ms 3.71ms    9.11MB   679   28         3s
 \end{Soutput}
@@
          expression     min  median mem_alloc n_itr n_gc total_time
-1      dplyr_median   4.05s   4.05s    52.8MB     1   86      4.05s
-2 data.table_median 21.45ms 22.51ms    18.9MB    86    8      2.01s
-3   collapse_median   9.7ms 10.34ms    11.1MB   171   11         2s
+1      dplyr_median    4.2s    4.2s    52.8MB     1   91       4.2s
+2 data.table_median 21.32ms 22.24ms    18.9MB   130   13         3s
+3   collapse_median  9.34ms  9.96ms    11.1MB   271   12         3s
 \end{Soutput}
@@
   expression     min   median mem_alloc n_itr n_gc total_time
-1      dplyr 90.25ms 99.01ms   20.18MB    20   44      2.09s
-2 data.table 59.54ms 63.14ms    5.77MB    32   27      2.04s
-3   collapse  5.32ms  5.88ms     6.8MB   303   11         2s
+1      dplyr    93ms 103.73ms   20.18MB    27   56      3.03s
+2 data.table 56.51ms  60.06ms    5.77MB    50   25      3.06s
+3   collapse  5.04ms   5.51ms     6.8MB   509   13         3s
 \end{Soutput}
@@
         expression      min  median mem_alloc n_itr n_gc total_time
-1      dplyr_joins 140.85ms 173.7ms   559.4MB    12   28      2.12s
-2 data.table_joins  99.45ms 132.9ms     491MB    15   36      2.11s
-3   collapse_joins   9.15ms  11.2ms    89.7MB   142   40      2.01s
+1      dplyr_joins 139.95ms   153ms   559.4MB    19   32      3.12s
+2 data.table_joins  97.39ms   129ms     491MB    19   39      3.11s
+3   collapse_joins   8.85ms  11.9ms    89.7MB   188   51      3.01s
 \end{Soutput}
@@
   expression    min median mem_alloc n_itr n_gc total_time
-1      tidyr 58.2ms 58.2ms     255MB     1   27     58.2ms
-2 data.table 41.9ms 43.9ms     209MB    14   22    613.6ms
-3   collapse   13ms 15.3ms     209MB    34   57    512.9ms
+1      tidyr 56.9ms 58.8ms     254MB     8   39      472ms
+2 data.table   40ms 42.6ms     209MB    22   32      962ms
+3   collapse   13ms   15ms     209MB    47   84      713ms
 \end{Soutput}
@@
      expression      min   median mem_alloc n_itr n_gc total_time
-1         tidyr 353.82ms 383.22ms     143MB     5   37         2s
-2    data.table 222.83ms 230.79ms    21.7MB     9   40      2.08s
-3 collapse_fsum   5.04ms   7.01ms    39.1MB   233   23      2.07s
-4 collapse_itnl   4.12ms   4.51ms    12.4MB   392    8      2.01s
+1         tidyr 379.96ms 397.91ms     143MB     8   59      3.27s
+2    data.table    227ms  231.1ms    21.7MB    13   47      3.03s
+3 collapse_fsum   5.61ms   7.18ms    39.1MB   327   26      3.01s
+4 collapse_itnl   3.77ms   4.45ms    12.4MB   611   15         3s
 \end{Soutput}

### 2025-05-25 misc/JSS article/final/article.Rnw
Minors.

@@
 \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations for vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
-It is also globally and interactively configurable, allowing changing defaults of key function arguments, such as the \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the package namespace to mask equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
+It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as the \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modification of the package namespace to mask equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
 See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline

### 2025-05-26 misc/JSS article/final/article.Rnw
Tweaks and hyphenation.

--- 2025-05-07 misc/JSS article/final/article.Rnw
@@
 \newcommand{\fct}[1]{\code{#1()}}
+
+% Hypthens
+\hyphenation{mod-i-fi-ca-tions ma-nip-u-la-tion trans-for-ma-tion per-for-mance de-vel-op-ment
+             ef-fi-cient in-tro-duces ca-pa-bil-i-ties gen-er-al-iz-ing en-abling pro-gram-ming
+             com-pu-ta-tions eval-u-at-ed fa-cil-i-tate im-ple-men-ta-tions re-place-ment
+             sub-set-ting de-com-po-si-tion al-go-rith-mic en-hanced mul-ti-ple coun-tries
+             com-pu-ta-tion-al par-tic-u-lar-ly}
+\usepackage[english]{babel}
+% \sloppy
+% \usepackage{microtype}
 
@@
 bmark <- function(...) {
-  bench::mark(..., min_time = 3, check = FALSE) |> janitor::clean_names() |>
+  bench::mark(..., min_time = 0.01, check = FALSE) |> janitor::clean_names() |>
   fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
@@
 \pkg{collapse} is a large \proglang{C/C++}-based infrastructure package facilitating complex statistical computing, data transformation, and exploration tasks in \proglang{R}---at outstanding levels of performance and memory efficiency.
-It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular extensions (\class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, \class{pdata.frame}), enabling its seamless integration with large parts of the \proglang{R} ecosystem.
+It also implements a class-agnostic approach to \proglang{R} programming, supporting vector, matrix and data frame-like objects and their popular extensions (\class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, \class{pdata.frame}), enabling its seamless integration with the bulk of the \proglang{R} ecosystem.
 This article introduces the package's key components and design principles in a structured way, supported by a rich set of examples.
@@
 %% - Should be comma-separated and in sentence case.
-\Keywords{statistical computing, vectorization, data manipulation and transformation, class-agnostic programming, summary statistics, \proglang{C/C++}, \proglang{R}}
+\Keywords{statistical computing, vectorization, data manipulation and transformation,\\\hphantom{aaaaaaaaa.}class-agnostic programming, summary statistics, \proglang{C/C++}, \proglang{R}}
 \Plainkeywords{statistical computing, vectorization, data transformation and manipulation, class-agnostic programming, summary statistics, C/C++, R}
@@
 % Most of these statistical functions are vectorized along multiple dimensions (notably along groups and columns) and perform high-cardinality operations\footnote{With many columns and/or groups relative to data size.} very efficiently.
-Its core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}).
+Core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}).
 It also provides powerful data manipulation functions, such as vectorized and verbose hash-joins or fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools.
-This functionality is supported by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels.
-It also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
-\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations for vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
-It is also globally and interactively configurable, allowing changing defaults of key function arguments, such as the \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the package namespace to mask equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
+This functionality is powered by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels.
+The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
+\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
+It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications to the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
 See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline
@@
 The more elaborate answer is to (1) facilitate complex data transformation, exploration, and computing tasks in \proglang{R}; (2) increase the performance and memory efficiency of \proglang{R} programs; % \footnote{Principally by avoiding \proglang{R}-level repetition such as applying \proglang{R} functions across columns/groups using a split-apply-combine logic, but also by avoiding object conversions and the need for certain classes to do certain things, such as converting matrices to \class{data.frame} or \class{data.table} just to compute statistics by groups.}
-and (3) create a new foundation package for statistics and data transformation that implements successful ideas in the \proglang{R} ecosystem and other programming environments in a stable, high performance, and broadly compatible manner.\footnote{Such ideas include \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\proglang{Python}, \pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}: \citet{STATA}), indexed time series and panel data (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), variable labels (\proglang{STATA}), recast pivots (\pkg{reshape2}), etc...} \newline
+and (3) create a new foundation package for statistics and data manipulation that implements many successful ideas in the \proglang{R} ecosystem and other programming environments in a stable, high performance, and broadly compatible manner.\footnote{Such ideas include \pkg{tidyverse} syntax, vectorized aggregations (\pkg{data.table}), data transformation by reference (\pkg{pandas}), vectorized and verbose joins (\pkg{polars}, \proglang{STATA}: \citet{STATA}), indexed time/panel series (\pkg{xts}, \pkg{plm}), summary statistics for panel data (\proglang{STATA}), variable labels (\proglang{STATA}), recast pivots (\pkg{reshape2}), etc...} \newline
 
@@
 
-Other programming environments such as \proglang{Python} and \proglang{Julia} now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars}, and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy}, or \pkg{StatsBase.jl} \citep{jlstatsbase}.
+Other programming environments such as \proglang{Python} and \proglang{Julia} by now also offer computationally very powerful libraries for tabular data such as \pkg{DataFrames.jl} \citep{jldataframes}, \pkg{Polars} \citep{pypolars}, and \pkg{Pandas} \citep{mckinney2010pandas, pypandas}, and supporting numerical libraries such as \pkg{Numpy} \citep{pynumpy}, or \pkg{StatsBase.jl} \citep{jlstatsbase}.
 % \pkg{NaNStatistics.jl} \citep{jlnanstatistics}
@@
 In short, it is very useful for complex statistical workflows, rich datasets (e.g., surveys), and for integrating with different parts of the \proglang{R} ecosystem.
-On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not highly competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes.
+On the other hand, \pkg{collapse}, for the most part, does not offer a sub-column-level parallel architecture and is thus not extremely competitive with top frameworks, including \pkg{data.table}, on aggregating billion-row datasets with few columns.\footnote{As can be seen in the \href{https://duckdblabs.github.io/db-benchmark/}{DuckDB Benchmarks}: \pkg{collapse} is highly competitive on the 10-100 million observations datasets, but deteriorates in performance at larger data sizes.
 There may be performance improvements for very "long data" in the future, but, at present, the treatment of columns as fundamental units of computation (in most cases) is a tradeoff for the highly flexible class-agnostic architecture.} Its vectorization capabilities are also limited to the statistical functions it provides and not, like \pkg{DataFrames.jl}, to any \proglang{Julia} function.
@@
 %
-This article cannot fully present \pkg{collapse}, but the following sections introduce its key components, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options.
+This article does not fully present \pkg{collapse}, but the following sections introduce its key components, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options.
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}, including the \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} $+$ \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
+For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} $+$ \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
 %
@@
 Additionally, a \code{set} argument can be passed to \emph{Fast Statistical Functions} to toggle transformation by reference.
-For example \code{fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill", set = TRUE)} would modify \code{Sepal.Length} in-place and return the result invisibly.
+For example \code{fmean(iris$Sepal.Length, g = iris$Species, TRA = "fill", set = TRUE)} would modify \code{Sepal.Length} in-place and return it invisibly.
 \newline
@@
 <<fscale>>=
-fscale(iris$Sepal.Length, g = iris$Species) |> fsd(iris$Species)
+fscale(iris$Sepal.Length, g = iris$Species) |> fsd(g = iris$Species)
 @
@@
 @
-Below, I track changes in Life Expectancy at the country level, computing the range, average change per year, and correlation with GDP per Capita in log-differences\footnote{\fct{Dlog} is a shorthand for \code{fdiff(log(x))} in \pkg{collapse}, \fct{pwcor} wraps \code{cor(..., use = "pairwise.complete.obs")}.} while also saving the latest measurements.\footnote{Note that \fct{fsummarise} never re-groups data, making a call to \fct{fungroup} redundant.} The \fct{collap} function called at the end to aggregate the country-level statistics across income groups is a convenience function for mixed-type aggregation.\footnote{See also \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{this blog post on aggregating survey data using \pkg{collapse}} which showcases more aspects of the \fct{collap} function using real census data.} By default, it uses \fct{fmean} for numeric data and \fct{fmode} for categorical data, here also applying population weights such that the most populous country in each group is kept.
+Below, I track changes in Life Expectancy at the country level, computing the range, average change per year, and the correlation with GDP per Capita in log-differences\footnote{\fct{Dlog} abbreviates \code{fdiff(log(x))}, \fct{pwcor} wraps \code{cor(..., use = "pairwise.complete.obs")}.} while also saving the latest measurements.\footnote{Note that \fct{fsummarise} never re-groups data, making a call to \fct{fungroup} redundant.} The \fct{collap} function called at the end to aggregate the country-level statistics across income groups is a convenience function for mixed-type aggregation.
+By default, it uses \fct{fmean} for numeric data and \fct{fmode} for categorical data, here also applying population weights such that the most populous country is selected.\footnote{See also \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{this blog post on aggregating survey data using \pkg{collapse}} which showcases more aspects of the \fct{collap} function using real census data.}
 %
@@
 @
-With the exception of \code{pwcor(Dlog(LIFEEX), Dlog(PCGDP))}, which is evaluated for every country, all other expressions in \fct{fsummarise} and also the final \fct{collap} call are fully vectorized, i.e., evaluated only once with grouping information passed directly to the \code{g} arguments of \emph{Fast Statistical and Transformation Functions}.
-Using subtraction by reference in the range is thus memory efficient as the executed expression is equivalent to \code{fmax(LIFFEX, country) \%-=\% fmin(LIFFEX, country)}.
-\fct{across} also directly invokes \fct{flast.data.frame} on the subset of columns.
+With the exception of \code{pwcor(Dlog(LIFEEX), Dlog(PCGDP))}, which is evaluated for every country, all other expressions in \fct{fsummarise} and also the final \fct{collap} call are fully vectorized, i.e., \emph{Fast Statistical and Transformation Functions} are evaluated only once with grouping information passed to their \code{g} arguments.
+Using subtraction by reference in the range is more memory efficient as the executed expression is equivalent to \code{fmax(LIFFEX, country) \%-=\% fmin(LIFFEX, country)}.
+\fct{across} here directly invokes \fct{flast.data.frame} on the subset of columns.
 %
@@
 @
-Using the \emph{Fast Statistical Functions} directly often results in more efficient code.
-For example, computing population shares for each year as \code{fmutate(wlddev, POP\_share = fsum(POP, year, TRA = "/"))} is considerably more efficient than \code{wlddev |> fgroup_by(year) |> fmutate(POP\_share = fsum(POP, TRA = "/")) |> fungroup()}, which in turn is more efficient than \code{POP / fsum(POP)} (extra allocation) or \code{proportions(POP)} (split-apply-combine).
+Emplyoing the \emph{Fast Statistical Functions} directly often results in more efficient code.
+For example, computing population shares for each year as \code{fmutate(wlddev, POP\_share = fsum\\(POP, year, TRA = "/"))} is considerably more efficient than \code{wlddev |> fgroup_by(year) |> fmutate(POP\_share = fsum(POP, TRA = "/")) |> fungroup()}, which in turn is more efficient than using \code{POP / fsum(POP)} (extra full-length allocation) or \code{proportions(POP)} (slow split-apply-combine evaluation logic) inside the grouped \fct{fmutate} call.
 \newline
@@
 \pkg{collapse} thus offers several ways to reach the same outcome.
-This may be confusing, but all options can be distinguished in efficiency terms.
-It is thus useful to be mindful of computer resources and attempt to craft minimalistic solutions.
+This may be confusing at first, but all options can be distinguished in terms of efficiency.
+It is thus helpful to be mindful of computer resources and attempt to craft minimalistic solutions to complex problems.
 For example, I recently combined multiple spatial datasets on points of interest (POIs) and needed to deduplicate them.
-I decided to keep the richest source for each location and POI type.
-After creating POI confidence, location, and type indicators comparable across sources (datasets), my deduplication expression ended up being a single line of the form \code{fsubset(data, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (i.e., richest) source by location and type.
-This is very efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}.
+I decided to keep the richest source for each location and POI type/category.
+After creating comparable POI confidence, location, and type indicators, my deduplication expression ended up being a single line of the form \code{fsubset(data, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (richest) source by location and type.
+This is efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}---\emph{Fast Statistical Functions} internal grouped logic is highly optimized.
 %
@@
 %
-\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, implying that using any \emph{Fast Statistical Function} in an expression causes the whole expression to be vectorized.
-This only applies to visible expressions, \pkg{collapse} cannot read the contents of custom functions, thus such functions are always evaluated split-apply-combine.
-The eager vectorization approach enables efficient grouped calculation of more complex statistics.
+\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, such that using any \emph{Fast Statistical Function} in an expression causes the entire expression to be vectorized (i.e., evaluated only once with \emph{Fast Statistical Function} returning grouped output).
+This only applies to visible expressions, \pkg{collapse} cannot read the contents of custom functions, thus such functions are always evaluated using split-apply-combine logic.
+This eager vectorization approach enables efficient grouped calculations of more complex statistics.
 Below, I forecast the population of each region via linear regression (\code{POP ~ year}) in a fully vectorized way.
@@
 Similarly, \fct{fmutate} sets \code{g} and \code{TRA = "fill"}, which can be overwritten by the user (here with \code{TRA = "-"}).
-The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)}, amounting to \code{cov(x, y)/var(y)}, is thus vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation.
+The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)} amounts to \code{cov(x, y)/var(y)}, but is vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation.
 In a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post}, I forecasted high-resolution population estimates for South Africa like this.
-Using 1 $km^2$ \href{https://www.worldpop.org/}{WorldPop} data available for years 2014-2020, running 1.6 million cell-regressions and obtaining 2 forecasts for 2021 and 2022 took only 0.3 seconds on my M1 Mac.
+Using 1 $km^2$ \href{https://www.worldpop.org/}{WorldPop} data available for years 2014-2020, I ran 1.6 million cell-regressions and obtained 2 forecasts for 2021 and 2022 in less than 0.3 seconds on my M1 Mac.
 Another neat example from the community, shared by Andrew Ghazi in a \href{https://andrewghazi.github.io/posts/collapse\_is\_sick/sick.html}{blog post}, vectorizes an expression to compute the $p$~value, \code{2 * pt(abs(fmean(x) * sqrt(6) / fsd(x)), 5, lower.tail = FALSE)}, across 300,000 groups for a simulation study, yielding a 70x performance increase over \pkg{dplyr}.
@@
 %
-Weighted quantiles have a sub-column parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \fct{fnth}/\fct{fmedian}/\fct{fmode} (default 1).} and, as shown above, can also harness an (optional) optimization utilizing an overall ordering vector (combining groups and the data column) to avoid repeated partial sorting of the same elements within groups.
+Weighted quantiles have a sub-column level parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \fct{fnth}/\fct{fmedian}/\fct{fmode} (default 1).} and, as shown above, can harness an (optional) optimization utilizing an overall ordering vector (combining groups and the data column) to avoid repeated sorting of the same elements within groups.
 % \newline
@@
 \class{GRP} objects make grouped statistical computations in \pkg{collapse} fully programmable.
-I can employ the object with the \emph{Fast Statistical Functions} and some utilities\footnote{\fct{add\_vars} is a fast \fct{cbind.data.frame} which also has an assignment method, and \fct{get\_vars} enables fast and secure extraction of data frame columns.} to efficiently aggregate GDP per capita, life expectancy, and country name, with population weights.
+I can employ the object with the \emph{Fast Statistical Functions} and some utilities\footnote{\fct{add\_vars} is a fast \fct{cbind.data.frame} which also has an assignment method, and \fct{get\_vars} enables fast and secure extraction of data frame columns.
+\code{use = FALSE} abbreviates \code{use.g.names = FALSE}.} to efficiently aggregate GDP per capita, life expectancy, and country name, again applying population weights.
 %
@@
 
-The low-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
+The lower-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
 Users should note that \pkg{collapse} does not provide metaprogramming capabilities in its non-standard evaluation functions---such as \href{https://rlang.r-lib.org/reference/topic-quosure.html}{quosures} or \href{https://dplyr.tidyverse.org/articles/programming.html}{indirection} familiar to \pkg{tidyverse} users.
-Instead, there exist standard-evaluation equivalents to some of these functions which typically end with a \code{v} (for 'variables'), such as \fct{collapv}, \fct{fslicev}, \fct{ftransformv}, etc.
-In other cases, such as those typically handled by \fct{fsummarise} or \fct{fmutate}, users may need to use the low-level API for programming, or resort to \fct{substitute} and friends.
+Instead, it has standard-evaluation equivalents to some of these functions which typically end with a \code{v} for 'variables', such as \fct{collapv}, \fct{fslicev}, \fct{ftransformv}, etc.
+In other cases, including those typically handled by \fct{fsummarise} or \fct{fmutate}, users need to use the lower-level API for programming, or resort to \fct{substitute} and friends.
 The main reason for not providing higher-level metaprogramming capabilities is to keep all functions as simple as possible.
-It also compels users to think deeper about their programs and devise more efficient solutions.
-\newline
-
-These metaprogramming restrictions are also vindicated by internally complex data manipulation functions producing greater overheads that can noticeably add up in larger scripts:\footnote{\fct{bmark} is a slim wrapper around \fct{bench::mark}.
+It also compels users to think deeply about their programs and devise more efficient solutions.
+\newline
+
+These metaprogramming restrictions are vindicated by internally complex data manipulation functions producing greater overheads that can noticeably add up in larger scripts:\footnote{\fct{bmark} is a slim wrapper around \fct{bench::mark}.
 See the Computational details section.}
@@
 Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices.
-In the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case) - see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
+For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 44.7 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
 %On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another recent example involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}.
@@
 %
+\newpage
 %
@@
 
-An example using the \code{exports} data follows. Note that data need not be sorted for indexing.
+An example using the \code{exports} data follows. \emph{Note} that data need not be sorted for indexing.
 <<indexing>>=
@@
 @
-\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work in arbitrary data masking environments:
+\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
 <<indexing_3>>=
@@
 Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series.
-The \class{index\_df} attached to these objects can also be used with other general tools such as \code{collapse::BY()} to perform grouped computations using 3rd-party functions.
-An example of calculating a 5-year rolling average is given below (\fct{ix} abbreviates \fct{findex}).
+The \class{index\_df} attached to these objects can also be used with other general tools such as \fct{collapse::BY} to perform grouped computations using 3rd-party functions.
+For example, below I compute a 5-year rolling average (\fct{ix} abbreviates \fct{findex}).
 %
@@
 The reshaping/pivoting functionality of both commercial and open-source software is also presently unsatisfactory for complex datasets such as surveys or disaggregated production, trade, or financial sector data, where variable names resemble codes and variable labels are essential to making sense of the data.
-Such datasets can presently only be reshaped by losing these labels or additional manual efforts to retain them.
+Such datasets can presently only be reshaped by losing these labels or manual efforts to retain them.
 Modern \proglang{R} packages also offer different reshaping functions, such as \fct{data.table::melt}/\fct{tidyr::pivot\_longer} to combine columns and \fct{data.table::dcast}/\fct{tidyr::pivot\_wider} to expand them, requiring users to learn both.
@@
 
-\code{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings.
+\fct{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings.
 It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information.
@@
 For example, the same result could have been obtained by specifying \code{values = c("AGR", "MIN", "MAN")} instead of \code{ids = 1:4}.
-An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", "wider")}, with \code{dl} the long data.
+An exact reverse operation can also be specified as \code{pivot(dl, 1:4, "Value", "Sectorcode", "Sector", how = "wider")}, where \code{dl} is the long data.
 \newline
@@
 The latter can be checked with \code{check.dups = TRUE}.
-There are a small number of internal functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}.
-These carry out computations 'on the fly' and are thus extremely fast.
-\fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but require a deep copy of the columns aggregated which is avoided using the internal functions.
+A small number extremely fast internal aggregation functions, \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"} operate 'on the fly' while reshaping.
+\fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but require a deep copy of the columns aggregated which is avoided by the internal functions.
 The following example performs aggregation across years with the internal mean function during a recast pivot.
@@
 The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
-Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.
+Notably, it can also perform longer and recast pivots without id variables, similar to \code{data.table::transpose()}.
 %
@@
 A typical use case in development involves serving different data according to user choices.
-Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data.
+Except for certain recursive functions in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data.
 \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit.
@@
 <<qsu_2>>=
-qsu(wlddev, by = LIFEEX ~ income, w = ~ POP)
+wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
+qsu(wlda15, by = LIFEEX ~ income, w = ~ POP)
 @
@@
 <<descr>>=
-wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
-wlda15 |> descr(by = income + LIFEEX ~ OECD)
+descr(wlda15, by = income + LIFEEX ~ OECD)
 @
@@
 I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.github.io/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages.
-These benchmarks align in showing that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and personal computers.
+These benchmarks align that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and on PCs/Macs.
 % \footnote{Reasons for the particularly strong performance of \pkg{collapse} on Windows may be that it is largely written in C and has limited multithreading in favor or highly efficient serial algorithms---there appear to be persistent obstacles to efficient (low-overhead) multithreading on Windows, implying that multithreaded query engines do not develop their full thrust on medium-sized ($\leq$100 million row) datasets.}

### 2025-05-26 misc/JSS article/final/article.tex
Tweaks and hyphenation.

--- 2025-05-07 misc/JSS article/final/article.tex
@@
 \newcommand{\fct}[1]{\code{#1()}}
+
+% Hypthens
+\hyphenation{mod-i-fi-ca-tions ma-nip-u-la-tion trans-for-ma-tion per-for-mance de-vel-op-ment
+             ef-fi-cient in-tro-duces ca-pa-bil-i-ties gen-er-al-iz-ing en-abling pro-gram-ming
+             com-pu-ta-tions eval-u-at-ed fa-cil-i-tate im-ple-men-ta-tions re-place-ment
+             sub-set-ting de-com-po-si-tion al-go-rith-mic en-hanced mul-ti-ple coun-tries com-pu-ta-tion-al}
+\usepackage[english]{babel}
+% \sloppy
+% \usepackage{microtype}
 
@@
 \begin{Sinput}
-R> fscale(iris$Sepal.Length, g = iris$Species) |> fsd(iris$Species)
+R> fscale(iris$Sepal.Length, g = iris$Species) |> fsd(g = iris$Species)
 \end{Sinput}
@@
 \end{Schunk}
-Below, I track changes in Life Expectancy at the country level, computing the range, average change per year, and correlation with GDP per Capita in log-differences\footnote{\fct{Dlog} is a shorthand for \code{fdiff(log(x))} in \pkg{collapse}, \fct{pwcor} wraps \code{cor(..., use = "pairwise.complete.obs")}.} while also saving the latest measurements.\footnote{Note that \fct{fsummarise} never re-groups data, making a call to \fct{fungroup} redundant.} The \fct{collap} function called at the end to aggregate the country-level statistics across income groups is a convenience function for mixed-type aggregation.\footnote{See also \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{this blog post on aggregating survey data using \pkg{collapse}} which showcases more aspects of the \fct{collap} function using real census data.} By default, it uses \fct{fmean} for numeric data and \fct{fmode} for categorical data, here also applying population weights such that the most populous country in each group is kept.
+Below, I track changes in Life Expectancy at the country level, computing the range, average change per year, and the correlation with GDP per Capita in log-differences\footnote{\fct{Dlog} abbreviates \code{fdiff(log(x))}, \fct{pwcor} wraps \code{cor(..., use = "pairwise.complete.obs")}.} while also saving the latest measurements.\footnote{Note that \fct{fsummarise} never re-groups data, making a call to \fct{fungroup} redundant.} The \fct{collap} function called at the end to aggregate the country-level statistics across income groups is a convenience function for mixed-type aggregation.
+By default, it uses \fct{fmean} for numeric data and \fct{fmode} for categorical data, here also applying population weights such that the most populous country is selected.\footnote{See also \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{this blog post on aggregating survey data using \pkg{collapse}} which showcases more aspects of the \fct{collap} function using real census data.}
 %
@@
 \end{Schunk}
-With the exception of \code{pwcor(Dlog(LIFEEX), Dlog(PCGDP))}, which is evaluated for every country, all other expressions in \fct{fsummarise} and also the final \fct{collap} call are fully vectorized, i.e., evaluated only once with grouping information passed directly to the \code{g} arguments of \emph{Fast Statistical and Transformation Functions}.
-Using subtraction by reference in the range is thus memory efficient as the executed expression is equivalent to \code{fmax(LIFFEX, country) \%-=\% fmin(LIFFEX, country)}.
-\fct{across} also directly invokes \fct{flast.data.frame} on the subset of columns.
+With the exception of \code{pwcor(Dlog(LIFEEX), Dlog(PCGDP))}, which is evaluated for every country, all other expressions in \fct{fsummarise} and also the final \fct{collap} call are fully vectorized, i.e., \emph{Fast Statistical and Transformation Functions} are evaluated only once with grouping information passed to their \code{g} arguments.
+Using subtraction by reference in the range is more memory efficient as the executed expression is equivalent to \code{fmax(LIFFEX, country) \%-=\% fmin(LIFFEX, country)}.
+\fct{across} here directly invokes \fct{flast.data.frame} on the subset of columns.
 %
@@
 \end{Schunk}
-Using the \emph{Fast Statistical Functions} directly often results in more efficient code.
-For example, computing population shares for each year as \code{fmutate(wlddev, POP\_share = fsum(POP, year, TRA = "/"))} is considerably more efficient than \code{wlddev |> fgroup_by(year) |> fmutate(POP\_share = fsum(POP, TRA = "/")) |> fungroup()}, which in turn is more efficient than \code{POP / fsum(POP)} (extra allocation) or \code{proportions(POP)} (split-apply-combine).
+Emplyoing the \emph{Fast Statistical Functions} directly often results in more efficient code.
+For example, computing population shares for each year as \code{fmutate(wlddev, POP\_share = fsum(\\POP, year, TRA = "/"))} is considerably more efficient than \code{wlddev |> fgroup_by(year) |> fmutate(POP\_share = fsum(POP, TRA = "/")) |> fungroup()}, which in turn is more efficient than using \code{POP / fsum(POP)} (extra full-length allocation) or \code{proportions(POP)} (slow split-apply-combine evaluation logic) inside the grouped \fct{fmutate} call.
 \newline
@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1   collapse   3.32µs   4.18µs   14.33KB  9999    1    43.08ms
-2      dplyr    200µs 238.91µs    1.53MB  9934   66      2.48s
+1   collapse   3.81µs   4.22µs        0B  2235    0     9.91ms
+2      dplyr 239.44µs 275.77µs    1.92KB    33    0    10.35ms
 \end{Soutput}
@@
 
-An example using the \code{exports} data follows. Note that data need not be sorted for indexing.
+An example using the \code{exports} data follows. \emph{Note} that data need not be sorted for indexing.
 \begin{Schunk}
@@
 \end{Schunk}
-\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work in arbitrary data masking environments:
+\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
 \begin{Schunk}
@@
 The \class{index\_df} attached to these objects can also be used with other general tools such as \code{collapse::BY()} to perform grouped computations using 3rd-party functions.
-An example of calculating a 5-year rolling average is given below (\fct{ix} abbreviates \fct{findex}).
+For example, below I compute a 5-year rolling average (\fct{ix} abbreviates \fct{findex}).
 %
@@
 \begin{Sinput}
-R> qsu(wlddev, by = LIFEEX ~ income, w = ~ POP)
+R> wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
+R> qsu(wlda15, by = LIFEEX ~ income, w = ~ POP)
 \end{Sinput}
@@
                       N       WeightSum     Mean      SD      Min     Max
-High income          3828  5.87959699e+10  75.6926  4.5346  42.672  85.4171
-Low income           1792  2.09491614e+10  53.5061  8.8691  26.172    74.43
-Lower middle income  2790  1.13837685e+11  60.5865  8.3607  18.907   76.699
-Upper middle income  3249  1.19295269e+11  68.2698  7.1889  36.535   80.279
+High income          68  1.19122607e+09   80.879   2.441  70.6224  85.078
+Low income           29      694'893773  63.8061  3.9266   53.283  72.697
+Lower middle income  47  3.06353648e+09  68.7599  4.7055   54.331  76.699
+Upper middle income  55  2.67050662e+09  75.9476  2.3895   58.735  80.279
 \end{Soutput}
@@
 \begin{Sinput}
-R> wlda15 <- wlddev |> fsubset(year >= 2015) |> fgroup_by(iso3c) |> flast()
-R> wlda15 |> descr(by = income + LIFEEX ~ OECD)
+R> descr(wlda15, by = income + LIFEEX ~ OECD)
 \end{Sinput}
@@
     expression    min median mem_alloc n_itr n_gc total_time
-1     base_int 43.78ms  45.3ms   166.2MB    65   65         3s
-2 collapse_int  6.25ms  8.24ms    38.2MB   353   71         3s
+1     base_int 46.5ms 46.5ms   166.2MB     1    0     46.5ms
+2 collapse_int 10.5ms 10.5ms    38.2MB     1    0     10.5ms
 \end{Soutput}
@@
      expression    min median mem_alloc n_itr n_gc total_time
-1     base_char 69.9ms 71.3ms   166.2MB    43   43      3.07s
-2 collapse_char 13.9ms 15.6ms    38.2MB   180   36      3.01s
+1     base_char 69.8ms 69.8ms   166.2MB     1    0     69.8ms
+2 collapse_char 20.6ms 20.6ms    38.2MB     1    0     20.6ms
 \end{Soutput}
@@
     expression     min  median mem_alloc n_itr n_gc total_time
-1     base_int 16.92ms  17.4ms    76.3MB    78   78      1.37s
-2 collapse_int  6.95ms  7.68ms    38.2MB   270   75      2.08s
+1     base_int 27.28ms 27.28ms    76.3MB     1    1     27.3ms
+2 collapse_int  7.93ms  8.01ms    38.2MB     2    0       16ms
 \end{Soutput}
@@
        expression    min median mem_alloc n_itr n_gc total_time
-1       base_char 58.8ms 58.8ms   114.5MB     1   48    58.78ms
-2 data.table_char 32.2ms 33.7ms    38.1MB    66   19      2.22s
-3   collapse_char 19.5ms 20.4ms    38.1MB   113   28      2.31s
+1       base_char 72.4ms 72.4ms   114.5MB     1    0     72.4ms
+2 data.table_char 46.5ms 46.5ms    38.1MB     1    1     46.5ms
+3   collapse_char   29ms   29ms    38.1MB     1    0       29ms
 \end{Soutput}
@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1          R   5.4ms   6.2ms    7.86KB   493    0         3s
-2   collapse 282.9µs 324.5µs    7.86KB  9150    1         3s
+1          R   6.07ms   6.09ms    7.86KB     2    0     12.2ms
+2   collapse 311.31µs 345.01µs    7.86KB    28    0     10.1ms
 \end{Soutput}
@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1          R   5.38ms   6.2ms   32.09KB   491    0         3s
-2   collapse 280.93µs 327.2µs    7.86KB  8975    0         3s
+1          R   6.22ms   6.22ms    7.86KB     2    0     12.4ms
+2   collapse 329.27µs 341.08µs    7.86KB    27    0     10.3ms
 \end{Soutput}
@@
   expression    min median mem_alloc n_itr n_gc total_time
-1         MS 77.8ms 79.4ms    89.9KB    38    0      3.02s
-2   collapse 19.1ms 19.5ms    27.2KB   154    0      3.01s
+1         MS 77.5ms 77.5ms   86.04KB     1    0     77.5ms
+2   collapse 19.5ms 19.5ms    7.86KB     1    0     19.5ms
 \end{Soutput}
@@
   expression    min median mem_alloc n_itr n_gc total_time
-1          R   5.4ms 5.62ms    7.87MB   459   35      2.61s
-2   collapse 900.2µs 1.13ms    7.71MB  1876  104      2.19s
+1          R 6.04ms 6.29ms    7.85MB     2    0     12.6ms
+2   collapse  1.3ms 1.39ms    7.67MB     8    0     11.2ms
 \end{Soutput}
@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1      dplyr 363.83ms 479.56ms    50.5MB     7   65      3.15s
-2 data.table   7.22ms   7.99ms    20.8MB   339   33         3s
-3   collapse   2.96ms    3.4ms     9.2MB   769   33         3s
+1      dplyr 369.09ms 369.09ms   48.56MB     1    6    369.1ms
+2 data.table   8.36ms   8.46ms   18.93MB     2    0     16.9ms
+3   collapse   4.24ms   4.29ms    9.11MB     3    0     12.9ms
 \end{Soutput}
@@
        expression    min median mem_alloc n_itr n_gc total_time
-1      dplyr_mean  1.16s  1.17s   48.57MB     3   78      3.52s
-2 data.table_mean 7.43ms 8.31ms   18.93MB   283   28         3s
-3   collapse_mean  3.1ms 3.71ms    9.11MB   679   28         3s
+1      dplyr_mean  1.28s  1.28s   48.56MB     1   23      1.28s
+2 data.table_mean  8.9ms 9.05ms   18.93MB     2    0    18.11ms
+3   collapse_mean 4.08ms 4.39ms    9.11MB     3    0    13.02ms
 \end{Soutput}
@@
          expression     min  median mem_alloc n_itr n_gc total_time
-1      dplyr_median    4.2s    4.2s    52.8MB     1   91       4.2s
-2 data.table_median 21.32ms 22.24ms    18.9MB   130   13         3s
-3   collapse_median  9.34ms  9.96ms    11.1MB   271   12         3s
+1      dplyr_median   4.51s   4.51s    52.8MB     1   90      4.51s
+2 data.table_median 23.72ms 23.72ms    18.9MB     1    0    23.72ms
+3   collapse_median 11.54ms 11.54ms    11.1MB     1    0    11.54ms
 \end{Soutput}
@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1      dplyr    93ms 103.73ms   20.18MB    27   56      3.03s
-2 data.table 56.51ms  60.06ms    5.77MB    50   25      3.06s
-3   collapse  5.04ms   5.51ms     6.8MB   509   13         3s
+1      dplyr 112.36ms 112.36ms   20.18MB     1    3    112.4ms
+2 data.table  63.78ms  63.78ms    5.77MB     1    1     63.8ms
+3   collapse   6.46ms   6.54ms     6.8MB     2    0     13.1ms
 \end{Soutput}
@@
         expression     min  median mem_alloc n_itr n_gc total_time
-1      dplyr_joins 139.95ms   153ms   559.4MB    19   32      3.12s
-2 data.table_joins  97.39ms   129ms     491MB    19   39      3.11s
-3   collapse_joins   8.85ms  11.9ms    89.7MB   188   51      3.01s
+1      dplyr_joins 172.7ms 172.7ms   558.9MB     1    2    172.7ms
+2 data.table_joins 204.5ms 204.5ms   490.4MB     1    1    204.5ms
+3   collapse_joins  59.1ms  59.1ms    89.7MB     1    2     59.1ms
 \end{Soutput}
@@
   expression     min  median mem_alloc n_itr n_gc total_time
-1      tidyr 56.9ms 58.8ms     254MB     8   39      472ms
-2 data.table   40ms 42.6ms     209MB    22   32      962ms
-3   collapse   13ms   15ms     209MB    47   84      713ms
+1      tidyr  82.7ms  82.7ms     251MB     1    2     82.7ms
+2 data.table 143.9ms 143.9ms     209MB     1    2    143.9ms
+3   collapse  17.7ms  17.7ms     209MB     1    0     17.7ms
 \end{Soutput}
@@
      expression      min   median mem_alloc n_itr n_gc total_time
-1         tidyr 379.96ms 397.91ms     143MB     8   59      3.27s
-2    data.table    227ms  231.1ms    21.7MB    13   47      3.03s
-3 collapse_fsum   5.61ms   7.18ms    39.1MB   327   26      3.01s
-4 collapse_itnl   3.77ms   4.45ms    12.4MB   611   15         3s
+1         tidyr 376.32ms 376.32ms   142.4MB     1    5    376.3ms
+2    data.table 237.61ms 237.61ms      21MB     1    4    237.6ms
+3 collapse_fsum    7.6ms   7.94ms    39.1MB     2    0     15.9ms
+4 collapse_itnl   5.05ms   5.08ms    12.4MB     2    0     10.2ms
 \end{Soutput}

### 2025-05-29 misc/JSS article/final/article.Rnw
Final article update.

@@
              sub-set-ting de-com-po-si-tion al-go-rith-mic en-hanced mul-ti-ple coun-tries
-             com-pu-ta-tion-al par-tic-u-lar-ly}
+             com-pu-ta-tion-al par-tic-u-lar-ly com-po-nents}
 \usepackage[english]{babel}
@@
 bmark <- function(...) {
-  bench::mark(..., min_time = 0.01, check = FALSE) |> janitor::clean_names() |>
+  bench::mark(..., min_time = 3, check = FALSE) |> janitor::clean_names() |>
   fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
@@
 Core functionality includes a rich set of S3 generic (grouped, weighted) statistical functions for vectors, matrices, and data frames, which provide efficient low-level vectorizations, OpenMP multithreading, and skip missing values by default (\code{na.rm = TRUE}).
-It also provides powerful data manipulation functions, such as vectorized and verbose hash-joins or fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools.
+It also provides powerful data manipulation functions, including vectorized and verbose hash-joins and fast (aggregation, recast) pivots, functions and classes for indexed (time-aware) computations on time series and panel data, recursive tools to deal with nested data, and advanced descriptive statistical tools.
 This functionality is powered by efficient algorithms for grouping, ordering, deduplication, and matching callable at both \proglang{R} and \proglang{C} levels.
@@
 \pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
-It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications to the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
+It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications of the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
 See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline
@@
 Its low-level vectorizations and workhorse algorithms are accessible at the \proglang{R} and \proglang{C}-levels, unlike \pkg{data.table}, where most vectorizations and algorithms are internal.
-It also supports variable labels and intelligently preserves attributes of all objects, complementing \pkg{labelled}.
+It also supports variable labels and intelligently preserves the attributes of all objects, complementing \pkg{labelled}.
 It provides novel recursive tools to deal with nested data, enhancing \pkg{tidyr}, \pkg{purrr}, and \pkg{rrapply}.
@@
 
-The package has a built-in structured \href{https://sebkrantz.github.io/collapse/reference/index.html}{documentation} facilitating its use.
-This documentation includes a central \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and supplementary topic pages which briefly describe related functionality.
+The package comes with a built-in structured \href{https://sebkrantz.github.io/collapse/reference/index.html}{documentation} facilitating its use.
+This includes a central \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and a set of supplementary topic pages which briefly summarize related functionality.
 The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:
@@
 %
-This article does not fully present \pkg{collapse}, but the following sections introduce its key components, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options.
+While this article cannot fully present \pkg{collapse}, the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options.
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} $+$ \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
+For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
 %
@@
 fmean(iris$Sepal.Length, g = iris$Species)
-fmean(num_vars(iris), g = iris$Species, nthreads = 4)
+fmean(num_vars(iris), g = iris$Species)
 @
@@
 Having grouping and data transformation functionality directly built into generic statistical functions facilitates and speeds up many common operations.
-It also avoids the need to convert atomic objects to data frames for grouped aggregations or transformations.
+It notably avoids the need to convert atomic objects to data frames for grouped aggregations or transformations.
 The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://sebkrantz.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
@@
 Using subtraction by reference in the range is more memory efficient as the executed expression is equivalent to \code{fmax(LIFFEX, country) \%-=\% fmin(LIFFEX, country)}.
-\fct{across} here directly invokes \fct{flast.data.frame} on the subset of columns.
-%
-\emph{Fast Statistical Functions} also have a method for grouped data, hence \fct{fsummarise} is not always needed.
-Below, I average the data with population weights.\footnote{Functions like \fct{fmean}, when called on a grouped data frame with weights, by default (\code{keep.w = TRUE}) sum the weights column and retain it after the grouping columns to permit further weighted operations.
+\fct{across} directly invokes \fct{flast.data.frame} on the subset of columns.
+\emph{Fast Statistical Functions} also have a method for grouped data, thus \fct{fsummarise} is not always needed.
+Below, I average recent data with population weights.\footnote{Functions like \fct{fmean}, when called on a grouped data frame with weights, by default (\code{keep.w = TRUE}) sum the weights column and retain it after the grouping columns to permit further weighted operations.
 \vspace{-10mm}}
@@
 This may be confusing at first, but all options can be distinguished in terms of efficiency.
-It is thus helpful to be mindful of computer resources and attempt to craft minimalistic solutions to complex problems.
+Hence, it is helpful to think about computer resources and attempt to craft minimalistic solutions to complex problems.
 For example, I recently combined multiple spatial datasets on points of interest (POIs) and needed to deduplicate them.
@@
 After creating comparable POI confidence, location, and type indicators, my deduplication expression ended up being a single line of the form \code{fsubset(data, source == fmode(source, list(location, type), confidence, "fill"))}---which retains POIs from the confidence-weighted most frequent (richest) source by location and type.
-This is efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}---\emph{Fast Statistical Functions} internal grouped logic is highly optimized.
+This is efficient because it avoids materializing intermediate datasets and relegates all computations to \fct{fmode}---\emph{Fast Statistical Functions} call a highly optimized internal grouping algorithm.
 %
@@
 Similarly, \fct{fmutate} sets \code{g} and \code{TRA = "fill"}, which can be overwritten by the user (here with \code{TRA = "-"}).
-The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)} amounts to \code{cov(x, y)/var(y)}, but is vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) on the fly and division by reference (\code{\%/=\%}) to avoid an additional allocation.
+The expression \code{fsum(x, dmy) \%/=\% fsum(dmy, dmy)} amounts to \code{cov(x, y)/var(y)}, but is vectorized across groups and memory efficient---leveraging the weights (\code{w}) argument to \fct{fsum} to compute products (\code{v * dmy} and \code{dmy * dmy}) internally and division by reference (\code{\%/=\%}) to avoid an additional allocation.
 In a 2023 \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{blog post}, I forecasted high-resolution population estimates for South Africa like this.
@@
 %
-Weighted quantiles have a sub-column level parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or the \code{nthreads} arguments to \fct{fnth}/\fct{fmedian}/\fct{fmode} (default 1).} and, as shown above, can harness an (optional) optimization utilizing an overall ordering vector (combining groups and the data column) to avoid repeated sorting of the same elements within groups.
+Weighted quantiles have a sub-column level parallel implementation,\footnote{Use \code{set\_collapse(nthreads = \#)} or set the \code{nthreads} arguments of \fct{fnth}/\fct{fmedian} (default 1).} and, as shown above, can also harness an (optional) optimization via an overall ordering vector---combining groups with the data column to avoid repeated sorting of the same elements by different functions.
 % \newline
@@
 The main reason for not providing higher-level metaprogramming capabilities is to keep all functions as simple as possible.
-It also compels users to think deeply about their programs and devise more efficient solutions.
-\newline
-
-These metaprogramming restrictions are vindicated by internally complex data manipulation functions producing greater overheads that can noticeably add up in larger scripts:\footnote{\fct{bmark} is a slim wrapper around \fct{bench::mark}.
+It also compels users to think more deeply about their programs and devise more efficient solutions.
+\newline
+
+As the small benchmark below illustrates, \pkg{dplyr}'s internally more complex data manipulation functions produce much greater overheads, which can noticeably add-up in longer scripts.\footnote{\fct{bmark} is a slim wrapper around \fct{bench::mark}.
 See the Computational details section.}
+%
 <<fsumm_bench>>=
- bmark(collapse = collapse::fsummarise(mtcars, sum = sum(mpg)),
-       dplyr = dplyr::summarise(mtcars, sum = sum(mpg)))
-@
+bmark(collapse = collapse::fsummarise(wlddev, mean = fmean(PCGDP)),
+      dplyr = dplyr::summarise(wlddev, mean = fmean(PCGDP)))
+@
+%
 Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices.
-For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 44.7 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
+For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions like \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 45 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
 %On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another recent example involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}.
@@
 \end{table}
-%
 \subsection{Ad-hoc computations}
@@
 Time series functions such as \fct{fgrowth} to compute growth rates are S3 generic and can be applied to most time series classes.
-In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation.
-If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a plain numeric vector representing calendar years.
-Users need to manually call \code{timeid()} on plain numeric vectors with decimals to yield an appropriate integer representation.
-If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id.
-For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}.} But first, a basic example:
+In addition to a \code{g} argument for grouping, these functions also have a \code{t} argument for indexation.
+But first, I provide a basic example of computing the annualized 10-year growth rates in miles flown by airlines in the USA from 1937-1960.
 %
 <<ts_example>>=
-fgrowth(airmiles) |> round(2)
-@
+fgrowth(airmiles, n = 10, power = 1/10) |> na.omit() |> round(1)
+@
+%
+The results show that the flight volume as been growing steadily but at a decreasing rate. \newline
+
 To illustrate the full capabilities of these time series functions, I generate a sector-level trade dataset of export values (v) by country (c), sector (s), and year (y).
@@
 The following extracts one country-sector series from the \code{exports} dataset.
-It is irregular, missing years 2003 and 2006.\footnote{\code{\%=\%} is an infix operator for the \fct{massign} function in \pkg{collapse} which is a multivariate version of \fct{assign}.} Indexation using the \code{t} argument then allows for correct (time-aware) computations in this context without the need to 'expand' the data/fill gaps.
+It is irregular, missing years 2003 and 2006.\footnote{\code{\%=\%} is an infix operator for the \fct{massign} function in \pkg{collapse} which is a multivariate version of \fct{assign}.} Indexation using the \code{t} argument still allows for correct (time-aware) computations in this context without the need to 'expand' the data/fill gaps.
 %
@@
 %
+If \code{t} is a plain numeric vector or factor as in this case, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a plain numeric vector representing calendar years.
+Users should manually call \code{timeid()} on plain vectors with decimals.} Time objects like \class{Date} or \class{POSIXct} on the other hand are internally passed through \fct{timeid} to generate an appropriate integer representation of them.\footnote{\fct{timeid} divides by the greatest common divisor (GCD) and subtracts the minimum to generate an integer-id (starting form 1).
+For this approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}.} \newline
+
 Functions \code{flag()/fdiff()/fgrowth()} also have associated 'operators' \code{L()/D()/G()} to facilitate their use inside formulas and provide an enhanced data frame interface for convenient ad-hoc computations.
@@
 %
-These functions and operators are also integrated with \fct{fgroup\_by} and \fct{fmutate} for vectorized computations. But, as mentioned earlier, ad-hoc grouping is always more efficient.
+These functions and operators are also integrated with \fct{fgroup\_by} and \fct{fmutate} for vectorized computations. As mentioned earlier, ad-hoc grouping is always more efficient.
 %
@@
 The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}.
-Their basic syntax is:
+Their basic usage is:
 %
@@
 Indexation can be undone using \code{unindex()} and redone with \code{reindex()} and a suitable \class{index\_df}.
-\class{indexes\_series} can be atomic vectors or matrices (including objects such as \class{ts} or \class{xts}) and can be created directly using \code{reindex()}.
+\class{indexes\_series} can be atomic vectors or matrices (including \class{ts} or \class{xts}) and can also be created directly using \code{reindex()}.
 %
@@
 It is worth highlighting that the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions.\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.} Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures or native handling of irregularity in basic operations.
-\class{indexed\_series} and \class{indexed\_frame}, on the other hand, work 'anywhere', and can be superimposed on any suitable object, as long as \pkg{collapse}'s functions (\code{flag()/L()}, etc.) are used to perform the time-based computations.
-\newline
-
-An example using the \code{exports} data follows. \emph{Note} that data need not be sorted for indexing.
+\class{indexed\_series} and \class{indexed\_frame}, on the other hand, work 'anywhere', and can be superimposed on any suitable object as long as \pkg{collapse}'s functions (\code{flag()/L()}, etc.) are used to perform the time-based computations.
+\newline
+
+An example follows using the \code{exports} data. \emph{Note} that data can be unsorted for indexation.
 <<indexing>>=
 exportsi <- exports |> findex_by(c, s, y)
-exportsi |> G() |> head(5)
+exportsi |> G(0:1) |> head(5)
 exportsi |> findex() |> print(2)
@@
 @
-\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
+\fct{psmat}, for panel-series to matrix, generates a matrix or array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
 <<indexing_3>>=
@@
 @
-Indexed series/frames also support transformation such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}.
+Indexed series/frames also support transformations such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}.
 Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series.
-The \class{index\_df} attached to these objects can also be used with other general tools such as \fct{collapse::BY} to perform grouped computations using 3rd-party functions.
-For example, below I compute a 5-year rolling average (\fct{ix} abbreviates \fct{findex}).
+The \class{index\_df} attached to these objects can also be used with other general tools such as \fct{collapse::BY} to perform grouped computations using third-party functions: %For example, below I compute a 5-year rolling average ().
 %
 <<rollmean>>=
-BY(vi, ix(vi)$c.s, data.table::frollmean, 5) |> head(10)
-@
-% Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
+BY(vi, findex(vi)$c.s, data.table::frollmean, 5) |> head(10)
+@
+%
+Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
 %
@@
 %
-Among all \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots.
-Both implementations provide outstanding computational performance, syntax, and memory efficiency.
+Among its suite of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots.
+Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
 %
@@
 \end{Code}
-It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}), thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases.
+It defaults to left join and only takes first matches from \code{y} (\code{multiple = FALSE}).
+Thus, it simply adds columns to \code{x}, which is efficient and sufficient/desired in many cases.
 By default (\code{verbose = 1}), it prints information about the join operation and number of records joined.
@@
 %
-Users can request the generation of a \code{.join} column (\code{column = "name"/TRUE}), akin to \proglang{STATA}'s \code{\_merge} column, to indicate the origin of records in the joined table---useful with a full join:
+Users can request the generation of a \code{.join} column (\code{column = "name"/TRUE}) akin to \proglang{STATA}'s \code{\_merge} column to indicate the origin of records in the joined table---useful with a full join:
 %
@@
 @
-Users can also invoke the \code{validate} argument to examine the uniqueness of the join keys in either table: passing a '1' produces an error if the key is not unique.
+Users can also invoke the \code{validate} argument to examine the uniqueness of the join keys in either table: passing a '1' produces an error if the respective key is not unique.
 %
@@
 First, \fct{join} is class-agnostic and preserves the attributes of \code{x} (any list-based object).
-It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"}, or \code{"anti"} join).
-This demonstrates the latter:
+It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"}, or \code{"anti"}).
+This demonstrates the latter two:
 <<joins_5>>=
@@
 By default (\code{sort = FALSE}), the order of rows in \code{x} is preserved.
-Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join (using \fct{fmatch}) followed by sorting, particularly if the data is already sorted on the keys.
+Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join using \fct{fmatch} followed by sorting, particularly if the data is already sorted on the keys.
 } The join relationship is indicated inside the \code{<>} as the number of records joined from each table divided by the number of unique matches.
 % In multi-match settings, this will be reflected by few records from \code{y} being used.
+\newpage
+\fct{join}'s handling of duplicate columns in both tables is also rather special.
+By default (\code{suffix = NULL}), \fct{join} extracts the name of the \code{y} table and appends \code{y}-columns with it.
+% \code{x}-columns are not renamed.
+%
 <<joins_6>>=
@@
 @
-As shown above, \fct{join}'s handling of duplicate columns in both tables is rather special.
-By default (\code{suffix = NULL}), \fct{join} extracts the name of the \code{y} table and appends \code{y}-columns with it.
-\code{x}-columns are not renamed.
-This is congruent to the principle of adding columns to \code{x} and altering this table as little as possible.
+%
+This is congruent to the principle of adding columns to \code{x} in the default first-match left join by altering this table as little as possible.
 Alternatively, option \code{drop.dup.cols = "x"/"y"} can be used to remove duplicate columns from either \code{x} or \code{y} before the join operation.
@@
 A final noteworthy feature is that \fct{fmatch} has a built-in overidentification check, which warns if more key columns than necessary to identify the records are provided.
-This check only triggers with 3+ id columns as for efficiency reasons the first two ids are hashed jointly.
+This check only triggers with 3+ id columns as for efficiency reasons the first two ids are jointly hashed.
 
@@
 It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information.
-Its basic syntax is:
+Its basic syntax is: \newpage
 %
@@
 The second option is a wider pivot with \code{how = "wider"}.
-Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{If multiple columns are selected, they are combined using \code{"\_"} for names and \code{" - "} for labels.} Note below how the labels are combined with existing labels such that also this operation is without loss of information.
-It is, however, a destructive operation---with 2 or more columns selected through \code{values}, \fct{pivot} is not able to reverse it.
-Further arguments like \code{na.rm}, \code{fill}, \code{drop}, \code{sort}, and \code{transpose} can be used to control the casting process/output.
+Here, \code{names} and \code{labels} can be used to select columns containing the names of new columns and their labels.\footnote{If multiple columns are selected, they are combined using \code{"\_"} for names and \code{" - "} for labels.
+\vspace{-5mm}} Note below how the labels are combined with existing labels such that also this operation is without loss of information.
+It is, however, a destructive operation, as with two or more columns selected through \code{values}, \fct{pivot} is not able to reverse it.
+Further arguments like \code{na.rm}, \code{fill}, \code{drop}, \code{sort}, and \code{transpose} can be invoked to control the casting process/output.
 %
@@
 For the recast pivot (\code{how = "recast"}), unless a column named \code{variable} exists in the data, the source and (optionally) destination of variable names needs to be specified using a list passed to \code{names}, and similarly for \code{labels}.
-Again, taking along labels is entirely optional---omitting either the labels-list's \code{from} or \code{to} element will omit the respective operation.
+Again, taking along labels is entirely optional---omitting either the labels-list's \code{from} or \code{to} elements will omit the respective operation.
 %
@@
 The latter can be checked with \code{check.dups = TRUE}.
-A small number extremely fast internal aggregation functions, \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"} operate 'on the fly' while reshaping.
-\fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but require a deep copy of the columns aggregated which is avoided by the internal functions.
+A small number extremely fast internal aggregation functions: \code{"first", "last", "sum", "mean", "min", "max",} and \code{"count"}, operate 'on the fly' during reshaping.
+\fct{pivot} also supports \emph{Fast Statistical Functions}, which will yield vectorized aggregations, but requires a deep copy of the columns aggregated which is avoided by the internal functions.
 The following example performs aggregation across years with the internal mean function during a recast pivot.
@@
 Often in programming, nested structures are needed.
-A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects (such as a model predictions and performance statistics) in a list.
+A typical use case involves running statistical procedures for multiple configurations of variables and parameters and saving multiple objects, such as a model predictions and performance statistics, in a list.
 Nested data is also often the result of web scraping or web APIs.
@@
 To create nested data, \fct{rsplit} generalizes \fct{split} and (recursively) splits up data frame-like objects into (nested) lists.
-For example, we can split the \code{GGDC10S} data by country and variable, such that, for example, agricultural employment in Argentina can be accessed as:\footnote{If a nested structure is not needed, \code{flatten = TRUE} lets \fct{rsplit} operate like a faster version of \fct{split}.}
+For example, we can split the \code{GGDC10S} data by country and variable, such that agricultural employment in Argentina can be accessed as:\footnote{If a nested structure is not needed, \code{flatten = TRUE} lets \fct{rsplit} operate like a faster version of \fct{split}.} \newpage
 %
 <<list_proc>>=
-dl <- GGDC10S |> rsplit( ~ Country + Variable)
-dl$ARG$EMP$AGR[1:12]
-@
-%
-This is a convenient data representation for \emph{Shiny Apps} where we can let the user choose data (e.g., \code{dl[[input$country]][[input$variable]][[input$sector]]}) without expensive subsetting operations.
+d_list <- GGDC10S |> rsplit( ~ Country + Variable)
+d_list$ARG$EMP$AGR[1:12]
+@
+%
+This is a convenient data representation for \emph{Shiny Apps} where we can let the user choose data (e.g., \code{d_list[[input$country]][[input$variable]][[input$sector]]}) without expensive subsetting operations.
 As mentioned, such data representation can also be the result of an API call parsing JSON or a nested loop or \fct{lapply} call.
@@
 <<list_proc_2>>=
-result <- list()
+results <- list()
 for (country in c("ARG", "BRA", "CHL")) {
@@
     m <- lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year,
-            data = dl[[country]][[variable]])
-    result[[country]][[variable]] <- list(model = m, BIC = BIC(m),
+            data = d_list[[country]][[variable]])
+    results[[country]][[variable]] <- list(model = m, BIC = BIC(m),
                                            summary = summary(m))
@@
 @
-This programming may not be ideal for this particular use case as I could have used data frame-based tools and saved the result in a column.\footnote{E.g., \code{GGDC10S |> fgroup\_by(Country, Variable) |> fsummarise(result = my\_fun(lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year)))} with \code{my\_fun <- function(m) list(list(m, BIC(m), summary(m)))}.}
+This programming may not be ideal for this particular use case as I could have used data frame-based tools and saved the result in a column.\footnote{E.g., \code{GGDC10S |> fgroup\_by(Country, Variable) |> fsummarise(results = my\_fun(lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year)))} with \code{my\_fun <- function(m) list(list(m, BIC(m), summary(m)))}.}
 However, there are limits to such workflows.
-For example, I recently trained a complex ML model for different variables and parameters, which involved loading a different dataset at each iteration.
+For example, I recently trained a complex ML model for different variables and parameters, while also loading a different dataset for each combination.
 Loops are useful in such cases, and lists a natural vehicle to structure complex outputs.
 The main issue with nested lists is that they are complex to query.
-What if we want to know just the $R^2$ of these 6 models?
-We would need to use, e.g., \code{result$ARG$EMP$summary$r.squared} for each model.\newline
+What if we wanted to know the $R^2$ of these 6 models?
+We would need to use, e.g., \code{results$ARG$EMP$summary$r.squared} for each model.\newline
 
@@
 <<list_proc_3>>=
-str(r_sq_l <- result |> get_elem("r.squared"))
-rowbind(r_sq_l, idcol = "Country", return = "data.frame")
+str(r_sq <- results |> get_elem("r.squared"))
+rowbind(r_sq, idcol = "Country", return = "data.frame")
 @
@@
 % Thus \code{get\_elem("R2")} returns the same nested list with the $R^2$ in all final nodes. Then \code{rowbind()} could be used to create a data frame:
-Note how the \code{"summary"} branch was eliminated since it is common to all final nodes; \code{result |> get\_elem("r.squared", keep.tree = TRUE)} could have been used to keep it.
+Note how the \code{"summary"} branch was eliminated since it is common to all final nodes; \code{results |> get\_elem("r.squared", keep.tree = TRUE)} could have been used to keep it.
 \fct{rowbind} then efficiently combines lists of lists.
@@
 <<list_proc_3.5>>=
-r_sq_l |> t_list() |> rowbind(idcol = "Variable", return = "data.frame")
-@
-\fct{rowbind} is limited if \fct{get\_elem} returns a more nested or asymmetric list, potentially with vectors/arrays in the final nodes. Suppose we want to extract the coefficient matrices:
+r_sq |> t_list() |> rowbind(idcol = "Variable", return = "data.frame")
+@
+\fct{rowbind} is limited if \fct{get\_elem} returns a more nested or asymmetric list, potentially with vectors/arrays in the final nodes. Suppose we wanted to extract the coefficient matrices:
 <<list_proc_4>>=
-result$ARG$EMP$summary$coefficients
+results$ARG$EMP$summary$coefficients
 @
@@
 <<list_proc_5>>=
-result |> get_elem("coefficients") |> get_elem(is.matrix) |>
+results |> get_elem("coefficients") |> get_elem(is.matrix) |>
   unlist2d(idcols = c("Country", "Variable"),
@@
 \pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets.
-A particular focus has been on providing tools for examining longitudinal (panel) data.
+A particular focus has been on providing tools to understand multilevel (panel) data.
 Recall the World Development panel dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{sec:integration}.
-The function \fct{varying} can be used to examine which of these variables are time-varying.
+The function \fct{varying} can be used to examine which of the variables are time-varying.
 %
@@
 A related exercise is to decompose the variance of a panel series into variation between countries and variation within countries over time.
-Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (see Table~\ref{tab:TSfun}), this is easily demonstrated:
+Using the (de-)meaning functions \fct{fbetween}/ \fct{fwithin} supporting \class{indexed\_series} (Table~\ref{tab:TSfun}), this is easily demonstrated:
 %
@@
 It can also be computed for different subgroups and with sampling weights.
-\fct{qsu} also has a data frame method, and by default computes simple statistics:\footnote{The \code{pid} argument to \fct{qsu} can also be used to pass identifiers for panel-decomposition, e.g., \code{pid = ~ iso3c}.
-With indexed data, it is automatically set to the first column in the index (\code{effect = 1}).}
+\fct{qsu} also has a data frame method, and by default computes simple statistics.\footnote{The \code{pid} argument to \fct{qsu} can also be used to manually pass identifiers for panel-decomposition, e.g., \code{pid = iso3c}.
+With indexed data, it is automatically set to the first column in the index (\code{effect = 1}).} Below, I take the latest post-2015 estimates and summarise LIFEEX by income groups with population weights.
+The \code{WeightSum} column thus records the total population in each group.
 %
@@
 For greater detail, \fct{descr} provides a rich (grouped, weighted) statistical description.
-It does not support panel-variance decompositions like \fct{qsu}, but also computes detailed (grouped, weighted) frequency tables for categorical data and is thus very useful with mixed-type data.
+It does not support panel-variance decompositions like \fct{qsu}, but also computes detailed frequency tables for categorical data.
+Below, I summarize income and LIFEEX by OECD membership, scaling the weights to limit printout and replacing missing weights with \code{0} (the default).\footnote{This is necessary in \fct{descr} because \fct{fquantile} does not support missing weights for non-missing \code{x}.\vspace{-5mm}}
 %
 <<descr>>=
-descr(wlda15, by = income + LIFEEX ~ OECD)
-@
-
-%
- It also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden.
+descr(wlda15, by = income + LIFEEX ~ OECD, w = ~ replace_na(POP / 1e6))
+@
+
+%
+\fct{descr} also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden.
 The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples.
@@
 A final noteworthy function from \pkg{collapse}'s descriptive statistics toolkit is \fct{qtab}, an enhanced drop-in replacement for \fct{table}.
-It is enhanced both in a statistical and a computational sense, providing a remarkable performance boost, an option \code{sort = FALSE} to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights---vectorized when using \emph{Fast Statistical Functions}, as demonstrated below.
+It is enhanced both in a statistical and a computational sense, providing a remarkable performance boost, an option \code{sort = FALSE} to preserve the first-appearance-order of vectors being cross-tabulated, support for frequency weights (\code{w}), and the ability to compute different statistics representing table entries using these weights---vectorized when using \emph{Fast Statistical Functions} as demonstrated below.
 %
@@
 <<qtab_2>>=
-wlda15 |> with(qtab(OECD, income, w = POP) / 1e6)
+wlda15 |> with(qtab(OECD, income, w = POP / 1e6))
 @
@@
 \class{qtab} objects inherit the \class{table} class, thus all \class{table} methods apply.
-Apart from the above functions, \pkg{collapse} also provides functions \fct{pwcor}, \fct{pwcov}, and \fct{pwnobs} for convenient (pairwise, weighted) correlations, covariances, and observations counts, respectively.
+Apart from the above summary functions, \pkg{collapse} also provides \fct{pwcor}, \fct{pwcov}, and \fct{pwnobs} for convenient (pairwise, weighted) correlations, covariances, and observations counts, respectively.
 %
@@
 
-As evident from previous sections, \pkg{collapse} provides performance-improved or otherwise enhanced versions of functionality already present in base \proglang{R} (like the \emph{Fast Statistical Functions}, \fct{funique}, \fct{fmatch}, \fct{fsubset}, \fct{ftransform}, etc.) and other packages (esp.
-\pkg{dplyr} \citep{rdplyr}: \fct{fselect}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, etc.).
-The objective of being namespace compatible warrants such a naming convention, but this has a syntactical cost, particularly when \pkg{collapse} is the primary data manipulation package.
+As evident from previous sections, \pkg{collapse} provides performance-improved or otherwise enhanced versions of functions already present in base \proglang{R} (like the \emph{Fast Statistical Functions}, \fct{funique}, \fct{fmatch}, \fct{fsubset}, \fct{ftransform}, etc.) or other packages (especially \pkg{dplyr} \citep{rdplyr}: \fct{fselect}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, etc.).
+The objective of being namespace compatible warrants such a naming convention, but this has a syntactical cost, particularly when \pkg{collapse} is used as the primary workhorse package.
 \newline
@@
 To reduce this cost, \pkg{collapse}'s \code{mask} option allows masking existing \proglang{R} functions with the faster \pkg{collapse} versions by creating additional functions in the namespace and instantly exporting them.
-All \pkg{collapse} functions starting with 'f' can be passed to the option (with or without the 'f'), e.g., \code{set\_collapse(mask = c("subset", "transform"))} creates \code{subset <- fsubset} and \code{transform <- ftransform} and exports them.
-Special functions are \code{"n"}, \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} (for use in \code{(f)summarise}/\code{(f)mutate}), \code{table <- qtab}, and replace \code{\%in\%} with a fast version using \fct{fmatch}, respectively.
-There are also several \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}.
-The most powerful of these is \code{"all"}, which masks all f-functions and special functions, as shown below.
+All \pkg{collapse} functions starting with an 'f' can be passed to the option (with or without the 'f'), e.g., \code{set\_collapse(mask = c("subset", "transform"))} creates \code{subset <- fsubset} and \code{transform <- ftransform} and exports them.
+Special functions are \code{"n"}, \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} for use in \code{(f)summarise}/\code{(f)mutate}, \code{table <- qtab}, and replace \code{\%in\%} with a faster version based on \fct{fmatch}, respectively.
+There also exist several \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}, such as \code{"manip"} (only data manipulation functions), or \code{"all"} (everything), as demonstrated below.
 %
@@
 They are executed on a 2024 Apple MacBook Pro with 48GB M4 Pro chip.
-It also discusses results from \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{3rd party benchmarks} involving \pkg{collapse}.
+It also discusses results from \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{third-party benchmarks} involving \pkg{collapse}.
 The first set of benchmarks show that \pkg{collapse} provides faster computationally intensive operations like unique values and matching on large integer and character vectors.
@@
 g_char <- sample(char <- outer(char, char, paste0), 1e7, TRUE)
-bmark(base_int = unique(g_int), collapse_int = funique(g_int))
-bmark(base_char = unique(g_char), collapse_char = funique(g_char))
-bmark(base_int = match(g_int, int), collapse_int = fmatch(g_int, int))
-bmark(base_char = match(g_char, char), data.table_char =
-      chmatch(g_char, char), collapse_char = fmatch(g_char, char))
-@
-%
-The second set below shows that \pkg{collapse}'s statistical functions are very efficient on aggregating a numeric matrix with 10,000 rows and 1000 columns.
+bmark(base = unique(g_int), collapse = funique(g_int))
+bmark(base = unique(g_char), collapse = funique(g_char))
+bmark(base = match(g_int, int), collapse = fmatch(g_int, int))
+bmark(base = match(g_char, char), data.table =
+      chmatch(g_char, char), collapse = fmatch(g_char, char))
+@
+%
+The second set below shows that \pkg{collapse}'s statistical functions are very efficient in aggregating a numeric matrix with 10,000 rows and 1000 columns.
 They are faster than base \proglang{R} even without multithreading, but using 4 threads in this case induces a sizeable difference.
@@
 m <- matrix(rnorm(1e7), ncol = 1000)
-bmark(R = colSums(m), collapse = fsum(m))
-bmark(R = colMeans(m), collapse = fmean(m))
-bmark(MS = matrixStats::colMedians(m), collapse = fmedian(m))
-@
-%
-Below, I also show a grouped version summing the columns within 1000 random groups.
+bmark(base = colSums(m), collapse = fsum(m))
+bmark(base = colMeans(m), collapse = fmean(m))
+bmark(matrixStats = matrixStats::colMedians(m), collapse = fmedian(m))
+@
+%
+Below, I also benchmark a grouped version summing the columns within 1000 random groups.
 %
@@
 g <- sample.int(1e3, 1e4, TRUE)
-bmark(R = rowsum(m, g), collapse = fsum(m, g))
-@
-%
-I now turn to basic operations on a medium sized real-world database recording all flights from New York City (EWR, JFK, and LGA) in 2023---provided by the \pkg{nycflights23} package.
+bmark(base = rowsum(m, g), collapse = fsum(m, g))
+@
+%
+I now turn to basic operations on a medium-sized real-world database recording all flights from New York City (EWR, JFK, and LGA) in 2023---provided by the \pkg{nycflights23} package.
 The \code{flights} table has 435k flights, and grouping it by day and route yields 76k unique trips.
@@
 The verbosity of \fct{join} is essential to understanding what has happened here---how many records from each table were matched and which duplicate non-id columns were suffixed with the (default) \code{y}-table name.
-Normally, I would set \code{drop.dup.cols = "y"} as it seems not useful to preserve them here, but the other packages don't have this option.
+Usually, I would set \code{drop.dup.cols = "y"} as keeping them is not helpful in this case, but the other packages don't have this option.
 For the benchmark, I set \code{verbose = 0} in \pkg{collapse} and employ the fastest syntax for \pkg{dplyr} and \pkg{data.table}:\footnote{\code{left\_join(..., multiple = "first")} for \pkg{dplyr} and \code{y[x, on = ids, mult = "first"]} for \pkg{data.table}.}
@@
 Evidently, the vectorized hash join provided by \pkg{collapse} is 10x faster than \pkg{data.table} on this database, at a substantially lower memory footprint.
-It remains competitive on \href{https://duckdblabs.github.io/db-benchmark/}{big data}.\footnote{\pkg{data.table} joins utilize multithreaded radix-ordering---a very different logic more useful for big data.}
+It remains competitive on \href{https://duckdblabs.github.io/db-benchmark/}{big data}.\footnote{\pkg{data.table} joins utilize multithreaded radix-ordering---a very different logic more useful for big data.\vspace{-5mm}} \newline
 
@@
 %
-Again, \pkg{collapse} is fastest, as it offers full vectorization, either via \fct{fsum}, which translates to \code{fsum(x, g, TRA = "fill")} before pivoting and thus entails a full deep copy of the \code{vars} columns, or via an optimized internal sum function which sums values 'on the fly' during the reshaping process.
+Again, \pkg{collapse} is fastest, as it offers full vectorization, either via \fct{fsum}, which translates to \code{fsum(x, g, TRA = "fill")} before pivoting and thus entails a full deep copy of the \code{vars} columns, or via the optimized internal sum function which sums values 'on the fly' during the reshaping process.
 \pkg{data.table} is not vectorized here but at least memory efficient.
@@
 These four are supported by (semi-)commercial entities, leaving \pkg{data.table} as the only fully community-led project, and \pkg{collapse} as the only project that is single-authored and without financial support.
-The benchmarks show that \pkg{collapse} achieves the highest relative performance on 'smaller' datasets (10-100 million rows) and performing advanced operations.
-% \newline
+The benchmarks show that \pkg{collapse} achieves the highest relative performance on 'smaller' (10-100 million row) datasets and performing advanced operations.
+\newline
 
@@
 I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.github.io/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages.
-These benchmarks align that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and on PCs/Macs.
+These benchmarks agree that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and on users PCs/Macs---which typically have smaller memory and parallel computing resources but faster chips than servers.
 % \footnote{Reasons for the particularly strong performance of \pkg{collapse} on Windows may be that it is largely written in C and has limited multithreading in favor or highly efficient serial algorithms---there appear to be persistent obstacles to efficient (low-overhead) multithreading on Windows, implying that multithreaded query engines do not develop their full thrust on medium-sized ($\leq$100 million row) datasets.}
@@
 It has become a new foundation package for statistical computing and data transformation in \proglang{R}---one that is statistically advanced, class-agnostic, flexible, fast, lightweight, stable, and able to manipulate complex scientific data with ease.
-As such, it opens up new possibilities for statistics, data manipulation, and package development in \proglang{R}.
+As such, it opens up new possibilities for statistics, research, production, and package development in \proglang{R}.
 \newline % It also sets an ambitious benchmark for high-quality software engineering on the language.
@@
 
-This article provided a quick guide to the package, articulating its key ideas and design principles and demonstrating all core features.
+This article provides a quick guide to the package, articulating its key ideas and design principles and demonstrating all core features.
 At this point the API is stable---it has changed very little over the 5 years and no further changes are planned.
@@
 
-Last but not least, \pkg{collapse} users are encouraged to also familiarize themselves with the \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse}},\footnote{Website: https://fastverse.github.io/fastverse/} a suite of complementary high-performance packages for statistical computing and data manipulation in \proglang{R} which offer more advanced tools in several statistical computing domains.
-The \pkg{fastverse} metapackage also provides a lightweight framework to jointly load and manage these packages, as well as to build customized and fully separate package verses.
+Finally, \pkg{collapse} users are also encouraged to familiarize themselves with the \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse}},\footnote{Website: https://fastverse.github.io/fastverse/} a suite of complementary high-performance packages for statistical computing and data manipulation in \proglang{R} that offer more advanced tools in several statistical computing domains.
+The \pkg{fastverse} metapackage additionally provides a lightweight framework to jointly load and manage these packages, as well as to build customized and fully separate package verses.
 

### 2025-05-29 misc/JSS article/final/article.tex
Final article update.

@@
              com-pu-ta-tions eval-u-at-ed fa-cil-i-tate im-ple-men-ta-tions re-place-ment
-             sub-set-ting de-com-po-si-tion al-go-rith-mic en-hanced mul-ti-ple coun-tries com-pu-ta-tion-al}
+             sub-set-ting de-com-po-si-tion al-go-rith-mic en-hanced mul-ti-ple coun-tries
+             com-pu-ta-tion-al par-tic-u-lar-ly com-po-nents}
 \usepackage[english]{babel}
@@
 \begin{Sinput}
-R> fmean(num_vars(iris), g = iris$Species, nthreads = 4)
+R> fmean(num_vars(iris), g = iris$Species)
 \end{Sinput}
@@
 Emplyoing the \emph{Fast Statistical Functions} directly often results in more efficient code.
-For example, computing population shares for each year as \code{fmutate(wlddev, POP\_share = fsum(\\POP, year, TRA = "/"))} is considerably more efficient than \code{wlddev |> fgroup_by(year) |> fmutate(POP\_share = fsum(POP, TRA = "/")) |> fungroup()}, which in turn is more efficient than using \code{POP / fsum(POP)} (extra full-length allocation) or \code{proportions(POP)} (slow split-apply-combine evaluation logic) inside the grouped \fct{fmutate} call.
+For example, computing population shares for each year as \code{fmutate(wlddev, POP\_share = fsum\\(POP, year, TRA = "/"))} is considerably more efficient than \code{wlddev |> fgroup_by(year) |> fmutate(POP\_share = fsum(POP, TRA = "/")) |> fungroup()}, which in turn is more efficient than using \code{POP / fsum(POP)} (extra full-length allocation) or \code{proportions(POP)} (slow split-apply-combine evaluation logic) inside the grouped \fct{fmutate} call.
 \newline
@@
 The main reason for not providing higher-level metaprogramming capabilities is to keep all functions as simple as possible.
-It also compels users to think deeply about their programs and devise more efficient solutions.
-\newline
-
-These metaprogramming restrictions are vindicated by internally complex data manipulation functions producing greater overheads that can noticeably add up in larger scripts:\footnote{\fct{bmark} is a slim wrapper around \fct{bench::mark}.
+It also compels users to think more deeply about their programs and devise more efficient solutions.
+\newline
+
+As the small benchmark below illustrates, \pkg{dplyr}'s internally more complex data manipulation functions produce much greater overheads, which can noticeably add-up in longer scripts.\footnote{\fct{bmark} is a slim wrapper around \fct{bench::mark}.
 See the Computational details section.}
-\begin{Schunk}
-\begin{Sinput}
-R>  bmark(collapse = collapse::fsummarise(mtcars, sum = sum(mpg)),
-+         dplyr = dplyr::summarise(mtcars, sum = sum(mpg)))
+%
+\begin{Schunk}
+\begin{Sinput}
+R> bmark(collapse = collapse::fsummarise(wlddev, mean = fmean(PCGDP)),
++        dplyr = dplyr::summarise(wlddev, mean = fmean(PCGDP)))
 \end{Sinput}
@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1   collapse   3.81µs   4.22µs        0B  2235    0     9.91ms
-2      dplyr 239.44µs 275.77µs    1.92KB    33    0    10.35ms
-\end{Soutput}
-\end{Schunk}
+1   collapse   6.68µs   8.86µs        0B  9999    1       94ms
+2      dplyr 210.82µs 248.66µs    2.15KB  9455   29       2.9s
+\end{Soutput}
+\end{Schunk}
+%
 Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices.
-For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 44.7 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
+For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions like \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 45 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
 %On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another recent example involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}.
@@
 Time series functions such as \fct{fgrowth} to compute growth rates are S3 generic and can be applied to most time series classes.
-In addition to a \code{g} argument for grouped computation, these functions also have a \code{t} argument for indexation.
-If \code{t} is a plain numeric vector or a factor, it is coerced to integer and interpreted as time steps.\footnote{This is premised on the observation that the most common form of temporal identifier is a plain numeric vector representing calendar years.
-Users need to manually call \code{timeid()} on plain numeric vectors with decimals to yield an appropriate integer representation.
-If \code{t} is a numeric time object (e.g., \class{Date}, \class{POSIXct}, etc.), then it is internally passed through \code{timeid()} which computes the greatest common divisor (GCD) and generates an integer time-id.
-For the GCD approach to work, \code{t} must have an appropriate class, e.g., for monthly/quarterly data, \code{zoo::yearmon()/zoo::yearqtr()} should be used instead of \class{Date} or \class{POSIXct}.} But first, a basic example:
-%
-\begin{Schunk}
-\begin{Sinput}
-R> fgrowth(airmiles) |> round(2)
+In addition to a \code{g} argument for grouping, these functions also have a \code{t} argument for indexation.
+But first, I provide a basic example of computing the annualized 10-year growth rates in miles flown by airlines in the USA from 1937-1960.
+%
+\begin{Schunk}
+\begin{Sinput}
+R> fgrowth(airmiles, n = 10, power = 1/10) |> na.omit() |> round(1)
 \end{Sinput}
@@
 Time Series:
-Start = 1937 
+Start = 1947 
 End = 1960 
 Frequency = 1 
- [1]    NA 16.50 42.29 54.03 31.65  2.38 15.23 33.29 54.36 76.92  2.71 -2.10
-[13] 12.91 18.51 32.03 18.57 17.82 13.61 18.19 12.83 13.32  0.01 15.49  4.25
-\end{Soutput}
-\end{Schunk}
+ [1] 31.0 28.7 25.7 22.5 22.5 24.3 24.6 22.6 19.4 14.2 15.3 15.5 15.8 14.3
+\end{Soutput}
+\end{Schunk}
+%
+The results show that the flight volume as been growing steadily but at a decreasing rate. \newline
+
 To illustrate the full capabilities of these time series functions, I generate a sector-level trade dataset of export values (v) by country (c), sector (s), and year (y).
@@
 It is worth highlighting that the flexibility of this architecture is new to the \proglang{R} ecosystem: A \class{pdata.frame} or \class{fixest\_panel} only works inside \pkg{plm}/\pkg{fixest} estimation functions.\footnote{And, in the case of \pkg{fixest}, inside \pkg{data.table} due to dedicated methods.} Time series classes like \class{xts} and \class{tsibble} also do not provide deeply indexed structures or native handling of irregularity in basic operations.
-\class{indexed\_series} and \class{indexed\_frame}, on the other hand, work 'anywhere', and can be superimposed on any suitable object, as long as \pkg{collapse}'s functions (\code{flag()/L()}, etc.) are used to perform the time-based computations.
-\newline
-
-An example using the \code{exports} data follows. \emph{Note} that data need not be sorted for indexing.
+\class{indexed\_series} and \class{indexed\_frame}, on the other hand, work 'anywhere', and can be superimposed on any suitable object as long as \pkg{collapse}'s functions (\code{flag()/L()}, etc.) are used to perform the time-based computations.
+\newline
+
+An example follows using the \code{exports} data. \emph{Note} that data can be unsorted for indexation.
 \begin{Schunk}
@@
 R> exportsi <- exports |> findex_by(c, s, y)
-R> exportsi |> G() |> head(5)
-\end{Sinput}
-\begin{Soutput}
-   c  s    y   G1.v
-1 c1 s1 2002     NA
-2 c1 s1 2003  22.17
-3 c1 s1 2004 -68.24
-4 c1 s1 2005  44.98
-5 c1 s1 2006 277.76
+R> exportsi |> G(0:1) |> head(5)
+\end{Sinput}
+\begin{Soutput}
+   c  s    y      v   G1.v
+1 c1 s1 2002 0.5525     NA
+2 c1 s1 2003 0.6749  22.17
+3 c1 s1 2004 0.2144 -68.24
+4 c1 s1 2005 0.3108  44.98
+5 c1 s1 2006 1.1740 277.76
 
@@
 \end{Schunk}
-\fct{psmat}, for panel-series to matrix, generates a matrix/array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
+\fct{psmat}, for panel-series to matrix, generates a matrix or array from panel data. Thanks to deep indexation, indexed computations work inside arbitrary data masking environments:
 \begin{Schunk}
@@
 \end{Schunk}
-Indexed series/frames also support transformation such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}.
+Indexed series/frames also support transformations such as grouped scaling with \fct{fscale} or demeaning with \fct{fwithin}.
 Functions \fct{psacf}/\fct{pspacf}/\fct{psccf} provide panel-data autocorrelation functions, which are computed using group-scaled and suitably lagged panel-series.
-The \class{index\_df} attached to these objects can also be used with other general tools such as \code{collapse::BY()} to perform grouped computations using 3rd-party functions.
-For example, below I compute a 5-year rolling average (\fct{ix} abbreviates \fct{findex}).
-%
-\begin{Schunk}
-\begin{Sinput}
-R> BY(vi, ix(vi)$c.s, data.table::frollmean, 5) |> head(10)
+The \class{index\_df} attached to these objects can also be used with other general tools such as \fct{collapse::BY} to perform grouped computations using third-party functions: %For example, below I compute a 5-year rolling average ().
+%
+\begin{Schunk}
+\begin{Sinput}
+R> BY(vi, findex(vi)$c.s, data.table::frollmean, 5) |> head(10)
 \end{Sinput}
@@
 \end{Schunk}
-% Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
+%
+Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
 %
@@
 \end{Schunk}
-Users can also invoke the \code{validate} argument to examine the uniqueness of the join keys in either table: passing a '1' produces an error if the key is not unique.
+Users can also invoke the \code{validate} argument to examine the uniqueness of the join keys in either table: passing a '1' produces an error if the respective key is not unique.
 %
@@
 First, \fct{join} is class-agnostic and preserves the attributes of \code{x} (any list-based object).
-It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"}, or \code{"anti"} join).
-This demonstrates the latter:
+It supports 6 different join operations (\code{"left"}, \code{"right"}, \code{"inner"}, \code{"full"}, \code{"semi"}, or \code{"anti"}).
+This demonstrates the latter two:
 \begin{Schunk}
@@
 By default (\code{sort = FALSE}), the order of rows in \code{x} is preserved.
-Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join (using \fct{fmatch}) followed by sorting, particularly if the data is already sorted on the keys.
+Setting \code{sort = TRUE} sorts all records in the joined table by the keys.\footnote{This is done using a separate sort-merge-join algorithm, so it is faster than performing a hash join using \fct{fmatch} followed by sorting, particularly if the data is already sorted on the keys.
 } The join relationship is indicated inside the \code{<>} as the number of records joined from each table divided by the number of unique matches.
 % In multi-match settings, this will be reflected by few records from \code{y} being used.
+\newpage
+\fct{join}'s handling of duplicate columns in both tables is also rather special.
+By default (\code{suffix = NULL}), \fct{join} extracts the name of the \code{y} table and appends \code{y}-columns with it.
+% \code{x}-columns are not renamed.
+%
 \begin{Schunk}
@@
 \end{Schunk}
-As shown above, \fct{join}'s handling of duplicate columns in both tables is rather special.
-By default (\code{suffix = NULL}), \fct{join} extracts the name of the \code{y} table and appends \code{y}-columns with it.
-\code{x}-columns are not renamed.
-This is congruent to the principle of adding columns to \code{x} and altering this table as little as possible.
+%
+This is congruent to the principle of adding columns to \code{x} in the default first-match left join by altering this table as little as possible.
 Alternatively, option \code{drop.dup.cols = "x"/"y"} can be used to remove duplicate columns from either \code{x} or \code{y} before the join operation.
@@
 
-\code{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings.
+\fct{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings.
 It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information.
-Its basic syntax is:
+Its basic syntax is: \newpage
 %
@@
 To create nested data, \fct{rsplit} generalizes \fct{split} and (recursively) splits up data frame-like objects into (nested) lists.
-For example, we can split the \code{GGDC10S} data by country and variable, such that, for example, agricultural employment in Argentina can be accessed as:\footnote{If a nested structure is not needed, \code{flatten = TRUE} lets \fct{rsplit} operate like a faster version of \fct{split}.}
-%
-\begin{Schunk}
-\begin{Sinput}
-R> dl <- GGDC10S |> rsplit( ~ Country + Variable)
-R> dl$ARG$EMP$AGR[1:12]
+For example, we can split the \code{GGDC10S} data by country and variable, such that agricultural employment in Argentina can be accessed as:\footnote{If a nested structure is not needed, \code{flatten = TRUE} lets \fct{rsplit} operate like a faster version of \fct{split}.} \newpage
+%
+\begin{Schunk}
+\begin{Sinput}
+R> d_list <- GGDC10S |> rsplit( ~ Country + Variable)
+R> d_list$ARG$EMP$AGR[1:12]
 \end{Sinput}
@@
 %
-This is a convenient data representation for \emph{Shiny Apps} where we can let the user choose data (e.g., \code{dl[[input$country]][[input$variable]][[input$sector]]}) without expensive subsetting operations.
+This is a convenient data representation for \emph{Shiny Apps} where we can let the user choose data (e.g., \code{d_list[[input$country]][[input$variable]][[input$sector]]}) without expensive subsetting operations.
 As mentioned, such data representation can also be the result of an API call parsing JSON or a nested loop or \fct{lapply} call.
@@
 \begin{Sinput}
-R> result <- list()
+R> results <- list()
 R> for (country in c("ARG", "BRA", "CHL")) {
@@
 +      m <- lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year,
-+              data = dl[[country]][[variable]])
-+      result[[country]][[variable]] <- list(model = m, BIC = BIC(m),
++              data = d_list[[country]][[variable]])
++      results[[country]][[variable]] <- list(model = m, BIC = BIC(m),
 +                                             summary = summary(m))
@@
 \end{Schunk}
-This programming may not be ideal for this particular use case as I could have used data frame-based tools and saved the result in a column.\footnote{E.g., \code{GGDC10S |> fgroup\_by(Country, Variable) |> fsummarise(result = my\_fun(lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year)))} with \code{my\_fun <- function(m) list(list(m, BIC(m), summary(m)))}.}
+This programming may not be ideal for this particular use case as I could have used data frame-based tools and saved the result in a column.\footnote{E.g., \code{GGDC10S |> fgroup\_by(Country, Variable) |> fsummarise(results = my\_fun(lm(log(AGR+1) ~ log(MIN+1) + log(MAN+1) + Year)))} with \code{my\_fun <- function(m) list(list(m, BIC(m), summary(m)))}.}
 However, there are limits to such workflows.
-For example, I recently trained a complex ML model for different variables and parameters, which involved loading a different dataset at each iteration.
+For example, I recently trained a complex ML model for different variables and parameters, while also loading a different dataset for each combination.
 Loops are useful in such cases, and lists a natural vehicle to structure complex outputs.
 The main issue with nested lists is that they are complex to query.
-What if we want to know just the $R^2$ of these 6 models?
-We would need to use, e.g., \code{result$ARG$EMP$summary$r.squared} for each model.\newline
+What if we wanted to know the $R^2$ of these 6 models?
+We would need to use, e.g., \code{results$ARG$EMP$summary$r.squared} for each model.\newline
 
@@
 \begin{Sinput}
-R> str(r_sq_l <- result |> get_elem("r.squared"))
+R> str(r_sq <- results |> get_elem("r.squared"))
 \end{Sinput}
@@
 \begin{Sinput}
-R> rowbind(r_sq_l, idcol = "Country", return = "data.frame")
+R> rowbind(r_sq, idcol = "Country", return = "data.frame")
 \end{Sinput}
@@
 \begin{Sinput}
-R> r_sq_l |> t_list() |> rowbind(idcol = "Variable", return = "data.frame")
+R> r_sq |> t_list() |> rowbind(idcol = "Variable", return = "data.frame")
 \end{Sinput}
@@
 \end{Schunk}
-\fct{rowbind} is limited if \fct{get\_elem} returns a more nested or asymmetric list, potentially with vectors/arrays in the final nodes. Suppose we want to extract the coefficient matrices:
-\begin{Schunk}
-\begin{Sinput}
-R> result$ARG$EMP$summary$coefficients
+\fct{rowbind} is limited if \fct{get\_elem} returns a more nested or asymmetric list, potentially with vectors/arrays in the final nodes. Suppose we wanted to extract the coefficient matrices:
+\begin{Schunk}
+\begin{Sinput}
+R> results$ARG$EMP$summary$coefficients
 \end{Sinput}
@@
 \begin{Sinput}
-R> result |> get_elem("coefficients") |> get_elem(is.matrix) |>
+R> results |> get_elem("coefficients") |> get_elem(is.matrix) |>
 +    unlist2d(idcols = c("Country", "Variable"),
@@
 For greater detail, \fct{descr} provides a rich (grouped, weighted) statistical description.
-It does not support panel-variance decompositions like \fct{qsu}, but also computes detailed (grouped, weighted) frequency tables for categorical data and is thus very useful with mixed-type data.
-%
-\begin{Schunk}
-\begin{Sinput}
-R> descr(wlda15, by = income + LIFEEX ~ OECD)
-\end{Sinput}
-\begin{Soutput}
-Dataset: wlda15, 2 Variables, N = 216
+It does not support panel-variance decompositions like \fct{qsu}, but also computes detailed frequency tables for categorical data.
+Below, I summarize income and LIFEEX by OECD membership, scaling the weights to limit printout and replacing missing weights with \code{0} (the default).\footnote{This is necessary in \fct{descr} because \fct{fquantile} does not support missing weights for non-missing \code{x}.\vspace{-5mm}}
+%
+\begin{Schunk}
+\begin{Sinput}
+R> descr(wlda15, by = income + LIFEEX ~ OECD, w = ~ replace_na(POP / 1e6))
+\end{Sinput}
+\begin{Soutput}
+Dataset: wlda15, 2 Variables, N = 216, WeightSum = 7620.902563
 Grouped by: OECD [2]
-         N   Perc
-FALSE  180  83.33
-TRUE    36  16.67
+         N   Perc  WeightSum   Perc
+FALSE  180  83.33    6311.15  82.81
+TRUE    36  16.67    1309.75  17.19
 -----------------------------------------------------------------------------
 income (factor): Income Level
-Statistics (N = 216)
-         N   Perc  Ndist
-FALSE  180  83.33      4
-TRUE    36  16.67      2
-
-Table (Freq Perc)
+Statistics (WeightSum = 7621, 0% NAs)
+       WeightSum   Perc  Ndist
+FALSE    6311.15  82.81      4
+TRUE     1309.75  17.19      2
+
+Table (WeightSum Perc)
                          FALSE       TRUE      Total
-High income          45 25.0  34 94.4  79 36.6
-Upper middle income  58 32.2   2  5.6  60 27.8
-Lower middle income  47 26.1   0  0.0  47 21.8
-Low income           30 16.7   0  0.0  30 13.9
+Lower middle income  3064 48.5     0  0.0  3064 40.2
+Upper middle income  2460 39.0   211 16.1  2671 35.0
+High income            93  1.5  1099 83.9  1192 15.6
+Low income            695 11.0     0  0.0   695  9.1
 -----------------------------------------------------------------------------
 LIFEEX (numeric): Life expectancy at birth, total (years)
-Statistics (N = 200, 7.41% NAs)
-         N  Perc  Ndist   Mean    SD    Min    Max   Skew  Kurt
-FALSE  164    82    164  71.25  7.06  53.28  85.08   -0.5  2.61
-TRUE    36    18     36  80.83  2.55  75.05  84.36  -0.92  2.72
+Statistics (N = 199, 7.87% NAs)
+         N  Ndist  WeightSum   Perc   Mean    SD    Min    Max   Skew  Kurt
+FALSE  163    164    6310.41  82.81  71.14  5.77  53.28  85.08  -0.99  3.76
+TRUE    36     36    1309.75  17.19  80.32  2.77  75.05  84.36  -0.29  2.11
 
@@
           1%     5%    10%    25%    50%    75%    90%    95%    99%
-FALSE   54.3  58.38  61.25  66.43  72.56  76.68  78.93  80.88  83.77
-TRUE   75.12  75.83   76.8  79.04  81.77  82.63  83.21  83.54  84.13
+FALSE  54.69  59.38  63.67  69.64  71.71  76.89  76.91  76.91  77.94
+TRUE   75.07  75.14  76.03  78.77  80.86  82.86  83.61  84.05   84.3
 -----------------------------------------------------------------------------
@@
 %
- It also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden.
+\fct{descr} also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden.
 The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples.
@@
 \begin{Sinput}
-R> wlda15 |> with(qtab(OECD, income, w = POP) / 1e6)
+R> wlda15 |> with(qtab(OECD, income, w = POP / 1e6))
 \end{Sinput}
@@
 R> g_char <- sample(char <- outer(char, char, paste0), 1e7, TRUE)
-R> bmark(base_int = unique(g_int), collapse_int = funique(g_int))
+R> bmark(base = unique(g_int), collapse = funique(g_int))
 \end{Sinput}
@@
   expression    min  median mem_alloc n_itr n_gc total_time
-1     base_int 46.5ms 46.5ms   166.2MB     1    0     46.5ms
-2 collapse_int 10.5ms 10.5ms    38.2MB     1    0     10.5ms
-\end{Soutput}
-\begin{Sinput}
-R> bmark(base_char = unique(g_char), collapse_char = funique(g_char))
+1       base 42.8ms 44.08ms   166.2MB    30   30      1.33s
+2   collapse  5.6ms  8.27ms    38.2MB   303   38      2.59s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(base = unique(g_char), collapse = funique(g_char))
 \end{Sinput}
@@
   expression    min median mem_alloc n_itr n_gc total_time
-1     base_char 69.8ms 69.8ms   166.2MB     1    0     69.8ms
-2 collapse_char 20.6ms 20.6ms    38.2MB     1    0     20.6ms
-\end{Soutput}
-\begin{Sinput}
-R> bmark(base_int = match(g_int, int), collapse_int = fmatch(g_int, int))
+1       base 70.2ms 71.5ms   166.2MB    19   21      1.36s
+2   collapse 15.5ms 18.8ms    38.2MB   140   17      2.61s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(base = match(g_int, int), collapse = fmatch(g_int, int))
 \end{Sinput}
@@
   expression     min  median mem_alloc n_itr n_gc total_time
-1     base_int 27.28ms 27.28ms    76.3MB     1    1     27.3ms
-2 collapse_int  7.93ms  8.01ms    38.2MB     2    0       16ms
-\end{Soutput}
-\begin{Sinput}
-R> bmark(base_char = match(g_char, char), data.table_char =
-+        chmatch(g_char, char), collapse_char = fmatch(g_char, char))
+1       base 17.26ms 18.07ms    76.3MB   111   37      2.02s
+2   collapse  7.56ms  8.16ms    38.2MB   289   41      2.36s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(base = match(g_char, char), data.table =
++        chmatch(g_char, char), collapse = fmatch(g_char, char))
 \end{Sinput}
@@
   expression    min median mem_alloc n_itr n_gc total_time
-1       base_char 72.4ms 72.4ms   114.5MB     1    0     72.4ms
-2 data.table_char 46.5ms 46.5ms    38.1MB     1    1     46.5ms
-3   collapse_char   29ms   29ms    38.1MB     1    0       29ms
-\end{Soutput}
-\end{Schunk}
-%
-The second set below shows that \pkg{collapse}'s statistical functions are very efficient on aggregating a numeric matrix with 10,000 rows and 1000 columns.
+1       base 52.7ms 54.8ms   114.5MB    27   25      1.49s
+2 data.table 32.9ms   34ms    38.1MB    73   13      2.48s
+3   collapse 16.2ms 17.1ms    38.1MB   143   23      2.46s
+\end{Soutput}
+\end{Schunk}
+%
+The second set below shows that \pkg{collapse}'s statistical functions are very efficient in aggregating a numeric matrix with 10,000 rows and 1000 columns.
 They are faster than base \proglang{R} even without multithreading, but using 4 threads in this case induces a sizeable difference.
@@
 R> m <- matrix(rnorm(1e7), ncol = 1000)
-R> bmark(R = colSums(m), collapse = fsum(m))
+R> bmark(base = colSums(m), collapse = fsum(m))
 \end{Sinput}
@@
   expression     min   median mem_alloc n_itr n_gc total_time
-1          R   6.07ms   6.09ms    7.86KB     2    0     12.2ms
-2   collapse 311.31µs 345.01µs    7.86KB    28    0     10.1ms
-\end{Soutput}
-\begin{Sinput}
-R> bmark(R = colMeans(m), collapse = fmean(m))
+1       base   5.4ms   6.11ms    7.86KB   494    0         3s
+2   collapse   296µs 386.88µs    7.86KB  7172    0         3s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(base = colMeans(m), collapse = fmean(m))
 \end{Sinput}
@@
   expression      min  median mem_alloc n_itr n_gc total_time
-1          R   6.22ms   6.22ms    7.86KB     2    0     12.4ms
-2   collapse 329.27µs 341.08µs    7.86KB    27    0     10.3ms
-\end{Soutput}
-\begin{Sinput}
-R> bmark(MS = matrixStats::colMedians(m), collapse = fmedian(m))
+1       base   5.43ms   5.9ms    7.86KB   505    0         3s
+2   collapse 303.24µs 377.2µs    7.86KB  7608    1      2.99s
+\end{Soutput}
+\begin{Sinput}
+R> bmark(matrixStats = matrixStats::colMedians(m), collapse = fmedian(m))
 \end{Sinput}
@@
    expression    min median mem_alloc n_itr n_gc total_time
-1         MS 77.5ms 77.5ms   86.04KB     1    0     77.5ms
-2   collapse 19.5ms 19.5ms    7.86KB     1    0     19.5ms
-\end{Soutput}
-\end{Schunk}
-%
-Below, I also show a grouped version summing the columns within 1000 random groups.
+1 matrixStats 76.6ms 77.2ms   86.04KB    39    0      3.01s
+2    collapse 19.4ms 19.6ms    7.86KB   153    0         3s
+\end{Soutput}
+\end{Schunk}
+%
+Below, I also benchmark a grouped version summing the columns within 1000 random groups.
 %
@@
 R> g <- sample.int(1e3, 1e4, TRUE)
-R> bmark(R = rowsum(m, g), collapse = fsum(m, g))
+R> bmark(base = rowsum(m, g), collapse = fsum(m, g))
 \end{Sinput}
@@
   expression     min median mem_alloc n_itr n_gc total_time
-1          R 6.04ms 6.29ms    7.85MB     2    0     12.6ms
-2   collapse  1.3ms 1.39ms    7.67MB     8    0     11.2ms
-\end{Soutput}
-\end{Schunk}
-%
-I now turn to basic operations on a medium sized real-world database recording all flights from New York City (EWR, JFK, and LGA) in 2023---provided by the \pkg{nycflights23} package.
+1       base   5.1ms 5.74ms    7.85MB   456   16      2.72s
+2   collapse 905.8µs  1.2ms    7.67MB  1781   54      2.39s
+\end{Soutput}
+\end{Schunk}
+%
+I now turn to basic operations on a medium-sized real-world database recording all flights from New York City (EWR, JFK, and LGA) in 2023---provided by the \pkg{nycflights23} package.
 The \code{flights} table has 435k flights, and grouping it by day and route yields 76k unique trips.
@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1      dplyr 369.09ms 369.09ms   48.56MB     1    6    369.1ms
-2 data.table   8.36ms   8.46ms   18.93MB     2    0     16.9ms
-3   collapse   4.24ms   4.29ms    9.11MB     3    0     12.9ms
+1      dplyr 374.95ms 499.62ms   48.56MB     7   65      3.33s
+2 data.table   7.32ms   9.35ms   18.93MB   282   28         3s
+3   collapse      3ms   3.93ms    9.11MB   670   31      3.01s
 \end{Soutput}
@@
        expression    min median mem_alloc n_itr n_gc total_time
-1      dplyr_mean  1.28s  1.28s   48.56MB     1   23      1.28s
-2 data.table_mean  8.9ms 9.05ms   18.93MB     2    0    18.11ms
-3   collapse_mean 4.08ms 4.39ms    9.11MB     3    0    13.02ms
+1      dplyr_mean  1.24s   1.3s   48.56MB     3   74      4.05s
+2 data.table_mean 7.58ms 9.03ms   18.93MB   288   29      3.01s
+3   collapse_mean 3.14ms    4ms    9.11MB   650   31         3s
 \end{Soutput}
@@
          expression     min  median mem_alloc n_itr n_gc total_time
-1      dplyr_median   4.51s   4.51s    52.8MB     1   90      4.51s
-2 data.table_median 23.72ms 23.72ms    18.9MB     1    0    23.72ms
-3   collapse_median 11.54ms 11.54ms    11.1MB     1    0    11.54ms
+1      dplyr_median   4.46s   4.46s    52.8MB     1   76      4.46s
+2 data.table_median 22.12ms 23.58ms    18.9MB   123   14      3.02s
+3   collapse_median  9.37ms 10.83ms    11.1MB   247   18         3s
 \end{Soutput}
@@
   expression     min   median mem_alloc n_itr n_gc total_time
-1      dplyr 112.36ms 112.36ms   20.18MB     1    3    112.4ms
-2 data.table  63.78ms  63.78ms    5.77MB     1    1     63.8ms
-3   collapse   6.46ms   6.54ms     6.8MB     2    0     13.1ms
+1      dplyr 93.15ms 107.41ms   20.18MB    25   57      3.04s
+2 data.table 57.33ms  62.35ms    5.77MB    48   25      3.02s
+3   collapse  5.26ms   5.82ms     6.8MB   477   16         3s
 \end{Soutput}
@@
         expression      min  median mem_alloc n_itr n_gc total_time
-1      dplyr_joins 172.7ms 172.7ms   558.9MB     1    2    172.7ms
-2 data.table_joins 204.5ms 204.5ms   490.4MB     1    1    204.5ms
-3   collapse_joins  59.1ms  59.1ms    89.7MB     1    2     59.1ms
+1      dplyr_joins 148.67ms 163.5ms   558.9MB    17   36      3.12s
+2 data.table_joins  109.4ms 197.5ms   490.4MB    16   31      3.08s
+3   collapse_joins   8.09ms  11.9ms    89.7MB   198   43         3s
 \end{Soutput}
@@
   expression    min median mem_alloc n_itr n_gc total_time
-1      tidyr  82.7ms  82.7ms     251MB     1    2     82.7ms
-2 data.table 143.9ms 143.9ms     209MB     1    2    143.9ms
-3   collapse  17.7ms  17.7ms     209MB     1    0     17.7ms
+1      tidyr 58.7ms 59.8ms     251MB     8   38   480.69ms
+2 data.table 35.2ms 38.5ms     209MB    38   27      1.46s
+3   collapse 13.6ms 16.6ms     209MB    63   64      1.04s
 \end{Soutput}
@@
      expression      min   median mem_alloc n_itr n_gc total_time
-1         tidyr 376.32ms 376.32ms   142.4MB     1    5    376.3ms
-2    data.table 237.61ms 237.61ms      21MB     1    4    237.6ms
-3 collapse_fsum    7.6ms   7.94ms    39.1MB     2    0     15.9ms
-4 collapse_itnl   5.05ms   5.08ms    12.4MB     2    0     10.2ms
-\end{Soutput}
-\end{Schunk}
-%
-Again, \pkg{collapse} is fastest, as it offers full vectorization, either via \fct{fsum}, which translates to \code{fsum(x, g, TRA = "fill")} before pivoting and thus entails a full deep copy of the \code{vars} columns, or via an optimized internal sum function which sums values 'on the fly' during the reshaping process.
+1         tidyr 376.05ms 400.31ms   142.4MB     8   65      3.33s
+2    data.table 221.19ms 228.87ms      21MB    13   43      3.13s
+3 collapse_fsum   5.64ms   7.45ms    39.1MB   324   27      3.01s
+4 collapse_itnl   3.78ms   4.83ms    12.4MB   556   17         3s
+\end{Soutput}
+\end{Schunk}
+%
+Again, \pkg{collapse} is fastest, as it offers full vectorization, either via \fct{fsum}, which translates to \code{fsum(x, g, TRA = "fill")} before pivoting and thus entails a full deep copy of the \code{vars} columns, or via the optimized internal sum function which sums values 'on the fly' during the reshaping process.
 \pkg{data.table} is not vectorized here but at least memory efficient.

### 2025-05-29 misc/JSS article/final/article_replication.R

@@ -33,7 +33,7 @@ vlabels(GGDC10S) <- trimws(vlabels(GGDC10S))
 
 # This is the benchmark function
 bmark <- function(...) {
-  bench::mark(..., min_time = 2, check = FALSE) |> janitor::clean_names() |>
+  bench::mark(..., min_time = 3, check = FALSE) |> janitor::clean_names() |>
   fselect(expression, min, median, mem_alloc, n_itr, n_gc, total_time) |>
   fmutate(expression = names(expression)) |> dapply(as.character) |> qDF()
 }
@@ -53,7 +53,7 @@ fmean(iris$Sepal.Length)
 fmean(num_vars(iris))
 identical(fmean(num_vars(iris)), fmean(as.matrix(num_vars(iris))))
 fmean(iris$Sepal.Length, g = iris$Species)
-fmean(num_vars(iris), g = iris$Species, nthreads = 4)
+fmean(num_vars(iris), g = iris$Species)
 
 
 ###################################################
@@ -71,7 +71,7 @@ BY(num_vars(iris), g = iris$Species, FUN = mad) |> head(1)
 ###################################################
 ### code chunk number 6: fscale
 ###################################################
-fscale(iris$Sepal.Length, g = iris$Species) |> fsd(iris$Species)
+fscale(iris$Sepal.Length, g = iris$Species) |> fsd(g = iris$Species)
 
 
 ###################################################
@@ -135,7 +135,7 @@ str(g <- GRP(wlddev, ~ income + OECD))
 ###################################################
 add_vars(g$groups,
          get_vars(wlddev, "country") |> fmode(g, wlddev$POP, use = FALSE),
-         get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = F),
+         get_vars(wlddev, c("PCGDP", "LIFEEX")) |> fmean(g, wlddev$POP, use = FALSE),
          get_vars(wlddev, "POP") |> fsum(g, use = FALSE))
 
 
@@ -156,7 +156,7 @@ bmark(collapse = collapse::fsummarise(mtcars, sum = sum(mpg)),
 ###################################################
 ### code chunk number 16: ts_example
 ###################################################
-fgrowth(airmiles) |> round(2)
+fgrowth(airmiles, n = 10, power = 1/10) |> na.omit() |> round(1)
 
 
 ###################################################
@@ -196,7 +196,7 @@ identical(A, B)
 ### code chunk number 21: indexing
[…]

### 2025-05-30 misc/JSS article/final/article.tex
Add some gc()'s to benchmarks.

@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1   collapse   6.68µs   8.86µs        0B  9999    1       94ms
-2      dplyr 210.82µs 248.66µs    2.15KB  9455   29       2.9s
+1   collapse   6.68µs   8.04µs    1.05KB  9999    1    83.59ms
+2      dplyr 215.13µs 254.04µs    1.54MB  9936   64      2.78s
 \end{Soutput}
@@
 \begin{Sinput}
-R> set.seed(101)
+R> set.seed(101);
 R> int <- 1:1000; g_int <- sample.int(1000, 1e7, replace = TRUE)
@@
   expression     min  median mem_alloc n_itr n_gc total_time
-1       base 42.8ms 44.08ms   166.2MB    30   30      1.33s
-2   collapse  5.6ms  8.27ms    38.2MB   303   38      2.59s
+1       base 45.15ms 48.17ms   166.2MB    62   62      3.02s
+2   collapse  5.52ms  7.74ms    38.2MB   371   75         3s
 \end{Soutput}
@@
   expression    min median mem_alloc n_itr n_gc total_time
-1       base 70.2ms 71.5ms   166.2MB    19   21      1.36s
-2   collapse 15.5ms 18.8ms    38.2MB   140   17      2.61s
+1       base 71.9ms 73.4ms   166.2MB    41   41      3.02s
+2   collapse 14.8ms 17.7ms    38.2MB   165   33      3.01s
 \end{Soutput}
@@
   expression     min  median mem_alloc n_itr n_gc total_time
-1       base 17.26ms 18.07ms    76.3MB   111   37      2.02s
-2   collapse  7.56ms  8.16ms    38.2MB   289   41      2.36s
+1       base  17.1ms 17.58ms    76.3MB    79   78       1.4s
+2   collapse  7.48ms  8.12ms    38.2MB   260   70      2.11s
 \end{Soutput}
@@
   expression    min median mem_alloc n_itr n_gc total_time
-1       base 52.7ms 54.8ms   114.5MB    27   25      1.49s
-2 data.table 32.9ms   34ms    38.1MB    73   13      2.48s
-3   collapse 16.2ms 17.1ms    38.1MB   143   23      2.46s
+1       base 58.6ms 60.6ms   114.5MB    49   49      3.02s
+2 data.table 32.8ms 34.2ms    38.1MB    86   22      3.01s
+3   collapse 19.4ms 20.3ms    38.1MB   140   29         3s
 \end{Soutput}
@@
   expression     min   median mem_alloc n_itr n_gc total_time
-1       base   5.4ms   6.11ms    7.86KB   494    0         3s
-2   collapse   296µs 386.88µs    7.86KB  7172    0         3s
+1       base   5.4ms   6.12ms    7.86KB   492    0      3.01s
+2   collapse 292.5µs 387.94µs    7.86KB  7352    1         3s
 \end{Soutput}
@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1       base   5.43ms   5.9ms    7.86KB   505    0         3s
-2   collapse 303.24µs 377.2µs    7.86KB  7608    1      2.99s
+1       base   5.69ms   6.02ms   32.09KB   499    0         3s
+2   collapse 285.24µs 393.35µs    7.86KB  7200    0         3s
 \end{Soutput}
@@
    expression    min median mem_alloc n_itr n_gc total_time
-1 matrixStats 76.6ms 77.2ms   86.04KB    39    0      3.01s
-2    collapse 19.4ms 19.6ms    7.86KB   153    0         3s
+1 matrixStats 77.1ms   78ms    89.9KB    38    1      2.98s
+2    collapse 19.4ms 19.6ms    27.2KB   154    0      3.01s
 \end{Soutput}
@@
   expression      min median mem_alloc n_itr n_gc total_time
-1       base   5.1ms 5.74ms    7.85MB   456   16      2.72s
-2   collapse 905.8µs  1.2ms    7.67MB  1781   54      2.39s
+1       base   5.38ms 5.63ms    7.87MB   466   35      2.63s
+2   collapse 918.11µs 1.19ms    7.71MB  1714  127       2.2s
 \end{Soutput}
@@
   expression      min   median mem_alloc n_itr n_gc total_time
-1      dplyr 374.95ms 499.62ms   48.56MB     7   65      3.33s
-2 data.table   7.32ms   9.35ms   18.93MB   282   28         3s
-3   collapse      3ms   3.93ms    9.11MB   670   31      3.01s
+1      dplyr 356.33ms 490.66ms    50.5MB     7   69      3.24s
+2 data.table   7.37ms    8.5ms    20.8MB   315   33         3s
+3   collapse   3.21ms   3.78ms     9.2MB   696   27         3s
 \end{Soutput}
@@
        expression    min median mem_alloc n_itr n_gc total_time
-1      dplyr_mean  1.24s   1.3s   48.56MB     3   74      4.05s
-2 data.table_mean 7.58ms 9.03ms   18.93MB   288   29      3.01s
-3   collapse_mean 3.14ms    4ms    9.11MB   650   31         3s
+1      dplyr_mean  1.14s  1.25s   48.57MB     3   69      3.69s
+2 data.table_mean  7.5ms 8.69ms   18.93MB   314   32      3.01s
+3   collapse_mean 3.27ms 3.93ms    9.11MB   643   27         3s
 \end{Soutput}
@@
          expression     min  median mem_alloc n_itr n_gc total_time
-1      dplyr_median   4.46s   4.46s    52.8MB     1   76      4.46s
-2 data.table_median 22.12ms 23.58ms    18.9MB   123   14      3.02s
-3   collapse_median  9.37ms 10.83ms    11.1MB   247   18         3s
+1      dplyr_median   4.21s   4.21s    52.8MB     1   84      4.21s
+2 data.table_median 21.92ms 23.71ms    18.9MB   122   13         3s
+3   collapse_median  9.37ms 10.57ms    11.1MB   255   13      3.01s
 \end{Soutput}
@@
   expression     min  median mem_alloc n_itr n_gc total_time
-1      dplyr 93.15ms 107.41ms   20.18MB    25   57      3.04s
-2 data.table 57.33ms  62.35ms    5.77MB    48   25      3.02s
-3   collapse  5.26ms   5.82ms     6.8MB   477   16         3s
+1      dplyr 91.78ms 107.7ms   20.18MB    27   55      3.14s
+2 data.table 55.02ms  67.2ms    5.77MB    45   27      3.05s
+3   collapse  5.17ms   5.7ms     6.8MB   491   14      3.01s
 \end{Soutput}
@@
         expression      min median mem_alloc n_itr n_gc total_time
-1      dplyr_joins 148.67ms 163.5ms   558.9MB    17   36      3.12s
-2 data.table_joins  109.4ms 197.5ms   490.4MB    16   31      3.08s
-3   collapse_joins   8.09ms  11.9ms    89.7MB   198   43         3s
+1      dplyr_joins 213.13ms  265ms   559.4MB    12   58      3.14s
+2 data.table_joins 173.31ms  229ms     491MB    14   62      3.14s
+3   collapse_joins   9.94ms   14ms    89.7MB   182  100      3.05s
 \end{Soutput}
@@
   expression    min median mem_alloc n_itr n_gc total_time
-1      tidyr 58.7ms 59.8ms     251MB     8   38   480.69ms
-2 data.table 35.2ms 38.5ms     209MB    38   27      1.46s
-3   collapse 13.6ms 16.6ms     209MB    63   64      1.04s
+1      tidyr 72.4ms  134ms     254MB    25   54      3.12s
+2 data.table   43ms   59ms     209MB    45   34      3.11s
+3   collapse 14.1ms   25ms     209MB    77   91      3.02s
 \end{Soutput}
@@
      expression      min   median mem_alloc n_itr n_gc total_time
-1         tidyr 376.05ms 400.31ms   142.4MB     8   65      3.33s
-2    data.table 221.19ms 228.87ms      21MB    13   43      3.13s
-3 collapse_fsum   5.64ms   7.45ms    39.1MB   324   27      3.01s
-4 collapse_itnl   3.78ms   4.83ms    12.4MB   556   17         3s
+1         tidyr 365.57ms 404.26ms     143MB     8   62      3.25s
+2    data.table 222.76ms 231.33ms    21.7MB    13   44      3.08s
+3 collapse_fsum   6.31ms   8.45ms    39.1MB   269   68         3s
+4 collapse_itnl   4.03ms   5.19ms    12.4MB   513   38         3s
 \end{Soutput}
@@
 
+\begin{Schunk}
+\begin{Soutput}
+          used (Mb) gc trigger  (Mb) limit (Mb) max used  (Mb)
+Ncells 1330529 71.1    3842589 205.3         NA  3842589 205.3
+Vcells 9393515 71.7   34009060 259.5      49152 81718535 623.5
+\end{Soutput}
+\end{Schunk}
+
 \section*{Acknowledgments}

### 2025-05-30 misc/JSS article/final/article.Rnw
Add some gc()'s to benchmarks.
Visual tweaks.

@@
 The main reason for not providing higher-level metaprogramming capabilities is to keep all functions as simple as possible.
-It also compels users to think more deeply about their programs and devise more efficient solutions.
+It also compels users to think deeply about their programs and devise more efficient solutions.
 \newline
@@
 
-\fct{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings.
+\fct{collapse::pivot} provides a powerful new implementation of reshaping for \proglang{R} that addresses these shortcomings.
 It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information.
@@
 %
+<<bench_pre_clear, echo=FALSE, include=FALSE>>=
+rm(list = setdiff(ls(), c("bmark", "vars", "r_opts", "oldopts")))
+invisible(gc())
+@
 <<bench_1, eval=TRUE>>=
-set.seed(101)
+set.seed(101);
 int <- 1:1000; g_int <- sample.int(1000, 1e7, replace = TRUE)
@@
 <<bench_flights_join, eval=TRUE, cache=TRUE, echo=FALSE>>=
+rm(list = setdiff(ls(), c("bmark", "vars", "r_opts", "oldopts")))
+invisible(gc())
 bmark(
@@
 set_collapse(oldopts)
+rm(list = ls())
+invisible(gc())
 @

### 2025-05-30 misc/JSS article/final/article.tex
Visual tweaks.

@@
 
-\fct{collapse::pivot} provides a class-agnostic implementation of reshaping for \proglang{R} that addresses these shortcomings.
+\fct{collapse::pivot} provides a powerful new implementation of reshaping for \proglang{R} addressing these shortcomings.
 It has a single intuitive syntax to perform 'longer', 'wider', and 'recast' pivots and supports complex labelled data without loss of information.
@@
 
-\begin{Schunk}
-\begin{Soutput}
-          used (Mb) gc trigger  (Mb) limit (Mb) max used  (Mb)
-Ncells 1330529 71.1    3842589 205.3         NA  3842589 205.3
-Vcells 9393515 71.7   34009060 259.5      49152 81718535 623.5
-\end{Soutput}
-\end{Schunk}
-
 \section*{Acknowledgments}

### 2025-05-30 misc/JSS article/final/jss5278/jss5278.tex
Submission package.

--- 2025-05-30 misc/JSS article/final/article.tex
@@
-\documentclass[nojss]{jss} % article
+\documentclass[article]{jss} % article
 
@@
 The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
-\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
-It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications of the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
-See Section~\ref{sec:glob_opt}.\vspace{-5mm}} % \newline
+\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames, and seamlessly supporting popular extensions to these objects such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
+It is also extensively globally and interactively configurable.
+\newline
 
@@
 
-In summary, \pkg{collapse} is a new foundation package for advanced statistical computing and data transformation in \proglang{R} that integrates seamlessly with the ecosystem and offers outstanding speed and memory efficiency.
+% In summary, \pkg{collapse} is a new foundation package for advanced statistical computing and data transformation in \proglang{R} that integrates seamlessly with the ecosystem and offers outstanding speed and memory efficiency.
 Thus, many core tasks can be done with \pkg{collapse}, and easily extended by specialized packages, yielding more lightweight, faster, and shorter \proglang{R} programs.
@@
 %
-While this article cannot fully present \pkg{collapse}, the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options.
+This article cannot fully present \pkg{collapse}, but the following sections introduce its key features, starting with (\ref{sec:fast_stat_fun}) the \emph{Fast Statistical Functions} and their (\ref{sec:integration}) integration with data manipulation functions; (\ref{sec:ts_ps}) architecture for time series and panel data; (\ref{sec:join_pivot}) table joins and pivots; (\ref{sec:list_proc}) list processing functions; (\ref{sec:summ_stat}) descriptive tools; and (\ref{sec:glob_opt}) global options.
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
+For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
 %

### 2025-05-31 misc/JSS article/final/article.Rnw
Spelling.

@@
 %
-\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, such that using any \emph{Fast Statistical Function} in an expression causes the entire expression to be vectorized (i.e., evaluated only once with \emph{Fast Statistical Function} returning grouped output).
+\fct{fsummarise} and \fct{fmutate} follow an eager vectorization approach, such that using any \emph{Fast Statistical Function} in an expression causes the entire expression to be vectorized (i.e., evaluated only once with \emph{Fast Statistical Functions} returning grouped output).
 This only applies to visible expressions, \pkg{collapse} cannot read the contents of custom functions, thus such functions are always evaluated using split-apply-combine logic.

### 2025-12-09 misc/JSS article/R&R/article.Rnw
Replace sebkrantz.github.io/collapse -> fastverse.github.io/collapse.

@@
 %
-\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/.
+\href{https://fastverse.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://fastverse.github.io/collapse/.
 Linecount (v2.0.19): R: 13,646, C: 18,594, C++: 9,844.
@@
 The package also provides efficient data object conversions, functions for memory efficient R programming, such as (grouped) transformation and math by reference, and helpers to effectively deal with variable labels, attributes, and missing data.
-\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem---notably, \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
+\pkg{collapse} is \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem---notably, \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
 It is globally and interactively configurable, which includes setting different defaults for key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the package namespace to mask equivalent but slower base \proglang{R} or \pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
@@
 
-The package has a built-in structured \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{documentation} facilitating its use.
-This documentation includes a central \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and supplementary topic pages which briefly describe related functionality.
+The package has a built-in structured \href{https://fastverse.github.io/collapse/reference/collapse-documentation.html}{documentation} facilitating its use.
+This documentation includes a central \href{https://fastverse.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and supplementary topic pages which briefly describe related functionality.
 The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:
@@
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} (\href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}).
+For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} (\href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}).
 %
@@
 %
-The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R} means that these operations are implemented using compiled C/C++ code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
+The \href{https://fastverse.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R} means that these operations are implemented using compiled C/C++ code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
 The functions basic syntax is
@@
 All examples support weights (\code{w}), and \fct{fmean} can also be multithreaded across columns (\code{nthreads}).\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions, as detailed in the respective function documentation.
-The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes.
+The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes.
 Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods.
@@
 \end{Code}
- Below, I create a \class{GRP} object from the included World Development Dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}).
+ Below, I create a \class{GRP} object from the included World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}).
 The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.
@@
 
-Exempting \fct{collap}, these examples may seem bulky for quick analysis, but a robust low-level API is very useful for package development, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
+Exempting \fct{collap}, these examples may seem bulky for quick analysis, but a robust low-level API is very useful for package development, as further elucidated in the \href{https://fastverse.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
 I also wrote a blog post on \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{aggregating survey data using \pkg{collapse}}, which showcases more aspects of the \fct{collap} function using real census data.
@@
 %
-\pkg{collapse} also provides a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{fast data manipulation functions} akin to base \proglang{R} and \pkg{tidyverse} functions, including \fct{fselect}, \fct{fsubset}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{ftransform}, \fct{fmutate}, \fct{across}, \fct{frename}, \fct{fcount}, etc.
+\pkg{collapse} also provides a broad set of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{fast data manipulation functions} akin to base \proglang{R} and \pkg{tidyverse} functions, including \fct{fselect}, \fct{fsubset}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{ftransform}, \fct{fmutate}, \fct{across}, \fct{frename}, \fct{fcount}, etc.
 These are integrated with the \emph{Fast Statistical Functions} to enable vectorized statistical operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow.
@@
 The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} calculates weighted quantiles in a theoretically consistent way by applying the probability measure to the sum of weights to create a target sum and cumulatively summing weights to find order statistics which are then combined following \citet{hyndman1996sample}.
-See \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for details.}
+See \href{https://fastverse.github.io/collapse/reference/fquantile.html}{fquantile} for details.}
 %
@@
 Any data frame-like object can become an \class{indexed\_frame} and function as usual for other operations.
-The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}.
+The technical implementation of these classes is described in the \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://fastverse.github.io/collapse/reference/indexing.html}{documentation}.
 Their basic syntax is:
@@
 Unlike \class{pdata.frame}, an \\ \class{indexed\_frame} is a deeply indexed structure---every series inside the frame is already an \class{indexes\_series}.
-A comprehensive set of \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()}, ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.).
+A comprehensive set of \href{https://fastverse.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()}, ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.).
 % A basic demonstration with World Bank panel data showcases the flexibility of these classes.
@@
 %
-Among all \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and 'recast' pivots.
+Among all \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://fastverse.github.io/collapse/reference/join.html}{joins} and \href{https://fastverse.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and 'recast' pivots.
 Both implementations provide outstanding computational performance, syntax, and memory efficiency.
@@
 %
-The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Databas (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}), providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America.
+The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Databas (\href{https://fastverse.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}), providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America.
 While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes.
@@
 @
-More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}.
+More features of \fct{pivot} are demonstrated in the \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}.
 Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.
@@
 Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data.
-\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit.
+\pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit.
 \newline
@@
 %
-\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious and powerful toolkit to examine complex datasets.
+\pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious and powerful toolkit to examine complex datasets.
 A particular focus has been on providing tools for examining longitudinal (panel) data.
-Recall the indexed World Development Panel (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{ssec:gopt}.
+Recall the indexed World Development Panel (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{ssec:gopt}.
 The function \fct{varying} can be used to examine which of these variables are time-varying.
@@
 A \code{stepwise} argument toggles describing one variable at a time, allowing users to naturally 'click-through' a large dataset rather than printing a massive output to the console.
-The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples.
+The \href{https://fastverse.github.io/collapse/reference/descr.html}{documentation} provides more details and examples.
 Both \fct{qsu} and \fct{descr} provide an \code{as.data.frame()} method for efficient tidying and further analysis.
@@
 %
-\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function.
+\pkg{collapse} is \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function.
 These options are saved in an internal environment called \code{.op}.
@@
 Special functions are \code{"n"}, \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} (for use in \code{(f)summarise}/\code{(f)mutate}), \code{table <- qtab}, and replace \code{\%in\%} with a fast version using \fct{fmatch}, respectively.
-There are also several \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}.
+There are also several \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}.
 The most powerful of these is \code{"all"}, which masks all f-functions and special functions, as shown below.
@@
 
-For deeper engagement with \pkg{collapse}, visit its \href{https://sebkrantz.github.io/collapse/index.html}{website} or start with the vignette summarizing all available \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}.
+For deeper engagement with \pkg{collapse}, visit its \href{https://fastverse.github.io/collapse/index.html}{website} or start with the vignette summarizing all available \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}.
 Users can also follow \pkg{collapse} on \href{https://x.com/collapse_R}{Twitter/X} and \href{https://bsky.app/profile/rcollapse.bsky.social}{Bluesky} to be notified about major updates and participate in community discussions.

### 2025-12-09 misc/JSS article/R&R/article.tex
Replace sebkrantz.github.io/collapse -> fastverse.github.io/collapse.

@@
 \end{Schunk}
-More features of \fct{pivot} are demonstrated in the \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}.
+More features of \fct{pivot} are demonstrated in the \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}.
 Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.

### 2025-12-09 misc/JSS article/final/article.Rnw
Replace sebkrantz.github.io/collapse -> fastverse.github.io/collapse.

@@
 %
-\href{https://sebkrantz.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://sebkrantz.github.io/collapse/.
+\href{https://fastverse.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://fastverse.github.io/collapse/.
 Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844.
@@
 The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
-\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
+\pkg{collapse} is \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
 It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications of the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
@@
 
-The package comes with a built-in structured \href{https://sebkrantz.github.io/collapse/reference/index.html}{documentation} facilitating its use.
-This includes a central \href{https://sebkrantz.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and a set of supplementary topic pages which briefly summarize related functionality.
+The package comes with a built-in structured \href{https://fastverse.github.io/collapse/reference/index.html}{documentation} facilitating its use.
+This includes a central \href{https://fastverse.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and a set of supplementary topic pages which briefly summarize related functionality.
 The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:
@@
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
+For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
 %
@@
 %
-The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R}/\pkg{collapse} means that these operations are implemented using compiled \proglang{C/C++} code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
+The \href{https://fastverse.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R}/\pkg{collapse} means that these operations are implemented using compiled \proglang{C/C++} code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
 Their basic syntax is
@@
 The following examples with \fct{fmean} demonstrate their basic usage on the familiar \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/iris}{\code{iris}} dataset recording 50 measurements of 4 variables for 3 species of iris flowers.\footnote{\fct{num\_vars} returns the numeric variables in a data frame-like object; \fct{cat\_vars} the categorical ones.} All examples support weights (\code{w}), and \fct{fmean} can also be multithreaded across columns (\code{nthreads}).\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions, as detailed in the respective function documentation.
-The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes.
+The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes.
 Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods.
@@
 It notably avoids the need to convert atomic objects to data frames for grouped aggregations or transformations.
-The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://sebkrantz.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
+The \href{https://fastverse.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://fastverse.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
 <<BY>>=
@@
 %
-\pkg{collapse} also provides a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc.
+\pkg{collapse} also provides a broad set of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc.
 as optimized analogues to base \proglang{R}/\pkg{tidyverse} functions.
 These are integrated with the \emph{Fast Statistical and Transformation Functions} to enable vectorized operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow.
-I illustrate this using the included World Development Dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.
+I illustrate this using the included World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.
 <<wlddev_distinct>>=
@@
 \pkg{collapse} also vectorizes advanced statistics.
-The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} computes weighted quantiles in a theoretically consistent way, see \href{https://sebkrantz.github.io/collapse/reference/fquantile.html}{fquantile} for details.}
+The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} computes weighted quantiles in a theoretically consistent way, see \href{https://fastverse.github.io/collapse/reference/fquantile.html}{fquantile} for details.}
 %
@@
 \end{Code}
- Below, I create a \class{GRP} object from two columns in the World Development Dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}).
+ Below, I create a \class{GRP} object from two columns in the World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}).
 The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.
@@
 
-The lower-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
+The lower-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://fastverse.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
 Users should note that \pkg{collapse} does not provide metaprogramming capabilities in its non-standard evaluation functions---such as \href{https://rlang.r-lib.org/reference/topic-quosure.html}{quosures} or \href{https://dplyr.tidyverse.org/articles/programming.html}{indirection} familiar to \pkg{tidyverse} users.
@@
 Any data frame-like object can become an \class{indexed\_frame} and behave as usual for other operations.
-The technical implementation of these classes is described in the \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{documentation}.
+The technical implementation of these classes is described in the \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://fastverse.github.io/collapse/reference/indexing.html}{documentation}.
 Their basic usage is:
@@
 Unlike \class{pdata.frame}, an \\ \class{indexed\_frame} is a deeply indexed structure---every series inside the frame is already an \class{indexes\_series}.
-A comprehensive set of \href{https://sebkrantz.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()} ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.).
+A comprehensive set of \href{https://fastverse.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()} ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.).
 % A basic demonstration with World Bank panel data showcases the flexibility of these classes.
@@
 %
-Among its suite of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://sebkrantz.github.io/collapse/reference/join.html}{joins} and \href{https://sebkrantz.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots.
+Among its suite of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://fastverse.github.io/collapse/reference/join.html}{joins} and \href{https://fastverse.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots.
 Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
@@
 %
-The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://sebkrantz.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America.
+The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://fastverse.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America.
 While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes.
@@
 @
-The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
+The \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
 Notably, it can also perform longer and recast pivots without id variables, similar to \code{data.table::transpose()}.
@@
 Except for certain recursive functions in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data.
-\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit.
+\pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit.
 \newline
@@
 %
-\pkg{collapse}'s \href{https://sebkrantz.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets.
+\pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets.
 A particular focus has been on providing tools to understand multilevel (panel) data.
-Recall the World Development panel dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{sec:integration}.
+Recall the World Development panel dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{sec:integration}.
 The function \fct{varying} can be used to examine which of the variables are time-varying.
@@
 \fct{descr} also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden.
-The \href{https://sebkrantz.github.io/collapse/reference/descr.html}{documentation} provides more details and examples.
+The \href{https://fastverse.github.io/collapse/reference/descr.html}{documentation} provides more details and examples.
 Both \fct{qsu} and \fct{descr} come with an \code{as.data.frame()} method for efficient tidying and easy further analysis.
@@
 %
-\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default values of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function.
+\pkg{collapse} is \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default values of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function.
 These options are saved in an internal environment called \code{.op}.
@@
 Special functions are \code{"n"}, \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} for use in \code{(f)summarise}/\code{(f)mutate}, \code{table <- qtab}, and replace \code{\%in\%} with a faster version based on \fct{fmatch}, respectively.
-There also exist several \href{https://sebkrantz.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}, such as \code{"manip"} (only data manipulation functions), or \code{"all"} (everything), as demonstrated below.
+There also exist several \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}, such as \code{"manip"} (only data manipulation functions), or \code{"all"} (everything), as demonstrated below.
 %
@@
 
-Such global powers confer responsibilities upon package developers, as further elucidated in the \href{https://sebkrantz.github.io/collapse/articles/developing_with_collapse.html#some-notes-on-global-options}{vignette on developing with \pkg{collapse}}.
+Such global powers confer responsibilities upon package developers, as further elucidated in the \href{https://fastverse.github.io/collapse/articles/developing_with_collapse.html#some-notes-on-global-options}{vignette on developing with \pkg{collapse}}.
 As a general rule, options \code{mask} and \code{remove} should be off-limits inside packages, and other options need to be reset immediately using \fct{on.exit}.

### 2025-12-09 misc/JSS article/final/article.tex
Replace sebkrantz.github.io/collapse -> fastverse.github.io/collapse.

@@
 It notably avoids the need to convert atomic objects to data frames for grouped aggregations or transformations.
-The \href{https://sebkrantz.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://sebkrantz.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
+The \href{https://fastverse.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://fastverse.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
 \begin{Schunk}
@@
 %
-\pkg{collapse} also provides a broad set of \href{https://sebkrantz.github.io/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc.
+\pkg{collapse} also provides a broad set of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc.
 as optimized analogues to base \proglang{R}/\pkg{tidyverse} functions.
 These are integrated with the \emph{Fast Statistical and Transformation Functions} to enable vectorized operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow.
-I illustrate this using the included World Development Dataset (\href{https://sebkrantz.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.
+I illustrate this using the included World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.
 \begin{Schunk}
@@
 \end{Schunk}
-The \href{https://sebkrantz.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
+The \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
 Notably, it can also perform longer and recast pivots without id variables, similar to \code{data.table::transpose()}.

### 2025-12-09 misc/JSS article/final/jss5278/jss5278.tex
Replace sebkrantz.github.io/collapse -> fastverse.github.io/collapse.

@@
 The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
-\pkg{collapse} is \href{https://sebkrantz.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames, and seamlessly supporting popular extensions to these objects such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
+\pkg{collapse} is \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames, and seamlessly supporting popular extensions to these objects such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
 It is also extensively globally and interactively configurable.
@@
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://sebkrantz.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://sebkrantz.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
+For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
 %

### 2025-12-09 misc/collapse cheat sheet/collapse_cheat_sheet.Rnw
Replace sebkrantz.github.io/collapse -> fastverse.github.io/collapse.

@@
 
-Calling \code{help("collapse-documentation")} brings up a detailed documentation, which is also available \href{https://sebkrantz.github.io/collapse/reference/index.html}{online}.
+Calling \code{help("collapse-documentation")} brings up a detailed documentation, which is also available \href{https://fastverse.github.io/collapse/reference/index.html}{online}.
 See also the \href{https://fastverse.github.io/fastverse/}{\emph{fastverse}} package/project for a recommended set of complimentary packages and easy package management.
@@
 # Grouped linear models (one way of doing it)
-qTBL(mtcars) |> fgroup_by(vs) |> fsummarise(reg = list(lm(disp ~ drat)))
+qTBL(mtcars) |> fgroup_by(vs) |> fsummarise(reg = list(lm(mpg ~ carb)))
 # Adding some columns. Use ftransform<- to also replace existing ones
@@
 {\scriptsize \vspace{-0.5mm}
- Page 1 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://sebkrantz.github.io/collapse/}{sebkrantz.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.0\ \textbullet\ Updated: 2023-10
+ Page 1 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
 }
@@
 
-\code{any|all[v|NA]}, \code{which[v|NA]}, \code{\%[=|!]=\%}, \code{copyv}, \code{setv}, \code{alloc} \code{missing\_cases}, \code{na\_[insert|rm|omit]}, \code{vlengths}, \code{vtypes}, \code{vgcd}, \code{frange}, \code{fnlevels}, \code{fn[row|col]}, \code{fdim}, \code{seq\_[row|col]}\\
+\code{any|all[v|NA]}, \code{which[v|NA]}, \code{\%[=|!]=\%}, \code{copyv}, \code{setv}, \code{alloc} \code{missing\_cases}, \code{na\_[insert|rm|omit]}, \code{vlengths}, \code{vtypes}, \code{vgcd}, \code{fnlevels}, \code{fn[row|col]}, \code{fdim}, \code{seq\_[row|col]}, \code{vec}\\
 <<eval = FALSE>>=
@@
 \itxt{Functions for (meta-)programming and attributes}\\ [0.5em]
-\code{.c}, \code{massign}, \code{\%=\%}, \code{vlabels[<-]}, \code{setLabels}, \code{vclasses}, \code{namlab}, \code{[add|rm]\_stub}, \code{\%!in\%}, \code{ckmatch}, \code{all\_identical}, \code{all\_obj\_equal}, \code{all\_funs}, \code{set[Dim|Row|Col]names}, \code{unattrib}, \code{setAttrib}, \code{copyAttrib}, \code{copyMostAttrib} %, \code{is\_categorical}, \code{is\_date}
+\code{.c}, \code{massign}, \code{\%=\%}, \code{vlabels[<-]}, \code{setLabels}, \code{vclasses}, \code{namlab}, \code{[add|rm]\_stub}, \code{all\_identical}, \code{all\_obj\_equal}, \code{all\_funs}, \code{set[Dim|Row|Col]names}, \code{unattrib}, \code{setAttrib}, \code{copyAttrib}, \code{copyMostAttrib}, \code{is\_categorical}, \code{is\_date}
 
@@
 \hrrule
-\section{API Extensions}
+\section{API Extensions and Global Options}
 \itxt{Shorthands for frequently used functions}\\ [0.5em]
@@
 
-\itxt{Namespace masking}\\ [0.5em]
-Can set \code{option(collpse\_mask = c(...))} with a vector of functions starting with f-, to export versions without f-, masking base R or \emph{dplyr}.
+\itxt{Namespace masking and other global options}\\ [0.5em]
+Use \code{set\_collpse(mask = c(...))} with a vector of functions starting with f-, to export versions without f-, masking base R and/or \emph{dplyr}.
 A few keywords exist to mask multiple functions, see \code{help("collapse-options")}.
-This allows clean \& fast code, but poses additional namespace challenges:
+There are also many other global defaults and optimizations that can be controlled with \code{set\_collapse(...)}.
+Retrieve options using \code{get\_collapse()}.
 
 <<eval = FALSE>>=
-# Masking all f- functions and specials n = GRPN and table = qtab
-options(collapse_mask = "all")
+# Masking all (f-)functions and changing some defaults (=optimizing)
 library(collapse)
-# The folowing is 100% collapse code, apart from the base pipe
-
+set_collapse(mask = "all", na.rm = FALSE, sort = FALSE, nthreads = 4)
+# The following is now 100% collapse code and executed without regard for
+# missing values, using unsorted grouping and 4 threads (where applicable)
 wlddev |>
-  subset(year >= 1990) |>
+  subset(year >= 1990 & is.finite(GINI)) |>
   group_by(year) |>
@@
 @
-The best way to set this option is inside an \code{.Rprofile} file placed in the user or project directory. Use it carefully.
 
@@
 {\scriptsize \vspace{-0.5mm}
- Page 2 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://sebkrantz.github.io/collapse/}{sebkrantz.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 1.8.8\ \textbullet\ Updated: 2022-08
+ Page 2 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
 }

### 2025-12-09 misc/JSS article/R&R/article.Rnw
Rename SebKrantz/collapse -> fastverse/collapse.

@@
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} (\href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}).
+For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} (\href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}).
 %
@@
 Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices.
-For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} I aggregate global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}.\footnote{A recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case) - see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
+For example, in the \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} I aggregate global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object and expressions of the form \code{x |> fsum(g) |> t() |> fsum(g) |> t()}.\footnote{A recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case) - see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
 %On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another recent example involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}.
@@
 @
-% Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
+% Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
 %

### 2025-12-09 misc/JSS article/R&R/article.tex
Rename SebKrantz/collapse -> fastverse/collapse.

@@
 \end{Schunk}
-% Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
+% Last but not least, the performance of these classes is second to none.\footnote{See the \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation} slide 40 for a small benchmark.}
 %

### 2025-12-09 R/GRP.R

@@ -404,7 +404,7 @@ fgroup_by <- function(.X, ..., sort = .op[["sort"]], decreasing = FALSE, na.last
     e <- eval(dots, .X, parent.frame())
     name <- names(e)
     vars <- all.vars(dots, unique = FALSE)
-    # If something else than NSE cols is supplied, see https://github.com/SebKrantz/collapse/issues/320
+    # If something else than NSE cols is supplied, see https://github.com/fastverse/collapse/issues/320
     # Note: doesn't support fgroup_by(mtcars, cyl / vs), but ok, this should be named...
     # fgroup_by(mtcars, c("cyl", "vs")) gives vars == character(0)
     if(length(e) == 1L && is.null(name) && (length(vars) != 1L || !anyv(names(.X), vars))) { # !is.symbol(dots[[2L]]) || length(e[[1L]]) != length(.X[[1L]]) || is.function(e[[1L]] # Fixes #320

### 2025-12-09 tests/testthat/test-misc.R

@@ -395,7 +395,7 @@ test_that("rowbind", {
 })
 
 if (requireNamespace("bit64", quietly = TRUE)) test_that("rowbind + integer64", {
-  # https://github.com/SebKrantz/collapse/issues/697
+  # https://github.com/fastverse/collapse/issues/697
   x <- data.frame(a = bit64::as.integer64(1))
   xi <- data.frame(a = 1L)
   xd <- data.frame(a = 1)

### 2025-12-09 misc/JSS article/final/article.Rnw

@@
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
+For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
 %
@@
 Grouped programming using 'GRP' objects and \emph{Fast Statistical Functions} is also particularly powerful with vectors and matrices.
-For example, in the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions like \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 45 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
+For example, in the \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}, I aggregate 32 global input-output tables stored as matrices (\code{x}) from the country to the region level using a single grouping object (\code{g}) and expressions like \code{x |> fsum(g) |> t() |> fsum(g) |> t()}---computing 45 million sums crunching 5.7GB of data in $\sim$0.3 seconds on my M1.\footnote{Another recent application with vectors involved numerically optimizing a parameter $a$ in an equation of the form $\sum_i x_{ij}^a\ \forall j\in J$ so as to minimize the deviation from a target $y_j$ where there are $J$ groups (1 million in my case)---see the first example in \href{https://sebkrantz.github.io/Rblog/2023/04/12/collapse-and-the-fastverse-reflecting-the-past-present-and-future/}{this blog post} for an illustration.}
 %On an M1 Mac using 4 threads, this computation, involving 44.7 million summations and 2.6 million weighted means, takes only 0.33 seconds.\footnote{Another recent example involved numerically optimizing a parameter $a$ in an equation of the form $y_j = \sum_i x_{ij}^a\ \forall j\in J$ where there are $J$ groups (1 million in my case), and the optimal value of $a$ is determined by the proximity of the aggregated vector \textbf{y} to another vector \textbf{z}.
@@
 %
-Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
+Last but not least, the computational performance of these classes is second to none.\footnote{See, e.g., the small benchmark presented on slide 40 of the \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{useR 2022 presentation}.}
 %

### 2025-12-09 misc/JSS article/final/jss5278/jss5278.tex

@@
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/SebKrantz/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
+For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
 %

### 2025-12-09 misc/collapse cheat sheet/collapse_cheat_sheet.Rnw

@@
 {\scriptsize \vspace{-0.5mm}
- Page 1 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
+ Page 1 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/fastverse/collapse}{github.com/fastverse/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
 }
@@
 {\scriptsize \vspace{-0.5mm}
- Page 2 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/SebKrantz/collapse}{github.com/SebKrantz/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
+ Page 2 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/fastverse/collapse}{github.com/fastverse/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
 }

### 2025-12-21 misc/JSS article/R&R/article.Rnw
Update package website url to fastverse.org/collapse.

@@
 %
-\href{https://fastverse.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://fastverse.github.io/collapse/.
+\href{https://fastverse.org/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://fastverse.org/collapse/.
 Linecount (v2.0.19): R: 13,646, C: 18,594, C++: 9,844.
@@
 The package also provides efficient data object conversions, functions for memory efficient R programming, such as (grouped) transformation and math by reference, and helpers to effectively deal with variable labels, attributes, and missing data.
-\pkg{collapse} is \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem---notably, \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
+\pkg{collapse} is \href{https://fastverse.org/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem---notably, \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
 It is globally and interactively configurable, which includes setting different defaults for key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifying the package namespace to mask equivalent but slower base \proglang{R} or \pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
@@
 
-The package has a built-in structured \href{https://fastverse.github.io/collapse/reference/collapse-documentation.html}{documentation} facilitating its use.
-This documentation includes a central \href{https://fastverse.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and supplementary topic pages which briefly describe related functionality.
+The package has a built-in structured \href{https://fastverse.org/collapse/reference/collapse-documentation.html}{documentation} facilitating its use.
+This documentation includes a central \href{https://fastverse.org/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and supplementary topic pages which briefly describe related functionality.
 The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:
@@
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation and resources} (\href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}).
+For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.org/collapse/articles/collapse_documentation.html}{documentation and resources} (\href{https://fastverse.org/collapse/articles/index.html}{vignettes}/\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}/\href{https://sebkrantz.github.io/Rblog/}{blog}/\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}/\href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}).
 %
@@
 %
-The \href{https://fastverse.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R} means that these operations are implemented using compiled C/C++ code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
+The \href{https://fastverse.org/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R} means that these operations are implemented using compiled C/C++ code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
 The functions basic syntax is
@@
 All examples support weights (\code{w}), and \fct{fmean} can also be multithreaded across columns (\code{nthreads}).\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions, as detailed in the respective function documentation.
-The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes.
+The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://fastverse.org/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes.
 Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods.
@@
 \end{Code}
- Below, I create a \class{GRP} object from the included World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}).
+ Below, I create a \class{GRP} object from the included World Development Dataset (\href{https://fastverse.org/collapse/reference/wlddev.html}{\code{wlddev}}).
 The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.
@@
 
-Exempting \fct{collap}, these examples may seem bulky for quick analysis, but a robust low-level API is very useful for package development, as further elucidated in the \href{https://fastverse.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
+Exempting \fct{collap}, these examples may seem bulky for quick analysis, but a robust low-level API is very useful for package development, as further elucidated in the \href{https://fastverse.org/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
 I also wrote a blog post on \href{https://sebkrantz.github.io/Rblog/2021/01/09/advanced-data-aggregation/}{aggregating survey data using \pkg{collapse}}, which showcases more aspects of the \fct{collap} function using real census data.
@@
 %
-\pkg{collapse} also provides a broad set of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{fast data manipulation functions} akin to base \proglang{R} and \pkg{tidyverse} functions, including \fct{fselect}, \fct{fsubset}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{ftransform}, \fct{fmutate}, \fct{across}, \fct{frename}, \fct{fcount}, etc.
+\pkg{collapse} also provides a broad set of \href{https://fastverse.org/collapse/reference/fast-data-manipulation.html}{fast data manipulation functions} akin to base \proglang{R} and \pkg{tidyverse} functions, including \fct{fselect}, \fct{fsubset}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{ftransform}, \fct{fmutate}, \fct{across}, \fct{frename}, \fct{fcount}, etc.
 These are integrated with the \emph{Fast Statistical Functions} to enable vectorized statistical operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow.
@@
 The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} calculates weighted quantiles in a theoretically consistent way by applying the probability measure to the sum of weights to create a target sum and cumulatively summing weights to find order statistics which are then combined following \citet{hyndman1996sample}.
-See \href{https://fastverse.github.io/collapse/reference/fquantile.html}{fquantile} for details.}
+See \href{https://fastverse.org/collapse/reference/fquantile.html}{fquantile} for details.}
 %
@@
 Any data frame-like object can become an \class{indexed\_frame} and function as usual for other operations.
-The technical implementation of these classes is described in the \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://fastverse.github.io/collapse/reference/indexing.html}{documentation}.
+The technical implementation of these classes is described in the \href{https://fastverse.org/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://fastverse.org/collapse/reference/indexing.html}{documentation}.
 Their basic syntax is:
@@
 Unlike \class{pdata.frame}, an \\ \class{indexed\_frame} is a deeply indexed structure---every series inside the frame is already an \class{indexes\_series}.
-A comprehensive set of \href{https://fastverse.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()}, ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.).
+A comprehensive set of \href{https://fastverse.org/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()}, ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.).
 % A basic demonstration with World Bank panel data showcases the flexibility of these classes.
@@
 %
-Among all \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://fastverse.github.io/collapse/reference/join.html}{joins} and \href{https://fastverse.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and 'recast' pivots.
+Among all \href{https://fastverse.org/collapse/reference/fast-data-manipulation.html}{data manipulation functions} \pkg{collapse} provides, its implementations of table \href{https://fastverse.org/collapse/reference/join.html}{joins} and \href{https://fastverse.org/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and 'recast' pivots.
 Both implementations provide outstanding computational performance, syntax, and memory efficiency.
@@
 %
-The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Databas (\href{https://fastverse.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}), providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America.
+The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Databas (\href{https://fastverse.org/collapse/reference/GGDC10S.html}{\code{GGDC10S}}), providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America.
 While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes.
@@
 @
-More features of \fct{pivot} are demonstrated in the \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}.
+More features of \fct{pivot} are demonstrated in the \href{https://fastverse.org/collapse/reference/pivot.html#ref-examples}{documentation examples}.
 Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.
@@
 Except for certain recursive functions found in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data.
-\pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit.
+\pkg{collapse}'s \href{https://fastverse.org/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit.
 \newline
@@
 %
-\pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious and powerful toolkit to examine complex datasets.
+\pkg{collapse}'s \href{https://fastverse.org/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious and powerful toolkit to examine complex datasets.
 A particular focus has been on providing tools for examining longitudinal (panel) data.
-Recall the indexed World Development Panel (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{ssec:gopt}.
+Recall the indexed World Development Panel (\href{https://fastverse.org/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{ssec:gopt}.
 The function \fct{varying} can be used to examine which of these variables are time-varying.
@@
 A \code{stepwise} argument toggles describing one variable at a time, allowing users to naturally 'click-through' a large dataset rather than printing a massive output to the console.
-The \href{https://fastverse.github.io/collapse/reference/descr.html}{documentation} provides more details and examples.
+The \href{https://fastverse.org/collapse/reference/descr.html}{documentation} provides more details and examples.
 Both \fct{qsu} and \fct{descr} provide an \code{as.data.frame()} method for efficient tidying and further analysis.
@@
 %
-\pkg{collapse} is \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function.
+\pkg{collapse} is \href{https://fastverse.org/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default value of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function.
 These options are saved in an internal environment called \code{.op}.
@@
 Special functions are \code{"n"}, \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} (for use in \code{(f)summarise}/\code{(f)mutate}), \code{table <- qtab}, and replace \code{\%in\%} with a fast version using \fct{fmatch}, respectively.
-There are also several \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}.
+There are also several \href{https://fastverse.org/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}.
 The most powerful of these is \code{"all"}, which masks all f-functions and special functions, as shown below.
@@
 Option \code{remove} can further be used to remove (un-export) any \pkg{collapse} function, allowing manual conflict management.
-Function \code{fastverse::fastverse\_conflicts()} from the related \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse} project} \citep{rfastverse} can be used to display namespace conflicts with \pkg{collapse}.
+Function \code{fastverse::fastverse\_conflicts()} from the related \href{https://fastverse.org/fastverse/}{\pkg{fastverse} project} \citep{rfastverse} can be used to display namespace conflicts with \pkg{collapse}.
 Invoking either \code{mask} or \code{remove} detaches \pkg{collapse} and re-attaches it at the top of the search path, letting its namespace to take precedence over other packages.
@@
 Since June 2024, there is also an independent \href{https://github.com/AdrianAntico/Benchmarks?tab=readme-ov-file#background}{database-like operations benchmark by Adrian Antico} using a windows server and executing scripts inside IDEs (VScode, Rstudio), on which \pkg{collapse} achieved the overall fastest runtimes.
-I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.github.io/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages.
+I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.org/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages.
 These benchmarks align in showing that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and on windows systems.\footnote{Reasons for the particularly strong performance of \pkg{collapse} on Windows may be that it is largely written in C and has limited multithreading in favor or highly efficient serial algorithms---there appear to be persistent obstacles to efficient (low-overhead) multithreading on Windows, implying that multithreaded query engines do not develop their full thrust on medium-sized ($\leq$100 million row) datasets.}
@@
 
-For deeper engagement with \pkg{collapse}, visit its \href{https://fastverse.github.io/collapse/index.html}{website} or start with the vignette summarizing all available \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation and resources}.
+For deeper engagement with \pkg{collapse}, visit its \href{https://fastverse.org/collapse/index.html}{website} or start with the vignette summarizing all available \href{https://fastverse.org/collapse/articles/collapse_documentation.html}{documentation and resources}.
 Users can also follow \pkg{collapse} on \href{https://x.com/collapse_R}{Twitter/X} and \href{https://bsky.app/profile/rcollapse.bsky.social}{Bluesky} to be notified about major updates and participate in community discussions.

### 2025-12-21 misc/JSS article/final/article.Rnw
Update package website url to fastverse.org/collapse.

@@
 %
-\href{https://fastverse.github.io/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://fastverse.github.io/collapse/.
+\href{https://fastverse.org/collapse/}{\pkg{collapse}} is a large \proglang{C/C++}-based \proglang{R} package that provides an integrated suite of statistical and data manipulation functions.\footnote{Website: https://fastverse.org/collapse/.
 Linecount (v2.1.2): \proglang{R}: 13,785, \proglang{C}: 19,013, \proglang{C++}: 9,844.
@@
 The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
-\pkg{collapse} is \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
+\pkg{collapse} is \href{https://fastverse.org/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames/lists, and seamlessly supporting extensions to these objects popular in the \proglang{R} ecosystem such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
 It is globally and interactively configurable, allowing changes to defaults of key function arguments, such as \code{na.rm} arguments to statistical functions or \code{sort} arguments to grouping algorithms (default \code{TRUE}), and modifications of the namespace to mask/replace equivalent but less performant base \proglang{R}/\pkg{tidyverse} functions.\footnote{\pkg{collapse}'s namespace is fully compatible with base \proglang{R} and the \pkg{tidyverse} \citep{rtidyverse}, but can be interactively modified to mask/overwrite key functions with the much faster \pkg{collapse} equivalents.
@@
 
-The package comes with a built-in structured \href{https://fastverse.github.io/collapse/reference/index.html}{documentation} facilitating its use.
-This includes a central \href{https://fastverse.github.io/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and a set of supplementary topic pages which briefly summarize related functionality.
+The package comes with a built-in structured \href{https://fastverse.org/collapse/reference/index.html}{documentation} facilitating its use.
+This includes a central \href{https://fastverse.org/collapse/reference/collapse-documentation.html}{overview page} linking to all other documentation pages and a set of supplementary topic pages which briefly summarize related functionality.
 The names of these extra pages are collected in a global macro \code{.COLLAPSE\_TOPICS} and can be called directly with \code{help()}:
@@
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
+For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.org/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.org/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk}$+$\href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
 %
@@
 %
-The \href{https://fastverse.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R}/\pkg{collapse} means that these operations are implemented using compiled \proglang{C/C++} code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
+The \href{https://fastverse.org/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}}, comprising \fct{fsum}, \fct{fprod}, \fct{fmean}, \fct{fmedian}, \fct{fmode}, \fct{fvar}, \fct{fsd}, \fct{fmin}, \fct{fmax}, \fct{fnth}, \fct{ffirst}, \fct{flast}, \fct{fnobs}, and \fct{fndistinct}, are a consistent set of S3-generic statistical functions providing fully vectorized statistical operations in R.\footnote{'Vectorization' in \proglang{R}/\pkg{collapse} means that these operations are implemented using compiled \proglang{C/C++} code.} Specifically, operations are vectorized across columns and groups, and may also involve weights or transformations of the input data.
 Their basic syntax is
@@
 The following examples with \fct{fmean} demonstrate their basic usage on the familiar \href{https://www.rdocumentation.org/packages/datasets/versions/3.6.2/topics/iris}{\code{iris}} dataset recording 50 measurements of 4 variables for 3 species of iris flowers.\footnote{\fct{num\_vars} returns the numeric variables in a data frame-like object; \fct{cat\_vars} the categorical ones.} All examples support weights (\code{w}), and \fct{fmean} can also be multithreaded across columns (\code{nthreads}).\footnote{Not all functions are multithreaded, and parallelism is implemented differently for different functions, as detailed in the respective function documentation.
-The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes.
+The default use of single instruction multiple data (SIMD) parallelism also implies limited gains from multithreading for simple (non-grouped) operations.}  % As laid out in the \href{https://fastverse.org/collapse/articles/collapse_object_handling.html}{vignette on object handling}, statistical functions have basis S3 methods for vectors (\class{default}), \class{matrix}, and \class{data.frame}, which call corresponding \proglang{C} implementations that intelligently preserve object attributes.
 Thus, the functions can be applied to a broad set of \class{matrix} or \class{data.frame}-based objects without the need to define explicit methods.
@@
 It notably avoids the need to convert atomic objects to data frames for grouped aggregations or transformations.
-The \href{https://fastverse.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://fastverse.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
+The \href{https://fastverse.org/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://fastverse.org/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
 <<BY>>=
@@
 %
-\pkg{collapse} also provides a broad set of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc.
+\pkg{collapse} also provides a broad set of \href{https://fastverse.org/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc.
 as optimized analogues to base \proglang{R}/\pkg{tidyverse} functions.
 These are integrated with the \emph{Fast Statistical and Transformation Functions} to enable vectorized operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow.
-I illustrate this using the included World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.
+I illustrate this using the included World Development Dataset (\href{https://fastverse.org/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.
 <<wlddev_distinct>>=
@@
 \pkg{collapse} also vectorizes advanced statistics.
-The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} computes weighted quantiles in a theoretically consistent way, see \href{https://fastverse.github.io/collapse/reference/fquantile.html}{fquantile} for details.}
+The following calculates a weighted set of summary statistics by groups, with weighted quantiles type 8 following \citet{hyndman1996sample}.\footnote{\pkg{collapse} computes weighted quantiles in a theoretically consistent way, see \href{https://fastverse.org/collapse/reference/fquantile.html}{fquantile} for details.}
 %
@@
 \end{Code}
- Below, I create a \class{GRP} object from two columns in the World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}).
+ Below, I create a \class{GRP} object from two columns in the World Development Dataset (\href{https://fastverse.org/collapse/reference/wlddev.html}{\code{wlddev}}).
 The \code{by} argument also supports column names/indices, and \code{X} could also be an atomic vector.
@@
 
-The lower-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://fastverse.github.io/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
+The lower-level API is useful for package development and standard-evaluation programming, as further elucidated in the \href{https://fastverse.org/collapse/articles/developing_with_collapse.html}{vignette on developing with \pkg{collapse}}.
 Users should note that \pkg{collapse} does not provide metaprogramming capabilities in its non-standard evaluation functions---such as \href{https://rlang.r-lib.org/reference/topic-quosure.html}{quosures} or \href{https://dplyr.tidyverse.org/articles/programming.html}{indirection} familiar to \pkg{tidyverse} users.
@@
 Any data frame-like object can become an \class{indexed\_frame} and behave as usual for other operations.
-The technical implementation of these classes is described in the \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://fastverse.github.io/collapse/reference/indexing.html}{documentation}.
+The technical implementation of these classes is described in the \href{https://fastverse.org/collapse/articles/collapse_object_handling.html#class-agnostic-grouped-and-indexed-data-frames}{vignette on object handling} and, in more detail, in the \href{https://fastverse.org/collapse/reference/indexing.html}{documentation}.
 Their basic usage is:
@@
 Unlike \class{pdata.frame}, an \\ \class{indexed\_frame} is a deeply indexed structure---every series inside the frame is already an \class{indexes\_series}.
-A comprehensive set of \href{https://fastverse.github.io/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()} ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.).
+A comprehensive set of \href{https://fastverse.org/collapse/reference/indexing.html}{methods for subsetting and manipulation}, and applicable \class{pseries} and \class{pdata.frame} methods for time series and transformation functions like \code{flag()/L()} ensure that these objects behave in a time-/panel-aware manner in any caller environment (\fct{with}, \fct{lm}, etc.).
 % A basic demonstration with World Bank panel data showcases the flexibility of these classes.
@@
 %
-Among its suite of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://fastverse.github.io/collapse/reference/join.html}{joins} and \href{https://fastverse.github.io/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots.
+Among its suite of \href{https://fastverse.org/collapse/reference/fast-data-manipulation.html}{data manipulation functions}, \pkg{collapse}'s implementations of table \href{https://fastverse.org/collapse/reference/join.html}{joins} and \href{https://fastverse.org/collapse/reference/pivot.html}{pivots} are particularly noteworthy since they offer several new features, including rich verbosity for table joins, pivots supporting variable labels, and recast pivots.
 Both implementations also offer outstanding computational performance, syntax, and memory efficiency.
@@
 %
-The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://fastverse.github.io/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America.
+The demonstration below employs the included Groningen Growth and Development Centre 10-Sector Database (\href{https://fastverse.org/collapse/reference/GGDC10S.html}{\code{GGDC10S}}) providing long-run internationally comparable data on sectoral productivity performance in Africa, Asia, and Latin America.
 While the database covers 10 sectors, for the demonstration I only retain Agriculture, Mining, and Manufacturing.\footnote{The \code{"Label"} column is added for demonstration purposes.
@@
 @
-The \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
+The \href{https://fastverse.org/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
 Notably, it can also perform longer and recast pivots without id variables, similar to \code{data.table::transpose()}.
@@
 Except for certain recursive functions in packages such as \pkg{purr}, \pkg{tidyr}, or \pkg{rrapply}, \proglang{R} lacks a general recursive toolkit to create, query, and tidy nested data.
-\pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit.
+\pkg{collapse}'s \href{https://fastverse.org/collapse/reference/list-processing.html}{list processing functions} attempt to provide a basic toolkit.
 \newline
@@
 %
-\pkg{collapse}'s \href{https://fastverse.github.io/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets.
+\pkg{collapse}'s \href{https://fastverse.org/collapse/reference/summary-statistics.html}{summary statistics functions} offer a parsimonious yet powerful toolkit to examine complex datasets.
 A particular focus has been on providing tools to understand multilevel (panel) data.
-Recall the World Development panel dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{sec:integration}.
+Recall the World Development panel dataset (\href{https://fastverse.org/collapse/reference/wlddev.html}{\code{wlddev}}) from Section~\ref{sec:integration}.
 The function \fct{varying} can be used to examine which of the variables are time-varying.
@@
 \fct{descr} also has a \code{stepwise} argument to describe one variable at a time, allowing users to naturally 'step-through' the variables in a large dataset while spreading the computational burden.
-The \href{https://fastverse.github.io/collapse/reference/descr.html}{documentation} provides more details and examples.
+The \href{https://fastverse.org/collapse/reference/descr.html}{documentation} provides more details and examples.
 Both \fct{qsu} and \fct{descr} come with an \code{as.data.frame()} method for efficient tidying and easy further analysis.
@@
 %
-\pkg{collapse} is \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default values of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function.
+\pkg{collapse} is \href{https://fastverse.org/collapse/reference/collapse-options.html}{globally configurable} to an extent few packages are: the default values of key function arguments governing the behavior of its algorithms, and the exported namespace, can be adjusted interactively through the \fct{set\_collapse} function.
 These options are saved in an internal environment called \code{.op}.
@@
 Special functions are \code{"n"}, \code{"table"/"qtab"}, and \code{"\%in\%"}, which create \code{n <- GRPN} for use in \code{(f)summarise}/\code{(f)mutate}, \code{table <- qtab}, and replace \code{\%in\%} with a faster version based on \fct{fmatch}, respectively.
-There also exist several \href{https://fastverse.github.io/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}, such as \code{"manip"} (only data manipulation functions), or \code{"all"} (everything), as demonstrated below.
+There also exist several \href{https://fastverse.org/collapse/reference/collapse-options.html}{convenience keywords to mask related groups of functions}, such as \code{"manip"} (only data manipulation functions), or \code{"all"} (everything), as demonstrated below.
 %
@@
 
-Such global powers confer responsibilities upon package developers, as further elucidated in the \href{https://fastverse.github.io/collapse/articles/developing_with_collapse.html#some-notes-on-global-options}{vignette on developing with \pkg{collapse}}.
+Such global powers confer responsibilities upon package developers, as further elucidated in the \href{https://fastverse.org/collapse/articles/developing_with_collapse.html#some-notes-on-global-options}{vignette on developing with \pkg{collapse}}.
 As a general rule, options \code{mask} and \code{remove} should be off-limits inside packages, and other options need to be reset immediately using \fct{on.exit}.
@@
 Since June 2024, there is also an independent \href{https://github.com/AdrianAntico/Benchmarks?tab=readme-ov-file#background}{database-like operations benchmark by Adrian Antico} using a windows server and executing scripts inside IDEs (VScode, Rstudio), on which \pkg{collapse} achieved the overall fastest runtimes.
-I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.github.io/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages.
+I also very recently started a \href{https://github.com/fastverse/fastverse/wiki/Benchmarks}{user-contributed benchmark Wiki} as part of the \href{https://fastverse.org/fastverse/}{fastverse project} promoting high-performance software for R, where users can freely contribute benchmarks involving, but not limited to, \pkg{fastverse} packages.
 These benchmarks agree that \pkg{collapse} offers a computationally outstanding experience, particularly for medium-sized datasets, complex tasks, and on users PCs/Macs---which typically have smaller memory and parallel computing resources but faster chips than servers.
@@
 
-Finally, \pkg{collapse} users are also encouraged to familiarize themselves with the \href{https://fastverse.github.io/fastverse/}{\pkg{fastverse}},\footnote{Website: https://fastverse.github.io/fastverse/} a suite of complementary high-performance packages for statistical computing and data manipulation in \proglang{R} that offer more advanced tools in several statistical computing domains.
+Finally, \pkg{collapse} users are also encouraged to familiarize themselves with the \href{https://fastverse.org/fastverse/}{\pkg{fastverse}},\footnote{Website: https://fastverse.org/fastverse/} a suite of complementary high-performance packages for statistical computing and data manipulation in \proglang{R} that offer more advanced tools in several statistical computing domains.
 The \pkg{fastverse} metapackage additionally provides a lightweight framework to jointly load and manage these packages, as well as to build customized and fully separate package verses.

### 2025-12-21 misc/collapse cheat sheet/collapse_cheat_sheet.Rnw
Update package website url to fastverse.org/collapse.

@@
 
-Calling \code{help("collapse-documentation")} brings up a detailed documentation, which is also available \href{https://fastverse.github.io/collapse/reference/index.html}{online}.
-See also the \href{https://fastverse.github.io/fastverse/}{\emph{fastverse}} package/project for a recommended set of complimentary packages and easy package management.
+Calling \code{help("collapse-documentation")} brings up a detailed documentation, which is also available \href{https://fastverse.org/collapse/reference/index.html}{online}.
+See also the \href{https://fastverse.org/fastverse/}{\emph{fastverse}} package/project for a recommended set of complimentary packages and easy package management.
 %}
@@
 {\scriptsize \vspace{-0.5mm}
- Page 1 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/fastverse/collapse}{github.com/fastverse/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
+ Page 1 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.org/collapse/}{fastverse.org/collapse}\ \textbullet\ Source code at \href{https://github.com/fastverse/collapse}{github.com/fastverse/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
 }
@@
 {\scriptsize \vspace{-0.5mm}
- Page 2 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.github.io/collapse/}{fastverse.github.io/collapse}\ \textbullet\ Source code at \href{https://github.com/fastverse/collapse}{github.com/fastverse/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
+ Page 2 of 2 \hfill \href{https://creativecommons.org/licenses/by-sa/4.0/}{CC-BY-SA}\ Sebastian Krantz\ \textbullet\ Learn more at \href{https://fastverse.org/collapse/}{fastverse.org/collapse}\ \textbullet\ Source code at \href{https://github.com/fastverse/collapse}{github.com/fastverse/collapse}\ \textbullet\ Updates announced at \href{https://twitter.com/collapse\_R}{twitter.com/collapse\_R} - \#rcollapse\ \textbullet\ Cheatsheet created for \emph{collapse} version 2.0.3\ \textbullet\ Updated: 2023-10
 }

### 2025-12-22 misc/JSS article/R&R/article.tex

@@
 \end{Schunk}
-More features of \fct{pivot} are demonstrated in the \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples}.
+More features of \fct{pivot} are demonstrated in the \href{https://fastverse.org/collapse/reference/pivot.html#ref-examples}{documentation examples}.
 Notably, it can also perform longer and recast pivots without id variables, like \code{data.table::transpose()}.

### 2025-12-22 misc/JSS article/final/article.tex

@@
 It notably avoids the need to convert atomic objects to data frames for grouped aggregations or transformations.
-The \href{https://fastverse.github.io/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://fastverse.github.io/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
+The \href{https://fastverse.org/collapse/reference/fast-statistical-functions.html}{\emph{Fast Statistical Functions}} are complemented by a smaller set of \href{https://fastverse.org/collapse/reference/data-transformations.html}{\emph{Data Transformation Functions}}, including \fct{TRA}, infix functions such as \code{\%r+\%}, \code{\%c+\%}, \code{\%+=\%} and \fct{setop} for row/column-wise arithmetic operations (by reference), \fct{dapply} to apply functions across rows or columns of matrices/data frames, \fct{BY} for general split-apply-combine computing, and specialized functions such as \fct{fscale} or \fct{f[hd]within} for (grouped, weighted) scaling and centering:
 \begin{Schunk}
@@
 %
-\pkg{collapse} also provides a broad set of \href{https://fastverse.github.io/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc.
+\pkg{collapse} also provides a broad set of \href{https://fastverse.org/collapse/reference/fast-data-manipulation.html}{Fast Data Manipulation Functions}, including \fct{fselect}, \fct{fsubset}, \fct{fslice}, \fct{fgroup\_by}, \fct{fsummarise}, \fct{fmutate}, \fct{frename}, \fct{fcount}, etc.
 as optimized analogues to base \proglang{R}/\pkg{tidyverse} functions.
 These are integrated with the \emph{Fast Statistical and Transformation Functions} to enable vectorized operations in a familiar data frame oriented and \pkg{tidyverse}-like workflow.
-I illustrate this using the included World Development Dataset (\href{https://fastverse.github.io/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.
+I illustrate this using the included World Development Dataset (\href{https://fastverse.org/collapse/reference/wlddev.html}{\code{wlddev}}) recording five key indicators for 216 economies and years 1960-2020.
 \begin{Schunk}
@@
 \end{Schunk}
-The \href{https://fastverse.github.io/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
+The \href{https://fastverse.org/collapse/reference/pivot.html#ref-examples}{documentation examples} demonstrate more features of \fct{pivot}.
 Notably, it can also perform longer and recast pivots without id variables, similar to \code{data.table::transpose()}.

### 2025-12-22 misc/JSS article/final/jss5278/jss5278.tex

@@
 The package also includes efficient object converters, functions for memory efficient \proglang{R} programming, such as (grouped) transformation and math by reference, and helper functions to handle variable labels, attributes, and missing data.
-\pkg{collapse} is \href{https://fastverse.github.io/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames, and seamlessly supporting popular extensions to these objects such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
+\pkg{collapse} is \href{https://fastverse.org/collapse/articles/collapse_object_handling.html}{class-agnostic}, providing statistical operations on vectors, matrices, and data frames, and seamlessly supporting popular extensions to these objects such as \class{units}, \class{integer64}, \class{xts}, \class{tibble}, \class{data.table}, \class{sf}, and \class{pdata.frame}.
 It is also extensively globally and interactively configurable.
@@
 Section~\ref{sec:bench} provides a small benchmark, Section~\ref{sec:conclusion} concludes.
-For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.github.io/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.github.io/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
+For deeper engagement with \pkg{collapse}, consult the \href{https://fastverse.org/collapse/articles/collapse_documentation.html}{documentation resources}, including \href{https://fastverse.org/collapse/articles/index.html}{vignettes}, \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/collapse\%20cheat\%20sheet/collapse_cheat_sheet.pdf}{cheatsheet}, \href{https://sebkrantz.github.io/Rblog/}{blog}, and \href{https://www.youtube.com/watch?v=OwWT1-dSEts}{talk} with \href{https://raw.githubusercontent.com/fastverse/collapse/master/misc/useR2022\%20presentation/collapse_useR2022_final.pdf}{slides}.
 %
